### 2019-01-20 paper.tex
update

\documentclass[11pt,letterpaper]{article}

\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{amssymb}

\usepackage{tikz-dependency}
\usepackage{longtable}

\newcommand{\R}[0]{\mathbb{R}}
\newcommand{\E}[0]{\mathbb{E}}
\newcommand{\Ff}[0]{\mathcal{F}}

\usepackage{multirow}

\newcommand{\soft}[1]{}
\newcommand{\nopreview}[1]{}
\newcommand\comment[1]{{\color{red}#1}}
\newcommand\mhahn[1]{{\color{red}(#1)}}

\usepackage{amsthm}

\newcommand{\thetad}[0]{{\theta_d}}
\newcommand{\thetal}[0]{{\theta_{LM}}}

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{thm}[theorem]{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}

\frenchspacing
%\def\baselinestretch{0.975}

%\emnlpfinalcopy
%\def\emnlppaperid{496}

\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
\author{Michael Hahn}
\date{2018}

\begin{document}

\maketitle

%
\begin{abstract}

Are languages optimized for processing with limited memory?
Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners. %assuming only that listeners perform incremental prediction.
Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners. % producing language and listeners predicting input.
\end{abstract}

%
%
%\begin{thm}\label{prop:suboptimal}
%	For each positive integer $t$, define
%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
%bits of memory on average.
%Then this comprehender will incur average surprisal at least
%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
%\end{thm}
%

\section{Introduction}

A wide range of work has argued that natural language orders information in ways that reduce memory effort.
An early example is \cite{miller-finitary-1963}, who attributed the unaccebtability of multiple center embeddings in English to limitations of human working memory.
\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.

%TODO mention prior work
%
%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
%
%- explanation of greenberg universals
%
%- dependency length minimization
%
%- \cite{gibson-linguistic-1998}
%

%\paragraph{\cite{berwick-grammatical-1986}}
%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.

%also (CITE): operations in a specific incremental parsing model

All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.

We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
We will consider a general setting of a speaker producing sentences from a language, and a listener performing incremental prediction.
Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
Unlike many previous theories, our results apply to both speakers and listeners.
We will then use corpus data from over 50 languages to provide evidence that their word orders help lower memory cost.

\section{Two Memory-Surprisal Tradeoffs}

With minimal assumptions, we will use information theory to derive two tradeoffs:

(1) a tradeoff between \emph{speaker memory} and (listener) surprisal

(2) a tradeoff between \emph{listener memory} and (listener) surprisal

In the second part of the paper, we examine whether word orders in natural language optimise these two tradeoffs.

\subsection{Memory and Locality in General Sequence Models}

We consider a \emph{speaker} who produces sentences and a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
Informally, we expect that both speaker and listener will be affected by memory demands:
From the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
From the listener's perspective, predicting the next word well requires maintaining information about the past.
For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.

We now make this more formal:
We think of the language as a stochastic process $(X_t)_{t \in \mathbb{Z}}$, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
Second, we make the assumption that this process is \emph{stationary}.
Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.

The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
\begin{equation}\label{eq:markov-speaker}
X_{>t} \bot X_{\leq t} | S_t
\end{equation}
This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.

What can we say about the amount of information that the speaker has to store when producing utterances -- that is, how many bits does it take to encode the speaker's  memory states $S_t$ and $L_t$?
Our first theoretical result sheds light on this; we will show that the amount of bits needed to encode the memory state is related to the decay of mutual information over distances.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}
\end{figure*}

We will use the concept of \emph{conditional mutual information}.
For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
\begin{equation}
	I[X,Y] := H[X] - H[X|Y]
\end{equation}
For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
\begin{equation}
I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
\end{equation}
We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
\begin{equation}
	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
\end{equation}
This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.

Using this concept, we can now state our first main result, concerning the memory load of a speaker: %, and of a listener performing optimal incremental prediction:

\begin{proposition}\label{prop:lower-bound}
	A speaker producing samples from $(X_t)_{t \in \mathbb{Z}}$ needs to maintain at least
\begin{equation}\label{eq:memory-bound}
M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}
bits of memory on average.

\end{proposition}

Also can relax this to imprecise production: A speaker with low memory must deviate strongly from the language, measured in KL divergence.

%that the model does not have an `absolute clock' built into it.
The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).

Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
This formalizes a general, assumption-free, link between memory and locality in language production.
In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
%incremental prediction.

The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model. % and the mechanisms by which it computes predictions.
In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
Memory representations do not have to be rational or optimal for this bound to hold:
It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.

We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
%This confirms the intuition that finiteness of memory entails that the contribution of long-term

We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
The analysis is not even restricted to discrete timesteps: We can replace the sum in~(\ref{eq:memory-bound}) with an integral to get a continuous-time version.

We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
However, $I_t$ decays more quickly for Process A (red).
This means that predictive information about an observation is concentrated more strongly in the recent past.
In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
Note that the area under the curve is equal to (\ref{eq:memory-bound}).
This area is smaller for the red process, as $I_t$ decays more quickly there.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
%
	\caption{Left: $I_t$ as a function of $t$, for two different processes. $I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past. Left: $t \cdot I_t$ as a function of $t$ for the same processes. The area under the curves is given by (\ref{eq:memory-bound}). This area is smaller for the red process, as $I_t$ decays more quickly there.   }\label{fig:basic}
\end{figure*}

\subsection{Memory and Dependency Length}

We illustrate the linguistic predictions of Proposition \ref{prop:lower-bound} by reanalyzing the data from \cite{fedzechkina-human-2017}.
This is a miniature artificial language study that showed a bias for Dependency Length Minimization in production in artificial language learning.
Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in~(\ref{eq:memory-bound}).

As ~(\ref{eq:memory-bound}) is invariant under reversal of the language, we only consider the head-final version of her artificial language.
The language has consistent head-final order, and uses case marking on objects.
The relevant production targets are transitive sentences where one of the two arguments is much longer than the other, due to the presence of a PP modifier, as shown in Table~\ref{tab:artificial}.
The language has variable order of subjects and objects; for the production targets, the B versions produce much longer dependencies than the A versions.
Dependency Length Minimization thus predicts that speakers are more likely to use the A versions.
\cite{fedzechkina-human-2017} confirmed this experimentally.

\begin{table}
	\textbf{A Orders: Short Dependencies}

		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun\end{tabular}} & \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & Verb \\
		\end{tabular}
\\
\\

	\textbf{B Orders: Long Dependencies}

		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			 \fbox{\begin{tabular}{l}Noun\end{tabular}} &  \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
		\end{tabular}}  &  \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
		\end{tabular}} & Verb \\
		\end{tabular}

			\caption{Production targets in the artificial mini language from \cite{fedzechkina-human-2017}. The language has head-final order, with free variation between SO and OS orders. When one of the arguments is much longer than the other, placing the longer one first (`A' orders) shortens syntactic dependencies, compared to `B' orders.}\label{tab:artificial}

\end{table}

In this section, we show that our bound on speaker memory makes the same prediction, without reference to syntactic structure or specific memory architectures.

We constructed one language consisting of the A versions, and one language consisting of the B versions.
Following the experimental setup of \cite{fedzechkina-human-2017}, we assigned equal probability to the two possible configurations per language, and used a separate set of nouns (inanimate nouns) for the embedded noun in the long phrase.

We interpreted each of the two languages as a stationary processes, extending infinitely in both directions, by concatenating independent samples drawn from the language.
			We computed (\ref{eq:memory-bound}) from a chain of 1000 independently sampled sentences, for each of the two versions of the toy language.
			Figure~\ref{fig:toy-mis} (left) shows the curve of the conditional mutual information $I_t$ as a function of the distance $t$.
			The curves differ at $t=2$ and $t=5$: 
			About 0.073 nats of predictive information that are at distance $t=2$ in the A orders are moved to $t=5$ in the B orders.
%			In this sense, A orders have greater information locality than the B orders.
			%What is responsible for this difference?
			The source of the difference lies in predicting the presence and absence of a case marker on the second argument -- i.e., whether to anticipate a subject or object.
			In the A orders, considering the last two words is sufficient to make this decision.
			In the B orders, it is necessary to consider the word before the long second constituent, which is five words in the past.

			The total amounts of predictive information -- corresponding to the area under the curve -- are the same, indicating that both languages are equally predictable.
			However, we will see that the memory demands are different:
			Figure~\ref{fig:toy-mis} (right) shows $t\cdot I_t$ as a function of $t$.
			As $I_t$ decays faster in A orders, the total area under the curve now differs between A and B, and is larger in B.
			This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.
			While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.
It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/toy-mis.pdf}
\includegraphics[width=0.45\textwidth]{toy/toy-t-mis.pdf}
%
	\caption{Left: Decay of Conditional Mutual Information, as a function of the distance $t$, for the two versions in the artificial language. The areas under the two curves are identical, corresponding to the fact that both orders are equally predictable. However, mutual Information decays faster in Order A.\ \ \ Right: $t I_t$, as a function of $t$. The area under the B curve is larger, corresponding to larger memory demand for this order.}\label{fig:toy-mis}
\end{figure*}
%Both languages have the same overall entropy rate, but they differ in the distribution of predictive information.
%
%plot of $I_t$
%
%The areas under the curves are identical.
%
%good version
%
%%CONTEXT LENGTH 0   2.06856758681  9997.93143241   0.0
%%CONTEXT LENGTH 1   0.29195106479  1.77661652202   1.77661652202
%%CONTEXT LENGTH 2   0.0729865489103  0.21896451588   2.21454555377
%%CONTEXT LENGTH 3   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 4   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 5   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 6   0.0729865489103  0.0   2.21454555377
%
%bad version
%
%%CONTEXT LENGTH 0   2.06937301571  9997.93062698   0.0
%%CONTEXT LENGTH 1   0.291673373027  1.77769964269   1.77769964269
%%CONTEXT LENGTH 2   0.145830550934  0.145842822093   2.06938528687
%%CONTEXT LENGTH 3   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 4   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 5   0.0729152754672  0.0729152754672   2.43396166421
%%CONTEXT LENGTH 6   0.0729152754672  0.0   2.43396166421
%%CONTEXT LENGTH 7   0.0729152754672  0.0   2.43396166421
%%
%
%
%
%%
%%grammar:
%%
%%S $\rightarrow$ Obj Subj V (1/2) | Subj Obj V (1/2)
%%
%%Obj $\rightarrow$ NP di
%%
%%Subj $\rightarrow$ NP
%%
%%NP $\rightarrow$ N (3/4) | PP NP (1/8) | Adj NP (1/8)
%%
%%PP $\rightarrow$ NP P
%%
%
%

\subsection{Languages can trade off Memory Load and Predictability}
%\subsection{Speakers can can minimize Memory Load at Fixed Information Rate}

We have derived a general lower bound on the memory load of a speaker in Proposition~\ref{prop:lower-bound}.
We now relate the speaker's memory to the surprisal experienced by the listener.

The average surprisal of a language $(X_t)_{t \in \mathbb{Z}}$, as experienced by a listener performing optimal prediction, is equal to the entropy rate $H[X_t|X_{<t}]$.
This can be written as the unigram entropy minus the mutual information between a single observation $X_t$ and the past:
\begin{equation}
	H[X_t|X_{<t}] = H[X_t] - I[X_t, X_{<t}]
\end{equation}
where the unigram entropy $H[X_t]$ only depends on the frequencies of individual words, independently of the language's word order.

We will refer to the mutual information $I[X_t, X_{<t}]$ as the \emph{predictability} of the process.
The predictability $P := I[X_t, X_{<t}]$ is a lower bound on $M = I[X_{\geq t}, X_{<t}]$, and thus on speaker memory:
\begin{equation}
P \leq M
\end{equation}
and thus\footnote{More generally 
$H[X_t|X_{<t}] \geq \frac{1}{T} \left(H[X_{1\dots T}] - M\right)$ for any $T$. Thus, the following relation holds for any $T$: $$T \cdot \text{ListenerSurprisal} \geq H[X_{1\dots T}] - \text{SpeakerMemory}$$}
\begin{equation}
	Listener\ Surprisal \geq	H[X_t|X_{<t}] \geq H[X_t] - M \geq H[X_t] - Speaker\ Memory
\end{equation}
This means that, when we consider different languages that share a single unigram distribution, memory and surprisal trade off:
Languages that are highly predictable  require a good amount of speaker memory; a language that can be produced with very small memory cannot be very predictable.
In particular, a language that requires zero speaker memory must have zero predictability and maximal average surprisal.

We thus obtain a tradeoff between two different types of costs facing speakers and listeners: The \emph{speaker's} memory trades off with the \emph{listeners'} surprisal.

We can visualize this by placing languages into a \emph{memory-surprisal-plane} (Figure~\ref{fig:toy-mis}). Each language is assigned its entropy rate as its $y$ coordinate, and the memory load necessary to achieve optimal surprisal as its $x$ coordinate.
We will continue to assume that all languages share a single unigram distribution -- this will in particular apply when studying different word order variations of the same language.

The plane can be separated into an \emph{achievable} region, populated by languages, and an \emph{unachievable} region in which no languages can occur.
No language can have zero surprisal with zero memory -- the corner of the plane thus must fall in the unachievable region.
The boundary between the two regions is the \emph{Pareto frontier}.
Languages that fall on the boundary optimally trade off surprisal and memory; they cannot further decrease one without increasing the other.
The smallest possible values for this boundary consist of the points satisfying $P=M$, indicated by a straight diagonal line in Figure~\ref{fig:toy-mis}.
Only part of the achievable region will be populated by languages that could plausibly be natural languages: For instance, we may expect natural languages to have long-term dependencies that will prevent them from being at the absolute Pareto frontier.
We hypothesize that, if languages optimize the tradeoff between memory and surprisal, they will be close to the optimal boundary of a region populated by a `comparison class' of languages that share essential characteristics of natural language, but have other word order rules.
%When We will specify a class of baseline languages that we will argue to form a reasonable comparison class for natural language.
%Within this, we hypothesize that natural language is close to the Pareto frontier.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/pareto-scheme.pdf}
%
	\caption{Memory-Surprisal Tradeoff. The straight line represents the Pareto frontier for unconstrained languages. Languages resembling natural languages inhabit a subset of the achievable region (grey oval). Its lower-left boundary represents the Pareto frontier relevant for natural language. We hypothesize that natural language lies close to this frontier.   }\label{fig:toy-mis}
\end{figure}

\subsection{Listeners can trade off Memory Load and Surprisal}\label{sec:listener}

%TODO redo curves so that x-axis is surprisal, y-axis the necessary memory

We have analyzed the memory load of a speaker and related it to the listener's surprisal.
We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
Similar to the listener maintains a memory state $L_t$; again, there are no assumptions about the memory architecture and the nature of its computations.
%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
We only make two basic assumptions about the flow of information:

First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
Formally: 
\begin{equation}\label{eq:listener-markov-1}
L_t \bot X_{>t} | X_{\leq t}
\end{equation}
This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.

Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
\begin{equation}\label{eq:listener-markov-2}
L_t \bot X_{<t} | X_{t-1}, L_{t-1}
\end{equation}
This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.

The listener can trade off memory and future surprisal:
A listener who chooses to store less memory will exerience higher surprisal in the future.
A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.

We will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
Consider a listener who uses $J$ bits of memory on average.
What can we say about the listener's surprisal?
We return to the information and memory curves in Figure~\ref{fig:basic}.
In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
Then

\begin{proposition}\label{prop:suboptimal}
	Let $T$ be a positive integer, and consider a listener using at most
$$\sum_{t=1}^T t I_t$$
bits of memory on average.
Then this listener will incur surprisal at least
	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
	on average.
\end{proposition}

The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$, we obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
	\caption{Illustration for Proposition~\ref{prop:suboptimal}. Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal. In the given example, $T=4$. By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}. Recall that the red process had a faster decay of conditional mutual information. Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
\end{figure}

In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
For reaching optimal surprisal, Order A requires strictly less memory.
Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}. Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
\end{figure}

\paragraph{Center Embeddings}

\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.

%\cite{gibson-linguistic-1998}

%first center-embeddings, POS level model

%Later work found evidence that cross-serial dependencies are less taxing.

%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies

\subsection{Relation to Theories of Sentence Processing}

\paragraph{Lossy-Context Surprisal}
\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
In particular, they consider loss models that delete, erase, or replace words in the past.
Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
A listener will incur surprisal
$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.

\paragraph{Relation to Dependency Locality}
The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.

\paragraph{Center Embeddings, ...}
Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures

Miller and Chomsky 1963

Frzier 1985 local nonterminal count

Rambow and Joshi 1994 using TAG

Marcus 1980 deterministic parsing

(Sabrina Gerth, Memory Limitations in Sentence Processing)

\subsection{Relation to other work}

\paragraph{Statistical Complexity}
There are deep connections between our formalization of speaker and listener memory and studies of dynamic systems in the Physics literature.
Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
The quantity in (\ref{eq:memory-bound}) is equal to the \emph{excess entropy}, and it has long been known to bound statistical and generative complexity \citep{crutchfield-inferring-1989}.
However, the link between memory and locality provided by our series expansion in (\ref{eq:memory-bound}) appears to be a novel contribution.

\cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.

\paragraph{Decay of Mutual Information}
In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.

The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$. %%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR. This one has bounded nonzero $J$, but zero unconditional MI. To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal. The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}

\paragraph{Long-range dependencies in text}    % excess entropy
\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
Our work contrasts with this work in that we are interested in dependencies within sentences.

\subsection{Discussion}

\paragraph{Speakers vs Listeners}
Our results apply both to speakers and listeners.

\paragraph{Decay vs Interference}
Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
This is not the case:
A long dependency can be maintained perfectly with low average memory:
Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.

\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
In fact, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory (and thus in particular finite excess entropy):

\begin{proposition}
Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
\end{proposition}

\begin{proof}[Proof Sketch]
	We provide an algorithm that performs generation or optimal prediction and has finite average memory use. We consider incremental Earley parsing \citep{earley-efficient-1970}. Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$). For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
\end{proof}

\section{Heavy NP Shift}

Phenomenon linked to locality/memory (Hawkins)

Is the listener's memory-surprisal tradeoff optimized by Heavy NP Shift?

Penn Treebank

Trained 20 language models counterfactual version

then estimate memory on real and reverse orderings

3) Describe the key dependent variable(s) specifying how they will be measured.
Collect all instances of a VP consisting of a verb, and (in some order) an NP and a PP, from the Penn Treebank (WSJ) training set. Calculate the listener's memory-surprisal tradeoff on the words belonging to the NP and the PP.

5) Specify exactly which analyses you will conduct to examine the main question/hypothesis.
(1) Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs? Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP? This will be tested by bootstrapping over language models and examples.
(2) Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?
Also tested by bootstrapping over language models and examples.

6) Describe exactly how outliers will be defined and handled, and your precise rule(s) for excluding observations.
Language models with overall validation average surprisal > 10 will be excluded (this didn't happen).

7) How many observations will be collected or what will determine sample size?
20 LSTM language models are constructed on a version of Penn Treebank where for all verb-initial VPs, the order of the remaining daughter constituents is shuffled.

Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs?

Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP?

Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?

%
%
%\section{NP Structure}
%
%- adjective order
%
%- noted by ...: Det - Num - Adj
%
%
%This is a case where the predictions differ from DepL
%
%
%E.g., if the choice of determiners, numerals, and adjectives are statistically independent when conditioning on the noun, will predict the effect.
%Computationally verify the effect.

\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}

We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
We will define a class of baseline languages that serve as a comparison class of word orders that natural language can be compared to.

Ordering grammars are defined as in the Optimization paper: For each UD relation type, define a weight. To get random baseline languages, randomize the weights.
All grammars are valuated deterministically.

We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
Thus, random grammars with by-relation weights constitute a stronger baseline.

%Things that one may ponder: (1) what are even stronger baselines? (2) probably should do a control experiment using maximum-likelihood fits

%\subsection{Prediction on the level of POS}
%As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences. %, following~\cite{futrell-noisy-context-2017}.
%As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.
%
%%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
%%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
%%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
%%
%%%\mhahn{TODO results across languages}
%%
%%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
%
%Results for four languages are shown in Figure~\ref{fig:memory-results}.
%Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
%%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
%
%
%%%For half of the languages
%%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
%%For 37 \% of the languages, it also requires less memory than the original language.
%
%%We found that, as predicted, memory usage in real languages was correlated with word order freedom:
%
%
%%\begin{figure*}
%%\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%%%
%%\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
%%\end{figure*}
%%

\subsection{Method}
The small size of available syntactic data for many languages means that estimating mutual information is problematic.

In general, we expect a bias-variance tradeoff:

A method that estimates on a coarse-grained level (e.g. POS) will result in estimates with relatively higher fidelity and lower variance, potentially making it easier to localize the actual language in the space of counterfactual languages.
However, it introduces \emph{bias} by describing a simplified version of the language.

We propose to therefore compute the quantities on levels less likely to suffer from data sparseness: part-of-speech and morphology.

In the UD treebanks, each word has part-of-speech and morphological information.
The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token that indicates that information related to the word ends.
We further provide the word's lemma after its POS tag, if it is one of the 50 most frequent lemmas of the language.
Each POS tag and morphological feature-value pair is encoded as an atomic token.
The rationale for including frequent lemmas is that such words often are strongly grammaticalized and may provide grammatical information (and frequent words are less likely to be affected by data sparseness).

To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling sequences such as natural language.
We will discuss alternative estimation methods below.

During incremental prediction, the LSTM sequentially reads these tokens and computes a distribution over the next word.
The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).

Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.

Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
Therefore, to compute this quantity, we compute the differences between the surprisals
$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.
We cut $T$ off at 30.

\paragraph{Alternative Models}
Recent research has proven strong theoretical learnability guarantees for neural network models (CITE), supporting their success in practice.
They also provide the strongest predictors of the surprisal effect on human reading times \citep{frank-insensitivity-2011, goodkind-predictive-2018}.

In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:

First, an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations. Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent. Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.

A statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar. The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model. One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations. The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.

\subsection{Data}
We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
We excluded four languages for a later confirmatory experiment (English, Korean, Russian). % Irish
This resulted in X languages.

For most language, we used the predefined data split.
For some languages, where there was no training partition, or where the training partition was smaller than the test partition, we redefined the split to obtain more training data:
For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}

\subsection{Setup}
The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
This biases the hyperparameters towards favoring counterfactual grammars.

This experiment is not fully confirmatory, since some data was collected for some languages before the predictions and the setup of the experiment were finalized.

For each language, we collected data from the actual orderings and from random grammars.
We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
We collected data from the actual and random orderings in proportion one to three.
The stopping criterion will be described below.

\paragraph{Speaker Memory: Distance from Pareto Frontier}
We want to test: Are languages closer to the Pareto frontier than baseline languages tend to be?

For this, we need a way of quantifying distance to the Pareto frontier.
We suggest to measure the distance as the fraction of baseline languages that are better on both dimensions.

More formally, fixing a natural language, let $P_0$ be the distribution of baseline languages mapped to the surprisal-memory plane, viewed as a probability distribution over the positive quadrant of the plane $\mathbb{R}_+^2$.
For any point $x \in \mathbb{R}_+^2$, let $A_x$ be the fraction of baseline languages that are better on both dimensions.
Similarly, let $B_x$ be the fraction of baseline languages that are worse on both dimensions.
Using the probability distribution $P_0$, we can write:
$$A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$$
$$B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$$
%Ways of using this to test and quantify for distance (I think they are all reasonable and worth reporting in SI. Question is which ones to rely on mostly for the main part):
Futher, let $P_1$ be the distribution in the surprisal-memory plane that is generated by the results obtained for the real orderings, with different random initializations.
For each language, we \emph{test} the hypothesis that the language optimizes the tradeoff by considering the null hypothesis that the real orderings do not Pareto-dominate more baseline orderings than there are baseline orderings dominating them:
$$\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$$

We propose to \emph{quantify} closeness to the frontier using the following two statistics:

\begin{enumerate}
%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$

		This has the following interpretation: Among orderings that either dominate or are dominated by the real orderings, what fraction is dominated by the real orerings?
		This quantity is $1$ if the real orderings are not Pareto-dominated by any baseline order, but do dominate some baseline orders.
		The null hypothesis defined in our test amounts to the inequality $\E_{x \sim P_1}E_x \leq 0.5$.

A drawback of this quantity is that it is meaningless when the real orders do not dominate, and are not dominated by, any real order.
The following quantity is always maningful:

	\item Compute $A_{x'}$ for each $x'$ from the baseline distribution $P_0$, and then calculate the percentile rank of $\E_{x \sim P_1}[A_x]$ among these values:
		$$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$$
This statistic describes how many random orders achieve closer distance to the frontier as measured by $A_x$.
		Stated differently, it quantifies how `extraordinary' the value $\E_{x \sim P_1}[A_x]$ is in relation to values typically attained for $x' \sim P_0$.
The statistic is close to zero if $\E_{x \sim P_1}[A_x]$ is smaller than for almost all samples from $P_0$.

\end{enumerate}

We will obtain confidence intervals for the statistics via bootstrapping, and use a confidence interval for $B_x-A_x$ for testing $B_x \leq A_x$.
In the appendix, we report simulations that justify the use of bootstrapping in this situation.

\paragraph{Listener Memory: Surprisal-Memory Tradeoff}
We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
For each model for each language, we interpolate the surprisal-memory tradeoff curves given by Proposition~\ref{prop:suboptimal} linearly.
For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.

We test by considering the null hypothesis that $\E_{x \sim P_1}[W(x)] \leq 1/2$ where $P_1$ is the distribution over values obtained for real orderings.

We use a bootstrapped confidence interval for $\E[W]$ for testing and for quantifying the degree of optimization.

%- bootstrapping
%- subsampling
%- permutation test / rank test ??
%

\paragraph{Stopping Criterion}

For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $D$ had width $\leq 0.3$, and the CI for $W$ had width $\leq 0.15$.
We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.

%- at least 30 samples from both baseline and real
%
%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.

%
%(1) is bigram MI always greater in real languages?
%
%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?

\subsection{Results}

Results are reported in Appendix~\ref{sec:results-table}.

\paragraph{Surprisal and Speaker Memory}
In all but X languages, the null hypothesis $\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$ was rejected.
Also, in all but X languages, $D < 0.5$.

\paragraph{Surprisal and Listener Memory}
In all but X languages $\E[W] > 0.5$ significantly.

\section{Experiments on Further Treebanks}

Held-out:

- UD languages:

-- English, Korean, Russian

-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment)

-- character-level Russian

- other treebanks

-- Penn treebank \citep{marcus-building-1993}

-- spoken English (T{\"u}ba-E/S)

-- spoken German (T{\"u}ba-D/S)

-- spoken Japanese (T{\"u}ba-J/S)

-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017}

-- Chinese treebank \citep{xue-chinese-2013}

-- furher (2.2/2.3 Japanese treebanks)

Due to the sizes of these treebanks, can also do experiment with full word forms.

\bibliographystyle{apalike}
\bibliography{literature}

\appendix

\section{Proofs}

\begin{proof}
	The number of bits of memory is given by the entropy $H[S_t]$.
	In view of (\ref{eq:markov-speaker}) and the Data Processing Inequality, this is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
	I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I[X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}] \\
	&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I[X_i, X_j|X_{j+1}...X_{i-1}] \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
where we used stationarity of the process in the last step.
\end{proof}

\begin{proof}
We denote the listener's memory state after hearing $X_{<t}$ by $L_t$.
The average number of bits required to encode this state is $H[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $H[X_t | L_t]$.
The difference between the listener's surprisal and optimal surprisal is thus at least $H[X_t | L_t] - H[X_t | X_{<t}]$.
By stationarity, we can rewrite this expression as
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | L_t] - H[X_t | X_{<t}]\right) 
\end{align*}
For any $t$,
	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots T}, L_0]$$
	This can be derived from the two conditions \ref{eq:listener-markov-1} and \ref{eq:listener-markov-2} (TODO explain how).

Therefore
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq   \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
\end{align*}
	The first term $I[X_{1\dots T}|X_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
\end{align*}
	$I[X_{1\dots T}|L_0]$ is at most $H[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
	\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
		&= \sum_{t > T} I_t
\end{align*}
	Rearranging shows that the listener's surprisal is at least $H[X_t|L_t] \geq H[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
\end{proof}

\section{Bootstrapping for Distance to Frontier}

TODO write up

We conducted simulations in the following setting:

sample two ellipses such that

(1) $\mu_1 \sim Unif([-0.5, 0.5]^2)$

(2) $\mu_2 \sim Unif([-1.5, 1.5]^2)$

$A_1, A_2 \sim Unif(\mathbb{R}^2 \times \mathbb{R}^2)$, $\Sigma_i = A_i A_i^T$ subject to the wo eigenvalues differing by a factor of $\leq 2$, and $\Sigma_1$ having area 1, $\Sigma_2$ having area 1/20.

Computed ground-truth $A_x, B_x$ from 3000 samples from both ellipses.

Bootstrapping: get 0.001, 0.999

Results:

For $A_x-B_x$: good coverage

For $E_x$: intervals are unreliable if $A_x + B_x$ is small

For $D_x$: good coverage

\section{Results (Morphology on UD Treebanks)}\label{sec:results-table}

Statistics reported:

\begin{enumerate}
%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
	\item $A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$
	\item $B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$
	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$
	\item Compute $A_{x'}$ for each $x'$ in the baseline distribution, and then calculate the percentile rank of $A_x$ among these values:
$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$
\item For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.

\end{enumerate}

For each language, we show the number of sentences in the training partition, the surprisal-speaker memory tradeoff plane, the listener's surprisal-memory tradeoff curves (smoothed means), and the statistics defined above with bootstraped CIs.

Languages where the stopping criterion hasn't been reached are marked with asterisks.

\begin{longtable}{lllllllllllllll}
Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.28]  \\ 
1315  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
  &    &    &  $E_x$  &  0.42  &  [0.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.99  &  [0.96, 1.0]  \\ [10.25ex] \hline
Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.16  &  [0.06, 0.29]  \\ 
974  &    &    &  $B_x-A_x$  &  0.1  &  [0.04, 0.18]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Arabic*  &\\ [10.25ex] \hline
Armenian-Adap*  &\\ [10.25ex] \hline
Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.03, 0.15]  \\ 
5396  &    &    &  $B_x-A_x$  &  0.82  &  [0.76, 0.88]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.03, 0.22]  \\ 
788  &    &    &  $B_x-A_x$  &  0.32  &  [0.2, 0.42]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.22]  \\ 
8907  &    &    &  $B_x-A_x$  &  0.56  &  [0.44, 0.7]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.02, 0.19]  \\ 
808  &    &    &  $B_x-A_x$  &  0.28  &  [0.19, 0.42]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.08, 0.31]  \\ 
550  &    &    &  $B_x-A_x$  &  0.25  &  [0.17, 0.36]  \\ 
  &    &    &  $E_x$  &  0.92  &  [0.85, 0.98]  \\ 
  &    &    &  $W_x$  &  0.94  &  [0.86, 1.0]  \\ [10.25ex] \hline
Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.03, 0.25]  \\ 
13123  &    &    &  $B_x-A_x$  &  0.44  &  [0.33, 0.57]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.04, 0.23]  \\ 
3997  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.19]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.03, 0.23]  \\ 
7689  &    &    &  $B_x-A_x$  &  0.13  &  [0.05, 0.23]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.05, 0.27]  \\ 
102993  &    &    &  $B_x-A_x$  &  0.66  &  [0.56, 0.77]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.94  &  [0.86, 0.99]  \\ [10.25ex] \hline
Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.26]  \\ 
4383  &    &    &  $B_x-A_x$  &  0.15  &  [0.06, 0.25]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.01, 0.15]  \\ 
18310  &    &    &  $B_x-A_x$  &  0.04  &  [0.01, 0.08]  \\ 
  &    &    &  $E_x$  &  0.88  &  [0.73, 1.0]  \\ 
  &    &    &  $W_x$  &  0.98  &  [0.96, 1.0]  \\ [10.25ex] \hline
Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.48]  \\ 
6959  &    &    &  $B_x-A_x$  &  0.26  &  [0.18, 0.36]  \\ 
  &    &    &  $E_x$  &  0.76  &  [0.67, 0.84]  \\ 
  &    &    &  $W_x$  &  0.72  &  [0.65, 0.79]  \\ [10.25ex] \hline
Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.26]  \\ 
1108  &    &    &  $B_x-A_x$  &  0.34  &  [0.24, 0.45]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.01, 0.12]  \\ 
27198  &    &    &  $B_x-A_x$  &  0.95  &  [0.91, 0.99]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
French*  &\\ [10.25ex] \hline
German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.44  &  [0.37, 0.51]  \\ 
13814  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.29]  \\ 
  &    &    &  $E_x$  &  0.66  &  [0.6, 0.74]  \\ 
  &    &    &  $W_x$  &  0.63  &  [0.57, 0.69]  \\ [10.25ex] \hline
Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.2]  \\ 
1662  &    &    &  $B_x-A_x$  &  0.59  &  [0.48, 0.73]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\ 
5241  &    &    &  $B_x-A_x$  &  0.19  &  [0.08, 0.3]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.24]  \\ 
13304  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.55  &  [0.44, 0.65]  \\ 
910  &    &    &  $B_x-A_x$  &  0.07  &  [0.0, 0.15]  \\ 
  &    &    &  $E_x$  &  0.62  &  [0.51, 0.74]  \\ 
  &    &    &  $W_x$  &  0.83  &  [0.76, 0.89]  \\ [10.25ex] \hline
Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.43]  \\ 
4477  &    &    &  $B_x-A_x$  &  0.27  &  [0.22, 0.33]  \\ 
  &    &    &  $E_x$  &  0.82  &  [0.77, 0.87]  \\ 
  &    &    &  $W_x$  &  0.62  &  [0.57, 0.66]  \\ [10.25ex] \hline
Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.0, 0.18]  \\ 
7164  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.08  &  [0.0, 0.24]  \\ 
947  &    &    &  $B_x-A_x$  &  0.29  &  [0.2, 0.38]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kurmanji-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.09, 0.36]  \\ 
634  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.16]  \\ 
  &    &    &  $E_x$  &  0.86  &  [0.72, 0.95]  \\ 
  &    &    &  $W_x$  &  0.96  &  [0.92, 1.0]  \\ [10.25ex] \hline
Latvian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.51  &  [0.41, 0.62]  \\ 
4124  &    &    &  $B_x-A_x$  &  -0.05  &  [-0.11, 0.01]  \\ 
  &    &    &  $E_x$  &  0.37  &  [0.24, 0.53]  \\ 
  &    &    &  $W_x$  &  0.66  &  [0.58, 0.74]  \\ [10.25ex] \hline
Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.02, 0.25]  \\ 
848  &    &    &  $B_x-A_x$  &  0.33  &  [0.21, 0.47]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.45]  \\ 
2257  &    &    &  $B_x-A_x$  &  0.15  &  [0.08, 0.23]  \\ 
  &    &    &  $E_x$  &  0.73  &  [0.62, 0.81]  \\ 
  &    &    &  $W_x$  &  0.36  &  [0.28, 0.43]  \\ [10.25ex] \hline
Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.04, 0.3]  \\ 
29870  &    &    &  $B_x-A_x$  &  0.63  &  [0.48, 0.78]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.07, 0.32]  \\ 
4798  &    &    &  $B_x-A_x$  &  0.12  &  [0.06, 0.2]  \\ 
  &    &    &  $E_x$  &  0.93  &  [0.77, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [0.99, 1.0]  \\ [10.25ex] \hline
Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.5]  \\ 
6100  &    &    &  $B_x-A_x$  &  0.08  &  [-0.0, 0.15]  \\ 
  &    &    &  $E_x$  &  0.6  &  [0.5, 0.69]  \\ 
  &    &    &  $W_x$  &  0.31  &  [0.23, 0.39]  \\ [10.25ex] \hline
Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\ 
8664  &    &    &  $B_x-A_x$  &  0.15  &  [0.07, 0.24]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.06  &  [0.01, 0.15]  \\ 
2935  &    &    &  $B_x-A_x$  &  0.19  &  [0.13, 0.26]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.13, 0.3]  \\ 
8483  &    &    &  $B_x-A_x$  &  0.48  &  [0.39, 0.56]  \\ 
  &    &    &  $E_x$  &  0.94  &  [0.9, 0.97]  \\ 
  &    &    &  $W_x$  &  0.89  &  [0.83, 0.94]  \\ [10.25ex] \hline
Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.31  &  [0.22, 0.39]  \\ 
7532  &    &    &  $B_x-A_x$  &  0.44  &  [0.35, 0.56]  \\ 
  &    &    &  $E_x$  &  0.85  &  [0.79, 0.92]  \\ 
  &    &    &  $W_x$  &  0.79  &  [0.72, 0.85]  \\ [10.25ex] \hline
Spanish* &\\ [10.25ex] \hline
Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.29]  \\ 
7041  &    &    &  $B_x-A_x$  &  0.06  &  [0.01, 0.12]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Tamil  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.04, 0.29]  \\ 
400  &    &    &  $B_x-A_x$  &  0.17  &  [0.1, 0.28]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.06, 0.28]  \\ 
900  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.27]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.18  &  [0.05, 0.35]  \\ 
3685  &    &    &  $B_x-A_x$  &  0.03  &  [0.01, 0.06]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.22]  \\ 
4506  &    &    &  $B_x-A_x$  &  0.18  &  [0.1, 0.24]  \\ 
  &    &    &  $E_x$  &  0.94  &  [0.87, 1.0]  \\ 
  &    &    &  $W_x$  &  0.99  &  [0.97, 1.0]  \\ [10.25ex] \hline
Uyghur-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.06, 0.3]  \\ 
1656  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
  &    &    &  $E_x$  &  0.9  &  [0.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.0  &  [0.0, 0.0]  \\ [10.25ex] \hline
Vietnamese* &\\ [10.25ex] \hline
\end{longtable}

\section{Thoughts}
Upper bound: Use $TH[X_t] - \sum_{t=1}^T (T-t) I_t = T (H[X_t] - \sum_{t=1}^T I_t) + \sum_{t=1}^T t I_t$ bits of memory and incur $\sum_{t>T} I_t$ in additioal surprisal.

\begin{tabular}{cccccccccccccc}
Afrikaans & Amharic & Arabic & Armenian & Basque & Breton & Bulgarian & Buryat \\
\includegraphics[width=0.14\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Arabic-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Basque-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}
\\
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & Estonian \\
\includegraphics[width=0.14\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Catalan-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Chinese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Croatian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Czech-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Danish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Dutch-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}
\\
Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian \\
\includegraphics[width=0.14\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Finnish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/French-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/German-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Greek-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hindi-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}
\\
Indonesian & Japanese & Kazakh & Kurmanji & Latvian & Naija & North Sami & Norwegian \\
\includegraphics[width=0.14\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Japanese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Latvian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}
\\
Persian & Polish & Romanian & Serbian & Slovak & Slovenian & Spanish & Swedish \\
\includegraphics[width=0.14\textwidth]{figures/Persian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Polish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Romanian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Serbian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovak-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Spanish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}
\\
Tamil & Thai & Turkish & Ukrainian & Uyghur & Vietnamese \\
\includegraphics[width=0.14\textwidth]{figures/Tamil-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Turkish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}
\\\end{tabular}

\end{document}

\appendix
\section{Bounding Memory in Imperfect Modeling}
We derived the nonlocality $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?

The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
(Other models would incur even higher average surprisal, without reduction in memory cost.)
The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.

This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
%Naturally, we expect that lower surprisal rates will require more memory.

%\subsection{Lower bound on approximation with n-th order Markov models}
%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
%We know $\sum_{t>T} I_{t}$ 
%If a Markov model has order $n$

There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
\begin{proposition}\label{prop:markov}
The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
\end{proposition}

In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.

\subsection{Lower bound on imperfect approximation} 
Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.

As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
How much greater will their surprisal be? In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$. We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1. In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$

The following result holds for any process:
\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%Choose any $T \in \mathbb{N}$.
Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
\end{proposition}
%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$. It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
%An equivalent way of stating this bound is:
Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}

\begin{proof}
Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
As explained above, without loss of generality, we can assume that the model is Bayes optimal. %: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
By definition, the difference between the model's surprisal and optimal surprisal is
\footnote{The first inequality seems really crude. In general, we cannot do better. If we know more about the approximating model, we might be able to improve this. Also, consider the relation to the other inequality. Can they be attained at the same time?}
\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
\end{align*}
\end{proof}
%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$. Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
%
%\begin{corollary}
%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%
%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%
%\end{corollary}
%
%

Here is way of phrasing the same statement in terms of memory:
\begin{corollary}
Let $\epsilon > 0$. Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.

\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
%What is the optimal $T$?
%It turns out that the optimal $T$ 
Searching for the optimal $T$ in the bound, we find:
\begin{corollary}
Consider any memory requirement $J \in [0,M)$.
Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
That is, it cannot do better than a $T$-th order Markov model.
\end{corollary}
Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.

% a $T$-th order Markov

Here is a way of rephrasing this in terms of memory demand:
\begin{corollary}
Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$. Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
%
%Do these results agree with the idea of information locality decreasing memory load?
%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
%

\begin{example}
%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
%
Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.

Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
and, plugging back into the bound, get that the model's surprisal is at least
$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
\end{example}

%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns

%\begin{corollary}[Approximation by N-Gram Models, lower bound]
%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
%Then
%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
%
%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
%\end{corollary}
%
%\begin{proof}
%
%\end{proof}
%
%

\subsection{Upper Bounds on imperfect approximation}
$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy. Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$. Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else. We consider the problem of coding $X$ with small memory to predict $Y$ optimally. We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$. In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large. The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}

However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.

%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).

\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
\end{proposition}
\begin{proof}
In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
\end{proof}
%Compared with the lower bound from the previous section, we see:
Computing the memory cost of a an n-th order Markov model, we can conclude:
\begin{corollary}
(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.

(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$

\end{corollary}
\begin{proof}
By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
This model can be implemented using 
$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
This bound is really crude. We can compute
\begin{align*}
H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
%&\leq I/\epsilon H[X] + I \\
%& = I(1+H[X]/\epsilon) \\
\end{align*}
but does this gives us anything?}
The second statement is a rephrasing of the first.
\end{proof}
As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
In the case of natural language, $H[X]$ is a reasonably small constant.

%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
%\begin{corollary}[Approximation by $n$-gram models]
%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
%\end{corollary}

%\subsection{Summary}

%
%\paragraph{Invariance properties under certain transformations of the sequence}
%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
%%Second, 
%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
%If we measure memory also at all new boundaries, $I$ can at most increase.
%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
%
%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
%

\section{Crypticity}

%We let $X_t$ be a process of observations (e.g. words), and $M_t$ be the internal memory states of the listener, which we can take to be a function of $X_{\leq t}$.
Under the assumption of optimality oif predictions and minimality of memory cost, we can take $M_t$ to be an equivalence class of pasts that induce the same distribution over futures (need to be careful for general processes. we here want to add the assumption that, since we are interested in sentence processing, memory is reset entirely at sentence or discourse boundaries, avoiding cases where the math wouldn't work so nicely).

Nonlocality:
\begin{equation}\label{eq:locality-cost}
L := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}

Crypticity:
\begin{equation}\label{eq:ambiguity-cost}
A := H[M_0|X_{> 0}]
\end{equation}

Note that both terms are nonnegative by definition.

\begin{proof}
By the definitions of conditional entropy and mutual information,
$H[M_0] = H[M_0 | X_{>0}] + I[M_0, X_{>0}]$
The first part is Ambiguity cost.
Now we show that the second part is equal to Locality cost.
As we assume that the listeners predictions are optimal, $I[M_0, X_{>0}] = I[X_{\leq 0}, X_{>0}]$.
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
\end{proof}

\end{document}

\section{Normalized Excess Entropy}

We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$

Why?

(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity

(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?

\section{Computing a tighter lower bound}

Here is a method that takes a quadratic amount of time to compute:

Take all pairs of positions in the corpus, and for each pair $s, t$ compute
$$P(X_{t, ..., t+T}|T_{s-T, ..., s-1})$$
After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t, ..., t+T}|T_{s-T, ..., s-1}) + \log P(X_{t, ..., t+T}|T_{t-T, ..., t-1})\right) < \epsilon$$
Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) lower bound of memory for modeling with KL loss $\epsilon$.
However, this bound might still not be tight, unless we require that $\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$.

Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).

It should be feasible to apply this to POS-level language modeling.

ACTUALLY

Take all pairs of positions in the corpus, and for each pair $s, t$ compute
$$P(X_{t}|T_{s-T, ..., s-1})$$
After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t}|T_{s-T, ..., s-1}) + \log P(X_{t}|T_{t-T, ..., t-1})\right) < \epsilon$$
and
$\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$
Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) estimator of a tight lower bound of memory for modeling with KL loss $\epsilon$.

Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).

It should be feasible to apply this to POS-level language modeling.

Is this a consistent estimator?

\section{Pointwise Bounds}

\begin{proposition}
Let $F := X_{>0}$, $P := X_{\leq X}$. Then pointwise cost for an optimal representation (in terms of average memory cost) is
%$$C(p) := \log P\left{p' : P(F|p') = P(F|p)$$
\end{proposition}

There are several pointwise versions of excess entropy.
Here is one:
$$D(p) := \E_{f|p} pmi(f,p) = D_{KL}(P(f|p) || P(f))$$

\begin{proposition}
$D(p) \leq C(p)$
\end{proposition}

\begin{proof}
\begin{align*}
D(p) &= \sum_f p(f|p) \log \frac{p(f|p)}{p(f)} \\
&= \sum_f p(f|p) (\log p(p|f) - \log p(p)) \\
&= \left( \sum_{f} p(f|p) \log p(p|f)\right) - \log p(p)
\end{align*}
and the left quantity is $\leq 0$.
\end{proof}

$D$ can be estimated at each specific point by sampling continuations.

TODO $D(p)$ can be seen as a formalization of DLT storage cost

\begin{corollary}
If $\pi(p)$ is the equivalence class of the past, then
pointwise memory cost is
$$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
\end{corollary}
Taking the expectation over $p$, the left component is the excess entropy, while the right component is the gap between excess entropy and state entropy:
$$\E{f,p} \log(\pi(p)|f) = H[\pi(P) | F]$$

$H[\pi(P)] = I[P,F] + H[\pi(P) | F] = I[\pi(P), F] + H[\pi(p) | F]$

Memory = Non-Locality + Difficulty of Guessing Past from Future

I will call excess entropy `locality cost', and the second term `ambiguity cost'.

This decomposition also holds if $\pi$ is lossy. If $\pi$ is lossy, the non-locality term has to decrease.

Informally $$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
decomposes into a locality/storage cost term $\E_{f|p} pmi(f,p)$ (can be seen as a formalization of DLT storage cost), and $- \E_{f|p} \log(\pi(p)|f)$.
The latter component quantifies how hard it is to guess the storage content from the future (difficulty of guessing past from future, given the past).
One possible interpretation is in terms of noisy-channel memory:
Under memory constraints, more frequent/typical pasts will be more likely to be imputed based on the more recent observations. `Preventing a wrong noisy-channel inference takes extra memory effort.'
Can also relate this to prefix ambiguity, in the case where the ambiguity doesn't always turn up: E.g., if it is hard to predict from a continuation whether the prefix had been temporarily ambiguous.

Interpreting Ambiguity Cost:

- not knowing which things will be relevant later is costly

- one instantiation: not knowing which referents will be referred back to later

- one instantiation: ambiguous prefixes are costly

can you show that languages also optimize ambiguity cost?

\section{Masha's experiments}

- DepL experiment: excess entropy, we can even trace the pointwise profile

- MaxOP: initial NP can give rise to three memory states: subj, obj, indeterminate. The remainder of the sentence does not reveal whether the first NP had a case marker (if it did reveal that, this would increase locality cost). Ambiguity cost can be decreased by avoiding the indeterminate state (but there are other, even better minima! -- so this doesn't predict as much as MaxOP).

\section{Estimating Memory Cost and Ambiguity Cost}

by actually estimating a memory representation: estimate a variable that is independent of the future when conditioning on the past, and also: independent of past when conditioning on last word and last state of memory

if memory is deterministic, just say that it is a function of $X_t, M_{t-1}$.

if memory is randomized:

formally: $M_t$ must have the properties (1) $M_t \bot X_{<t} | (M_{t-1}, X_t)$, (2) $M_t \bot X_{>t} | X_{\leq t}$.

In this case, cost measure should be MI with past, not entropy

randomized memory is useful because we can apply gradient descent, solving an optimization problem to provide an upper bound on memory cost

- solving a discrete optimization problem (probably not super scalable)

- variational: stochastic continuous memory representations and take memory cost to be $I[Memory, Past]$ (also done in IB)

(noise the LSTM hidden state in each time step)

calculate upper bound $D_{KL}(P(Memory|Past) || P_{prior}(Memory)) \geq I[Memory, Past]$, estimate the tradeoff curve between this and cross-entropy (the resulting curve is a variational upper bound on the real curve).

- other ways of minimizing MI with past (don't require noising of the hidden state):

-- adversarial: make the past hard to predict

-- dual lower bound on MI

both of these methods are ways of doing IB

\section{Incremental Prediction under Memory Limitations}

%\subsection{Noisy-Context Surprisal}
%
%\citet{futrell-noisy-context-2017} describe a psycholinguistic noisy-channel model of human language processing under memory limitations where words are progressively deleted from memory. %with progressive probability.
%They show that processing as described by this model favors word order where linguistic material that has high mutual information will appear closer together, in the sense that the processing cost, measured by the surprisal $-\log P(w_t | w_1 ... w_{t-1})$ will be
%$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
%where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
%Given that $f$ is decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
%
%%In the most extreme case of such a model, a bigram model, we have:
%%$$-\log P_{Model}(w_t | w_1 ... w_{t-1}) = -\log P(w_t) - pmi(w_t, w_{t-1})$$
%%where $PMI(X,Y) = \log P(X,Y) - \log P(X) - \log P(Y)$ is the pointwise mutual information.
%%
%%This directly applies to n-gram models: Here, words are indeed `forgotten' deterministically when they are more than $n$ words in the past.
%
%%\citet{futrell-noisy-context-2017} show that this model derives a range of psycholinguistic locality effects.
%%It also applies to models with strictly bounded context, such as n-gram models -- in this case, past words beyond a fixed distance threshold are entirely erased from memory.
%%However, it is less clear that human language processing (or state-of-the-art NLP techniques such as LSTMs) are affected by such loss -- it is not clear that humans entirely forget previous words, and perfectly remember the other words.
%
%\subsection{Memory in General Sequence Models}

We consider a general incremental model that maintains a memory state and, in each time step, computes a distribution over the vocabulary based on its memory state, stochastically emits a word from this distribution, and updates its memory state depending on the chosen word.
We are making the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
This model describes aspects of human language processing, if we accept that humans perform incremental prediction.
(Typical NLP models such as n-gram models and LSTM language models also fit into this framework.)
%N-gram models and LSTM language models both fit into this framework.
%On the other hand, LSTM models with random access to prior words do not fit into this framework, as they rely on past states in addition to their internal memory state at time $t$.

In this general framework, the minimal amount of bits needed to encode the model's internal memory state is directly determined by the decay of mutual information over distances:
The faster mutual information between words decays as their distance increases, the less memory is required.
More precisely:

\begin{proposition}
Modeling a stationary process  $(X_t)_{t \in \mathbb{Z}}$ in this framework %of incremental modeling described above,
 requires maintaining at least
\begin{equation}\label{eq:memory-bound}
M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}
bits of memory,
where
$\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ is the mutual information of the observations $X_t$ and $X_0$, conditioned on the intervening observations.
\end{proposition}
%that the model does not have an `absolute clock' built into it.
This quantity is also known as \emph{excess entropy} (CITE).

\begin{proof}
%We assume that the process is stationary, that is, that there is no `absolute clock' built into the model. Formally, this means that, for any set of indices $t_1 < t_2 < \dots < t_n$, the joint distribution $(X_{t_1}, X_{t_2}, ..., X_{t_n})$ only depends on the differences $(t_{i+1}-t_i)_{i=1,...,n-1}$, not on the absolute values $t_i$.
The number of bits of memory is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
%In the last step, we have made the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
%that the model does not have an `absolute clock' built into it.
\end{proof}
That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.

Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.

The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model and the mechanisms by which it computes predictions.
In particular, while the relation to psycholinguistic and psychological models of how memory works would be interesting to explore, our result applies to any such model.
Memory representations do not have to be rational or optimal for this bound to hold:
It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.

%as long the model describes the process $(X_t)_{t \in \mathbb{Z}}$

%\section{Estimating Memory Load in Prediction}

%We want to apply (\ref{eq:memory-bound}) on corpus data to compute the memory load of language for incremental prediction.

For our purposes, it is important to note that we can estimate (\ref{eq:memory-bound}) from a statistical model assigning probabilities to sentences.
(TODO summarise prior work attemtpting to estimate the same quantity, if there is any such work).

\section{Bounding Memory in Imperfect Modeling}
We derived $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?

The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
(Other models would incur even higher average surprisal, without reduction in memory cost.)
The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.

This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
%Naturally, we expect that lower surprisal rates will require more memory.

%\subsection{Lower bound on approximation with n-th order Markov models}
%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
%We know $\sum_{t>T} I_{t}$ 
%If a Markov model has order $n$

There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
\begin{proposition}\label{prop:markov}
The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
\end{proposition}

In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.

\subsection{Lower bound on imperfect approximation} 
Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.

As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
How much greater will their surprisal be? In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$. We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1. In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$

The following result holds for any process:
\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%Choose any $T \in \mathbb{N}$.
Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
\end{proposition}
%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$. It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
%An equivalent way of stating this bound is:
Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}

\begin{proof}
Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
As explained above, without loss of generality, we can assume that the model is Bayes optimal. %: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
By definition, the difference between the model's surprisal and optimal surprisal is
\footnote{The first inequality seems really crude. In general, we cannot do better. If we know more about the approximating model, we might be able to improve this. Also, consider the relation to the other inequality. Can they be attained at the same time?}
\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
\end{align*}
\end{proof}
%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$. Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
%
%\begin{corollary}
%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%
%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%
%\end{corollary}
%
%

Here is way of phrasing the same statement in terms of memory:
\begin{corollary}
Let $\epsilon > 0$. Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.

\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
%What is the optimal $T$?
%It turns out that the optimal $T$ 
Searching for the optimal $T$ in the bound, we find:
\begin{corollary}
Consider any memory requirement $J \in [0,M)$.
Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
That is, it cannot do better than a $T$-th order Markov model.
\end{corollary}
Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.

% a $T$-th order Markov

Here is a way of rephrasing this in terms of memory demand:
\begin{corollary}
Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$. Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
%
%Do these results agree with the idea of information locality decreasing memory load?
%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
%

\begin{example}
%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
%
Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.

Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
and, plugging back into the bound, get that the model's surprisal is at least
$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
\end{example}

%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns

%\begin{corollary}[Approximation by N-Gram Models, lower bound]
%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
%Then
%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
%
%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
%\end{corollary}
%
%\begin{proof}
%
%\end{proof}
%
%

\subsection{Upper Bounds on imperfect approximation}
$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy. Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$. Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else. We consider the problem of coding $X$ with small memory to predict $Y$ optimally. We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$. In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large. The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}

However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.

%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).

\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
\end{proposition}
\begin{proof}
In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
\end{proof}
%Compared with the lower bound from the previous section, we see:
Computing the memory cost of a an n-th order Markov model, we can conclude:
\begin{corollary}
(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.

(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$

\end{corollary}
\begin{proof}
By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
This model can be implemented using 
$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
This bound is really crude. We can compute
\begin{align*}
H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
%&\leq I/\epsilon H[X] + I \\
%& = I(1+H[X]/\epsilon) \\
\end{align*}
but does this gives us anything?}
The second statement is a rephrasing of the first.
\end{proof}
As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
In the case of natural language, $H[X]$ is a reasonably small constant.

%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
%\begin{corollary}[Approximation by $n$-gram models]
%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
%\end{corollary}

%\subsection{Summary}

%
%\paragraph{Invariance properties under certain transformations of the sequence}
%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
%%Second, 
%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
%If we measure memory also at all new boundaries, $I$ can at most increase.
%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
%
%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
%

\section{Normalized Excess Entropy}

We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$

Why?

(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity

(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?

\section{Excess Entropy and Hierarchical Structure}

Finiteness of excess entropy is not in conflict with the presence of hierarchical structure and long-range dependencies in language:
Average memory demand can be small if the probability mass assigned to long dependencies is small.

Indeed, perhaps surprisingly, languages defined by PCFGs always have finite average memory (and thus in particular finite excess entropy):

\begin{proposition}
Any language generated by a PCFG\footnote{With rational weights, and finitely many terminals and nonterminals} has finite average memory demand.

\end{proposition}

\begin{proof}
We can consider incremental Earley parsing. Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory on average. For storing probabilities, the storage will be similarly bounded.
The result follows from (CITE).
\end{proof}

\begin{example}
Consider $a^nb^n$ with the probability assignment given by the PCFG $X \rightarrow \epsilon$ ($1-p$), $X \rightarrow aXb$ ($p$), for some $0<p<1$.

To compute  excess entropy, we construct a stationary process by considering concatenations of i.i.d. samples from the language.

%Also, the associated stationary process has finite excess entropy.

\end{example}

\section{Related Work}

derivation of information locality from lossy-context surprisal

Crutchfield, L{\"o}hr: construct HMMs that provide upper bounds on memory cost.
Should we be studying such upper bounds? I think that the lower bound that we study here is more useful due to the bounds it provides on imperfect modeling. 
The example from \cite{harsha-communication-nodate} shows that the true minimum memory cost may be a lot larger than the lower bound $M$ (it is exponential in $M$ for the family of processes that they considered), while -- for fixed KL loss and unigram entropy -- the memory required for imperfect modeling is $O(M)$, as we saw in the previous section.

Debowski: compute excess entropy in text

\cite{sharan-prediction-2016}: link between excess entropy and approximability by $n$-th order Markov models %study approximability by n-th order Markov models

\bibliographystyle{apalike}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

\section{Memory and Cross-Linguistic Word-Orders}

Do word orders in language minimize memory need as quantified by (\ref{eq:memory-bound}), when compared to other possible orderings?

\subsection{Prediction on the level of POS}
As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences. %, following~\cite{futrell-noisy-context-2017}.
As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.

%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
%
%%\mhahn{TODO results across languages}
%
%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.

Results for four languages are shown in Figure~\ref{fig:memory-results}.
Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).

%%For half of the languages
%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
%For 37 \% of the languages, it also requires less memory than the original language.

%We found that, as predicted, memory usage in real languages was correlated with word order freedom:

\begin{figure*}
\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%
\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
\end{figure*}

\subsection{Incorporating Morphology}
In many languages, morphology may have a big impact on prediction, but it is entirely ignored if we compute memory need for predicting POS tags.
Accounting for morphology requires a more sophisticated statistical model of language that is able to encode morphosyntatic generalizations.

The following models might be options, considering the NLP literature:

\begin{enumerate}
\item an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations. Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent. Both the small scale of many UD corpora and free word order in many languages with rich morphology thus seem to make such models unattractive.

\item a statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar. The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model. One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations. The second option will make this a machine learning model that, for our purposes, does not seem to be superior to (3). % (it kept many NLP people busy for the 2000s...)
\item a recurrent neural network (in particular, LSTM): NLP research in the past years shows that such models consistently fit language data better than (1) and (2) (and any other model that does not have a neural component). The model family seems rich enough to learn morphosyntactic generalizations and deal with free word order.
\end{enumerate}

(3) clearly seems to be the best choice.

As LSTMs are sequential models, the question is how to encode the input, including morphological information, into a sequence of tokens.
%On option would be to encode words as sequences of characters, or morphemes.
The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token.
Each POS tag and morphological feature-value pair is encoded as an atomic token.

During incremental prediction, the LSTM sequentially reads these tokens and predicts the next one.
The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).

Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.

Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
Therefore, to compute this quantity, we compute the differences between the surprisals
$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.

\section{Measuring Information Locality via Normalized Excess Entropy}

%We consider the following question: For a fixed budget of predictability, $I[X_t|X_{<t}]$, we want to minimize excess entropy -- that is, we want to

How do we compare excess entropies for models with different predictabilities?
We can take the quotient of excess entropy and predictability:
$$\frac{\sum_{t=1}^\infty t \cdot I[X_t,X_0 | X_{1...t-1}]}{\sum_{t=1}^\infty I[X_t,X_0|X_{1...t-1}]}$$
Why is this a good measure of locality?

\begin{enumerate}
\item Minimizing this quantity (subject to constraints) is equivalent to minizing excess entropy subject to fixed predictability (and these constraints).
\item This quantity stays invariant when we multiply each $I[X_t,X_0 | X_{1...t-1}]$ with a fixed factor. This makes it a natural measure of the decay of $I[X_t,X_0 | X_{1...t-1}]$ as $t \rightarrow \infty$.
\item This quantity is $\geq 1$, and it takes on the value $1$ if and only if the process is modeled perfectly by a bigram model (is a first-order Markov process).

For fixed entropy rate, the excess entropy is minimized by exactly the Markov processes of this entropy rate.
\end{enumerate}

\bibliographystyle{acl}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

Research has shown that dependency lengths are minimized across a wide range of languages, and it has been suggested that this may be a universal.

This raises the question of what causes this phenomenon.

What explains this?

Hawkins: TODO

Gildea and Jaeger: Gildea & Jaeger (2015) also perform optimization experiments, searching over possible grammars
optimized for dependency length, average log probability, or some weighted combination of the two.
The authors find that optimizing one criterion does not generally lead to good performance on the
other,  but that the observed languages are generally closer to some optimal frontier that trades off
dependency length against average log probability.  (from Temperley 2018)

Futrell and Levy: noisy context surprisal explains

Here, we use state-of-the art language modeling techniques to address this question:
On 50 diverse languages, we ... 
Resulting languages show shorter dependency length.

\subsection{Ordering Model}

As in work studying dependency length minimization in language (e.g. Futrell et al. 2015, Gildea and Jaeger 2015), we specify a parametric dependency tree ordering model.
%While the precise parameterization of the models is not necessarily crucial, it is crucial that the models are probabilistic, as this allows us to use gradient-based optimization.
%As we want to compare models optimized for language modeling with models 
We design our model to be particularly simple to make its parameters interpretable and comparable across models.

For each triple $\tau = $(Head POS, Dependency Label, Dependent POS), we have (1) a Direction Parameter $a_\tau$, and (2) a Distance Parameter $b_\tau$.
%An Ordering Model is an assignment of values $a_\tau, b_\tau \in \R$ to each of these triples.
Given a dependency tree, define a distribution over linearizations as follows:

\begin{enumerate}
\item Choose which dependents follow or precede their heads: Given a head $h$ and dependents $\{d_1, ..., d_k\}$, independently sample $O_i \sim Bernoulli(\sigma(a_{\tau_i}))$, where $i = 1, ..., k$, $\sigma$ is the inverse-logit function, and $\tau_i = (\text{POS of }h, \text{Label of }h \rightarrow d_i, \text{POS of }d_i)$. Order $d_i$ to the left of $h$ if $O_i = 0$, otherwise to the right of $h$.

\item %For any set of siblings $d_1, ..., d_k$ that are on the same side of their common head, the probability that $d_i$ is closest to the head, among these siblings, is $$\frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$$

Siblings $d_1, ..., d_k$ that were assigned to the same side of their head are ordered from the head outwards by iterating the following procedure:

Sample $i \in \{1, ..., k\}$ from the distribution $p_i = \frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$.
Then put $d_i$ closest to the head, and remove $d_i$ from the set, so that $k-1$ siblings remain.
Repeat this procedure until only one sibling remains, which is then ordered farthest away from the head.
\end{enumerate}

\mhahn{TODO: Maybe I should create a figure to illustrate this?}

As (Futrell 2015, Gildea and Jaeger 2015),  we only consider projective linearizations.

one similar model is the one from Gildea \& Temperley (2007), where each dependent has a weight in [-1,1] which defines its position wrt the head and the other dependents.

Yours is more general in that it’s probabilistic; theirs is deterministic.

I also have a short EMNLP paper Futrell & Gibson (2015) which derives ordering models from generative dependency models such as Eisner Model C and the Dependency Model with Valence: the Eisner Model C linearization model seems like it would end up being the same as yours with $a_tau$ but not $b_tau$

Kristina Gulordava made an improved version of those models but I can’t find her paper right now; I’m pretty sure it was at CMCL

Another previous work is the reordering model from Galactic Dependencies: Wang & Eisner (2016)

There’s also a small but focused literature on more general linearization models, with a shared task in Belz et al. (2011)

I think the state of the art linearization model in terms of BLEU score is ZGen: https://github.com/SUTDNLP/ZGen but these highly optimized NLP models probably won’t work for a scientific study like you’re doing because of their complexity.

%One issue with most of these models (glidea & temperley, my model, galactic dependencies) is they only model order constraints in “local subtrees” consisting of heads and their immediate dependents. That may not always work (but it’s hard to think of a counterexample other than heavy NP shift!). Also, nearly all models that I’m aware of, including yours, only generate projective linearizations.

The only exceptions I know of are some things by Bernd Bohnet, who’s now at Google: http://www.aclweb.org/anthology/D12-1085. And the way his system does it strikes me as hacky. %Another issue with some models (glidea & temperley, my model based on Eisner Model C, your model) is they don’t model interactions among dependents: it may be that if some dependent A is present that changes the order preference for some other dependent B, but this isn’t captured. 
%
%There’s not a lot of work using linearization models to test scientific hypotheses. The only ones I can really think of are Gildea & Temperley and my stuff and other people who have used that method.

A big problem with that work is that there’s no way to quantify goodness of fit: I can say that real dependency length is minimized beyond some baseline but what I’d like to do is make a model that incorporates dependency length minimization and show that it fits the real orders better than a model without dependency length minimization. It seems like your method provides a way to do that. 
%
%On minimizing dependency length, I don’t know if you know about the algorithmic work on this problem. There’s polynomial time solutions to the minimal-dependency-length linearization problem subject to various constraints. Chung (1984) has the solution when there are no constraints, Gildea & Temperley (2007) have it under the constraint of projectivity, and Park & Levy (2009) do it under the constraint of mild nonprojectivity (gap degree 1). These wouldn’t fit into your gradient descent framework though.
%
%Future work may extend this to non-projective ones.

\subsection{Optimization Framework}

We are interested in finding ordering models $\thetad$ that minimize the cross-entropy loss:
$$-\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P(w_{i+1}|w_i)$$
where linearized sentences ${\bf w}$ are obained by ordering trees from the treebank according to the ordering model $\thetad$, and $P$ is a language model appropriate to this distribution of linearized sentences. %the distribution defined by $\thetad$.
%In the limiting case where there are no memory limitations, and $P$ is exactly the distribution defined by $\thetad$, this is the entropy of the language, and does not actually depend on the ordering model.
Note that the original word orders from the treebank do not enter this objective -- they in particular do not influence the language model $P$.

As we simultaneously look for an ordering model optimized for surprisal of the language model, and a language model that optimizes surprisal given the ordering model, this looks like a chicken-egg problem.
Fortunately, we can reduce this to an unconstrained optimization problem:
The appropriate language model for a given language is the one that minimizes expected surprisal.
Therefore, our minimization problem is equivalent to jointly minimizing over ordering models $\thetad$ and language models $\thetal$:
\begin{equation}\label{eq:surp-final-obj}
\min_{\thetad, \thetal} -\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})
\end{equation}

We solve~(\ref{eq:surp-final-obj}) by performing stochastic gradient descent over both sets of parameters $\thetad, \thetal$ simultaneously:
In each step, we sample a dependency tree from the treebank, then sample an ordering from the current setting of $\thetad$ to obtain a linearized sentence ${\bf w}$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator
$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
for the gradient ${ \partial_\thetad R(\thetad, \thetal) \choose \partial_\thetal R(\thetad, \thetal)}$.
This unbiased estimator is the ordinary gradient estimator for $\thetal$, and the REINFORCE estimator for $\thetad$~\cite{williams-simple-1992}.

\paragraph{Implemention}
To reduce the variance of the gradient estimator, we apply reward attribution (CITE) and use a word-dependent baseline.
We add an entropy reward (CITE) for the direction parameters $a_\tau$, weighted with $\alpha$.
In preliminary experiments, we found $\alpha=0.001$ and a learning rate of $0.01$ with momentum $0.9$ for $\thetad$ to give stable results.
We apply early stopping using a Monte Carlo estimate of the loss  (\ref{eq:surp-final-obj}) on the development set.
The hyperparameters for the LSTMs were as in the first experiment; however, we explicitly added input POS embeddings with 20 dimensions to encourage generalization in languages with small training corpora.
Embeddings were randomly initialized, as before.

\paragraph{Setup}
We carried this out on all Universal Dependencies\footnote{\url{http://universaldependencies.org/}} 2.1 treebanks for which a training set was available, a total of 50 languages.
For each language where multiple treebanks with training sets were available, we pooled their training sets; similarly for development sets.

\section{Results}

%\subsection{Correlation with Maximum-Likelihood Fit}
\subsection{Comparison with Ground-Truth Orderings}

\begin{figure}
\includegraphics[width=0.48\textwidth]{figures/freedom-lstm.pdf}
\caption{Optimizing word order for LSTM language modeling explains more of the language's actual rules in languages that have constrained word order.}\label{fig:lstm-languages}
\end{figure}

We first ask to what extent the ordering rules of the actual languages, as manifested in the corpora, can be recovered by optimizing for predictability.

Since we are interested in locality effects, we compare the distance parameters inferred by optimizing for language modeling with distance parameters inferred from the actual orderings via maximum-likelihood with an $\ell_2$ penalty.

We correlate the weights both at the level of dependency triples, and averaged at the levels of dependency labels, head POS, and dependent POS.
The correlations are significant at $p < 0.01$ for X languages.
In Figure~\ref{fig:lstm-languages}, we plot the per-language correlations on the level of triples against word order freedom.
The optimized languages explain more of the ground-truth ordering preferences in languages with low word order freedom ($r =$TODO, $p <$ TODO), in accordance with our previous result that LSTMs showed a stronger bias towards languages with fixed word order.

\subsection{Information Locality}

We predict that orderings optimized for prediction show information locality. %ease prediction with a limited memory. %show steeper decay of mutual information than random orderings.
We quantify Information Locality by computing the memory needs by the formula~(\ref{eq:memory-bound}).
As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we compute MI on the level of POS tags, following~\cite{futrell-noisy-context-2017}.
As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly.

We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.

Results for four languages are shown in Figure~\ref{fig:memory-results}.
We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
In comparison, again averaged across languages, natural language requires less memory than 87 \% of baseline models.
%For half of the languages
For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
For 37 \% of the languages, it also requires less memory than the original language.

We found that, as predicted, memory usage in real languages was correlated with word order freedom:
Languages which took less memory than all associated baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).

\begin{figure*}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Arabic.pdf}
%%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Japanese.pdf}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-English.pdf}
%
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Korean.pdf}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Basque.pdf}
\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%
\caption{Memory demands of real, optimized, and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
\end{figure*}

\subsection{Dependency Length Minimization}

\begin{figure}
\includegraphics[width=0.48\textwidth]{figures/depLength-facet.pdf}
\caption{Optimizing for LSTM language modeling (blue) shortens dependency lengths compared to baseline (green), though not always as strongly as natural language (red) does.}\label{fig:dep-lengths}
% Optimizing for LSTM modeling (green) tends to shorten more strongly than optimizing for bigram surprisal (red). \mhahn{choose languages}}
\end{figure}

A prominent manifestation of locality in language is dependency length minimization:
Languages have shorter dependencies than would be expected at random~\cite{futrell-large-scale-2015}. %, and under reasonable baselines.

%\cite{futrell-noisy-context-2017} 
As previous work (CITE) has argued that dependency length minimization can be derived from human memory limitations, we expect that languages optimized for language modeling should show short dependencies.
We computed per-sentence dependency lengths for the optimized ordering rules, creating a counterfactual corpus by sampling a linearization for each tree.
%As before, we sample linearizations at zero tempterature to eliminate nondeterminism that may lengthen average dependency length.
%We sampled trees with temperature $\tau = 1/100$ to eliminate nondeterminism (which may lengthen average dependency length).

%As a baseline, we also computed dependency lengths for ordering models that were trained with dependency length as the objective.
%(This differs from (CITE), who computed the per-sentence absolute minimum. Our baseline is more realistic in that it respects grammatical consistency across sentences).
As in the previous experiment and similar to \cite{futrell-large-scale-2015}, we considered ordering models with randomized $a_\tau, b_\tau$ as baselines.

In every one of the 50 languages, the optimized models show shorter dependencies than the baseline languages.
Results from four languages are shown in Figure~\ref{fig:dep-lengths}.
In 20 \% of the languages, they exceed the actual ground-truth orderings in dependency length minimization.

\subsection{Qualitative Predictions Across Languages}

We have seen that, for languages with fixed word order, optimizing for language models explains a significant amount of word order preferences.
We now examine what kinds of rules LSTMs prefer across languages. %take the grand average across languages.

We averaged distance weights across all languages, both for the optimized models, and for the maximum-likelihood models.
Within each language, we averaged across dependents, weighted by frequency of the dependency-dependent combination in the language.

The correlations between these averages across 33 dependency types was 0.50 \mhahn{TODO: I'll bootstrap over languages}, i.e., half of the variance in cross-linguistic ordering preferences as captured by $b_\tau$ could be explained as being optimized for language modeling. 
We then considered specifically the averages separately for dependencies headed by nouns and verbs, respectively.
We considered the five dependencies that bind most strongly and least strongly, that is, where $b_\tau$ was smallest or greatest.

%For nouns, we found that dependents typically considered to be part of the noun phrase (determiners, numerals, adjectives, modifiers, compound parts) bind strongly to the noun both in optimized and real languages.
%Copulas and adpositions, which in UD are considered dependents of the noun, bind very strongly in optimized models, while they are generally not considered part of the noun phrase in the linguistic literature -- nonetheless they tend to bind relatively close to the noun even in natural language.
%Intriguingly, the functions of adpositions and determiners, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.
%In both optimized and real languages, among the dependents that tend to be less close to the noun are relative clauses, arguments of copulas, adverbs, and coordination -- these are elements that are not part of the noun phrase as typically understood in the linguistic literature, and they also tend to be longer.

The closest dependencies were determiners, classifiers, case markers/adpositions, and numerals.
The farthest dependencies were relative clauses, conjuncts, adjectives, and adverbial clauses.
Intriguingly, the functions of determiners, classifiers, and adpositions, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.

Turning now to dependencies headed by verbs, expletives, auxiliaries, markers, and adjectives bind most strongly. % bind very strongly to the verb both in optimized and in real languages, even they are often not understood as dependents of the verb in the linguistic literature.
Again, intriguingly, typical functions of auxiliaries and markers are expressed morphologically on the verb in many languages. %, suggesting again that grammaticalization into morphological markers may be favored by locality.
Perhaps counter to expectation, arguments (subjects, objects) are not generally closer than adjuncts, neither in optimized nor in real languages.
Similar to noun phrases, adjunct clauses stand out farther in optimized languages.
The farthest dependents are two kinds of nominal modifiers, dislocated elements, and objects.

\section{Related Work and Discussion}

\citet{gildea-human-2015} showed that a sample of five natural languages optimizes trigram surprisal and dependency length compared to random baselines, and also performed optimization with these objectives.
Our work links these findings to general properties of sequence models including LSTMs, and studies a much broader sample of languages. % and studies a much broader sample of languages. %, and (3) investigating qualitative linguistic predictions.

Our work relates to recent work showing that neural language models tend to focus strongly on recent words for prediction~\cite{daniluk-frustratingly-2017, khandelwal-sharp-2018}.
Our results suggests that this is explained both by general properties of sequence models with bounded memory, and by natural language favoring incremental prediction with such models.

Recent work has disucssed the adequacy of sequence models for language with its hierarchical structure (CITE). Our work adds to this by showing that some of the architectural biases in linear sequence modeling turn out to favor natural language, arguably because they are also present in human language processing. %natural language.

%Our findings highlight that there is a psycholinguistic explanation for the reasonable success of local methods, such as n-gram models, in NLP.
%We have shown that word order regularities in languages can partly be recovered from unordered data by optimizing for predictability.

\section{Conclusion}

We first showed that LSTM language models are biased towards orderings of information as found in natural language, and that this bias is stronger for languages with fixed word orders.
This is grounded in general properties of sequence modeling, which favors sequences exhibiting locality as found in natural language.
Asking which orderings would be optimal for LSTM language modeling, we then used reinforcement learning to create counterfactual versions of 50 diverse languages that are optimized for ease of language modeling.
We showed that, across corpora, LSTMs favor word orders that exhibit information locality and have short syntactic dependencies, similar to natural languages.
In languages with fixed word order, a significant amount of the actual word order preferences can be recovered by optimizing for language modeling, suggesting that word order rules in these languages can partly be explained as making prediction easier.

%- If we could adapt the language's word order rules to make LSTM language modeling easiest, what would the resulting language look like?

%- previous work has shown language has locality

%- we showed that LSTM language models are biased towards orders typically chosen in natural language, and that this preference increases as memory is restricted, and increases if memory is unbounded

%- this bias was shown to be caused particularly by function words -- words primarily carrying grammatical information.

%- we then used reinforcement learning to directly investigate what kinds of word orders are preferred by incremental language models. We chose LSTMs and bigram models as two extremes.

%- Across 50 languages, we showed that LSTMs prefer word orders that require little memory to model and have short syntactic dependencies, similar to natural language.

%- Using an information theoretic measure of memory need, we showed that word orders chosen in many natural language require less memory than would be expected at random

%- Our quantification of memory demands is entirely information-theoretic and makes no architectural assumptions beyond incrementality of sequence modeling.
%
%- We also showed that inferred ordering preferences explain a significant amount of ordering preferences in natural language, both for specific languages and across languages.
%
%- Our work has implications both for NLP and for syntactic typology. 
%
%
%Bounded context representations
%
%
%We observe that incrementally predicting language under memory limitations -- as instantiated by both recurrent language models and, arguably, human language processing, favors word orders that exhibit Information Locality: 
%
%We showed that recurrent language models indeed show a bias towards information locality.
%We then showed that information locality makes nontrivial and partly accurate typological predictions for word order across languages.

\section{More on Memory}

- tradeoff between memory and cross-entropy: let $M$ be a model with memory $Mem(M)$. For all $t$,

$Mem(M) + t CE[M,X] \geq Mem(X) + t H_{rate}[X]$

since there is a compression scheme for the sequences in $X$ of average length $t$ with size $Mem(M) + t CE[M,X]$, but the optimal scheme has length $Mem(X) + t H_{rate}[X]$

In particular, if $M$ is a lossy context encoding scheme and word-by-word prediction is optimal w.r.t. this, the word-by-word KL loss is lower bounded as
$$(CE[M,X] - H_{rate}[X]) \geq (Mem(X) - Mem(M))$$
Of course, $Mem(M)$ might be large, in which case this bound will be vacuous.
We could define the actual information content in $Mem(X)$ that is relevant to $X$ using cross-entropies -- this might then be larger or smaller than the actual information content, depending on the efficiency of the encoding (e.g., it could ignore certain kinds of context observations and thus require unnecessarily much memory).

\bibliographystyle{acl}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

\begin{align*}
& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
\end{align*}

Is this useful?
If we know $I_t$, we can choose $T$ so as to make this bound as tight as possible.
To show the usefulness of this method, we consider the case where $I_t = \frac{\alpha}{t^\beta}$, for $\alpha > 0$, $\beta > 2$ (for smaller $\beta$, $M$ evaluates to $+\infty$).
Then $M = \alpha \sum_{t=1}^\infty \frac{1}{t^{\beta-1}}$, which is bounded as $\frac{\alpha}{\beta-1} = \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt \leq M \leq \alpha + \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt = \alpha + \frac{\alpha}{\beta-1} = \frac{\alpha\beta}{\beta-1}$.
The bound from the proposition evaluates to
\begin{align*}
\sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T &= \sum_{t \leq T} \frac{\alpha}{T t^{\beta-1}} + \sum_{t > T} \frac{\alpha}{t^\beta} - J/T \\
& \geq \int_1^T \frac{\alpha}{T t^{\beta-1}} dt + \int_T^\infty \frac{\alpha}{t^\beta} dt - J/T \\
& = \frac{\alpha (\beta-1)}{T} - \frac{\alpha(\beta-1)}{T^{\beta-1}} + \frac{\alpha\beta}{T^{\beta-1}} - J/T \\
&= \frac{1}{T} (\alpha(\beta-1) - J) + \frac{\alpha}{T^{\beta-1}}  \\
&\geq \frac{1}{T} (M(\beta-1) - \alpha - J) + \frac{\alpha}{T^{\beta-1}}  \\
%&\geq \frac{1}{T} (M - J) + \frac{M}{(\beta-1)T^{\beta-1}} \\
%& \geq M-J \ \ \ \ \ \ \  \text{(by setting $T=1$)} 
\end{align*}
%whose derivative by $T$ is
%$$\frac{-1}{T^2} (\alpha(\beta-1) - J) - \frac{\alpha(\beta-1)}{T^{\beta}}  $$
%which is zero at
%$$T = \left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)} $$
%Substituting back, we get the bound
%$$\frac{1}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)}} (\alpha(\beta-1) - J) + \frac{\alpha}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{(\beta-1)/(\beta-2)}}$$

%
%\begin{proposition}
%Let $\epsilon > 0$, $T \in \mathbb{N}$.
%If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$. TODO
%\end{proposition}
%\begin{proof}
%Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
%Without loss of generality, we can assume that the model is Bayes optimal: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
%
%By definition, the difference between the model's surprisal and optimal surprisal is
%\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
%& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
%& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
%& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
%& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
%& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
%%& \geq \frac{1}{T} \sum_{t \leq T}  I_t + \sum_{t > T} I_t - J/T \\
%%& \geq \frac{1}{T} (TODO - I[X_{1\dots T}|f(X_{<0})]) \\
%%& \geq \frac{1}{T} (TODO M - J - \epsilon) \\
%\end{align*}
%\end{proof}

%If we consider ratios between the memory bound and the predictability $I[X_t, X_{<t}]$, we can also express this as follows:
%\begin{corollary}
%Let $\epsilon > 0$, and let $P := I[X_t, X_{<t}]$ the predictability.
%Then If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$.
%\end{corollary}
%
%\begin{proof}
%$\frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{P} = \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > 0} I[X_t, X_0 | X_{1\dots t-1}]}  \leq \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > M/\epsilon} I[X_t, X_0 | X_{1\dots t-1}]}< \epsilon$
%
%Let $T := M/\epsilon$.
%Then 
%\end{proof}
%
%

%\subsection{Incorporating All Lemmas}

%\subsection{Prediction on Raw Word Forms}

%\section{information locality and dependency length minimization}

%\subsection{Languages can trade off Memory Load and Predictability; Information Locality}
%
%%The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$
%
%We will describe a geometric view on $M$ that gives us a  quantitative measure of Information Locality.
%
%By information locality, we refer to the degree to which words that are predictive of each other are closer together.
%Stated differently, information in past observations that helps predict $X_t$ should be concentrated in recent observations $X_{t'}$ ($t' < t$) with $t'$ close to $t$.
%
%Information locality is maximised if all contextual information predicting $X_t$ is contained in $X_{t-1}$, in which case $J = I[X_t, X_{t-1}]$.
%
%To quantify information locality more generally, we consider the relation between the quantity
%$$J = I[X_{\geq t}, X_{<t}] = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
%and the quantity
%$$P = I[X_t, X_{< t}] = \sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
%The first one quantifies how much information is transmitted between past and future, the second one captures the average predictability of single words from their prior context.
%Evidently, $P \leq J$ -- higher degrees of predictability will require more information to be stored from past observations.
%Stated differently, there is a tradeoff between predictability and memory: If the future is easier to predict from the past (at the same unconditional entropy), more memory is required to store all the relevant predictive information from the past.
%
%Minimizing $J$ unconditionally corresponds to an entirely unpredictable process.
%In contrast, if we minimize $J$ while holding $P$ constant, we obtain a formalization of the requirement of optimal information locality: This requires placing the predictive information closer to the present observation, keeping the total amount of predictive information constant..
%
%If there are no further constraints on the minimization, the minimum is attained when $P = J$.
%This is the case when all contextual information predicting $X_t$ is contained in $X_{t-1}$ -- this minimizes memory load, as only information from the last word needs to be stored.
%If $P$ is fixed, and $J$ increases, information locality decreases -- corresponding to increasing memory load.
%
%To understand this more generally, let us consider $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ as a function of $t$.
%This function describes the amount of bits predictive of $X_t$ contained in $X_0$, the observation $t$ steps in the fast.
%The faster this function decays, the lower $J$ is.
%Information locality can be interpreted as requiring this function to decay fast -- $J$ is absolutely minimized if the function jumps to zero for $t > 1$.
%
%If this is a principle describing natural language, we expect that there will be further constraints, resulting in values of $J$ larger than $P$.
%We will study this in detail in the next section.
%
%
%We can observe that the area under the curve is equal to $P$ -- this quantifies the total amount of contextual information predicting the current observation.
%This suggests to consider the ratio
%$$L := \frac{J}{P}$$
%First, note that minimizing $J$ with $P$ fixed is equivalent to minimizing $L$.
%Second, expanding this quantity, we see
%$$L = \sum_{t=1}^\infty t \frac{\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}{\sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}$$
%That is, $L$ describes the $J$-coordinate of the center of mass of the area under the curve in (REF).
%A different way of understanding it is as the average distance of past information predictive of the current observation.
%With this interpretation of $L$, we can understand it as a direct fornalization of the notion of information locality.
%
%Let  us consider a second visualization by plotting $J$ against $P$.
%The area that is achievable by processes is the area below the diagonal.
%Information locality is maximized by processes on the diagonal; the distance from the diagonal quantifies the ratio $J/P$ and thus information locality.
%
%

### 2019-01-20 toy/analyze.R

@@ -1,0 +1,265 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+
+data = read.csv("results.tsv", sep="\t")
+
+data = data %>% mutate(Order = ifelse(Type=="good", "A", "B"))
+
+
+plot = ggplot(data=data, aes(x=Distance, y=MI, color=Order)) + geom_step(size=2) + theme_bw() + theme(axis.text=element_text(size=18), axis.title=element_text(size=18), legend.title=element_text(size=18), legend.text=element_text(size=18))
+
+ggsave("toy-mis.pdf")
+
+
+
+
+plot = ggplot(data=data, aes(x=Distance, y=Distance * MI, color=Order)) + geom_step(size=2) + theme_bw() + theme(axis.text=element_text(size=18), axis.title=element_text(size=18), legend.title=element_text(size=18), legend.text=element_text(size=18))
+
+ggsave("toy-t-mis.pdf")
+
+
+
+data$MI[is.na(data$MI)] = 0
+data = data %>% group_by(Order) %>% mutate(CumMI = cumsum(MI), Memory = cumsum(Distance*MI)) %>% mutate(Surprisal = 2.07-CumMI)
+plot = ggplot(data=data %>% filter(Surprisal < 2.0), aes(x=Surprisal, y=Memory, color=Order)) + geom_line(size=2) + theme_bw() + theme(axis.text=element_text(size=18), axis.title=element_text(size=18), legend.title=element_text(size=18), legend.text=element_text(size=18))
+
+
+ggsave("toy-surp-mem.pdf")
+
+
+plot = ggplot(data=data %>% filter(Surprisal < 2.0), aes(x=Memory, y=Surprisal, color=Order)) + geom_line(size=2) + theme_bw() + theme(axis.text=element_text(size=18), axis.title=element_text(size=18), legend.title=element_text(size=18), legend.text=element_text(size=18))
+
+
+ggsave("toy-mem-surp.pdf")
+
+
+
+
+
+
+
+#plot = ggplot(data=data, aes(x=Distance, y=Distance*MI, color=Type)) + geom_line()
+
+
+x = (1:10)
+y = 5 * x^(-2)
+
+data = data.frame(Distance=x, MI=y)
+
[…]

### 2019-01-20 toy/simulation-2.py

@@ -1,0 +1,266 @@
+
+
+productions = {}
+
+#productions["S"] = [(0.5, ("NP-heavy", "di", "N_a", "V", ".")), (0.5, ("NP-heavy", "N_a", "di", "V", "."))]
+productions["S"] = [(0.5, ("N_a", "di", "NP-heavy", "V", ".")), (0.5, ("N_a", "NP-heavy", "di", "V", "."))]
+
+productions["NP-heavy"] = [(1.0, ("Adj", "N_i", "P", "N_a"))]
+
+
+import numpy as np
+import random
+
+def sample(x):
+    if x in productions:
+        probs = [y[0] for y in productions[x]]
+        return " ".join([sample(z) for z in productions[x][np.random.choice(range(len(probs)), p=probs)][1]])
+    return x
+
+vocabulary = set()
+for x in productions:
+      for _, rhs in productions[x]:
+          for word in rhs:
+              if word not in productions:
+                  vocabulary.add(word)
+itos = list(vocabulary)
+stoi = dict(zip(itos, range(len(itos))))
+print(itos)
+
+print(sample("S"))
+
+
+
+
+import random
+import sys
+from collections import deque
+
+lengthForPrediction = 20 #int(sys.argv[4]) #20
+contextLength = 20 #int(sys.argv[5]) #20
+
+
+
+
+myID = random.randint(0,10000000)
+
+posUni = set() #[ "ADJ", "ADP", "ADV", "AUX", "CONJ", "DET", "INTJ", "NOUN", "NUM", "PART", "PRON", "PROPN", "PUNCT", "SCONJ", "SYM", "VERB", "X"] 
+
+posFine = set() #[ "``", ",", ":", ".", "''", "$", "ADD", "AFX", "CC",  "CD", "DT", "EX", "FW", "GW", "HYPH", "IN", "JJ", "JJR",  "JJS", "-LRB-", "LS", "MD", "NFP", "NN", "NNP", "NNPS", "NNS",  "PDT", "POS", "PRP", "PRP$", "RB", "RBR", "RBS", "RP", "-RRB-", "SYM", "TO", "UH", "VB", "VBD", "VBG", "VBN", "VBP", "VBZ",  "WDT", "WP", "WP$", "WRB", "XX" ]
+
+
+
+
+from math import log, exp
+from random import random, shuffle
+import os
+
+
+originalDistanceWeights = {}
+
[…]

### 2019-01-20 toy/simulation-3.py

@@ -1,0 +1,426 @@
+
+
+productions = {}
+
+import random
+
+basic = random.choice(["verb-final", "verb-medial", "half-half"])
+case = random.choice(["before", "after"])
+npOrder = random.choice(["APN", "ANP", "PAN", "PNA", "NAP", "NPA"])
+
+productions["S"] = [(0.5, ("NP-heavy", "di", "N_a", "V", ".")), (0.5, ("NP-heavy", "N_a", "di", "V", "."))]
+#productions["S"] += [(0.25, ("N_a", "di", "NP-heavy", "V", ".")), (0.25, ("N_a", "NP-heavy", "di" "V", "."))]
+
+productions["NP-heavy"] = [(1.0, ("Adj", "N_i", "P", "N_a"))]
+
+
+
+sentence1 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "obj_heavy"}, {"index" : 5, "word" : "di", "head" : 4, "dep" : "case"}, {"index" : 6, "word" : "N_a", "head" : 7, "dep" : "nsubj"}, {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+sentence2 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "nsubj_heavy"}, {"index" : 5, "word" : "N_a", "head" : 7, "dep" : "obj"}, {"index" : 6, "word" : "di", "head" : 5, "dep" : "case"} ,    {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+
+
+sentences = [sentence1, sentence2]
+
+itos_deps = list(set([x["dep"] for x in sentence1+sentence2]))
+stoi_deps = dict(zip(itos_deps, range(len(itos_deps))))
+
[…]

### 2019-01-20 toy/simulation-4.py

@@ -1,0 +1,434 @@
+
+
+productions = {}
+
+import random
+
+basic = random.choice(["verb-final", "verb-medial", "half-half"])
+case = random.choice(["before", "after"])
+npOrder = random.choice(["APN", "ANP", "PAN", "PNA", "NAP", "NPA"])
+
+productions["S"] = [(0.5, ("NP-heavy", "di", "N_a", "V", ".")), (0.5, ("NP-heavy", "N_a", "di", "V", "."))]
+#productions["S"] += [(0.25, ("N_a", "di", "NP-heavy", "V", ".")), (0.25, ("N_a", "NP-heavy", "di" "V", "."))]
+
+productions["NP-heavy"] = [(1.0, ("Adj", "N_i", "P", "N_a"))]
+
+
+
+sentence1 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "obj_heavy"}, {"index" : 5, "word" : "di", "head" : 4, "dep" : "case"}, {"index" : 6, "word" : "N_a", "head" : 7, "dep" : "nsubj"}, {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+sentence2 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "nsubj_heavy"}, {"index" : 5, "word" : "N_a", "head" : 7, "dep" : "obj"}, {"index" : 6, "word" : "di", "head" : 5, "dep" : "case"} ,    {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+
+
+sentences = [sentence1, sentence2]
+
+itos_deps = list(set([x["dep"] for x in sentence1+sentence2]))
+stoi_deps = dict(zip(itos_deps, range(len(itos_deps))))
+
[…]

### 2019-01-20 toy/simulation-5.py

@@ -1,0 +1,434 @@
+
+
+productions = {}
+
+import random
+
+basic = random.choice(["verb-final", "verb-medial", "half-half"])
+case = random.choice(["before", "after"])
+npOrder = random.choice(["APN", "ANP", "PAN", "PNA", "NAP", "NPA"])
+
+productions["S"] = [(0.5, ("NP-heavy", "di", "N_a", "V", ".")), (0.5, ("NP-heavy", "N_a", "di", "V", "."))]
+#productions["S"] += [(0.25, ("N_a", "di", "NP-heavy", "V", ".")), (0.25, ("N_a", "NP-heavy", "di" "V", "."))]
+
+productions["NP-heavy"] = [(1.0, ("Adj", "N_i", "P", "N_a"))]
+
+
+
+sentence1 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "obj_heavy"}, {"index" : 5, "word" : "di", "head" : 4, "dep" : "case"}, {"index" : 6, "word" : "N_a", "head" : 7, "dep" : "nsubj"}, {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+sentence2 = [{"index" : 1, "word" : "Adj", "head" : 2, "dep" : "amod"}, {"index" : 2, "word" : "N_i", "head" : 4, "dep" : "nmod"}, {"index" : 3, "word" : "P", "head" : 2, "dep" : "case"}, {"index" : 4, "word" : "N_a", "head" : 7, "dep" : "nsubj_heavy"}, {"index" : 5, "word" : "N_a", "head" : 7, "dep" : "obj"}, {"index" : 6, "word" : "di", "head" : 5, "dep" : "case"} ,    {"index" : 7, "word" : "V", "head" : -1, "dep" : "root"}, {"index" : 8, "word" : ".", "head" : 7, "dep" : "punct"}]
+
+
+sentences = [sentence1, sentence2]
+
+itos_deps = list(set([x["dep"] for x in sentence1+sentence2]))
+stoi_deps = dict(zip(itos_deps, range(len(itos_deps))))
+
+import torch
[…]

### 2019-01-20 toy/simulation.py

@@ -1,0 +1,266 @@
+
+
+productions = {}
+
+productions["S"] = [(0.5, ("Obj", "Subj", "V", ".")), (0.5, ("Subj", "Obj", "V", "."))]
+productions["Obj"] = [(1.0, ("NP", "di"))]
+productions["Subj"] = [(1.0, ("NP"))]
+productions["NP"] = [(0.75, ("N")), (0.125, ("PP", "NP")), (0.125, ("Adj", "NP"))]
+productions["PP"] = [(1.0, ("NP", "P"))]
+
+
+import numpy as np
+import random
+
+def sample(x):
+    if x in productions:
+        probs = [y[0] for y in productions[x]]
+        return " ".join([sample(z) for z in productions[x][np.random.choice(range(len(probs)), p=probs)][1]])
+    return x
+
+vocabulary = set()
+for x in productions:
+      for _, rhs in productions[x]:
+          for word in rhs:
+              if word not in vocabulary:
+                  vocabulary.add(word)
+itos = list(vocabulary)
+stoi = dict(zip(itos, range(len(itos))))
+
+print(sample("S"))
+
+
+
+
+import random
+import sys
+from collections import deque
+
+lengthForPrediction = 20 #int(sys.argv[4]) #20
+contextLength = 20 #int(sys.argv[5]) #20
+
+
+
+
+myID = random.randint(0,10000000)
+
+posUni = set() #[ "ADJ", "ADP", "ADV", "AUX", "CONJ", "DET", "INTJ", "NOUN", "NUM", "PART", "PRON", "PROPN", "PUNCT", "SCONJ", "SYM", "VERB", "X"] 
+
+posFine = set() #[ "``", ",", ":", ".", "''", "$", "ADD", "AFX", "CC",  "CD", "DT", "EX", "FW", "GW", "HYPH", "IN", "JJ", "JJR",  "JJS", "-LRB-", "LS", "MD", "NFP", "NN", "NNP", "NNPS", "NNS",  "PDT", "POS", "PRP", "PRP$", "RB", "RBR", "RBS", "RP", "-RRB-", "SYM", "TO", "UH", "VB", "VBD", "VBG", "VBN", "VBP", "VBZ",  "WDT", "WP", "WP$", "WRB", "XX" ]
+
+
+
+
+from math import log, exp
+from random import random, shuffle
+import os
+
+
+originalDistanceWeights = {}
+
[…]

### 2019-02-17 paper.tex
updating
paper

@@
 \title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
-\author{Michael Hahn}
+\author{Michael Hahn, Judith Degen, Richard Futrell}
 \date{2018}
@@
 
-\section{Two Memory-Surprisal Tradeoffs}
-
-With minimal assumptions, we will use information theory to derive two tradeoffs:
-
-(1) a tradeoff between \emph{speaker memory} and (listener) surprisal
-
-(2) a tradeoff between \emph{listener memory} and (listener) surprisal
-
-In the second part of the paper, we examine whether word orders in natural language optimise these two tradeoffs.
+\section{Memory-Surprisal Tradeoff}
+
+With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal
+
+In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.
 
@@
 
-What can we say about the amount of information that the speaker has to store when producing utterances -- that is, how many bits does it take to encode the speaker's  memory states $S_t$ and $L_t$?
-Our first theoretical result sheds light on this; we will show that the amount of bits needed to encode the memory state is related to the decay of mutual information over distances.
+We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
+The listener maintains a memory state $L_t$; there are no assumptions about the memory architecture and the nature of its computations.
+%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
+We only make two basic assumptions about the flow of information:
+
+First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
+Formally: 
+\begin{equation}\label{eq:listener-markov-1}
+L_t \bot X_{>t} | X_{\leq t}
+\end{equation}
+This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
+
+Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
+\begin{equation}\label{eq:listener-markov-2}
+L_t \bot X_{<t} | X_{t-1}, L_{t-1}
+\end{equation}
+This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
+
+The listener can trade off memory and future surprisal:
+A listener who chooses to store less memory will exerience higher surprisal in the future.
+A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
+
+We will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
+Consider a listener who uses $J$ bits of memory on average.
+What can we say about the listener's surprisal?
+We return to the information and memory curves in Figure~\ref{fig:basic}.
+In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
+This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
+Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
+We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
+As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
+Then
+
+\begin{proposition}\label{prop:suboptimal}
+	Let $T$ be a positive integer, and consider a listener using at most
+$$\sum_{t=1}^T t I_t$$
+bits of memory on average.
+Then this listener will incur surprisal at least
+	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
+	on average.
+\end{proposition}
+
+The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$, we obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
+We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
+\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
+	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
+Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
+In the given example, $T=4$.
+By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
+	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
+Recall that the red process had a faster decay of conditional mutual information.
+Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
+\end{figure}
+
+In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
+In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
+The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
+For reaching optimal surprisal, Order A requires strictly less memory.
+Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
+	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
+Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
+\end{figure}
 
@@
 
-Using this concept, we can now state our first main result, concerning the memory load of a speaker: %, and of a listener performing optimal incremental prediction:
-
-\begin{proposition}\label{prop:lower-bound}
-	A speaker producing samples from $(X_t)_{t \in \mathbb{Z}}$ needs to maintain at least
-\begin{equation}\label{eq:memory-bound}
-M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{equation}
-bits of memory on average.
-
-\end{proposition}
-
-Also can relax this to imprecise production: A speaker with low memory must deviate strongly from the language, measured in KL divergence.
-
 %that the model does not have an `absolute clock' built into it.
@@
 
-\subsection{Languages can trade off Memory Load and Predictability}
-%\subsection{Speakers can can minimize Memory Load at Fixed Information Rate}
-
-We have derived a general lower bound on the memory load of a speaker in Proposition~\ref{prop:lower-bound}.
-We now relate the speaker's memory to the surprisal experienced by the listener.
-
-The average surprisal of a language $(X_t)_{t \in \mathbb{Z}}$, as experienced by a listener performing optimal prediction, is equal to the entropy rate $H[X_t|X_{<t}]$.
-This can be written as the unigram entropy minus the mutual information between a single observation $X_t$ and the past:
-\begin{equation}
-	H[X_t|X_{<t}] = H[X_t] - I[X_t, X_{<t}]
-\end{equation}
-where the unigram entropy $H[X_t]$ only depends on the frequencies of individual words, independently of the language's word order.
-
-We will refer to the mutual information $I[X_t, X_{<t}]$ as the \emph{predictability} of the process.
-The predictability $P := I[X_t, X_{<t}]$ is a lower bound on $M = I[X_{\geq t}, X_{<t}]$, and thus on speaker memory:
-\begin{equation}
-P \leq M
-\end{equation}
-and thus\footnote{More generally 
-$H[X_t|X_{<t}] \geq \frac{1}{T} \left(H[X_{1\dots T}] - M\right)$ for any $T$.
-Thus, the following relation holds for any $T$: $$T \cdot \text{ListenerSurprisal} \geq H[X_{1\dots T}] - \text{SpeakerMemory}$$}
-\begin{equation}
-	Listener\ Surprisal \geq	H[X_t|X_{<t}] \geq H[X_t] - M \geq H[X_t] - Speaker\ Memory
-\end{equation}
-This means that, when we consider different languages that share a single unigram distribution, memory and surprisal trade off:
-Languages that are highly predictable  require a good amount of speaker memory; a language that can be produced with very small memory cannot be very predictable.
-In particular, a language that requires zero speaker memory must have zero predictability and maximal average surprisal.
-
-We thus obtain a tradeoff between two different types of costs facing speakers and listeners: The \emph{speaker's} memory trades off with the \emph{listeners'} surprisal.
-
-We can visualize this by placing languages into a \emph{memory-surprisal-plane} (Figure~\ref{fig:toy-mis}).
-Each language is assigned its entropy rate as its $y$ coordinate, and the memory load necessary to achieve optimal surprisal as its $x$ coordinate.
-We will continue to assume that all languages share a single unigram distribution -- this will in particular apply when studying different word order variations of the same language.
-
-The plane can be separated into an \emph{achievable} region, populated by languages, and an \emph{unachievable} region in which no languages can occur.
-No language can have zero surprisal with zero memory -- the corner of the plane thus must fall in the unachievable region.
-The boundary between the two regions is the \emph{Pareto frontier}.
-Languages that fall on the boundary optimally trade off surprisal and memory; they cannot further decrease one without increasing the other.
-The smallest possible values for this boundary consist of the points satisfying $P=M$, indicated by a straight diagonal line in Figure~\ref{fig:toy-mis}.
-Only part of the achievable region will be populated by languages that could plausibly be natural languages: For instance, we may expect natural languages to have long-term dependencies that will prevent them from being at the absolute Pareto frontier.
-We hypothesize that, if languages optimize the tradeoff between memory and surprisal, they will be close to the optimal boundary of a region populated by a `comparison class' of languages that share essential characteristics of natural language, but have other word order rules.
-%When We will specify a class of baseline languages that we will argue to form a reasonable comparison class for natural language.
-%Within this, we hypothesize that natural language is close to the Pareto frontier.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/pareto-scheme.pdf}
-%
-	\caption{Memory-Surprisal Tradeoff.
-The straight line represents the Pareto frontier for unconstrained languages.
-Languages resembling natural languages inhabit a subset of the achievable region (grey oval).
-Its lower-left boundary represents the Pareto frontier relevant for natural language.
-We hypothesize that natural language lies close to this frontier.
-}\label{fig:toy-mis}
-\end{figure}
-
-\subsection{Listeners can trade off Memory Load and Surprisal}\label{sec:listener}
-
-%TODO redo curves so that x-axis is surprisal, y-axis the necessary memory
-
-We have analyzed the memory load of a speaker and related it to the listener's surprisal.
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-Similar to the listener maintains a memory state $L_t$; again, there are no assumptions about the memory architecture and the nature of its computations.
-%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-We only make two basic assumptions about the flow of information:
-
-First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-Formally: 
-\begin{equation}\label{eq:listener-markov-1}
-L_t \bot X_{>t} | X_{\leq t}
-\end{equation}
-This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-
-Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-\begin{equation}\label{eq:listener-markov-2}
-L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-\end{equation}
-This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
-
-The listener can trade off memory and future surprisal:
-A listener who chooses to store less memory will exerience higher surprisal in the future.
-A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-We will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-Consider a listener who uses $J$ bits of memory on average.
-What can we say about the listener's surprisal?
-We return to the information and memory curves in Figure~\ref{fig:basic}.
-In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
-This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
-Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
-We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
-As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-Then
-
-\begin{proposition}\label{prop:suboptimal}
-	Let $T$ be a positive integer, and consider a listener using at most
-$$\sum_{t=1}^T t I_t$$
-bits of memory on average.
-Then this listener will incur surprisal at least
-	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-	on average.
-\end{proposition}
-
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$, we obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
-We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
-Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
-In the given example, $T=4$.
-By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
-	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
-Recall that the red process had a faster decay of conditional mutual information.
-Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
-\end{figure}
-
-In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
-The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
-For reaching optimal surprisal, Order A requires strictly less memory.
-Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
-	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
-\end{figure}
-
 \paragraph{Center Embeddings}
@@
 
-\subsection{Relation to Theories of Sentence Processing}
+\section{Relation to Theories of Sentence Processing}
 
@@
 \paragraph{Statistical Complexity}
-There are deep connections between our formalization of speaker and listener memory and studies of dynamic systems in the Physics literature.
+There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
 Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
@@
 
-\section{Heavy NP Shift}
-
-Phenomenon linked to locality/memory (Hawkins)
-
-Is the listener's memory-surprisal tradeoff optimized by Heavy NP Shift?
-
-Penn Treebank
-
-Trained 20 language models counterfactual version
-
-then estimate memory on real and reverse orderings
-
-3) Describe the key dependent variable(s) specifying how they will be measured.
-Collect all instances of a VP consisting of a verb, and (in some order) an NP and a PP, from the Penn Treebank (WSJ) training set.
-Calculate the listener's memory-surprisal tradeoff on the words belonging to the NP and the PP.
-
-5) Specify exactly which analyses you will conduct to examine the main question/hypothesis.
-(1) Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs?
-Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP?
-This will be tested by bootstrapping over language models and examples.
-(2) Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?
-Also tested by bootstrapping over language models and examples.
-
-6) Describe exactly how outliers will be defined and handled, and your precise rule(s) for excluding observations.
-Language models with overall validation average surprisal > 10 will be excluded (this didn't happen).
-
-7) How many observations will be collected or what will determine sample size?
-20 LSTM language models are constructed on a version of Penn Treebank where for all verb-initial VPs, the order of the remaining daughter constituents is shuffled.
-
-Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs?
-
-Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP?
-
-Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?
-
-%
-%
-%\section{NP Structure}
-%
-%- adjective order
-%
-%- noted by ...: Det - Num - Adj
-%
-%
-%This is a case where the predictions differ from DepL
-%
-%
-%E.g., if the choice of determiners, numerals, and adjectives are statistically independent when conditioning on the noun, will predict the effect.
-%Computationally verify the effect.
-
 \section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
@@
 
-Ordering grammars are defined as in the Optimization paper: For each UD relation type, define a weight. To get random baseline languages, randomize the weights.
-All grammars are valuated deterministically.
-
-We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
-Thus, random grammars with by-relation weights constitute a stronger baseline.
-
-%Things that one may ponder: (1) what are even stronger baselines? (2) probably should do a control experiment using maximum-likelihood fits
-
-%\subsection{Prediction on the level of POS}
-%As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences.
-%, following~\cite{futrell-noisy-context-2017}.
-%As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.
-%
-%%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages.
-For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
-%%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
-%%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
-%%
-%%%\mhahn{TODO results across languages}
-%%
-%%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
-%
-%Results for four languages are shown in Figure~\ref{fig:memory-results}.
-%Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
-%%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
-%
-%
-%%%For half of the languages
-%%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
-%%For 37 \% of the languages, it also requires less memory than the original language.
-%
-%%We found that, as predicted, memory usage in real languages was correlated with word order freedom:
-%
-%
-%%\begin{figure*}
-%%\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
-%%%
-%%\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction.
-On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
-%%\end{figure*}
-%%
+\paragraph{Ordering Grammrs}
+We define ordering grammars, small interpretable models of dependency tree linearization.
+
+Every dependency type $\tau$ is assigned a parameter $a_\tau \in [0,1]$.
+Dependency trees are linearized by the following algorithm:
+
+%Every dependency type $\tau$ is assigned a direction parameter $\alpha_\tau$ and a distance parameter $\beta_\tau$.
+
+\paragraph{Baseline Languages}
+We define baseline grammars by randomly sampling the parameters $a_\tau$.
 
@@
 
-This experiment is not fully confirmatory, since some data was collected for some languages before the predictions and the setup of the experiment were finalized.
-
 For each language, we collected data from the actual orderings and from random grammars.
@@
 
-\paragraph{Speaker Memory: Distance from Pareto Frontier}
-We want to test: Are languages closer to the Pareto frontier than baseline languages tend to be?
-
-For this, we need a way of quantifying distance to the Pareto frontier.
-We suggest to measure the distance as the fraction of baseline languages that are better on both dimensions.
-
-More formally, fixing a natural language, let $P_0$ be the distribution of baseline languages mapped to the surprisal-memory plane, viewed as a probability distribution over the positive quadrant of the plane $\mathbb{R}_+^2$.
-For any point $x \in \mathbb{R}_+^2$, let $A_x$ be the fraction of baseline languages that are better on both dimensions.
-Similarly, let $B_x$ be the fraction of baseline languages that are worse on both dimensions.
-Using the probability distribution $P_0$, we can write:
-$$A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$$
-$$B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$$
-%Ways of using this to test and quantify for distance (I think they are all reasonable and worth reporting in SI. Question is which ones to rely on mostly for the main part):
-Futher, let $P_1$ be the distribution in the surprisal-memory plane that is generated by the results obtained for the real orderings, with different random initializations.
-For each language, we \emph{test} the hypothesis that the language optimizes the tradeoff by considering the null hypothesis that the real orderings do not Pareto-dominate more baseline orderings than there are baseline orderings dominating them:
-$$\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$$
-
-We propose to \emph{quantify} closeness to the frontier using the following two statistics:
-
-\begin{enumerate}
-%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
-	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$
-
-		This has the following interpretation: Among orderings that either dominate or are dominated by the real orderings, what fraction is dominated by the real orerings?
-		This quantity is $1$ if the real orderings are not Pareto-dominated by any baseline order, but do dominate some baseline orders.
-		The null hypothesis defined in our test amounts to the inequality $\E_{x \sim P_1}E_x \leq 0.5$.
-
-A drawback of this quantity is that it is meaningless when the real orders do not dominate, and are not dominated by, any real order.
-The following quantity is always maningful:
-
-	\item Compute $A_{x'}$ for each $x'$ from the baseline distribution $P_0$, and then calculate the percentile rank of $\E_{x \sim P_1}[A_x]$ among these values:
-		$$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$$
-This statistic describes how many random orders achieve closer distance to the frontier as measured by $A_x$.
-		Stated differently, it quantifies how `extraordinary' the value $\E_{x \sim P_1}[A_x]$ is in relation to values typically attained for $x' \sim P_0$.
-The statistic is close to zero if $\E_{x \sim P_1}[A_x]$ is smaller than for almost all samples from $P_0$.
-
-\end{enumerate}
-
-We will obtain confidence intervals for the statistics via bootstrapping, and use a confidence interval for $B_x-A_x$ for testing $B_x \leq A_x$.
-In the appendix, we report simulations that justify the use of bootstrapping in this situation.
-
 \paragraph{Listener Memory: Surprisal-Memory Tradeoff}
@@
 
-For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $D$ had width $\leq 0.3$, and the CI for $W$ had width $\leq 0.15$.
+For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $W$ had width $\leq 0.15$.
 We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
@@
 
-Results are reported in Appendix~\ref{sec:results-table}.
-
-\paragraph{Surprisal and Speaker Memory}
-In all but X languages, the null hypothesis $\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$ was rejected.
-Also, in all but X languages, $D < 0.5$.
-
-\paragraph{Surprisal and Listener Memory}
 In all but X languages $\E[W] > 0.5$ significantly.
@@
 \section{Experiments on Further Treebanks}
-
-Held-out:
-
-- UD languages:
-
--- English, Korean, Russian
-
--- UD$\_$Polish-LFG (released in 2.2, not included in original experiment)
-
--- character-level Russian
-
-- other treebanks
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017}
-
--- Chinese treebank \citep{xue-chinese-2013}
-
--- furher (2.2/2.3 Japanese treebanks)
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
 
@@
 
+\section{Experiment with MLE Grammars}
+
+\paragraph{Maximum-Likelihood Grammars}
+We additionally create ordering grammars that are fit to the actual orderings of each language.
+These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
+
+We construct these grammars by interpreting ordering grammars probabilistically, and setting the parameters
+
+We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
+Thus, random grammars with by-relation weights constitute a stronger baseline.
+
+\section{N-Gram Models}
+
+\section{Morphology-Based Encoding}
+
+Held-out:
+
+- UD languages:
+
+-- English, Korean, Russian
+
+-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
+
+-- character-level Russian
+
+\section{Character-Level Modeling}
+
+\section{Non-UD Dependency Treebanks}
+
+- other treebanks
+
+-- spoken Japanese (T{\"u}ba-J/S)
+
+-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
+
+-- another Chinese dependency treebank LDC2012T05
+
+Due to the sizes of these treebanks, can also do experiment with full word forms.
+
+\section{Constituency Treebank}
+
+-- Penn treebank \citep{marcus-building-1993}
+
+-- spoken English (T{\"u}ba-E/S)
+
+-- spoken German (T{\"u}ba-D/S)
+
+-- Chinese treebank \citep{xue-chinese-2013}
+
 \section{Proofs}
-
-\begin{proof}
-	The number of bits of memory is given by the entropy $H[S_t]$.
-	In view of (\ref{eq:markov-speaker}) and the Data Processing Inequality, this is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
-Then the formula is a consequence of the chain rule for mutual information:
-\begin{align*}
-	I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I[X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}] \\
-	&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I[X_i, X_j|X_{j+1}...X_{i-1}] \\
-& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{align*}
-where we used stationarity of the process in the last step.
-\end{proof}
 
@@
 
-\section{Bootstrapping for Distance to Frontier}
-
-TODO write up
-
-We conducted simulations in the following setting:
-
-sample two ellipses such that
-
-(1) $\mu_1 \sim Unif([-0.5, 0.5]^2)$
-
-(2) $\mu_2 \sim Unif([-1.5, 1.5]^2)$
-
-$A_1, A_2 \sim Unif(\mathbb{R}^2 \times \mathbb{R}^2)$, $\Sigma_i = A_i A_i^T$ subject to the wo eigenvalues differing by a factor of $\leq 2$, and $\Sigma_1$ having area 1, $\Sigma_2$ having area 1/20.
-
-Computed ground-truth $A_x, B_x$ from 3000 samples from both ellipses.
-
-Bootstrapping: get 0.001, 0.999
-
-Results:
-
-For $A_x-B_x$: good coverage
-
-For $E_x$: intervals are unreliable if $A_x + B_x$ is small
-
-For $D_x$: good coverage
-
 \section{Results (Morphology on UD Treebanks)}\label{sec:results-table}
@@
 
-\section{Thoughts}
-Upper bound: Use $TH[X_t] - \sum_{t=1}^T (T-t) I_t = T (H[X_t] - \sum_{t=1}^T I_t) + \sum_{t=1}^T t I_t$ bits of memory and incur $\sum_{t>T} I_t$ in additioal surprisal.
-
-\begin{tabular}{cccccccccccccc}
-Afrikaans & Amharic & Arabic & Armenian & Basque & Breton & Bulgarian & Buryat \\
-\includegraphics[width=0.14\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Arabic-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Basque-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}
-\\
-Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & Estonian \\
-\includegraphics[width=0.14\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Catalan-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Chinese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Croatian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Czech-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Danish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Dutch-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}
-\\
-Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian \\
-\includegraphics[width=0.14\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Finnish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/French-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/German-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Greek-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hindi-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}
-\\
-Indonesian & Japanese & Kazakh & Kurmanji & Latvian & Naija & North Sami & Norwegian \\
-\includegraphics[width=0.14\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Japanese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Latvian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}
-\\
-Persian & Polish & Romanian & Serbian & Slovak & Slovenian & Spanish & Swedish \\
-\includegraphics[width=0.14\textwidth]{figures/Persian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Polish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Romanian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Serbian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovak-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Spanish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}
-\\
-Tamil & Thai & Turkish & Ukrainian & Uyghur & Vietnamese \\
-\includegraphics[width=0.14\textwidth]{figures/Tamil-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Turkish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}
-\\\end{tabular}
-
 \end{document}
-
-\appendix
-\section{Bounding Memory in Imperfect Modeling}
-We derived the nonlocality $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
-What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?
-
-The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
-To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
-The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
-In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
-However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
-(Other models would incur even higher average surprisal, without reduction in memory cost.)
-The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.
-
-This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
-%Naturally, we expect that lower surprisal rates will require more memory.
-
-%\subsection{Lower bound on approximation with n-th order Markov models}
-%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
-%We know $\sum_{t>T} I_{t}$ 
-%If a Markov model has order $n$
-
-There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
-\begin{proposition}\label{prop:markov}
-The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
-\end{proposition}
-
-In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.
-
-\subsection{Lower bound on imperfect approximation} 
-Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.
-
-As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
-How much greater will their surprisal be?
-In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$.
-We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1.
-In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
-However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
-%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
-%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
-We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-
-The following result holds for any process:
-\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
-%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
-%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
-%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
-%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
-%Choose any $T \in \mathbb{N}$.
-Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
-$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
-%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
-\end{proposition}
-%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$.
-It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
-%An equivalent way of stating this bound is:
-Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
-Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
-I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}
-
-\begin{proof}
-Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
-As explained above, without loss of generality, we can assume that the model is Bayes optimal.
-%: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
-By definition, the difference between the model's surprisal and optimal surprisal is
-\footnote{The first inequality seems really crude.
-In general, we cannot do better.
-If we know more about the approximating model, we might be able to improve this.
-Also, consider the relation to the other inequality.
-Can they be attained at the same time?}
-\begin{align*}
-%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
-%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
-%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
-& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
-& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
-& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
-& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
-\end{align*}
-\end{proof}
-%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$.
-Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
-%
-%\begin{corollary}
-%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
-%
-%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
-%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
-%
-%\end{corollary}
-%
-%
-
-Here is way of phrasing the same statement in terms of memory:
-\begin{corollary}
-Let $\epsilon > 0$.
-Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
-\end{corollary}
-
-\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
-Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
-Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
-Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
-While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
-If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.
-
-\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
-%What is the optimal $T$?
-%It turns out that the optimal $T$ 
-Searching for the optimal $T$ in the bound, we find:
-\begin{corollary}
-Consider any memory requirement $J \in [0,M)$.
-Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
-Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
-Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-That is, it cannot do better than a $T$-th order Markov model.
-\end{corollary}
-Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.
-
-% a $T$-th order Markov
-
-Here is a way of rephrasing this in terms of memory demand:
-\begin{corollary}
-Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$.
-Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
-\end{corollary}
-
-%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
-%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
-%
-%Do these results agree with the idea of information locality decreasing memory load?
-%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
-%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
-%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
-%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
-%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
-%
-
-\begin{example}
-%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
-%
-Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
-We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.
-
-Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
-$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
-To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
-$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
-and, plugging back into the bound, get that the model's surprisal is at least
-$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
-which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
-Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
-\end{example}
-
-%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns
-
-%\begin{corollary}[Approximation by N-Gram Models, lower bound]
-%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
-%Then
-%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
-%
-%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
-%\end{corollary}
-%
-%\begin{proof}
-%
-%\end{proof}
-%
-%
-
-\subsection{Upper Bounds on imperfect approximation}
-$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy.
-Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$.
-Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else.
-We consider the problem of coding $X$ with small memory to predict $Y$ optimally.
-We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$.
-In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large.
-The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}
-
-However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
-We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.
-
-%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).
-
-\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
-The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
-%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
-\end{proposition}
-\begin{proof}
-In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
-(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
-\end{proof}
-%Compared with the lower bound from the previous section, we see:
-Computing the memory cost of a an n-th order Markov model, we can conclude:
-\begin{corollary}
-(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.
-
-(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$
-
-\end{corollary}
-\begin{proof}
-By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
-This model can be implemented using 
-$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
-This bound is really crude. We can compute
-\begin{align*}
-H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
-%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
-%&\leq I/\epsilon H[X] + I \\
-%& = I(1+H[X]/\epsilon) \\
-\end{align*}
-but does this gives us anything?}
-The second statement is a rephrasing of the first.
-\end{proof}
-As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
-In the case of natural language, $H[X]$ is a reasonably small constant.
-
-%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
-%\begin{corollary}[Approximation by $n$-gram models]
-%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
-%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
-%\end{corollary}
-
-%\subsection{Summary}
-
-%
-%\paragraph{Invariance properties under certain transformations of the sequence}
-%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
-%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
-%%Second, 
-%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
-%If we measure memory also at all new boundaries, $I$ can at most increase.
-%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
-%
-%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
-%
-
-\section{Crypticity}
-
-%We let $X_t$ be a process of observations (e.g. words), and $M_t$ be the internal memory states of the listener, which we can take to be a function of $X_{\leq t}$.
-Under the assumption of optimality oif predictions and minimality of memory cost, we can take $M_t$ to be an equivalence class of pasts that induce the same distribution over futures (need to be careful for general processes.
-we here want to add the assumption that, since we are interested in sentence processing, memory is reset entirely at sentence or discourse boundaries, avoiding cases where the math wouldn't work so nicely).
-
-Nonlocality:
-\begin{equation}\label{eq:locality-cost}
-L := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{equation}
-
-Crypticity:
-\begin{equation}\label{eq:ambiguity-cost}
-A := H[M_0|X_{> 0}]
-\end{equation}
-
-Note that both terms are nonnegative by definition.
-
-\begin{proof}
-By the definitions of conditional entropy and mutual information,
-$H[M_0] = H[M_0 | X_{>0}] + I[M_0, X_{>0}]$
-The first part is Ambiguity cost.
-Now we show that the second part is equal to Locality cost.
-As we assume that the listeners predictions are optimal, $I[M_0, X_{>0}] = I[X_{\leq 0}, X_{>0}]$.
-Then the formula is a consequence of the chain rule for mutual information:
-\begin{align*}
-I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
-&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
-& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{align*}
-\end{proof}
-
-\end{document}
-
-\section{Normalized Excess Entropy}
-
-We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$
-
-Why?
-
-(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity
-
-(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?
-
-\section{Computing a tighter lower bound}
-
-Here is a method that takes a quadratic amount of time to compute:
-
-Take all pairs of positions in the corpus, and for each pair $s, t$ compute
-$$P(X_{t, ..., t+T}|T_{s-T, ..., s-1})$$
-After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
-$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t, ..., t+T}|T_{s-T, ..., s-1}) + \log P(X_{t, ..., t+T}|T_{t-T, ..., t-1})\right) < \epsilon$$
-Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) lower bound of memory for modeling with KL loss $\epsilon$.
-However, this bound might still not be tight, unless we require that $\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$.
-
-Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).
-
-It should be feasible to apply this to POS-level language modeling.
-
-ACTUALLY
-
-Take all pairs of positions in the corpus, and for each pair $s, t$ compute
-$$P(X_{t}|T_{s-T, ..., s-1})$$
-After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
-$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t}|T_{s-T, ..., s-1}) + \log P(X_{t}|T_{t-T, ..., t-1})\right) < \epsilon$$
-and
-$\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$
-Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) estimator of a tight lower bound of memory for modeling with KL loss $\epsilon$.
-
-Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).
-
-It should be feasible to apply this to POS-level language modeling.
-
-Is this a consistent estimator?
-
-\section{Pointwise Bounds}
-
-\begin{proposition}
-Let $F := X_{>0}$, $P := X_{\leq X}$. Then pointwise cost for an optimal representation (in terms of average memory cost) is
-%$$C(p) := \log P\left{p' : P(F|p') = P(F|p)$$
-\end{proposition}
-
-There are several pointwise versions of excess entropy.
-Here is one:
-$$D(p) := \E_{f|p} pmi(f,p) = D_{KL}(P(f|p) || P(f))$$
-
-\begin{proposition}
-$D(p) \leq C(p)$
-\end{proposition}
-
-\begin{proof}
-\begin{align*}
-D(p) &= \sum_f p(f|p) \log \frac{p(f|p)}{p(f)} \\
-&= \sum_f p(f|p) (\log p(p|f) - \log p(p)) \\
-&= \left( \sum_{f} p(f|p) \log p(p|f)\right) - \log p(p)
-\end{align*}
-and the left quantity is $\leq 0$.
-\end{proof}
-
-$D$ can be estimated at each specific point by sampling continuations.
-
-TODO $D(p)$ can be seen as a formalization of DLT storage cost
-
-\begin{corollary}
-If $\pi(p)$ is the equivalence class of the past, then
-pointwise memory cost is
-$$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
-\end{corollary}
-Taking the expectation over $p$, the left component is the excess entropy, while the right component is the gap between excess entropy and state entropy:
-$$\E{f,p} \log(\pi(p)|f) = H[\pi(P) | F]$$
-
-$H[\pi(P)] = I[P,F] + H[\pi(P) | F] = I[\pi(P), F] + H[\pi(p) | F]$
-
-Memory = Non-Locality + Difficulty of Guessing Past from Future
-
-I will call excess entropy `locality cost', and the second term `ambiguity cost'.
-
-This decomposition also holds if $\pi$ is lossy. If $\pi$ is lossy, the non-locality term has to decrease.
-
-Informally $$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
-decomposes into a locality/storage cost term $\E_{f|p} pmi(f,p)$ (can be seen as a formalization of DLT storage cost), and $- \E_{f|p} \log(\pi(p)|f)$.
-The latter component quantifies how hard it is to guess the storage content from the future (difficulty of guessing past from future, given the past).
-One possible interpretation is in terms of noisy-channel memory:
-Under memory constraints, more frequent/typical pasts will be more likely to be imputed based on the more recent observations. `Preventing a wrong noisy-channel inference takes extra memory effort.'
-Can also relate this to prefix ambiguity, in the case where the ambiguity doesn't always turn up: E.g., if it is hard to predict from a continuation whether the prefix had been temporarily ambiguous.
-
-Interpreting Ambiguity Cost:
-
-- not knowing which things will be relevant later is costly
-
-- one instantiation: not knowing which referents will be referred back to later
-
-- one instantiation: ambiguous prefixes are costly
-
-can you show that languages also optimize ambiguity cost?
-
-\section{Masha's experiments}
-
-- DepL experiment: excess entropy, we can even trace the pointwise profile
-
-- MaxOP: initial NP can give rise to three memory states: subj, obj, indeterminate.
-The remainder of the sentence does not reveal whether the first NP had a case marker (if it did reveal that, this would increase locality cost).
-Ambiguity cost can be decreased by avoiding the indeterminate state (but there are other, even better minima!
--- so this doesn't predict as much as MaxOP).
-
-\section{Estimating Memory Cost and Ambiguity Cost}
-
-by actually estimating a memory representation: estimate a variable that is independent of the future when conditioning on the past, and also: independent of past when conditioning on last word and last state of memory
-
-if memory is deterministic, just say that it is a function of $X_t, M_{t-1}$.
-
-if memory is randomized:
-
-formally: $M_t$ must have the properties (1) $M_t \bot X_{<t} | (M_{t-1}, X_t)$, (2) $M_t \bot X_{>t} | X_{\leq t}$.
-
-In this case, cost measure should be MI with past, not entropy
-
-randomized memory is useful because we can apply gradient descent, solving an optimization problem to provide an upper bound on memory cost
-
-- solving a discrete optimization problem (probably not super scalable)
-
-- variational: stochastic continuous memory representations and take memory cost to be $I[Memory, Past]$ (also done in IB)
-
-(noise the LSTM hidden state in each time step)
-
-calculate upper bound $D_{KL}(P(Memory|Past) || P_{prior}(Memory)) \geq I[Memory, Past]$, estimate the tradeoff curve between this and cross-entropy (the resulting curve is a variational upper bound on the real curve).
-
-- other ways of minimizing MI with past (don't require noising of the hidden state):
-
--- adversarial: make the past hard to predict
-
--- dual lower bound on MI
-
-both of these methods are ways of doing IB
-
-\section{Incremental Prediction under Memory Limitations}
-
-%\subsection{Noisy-Context Surprisal}
-%
-%\citet{futrell-noisy-context-2017} describe a psycholinguistic noisy-channel model of human language processing under memory limitations where words are progressively deleted from memory.
-%with progressive probability.
-%They show that processing as described by this model favors word order where linguistic material that has high mutual information will appear closer together, in the sense that the processing cost, measured by the surprisal $-\log P(w_t | w_1 ...
-w_{t-1})$ will be
-%$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-%where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-%Given that $f$ is decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-%
-%%In the most extreme case of such a model, a bigram model, we have:
-%%$$-\log P_{Model}(w_t | w_1 ... w_{t-1}) = -\log P(w_t) - pmi(w_t, w_{t-1})$$
-%%where $PMI(X,Y) = \log P(X,Y) - \log P(X) - \log P(Y)$ is the pointwise mutual information.
-%%
-%%This directly applies to n-gram models: Here, words are indeed `forgotten' deterministically when they are more than $n$ words in the past.
-%
-%%\citet{futrell-noisy-context-2017} show that this model derives a range of psycholinguistic locality effects.
-%%It also applies to models with strictly bounded context, such as n-gram models -- in this case, past words beyond a fixed distance threshold are entirely erased from memory.
-%%However, it is less clear that human language processing (or state-of-the-art NLP techniques such as LSTMs) are affected by such loss -- it is not clear that humans entirely forget previous words, and perfectly remember the other words.
-%
-%\subsection{Memory in General Sequence Models}
-
-We consider a general incremental model that maintains a memory state and, in each time step, computes a distribution over the vocabulary based on its memory state, stochastically emits a word from this distribution, and updates its memory state depending on the chosen word.
-We are making the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
-This model describes aspects of human language processing, if we accept that humans perform incremental prediction.
-(Typical NLP models such as n-gram models and LSTM language models also fit into this framework.)
-%N-gram models and LSTM language models both fit into this framework.
-%On the other hand, LSTM models with random access to prior words do not fit into this framework, as they rely on past states in addition to their internal memory state at time $t$.
-
-In this general framework, the minimal amount of bits needed to encode the model's internal memory state is directly determined by the decay of mutual information over distances:
-The faster mutual information between words decays as their distance increases, the less memory is required.
-More precisely:
-
-\begin{proposition}
-Modeling a stationary process  $(X_t)_{t \in \mathbb{Z}}$ in this framework %of incremental modeling described above,
- requires maintaining at least
-\begin{equation}\label{eq:memory-bound}
-M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{equation}
-bits of memory,
-where
-$\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ is the mutual information of the observations $X_t$ and $X_0$, conditioned on the intervening observations.
-\end{proposition}
-%that the model does not have an `absolute clock' built into it.
-This quantity is also known as \emph{excess entropy} (CITE).
-
-\begin{proof}
-%We assume that the process is stationary, that is, that there is no `absolute clock' built into the model.
-Formally, this means that, for any set of indices $t_1 < t_2 < \dots < t_n$, the joint distribution $(X_{t_1}, X_{t_2}, ..., X_{t_n})$ only depends on the differences $(t_{i+1}-t_i)_{i=1,...,n-1}$, not on the absolute values $t_i$.
-The number of bits of memory is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
-Then the formula is a consequence of the chain rule for mutual information:
-\begin{align*}
-I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
-&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
-& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
-\end{align*}
-%In the last step, we have made the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
-%that the model does not have an `absolute clock' built into it.
-\end{proof}
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-
-The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works would be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-%as long the model describes the process $(X_t)_{t \in \mathbb{Z}}$
-
-%\section{Estimating Memory Load in Prediction}
-
-%We want to apply (\ref{eq:memory-bound}) on corpus data to compute the memory load of language for incremental prediction.
-
-For our purposes, it is important to note that we can estimate (\ref{eq:memory-bound}) from a statistical model assigning probabilities to sentences.
-(TODO summarise prior work attemtpting to estimate the same quantity, if there is any such work).
-
-\section{Bounding Memory in Imperfect Modeling}
-We derived $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
-What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?
-
-The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
-To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
-The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
-In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
-However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
-(Other models would incur even higher average surprisal, without reduction in memory cost.)
-The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.
-
-This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
-%Naturally, we expect that lower surprisal rates will require more memory.
-
-%\subsection{Lower bound on approximation with n-th order Markov models}
-%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
-%We know $\sum_{t>T} I_{t}$ 
-%If a Markov model has order $n$
-
-There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
-\begin{proposition}\label{prop:markov}
-The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
-\end{proposition}
-
-In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.
-
-\subsection{Lower bound on imperfect approximation} 
-Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.
-
-As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
-How much greater will their surprisal be?
-In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$.
-We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1.
-In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
-However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
-%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
-%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
-We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-
-The following result holds for any process:
-\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
-%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
-%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
-%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
-%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
-%Choose any $T \in \mathbb{N}$.
-Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
-$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
-%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
-\end{proposition}
-%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$.
-It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
-%An equivalent way of stating this bound is:
-Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
-Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
-I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}
-
-\begin{proof}
-Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
-As explained above, without loss of generality, we can assume that the model is Bayes optimal.
-%: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
-By definition, the difference between the model's surprisal and optimal surprisal is
-\footnote{The first inequality seems really crude.
-In general, we cannot do better.
-If we know more about the approximating model, we might be able to improve this.
-Also, consider the relation to the other inequality.
-Can they be attained at the same time?}
-\begin{align*}
-%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
-%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
-%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
-& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
-& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
-& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
-& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
-\end{align*}
-\end{proof}
-%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$.
-Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
-%
-%\begin{corollary}
-%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
-%
-%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
-%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
-%
-%\end{corollary}
-%
-%
-
-Here is way of phrasing the same statement in terms of memory:
-\begin{corollary}
-Let $\epsilon > 0$.
-Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
-\end{corollary}
-
-\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
-Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
-Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
-Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
-While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
-If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.
-
-\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
-%What is the optimal $T$?
-%It turns out that the optimal $T$ 
-Searching for the optimal $T$ in the bound, we find:
-\begin{corollary}
-Consider any memory requirement $J \in [0,M)$.
-Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
-Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
-Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-That is, it cannot do better than a $T$-th order Markov model.
-\end{corollary}
-Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.
-
-% a $T$-th order Markov
-
-Here is a way of rephrasing this in terms of memory demand:
-\begin{corollary}
-Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$.
-Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
-\end{corollary}
-
-%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
-%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
-%
-%Do these results agree with the idea of information locality decreasing memory load?
-%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
-%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
-%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
-%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
-%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
-%
-
-\begin{example}
-%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
-%
-Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
-We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.
-
-Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
-$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
-To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
-$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
-and, plugging back into the bound, get that the model's surprisal is at least
-$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
-which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
-Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
-\end{example}
-
-%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns
-
-%\begin{corollary}[Approximation by N-Gram Models, lower bound]
-%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
-%Then
-%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
-%
-%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
-%\end{corollary}
-%
-%\begin{proof}
-%
-%\end{proof}
-%
-%
-
-\subsection{Upper Bounds on imperfect approximation}
-$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy.
-Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$.
-Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else.
-We consider the problem of coding $X$ with small memory to predict $Y$ optimally.
-We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$.
-In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large.
-The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}
-
-However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
-We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.
-
-%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).
-
-\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
-The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
-%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
-\end{proposition}
-\begin{proof}
-In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
-(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
-\end{proof}
-%Compared with the lower bound from the previous section, we see:
-Computing the memory cost of a an n-th order Markov model, we can conclude:
-\begin{corollary}
-(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.
-
-(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$
-
-\end{corollary}
-\begin{proof}
-By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
-This model can be implemented using 
-$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
-This bound is really crude. We can compute
-\begin{align*}
-H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
-%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
-%&\leq I/\epsilon H[X] + I \\
-%& = I(1+H[X]/\epsilon) \\
-\end{align*}
-but does this gives us anything?}
-The second statement is a rephrasing of the first.
-\end{proof}
-As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
-In the case of natural language, $H[X]$ is a reasonably small constant.
-
-%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
-%\begin{corollary}[Approximation by $n$-gram models]
-%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
-%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
-%\end{corollary}
-
-%\subsection{Summary}
-
-%
-%\paragraph{Invariance properties under certain transformations of the sequence}
-%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
-%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
-%%Second, 
-%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
-%If we measure memory also at all new boundaries, $I$ can at most increase.
-%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
-%
-%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
-%
-
-\section{Normalized Excess Entropy}
-
-We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$
-
-Why?
-
-(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity
-
-(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?
-
-\section{Excess Entropy and Hierarchical Structure}
-
-Finiteness of excess entropy is not in conflict with the presence of hierarchical structure and long-range dependencies in language:
-Average memory demand can be small if the probability mass assigned to long dependencies is small.
-
-Indeed, perhaps surprisingly, languages defined by PCFGs always have finite average memory (and thus in particular finite excess entropy):
-
-\begin{proposition}
-Any language generated by a PCFG\footnote{With rational weights, and finitely many terminals and nonterminals} has finite average memory demand.
-
-\end{proposition}
-
-\begin{proof}
-We can consider incremental Earley parsing.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory on average.
-For storing probabilities, the storage will be similarly bounded.
-The result follows from (CITE).
-\end{proof}
-
-\begin{example}
-Consider $a^nb^n$ with the probability assignment given by the PCFG $X \rightarrow \epsilon$ ($1-p$), $X \rightarrow aXb$ ($p$), for some $0<p<1$.
-
-To compute  excess entropy, we construct a stationary process by considering concatenations of i.i.d. samples from the language.
-
-%Also, the associated stationary process has finite excess entropy.
-
-\end{example}
-
-\section{Related Work}
-
-derivation of information locality from lossy-context surprisal
-
-Crutchfield, L{\"o}hr: construct HMMs that provide upper bounds on memory cost.
-Should we be studying such upper bounds? I think that the lower bound that we study here is more useful due to the bounds it provides on imperfect modeling. 
-The example from \cite{harsha-communication-nodate} shows that the true minimum memory cost may be a lot larger than the lower bound $M$ (it is exponential in $M$ for the family of processes that they considered), while -- for fixed KL loss and unigram entropy -- the memory required for imperfect modeling is $O(M)$, as we saw in the previous section.
-
-Debowski: compute excess entropy in text
-
-\cite{sharan-prediction-2016}: link between excess entropy and approximability by $n$-th order Markov models %study approximability by n-th order Markov models
-
-\bibliographystyle{apalike}
-%\bibliography{extracted}
-\bibliography{literature}
-
-\end{document}
-
-\section{Memory and Cross-Linguistic Word-Orders}
-
-Do word orders in language minimize memory need as quantified by (\ref{eq:memory-bound}), when compared to other possible orderings?
-
-\subsection{Prediction on the level of POS}
-As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences.
-%, following~\cite{futrell-noisy-context-2017}.
-As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.
-
-%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages.
-For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
-%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
-%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
-%
-%%\mhahn{TODO results across languages}
-%
-%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
-
-Results for four languages are shown in Figure~\ref{fig:memory-results}.
-Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
-%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
-
-%%For half of the languages
-%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
-%For 37 \% of the languages, it also requires less memory than the original language.
-
-%We found that, as predicted, memory usage in real languages was correlated with word order freedom:
-
-\begin{figure*}
-\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
-%
-\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction.
-On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
-\end{figure*}
-
-\subsection{Incorporating Morphology}
-In many languages, morphology may have a big impact on prediction, but it is entirely ignored if we compute memory need for predicting POS tags.
-Accounting for morphology requires a more sophisticated statistical model of language that is able to encode morphosyntatic generalizations.
-
-The following models might be options, considering the NLP literature:
-
-\begin{enumerate}
-\item an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of many UD corpora and free word order in many languages with rich morphology thus seem to make such models unattractive.
-
-\item a statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar.
-The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-The second option will make this a machine learning model that, for our purposes, does not seem to be superior to (3).
-% (it kept many NLP people busy for the 2000s...)
-\item a recurrent neural network (in particular, LSTM): NLP research in the past years shows that such models consistently fit language data better than (1) and (2) (and any other model that does not have a neural component).
-The model family seems rich enough to learn morphosyntactic generalizations and deal with free word order.
-\end{enumerate}
-
-(3) clearly seems to be the best choice.
-
-As LSTMs are sequential models, the question is how to encode the input, including morphological information, into a sequence of tokens.
-%On option would be to encode words as sequences of characters, or morphemes.
-The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
-We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token.
-Each POS tag and morphological feature-value pair is encoded as an atomic token.
-
-During incremental prediction, the LSTM sequentially reads these tokens and predicts the next one.
-The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).
-
-Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
-Therefore, to compute this quantity, we compute the differences between the surprisals
-$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.
-
-\section{Measuring Information Locality via Normalized Excess Entropy}
-
-%We consider the following question: For a fixed budget of predictability, $I[X_t|X_{<t}]$, we want to minimize excess entropy -- that is, we want to
-
-How do we compare excess entropies for models with different predictabilities?
-We can take the quotient of excess entropy and predictability:
-$$\frac{\sum_{t=1}^\infty t \cdot I[X_t,X_0 | X_{1...t-1}]}{\sum_{t=1}^\infty I[X_t,X_0|X_{1...t-1}]}$$
-Why is this a good measure of locality?
-
-\begin{enumerate}
-\item Minimizing this quantity (subject to constraints) is equivalent to minizing excess entropy subject to fixed predictability (and these constraints).
-\item This quantity stays invariant when we multiply each $I[X_t,X_0 | X_{1...t-1}]$ with a fixed factor.
-This makes it a natural measure of the decay of $I[X_t,X_0 | X_{1...t-1}]$ as $t \rightarrow \infty$.
-\item This quantity is $\geq 1$, and it takes on the value $1$ if and only if the process is modeled perfectly by a bigram model (is a first-order Markov process).
-
-For fixed entropy rate, the excess entropy is minimized by exactly the Markov processes of this entropy rate.
-\end{enumerate}
-
-\bibliographystyle{acl}
-%\bibliography{extracted}
-\bibliography{literature}
-
-\end{document}
-
-Research has shown that dependency lengths are minimized across a wide range of languages, and it has been suggested that this may be a universal.
-
-This raises the question of what causes this phenomenon.
-
-What explains this?
-
-Hawkins: TODO
-
-Gildea and Jaeger: Gildea & Jaeger (2015) also perform optimization experiments, searching over possible grammars
-optimized for dependency length, average log probability, or some weighted combination of the two.
-The authors find that optimizing one criterion does not generally lead to good performance on the
-other,  but that the observed languages are generally closer to some optimal frontier that trades off
-dependency length against average log probability.  (from Temperley 2018)
-
-Futrell and Levy: noisy context surprisal explains
-
-Here, we use state-of-the art language modeling techniques to address this question:
-On 50 diverse languages, we ... 
-Resulting languages show shorter dependency length.
-
-\subsection{Ordering Model}
-
-As in work studying dependency length minimization in language (e.g. Futrell et al. 2015, Gildea and Jaeger 2015), we specify a parametric dependency tree ordering model.
-%While the precise parameterization of the models is not necessarily crucial, it is crucial that the models are probabilistic, as this allows us to use gradient-based optimization.
-%As we want to compare models optimized for language modeling with models 
-We design our model to be particularly simple to make its parameters interpretable and comparable across models.
-
-For each triple $\tau = $(Head POS, Dependency Label, Dependent POS), we have (1) a Direction Parameter $a_\tau$, and (2) a Distance Parameter $b_\tau$.
-%An Ordering Model is an assignment of values $a_\tau, b_\tau \in \R$ to each of these triples.
-Given a dependency tree, define a distribution over linearizations as follows:
-
-\begin{enumerate}
-\item Choose which dependents follow or precede their heads: Given a head $h$ and dependents $\{d_1, ..., d_k\}$, independently sample $O_i \sim Bernoulli(\sigma(a_{\tau_i}))$, where $i = 1, ..., k$, $\sigma$ is the inverse-logit function, and $\tau_i = (\text{POS of }h, \text{Label of }h \rightarrow d_i, \text{POS of }d_i)$.
-Order $d_i$ to the left of $h$ if $O_i = 0$, otherwise to the right of $h$.
-
-\item %For any set of siblings $d_1, ..., d_k$ that are on the same side of their common head, the probability that $d_i$ is closest to the head, among these siblings, is $$\frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$$
-
-Siblings $d_1, ..., d_k$ that were assigned to the same side of their head are ordered from the head outwards by iterating the following procedure:
-
-Sample $i \in \{1, ..., k\}$ from the distribution $p_i = \frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$.
-Then put $d_i$ closest to the head, and remove $d_i$ from the set, so that $k-1$ siblings remain.
-Repeat this procedure until only one sibling remains, which is then ordered farthest away from the head.
-\end{enumerate}
-
-\mhahn{TODO: Maybe I should create a figure to illustrate this?}
-
-As (Futrell 2015, Gildea and Jaeger 2015),  we only consider projective linearizations.
-
-one similar model is the one from Gildea \& Temperley (2007), where each dependent has a weight in [-1,1] which defines its position wrt the head and the other dependents.
-
-Yours is more general in that it’s probabilistic; theirs is deterministic.
-
-I also have a short EMNLP paper Futrell & Gibson (2015) which derives ordering models from generative dependency models such as Eisner Model C and the Dependency Model with Valence: the Eisner Model C linearization model seems like it would end up being the same as yours with $a_tau$ but not $b_tau$
-
-Kristina Gulordava made an improved version of those models but I can’t find her paper right now; I’m pretty sure it was at CMCL
-
-Another previous work is the reordering model from Galactic Dependencies: Wang & Eisner (2016)
-
-There’s also a small but focused literature on more general linearization models, with a shared task in Belz et al. (2011)
-
-I think the state of the art linearization model in terms of BLEU score is ZGen: https://github.com/SUTDNLP/ZGen but these highly optimized NLP models probably won’t work for a scientific study like you’re doing because of their complexity.
-
-%One issue with most of these models (glidea & temperley, my model, galactic dependencies) is they only model order constraints in “local subtrees” consisting of heads and their immediate dependents.
-That may not always work (but it’s hard to think of a counterexample other than heavy NP shift!).
-Also, nearly all models that I’m aware of, including yours, only generate projective linearizations.
-
-The only exceptions I know of are some things by Bernd Bohnet, who’s now at Google: http://www.aclweb.org/anthology/D12-1085.
-And the way his system does it strikes me as hacky.
-%Another issue with some models (glidea & temperley, my model based on Eisner Model C, your model) is they don’t model interactions among dependents: it may be that if some dependent A is present that changes the order preference for some other dependent B, but this isn’t captured.
-%
-%There’s not a lot of work using linearization models to test scientific hypotheses. The only ones I can really think of are Gildea & Temperley and my stuff and other people who have used that method.
-
-A big problem with that work is that there’s no way to quantify goodness of fit: I can say that real dependency length is minimized beyond some baseline but what I’d like to do is make a model that incorporates dependency length minimization and show that it fits the real orders better than a model without dependency length minimization.
-It seems like your method provides a way to do that.
-%
-%On minimizing dependency length, I don’t know if you know about the algorithmic work on this problem.
-There’s polynomial time solutions to the minimal-dependency-length linearization problem subject to various constraints.
-Chung (1984) has the solution when there are no constraints, Gildea & Temperley (2007) have it under the constraint of projectivity, and Park & Levy (2009) do it under the constraint of mild nonprojectivity (gap degree 1).
-These wouldn’t fit into your gradient descent framework though.
-%
-%Future work may extend this to non-projective ones.
-
-\subsection{Optimization Framework}
-
-We are interested in finding ordering models $\thetad$ that minimize the cross-entropy loss:
-$$-\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P(w_{i+1}|w_i)$$
-where linearized sentences ${\bf w}$ are obained by ordering trees from the treebank according to the ordering model $\thetad$, and $P$ is a language model appropriate to this distribution of linearized sentences.
-%the distribution defined by $\thetad$.
-%In the limiting case where there are no memory limitations, and $P$ is exactly the distribution defined by $\thetad$, this is the entropy of the language, and does not actually depend on the ordering model.
-Note that the original word orders from the treebank do not enter this objective -- they in particular do not influence the language model $P$.
-
-As we simultaneously look for an ordering model optimized for surprisal of the language model, and a language model that optimizes surprisal given the ordering model, this looks like a chicken-egg problem.
-Fortunately, we can reduce this to an unconstrained optimization problem:
-The appropriate language model for a given language is the one that minimizes expected surprisal.
-Therefore, our minimization problem is equivalent to jointly minimizing over ordering models $\thetad$ and language models $\thetal$:
-\begin{equation}\label{eq:surp-final-obj}
-\min_{\thetad, \thetal} -\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})
-\end{equation}
-
-We solve~(\ref{eq:surp-final-obj}) by performing stochastic gradient descent over both sets of parameters $\thetad, \thetal$ simultaneously:
-In each step, we sample a dependency tree from the treebank, then sample an ordering from the current setting of $\thetad$ to obtain a linearized sentence ${\bf w}$.
-Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
-do a gradient descent step using the estimator
-$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
-for the gradient ${ \partial_\thetad R(\thetad, \thetal) \choose \partial_\thetal R(\thetad, \thetal)}$.
-This unbiased estimator is the ordinary gradient estimator for $\thetal$, and the REINFORCE estimator for $\thetad$~\cite{williams-simple-1992}.
-
-\paragraph{Implemention}
-To reduce the variance of the gradient estimator, we apply reward attribution (CITE) and use a word-dependent baseline.
-We add an entropy reward (CITE) for the direction parameters $a_\tau$, weighted with $\alpha$.
-In preliminary experiments, we found $\alpha=0.001$ and a learning rate of $0.01$ with momentum $0.9$ for $\thetad$ to give stable results.
-We apply early stopping using a Monte Carlo estimate of the loss  (\ref{eq:surp-final-obj}) on the development set.
-The hyperparameters for the LSTMs were as in the first experiment; however, we explicitly added input POS embeddings with 20 dimensions to encourage generalization in languages with small training corpora.
-Embeddings were randomly initialized, as before.
-
-\paragraph{Setup}
-We carried this out on all Universal Dependencies\footnote{\url{http://universaldependencies.org/}} 2.1 treebanks for which a training set was available, a total of 50 languages.
-For each language where multiple treebanks with training sets were available, we pooled their training sets; similarly for development sets.
-
-\section{Results}
-
-%\subsection{Correlation with Maximum-Likelihood Fit}
-\subsection{Comparison with Ground-Truth Orderings}
-
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{figures/freedom-lstm.pdf}
-\caption{Optimizing word order for LSTM language modeling explains more of the language's actual rules in languages that have constrained word order.}\label{fig:lstm-languages}
-\end{figure}
-
-We first ask to what extent the ordering rules of the actual languages, as manifested in the corpora, can be recovered by optimizing for predictability.
-
-Since we are interested in locality effects, we compare the distance parameters inferred by optimizing for language modeling with distance parameters inferred from the actual orderings via maximum-likelihood with an $\ell_2$ penalty.
-
-We correlate the weights both at the level of dependency triples, and averaged at the levels of dependency labels, head POS, and dependent POS.
-The correlations are significant at $p < 0.01$ for X languages.
-In Figure~\ref{fig:lstm-languages}, we plot the per-language correlations on the level of triples against word order freedom.
-The optimized languages explain more of the ground-truth ordering preferences in languages with low word order freedom ($r =$TODO, $p <$ TODO), in accordance with our previous result that LSTMs showed a stronger bias towards languages with fixed word order.
-
-\subsection{Information Locality}
-
-We predict that orderings optimized for prediction show information locality. %ease prediction with a limited memory. %show steeper decay of mutual information than random orderings.
-We quantify Information Locality by computing the memory needs by the formula~(\ref{eq:memory-bound}).
-As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we compute MI on the level of POS tags, following~\cite{futrell-noisy-context-2017}.
-As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly.
-
-We computed these quantitities for the optimized languages, the original languages, and for baseline  languages.
-For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
-As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
-This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
-
-Results for four languages are shown in Figure~\ref{fig:memory-results}.
-We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
-In comparison, again averaged across languages, natural language requires less memory than 87 \% of baseline models.
-%For half of the languages
-For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
-For 37 \% of the languages, it also requires less memory than the original language.
-
-We found that, as predicted, memory usage in real languages was correlated with word order freedom:
-Languages which took less memory than all associated baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
-
-\begin{figure*}
-%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Arabic.pdf}
-%%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Japanese.pdf}
-%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-English.pdf}
-%
-%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Korean.pdf}
-%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Basque.pdf}
-\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
-%
-\caption{Memory demands of real, optimized, and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction.
-On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
-\end{figure*}
-
-\subsection{Dependency Length Minimization}
-
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{figures/depLength-facet.pdf}
-\caption{Optimizing for LSTM language modeling (blue) shortens dependency lengths compared to baseline (green), though not always as strongly as natural language (red) does.}\label{fig:dep-lengths}
-% Optimizing for LSTM modeling (green) tends to shorten more strongly than optimizing for bigram surprisal (red). \mhahn{choose languages}}
-\end{figure}
-
-A prominent manifestation of locality in language is dependency length minimization:
-Languages have shorter dependencies than would be expected at random~\cite{futrell-large-scale-2015}. %, and under reasonable baselines.
-
-%\cite{futrell-noisy-context-2017} 
-As previous work (CITE) has argued that dependency length minimization can be derived from human memory limitations, we expect that languages optimized for language modeling should show short dependencies.
-We computed per-sentence dependency lengths for the optimized ordering rules, creating a counterfactual corpus by sampling a linearization for each tree.
-%As before, we sample linearizations at zero tempterature to eliminate nondeterminism that may lengthen average dependency length.
-%We sampled trees with temperature $\tau = 1/100$ to eliminate nondeterminism (which may lengthen average dependency length).
-
-%As a baseline, we also computed dependency lengths for ordering models that were trained with dependency length as the objective.
-%(This differs from (CITE), who computed the per-sentence absolute minimum. Our baseline is more realistic in that it respects grammatical consistency across sentences).
-As in the previous experiment and similar to \cite{futrell-large-scale-2015}, we considered ordering models with randomized $a_\tau, b_\tau$ as baselines.
-
-In every one of the 50 languages, the optimized models show shorter dependencies than the baseline languages.
-Results from four languages are shown in Figure~\ref{fig:dep-lengths}.
-In 20 \% of the languages, they exceed the actual ground-truth orderings in dependency length minimization.
-
-\subsection{Qualitative Predictions Across Languages}
-
-We have seen that, for languages with fixed word order, optimizing for language models explains a significant amount of word order preferences.
-We now examine what kinds of rules LSTMs prefer across languages. %take the grand average across languages.
-
-We averaged distance weights across all languages, both for the optimized models, and for the maximum-likelihood models.
-Within each language, we averaged across dependents, weighted by frequency of the dependency-dependent combination in the language.
-
-The correlations between these averages across 33 dependency types was 0.50 \mhahn{TODO: I'll bootstrap over languages}, i.e., half of the variance in cross-linguistic ordering preferences as captured by $b_\tau$ could be explained as being optimized for language modeling.
-We then considered specifically the averages separately for dependencies headed by nouns and verbs, respectively.
-We considered the five dependencies that bind most strongly and least strongly, that is, where $b_\tau$ was smallest or greatest.
-
-%For nouns, we found that dependents typically considered to be part of the noun phrase (determiners, numerals, adjectives, modifiers, compound parts) bind strongly to the noun both in optimized and real languages.
-%Copulas and adpositions, which in UD are considered dependents of the noun, bind very strongly in optimized models, while they are generally not considered part of the noun phrase in the linguistic literature -- nonetheless they tend to bind relatively close to the noun even in natural language.
-%Intriguingly, the functions of adpositions and determiners, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.
-%In both optimized and real languages, among the dependents that tend to be less close to the noun are relative clauses, arguments of copulas, adverbs, and coordination -- these are elements that are not part of the noun phrase as typically understood in the linguistic literature, and they also tend to be longer.
-
-The closest dependencies were determiners, classifiers, case markers/adpositions, and numerals.
-The farthest dependencies were relative clauses, conjuncts, adjectives, and adverbial clauses.
-Intriguingly, the functions of determiners, classifiers, and adpositions, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.
-
-Turning now to dependencies headed by verbs, expletives, auxiliaries, markers, and adjectives bind most strongly.
-% bind very strongly to the verb both in optimized and in real languages, even they are often not understood as dependents of the verb in the linguistic literature.
-Again, intriguingly, typical functions of auxiliaries and markers are expressed morphologically on the verb in many languages.
-%, suggesting again that grammaticalization into morphological markers may be favored by locality.
-Perhaps counter to expectation, arguments (subjects, objects) are not generally closer than adjuncts, neither in optimized nor in real languages.
-Similar to noun phrases, adjunct clauses stand out farther in optimized languages.
-The farthest dependents are two kinds of nominal modifiers, dislocated elements, and objects.
-
-\section{Related Work and Discussion}
-
-\citet{gildea-human-2015} showed that a sample of five natural languages optimizes trigram surprisal and dependency length compared to random baselines, and also performed optimization with these objectives.
-Our work links these findings to general properties of sequence models including LSTMs, and studies a much broader sample of languages.
-% and studies a much broader sample of languages.
-%, and (3) investigating qualitative linguistic predictions.
-
-Our work relates to recent work showing that neural language models tend to focus strongly on recent words for prediction~\cite{daniluk-frustratingly-2017, khandelwal-sharp-2018}.
-Our results suggests that this is explained both by general properties of sequence models with bounded memory, and by natural language favoring incremental prediction with such models.
-
-Recent work has disucssed the adequacy of sequence models for language with its hierarchical structure (CITE).
-Our work adds to this by showing that some of the architectural biases in linear sequence modeling turn out to favor natural language, arguably because they are also present in human language processing.
-%natural language.
-
-%Our findings highlight that there is a psycholinguistic explanation for the reasonable success of local methods, such as n-gram models, in NLP.
-%We have shown that word order regularities in languages can partly be recovered from unordered data by optimizing for predictability.
-
-\section{Conclusion}
-
-We first showed that LSTM language models are biased towards orderings of information as found in natural language, and that this bias is stronger for languages with fixed word orders.
-This is grounded in general properties of sequence modeling, which favors sequences exhibiting locality as found in natural language.
-Asking which orderings would be optimal for LSTM language modeling, we then used reinforcement learning to create counterfactual versions of 50 diverse languages that are optimized for ease of language modeling.
-We showed that, across corpora, LSTMs favor word orders that exhibit information locality and have short syntactic dependencies, similar to natural languages.
-In languages with fixed word order, a significant amount of the actual word order preferences can be recovered by optimizing for language modeling, suggesting that word order rules in these languages can partly be explained as making prediction easier.
-
-%- If we could adapt the language's word order rules to make LSTM language modeling easiest, what would the resulting language look like?
-
-%- previous work has shown language has locality
-
-%- we showed that LSTM language models are biased towards orders typically chosen in natural language, and that this preference increases as memory is restricted, and increases if memory is unbounded
-
-%- this bias was shown to be caused particularly by function words -- words primarily carrying grammatical information.
-
-%- we then used reinforcement learning to directly investigate what kinds of word orders are preferred by incremental language models. We chose LSTMs and bigram models as two extremes.
-
-%- Across 50 languages, we showed that LSTMs prefer word orders that require little memory to model and have short syntactic dependencies, similar to natural language.
-
-%- Using an information theoretic measure of memory need, we showed that word orders chosen in many natural language require less memory than would be expected at random
-
-%- Our quantification of memory demands is entirely information-theoretic and makes no architectural assumptions beyond incrementality of sequence modeling.
-%
-%- We also showed that inferred ordering preferences explain a significant amount of ordering preferences in natural language, both for specific languages and across languages.
-%
-%- Our work has implications both for NLP and for syntactic typology. 
-%
-%
-%Bounded context representations
-%
-%
-%We observe that incrementally predicting language under memory limitations -- as instantiated by both recurrent language models and, arguably, human language processing, favors word orders that exhibit Information Locality:
-%
-%We showed that recurrent language models indeed show a bias towards information locality.
-%We then showed that information locality makes nontrivial and partly accurate typological predictions for word order across languages.
-
-\section{More on Memory}
-
-- tradeoff between memory and cross-entropy: let $M$ be a model with memory $Mem(M)$. For all $t$,
-
-$Mem(M) + t CE[M,X] \geq Mem(X) + t H_{rate}[X]$
-
-since there is a compression scheme for the sequences in $X$ of average length $t$ with size $Mem(M) + t CE[M,X]$, but the optimal scheme has length $Mem(X) + t H_{rate}[X]$
-
-In particular, if $M$ is a lossy context encoding scheme and word-by-word prediction is optimal w.r.t. this, the word-by-word KL loss is lower bounded as
-$$(CE[M,X] - H_{rate}[X]) \geq (Mem(X) - Mem(M))$$
-Of course, $Mem(M)$ might be large, in which case this bound will be vacuous.
-We could define the actual information content in $Mem(X)$ that is relevant to $X$ using cross-entropies -- this might then be larger or smaller than the actual information content, depending on the efficiency of the encoding (e.g., it could ignore certain kinds of context observations and thus require unnecessarily much memory).
-
-\bibliographystyle{acl}
-%\bibliography{extracted}
-\bibliography{literature}
-
-\end{document}
-
-\begin{align*}
-& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
-& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
-& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
-& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
-& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
-& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
-& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
-& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
-\end{align*}
-
-Is this useful?
-If we know $I_t$, we can choose $T$ so as to make this bound as tight as possible.
-To show the usefulness of this method, we consider the case where $I_t = \frac{\alpha}{t^\beta}$, for $\alpha > 0$, $\beta > 2$ (for smaller $\beta$, $M$ evaluates to $+\infty$).
-Then $M = \alpha \sum_{t=1}^\infty \frac{1}{t^{\beta-1}}$, which is bounded as $\frac{\alpha}{\beta-1} = \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt \leq M \leq \alpha + \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt = \alpha + \frac{\alpha}{\beta-1} = \frac{\alpha\beta}{\beta-1}$.
-The bound from the proposition evaluates to
-\begin{align*}
-\sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T &= \sum_{t \leq T} \frac{\alpha}{T t^{\beta-1}} + \sum_{t > T} \frac{\alpha}{t^\beta} - J/T \\
-& \geq \int_1^T \frac{\alpha}{T t^{\beta-1}} dt + \int_T^\infty \frac{\alpha}{t^\beta} dt - J/T \\
-& = \frac{\alpha (\beta-1)}{T} - \frac{\alpha(\beta-1)}{T^{\beta-1}} + \frac{\alpha\beta}{T^{\beta-1}} - J/T \\
-&= \frac{1}{T} (\alpha(\beta-1) - J) + \frac{\alpha}{T^{\beta-1}}  \\
-&\geq \frac{1}{T} (M(\beta-1) - \alpha - J) + \frac{\alpha}{T^{\beta-1}}  \\
-%&\geq \frac{1}{T} (M - J) + \frac{M}{(\beta-1)T^{\beta-1}} \\
-%& \geq M-J \ \ \ \ \ \ \  \text{(by setting $T=1$)} 
-\end{align*}
-%whose derivative by $T$ is
-%$$\frac{-1}{T^2} (\alpha(\beta-1) - J) - \frac{\alpha(\beta-1)}{T^{\beta}}  $$
-%which is zero at
-%$$T = \left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)} $$
-%Substituting back, we get the bound
-%$$\frac{1}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)}} (\alpha(\beta-1) - J) + \frac{\alpha}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{(\beta-1)/(\beta-2)}}$$
-
-%
-%\begin{proposition}
-%Let $\epsilon > 0$, $T \in \mathbb{N}$.
-%If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$.
-TODO
-%\end{proposition}
-%\begin{proof}
-%Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
-%Without loss of generality, we can assume that the model is Bayes optimal: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
-%
-%By definition, the difference between the model's surprisal and optimal surprisal is
-%\begin{align*}
-%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
-%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
-%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
-%& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
-%& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
-%& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
-%& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
-%& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
-%%& \geq \frac{1}{T} \sum_{t \leq T}  I_t + \sum_{t > T} I_t - J/T \\
-%%& \geq \frac{1}{T} (TODO - I[X_{1\dots T}|f(X_{<0})]) \\
-%%& \geq \frac{1}{T} (TODO M - J - \epsilon) \\
-%\end{align*}
-%\end{proof}
-
-%If we consider ratios between the memory bound and the predictability $I[X_t, X_{<t}]$, we can also express this as follows:
-%\begin{corollary}
-%Let $\epsilon > 0$, and let $P := I[X_t, X_{<t}]$ the predictability.
-%Then If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$.
-%\end{corollary}
-%
-%\begin{proof}
-%$\frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{P} = \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > 0} I[X_t, X_0 | X_{1\dots t-1}]}  \leq \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > M/\epsilon} I[X_t, X_0 | X_{1\dots t-1}]}< \epsilon$
-%
-%Let $T := M/\epsilon$.
-%Then 
-%\end{proof}
-%
-%
-
-%\subsection{Incorporating All Lemmas}
-
-%\subsection{Prediction on Raw Word Forms}
-
-%\section{information locality and dependency length minimization}
-
-%\subsection{Languages can trade off Memory Load and Predictability; Information Locality}
-%
-%%The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$
-%
-%We will describe a geometric view on $M$ that gives us a  quantitative measure of Information Locality.
-%
-%By information locality, we refer to the degree to which words that are predictive of each other are closer together.
-%Stated differently, information in past observations that helps predict $X_t$ should be concentrated in recent observations $X_{t'}$ ($t' < t$) with $t'$ close to $t$.
-%
-%Information locality is maximised if all contextual information predicting $X_t$ is contained in $X_{t-1}$, in which case $J = I[X_t, X_{t-1}]$.
-%
-%To quantify information locality more generally, we consider the relation between the quantity
-%$$J = I[X_{\geq t}, X_{<t}] = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
-%and the quantity
-%$$P = I[X_t, X_{< t}] = \sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
-%The first one quantifies how much information is transmitted between past and future, the second one captures the average predictability of single words from their prior context.
-%Evidently, $P \leq J$ -- higher degrees of predictability will require more information to be stored from past observations.
-%Stated differently, there is a tradeoff between predictability and memory: If the future is easier to predict from the past (at the same unconditional entropy), more memory is required to store all the relevant predictive information from the past.
-%
-%Minimizing $J$ unconditionally corresponds to an entirely unpredictable process.
-%In contrast, if we minimize $J$ while holding $P$ constant, we obtain a formalization of the requirement of optimal information locality: This requires placing the predictive information closer to the present observation, keeping the total amount of predictive information constant..
-%
-%If there are no further constraints on the minimization, the minimum is attained when $P = J$.
-%This is the case when all contextual information predicting $X_t$ is contained in $X_{t-1}$ -- this minimizes memory load, as only information from the last word needs to be stored.
-%If $P$ is fixed, and $J$ increases, information locality decreases -- corresponding to increasing memory load.
-%
-%To understand this more generally, let us consider $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ as a function of $t$.
-%This function describes the amount of bits predictive of $X_t$ contained in $X_0$, the observation $t$ steps in the fast.
-%The faster this function decays, the lower $J$ is.
-%Information locality can be interpreted as requiring this function to decay fast -- $J$ is absolutely minimized if the function jumps to zero for $t > 1$.
-%
-%If this is a principle describing natural language, we expect that there will be further constraints, resulting in values of $J$ larger than $P$.
-%We will study this in detail in the next section.
-%
-%
-%We can observe that the area under the curve is equal to $P$ -- this quantifies the total amount of contextual information predicting the current observation.
-%This suggests to consider the ratio
-%$$L := \frac{J}{P}$$
-%First, note that minimizing $J$ with $P$ fixed is equivalent to minimizing $L$.
-%Second, expanding this quantity, we see
-%$$L = \sum_{t=1}^\infty t \frac{\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}{\sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}$$
-%That is, $L$ describes the $J$-coordinate of the center of mass of the area under the curve in (REF).
-%A different way of understanding it is as the average distance of past information predictive of the current observation.
-%With this interpretation of $L$, we can understand it as a direct fornalization of the notion of information locality.
-%
-%Let  us consider a second visualization by plotting $J$ against $P$.
-%The area that is achievable by processes is the area below the diagonal.
-%Information locality is maximized by processes on the diagonal; the distance from the diagonal quantifies the ratio $J/P$ and thus information locality.
-%
-%

### 2019-02-17 figures/getFigures.sh

@@ -1,0 +1,1 @@
+cp ~/CS_SCR/*mem*pdf .

### 2019-02-28 paper.tex
lstm

@@
 
+\paragraph{Dependency Data}
+We draw on dependency corpora.
+
+Universal Dependencies has compiled dependency corpora for several dozen languages
+
 \paragraph{Ordering Grammrs}
 We define ordering grammars, small interpretable models of dependency tree linearization.
+Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
 
-Every dependency type $\tau$ is assigned a parameter $a_\tau \in [0,1]$.
-Dependency trees are linearized by the following algorithm:
+Universal dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
+These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
+We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
+Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.
 
-%Every dependency type $\tau$ is assigned a direction parameter $\alpha_\tau$ and a distance parameter $\beta_\tau$.
+This parameter defines how dependents are ordered relative to their head:
+Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
+Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
+
+Ordering grammars describe languages that have consistent word order:
+For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
 
@@
 
+Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
+
 \subsection{Method}
-The small size of available syntactic data for many languages means that estimating mutual information is problematic.
-
-In general, we expect a bias-variance tradeoff:
-
-A method that estimates on a coarse-grained level (e.g.
+%The small size of available syntactic data for many languages means that estimating mutual information is problematic.
+%
+%
+%In general, we expect a bias-variance tradeoff:
+%
+%A method that estimates on a coarse-grained level (e.g.
 POS) will result in estimates with relatively higher fidelity and lower variance, potentially making it easier to localize the actual language in the space of counterfactual languages.
-However, it introduces \emph{bias} by describing a simplified version of the language.
-
-We propose to therefore compute the quantities on levels less likely to suffer from data sparseness: part-of-speech and morphology.
-
-In the UD treebanks, each word has part-of-speech and morphological information.
-The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
-We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token that indicates that information related to the word ends.
-We further provide the word's lemma after its POS tag, if it is one of the 50 most frequent lemmas of the language.
-Each POS tag and morphological feature-value pair is encoded as an atomic token.
-The rationale for including frequent lemmas is that such words often are strongly grammaticalized and may provide grammatical information (and frequent words are less likely to be affected by data sparseness).
+%However, it introduces \emph{bias} by describing a simplified version of the language.
+%
+%
+%
+%
+%
+%We propose to therefore compute the quantities on levels less likely to suffer from data sparseness: part-of-speech and morphology.
+%
+%In the UD treebanks, each word has part-of-speech and morphological information.
+%The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
+%We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token that indicates that information related to the word ends.
+%We further provide the word's lemma after its POS tag, if it is one of the 50 most frequent lemmas of the language.
+%Each POS tag and morphological feature-value pair is encoded as an atomic token.
+%The rationale for including frequent lemmas is that such words often are strongly grammaticalized and may provide grammatical information (and frequent words are less likely to be affected by data sparseness).
+%
 
 To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling sequences such as natural language.
-We will discuss alternative estimation methods below.
+We model language on the level of individual word forms.
+We provide data from alternative estimation methods in the SI.
 
-During incremental prediction, the LSTM sequentially reads these tokens and computes a distribution over the next word.
-The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).
+\paragraph{Data Split}
+
+UD models predefined with train, dev, and test splits.
+Small datasets sometimes not.
+
+\paragraph{Model}
+We use a recurrent neural network with Long-Short-Term Memory cells.
+%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
+%
+The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
+This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
+The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
+The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
+
+%Such networks estimate the probability of a word in context as follows.
+Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
+It then computes a sequence of hidden and cell states by the following recurrent computation:
+\begin{align*}
+	h_1, c_1 &:= 0 \\
+	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
+	\dots \\
+	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
+\end{align*}
+The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
+We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
+
+The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
+We obtain per-word predictions of the next word by computing
+\begin{align*}
+	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
+	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
+\end{align*}
+where the softmax transformation normalizes vectors into probability distributions as follows
+\begin{equation}
+	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
+\end{equation}
+Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
+\begin{equation}
+	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+\end{equation}
+and thus the surprisal is estimated as
+\begin{equation}
+- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+\end{equation}
+We discuss the choice of the numerical parameters in the next section.
+
+\paragraph{Training}
+Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
+We can think of the LSTM parameters as forming one large vector $\theta$.
+At the beginning of training, the parameters $\theta$ are randomly initialized.
+If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
+w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
+We then update the parameter vector:
+\begin{equation}\label{eq:train}
+	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
+\end{equation}
+where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
+
+Word sequences are sampled without replacement.
+Once all sequences have been processed, we start another pass through the training data.
+
+After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the development partition.
+We terminate training once this development partition surprisal does not improve over the one computed after the previous pass any more.
+
+\paragraph{Regularization}
+%Dropout, Word dropout, Word noising
+%Any statistical estimation problem faces the 
+The quality of neural network models is improved by regularization, improving generalization to the full data distribution.
+In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
+We use three common methods of regularization.
+First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
+
+\paragraph{Choice of Hyperparameters}
+
+The LSTM model has a set of numerical parameters that need to be specified before training, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters $p_{embedding}, p_{word}, p_{noising}$.
+%We do this using the standard method
 
@@
 This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
+
+\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
 

### 2019-03-01 langs-figure.tex

\documentclass[8pt,letterpaper]{article}

\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{amssymb}

\usepackage{tikz-dependency}
\usepackage{longtable}

\newcommand{\R}[0]{\mathbb{R}}
\newcommand{\E}[0]{\mathbb{E}}
\newcommand{\Ff}[0]{\mathcal{F}}

\usepackage{multirow}

\newcommand{\soft}[1]{}
\newcommand{\nopreview}[1]{}
\newcommand\comment[1]{{\color{red}#1}}
\newcommand\mhahn[1]{{\color{red}(#1)}}

\usepackage{amsthm}

\newcommand{\thetad}[0]{{\theta_d}}
\newcommand{\thetal}[0]{{\theta_{LM}}}

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}

\frenchspacing
%\def\baselinestretch{0.975}

%\emnlpfinalcopy
%\def\emnlppaperid{496}

\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
\author{Michael Hahn}
\date{2018}

\begin{document}

\maketitle

\begin{tabular}{cccccccccccccc}
Afrikaans & Amharic & Arabic & Armenian & Basque & Breton & Bulgarian & Buryat \\
\includegraphics[width=0.12\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Arabic-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Basque-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}
\\
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & Estonian \\
\includegraphics[width=0.12\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Catalan-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Chinese-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Croatian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Czech-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Danish-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Dutch-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}
\\
Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian \\
\includegraphics[width=0.12\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Finnish-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/French-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/German-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Greek-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Hindi-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}
\\
Indonesian & Japanese & Kazakh & Kurmanji & Latvian & Naija & North Sami & Norwegian \\
\includegraphics[width=0.12\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Japanese-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Latvian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}
\\
Persian & Polish & Romanian & Serbian & Slovak & Slovenian & Spanish & Swedish \\
\includegraphics[width=0.12\textwidth]{figures/Persian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Polish-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Romanian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Serbian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Slovak-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Spanish-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}
\\
Tamil & Thai & Turkish & Ukrainian & Uyghur & Vietnamese \\
\includegraphics[width=0.12\textwidth]{figures/Tamil-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Turkish-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.12\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}
\\\end{tabular}

\end{document}

### 2019-03-01 paper.tex

@@
 
-\paragraph{Data Split}
+\paragraph{Data}
+
+We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
+We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
+We excluded four languages for a later confirmatory experiment (English, Korean, Russian). % Irish
+This resulted in X languages.
+
+For most language, we used the predefined data split.
+For some languages, where there was no training partition, or where the training partition was smaller than the test partition, we redefined the split to obtain more training data:
+For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
 
@@
 %Any statistical estimation problem faces the 
-The quality of neural network models is improved by regularization, improving generalization to the full data distribution.
-In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-We use three common methods of regularization.
-First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
+The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
+%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
+We use three standard methods of regularization that have been shown to improve neural language modeling: dropout (CITE), word dropout (CITE), and word noising (CITE).
+%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
+%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
 
@@
 
-The LSTM model has a set of numerical parameters that need to be specified before training, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters $p_{embedding}, p_{word}, p_{noising}$.
-%We do this using the standard method
-
-Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
+The LSTM model has a set of numerical parameters that need to be specified before training, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$.
+We choose these parameters so as to minimize the average surprisal on the held-out partition resulting from a fully trained model.
+
+For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
+This biases the hyperparameters towards modeling counterfactual grammars better, biasing them against our hypothesis.
+
+%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
+%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
+%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
 
@@
 
-Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
-Therefore, to compute this quantity, we compute the differences between the surprisals
-$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.
+The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
+\begin{equation}
+H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]
+\end{equation}
+For each word in the held-out partition, we compute the difference
+\begin{equation}
+	-\log P_\theta[X_t | X_0, X_1, ..., X_{t-1}] - P_\theta[X_t | X_1, ..., X_{t-1}]
+\end{equation}
+and take the average over these.
 We cut $T$ off at 30.
 
-\paragraph{Alternative Models}
+\subsection{Discussion: Alternative Models}
 Recent research has proven strong theoretical learnability guarantees for neural network models (CITE), supporting their success in practice.
-They also provide the strongest predictors of the surprisal effect on human reading times \citep{frank-insensitivity-2011, goodkind-predictive-2018}.
+They also provide strong predictors of the surprisal effect on human reading times \citep{frank-insensitivity-2011, goodkind-predictive-2018}.
 
@@
 
-First, an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations.
+A traditional model uses n-gram models.
+A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
 Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
 Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-
-A statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar.
+We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
+
+A second option is to construct a statistical grammar, such as PCFG.
 The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
@@
 
-\subsection{Data}
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-We excluded four languages for a later confirmatory experiment (English, Korean, Russian). % Irish
-This resulted in X languages.
-
-For most language, we used the predefined data split.
-For some languages, where there was no training partition, or where the training partition was smaller than the test partition, we redefined the split to obtain more training data:
-For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
+%\subsection{Data}
+%\subsection{Setup}
+%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
+%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
+%This biases the hyperparameters towards favoring counterfactual grammars.
 
 \subsection{Setup}
-The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-This biases the hyperparameters towards favoring counterfactual grammars.
 
@@
 For each model for each language, we interpolate the surprisal-memory tradeoff curves given by Proposition~\ref{prop:suboptimal} linearly.
-For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.
-
-We test by considering the null hypothesis that $\E_{x \sim P_1}[W(x)] \leq 1/2$ where $P_1$ is the distribution over values obtained for real orderings.
-
-We use a bootstrapped confidence interval for $\E[W]$ for testing and for quantifying the degree of optimization.
+We estimate the unigram entropy $H[X_0]$ by averaging over all models.
+We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
+For each sample $x$ from real orderings, we look at the proportions $W_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $W_-(x)$ of baseline samples that are consistently less optimal.
+
+We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
+\begin{equation}
+	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
+\end{equation}
+where $P_1$ is the distribution over values obtained for real orderings.
+
+We estimate the quotient
+\begin{equation}
+	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
+\end{equation}
+We use a bootstrapped confidence interval for $\E[G]$ for testing and for quantifying the degree of optimization.
 
@@
 
-For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $W$ had width $\leq 0.15$.
-We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
+For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
+We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
+Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
+%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
 
@@
 
-In all but X languages $\E[W] > 0.5$ significantly.
-
-\section{Experiments on Further Treebanks}
+In all but X languages $\E[G] > 0.5$ significantly.
+
+\subsection{Discussion}
+
+Word Order Freedom and Memory Efficiency
 

### 2019-03-01 results-table-word-level.tex
results

Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Arabic  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Armenian-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.91  &  [0.83, 0.98]  \\ [10.25ex] \hline
Bambara-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.96  &  [0.86, 1.0]  \\ [10.25ex] \hline
Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
English  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Erzya-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.99  &  [0.98, 1.0]  \\ [10.25ex] \hline
Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.83  &  [0.76, 0.9]  \\ [10.25ex] \hline
Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
French  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.89  &  [0.81, 0.96]  \\ [10.25ex] \hline
Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kurmanji-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.92  &  [0.83, 0.99]  \\ [10.25ex] \hline
Latvian*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.48  &  [0.4, 0.57]  \\ [10.25ex] \hline
Maltese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.37  &  [0.3, 0.44]  \\ [10.25ex] \hline
Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.07  &  [0.01, 0.15]  \\ [10.25ex] \hline
Portuguese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.07  &  [0.03, 0.12]  \\ [10.25ex] \hline
Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.82  &  [0.74, 0.89]  \\ [10.25ex] \hline
Spanish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Urdu  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Uyghur-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  0.65  &  [0.57, 0.74]  \\ [10.25ex] \hline
Vietnamese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}}  &  $D_x$  \\ 
  &    &    &  $W_x$  &  1.0  &  [0.98, 1.0]  \\ [10.25ex] \hline

### 2019-03-01 code/tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py

@@ -1,4 +1,4 @@
-# tradeoff/effectSize.tsv
+# ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
 
 with open("../ud_languages.txt", "r") as inFile:
    languages = inFile.read().strip().split("\n")
@@ -43,10 +43,6 @@ with open("../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv", "r") 
 languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
 
 
-effectSize = (effectSize[0], sorted(effectSize[1], key=lambda x:x[0]))
-languageKey = dict([(h(effectSize, line, "Language"), line) for line in effectSize[1]])
-
-languageKey_diff = dict([(h(effectSize_diff, line, "Language"), line) for line in effectSize_diff[1]])
 
 with open("../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
    listener_curve = readTSV(inFile)
@@ -54,39 +50,17 @@ with open("../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.ts
 languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
 
 for language in languages:
-   line = languageKey[language]
-   line_diff = languageKey_diff[language]
+#   line = languageKey[language]
    line_listener = languageKey_listener_curve[language]
    line_stats = languageKey_stats[language]
 
-   assert language == h(effectSize, line, "Language")
 
-   satisfied = h(stats, line_stats, "RANDOM_BY_TYPE") >= 20 and h(stats, line_stats, "REAL_REAL") >= 20 and (h(listener_curve, line_listener, "result1High") - h(listener_curve, line_listener, "result1Low") <= 0.15) and (h(effectSize, line, "hi2") - h(effectSize, line, "lo2") <= 0.3) # and (h(effectSize_diff, line_diff, "hi1") - h(effectSize_diff, line_diff, "lo1") <= 0.2)
+   satisfied = h(stats, line_stats, "RANDOM_BY_TYPE") >= 10 and h(stats, line_stats, "REAL_REAL") >= 5 and (h(listener_curve, line_listener, "result1High") - h(listener_curve, line_listener, "result1Low") <= 0.15)
 
    components = [language.replace("_"," ")+("*" if not satisfied else "")]
[…]

### 2019-03-01 code/tradeoffStats_OnlyWordForms_BoundedVocab.py

@@ -21,7 +21,7 @@ def readTSV(x):
           data[i][column] = vals[i]
     return (header, data)
 try:
-  with open("/u/scr/mhahn/"+language+"_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+  with open("../results/raw/word-level/"+language+"_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
      data = readTSV(inFile)
 except IOError:
    print("\t".join(map(str, [language, 0,0  ])))
@@ -31,8 +31,6 @@ except IOError:
 
 
 #data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
-
-import torch
 
 def g(frame, name, i):
     return frame[1][i][frame[0][name]]

### 2019-03-01 code/tradeoffStats_OnlyWordForms_BoundedVocab_FORALL.py

@@ -1,14 +1,15 @@
 import sys
 
-#python ~/scr/CODE/deps/yStudyTradeoffPerLang_FORALL.py RANDOM_BY_TYPE > ~/scr/tradeoff-perLanguage-RANDOM_BY_TYPE.tsv
-#python ~/scr/CODE/deps/yStudyTradeoffPerLang_FORALL.py RANDOM_MODEL > ~/scr/tradeoff-perLanguage-RANDOM_MODEL.tsv
 from ud_languages import languages
 
 
-   
+# ./python27 tradeoffStats_OnlyWordForms_BoundedVocab_FORALL.py REAL_REAL > ../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv
+# ./python27 tradeoffStats_OnlyWordForms_BoundedVocab_FORALL.py GROUND > ../results/tradeoff/stats-onlyWordForms-boundedVocab_GROUND.tsv
+
+real_type = sys.argv[1] 
 
 import subprocess
-print("\t".join(["Language", "REAL_REAL", "RANDOM_BY_TYPE"]))
+print("\t".join(["Language", real_type, "RANDOM_BY_TYPE"]))
 for language in languages:
-    print(subprocess.check_output(["/u/nlp/anaconda/ubuntu_16/envs/py27-mhahn/bin/python2.7", "tradeoffStats_OnlyWordForms_BoundedVocab.py", language]).strip())
+    print(subprocess.check_output(["/u/nlp/anaconda/ubuntu_16/envs/py27-mhahn/bin/python2.7", "tradeoffStats_OnlyWordForms_BoundedVocab.py", language, real_type]).strip())
 

### 2019-03-02 paper.tex
results
paper

@@
 
-\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
+\title{Crosslinguistic Word Orders Enable and Efficient Tradeoff between Memory and Surprisal}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
-Are languages optimized for processing with limited memory?
-Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
+Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
+Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
+We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
+We show that the shape of the tradeoff is determined in part by word order.
+In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
+We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
+
+%Are languages optimized for processing with limited memory?
+%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
+%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
+%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
 %assuming only that listeners perform incremental prediction.
-Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
+%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
 % producing language and listeners predicting input.
@@
 
+\begin{longtable}{lllllllllllllll}
+\input{results-table-word-level.tex}
+\end{longtable}
+
 \subsection{Discussion}
@@
 
-\section{Results (Morphology on UD Treebanks)}\label{sec:results-table}
-
-Statistics reported:
-
-\begin{enumerate}
-%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
-	\item $A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$
-	\item $B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$
-	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$
-	\item Compute $A_{x'}$ for each $x'$ in the baseline distribution, and then calculate the percentile rank of $A_x$ among these values:
-$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$
-\item For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.
-
-\end{enumerate}
-
-For each language, we show the number of sentences in the training partition, the surprisal-speaker memory tradeoff plane, the listener's surprisal-memory tradeoff curves (smoothed means), and the statistics defined above with bootstraped CIs.
-
-Languages where the stopping criterion hasn't been reached are marked with asterisks.
-
-\begin{longtable}{lllllllllllllll}
-Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.28]  \\
-1315  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
-  &    &    &  $E_x$  &  0.42  &  [0.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  0.99  &  [0.96, 1.0]  \\ [10.25ex] \hline
-Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.16  &  [0.06, 0.29]  \\
-974  &    &    &  $B_x-A_x$  &  0.1  &  [0.04, 0.18]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Arabic*  &\\ [10.25ex] \hline
-Armenian-Adap*  &\\ [10.25ex] \hline
-Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.03, 0.15]  \\
-5396  &    &    &  $B_x-A_x$  &  0.82  &  [0.76, 0.88]  \\ 
-  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.03, 0.22]  \\
-788  &    &    &  $B_x-A_x$  &  0.32  &  [0.2, 0.42]  \\ 
-  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.22]  \\
-8907  &    &    &  $B_x-A_x$  &  0.56  &  [0.44, 0.7]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.02, 0.19]  \\
-808  &    &    &  $B_x-A_x$  &  0.28  &  [0.19, 0.42]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.08, 0.31]  \\
-550  &    &    &  $B_x-A_x$  &  0.25  &  [0.17, 0.36]  \\ 
-  &    &    &  $E_x$  &  0.92  &  [0.85, 0.98]  \\ 
-  &    &    &  $W_x$  &  0.94  &  [0.86, 1.0]  \\ [10.25ex] \hline
-Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.03, 0.25]  \\
-13123  &    &    &  $B_x-A_x$  &  0.44  &  [0.33, 0.57]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.04, 0.23]  \\
-3997  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.19]  \\ 
-  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.03, 0.23]  \\
-7689  &    &    &  $B_x-A_x$  &  0.13  &  [0.05, 0.23]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.05, 0.27]  \\
-102993  &    &    &  $B_x-A_x$  &  0.66  &  [0.56, 0.77]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  0.94  &  [0.86, 0.99]  \\ [10.25ex] \hline
-Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.26]  \\
-4383  &    &    &  $B_x-A_x$  &  0.15  &  [0.06, 0.25]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.01, 0.15]  \\
-18310  &    &    &  $B_x-A_x$  &  0.04  &  [0.01, 0.08]  \\ 
-  &    &    &  $E_x$  &  0.88  &  [0.73, 1.0]  \\ 
-  &    &    &  $W_x$  &  0.98  &  [0.96, 1.0]  \\ [10.25ex] \hline
-Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.48]  \\
-6959  &    &    &  $B_x-A_x$  &  0.26  &  [0.18, 0.36]  \\ 
-  &    &    &  $E_x$  &  0.76  &  [0.67, 0.84]  \\ 
-  &    &    &  $W_x$  &  0.72  &  [0.65, 0.79]  \\ [10.25ex] \hline
-Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.26]  \\
-1108  &    &    &  $B_x-A_x$  &  0.34  &  [0.24, 0.45]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.01, 0.12]  \\
-27198  &    &    &  $B_x-A_x$  &  0.95  &  [0.91, 0.99]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-French*  &\\ [10.25ex] \hline
-German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.44  &  [0.37, 0.51]  \\
-13814  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.29]  \\ 
-  &    &    &  $E_x$  &  0.66  &  [0.6, 0.74]  \\ 
-  &    &    &  $W_x$  &  0.63  &  [0.57, 0.69]  \\ [10.25ex] \hline
-Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.2]  \\
-1662  &    &    &  $B_x-A_x$  &  0.59  &  [0.48, 0.73]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\
-5241  &    &    &  $B_x-A_x$  &  0.19  &  [0.08, 0.3]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.24]  \\
-13304  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.55  &  [0.44, 0.65]  \\
-910  &    &    &  $B_x-A_x$  &  0.07  &  [0.0, 0.15]  \\ 
-  &    &    &  $E_x$  &  0.62  &  [0.51, 0.74]  \\ 
-  &    &    &  $W_x$  &  0.83  &  [0.76, 0.89]  \\ [10.25ex] \hline
-Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.43]  \\
-4477  &    &    &  $B_x-A_x$  &  0.27  &  [0.22, 0.33]  \\ 
-  &    &    &  $E_x$  &  0.82  &  [0.77, 0.87]  \\ 
-  &    &    &  $W_x$  &  0.62  &  [0.57, 0.66]  \\ [10.25ex] \hline
-Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.0, 0.18]  \\
-7164  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.08  &  [0.0, 0.24]  \\
-947  &    &    &  $B_x-A_x$  &  0.29  &  [0.2, 0.38]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kurmanji-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.09, 0.36]  \\
-634  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.16]  \\ 
-  &    &    &  $E_x$  &  0.86  &  [0.72, 0.95]  \\ 
-  &    &    &  $W_x$  &  0.96  &  [0.92, 1.0]  \\ [10.25ex] \hline
-Latvian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.51  &  [0.41, 0.62]  \\
-4124  &    &    &  $B_x-A_x$  &  -0.05  &  [-0.11, 0.01]  \\ 
-  &    &    &  $E_x$  &  0.37  &  [0.24, 0.53]  \\ 
-  &    &    &  $W_x$  &  0.66  &  [0.58, 0.74]  \\ [10.25ex] \hline
-Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.02, 0.25]  \\
-848  &    &    &  $B_x-A_x$  &  0.33  &  [0.21, 0.47]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.45]  \\
-2257  &    &    &  $B_x-A_x$  &  0.15  &  [0.08, 0.23]  \\ 
-  &    &    &  $E_x$  &  0.73  &  [0.62, 0.81]  \\ 
-  &    &    &  $W_x$  &  0.36  &  [0.28, 0.43]  \\ [10.25ex] \hline
-Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.04, 0.3]  \\
-29870  &    &    &  $B_x-A_x$  &  0.63  &  [0.48, 0.78]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.07, 0.32]  \\
-4798  &    &    &  $B_x-A_x$  &  0.12  &  [0.06, 0.2]  \\ 
-  &    &    &  $E_x$  &  0.93  &  [0.77, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [0.99, 1.0]  \\ [10.25ex] \hline
-Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.5]  \\
-6100  &    &    &  $B_x-A_x$  &  0.08  &  [-0.0, 0.15]  \\ 
-  &    &    &  $E_x$  &  0.6  &  [0.5, 0.69]  \\ 
-  &    &    &  $W_x$  &  0.31  &  [0.23, 0.39]  \\ [10.25ex] \hline
-Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\
-8664  &    &    &  $B_x-A_x$  &  0.15  &  [0.07, 0.24]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.06  &  [0.01, 0.15]  \\
-2935  &    &    &  $B_x-A_x$  &  0.19  &  [0.13, 0.26]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.13, 0.3]  \\
-8483  &    &    &  $B_x-A_x$  &  0.48  &  [0.39, 0.56]  \\ 
-  &    &    &  $E_x$  &  0.94  &  [0.9, 0.97]  \\ 
-  &    &    &  $W_x$  &  0.89  &  [0.83, 0.94]  \\ [10.25ex] \hline
-Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.31  &  [0.22, 0.39]  \\
-7532  &    &    &  $B_x-A_x$  &  0.44  &  [0.35, 0.56]  \\ 
-  &    &    &  $E_x$  &  0.85  &  [0.79, 0.92]  \\ 
-  &    &    &  $W_x$  &  0.79  &  [0.72, 0.85]  \\ [10.25ex] \hline
-Spanish* &\\ [10.25ex] \hline
-Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.29]  \\
-7041  &    &    &  $B_x-A_x$  &  0.06  &  [0.01, 0.12]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Tamil  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.04, 0.29]  \\
-400  &    &    &  $B_x-A_x$  &  0.17  &  [0.1, 0.28]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.06, 0.28]  \\
-900  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.27]  \\ 
-  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.18  &  [0.05, 0.35]  \\
-3685  &    &    &  $B_x-A_x$  &  0.03  &  [0.01, 0.06]  \\ 
-  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.22]  \\
-4506  &    &    &  $B_x-A_x$  &  0.18  &  [0.1, 0.24]  \\ 
-  &    &    &  $E_x$  &  0.94  &  [0.87, 1.0]  \\ 
-  &    &    &  $W_x$  &  0.99  &  [0.97, 1.0]  \\ [10.25ex] \hline
-Uyghur-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.06, 0.3]  \\
-1656  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
-  &    &    &  $E_x$  &  0.9  &  [0.0, 1.0]  \\ 
-  &    &    &  $W_x$  &  0.0  &  [0.0, 0.0]  \\ [10.25ex] \hline
-Vietnamese* &\\ [10.25ex] \hline
-\end{longtable}
-
 \end{document}

### 2019-03-02 code/corpusSizes.py

@@ -1,68 +1,13 @@
+# ./python27 corpusSizes.py > corpusSizes.tsv
 
-finalShib = {"Estonian" : "Estonian 0.16 200 256 1 0.0055",
-             "Vietnamese" : "vi 0.33 50 128 1 0.02",
-             "Slovenian" : "Slovenian 0.16 100 256 1 0.01",
-             "Finnish" : "Finnish 0.16 100 512 1 0.01",
-             "Polish" : "Polish 0.33 150 512 3 0.05",
-             "Ancient_Greek" : "Ancient_Greek 0.1 100 512 3 0.01",
-             "Old_Church_Slavonic" : "Old_Church_Slavonic 0.35 200 256 2 0.01",
-             "Czech" : "Czech 0.2 250 512 2 0.05",
-             "Swedish" : "Swedish 0.33 150 256 1 0.01",
-             "Hindi" : "Hindi 0.25 300 256 3 0.01",
-             "Basque" : "",
-             "Latin" : "Latin 0.25 200 512 2 0.01",
-             "German" : "German 0.2 150 512 2 0.01",
-             "Arabic" : "Arabic 0.2 300 1024 3 0.05",
-             "Turkish" : "Turkish 0.3 250 512 1 0.05",
-             "Old_French" : "Old_French 0.25 300 1024 1 0.005",
-             "Spanish" : "Spanish 0.25 200 1024 3 0.01",
-             "Latvian" : "Latvian 0.35 100 128 2 0.005",
-             "Indonesian" : "Indonesian 0.45 50 256 2 0.005",
-             "Hungarian" : "Hungarian 0.35 100 128 1 0.005",
-             "Persian" : "Persian 0.4 300 256 2 0.005",
-             "Croatian" : "Croatian 0.2 50 128 3 0.005",
-             "French" : "French 0.4 100 1024 2 0.005",
-             "ISWOC_Old_English" : "ISWOC_Old_English 0.3 50 256 1 0.001",
-             "TOROT_Old_Russian" : "TOROT_Old_Russian 0.1 50 128 2 0.01",
-             "Chinese" : "Chinese 0.3 300 128 3 0.005",
-             "Serbian" : "Serbian 0.45 200 256 1 0.01",
-             "Slovak" :  "Slovak 0.2 150 128 1 0.005",
-             "North_Sami" : " North_Sami 0.25 50 64 1 0.01 ",
[…]

### 2019-03-02 code/tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py

@@ -5,10 +5,6 @@ with open("../ud_languages.txt", "r") as inFile:
 
 languages = sorted(list(set(languages)))
 
-with open("corpusSizes.tsv", "r") as inFile:
-  corpusSizes = dict([x.split("\t") for x in inFile.read().strip().split("\n")])
-
-
 def readTSV(x):
     header = next(x).strip().split("\t")
     header = dict(zip(header, range(len(header))))
@@ -44,6 +40,13 @@ languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1
 
 
 
+with open("corpusSizes.tsv", "r") as inFile:
+  corpusSizes = readTSV(inFile) #dict([x.split("\t") for x in inFile.read().strip().split("\n")])
+languageKey_corpusSizes = dict([(h(corpusSizes, line, "Language"), line) for line in corpusSizes[1]])
+
+
+
+
 with open("../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
    listener_curve = readTSV(inFile)
 
@@ -58,15 +61,17 @@ for language in languages:
    satisfied = h(stats, line_stats, "RANDOM_BY_TYPE") >= 10 and h(stats, line_stats, "REAL_REAL") >= 5 and (h(listener_curve, line_listener, "result1High") - h(listener_curve, line_listener, "result1Low") <= 0.15)
 
    components = [language.replace("_"," ")+("*" if not satisfied else "")]
-   components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{figures/"+language+"-entropy-memory.pdf}}")
-   components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{figures/"+language+"-listener-surprisal-memory.pdf}}" )
-   components.append("$D_x$")
+   #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
+   components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}" )
+   components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}" )
[…]

### 2019-03-02 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_ConditionalMeans_All.py

@@ -1,0 +1,9 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_ConditionalMeans_All.py > ../results/tradeoff/listener-curve-means.tsv
+
+print "\t".join(["Language", "Position", "Type", "Memory", "MI_Mean", "MI_SD", "MI_SE"])
+for language in languages:
+   subprocess.call(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_ConditionalMeans.py", language])

### 2019-03-02 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem.py

@@ -1,0 +1,111 @@
+import sys
+
+language = sys.argv[1]
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data) < 2:
+       return (header, [])
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+with open("../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+     data = readTSV(inFile)
+#print(len(data))
+
+
+
+#data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
+
+import torch
+
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+
+matrix = [[g(data, "ModelID", i), g(data, "Distance", i), g(data, "ConditionalMI", i), g(data, "UnigramCE", i)] for i in range(len(data[1]))]
+
+matrixByType = {}
+misByType = {}
+#unigramCEByType = {}
+for i in range(len(data[1])):
+    typ = g(data, "Type", i)
+#    print(i,typ, len(data[1]))
+    if typ not in matrixByType:
+        matrixByType[typ] = []
+    matrixByType[typ].append(matrix[i])
+for typ in matrixByType:
+   matrixByType[typ] = torch.FloatTensor(matrixByType[typ]).view(-1, 19, 4)
+   #print(matrixByType[typ][0])
+   misByType[typ] = matrixByType[typ][:,:,2]
+#print(misByType["RANDOM_BY_TYPE"])
+
+#data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
+distance = torch.FloatTensor(range(1,20))
+
[…]

### 2019-03-02 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Histograms_All.py

@@ -1,0 +1,9 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Histograms_All.py > ../results/tradeoff/listener-curve-histogram_byMI.tsv
+
+print "\t".join(["Language", "Type", "MI", "Memory"])
+for language in languages:
+   subprocess.call(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Histograms.py", language])

### 2019-03-02 neural/plot_wordsOnly2.R

@@ -1,0 +1,53 @@
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Basque")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Bulgarian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Croatian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Estonian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hebrew")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Japanese")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Polish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Romanian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Slovak")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Slovenian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Swedish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Afrikaans")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Chinese")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Danish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Greek")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hungarian")
[…]

### 2019-03-02 neural/plot_wordsOnly_histByMem.R

@@ -1,0 +1,75 @@
+
+
+fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    UnigramCE = mean(dataL$UnigramCE)
+    data = fullData %>% filter(Language == language)
+    barWidth = (max(data$MI) - min(data$MI))/30
+    plot = ggplot(data, aes(x=UnigramCE-MI, fill=Type, color=Type)) + theme_classic() + theme(legend.position="none")   + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled..))      + geom_bar(data = data %>% filter(Type %in% c("REAL_REAL", "GROUND")) %>% group_by(Type) %>% summarise(MI=mean(MI)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
+    ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
+    return(plot)
+}
+
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
[…]

### 2019-03-03 paper.tex
word order freedom

@@
 
-\title{Crosslinguistic Word Orders Enable and Efficient Tradeoff between Memory and Surprisal}
+\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
+$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
+
 \begin{longtable}{lllllllllllllll}
@@
 
+\begin{figure}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}
+	\caption{Histogram}\label{fig:hist-real}
+\end{figure}
+
 \subsection{Discussion}
 
-Word Order Freedom and Memory Efficiency
+All four languages have relatively strong word order freedom.
+We hypothesized that freedom of word order freedm may be responsible for the difference between these languages and the other languages.
+We quantified word order freedom as the branching direction entropy~\cite{futrell-quantifying-2015}.
+We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
+
+\begin{figure}
+\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
+	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
+\end{figure}
+
+\section{Fixed Word Orders}
+
+We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
+This enables us to tease apart the impact of the languages' word order rules from the impact of more specific ordering choices.
+
+\begin{figure}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}
+	\caption{Histogram}\label{fig:hist-real}
+\end{figure}
 

### 2019-03-03 results-table-word-level.tex
word order freedom

@@
-Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Arabic  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Armenian-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.91  &  [0.83, 0.98]  \\ [10.25ex] \hline
-Bambara-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.96  &  [0.86, 1.0]  \\ [10.25ex] \hline
-Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-English  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Erzya-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.99  &  [0.98, 1.0]  \\ [10.25ex] \hline
-Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.83  &  [0.76, 0.9]  \\ [10.25ex] \hline
-Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-French  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.89  &  [0.81, 0.96]  \\ [10.25ex] \hline
-Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kurmanji-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.92  &  [0.83, 0.99]  \\ [10.25ex] \hline
-Latvian*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.48  &  [0.4, 0.57]  \\ [10.25ex] \hline
-Maltese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.37  &  [0.3, 0.44]  \\ [10.25ex] \hline
-Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.07  &  [0.01, 0.15]  \\ [10.25ex] \hline
-Portuguese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.07  &  [0.03, 0.12]  \\ [10.25ex] \hline
-Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.82  &  [0.74, 0.89]  \\ [10.25ex] \hline
-Spanish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Urdu  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Uyghur-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.65  &  [0.57, 0.74]  \\ [10.25ex] \hline
-Vietnamese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [0.98, 1.0]  \\ [10.25ex] \hline
+Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1315  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+974  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Arabic  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+21864  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Armenian-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+514  &    &    &    &  0.91  &  [0.83, 0.98]  \\ [12.25ex] \hline
+Bambara-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+926  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+5396  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+788  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+8907  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+808  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+550  &    &    &    &  0.96  &  [0.86, 1.0]  \\ [12.25ex] \hline
+Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+13123  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+3997  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+7689  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+102993  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4383  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+18310  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+English  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+17062  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Erzya-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1450  &    &    &    &  0.99  &  [0.98, 1.0]  \\ [12.25ex] \hline
+Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+6959  &    &    &    &  0.83  &  [0.76, 0.9]  \\ [12.25ex] \hline
+Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1108  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+27198  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+French  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+32347  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+13814  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1662  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+5241  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+13304  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+910  &    &    &    &  0.89  &  [0.81, 0.96]  \\ [12.25ex] \hline
+Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4477  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Italian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+17427  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+7164  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+947  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Kurmanji-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+634  &    &    &    &  0.92  &  [0.83, 0.99]  \\ [12.25ex] \hline
+Latvian*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4124  &    &    &    &  0.48  &  [0.4, 0.57]  \\ [12.25ex] \hline
+Maltese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1123  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+848  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+2257  &    &    &    &  0.37  &  [0.3, 0.44]  \\ [12.25ex] \hline
+Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+29870  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4798  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+6100  &    &    &    &  0.07  &  [0.01, 0.15]  \\ [12.25ex] \hline
+Portuguese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+17995  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+8664  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Russian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+52664  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+2935  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+8483  &    &    &    &  0.07  &  [0.03, 0.12]  \\ [12.25ex] \hline
+Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+7532  &    &    &    &  0.82  &  [0.74, 0.89]  \\ [12.25ex] \hline
+Spanish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+28492  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+7041  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+900  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+3685  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4506  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Urdu  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+4043  &    &    &    &  1.0  &  [1.0, 1.0]  \\ [12.25ex] \hline
+Uyghur-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1656  &    &    &    &  0.65  &  [0.57, 0.74]  \\ [12.25ex] \hline
+Vietnamese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEANS_onlyWordForms_boundedVocab.pdf}}  &    \\
+1400  &    &    &    &  1.0  &  [0.98, 1.0]  \\ [12.25ex] \hline

### 2019-03-03 code/branching_direction_entropy.py

@@ -1,0 +1,73 @@
+#/u/nlp/bin/stake.py -g 11.5g -s run-stats-pretrain2.json "python readDataDistEnglishGPUFree.py"
+
+import random
+import sys
+
+from math import log, exp
+from random import random, shuffle
+
+from corpusIterator import CorpusIterator
+
+from ud_languages import languages
+
+with open("branching_entropy.tsv", "w") as outFile:
+ print >> outFile, "Language\tBranchingEntropy"
+ for language in languages:
+    
+    posUni = set() #[ "ADJ", "ADP", "ADV", "AUX", "CONJ", "DET", "INTJ", "NOUN", "NUM", "PART", "PRON", "PROPN", "PUNCT", "SCONJ", "SYM", "VERB", "X"] 
+    
+    posFine = set() #[ "``", ",", ":", ".", "''", "$", "ADD", "AFX", "CC",  "CD", "DT", "EX", "FW", "GW", "HYPH", "IN", "JJ", "JJR",  "JJS", "-LRB-", "LS", "MD", "NFP", "NN", "NNP", "NNPS", "NNS",  "PDT", "POS", "PRP", "PRP$", "RB", "RBR", "RBS", "RP", "-RRB-", "SYM", "TO", "UH", "VB", "VBD", "VBG", "VBN", "VBP", "VBZ",  "WDT", "WP", "WP$", "WRB", "XX" ]
+    
+    deps = ["acl", "acl:relcl", "advcl", "advmod", "amod", "appos", "aux", "auxpass", "case", "cc", "ccomp", "compound", "compound:prt", "conj", "conj:preconj", "cop", "csubj", "csubjpass", "dep", "det", "det:predet", "discourse", "dobj", "expl", "foreign", "goeswith", "iobj", "list", "mark", "mwe", "neg", "nmod", "nmod:npmod", "nmod:poss", "nmod:tmod", "nsubj", "nsubjpass", "nummod", "parataxis", "punct", "remnant", "reparandum", "root", "vocative", "xcomp"] 
+    
+    #deps = ["acl", " advcl", " advmod", " amod", " appos", " aux", " case cc", " ccompclf", " compound", " conj", " cop", " csubjdep", " det", " discourse", " dislocated", " expl", " fixed", " flat", " goeswith", " iobj", " list", " mark", " nmod", " nsubj", " nummod", " obj", " obl", " orphan", " parataxis", " punct", " reparandum", " root", " vocative", " xcomp"]
+    
[…]

### 2019-03-03 neural/plot_wordsOnly_histByMem_AcrossLangs.R

@@ -14,76 +14,32 @@ fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", se
     plot = ggplot(data, aes(x=-MI_z, fill=Type, color=Type)) 
     plot = plot + theme_classic() 
 #    plot = plot + theme(legend.position="none")   
-    plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weights=1/counts))      
+    plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weight=1/counts))      
     plot = plot + geom_bar(data = data %>% filter(Type %in% c("REAL_REAL")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
     ggsave(plot, file=paste("figures/full-REAL-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
 
     plot = ggplot(data, aes(x=-MI_z, fill=Type, color=Type)) 
     plot = plot + theme_classic() 
 #    plot = plot + theme(legend.position="none")   
-    plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weights=1/counts))      
+    plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weight=1/counts))      
     plot = plot + geom_bar(data = data %>% filter(Type %in% c("GROUND")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
     ggsave(plot, file=paste("figures/full-GROUND-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
 
 
+branching = read.csv("../code/branching_entropy_coarse.tsv", sep="\t")
+data = merge(data, branching, by=c("Language"))
+
[…]

### 2019-03-07 paper.tex
figures

@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 	\caption{Histogram}\label{fig:hist-real}
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 	\caption{Histogram}\label{fig:hist-real}

### 2019-03-07 neural/plot_wordsOnly_histByMem_AcrossLangs.R

@@ -9,37 +9,35 @@ fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", se
 
     transform = fullData %>% filter(Type == "RANDOM_BY_TYPE") %>% group_by(Language) %>% summarise(mean=mean(MI), sd=sd(MI), counts=NROW(MI))
     data = merge(fullData, transform, by=c("Language"))
-    data = data %>% mutate(MI_z = (MI-mean)/sd)
+    data = data %>% mutate(MI_z = (MI-mean)/sd, MIDiff = MI-mean)
     barWidth = 0.05 #(max(data$MI) - min(data$MI))/30
+
     plot = ggplot(data, aes(x=-MI_z, fill=Type, color=Type)) 
     plot = plot + theme_classic() 
 #    plot = plot + theme(legend.position="none")   
     plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weight=1/counts))      
     plot = plot + geom_bar(data = data %>% filter(Type %in% c("REAL_REAL")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
-    ggsave(plot, file=paste("figures/full-REAL-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
+    ggsave(plot, file=paste("figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
 
     plot = ggplot(data, aes(x=-MI_z, fill=Type, color=Type)) 
     plot = plot + theme_classic() 
 #    plot = plot + theme(legend.position="none")   
     plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weight=1/counts))      
     plot = plot + geom_bar(data = data %>% filter(Type %in% c("GROUND")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
[…]

### 2019-03-11 paper.tex
paper

--- 2019-03-02 paper.tex
@@
 
-\title{Crosslinguistic Word Orders Enable and Efficient Tradeoff between Memory and Surprisal}
+\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
+Are languages optimized for processing with limited memory?
+Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
+Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
+We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
 %assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
+Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
 % producing language and listeners predicting input.
@@
 
+\paragraph{Other Psycholinguistic Predictions}
+
+% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries.
+
+\paragraph{Speakers}
+
+% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
+This changes the interpretation of the mutual information.
+For the listener, it’s just redundancy.
+For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
+Then the speaker’s pressure is to keep conceptually related things close.
+
 \section{Relation to Theories of Sentence Processing}
@@
 
-\begin{longtable}{lllllllllllllll}
-\input{results-table-word-level.tex}
-\end{longtable}
-
 \subsection{Discussion}
@@
 
+\section{Results (Morphology on UD Treebanks)}\label{sec:results-table}
+
+Statistics reported:
+
+\begin{enumerate}
+%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
+	\item $A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$
+	\item $B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$
+	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$
+	\item Compute $A_{x'}$ for each $x'$ in the baseline distribution, and then calculate the percentile rank of $A_x$ among these values:
+$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$
+\item For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.
+
+\end{enumerate}
+
+For each language, we show the number of sentences in the training partition, the surprisal-speaker memory tradeoff plane, the listener's surprisal-memory tradeoff curves (smoothed means), and the statistics defined above with bootstraped CIs.
+
+Languages where the stopping criterion hasn't been reached are marked with asterisks.
+
+\begin{longtable}{lllllllllllllll}
+Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.28]  \\
+1315  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
+  &    &    &  $E_x$  &  0.42  &  [0.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  0.99  &  [0.96, 1.0]  \\ [10.25ex] \hline
+Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.16  &  [0.06, 0.29]  \\
+974  &    &    &  $B_x-A_x$  &  0.1  &  [0.04, 0.18]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Arabic*  &\\ [10.25ex] \hline
+Armenian-Adap*  &\\ [10.25ex] \hline
+Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.03, 0.15]  \\
+5396  &    &    &  $B_x-A_x$  &  0.82  &  [0.76, 0.88]  \\ 
+  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.03, 0.22]  \\
+788  &    &    &  $B_x-A_x$  &  0.32  &  [0.2, 0.42]  \\ 
+  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.22]  \\
+8907  &    &    &  $B_x-A_x$  &  0.56  &  [0.44, 0.7]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.02, 0.19]  \\
+808  &    &    &  $B_x-A_x$  &  0.28  &  [0.19, 0.42]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.08, 0.31]  \\
+550  &    &    &  $B_x-A_x$  &  0.25  &  [0.17, 0.36]  \\ 
+  &    &    &  $E_x$  &  0.92  &  [0.85, 0.98]  \\ 
+  &    &    &  $W_x$  &  0.94  &  [0.86, 1.0]  \\ [10.25ex] \hline
+Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.03, 0.25]  \\
+13123  &    &    &  $B_x-A_x$  &  0.44  &  [0.33, 0.57]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.04, 0.23]  \\
+3997  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.19]  \\ 
+  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.03, 0.23]  \\
+7689  &    &    &  $B_x-A_x$  &  0.13  &  [0.05, 0.23]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.05, 0.27]  \\
+102993  &    &    &  $B_x-A_x$  &  0.66  &  [0.56, 0.77]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  0.94  &  [0.86, 0.99]  \\ [10.25ex] \hline
+Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.26]  \\
+4383  &    &    &  $B_x-A_x$  &  0.15  &  [0.06, 0.25]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.01, 0.15]  \\
+18310  &    &    &  $B_x-A_x$  &  0.04  &  [0.01, 0.08]  \\ 
+  &    &    &  $E_x$  &  0.88  &  [0.73, 1.0]  \\ 
+  &    &    &  $W_x$  &  0.98  &  [0.96, 1.0]  \\ [10.25ex] \hline
+Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.48]  \\
+6959  &    &    &  $B_x-A_x$  &  0.26  &  [0.18, 0.36]  \\ 
+  &    &    &  $E_x$  &  0.76  &  [0.67, 0.84]  \\ 
+  &    &    &  $W_x$  &  0.72  &  [0.65, 0.79]  \\ [10.25ex] \hline
+Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.26]  \\
+1108  &    &    &  $B_x-A_x$  &  0.34  &  [0.24, 0.45]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.01, 0.12]  \\
+27198  &    &    &  $B_x-A_x$  &  0.95  &  [0.91, 0.99]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+French*  &\\ [10.25ex] \hline
+German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.44  &  [0.37, 0.51]  \\
+13814  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.29]  \\ 
+  &    &    &  $E_x$  &  0.66  &  [0.6, 0.74]  \\ 
+  &    &    &  $W_x$  &  0.63  &  [0.57, 0.69]  \\ [10.25ex] \hline
+Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.2]  \\
+1662  &    &    &  $B_x-A_x$  &  0.59  &  [0.48, 0.73]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\
+5241  &    &    &  $B_x-A_x$  &  0.19  &  [0.08, 0.3]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.24]  \\
+13304  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.55  &  [0.44, 0.65]  \\
+910  &    &    &  $B_x-A_x$  &  0.07  &  [0.0, 0.15]  \\ 
+  &    &    &  $E_x$  &  0.62  &  [0.51, 0.74]  \\ 
+  &    &    &  $W_x$  &  0.83  &  [0.76, 0.89]  \\ [10.25ex] \hline
+Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.43]  \\
+4477  &    &    &  $B_x-A_x$  &  0.27  &  [0.22, 0.33]  \\ 
+  &    &    &  $E_x$  &  0.82  &  [0.77, 0.87]  \\ 
+  &    &    &  $W_x$  &  0.62  &  [0.57, 0.66]  \\ [10.25ex] \hline
+Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.0, 0.18]  \\
+7164  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.08  &  [0.0, 0.24]  \\
+947  &    &    &  $B_x-A_x$  &  0.29  &  [0.2, 0.38]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Kurmanji-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.09, 0.36]  \\
+634  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.16]  \\ 
+  &    &    &  $E_x$  &  0.86  &  [0.72, 0.95]  \\ 
+  &    &    &  $W_x$  &  0.96  &  [0.92, 1.0]  \\ [10.25ex] \hline
+Latvian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.51  &  [0.41, 0.62]  \\
+4124  &    &    &  $B_x-A_x$  &  -0.05  &  [-0.11, 0.01]  \\ 
+  &    &    &  $E_x$  &  0.37  &  [0.24, 0.53]  \\ 
+  &    &    &  $W_x$  &  0.66  &  [0.58, 0.74]  \\ [10.25ex] \hline
+Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.02, 0.25]  \\
+848  &    &    &  $B_x-A_x$  &  0.33  &  [0.21, 0.47]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.45]  \\
+2257  &    &    &  $B_x-A_x$  &  0.15  &  [0.08, 0.23]  \\ 
+  &    &    &  $E_x$  &  0.73  &  [0.62, 0.81]  \\ 
+  &    &    &  $W_x$  &  0.36  &  [0.28, 0.43]  \\ [10.25ex] \hline
+Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.04, 0.3]  \\
+29870  &    &    &  $B_x-A_x$  &  0.63  &  [0.48, 0.78]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.07, 0.32]  \\
+4798  &    &    &  $B_x-A_x$  &  0.12  &  [0.06, 0.2]  \\ 
+  &    &    &  $E_x$  &  0.93  &  [0.77, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [0.99, 1.0]  \\ [10.25ex] \hline
+Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.5]  \\
+6100  &    &    &  $B_x-A_x$  &  0.08  &  [-0.0, 0.15]  \\ 
+  &    &    &  $E_x$  &  0.6  &  [0.5, 0.69]  \\ 
+  &    &    &  $W_x$  &  0.31  &  [0.23, 0.39]  \\ [10.25ex] \hline
+Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\
+8664  &    &    &  $B_x-A_x$  &  0.15  &  [0.07, 0.24]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.06  &  [0.01, 0.15]  \\
+2935  &    &    &  $B_x-A_x$  &  0.19  &  [0.13, 0.26]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.13, 0.3]  \\
+8483  &    &    &  $B_x-A_x$  &  0.48  &  [0.39, 0.56]  \\ 
+  &    &    &  $E_x$  &  0.94  &  [0.9, 0.97]  \\ 
+  &    &    &  $W_x$  &  0.89  &  [0.83, 0.94]  \\ [10.25ex] \hline
+Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.31  &  [0.22, 0.39]  \\
+7532  &    &    &  $B_x-A_x$  &  0.44  &  [0.35, 0.56]  \\ 
+  &    &    &  $E_x$  &  0.85  &  [0.79, 0.92]  \\ 
+  &    &    &  $W_x$  &  0.79  &  [0.72, 0.85]  \\ [10.25ex] \hline
+Spanish* &\\ [10.25ex] \hline
+Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.29]  \\
+7041  &    &    &  $B_x-A_x$  &  0.06  &  [0.01, 0.12]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Tamil  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.04, 0.29]  \\
+400  &    &    &  $B_x-A_x$  &  0.17  &  [0.1, 0.28]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.06, 0.28]  \\
+900  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.27]  \\ 
+  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.18  &  [0.05, 0.35]  \\
+3685  &    &    &  $B_x-A_x$  &  0.03  &  [0.01, 0.06]  \\ 
+  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.22]  \\
+4506  &    &    &  $B_x-A_x$  &  0.18  &  [0.1, 0.24]  \\ 
+  &    &    &  $E_x$  &  0.94  &  [0.87, 1.0]  \\ 
+  &    &    &  $W_x$  &  0.99  &  [0.97, 1.0]  \\ [10.25ex] \hline
+Uyghur-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.06, 0.3]  \\
+1656  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
+  &    &    &  $E_x$  &  0.9  &  [0.0, 1.0]  \\ 
+  &    &    &  $W_x$  &  0.0  &  [0.0, 0.0]  \\ [10.25ex] \hline
+Vietnamese* &\\ [10.25ex] \hline
+\end{longtable}
+
 \end{document}

### 2019-03-14 paper.tex
quantile plot

--- 2019-03-07 paper.tex
@@
 
-First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-Formally: 
-\begin{equation}\label{eq:listener-markov-1}
-L_t \bot X_{>t} | X_{\leq t}
+We assume that $L_t$ contains no information about the process beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
+	\begin{equation}\label{eq:listener-markov}
+p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
 	\end{equation}
-This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
 
-Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-\begin{equation}\label{eq:listener-markov-2}
-L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-\end{equation}
-This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
+%
+%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
+%Formally: 
+%\begin{equation}\label{eq:listener-markov-1}
+%L_t \bot X_{>t} | X_{\leq t}
+%\end{equation}
+%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
+%
+%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
+%\begin{equation}\label{eq:listener-markov-2}
+%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
+%\end{equation}
+%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
 
@@
 For any $t$,
-	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots T}, L_0]$$
+	\begin{equation}
+H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
+		\end{equation}
 	This can be derived from the two conditions \ref{eq:listener-markov-1} and \ref{eq:listener-markov-2} (TODO explain how).
+
+	\begin{equation}
+H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
+	\end{equation}
+is true because the following is a Markov chain:
+\begin{equation}
+(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
+\end{equation}
+because
+\begin{align*}
+p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) & = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) \\
+& = \sum_{X_{<1}} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1}) \\
+&= \sum_{X_{<1}} p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, X_1, X_{2 \dots t-1}) \\
+&= p(X_t|L_0,  X_1, X_{2 \dots t-1}) 
+\end{align*}
+because
+	\begin{align*}
+p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
+p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
+	\end{align*}
+because of Equation~\ref{eq:listener-markov}.
+	\begin{equation}
+p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
+	\end{equation}
+
+	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
+
+%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
 
@@
 \begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}]& \geq   \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
+	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( H[X_t|X_{1\dots t-1}, L_0] - H[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
+	& = \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
 	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 

### 2019-03-14 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py

@@ -118,14 +118,18 @@ import statsmodels.stats.proportion
 
 
 CIs = {}
+medians = {}
 
 for typ, interpolated in interpolatedByTypes.iteritems():  
     CIs[typ] = []
+    medians[typ] = []
 
     for i in range(39):
        
        hereReal = torch.sort(interpolated[:,i])[0]      
        realCount = hereReal.size()[0]
+
+       medians[typ].append(float(torch.median(hereReal)))
 
        probsBinomial = [scipy.stats.binom.pmf(j+1, realCount, 0.5) for j in range(realCount)]
        
@@ -146,7 +150,7 @@ for typ, interpolated in interpolatedByTypes.iteritems():
 
 for real in ["REAL_REAL", "GROUND"]:
     for i in range(39):
-       print "\t".join(map(str, [language, real, CIs[real][i][0] - CIs["RANDOM_BY_TYPE"][i][1], CIs[real][i][1] - CIs["RANDOM_BY_TYPE"][i][0], CIs[real][i][2] * CIs["RANDOM_BY_TYPE"][i][2]]))
+       print "\t".join(map(str, [language, real, float(xPoints[i]), medians[real][i] - medians["RANDOM_BY_TYPE"][i], CIs[real][i][0] - CIs["RANDOM_BY_TYPE"][i][1], CIs[real][i][1] - CIs["RANDOM_BY_TYPE"][i][0], CIs[real][i][2] * CIs["RANDOM_BY_TYPE"][i][2]]))
 
 
 

### 2019-03-14 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI_All.py

@@ -5,7 +5,7 @@ import subprocess
 # ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
 
 with open("../results/tradeoff/listener-curve-ci-median_diff.tsv", "w") as outFile:
-  print >> outFile, "\t".join(["Language", "Type", "MedianDiff_Lower", "MedianDiff_Upper", "Level"])
+  print >> outFile, "\t".join(["Language", "Type", "Memory", "EmpiricalMedianDiff", "MedianDiff_Lower", "MedianDiff_Upper", "Level"])
   for language in languages:
      print(language)
      print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py", language]).strip()

### 2019-03-14 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py

@@ -136,19 +136,22 @@ for real in ["REAL_REAL", "GROUND"]:
 
      # The goal here is to provide a confidence lower bound on worseRandomCount
      if worseRandomCount == 0:
+         print "\t".join(map(str,[language, real, i, 0.0, 0.0, float(xPoints[i])]))
          bound = 0
      else:
    #     print(hereRandom.size())
         largestPossibleValue = xPoints[i]
         #print(i, largestPossibleValue, medians[real][i], hereRandom)
         # if the median is at most the best sample
-        targetLevel = 0.05
         #unaccountedForTarget = 1-math.pow(targetLevel, 1/(float(hereRandom.size()[0])))
         randomBestWorse = float(hereRandom[int(worseRandomCount)-1])
         assert randomBestWorse < float(medians[real][i])
 #        print((largestPossibleValue, medians[real][i] , randomBestWorse,   (largestPossibleValue - medians[real][i] + 0.0001) / (largestPossibleValue - randomBestWorse + 0.0001)))
 
-        for unaccountedFor in [float(x)/200 for x in range(1,200)]:
+        hasFound = False # TODO in reality this should be binary search, as the map is monotonic, would be much faster
+        for limit in [200,800,2000,10000]:
+#         print(limit)
+         for unaccountedFor in [float(x)/limit for x in range(1,limit)]:
 #          print scipy.stats.binom_test(x=1, n=10, p=0.5, alternative="greater")
           assert unaccountedFor > 0
           assert unaccountedFor < 1
@@ -158,8 +161,14 @@ for real in ["REAL_REAL", "GROUND"]:
              p1 = math.pow(1-unaccountedFor, worseRandomCount)
           else:
              p1 = 1-(scipy.stats.binom_test(x=sameRandomCount+betterRandomCount, n=worseRandomCount+sameRandomCount+betterRandomCount, p=unaccountedFor, alternative="greater"))
-          if p1 < 0.05:
-             print "\t".join(map(str,[language, real, 1-float(percentile), p1, float(xPoints[i])]))
+        #  print(unaccountedFor, p1)
+          if p1 < 0.001:
+             print "\t".join(map(str,[language, real, i, 1-float(percentile), p1, float(xPoints[i])]))
+             hasFound =True 
              break
-
+         if hasFound:
+             break
+        assert hasFound, (worseRandomCount, sameRandomCount, betterRandomCount)
+        if not hasFound:
+             print "\t".join(map(str,[language, real, i, 0.0, 0.0, float(xPoints[i])]))
 

### 2019-03-14 code/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_All.py

@@ -5,7 +5,7 @@ import subprocess
 # ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
 
 with open("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", "w") as outFile:
-  print >> outFile, "\t".join(["Language", "Type", "LowerConfidenceBound", "Level", "Memory"])
+  print >> outFile, "\t".join(["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
   for language in languages:
      print(language)
      print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py", language]).strip()

### 2019-03-14 neural/order_freedom.R

@@ -1,0 +1,39 @@
+fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    transform = fullData %>% filter(Type == "RANDOM_BY_TYPE") %>% group_by(Language) %>% summarise(mean=mean(MI), sd=sd(MI), counts=NROW(MI))
+    data = merge(fullData, transform, by=c("Language"))
+    data = data %>% mutate(MI_z = (MI-mean)/sd)
+    barWidth = 0.05 #(max(data$MI) - min(data$MI))/30
+
+branching = read.csv("../code/branching_entropy.tsv", sep="\t")
+data = merge(data, branching, by=c("Language"))
+
+dataReal = data %>% filter(Type %in% c("REAL_REAL")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z), MIDiff = mean(MI-mean), BranchingEntropy=mean(BranchingEntropy)) %>% mutate(y=1)
+#cor.test(dataReal$MI_z, dataReal$BranchingEntropy, method="spearman")
+cor.test(dataReal$MIDiff, dataReal$BranchingEntropy, method="spearman")
+plot = ggplot(dataReal, aes(x=-MIDiff, y=BranchingEntropy)) + geom_point()
+plot = plot + theme_classic() 
+plot = plot + theme(legend.position="none")   
+plot = plot + geom_text(aes(label=Language), hjust=0.8, vjust=1.1)
+#plot = plot + xlim(-2.5, 0.5)
+
+ggsave(plot, file=paste("figures/surprisal-branching-entropy-REAL.pdf", sep=""))
+
+
+
+dataGround = data %>% filter(Type %in% c("GROUND")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z), MIDiff = mean(MI-mean), BranchingEntropy=mean(BranchingEntropy)) %>% mutate(y=1)
+#cor.test(dataGround$MI_z, dataGround$BranchingEntropy, method="spearman")
+cor.test(dataGround$MIDiff, dataGround$BranchingEntropy, method="spearman")
+plot = ggplot(dataGround, aes(x=-MIDiff, y=BranchingEntropy)) + geom_point()
+plot = plot + theme_classic() 
+plot = plot + theme(legend.position="none")   
[…]

### 2019-03-14 neural/plot_wordsOnly_medianDiff.R

@@ -1,46 +1,17 @@
 
+# Plots median difference with confidence intervals
+
+fullData = read.csv("../results/tradeoff/listener-curve-ci-median_diff.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-#    data = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-    data = read.csv(paste("~/CS_SCR/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-
     library(tidyr)
     library(dplyr)
     library(ggplot2)
-    data = data %>% filter(Type %in% c("RANDOM_BY_TYPE", "REAL_REAL", "GROUND"))
-
-
-    data2 = data %>% filter(Distance==1) %>% mutate(ConditionalMI=0, Distance=0)
-    data = rbind(data2, data)
-
-#data$UnigramCE = mean(data$UnigramCE, na.rm=TRUE)
-
-    data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI), Surprisal=UnigramCE-CumulativeMI)
-
-
-    
-
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=Type, fill=Type, color=Type, alpha=0.5)) + geom_smooth()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory_onlyWordForms_boundedVocab.pdf", sep=""))
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=ModelID, fill=Type, color=Type, alpha=0.5)) + geom_line()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf", sep=""))
-
-
-
-#    data3 = data %>% group_by(ModelID, Type) %>% summarise(Surprisal=min(Surprisal), CumulativeMI = max(CumulativeMI))
-#    data3$CumulativeMemory = max(data$CumulativeMemory)
[…]

### 2019-03-14 neural/plot_wordsOnly_medians.R

@@ -1,46 +1,17 @@
 
+# Plots medians with confidence intervals
+
+fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-#    data = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-    data = read.csv(paste("~/CS_SCR/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-
     library(tidyr)
     library(dplyr)
     library(ggplot2)
-    data = data %>% filter(Type %in% c("RANDOM_BY_TYPE", "REAL_REAL", "GROUND"))
-
-
-    data2 = data %>% filter(Distance==1) %>% mutate(ConditionalMI=0, Distance=0)
-    data = rbind(data2, data)
-
-#data$UnigramCE = mean(data$UnigramCE, na.rm=TRUE)
-
-    data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI), Surprisal=UnigramCE-CumulativeMI)
-
-
-    
-
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=Type, fill=Type, color=Type, alpha=0.5)) + geom_smooth()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory_onlyWordForms_boundedVocab.pdf", sep=""))
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=ModelID, fill=Type, color=Type, alpha=0.5)) + geom_line()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf", sep=""))
-
-
-
-#    data3 = data %>% group_by(ModelID, Type) %>% summarise(Surprisal=min(Surprisal), CumulativeMI = max(CumulativeMI))
-#    data3$CumulativeMemory = max(data$CumulativeMemory)
[…]

### 2019-03-14 neural/plot_wordsOnly_quantile.R

@@ -1,46 +1,34 @@
 
+# Plots CIs for the quantile
+
+fullData_ConfidenceLowerBound = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", sep="\t")
+fullData_ConfidenceLowerBound_05 = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05.tsv", sep="\t")
+
+fullData_BinomialTest = read.csv("../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-#    data = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-    data = read.csv(paste("~/CS_SCR/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
-
     library(tidyr)
     library(dplyr)
     library(ggplot2)
-    data = data %>% filter(Type %in% c("RANDOM_BY_TYPE", "REAL_REAL", "GROUND"))
-
-
-    data2 = data %>% filter(Distance==1) %>% mutate(ConditionalMI=0, Distance=0)
-    data = rbind(data2, data)
-
-#data$UnigramCE = mean(data$UnigramCE, na.rm=TRUE)
-
-    data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI), Surprisal=UnigramCE-CumulativeMI)
-
-
-    
-
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=Type, fill=Type, color=Type, alpha=0.5)) + geom_smooth()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory_onlyWordForms_boundedVocab.pdf", sep=""))
-#    plot = ggplot(data, aes(x=Surprisal, y=CumulativeMemory, group=ModelID, fill=Type, color=Type, alpha=0.5)) + geom_line()+ theme_classic() + theme(legend.position="none")
-#    ggsave(plot, file=paste("figures/",language,"-listener-reverse-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf", sep=""))
-
-
-
[…]

### 2019-03-16 medians_0.tex
revising paper

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-16 paper.tex
revising paper

@@
 
+\usepackage{fullpage}
 \usepackage{pslatex}
@@
 
+also surprisal -- integration cost
+
 \paragraph{Center Embeddings, ...}
@@
 \begin{proof}[Proof Sketch]
+	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
+
 	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
@@
 We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-We excluded four languages for a later confirmatory experiment (English, Korean, Russian). % Irish
-This resulted in X languages.
+This resulted in 52 languages.
+
+UD models predefined with train, dev, and test splits.
+Small datasets sometimes not.
 
@@
 For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-
-UD models predefined with train, dev, and test splits.
-Small datasets sometimes not.
+We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
 
@@
 
-\paragraph{Stopping Criterion}
-
-For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
+\paragraph{Number of Samples}
+Training neural language models is computationally costly.
+Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
+Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
+
+For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
 We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
+Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
+
+This procedure was parallelized on several machines.
+In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
+Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
+
+%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
+
 %We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
@@
 
+The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
+
+In Figrue~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves.
+
+In Figure~\ref{tab:medians}, we show the medians.
+
+Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
+
+- show the slice histograms
+
+- show the total histogram
+
+\paragraph{Inferential Statistics}
+
+inferential
+
+- report the bootstrapped thing G
+
+Median Differences \ref{tab:median_diffs}
+
+Quantiles \ref{tab:quantiles}
+
+\begin{table}
+\begin{longtable}{l|ll||l|llllllllllllll}
+	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
+\input{corpusSizes.tex}
+\end{longtable}
+	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
+\end{table}
+
+\begin{table}
+\begin{longtable}{l|ll||l|llllllllllllll}
+	Language & Base. & Real & Language & Base. & Real \\ \hline
+\input{samplesNumber.tex}
+\end{longtable}
+	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_0.tex}
+\end{longtable}
+	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample means, dashed lines indicate 95 \% confidence intervals.}\label{tab:medians}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_1.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_2.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_3.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccccll}
+\input{medianDiff_0.tex}
+\end{longtable}
+	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines.
+Lower values indicate lower surprisal compared to baselines.
+Solid lines indicate sample means.
+Dashed lines indicate 95 \% confidence intervals.}\label{tab:median_diffs}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccccll}
+\input{medianDiff_1.tex}
+\end{longtable}
+	\caption{Median Differences (Part 2)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccccll}
+\input{medianDiff_2.tex}
+\end{longtable}
+	\caption{Median Differences (Part 3)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccccll}
+\input{medianDiff_3.tex}
+\end{longtable}
+	\caption{Median Differences (Part 4)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_0.tex}
+\end{longtable}
+	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_1.tex}
+\end{longtable}
+	\caption{Quantiles (part 2)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_2.tex}
+\end{longtable}
+	\caption{Quantiles (part 3)}
+\end{table}
+
 In all but X languages $\E[G] > 0.5$ significantly.
@@
 
-\begin{longtable}{lllllllllllllll}
-\input{results-table-word-level.tex}
-\end{longtable}
+%\begin{longtable}{llllllllllllllllll}
+	%\input{results-table-word-level.tex}
+%\end{longtable}
 
@@
 
-All four languages have relatively strong word order freedom.
-We hypothesized that freedom of word order freedm may be responsible for the difference between these languages and the other languages.
-We quantified word order freedom as the branching direction entropy~\cite{futrell-quantifying-2015}.
+We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
+
+Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
+All four languages have strong word order freedom (CITE).
+Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
+We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
+%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
+
+To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
+To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
+This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
+%\cite{futrell-quantifying-2015} showed that this me
+These two quantities are plotted in Figure~\ref{fig:hist-real}.
 We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
@@
 
+\paragraph{Fitting Ordering Grammars to Actual Orders}
+We additionally create ordering grammars that are fit to the actual orderings of each language.
+These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
+
+We construct these grammars by interpreting ordering grammars probabilistically, and setting the parameters
+
+We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
+Thus, random grammars with by-relation weights constitute a stronger baseline.
+
 \bibliographystyle{apalike}
@@
 
-\section{Experiment with MLE Grammars}
-
-\paragraph{Maximum-Likelihood Grammars}
-We additionally create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by interpreting ordering grammars probabilistically, and setting the parameters
-
-We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
-Thus, random grammars with by-relation weights constitute a stronger baseline.
-
 \section{N-Gram Models}
 
-\section{Morphology-Based Encoding}
-
-Held-out:
-
-- UD languages:
-
--- English, Korean, Russian
-
--- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-
--- character-level Russian
+%-- English, Korean, Russian
+%
+%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
+%
+%-- character-level Russian
+%
 

### 2019-03-16 quantiles_0.tex
revising paper

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-16 code/bootstrap/yStudyTradeoff_Bootstrap_OnlyWordForms_BoundedVocab_FORALL.py

@@ -5,18 +5,13 @@
 import sys
 realType = sys.argv[1]
 
-languages = []
-languages += ["Arabic", "Catalan", "Czech", "Dutch", "Finnish", "French", "German", "Hindi", "Norwegian", "Spanish"]
-languages += ["Basque", "Bulgarian", "Croatian", "Estonian", "Hebrew", "Japanese", "Polish", "Romanian", "Slovak", "Slovenian", "Swedish"]
-languages += ["Afrikaans", "Chinese", "Danish", "Greek", "Hungarian",  "North_Sami", "Persian", "Serbian", "Turkish", "Ukrainian", "Vietnamese"]
-languages += ["Amharic-Adap", "Armenian-Adap",  "Breton-Adap",  "Buryat-Adap", "Cantonese-Adap","Faroese-Adap", "Kazakh-Adap", "Kurmanji-Adap", "Naija-Adap","Thai-Adap", "Uyghur-Adap"]
-languages += ["Bambara-Adap", "Erzya-Adap", "Maltese", "Latvian"]
+with open("../../ud_languages.txt", "r") as inFile:
+   languages = inFile.read().strip().split("\n")
 
-languages += ["Indonesian", "Urdu" , "Portuguese", "English", "Italian", "Russian"]
 languages = set(languages)
 
 import subprocess
-outPath = "../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_"+("REAL" if realType == "REAL_REAL" else "GROUND")+".tsv"
+outPath = "../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_"+("REAL" if realType == "REAL_REAL" else "GROUND")+".tsv"
 print(outPath)
 with open(outPath, "w") as outFile:
    print >> outFile, "\t".join(["language", "result1Mean", "result2Mean", "result1Low", "result1High", "result2Low", "result2High", "result3Mean", "result3Low", "result3High"])

### 2019-03-16 code/corpus_size/corpusSizes.py

@@ -1,13 +1,16 @@
 # ./python27 corpusSizes.py > corpusSizes.tsv
 
 
-from ud_languages import languages
-
 from corpusIterator import CorpusIterator
 
-with open("corpusSizes.tsv", "w") as outFile:
-  print >> outFile, ("Language\tTrainingSents")
+from ud_languages import languages
+
+
+with open("../corpusSizes.tsv", "w") as outFile:
+  print >> outFile, ("Language\tTrainingSents\tHeldoutSents")
   for language in languages:
     train = [x for x in CorpusIterator(language,"train", storeMorph=False).iterator()]
-    print >> outFile, (language+"\t"+str( len(train)))
+    heldout = [x for x in CorpusIterator(language,"dev", storeMorph=False).iterator()]
+
+    print >> outFile, (language+"\t"+str( len(train))+"\t"+str(len(heldout)))
 

### 2019-03-16 code/preparePaper/corpusSizesTable.py

@@ -45,9 +45,13 @@ with open("../corpusSizes.tsv", "r") as inFile:
 languageKey_corpusSizes = dict([(h(corpusSizes, line, "Language"), line) for line in corpusSizes[1]])
 
 
+output = []
 
-
-
+def pretty(x):
+    x = str(x)
+    if len(x) > 3:
+        x = x[:-3] + "," + x[-3:]
+    return x
 
 for language in languages:
 #   line = languageKey[language]
@@ -55,8 +59,16 @@ for language in languages:
 
 
    components = [language.replace("_"," ").replace("-Adap", "")]
-   components.append(h(corpusSizes,languageKey_corpusSizes[language], "TrainingSents"))
-   components.append(h(corpusSizes,languageKey_corpusSizes[language], "HeldoutSents"))
-   print("  &  ".join([str(x) for x in components]) + "  \\\\") # [10.25ex] \\hline" )
+   components.append(pretty(h(corpusSizes,languageKey_corpusSizes[language], "TrainingSents")))
+   components.append(pretty(h(corpusSizes,languageKey_corpusSizes[language], "HeldoutSents")))
+   output.append(components)
+
+if len(output)/2 * 2 < len(output):
+    output.append(["","",""])
+
+with open("../../corpusSizes.tex", "w") as outFile:
+ for i in range(len(output)/2):
+    here = output[i] + output[len(output)/2+i]
+    print >> outFile, ("  &  ".join([str(x) for x in here]) + "  \\\\") # [10.25ex] \\hline" )
 
 

### 2019-03-16 code/preparePaper/medianDifference.py

@@ -42,13 +42,41 @@ with open("../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL
 
 languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
 
+entries = []
+
 for language in languages:
 #   line = languageKey[language]
    line_listener = languageKey_listener_curve[language]
 
 
    components = [language.replace("_"," ").replace("-Adap", "")]
-   components.append( "\\multirow{4}{*}{\includegraphics[width=0.1\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}" )
-   print("  &  ".join([str(x) for x in components]) + "  \\\\ [10.25ex] \\hline" )
+   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}" )
+   entries.append(components)
+
+ROWS_PER_PART = 4
+COLUMNS = 4
+entries += [["",""] for x in range((COLUMNS-(len(entries)%COLUMNS))%COLUMNS)]
+
+outputRows = []
+if True:
+    languages = [x[0] for x in entries]
+    images = [x[1] for x in entries]
+
+    for i in range(len(entries)/COLUMNS):
+       outputRows.append( " & ".join(languages[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+       outputRows.append( " & ".join(images[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+
+for part in range(len(entries)/(COLUMNS*ROWS_PER_PART)+1):
+  with open("../../medianDiff_"+str(part)+".tex", "w") as outFile:
+      for row in range(part*ROWS_PER_PART, (part+1)*ROWS_PER_PART):
+          if 4*row >= len(outputRows):
+              break
+          print >> outFile, outputRows[4*row]
+          print >> outFile, outputRows[4*row+1]
+          print >> outFile, outputRows[4*row+2]
+          print >> outFile, outputRows[4*row+3]
+
 
 

### 2019-03-16 code/preparePaper/medianTables.py

@@ -39,16 +39,39 @@ with open("../../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv", "r
 languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
 
 
-
+entries = []
 
 for language in languages:
 #   line = languageKey[language]
-
-
    components = [language.replace("_"," ").replace("-Adap", "")]
    #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
-   components.append( "\\multirow{4}{*}{\includegraphics[width=0.1\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}" )
-   print("  &  ".join([str(x) for x in components]) + "  \\\\ " )
+   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}" )
+   entries.append(components)
+
+ROWS_PER_PART = 4
+COLUMNS = 4
+entries += [["",""] for x in range((COLUMNS-(len(entries)%COLUMNS))%COLUMNS)]
+
+outputRows = []
+if True:
+    languages = [x[0] for x in entries]
+    images = [x[1] for x in entries]
+
+    for i in range(len(entries)/COLUMNS):
+       outputRows.append( " & ".join(languages[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+       outputRows.append( " & ".join(images[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+
+for part in range(len(entries)/(COLUMNS*ROWS_PER_PART)+1):
+  with open("../../medians_"+str(part)+".tex", "w") as outFile:
+      for row in range(part*ROWS_PER_PART, (part+1)*ROWS_PER_PART):
+          if 4*row >= len(outputRows):
+              break
+          print >> outFile, outputRows[4*row]
+          print >> outFile, outputRows[4*row+1]
+          print >> outFile, outputRows[4*row+2]
+          print >> outFile, outputRows[4*row+3]
 
 
 

### 2019-03-16 code/preparePaper/quantiles.py

@@ -52,11 +52,39 @@ with open("../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL
 
 languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
 
+entries = []
+
 for language in languages:
 #   line = languageKey[language]
 
    components = [language.replace("_"," ").replace("-Adap", "")]
-   components.append( "\\multirow{4}{*}{\includegraphics[width=0.1\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}" )
-   print("  &  ".join([str(x) for x in components]) + "  \\\\ [10.25ex] \\hline" )
+   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}" )
+   entries.append(components) #   print >> outFile, ("  &  ".join([str(x) for x in components]) + "  \\\\ [10.25ex] \\hline" )
+
+ROWS_PER_PART = 5
+COLUMNS = 4
+entries += [["",""] for x in range((COLUMNS-(len(entries)%COLUMNS))%COLUMNS)]
+
+outputRows = []
+if True:
+    languages = [x[0] for x in entries]
+    images = [x[1] for x in entries]
+
+    for i in range(len(entries)/COLUMNS):
+       outputRows.append( " & ".join(languages[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+       outputRows.append( " & ".join(images[i*COLUMNS:(i+1)*COLUMNS]))
+       outputRows.append( " \\\\ ")
+
+for part in range(len(entries)/(COLUMNS*ROWS_PER_PART)+1):
+  with open("../../quantiles_"+str(part)+".tex", "w") as outFile:
+      for row in range(part*ROWS_PER_PART, (part+1)*ROWS_PER_PART):
+          if 4*row >= len(outputRows):
+              break
+          print >> outFile, outputRows[4*row]
+          print >> outFile, outputRows[4*row+1]
+          print >> outFile, outputRows[4*row+2]
+          print >> outFile, outputRows[4*row+3]
+
 
 

### 2019-03-16 code/preparePaper/samplesNumber.py

@@ -46,24 +46,32 @@ languageKey_corpusSizes = dict([(h(corpusSizes, line, "Language"), line) for lin
 
 
 
-
 with open("../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
    listener_curve = readTSV(inFile)
 
 languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
 
+output = []
+
 for language in languages:
-#   line = languageKey[language]
-   line_listener = languageKey_listener_curve[language]
-   line_stats = languageKey_stats[language]
+  #   line = languageKey[language]
+     line_listener = languageKey_listener_curve[language]
+     line_stats = languageKey_stats[language]
+  
+  
+     satisfied = h(stats, line_stats, "RANDOM_BY_TYPE") >= 10 and h(stats, line_stats, "REAL_REAL") >= 5 and (h(listener_curve, line_listener, "result1High") - h(listener_curve, line_listener, "result1Low") <= 0.15)
+  
+     components = [language.replace("_"," ").replace("-Adap", "")]
+     components.append( h(stats, line_stats, "RANDOM_BY_TYPE") )
+     components.append( h(stats, line_stats, "REAL_REAL") )
+  
+     output.append(components) # ("  &  ".join([str(x) for x in components]) + "  \\\\ " )
 
+if len(output)/2 * 2 < len(output):
+    output.append(["","",""])
 
-   satisfied = h(stats, line_stats, "RANDOM_BY_TYPE") >= 10 and h(stats, line_stats, "REAL_REAL") >= 5 and (h(listener_curve, line_listener, "result1High") - h(listener_curve, line_listener, "result1Low") <= 0.15)
-
-   components = [language.replace("_"," ").replace("-Adap", "")]
-   components.append( h(stats, line_stats, "RANDOM_BY_TYPE") )
-   components.append( h(stats, line_stats, "REAL_REAL") )
-
-   print("  &  ".join([str(x) for x in components]) + "  \\\\ [10.25ex] \\hline" )
-
+with open("../../samplesNumber.tex", "w") as outFile:
+ for i in range(len(output)/2):
+    here = output[i] + output[len(output)/2+i]
+    print >> outFile, ("  &  ".join([str(x) for x in here]) + "  \\\\") # [10.25ex] \\hline" )
 

### 2019-03-16 code/ud_languages.py

@@ -1,2 +1,2 @@
-with open("../ud_languages.txt", "r") as inFile:
+with open("../../ud_languages.txt", "r") as inFile:
    languages = inFile.read().strip().split("\n")

### 2019-03-16 neural/plot_wordsOnly_histByMem_AcrossLangs.R

@@ -41,3 +41,17 @@ fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", se
     ggsave(plot, file=paste("figures/full-GROUND-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf", sep=""))
 
 
+#library(ggrepel)
+    plot = ggplot(data, aes(x=-MI_z, fill=Type, color=Type)) 
+    plot = plot + theme_classic() 
+#    plot = plot + theme(legend.position="none")   
+    plot = plot + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled.., weight=1/counts))      
+    dataReal = data %>% filter(Type %in% c("REAL_REAL")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z)) %>% mutate(y=1)
+    plot = plot + geom_bar(data=dataReal,  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
+    plot = plot + geom_text(data=dataReal, aes(y=1, label=Language), angle=90, hjust=-0.1)
+    plot = plot + ylim(0, 1.5)
+    ggsave(plot, file=paste("figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab_labeled.pdf", sep=""))
+
+
+
+

### 2019-03-16 neural/plot_wordsOnly_medianDiff.R

@@ -70,6 +70,7 @@ plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Portuguese")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("English")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Italian")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Russian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Korean")
 
 
 

### 2019-03-16 neural/plot_wordsOnly_quantile.R

@@ -87,6 +87,7 @@ plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Portuguese")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("English")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Italian")
 plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Russian")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Korean")
 
 
 

### 2019-03-16 paper.tex
revision

--- 2019-03-14 paper.tex
@@
 
-We assume that $L_t$ contains no information about the process beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-	\begin{equation}\label{eq:listener-markov}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
+First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
+Formally: 
+\begin{equation}\label{eq:listener-markov-1}
+L_t \bot X_{>t} | X_{\leq t}
 \end{equation}
+This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
 
-%
-%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-%Formally: 
-%\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot X_{>t} | X_{\leq t}
-%\end{equation}
-%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-%
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-%\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-%\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
+Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
+\begin{equation}\label{eq:listener-markov-2}
+L_t \bot X_{<t} | X_{t-1}, L_{t-1}
+\end{equation}
+This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
 
@@
 %now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies
+
+\paragraph{Other Psycholinguistic Predictions}
+
+% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries.
+
+\paragraph{Speakers}
+
+% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
+This changes the interpretation of the mutual information.
+For the listener, it’s just redundancy.
+For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
+Then the speaker’s pressure is to keep conceptually related things close.
 
@@
 \paragraph{Listener Memory: Surprisal-Memory Tradeoff}
+
+\paragraph{Stopping Criterion}
+
+For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
+We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
+Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
+%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
+
+%- at least 30 samples from both baseline and real
+%
+%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
+
+%
+%(1) is bigram MI always greater in real languages?
+%
+%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
+
+\subsection{Statistics}
+
+We do these at each fixed memory value.
+
 We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
@@
 We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
+
+\paragraph{Confidence Interval for Medians}
+We create (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
+
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
+
+\paragraph{CI for Median Difference}
+We create (nonparametric and nonasymptotic) confidence interval for the difference median surprisal at each memory value.
+
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
+
+\paragraph{Significance Test}
+For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic) with the null hypothesis
+
+Let $W_-$ be the proportion of baseline languages that are less optimal than the real language throughout the entire range, and let $W_+$ be the proportion of baseline languages that are more optimal.
+
+We consider the null hypothesis that
+\begin{equation}
+	W_+ \geq W_-
+\end{equation}
+In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
+
+For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic) with the null hypothesis
+\begin{equation}
+	W_+(\mu) \geq W_-(\mu)
+\end{equation}
+using the empirical median for the real language.
+
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
+
+We take the REAL values to be estimated exactly by their medians.
+
 For each sample $x$ from real orderings, we look at the proportions $W_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $W_-(x)$ of baseline samples that are consistently less optimal.
@@
 
-\paragraph{Stopping Criterion}
+\paragraph{Confidence Interval for Quantile}
+CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
 
-For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
+We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
 
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
+We want to create a CI at each Memory value for the quantile.
 
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
+Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
+
+Let $\theta$ the best baseline sample that is worse than the real median.
+
+We want to get a confidence bound $q$ on $P(X < x_{real})$.
+
+Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
+Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
+
+We minimize $q$ subject to $p < 0.05$.
+
+Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
+
+This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
+
+Also the following does not assume unimodality, and ends up getting about the same intervals
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
+
+\paragraph{Confidence interval for quantile, for the entire range}
+
+we also apply bootstrapping
+
+We obtained a global measure as follows.
 
@@
 For any $t$,
-	\begin{equation}
-H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{equation}
+	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
 	This can be derived from the two conditions \ref{eq:listener-markov-1} and \ref{eq:listener-markov-2} (TODO explain how).
-
-	\begin{equation}
-H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
-	\end{equation}
-is true because the following is a Markov chain:
-\begin{equation}
-(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
-\end{equation}
-because
-\begin{align*}
-p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) & = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) \\
-& = \sum_{X_{<1}} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1}) \\
-&= \sum_{X_{<1}} p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, X_1, X_{2 \dots t-1}) \\
-&= p(X_t|L_0,  X_1, X_{2 \dots t-1}) 
-\end{align*}
-because
-	\begin{align*}
-p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
-p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
-	\end{align*}
-because of Equation~\ref{eq:listener-markov}.
-	\begin{equation}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-
-	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
 
@@
 \begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( H[X_t|X_{1\dots t-1}, L_0] - H[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
+	H[X_t | L_t] - H[X_t | X_{<t}]& \geq   \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
 	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 

### 2019-03-16 medianDiff_3.tex

Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} &  & 
 \\

### 2019-03-16 medians_1.tex

English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-16 medians_2.tex

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-16 medians_3.tex

Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} &  & 
 \\

### 2019-03-16 paper.tex

--- 2019-03-16 paper.tex
@@
 
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
+\paragraph{Confidence Interval for Quantile}
+CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
+
+We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
+
+We want to create a CI at each Memory value for the quantile.
+
+Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
+
+Let $\theta$ the best baseline sample that is worse than the real median.
+
+We want to get a confidence bound $q$ on $P(X < x_{real})$.
+
+Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
+Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
+
+We minimize $q$ subject to $p < 0.05$.
+
+Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
+
+This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
+
+Also the following does not assume unimodality, and ends up getting about the same intervals
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
+
+\paragraph{Confidence interval for quantile, for the entire range}
+
+we also apply bootstrapping
+
+We obtained a global measure as follows.
 
@@
 For any $t$,
+	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
 	\begin{equation}

### 2019-03-16 quantiles_2.tex

Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} &  & 
 \\

### 2019-03-16 results-table-word-level.tex

--- 2019-03-01 results-table-word-level.tex
@@
-Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Arabic  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Arabic-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Armenian-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.91  &  [0.83, 0.98]  \\ [10.25ex] \hline
-Bambara-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bambara-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.96  &  [0.86, 1.0]  \\ [10.25ex] \hline
-Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-English  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/English-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Erzya-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Erzya-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.99  &  [0.98, 1.0]  \\ [10.25ex] \hline
-Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.83  &  [0.76, 0.9]  \\ [10.25ex] \hline
-Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-French  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/French-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.89  &  [0.81, 0.96]  \\ [10.25ex] \hline
-Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Kurmanji-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.92  &  [0.83, 0.99]  \\ [10.25ex] \hline
-Latvian*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.48  &  [0.4, 0.57]  \\ [10.25ex] \hline
-Maltese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Maltese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.37  &  [0.3, 0.44]  \\ [10.25ex] \hline
-Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.07  &  [0.01, 0.15]  \\ [10.25ex] \hline
-Portuguese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Portuguese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.07  &  [0.03, 0.12]  \\ [10.25ex] \hline
-Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.82  &  [0.74, 0.89]  \\ [10.25ex] \hline
-Spanish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Spanish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Urdu  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Urdu-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
-Uyghur-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  0.65  &  [0.57, 0.74]  \\ [10.25ex] \hline
-Vietnamese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}}  &  $D_x$  \\
-  &    &    &  $W_x$  &  1.0  &  [0.98, 1.0]  \\ [10.25ex] \hline
+Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1315  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+974  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Arabic  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+21864  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Armenian-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+514  &    &    &    &    &    &  0.92  &  [0.87, 0.97]  \\ [10.25ex] \hline
+Bambara-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+926  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Basque  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+5396  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+788  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+8907  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+808  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+550  &    &    &    &    &    &  0.96  &  [0.86, 1.0]  \\ [10.25ex] \hline
+Catalan  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+13123  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Chinese  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+3997  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Croatian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+7689  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Czech  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+102993  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Danish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4383  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Dutch  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+18310  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+English  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+17062  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Erzya-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1450  &    &    &    &    &    &  0.99  &  [0.98, 1.0]  \\ [10.25ex] \hline
+Estonian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+6959  &    &    &    &    &    &  0.8  &  [0.72, 0.86]  \\ [10.25ex] \hline
+Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1108  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Finnish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+27198  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+French  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+32347  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+German  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+13814  &    &    &    &    &    &  1.0  &  [0.91, 1.0]  \\ [10.25ex] \hline
+Greek  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1662  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+5241  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hindi  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+13304  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+910  &    &    &    &    &    &  0.87  &  [0.8, 0.93]  \\ [10.25ex] \hline
+Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4477  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Italian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+17427  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Japanese  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+7164  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+947  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Korean  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+27410  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Kurmanji-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+634  &    &    &    &    &    &  0.93  &  [0.88, 0.98]  \\ [10.25ex] \hline
+Latvian*  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4124  &    &    &    &    &    &  0.49  &  [0.4, 0.57]  \\ [10.25ex] \hline
+Maltese  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1123  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+848  &    &    &    &    &    &  1.0  &  [0.99, 1.0]  \\ [10.25ex] \hline
+North Sami  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+2257  &    &    &    &    &    &  0.37  &  [0.3, 0.44]  \\ [10.25ex] \hline
+Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+29870  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Persian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4798  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Polish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+6100  &    &    &    &    &    &  0.1  &  [0.04, 0.17]  \\ [10.25ex] \hline
+Portuguese  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+17995  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Romanian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+8664  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Russian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+52664  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Serbian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+2935  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Slovak  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+8483  &    &    &    &    &    &  0.07  &  [0.03, 0.12]  \\ [10.25ex] \hline
+Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+7532  &    &    &    &    &    &  0.82  &  [0.77, 0.88]  \\ [10.25ex] \hline
+Spanish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+28492  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Swedish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+7041  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+900  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Turkish  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+3685  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4506  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Urdu  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+4043  &    &    &    &    &    &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
+Uyghur-Adap*  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1656  &    &    &    &    &    &  0.65  &  [0.57, 0.73]  \\ [10.25ex] \hline
+Vietnamese  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-by-run_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}}  &    \\
+1400  &    &    &    &    &    &  1.0  &  [0.98, 1.0]  \\ [10.25ex] \hline

### 2019-03-16 paper.tex
pictures

@@
 
-\subsection{Memory and Locality in General Sequence Models}
-
-We consider a \emph{speaker} who produces sentences and a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
-Informally, we expect that both speaker and listener will be affected by memory demands:
-From the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
-From the listener's perspective, predicting the next word well requires maintaining information about the past.
+We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
+%Informally, we expect that both speaker and listener will be affected by memory demands:
+%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
+For the listener, predicting the next word well requires maintaining information about the past.
 For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
@@
 
-We now make this more formal:
-We think of the language as a stochastic process $(X_t)_{t \in \mathbb{Z}}$, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
-As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
-This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
-Second, we make the assumption that this process is \emph{stationary}.
-Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
-Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
-In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
-
-The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
-We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
-The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
-\begin{equation}\label{eq:markov-speaker}
-X_{>t} \bot X_{\leq t} | S_t
-\end{equation}
-This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
+TODO say at some point that we're studying sentence-internal memory
+
+\subsection{Theoretical Results}
+
+We now introduce our main theoretical result on memory-surprisal tradeoffs.
+
+We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
+The symbols $x_i$ belong to a common set, representing the vocabulary of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+
+Our main tool is the conditional probability distribution $p(x_t|x_{<t})$.
+
+% $(X_t)_{t \in \mathbb{Z}}$
+%We think of the language as a stochastic process, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
+%As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
+%This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
+%We make the assumption that this process is \emph{stationary}.
+%Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
+%Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
+%In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
+%Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
+
+%The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
+%We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
+%The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
+%\begin{equation}\label{eq:markov-speaker}
+%X_{>t} \bot X_{\leq t} | S_t
+%\end{equation}
+%This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
 
 We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-The listener maintains a memory state $L_t$; there are no assumptions about the memory architecture and the nature of its computations.
-%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-We only make two basic assumptions about the flow of information:
-
-We assume that $L_t$ contains no information about the process beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
+As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
+
+There are no assumptions about the memory architecture and the nature of its computations.
+We only make a basic assumption about the flow of information:
+At a given point in time, the listener's memory state $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
+This is formalized as a statement about conditional independence:
 	\begin{equation}\label{eq:listener-markov}
@@
 	\end{equation}
+This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.
 
@@
 What can we say about the listener's surprisal?
+
+We will use the concept of \emph{conditional mutual information}.
+For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
+\begin{equation}
+	I[X,Y] := H[X] - H[X|Y]
+\end{equation}
+For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
+\begin{equation}
+I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
+\end{equation}
+We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
+\begin{equation}
+	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
+\end{equation}
+This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
+That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
+This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
+
 We return to the information and memory curves in Figure~\ref{fig:basic}.
@@
 \end{proposition}
+The proof is given in the appendix (REF).
 
@@
 
-In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
-The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
-For reaching optimal surprisal, Order A requires strictly less memory.
-Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
-
 \begin{figure}
@@
 
-We will use the concept of \emph{conditional mutual information}.
-For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
-\begin{equation}
-	I[X,Y] := H[X] - H[X|Y]
-\end{equation}
-For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-\begin{equation}
-I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-\end{equation}
-We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-\end{equation}
-This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
 %that the model does not have an `absolute clock' built into it.
-The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
 
@@
 
+In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
+In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
+The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
+For reaching optimal surprisal, Order A requires strictly less memory.
+Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
+
 \begin{figure*}
@@
 
-\paragraph{Dependency Data}
+We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
+
+We want to compare to counterfactual reorderings.
+We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
+
+\paragraph{Data}
 We draw on dependency corpora.
 
+TODO talk about dependency grammar
+
 Universal Dependencies has compiled dependency corpora for several dozen languages
 
-\paragraph{Ordering Grammrs}
+We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
+We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
+This resulted in 52 languages.
+
+\paragraph{Counterfactual Ordering Grammars}
 We define ordering grammars, small interpretable models of dependency tree linearization.
@@
 \subsection{Method}
-%The small size of available syntactic data for many languages means that estimating mutual information is problematic.
-%
-%
-%In general, we expect a bias-variance tradeoff:
-%
-%A method that estimates on a coarse-grained level (e.g.
-POS) will result in estimates with relatively higher fidelity and lower variance, potentially making it easier to localize the actual language in the space of counterfactual languages.
-%However, it introduces \emph{bias} by describing a simplified version of the language.
-%
-%
-%
-%
-%
-%We propose to therefore compute the quantities on levels less likely to suffer from data sparseness: part-of-speech and morphology.
-%
-%In the UD treebanks, each word has part-of-speech and morphological information.
-%The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
-%We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token that indicates that information related to the word ends.
-%We further provide the word's lemma after its POS tag, if it is one of the 50 most frequent lemmas of the language.
-%Each POS tag and morphological feature-value pair is encoded as an atomic token.
-%The rationale for including frequent lemmas is that such words often are strongly grammaticalized and may provide grammatical information (and frequent words are less likely to be affected by data sparseness).
-%
-
 To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling sequences such as natural language.
@@
 \paragraph{Data}
-
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 52 languages.
 
@@
 
-We do these at each fixed memory value.
+We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
+
+For each language, we viewed surprisal as a function of memory.
+We used linear interpolation to obtain surprisal values for all levels of memory.
 
@@
 
+We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
+We describe the statistics computed in the following paragraphs.
+
+We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
+Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
+
 \paragraph{Confidence Interval for Medians}
@@
 \paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference median surprisal at each memory value.
+We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
 
@@
 
-\paragraph{Significance Test}
-For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic) with the null hypothesis
-
-Let $W_-$ be the proportion of baseline languages that are less optimal than the real language throughout the entire range, and let $W_+$ be the proportion of baseline languages that are more optimal.
-
-We consider the null hypothesis that
-\begin{equation}
-	W_+ \geq W_-
+\paragraph{Pointwise Significance Test}
+For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
+Let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
+We take the real language to be represented by the \emph{population median} of the samples.
+For each level $\mu$ of memory, we consider the null hypothesis that
+\begin{equation}
+	W_-(\mu) \leq 0.5
 \end{equation}
@@
 
-For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic) with the null hypothesis
-\begin{equation}
-	W_+(\mu) \geq W_-(\mu)
-\end{equation}
-using the empirical median for the real language.
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.45\textwidth]{figures/nhst.png}
+\end{center}
+	\caption{Illustration for the pointwise null-hypothesis significance test.
+At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
+\end{figure}
+
+For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
+%\begin{equation}
+%	W_+(\mu) \geq W_-(\mu)
+%\end{equation}
+We use the empirical median for the real language.
 
@@
 We take the REAL values to be estimated exactly by their medians.
+
+\paragraph{Pointwise Quantile Estimate}
+%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
+
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.45\textwidth]{figures/quantile.png}
+\end{center}
+	\caption{Illustration for the quantile estimate.
+At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
+\end{figure}
+
+For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
+This is described in Figure~\ref{fig:quantile-pointwise}.
+
+We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
+
+We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
+
+We want to create a CI at each Memory value for the quantile.
+
+Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
+
+Let $\theta$ the best baseline sample that is worse than the real median.
+
+We want to get a confidence bound $q$ on $P(X < x_{real})$.
+
+Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
+Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
+
+We minimize $q$ subject to $p < 0.05$.
+
+Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
+
+This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
+
+Also the following does not assume unimodality, and ends up getting about the same intervals
+% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
+
+\paragraph{Global Quantile Estimate}
+
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
+\end{center}
+	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
+\end{figure}
 
@@
 %
+
+we also apply bootstrapping
+
+We obtained a global measure as follows.
 
@@
 %We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-\paragraph{Confidence Interval for Quantile}
-CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Confidence interval for quantile, for the entire range}
-
-we also apply bootstrapping
-
-We obtained a global measure as follows.
 
@@
 Thus, random grammars with by-relation weights constitute a stronger baseline.
+
+\section{Discussion}
+
+\subsection{Speaker Memory}
+
+The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
+
+\section{Conclusion}
 

### 2019-03-16 paper.tex
revising

@@
 
-For each sample $x$ from real orderings, we look at the proportions $W_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $W_-(x)$ of baseline samples that are consistently less optimal.
-
-We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-\begin{equation}
-	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
+For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
+
+%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
+%\begin{equation}
+%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
+%\end{equation}
+
+We estimate the quotient
+\begin{equation}
+	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
 \end{equation}
 where $P_1$ is the distribution over values obtained for real orderings.
-
-We estimate the quotient
-\begin{equation}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-We use a bootstrapped confidence interval for $\E[G]$ for testing and for quantifying the degree of optimization.
+We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
+For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
+
+Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
+Due to the use of bootstrapping, the confidence intervals are not exact.
 
@@
 %
-
-we also apply bootstrapping
-
-We obtained a global measure as follows.
 
@@
 
-In Figrue~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves.
+In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves.
 
@@
 - show the total histogram
+
+\begin{figure}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\caption{Histogram}\label{fig:hist-real}
+\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
 \subsection{Discussion}
@@
 
-\section{Fixed Word Orders}
+\subsection{Experiment 2b: Fixed Word Orders}
 
 We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of more specific ordering choices.
+This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
 
@@
 \paragraph{Fitting Ordering Grammars to Actual Orders}
-We additionally create ordering grammars that are fit to the actual orderings of each language.
+We create ordering grammars that are fit to the actual orderings of each language.
 These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
 
-We construct these grammars by interpreting ordering grammars probabilistically, and setting the parameters
-
-We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
-Thus, random grammars with by-relation weights constitute a stronger baseline.
+We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
+We parameterized probabilistic ordering grammars as follows.
+For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
+Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
+Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
+184) without replacement.
+
+Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
+We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
+
+We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
+Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
+Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
+
+Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
+
+We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
+Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
+However, discrete optimization is computationally challenging.
+
+\paragraph{Results}
 

### 2019-03-17 paper.tex
revising

@@
 
+\paragraph{Setting}
 We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the vocabulary of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-
-Our main tool is the conditional probability distribution $p(x_t|x_{<t})$.
+The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+
+We model the sequence as a probabilistic sequence; that is, given a context $x_{<t}$, the next word is distributed according to a distribution $p(x_t|x_{<t})$.
 
@@
 
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
+	\caption{Illustration of (\ref{eq:listener-markov}).
+As the utterance unfolds, the listener maintains a memory state.
+After receiving word $X_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $X_t$.}\label{fig:listener-markov}
+\end{figure}
+
 There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information:
-At a given point in time, the listener's memory state $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
-This is formalized as a statement about conditional independence:
+We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
+At a given point in time, the listener's memory state $L_t$ is determined by the last word $X_t$, and the prior memory state $L_{t-1}$.
+As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
+This is formalized as a statement about conditional probabilities:
 	\begin{equation}\label{eq:listener-markov}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
+p(L_1| (X_{t})_t, L_0)   = p((X_{t})_t| L_0, X_1)
+%p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
 	\end{equation}
 This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.	
+As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
+For the study of memory in sentence processing, this seems fair. Discuss this more.
 
@@
 
-The listener can trade off memory and future surprisal:
-A listener who chooses to store less memory will exerience higher surprisal in the future.
-A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-We will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-Consider a listener who uses $J$ bits of memory on average.
-What can we say about the listener's surprisal?
-
-We will use the concept of \emph{conditional mutual information}.
-For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
+%The listener can trade off memory and future surprisal:
+%A listener who chooses to store less memory will exerience higher surprisal in the future.
+%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
+
+%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
+%Consider a listener who uses $J$ bits of memory on average.
+%What can we say about the listener's surprisal?
+
+\paragraph{Conditional Mutual Information}
+
+We will use the concept of \emph{conditional mutual information} (Figure~\ref{fig:it-surprisal-intuition}).
+%For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
+%\begin{equation}
+%	I[X,Y] := H[X] - H[X|Y]
+%\end{equation}
+%For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
+%\begin{equation}
+%I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
+%\end{equation}
+%We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
 \begin{equation}
-	I[X,Y] := H[X] - H[X|Y]
-\end{equation}
-For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-\begin{equation}
-I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-\end{equation}
-We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
+%	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
+	I_t := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
 \end{equation}
@@
 This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
+
+\begin{figure*}
+\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
+	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}\label{fig:it-surprisal-intuition}
+\end{figure*}
 
@@
 \begin{proposition}\label{prop:suboptimal}
-	Let $T$ be a positive integer, and consider a listener using at most
-$$\sum_{t=1}^T t I_t$$
+	Let $T$ be any positive integer ($1, 2, 3, ...$), and consider a listener using at most
+	\begin{equation}\label{eq:memory}
+		\sum_{t=1}^T t I_t
+	\end{equation}
 bits of memory on average.
@@
 The proof is given in the appendix (REF).
+
+%An intuitive (non-rigorous) argument goes as follows:
+%For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
+
+\begin{figure*}
+\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
+\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
+%
+	\caption{Left: $I_t$ as a function of $t$, for two different processes.
+$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
+Left: $t \cdot I_t$ as a function of $t$ for the same processes.
+}\label{fig:basic}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
-	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}
-\end{figure*}
-
 %that the model does not have an `absolute clock' built into it.
+
+Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
+% and the mechanisms by which it computes predictions.
+In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
+Memory representations do not have to be rational or optimal for this bound to hold:
+It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
+
+\paragraph{Information Locality}
 
@@
 
-The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
-% and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
+%We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
 The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
@@
 
-We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-The analysis is not even restricted to discrete timesteps: We can replace the sum in~(\ref{eq:memory-bound}) with an integral to get a continuous-time version.
+%\paragraph{Discussion}
+%We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
+%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
 
@@
 
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
-%
-	\caption{Left: $I_t$ as a function of $t$, for two different processes.
-$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
-Left: $t \cdot I_t$ as a function of $t$ for the same processes.
-The area under the curves is given by (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-}\label{fig:basic}
-\end{figure*}
-
-\subsection{Memory and Dependency Length}
+\section{Experiment 1: Memory and Dependency Length}
 
@@
 			As $I_t$ decays faster in A orders, the total area under the curve now differs between A and B, and is larger in B.
-			This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.
-			While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
-That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.
-It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.
+			%This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.
+
+%While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
+%That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.
 
 In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
+%In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
 The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
@@
 Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
+
+It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.
 
@@
 
-\section{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-The quantity in (\ref{eq:memory-bound}) is equal to the \emph{excess entropy}, and it has long been known to bound statistical and generative complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and locality provided by our series expansion in (\ref{eq:memory-bound}) appears to be a novel contribution.
-
-\cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
-%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-Our results apply both to speakers and listeners.
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-In fact, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory (and thus in particular finite excess entropy):
-
-\begin{proposition}
-Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
-\end{proposition}
-
-\begin{proof}[Proof Sketch]
-	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
-
-	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
-We consider incremental Earley parsing \citep{earley-efficient-1970}.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
-For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
-	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
-\end{proof}
-
 \section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
@@
 We will define a class of baseline languages that serve as a comparison class of word orders that natural language can be compared to.
-
 We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
@@
 We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
+Therefore, we compare to counterfactual word order gramars.
 
@@
 We draw on dependency corpora.
-
-TODO talk about dependency grammar
-
-Universal Dependencies has compiled dependency corpora for several dozen languages
-
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
+The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
+
+In dependency corpora, sentences are annotated with \emph{dependency trees}. TODO a figure.
+These are directed trees describing the grammatical relations among words.
+For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
+the subject or an indirect object.
+A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
+Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
+Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
+
+We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 500 sentences of training data.
 We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
@@
 
+For each of these languages, we pooled all available corpora in one dataset.
+Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
+While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
+For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
+For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
+In these cases, we redefined the split to obtain more training data:
+For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
+For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
+We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
+
 \paragraph{Counterfactual Ordering Grammars}
-We define ordering grammars, small interpretable models of dependency tree linearization.
+We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
 Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
 
-Universal dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
+Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
 These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
@@
 
-\subsection{Method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling sequences such as natural language.
+\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
+To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE) and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
 We model language on the level of individual word forms.
@@
 
-\paragraph{Data}
-
-UD models predefined with train, dev, and test splits.
-Small datasets sometimes not.
-
-For most language, we used the predefined data split.
-For some languages, where there was no training partition, or where the training partition was smaller than the test partition, we redefined the split to obtain more training data:
-For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
+%\paragraph{Data}
 
 \paragraph{Model}
-We use a recurrent neural network with Long-Short-Term Memory cells.
+We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997}.
 %Given a sequence of input words $w_1, ..., w_n \in V$, the model 
 %
+\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
 The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
@@
 
-\subsection{Discussion: Alternative Models}
-Recent research has proven strong theoretical learnability guarantees for neural network models (CITE), supporting their success in practice.
-They also provide strong predictors of the surprisal effect on human reading times \citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-
-In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-
-A traditional model uses n-gram models.
+%
+%\subsection{Discussion: Alternative Models}
+%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
+%
+%A traditional model uses n-gram models.
 A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
@@
 Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-
-A second option is to construct a statistical grammar, such as PCFG.
-The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
+%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
+%
+%A second option is to construct a statistical grammar, such as PCFG.
+%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
+%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
+%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
+%
 
@@
 We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-We collected data from the actual and random orderings in proportion one to three.
+We collected data from the actual and random orderings in proportion one to two.
 The stopping criterion will be described below.
@@
 
-For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
 %We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
@@
 
-\paragraph{Number of Samples}
+\subsection{Number of Samples}
 Training neural language models is computationally costly.
@@
 %We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
+
+%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
+%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
+%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
 
@@
 
-The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
+%The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
+TODO
+
+\subsection{Relation to Theories of Sentence Processing}
+
+\paragraph{Lossy-Context Surprisal}
+\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
+In particular, they consider loss models that delete, erase, or replace words in the past.
+Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
+A listener will incur surprisal
+$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
+where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
+Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
+Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
+
+\paragraph{Relation to Dependency Locality}
+The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
+Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
+In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
+In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
+
+also surprisal -- integration cost
+
+\paragraph{Center Embeddings, ...}
+Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
+
+Miller and Chomsky 1963
+
+Frzier 1985 local nonterminal count
+
+Rambow and Joshi 1994 using TAG
+
+Marcus 1980 deterministic parsing
+
+(Sabrina Gerth, Memory Limitations in Sentence Processing)
+
+\subsection{Relation to other work}
+
+\paragraph{Statistical Complexity}
+There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
+%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
+The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
+In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
+In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
+However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
+Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
+
+also information-theoretic studies of memory capacity
+
+\paragraph{Decay of Mutual Information}
+In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
+Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
+
+The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
+While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
+Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
+%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
+The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
+Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
+This one has bounded nonzero $J$, but zero unconditional MI.
+To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
+The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
+
+\paragraph{Long-range dependencies in text}    % excess entropy
+\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
+Our work contrasts with this work in that we are interested in dependencies within sentences.
+
+\subsection{Discussion}
+
+\paragraph{Speakers vs Listeners}
+Our results apply both to speakers and listeners.
+
+\paragraph{Decay vs Interference}
+Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
+The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
+The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
+This is not the case:
+A long dependency can be maintained perfectly with low average memory:
+Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
+In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
+
+\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
+Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
+However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
+For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
+The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
+This implies that average memory demands are finite.}
+
+%\begin{proposition}
+%Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
+%\end{proposition}
+%
+%\begin{proof}[Proof Sketch]
+%	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
+%
+%	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
+We consider incremental Earley parsing \citep{earley-efficient-1970}.
+Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
+For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
+%	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
+%\end{proof}
+%
+%
+%
 

### 2019-03-17 paper.tex
revising

@@
 We estimate the quotient
-\begin{equation}
+\begin{equation}\label{eq:g}
 	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
@@
 
-In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves.
-
-In Figure~\ref{tab:medians}, we show the medians.
+In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
+
+In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
 
@@
 
-- show the slice histograms
-
-- show the total histogram
+In Figure~\ref{tab:slices}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
+
+In Figure~\ref{tab:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
+
+In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
+
+In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
+$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
+
+In Table~\ref{tab:quantiles}, we show the quantiles.
 
@@
 \end{figure}
-
-\paragraph{Inferential Statistics}
-
-inferential
-
-- report the bootstrapped thing G
-
-Median Differences \ref{tab:median_diffs}
-
-Quantiles \ref{tab:quantiles}
 
@@
 \end{table}
-
-In all but X languages $\E[G] > 0.5$ significantly.
-
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-%\begin{longtable}{llllllllllllllllll}
-	%\input{results-table-word-level.tex}
-%\end{longtable}
 

### 2019-03-17 paper.tex
font size

@@
 	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample means, dashed lines indicate 95 \% confidence intervals.}\label{tab:medians}
+Solid lines indicate sample medians, dashed lines indicate 95 \% confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
 \end{table}
@@
 \end{longtable}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines.
+	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
 Lower values indicate lower surprisal compared to baselines.

### 2019-03-17 neural/plot_wordsOnly_medianDiff.R

@@ -11,6 +11,8 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
  #   UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     plot = ggplot(data, aes(x=Memory, y=-EmpiricalMedianDiff, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=-MedianDiff_Lower), linetype="dashed") + geom_line(aes(x=Memory, y=-MedianDiff_Upper), linetype="dashed")
+    plot = plot + xlab("Memory") + ylab("Difference in Median Surprisal")
+    plot = plot + theme(text = element_text(size=30))
     ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf", sep=""))
     return(plot)
 }

### 2019-03-17 neural/plot_wordsOnly_medians.R

@@ -11,6 +11,8 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
+    plot = plot + xlab("Memory") + ylab("Median Surprisal")
+    plot = plot + theme(text = element_text(size=30))
     ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf", sep=""))
     return(plot)
 }

### 2019-03-17 neural/plot_wordsOnly_quantile.R

@@ -27,7 +27,8 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     plot = plot + ylim(0,1.1)
     plot = plot + ylab("Quantile")
     plot = plot + xlab("Memory")
-    # + geom_line(aes(x=Memory, y=-MedianDiff_Upper))
+    plot = plot + theme(text = element_text(size=20))
+   # + geom_line(aes(x=Memory, y=-MedianDiff_Upper))
     ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf", sep=""), height=3.5, width=4.5)
     return(plot)
 }

### 2019-03-19 paper.tex
ngram results

@@
 	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 \% confidence intervals for the population median.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
@@
 Solid lines indicate sample means.
-Dashed lines indicate 95 \% confidence intervals.}\label{tab:median_diffs}
+Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
 \end{table}
@@
 \section{N-Gram Models}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_ngrams_0.tex}
+\end{longtable}
+	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_ngrams_1.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_ngrams_2.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_ngrams_3.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
 

### 2019-03-19 ngrams/analyzeNgramsAll.py

@@ -1,65 +1,10 @@
 import os
 
-
-finalShib = {"Estonian" : "Estonian 0.16 200 256 1 0.0055",
-             "Vietnamese" : "vi 0.33 50 128 1 0.02",
-             "Slovenian" : "Slovenian 0.16 100 256 1 0.01",
-             "Finnish" : "Finnish 0.16 100 512 1 0.01",
-             "Polish" : "Polish 0.33 150 512 3 0.05",
-             "Czech" : "Czech 0.2 250 512 2 0.05",
-             "Swedish" : "Swedish 0.33 150 256 1 0.01",
-             "Hindi" : "Hindi 0.25 300 256 3 0.01",
-             "Basque" : "",
-             "German" : "German 0.2 150 512 2 0.01",
-             "Arabic" : "Arabic 0.2 300 1024 3 0.05",
-             "Turkish" : "Turkish 0.3 250 512 1 0.05",
-             "Spanish" : "Spanish 0.25 200 1024 3 0.01",
-             "Latvian" : "Latvian 0.35 100 128 2 0.005",
-             "Indonesian" : "Indonesian 0.45 50 256 2 0.005",
-             "Hungarian" : "Hungarian 0.35 100 128 1 0.005",
-             "Persian" : "Persian 0.4 300 256 2 0.005",
-             "Croatian" : "Croatian 0.2 50 128 3 0.005",
-             "French" : "French 0.4 100 1024 2 0.005",
-             "Chinese" : "Chinese 0.3 300 128 3 0.005",
-             "Serbian" : "Serbian 0.45 200 256 1 0.01",
-             "Slovak" :  "Slovak 0.2 150 128 1 0.005",
-             "North_Sami" : " North_Sami 0.25 50 64 1 0.01 ",
-             "Irish" : " Irish 0.1 200 64 2 0.001 ",
-             "Lithuanian" : "Lithuanian 0.15 200 64 1 0.001 ",
-             "Armenian" : "Armenian 0.3 150 64 1 0.05 ",
-             "Ukrainian" : "Ukrainian 0.25 100 128 1 0.005 ",
-             "Armenian-Adap" : "Armenian-Adap 0.45 200 64 2 0.005 ",
-             "Greek" : " Greek 0.35 300 512 3 0.01 ",
-             "Uyghur-Adap" : "Uyghur-Adap 0.4 50 128 1 0.01 ",
[…]

### 2019-03-19 ngrams/plotResults.R

@@ -1,7 +1,7 @@
 
 
 memSurpNgramsPlot = function(language) {
-    data = read.csv(paste(language,"_ngrams_after_tuning.tsv", sep=""), sep="\t")
+    data = read.csv(paste("../../../",language,"_ngrams_after_tuning.tsv", sep=""), sep="\t")
     library(tidyr)
     library(dplyr)
     library(ggplot2)
@@ -75,7 +75,7 @@ plot = memSurpNgramsPlot("Maltese")
 
 
 memListenerSurpNgramsPlot = function(language) {
-    data = read.csv(paste(language,"_ngrams_decay_after_tuning.tsv", sep=""), sep="\t")
+    data = read.csv(paste("../../../",language,"_ngrams_decay_after_tuning.tsv", sep=""), sep="\t")
     library(tidyr)
     library(dplyr)
     library(ggplot2)

### 2019-03-19 ngrams/yAnalyzeNgrams.py

@@ -8,6 +8,7 @@ import sys
 sortBy = int(sys.argv[1])
 #horizon = int(sys.argv[2])
 language = sys.argv[2]
+assert len(sys.argv) == 3
 #restrictToFinished = (sys.argv[4] == "True")
 #onlyOptimized = (not (len(sys.argv) > 5 and sys.argv[5] == "False"))
 def f(a):
@@ -97,6 +98,9 @@ finalShib = {"Estonian" : "Estonian 0.16 200 256 1 0.0055",
 
 
 filenames = [x for x in os.listdir("/u/scr/mhahn/deps/memory-need-ngrams/") if x.startswith("search-"+language+"_yWithMo") and len(open("/u/scr/mhahn/deps/memory-need-ngrams/"+x, "r").read().split("\n"))>=30]
+#print(filenames)
+#print([x for x in os.listdir("/u/scr/mhahn/deps/memory-need-ngrams/") if x.startswith("search-"+language+"_yWithMo")])
+#print("/u/scr/mhahn/deps/memory-need-ngrams/search-"+language+"yWithMo")
 if len(filenames) == 0:
    quit()
 with open("/u/scr/mhahn/deps/memory-need-ngrams/"+filenames[0], "r") as inFile:
@@ -117,6 +121,7 @@ correctParameters = " ".join(params2)
 types = [" REAL_REAL ", "RANDOM_MODEL ", "RANDOM_BY_TYPE "]
 
 for fileName in files:
+  #print(fileName)
   if language not in fileName:
        continue
   if not fileName.startswith("estimates"):
@@ -124,13 +129,16 @@ for fileName in files:
 
   with open(path+fileName, "r") as inFile:
      result = inFile.read().split("\n")
-     if (correctParameters not in result[0]):
+     if (correctParameters+" " not in result[0]+" "):
          continue
      typeOfResult = filter(lambda x:x in result[0], types)[0][:-1]
      if len(result) < 3:
          continue
      if typeOfResult not in resultsPerType:
         resultsPerType[typeOfResult] = []
+     if "idForProcess" in result[0]:
+        continue
+#     print(result)
      result[1] = list(map(lambda x:x if x=="" else float(x), result[1].replace("[","").replace("]","").split(" ")))
      balanced, memory, residual, mi, decay, unigramCE  = f(result[2])
[…]

### 2019-03-19 medians_REAL_0.tex
additional files

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-03-19 medians_REAL_1.tex
additional files

English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-03-19 medians_REAL_2.tex
additional files

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-03-19 medians_REAL_3.tex
additional files

Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} &  & 
 \\

### 2019-03-19 medians_ngrams_0.tex
additional files

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Czech & Danish & Dutch & Estonian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-19 medians_ngrams_1.tex
additional files

Faroese & Finnish & French & German
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Indonesian & Japanese & Kazakh & Kurmanji
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Latvian & Naija & North Sami & Norwegian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-03-19 quantiles_REAL_2.tex
additional files

Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} &  & 
 \\

### 2019-03-23 paper.tex
paper

@@
 \newtheorem{example}[theorem]{Example}
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a speaker producing sentences from a language, and a listener performing incremental prediction.
+We will consider a general setting of a listener performing incremental prediction.
 Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-Unlike many previous theories, our results apply to both speakers and listeners.
-We will then use corpus data from over 50 languages to provide evidence that their word orders help lower memory cost.
+We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
 
@@
 
-\begin{proposition}\label{prop:suboptimal}
-	Let $T$ be any positive integer ($1, 2, 3, ...$), and consider a listener using at most
+\begin{thm}\label{prop:suboptimal}
+	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
 	\begin{equation}\label{eq:memory}
@@
 	on average.
-\end{proposition}
+\end{thm}
 The proof is given in the appendix (REF).
@@
 \begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_0.tex}
+\end{longtable}
+	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_1.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_2.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_3.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_REAL_0.tex}
+\end{longtable}
+	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_REAL_1.tex}
+\end{longtable}
+	\caption{Quantiles (part 2)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{cccccccccccccccccc}
+\input{quantiles_REAL_2.tex}
+\end{longtable}
+	\caption{Quantiles (part 3)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
 \input{medians_0.tex}
@@
 
-\section{Proofs}
-
-\begin{proof}
-We denote the listener's memory state after hearing $X_{<t}$ by $L_t$.
-The average number of bits required to encode this state is $H[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $H[X_t | L_t]$.
-The difference between the listener's surprisal and optimal surprisal is thus at least $H[X_t | L_t] - H[X_t | X_{<t}]$.
-By stationarity, we can rewrite this expression as
+\section{Proof of the Theorem}
+
+%\begin{proof}
+	We denote the listener's memory state at time $t$, after hearing $X_{<t} = ... X_{t-2} X_{t-1}$ by $L_t$.
+	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
+	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[X_t | L_t]$.
+	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]$.
+By the assumption of stationarity, we can rewrite this expression as
 \begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | L_t] - H[X_t | X_{<t}]\right) 
+	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]\right) 
 \end{align*}
-For any $t$,
-	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
+	\begin{lemma}
+For any $t$, the following inequality holds:
+%	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
 	\begin{equation}
-H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{equation}
-	This can be derived from the two conditions \ref{eq:listener-markov-1} and \ref{eq:listener-markov-2} (TODO explain how).
-
+%H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
+H[X_t | L_t] \geq H[X_t|X_{1 \dots t-1}, L_0]
+		\end{equation}
+	\end{lemma}
+	\begin{proof}[Proof of the Lemma]
+
+By Equation~\ref{eq:listener-markov}:
 	\begin{equation}
-H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
-	\end{equation}
-is true because the following is a Markov chain:
-\begin{equation}
-(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
-\end{equation}
-because
+p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
+	\end{equation}
+we have
+	\begin{align*}
+p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
+p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
+	\end{align*}
+Therefore
 \begin{align*}
@@
 \end{align*}
-because
+Thus
+\begin{equation}
+(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
+\end{equation}
+Thus
+	\begin{equation}
+H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
+	\end{equation}
+
+Finally, iteratively applying this inequality, we get:
 		\begin{align*}
-p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
-p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
+		H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq H[X_t| X_{t-2, t-1}, L_{t-2}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
 		\end{align*}
-because of Equation~\ref{eq:listener-markov}.
-	\begin{equation}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
+
+	\end{proof}
 
@@
 
-Therefore
+	Plugging this inequality into the equality above:
 \begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( H[X_t|X_{1\dots t-1}, L_0] - H[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
+	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[X_t|X_{1\dots t-1}, L_0] - \operatorname{H}[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
+	& = \frac{1}{T} \left(\operatorname{H}[X_{1\dots T} | L_0] - \operatorname{H}[X_{1\dots T} | X_{\leq 0}]\right)  \\
 	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
@@
 	\begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
+		I[X_{1\dots T}|X_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[X_i, X_j | X_{j+1}...X_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
+	\end{align*}
+	Therefore
+\begin{align*}
+	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
 	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
 \end{align*}
-	$I[X_{1\dots T}|L_0]$ is at most $H[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
+	$I[X_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
 	\begin{align*}
-	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
+	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
 		&= \sum_{t > T} I_t
 \end{align*}
-	Rearranging shows that the listener's surprisal is at least $H[X_t|L_t] \geq H[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
-\end{proof}
+	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[X_t|L_t] \geq \operatorname{H}[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
+%\end{proof}
 

### 2019-03-24 paper.tex
revising

@@
 
+\usepackage{showlabels}
 \usepackage{fullpage}
@@
 
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$, we obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
+\end{center}
+	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
+\end{figure}
+
+The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$.
+Then interplate linearly (justify this in appendix).
+We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
 We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
@@
 
-We illustrate the linguistic predictions of Proposition \ref{prop:lower-bound} by reanalyzing the data from \cite{fedzechkina-human-2017}.
-This is a miniature artificial language study that showed a bias for Dependency Length Minimization in production in artificial language learning.
-Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in~(\ref{eq:memory-bound}).
-
-As ~(\ref{eq:memory-bound}) is invariant under reversal of the language, we only consider the head-final version of her artificial language.
-The language has consistent head-final order, and uses case marking on objects.
-The relevant production targets are transitive sentences where one of the two arguments is much longer than the other, due to the presence of a PP modifier, as shown in Table~\ref{tab:artificial}.
-The language has variable order of subjects and objects; for the production targets, the B versions produce much longer dependencies than the A versions.
-Dependency Length Minimization thus predicts that speakers are more likely to use the A versions.
-\cite{fedzechkina-human-2017} confirmed this experimentally.
-
-\begin{table}
-	\textbf{A Orders: Short Dependencies}
-
-		\begin{tabular}{ccc}
-			Object & Subject & Verb \\
-			\fbox{\begin{tabular}{llllll}
-				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
-					\end{tabular}} & \fbox{\begin{tabular}{l}Noun\end{tabular}} & \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
-		\end{tabular}
-\\
-		\begin{tabular}{ccc}
-			Subject & Object & Verb \\
-			\fbox{\begin{tabular}{llllll}
-				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
-					\end{tabular}} & \fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & Verb \\
-		\end{tabular}
-\\
-\\
-
-	\textbf{B Orders: Long Dependencies}
-
-		\begin{tabular}{ccc}
-			Subject & Object & Verb \\
-			 \fbox{\begin{tabular}{l}Noun\end{tabular}} &  \fbox{\begin{tabular}{llllll}
-				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
-		\end{tabular}}  &  \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
-		\end{tabular}
-\\
-		\begin{tabular}{ccc}
-			Object & Subject & Verb \\
-			\fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & \fbox{\begin{tabular}{llllll}
-				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
-		\end{tabular}} & Verb \\
-		\end{tabular}
-
-			\caption{Production targets in the artificial mini language from \cite{fedzechkina-human-2017}.
-The language has head-final order, with free variation between SO and OS orders.
-When one of the arguments is much longer than the other, placing the longer one first (`A' orders) shortens syntactic dependencies, compared to `B' orders.}\label{tab:artificial}
-
-\end{table}
-
-In this section, we show that our bound on speaker memory makes the same prediction, without reference to syntactic structure or specific memory architectures.
-
-We constructed one language consisting of the A versions, and one language consisting of the B versions.
-Following the experimental setup of \cite{fedzechkina-human-2017}, we assigned equal probability to the two possible configurations per language, and used a separate set of nouns (inanimate nouns) for the embedded noun in the long phrase.
-
-We interpreted each of the two languages as a stationary processes, extending infinitely in both directions, by concatenating independent samples drawn from the language.
-			We computed (\ref{eq:memory-bound}) from a chain of 1000 independently sampled sentences, for each of the two versions of the toy language.
-			Figure~\ref{fig:toy-mis} (left) shows the curve of the conditional mutual information $I_t$ as a function of the distance $t$.
-			The curves differ at $t=2$ and $t=5$: 
-			About 0.073 nats of predictive information that are at distance $t=2$ in the A orders are moved to $t=5$ in the B orders.
-%			In this sense, A orders have greater information locality than the B orders.
-			%What is responsible for this difference?
-			The source of the difference lies in predicting the presence and absence of a case marker on the second argument -- i.e., whether to anticipate a subject or object.
-			In the A orders, considering the last two words is sufficient to make this decision.
-			In the B orders, it is necessary to consider the word before the long second constituent, which is five words in the past.
-
-			The total amounts of predictive information -- corresponding to the area under the curve -- are the same, indicating that both languages are equally predictable.
-			However, we will see that the memory demands are different:
-			Figure~\ref{fig:toy-mis} (right) shows $t\cdot I_t$ as a function of $t$.
-			As $I_t$ decays faster in A orders, the total area under the curve now differs between A and B, and is larger in B.
-			%This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.
-
-%While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
-%That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.
-
-In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-%In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
-The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
-For reaching optimal surprisal, Order A requires strictly less memory.
-Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.
-
-It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.
-
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/toy-mis.pdf}
-\includegraphics[width=0.45\textwidth]{toy/toy-t-mis.pdf}
-%
-	\caption{Left: Decay of Conditional Mutual Information, as a function of the distance $t$, for the two versions in the artificial language.
-The areas under the two curves are identical, corresponding to the fact that both orders are equally predictable.
-However, mutual Information decays faster in Order A.\ \ \ Right: $t I_t$, as a function of $t$.
-The area under the B curve is larger, corresponding to larger memory demand for this order.}\label{fig:toy-mis}
-\end{figure*}
-%Both languages have the same overall entropy rate, but they differ in the distribution of predictive information.
-%
-%plot of $I_t$
-%
-%The areas under the curves are identical.
-%
-%good version
-%
-%%CONTEXT LENGTH 0   2.06856758681  9997.93143241   0.0
-%%CONTEXT LENGTH 1   0.29195106479  1.77661652202   1.77661652202
-%%CONTEXT LENGTH 2   0.0729865489103  0.21896451588   2.21454555377
-%%CONTEXT LENGTH 3   0.0729865489103  0.0   2.21454555377
-%%CONTEXT LENGTH 4   0.0729865489103  0.0   2.21454555377
-%%CONTEXT LENGTH 5   0.0729865489103  0.0   2.21454555377
-%%CONTEXT LENGTH 6   0.0729865489103  0.0   2.21454555377
-%
-%bad version
-%
-%%CONTEXT LENGTH 0   2.06937301571  9997.93062698   0.0
-%%CONTEXT LENGTH 1   0.291673373027  1.77769964269   1.77769964269
-%%CONTEXT LENGTH 2   0.145830550934  0.145842822093   2.06938528687
-%%CONTEXT LENGTH 3   0.145830550934  0.0   2.06938528687
-%%CONTEXT LENGTH 4   0.145830550934  0.0   2.06938528687
-%%CONTEXT LENGTH 5   0.0729152754672  0.0729152754672   2.43396166421
-%%CONTEXT LENGTH 6   0.0729152754672  0.0   2.43396166421
-%%CONTEXT LENGTH 7   0.0729152754672  0.0   2.43396166421
-%%
-%
-%
-%
-%%
-%%grammar:
-%%
-%%S $\rightarrow$ Obj Subj V (1/2) | Subj Obj V (1/2)
-%%
-%%Obj $\rightarrow$ NP di
-%%
-%%Subj $\rightarrow$ NP
-%%
-%%NP $\rightarrow$ N (3/4) | PP NP (1/8) | Adj NP (1/8)
-%%
-%%PP $\rightarrow$ NP P
-%%
-%
-%
+\input{toy-experiment.tex}
 
@@
 We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We will define a class of baseline languages that serve as a comparison class of word orders that natural language can be compared to.
 We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-
-We want to compare to counterfactual reorderings.
 We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order gramars.
-
-\paragraph{Data}
-We draw on dependency corpora.
+Therefore, we compare to counterfactual word order grammars.
+
+\begin{figure}
+\centering
+\begin{dependency}[theme = simple]
+   \begin{deptext}[column sep=1em]
+	   I \&	   wrote \& risāla \& li \& sadīq  \\
+   \end{deptext}
+	%   \deproot{3}{ROOT}
+   \depedge{1}{2}{obj}
+	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
+   \depedge{1}{4}{obl}
+   \depedge{4}{3}{case}
+   %\depedge[arc angle=50]{7}{6}{ATT}
+\end{dependency}
+	\caption{TODO Dependencies example}\label{fig:dependency}
+\end{figure}
+
+\subsection{Data}
+We draw on corpora annotated with syntactic structures.
 The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
 
-In dependency corpora, sentences are annotated with \emph{dependency trees}. TODO a figure.
+\paragraph{Dependency Grammar}
+In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
 These are directed trees describing the grammatical relations among words.
@@
 
+\paragraph{Corpora}
 We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 500 sentences of training data.
@@
 
-\paragraph{Counterfactual Ordering Grammars}
+\subsection{Counterfactual Ordering Grammars}
 We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
@@
 
-This parameter defines how dependents are ordered relative to their head:
+In our model, this parameter defines how dependents are ordered relative to their head:
 Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
@@
 
-\paragraph{Baseline Languages}
 We define baseline grammars by randomly sampling the parameters $a_\tau$.
-
 Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
+
+\paragraph{Discussion}
+In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
+However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
+Word order freedom also is not modeled.
+In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
 
@@
 To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE) and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-We model language on the level of individual word forms.
+%We model language on the level of individual word forms.
 We provide data from alternative estimation methods in the SI.
@@
 
-\paragraph{Model}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997}.
 %Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%
-\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-
-%Such networks estimate the probability of a word in context as follows.
-Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-It then computes a sequence of hidden and cell states by the following recurrent computation:
-\begin{align*}
-	h_1, c_1 &:= 0 \\
-	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-	\dots \\
-	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-\end{align*}
-The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-
-The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-We obtain per-word predictions of the next word by computing
-\begin{align*}
-	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-\end{align*}
-where the softmax transformation normalizes vectors into probability distributions as follows
-\begin{equation}
-	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-\end{equation}
-Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-\begin{equation}
-	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-\end{equation}
-and thus the surprisal is estimated as
-\begin{equation}
-- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-\end{equation}
-We discuss the choice of the numerical parameters in the next section.
-
-\paragraph{Training}
+%%
+%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
+%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
+%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
+%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
+%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
+%
+%%Such networks estimate the probability of a word in context as follows.
+%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
+%It then computes a sequence of hidden and cell states by the following recurrent computation:
+%\begin{align*}
+%	h_1, c_1 &:= 0 \\
+%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
+%	\dots \\
+%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
+%\end{align*}
+%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
+%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
+%
+%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
+%We obtain per-word predictions of the next word by computing
+%\begin{align*}
+%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
+%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
+%\end{align*}
+%where the softmax transformation normalizes vectors into probability distributions as follows
+%\begin{equation}
+%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
+%\end{equation}
+%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
+%\begin{equation}
+%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+%\end{equation}
+%and thus the surprisal is estimated as
+%\begin{equation}
+%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+%\end{equation}
+%We discuss the choice of the numerical parameters in the next section.
+%
+
+\paragraph{Model and Parameter Estimation}
+We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
+This architecture takes as input a sequence $x_1 ...
+x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
+w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
+
+The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
 Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-We can think of the LSTM parameters as forming one large vector $\theta$.
+%We can think of the LSTM parameters as forming one large vector $\theta$.
 At the beginning of training, the parameters $\theta$ are randomly initialized.
+
+The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
 If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
@@
 where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-
-Word sequences are sampled without replacement.
+When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
+In this process, the word sequences are sampled without replacement.
 Once all sequences have been processed, we start another pass through the training data.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the development partition.
-We terminate training once this development partition surprisal does not improve over the one computed after the previous pass any more.
-
-\paragraph{Regularization}
+Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
+
+After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
+We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
+
+%\paragraph{Regularization}
 %Dropout, Word dropout, Word noising
 %Any statistical estimation problem faces the 
-The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
+%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
 %In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-We use three standard methods of regularization that have been shown to improve neural language modeling: dropout (CITE), word dropout (CITE), and word noising (CITE).
 %First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
@@
 
-The LSTM model has a set of numerical parameters that need to be specified before training, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$.
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting from a fully trained model.
-
-For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them against our hypothesis.
+The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
+We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
+
+For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
+We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
+This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
+
+For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
+For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
+Following (CITE), we represented other words by their part-of-speech tags.
+This affected languages with larger corpora (TODO).
+We believe that this modeling limitation does not affect our results for the following reasons.
+First, this affects the same words in real and counterfactually ordered sentences.
+Second, excluded words are extremely infrequent, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
+Many of the excluded words occur only once in the dataset.
+This means that any model would only be able to extract very limited information on these words from the available data, likely \emph{less} than what is provided by the part-of-speech tag.
+Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
 
@@
 and take the average over these.
-We cut $T$ off at 30.
+We cut $T$ off at 20, as this is the length of the sequences processed by the model.
+
+We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
+This is justified theoretically (TODO maybe discuss when introducing the theorem).
+
+We estimate the unigram entropy $H[X_0]$ by averaging over all models.
 
@@
 
-\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from random grammars.
+%\subsection{Setup}
+
+For each language, we collected data from the actual orderings and from several random grammars.
 We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-We collected data from the actual and random orderings in proportion one to two.
-The stopping criterion will be described below.
-
-\paragraph{Listener Memory: Surprisal-Memory Tradeoff}
-
-\paragraph{Stopping Criterion}
+For each of the random grammars, we collect one sample.
+Data is collected according to a precision-based stopping criterion described in Section (REF).
+%We collected data from the actual and random orderings in proportion one to two.
+%The stopping criterion will be described below.
+
+%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
+%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
+%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
+%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
 
@@
 
-For each language, we viewed surprisal as a function of memory.
-We used linear interpolation to obtain surprisal values for all levels of memory.
+%For each sample, we estim
+%For each language, we viewed surprisal as a function of memory.
+%We used linear interpolation to obtain surprisal values for all levels of memory.
 
 We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-For each model for each language, we interpolate the surprisal-memory tradeoff curves given by Proposition~\ref{prop:suboptimal} linearly.
-We estimate the unigram entropy $H[X_0]$ by averaging over all models.
-We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-
+%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
 We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We describe the statistics computed in the following paragraphs.
-
+We now describe the statistics we use to quantifying the difference between real and baseline languages.
 We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
@@
 \paragraph{Confidence Interval for Medians}
-We create (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
+We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
+We consider the medians over all runs for the real language, and over all baselines grammars.
 
@@
 \paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-Let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{population median} of the samples.
+For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
+Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
+We take the real language to be represented by the \emph{sample median}.
 For each level $\mu$ of memory, we consider the null hypothesis that
@@
 \end{equation}
-In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
+%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
+We use the Binomial Test.
+
+%This statistic also should have a reasonable Bayesian interpretation:
+% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
 
@@
 
-For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
+%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
 %\begin{equation}
@@
 %\end{equation}
-We use the empirical median for the real language.
+%We use the empirical median for the real language.
 
@@
 
-We take the REAL values to be estimated exactly by their medians.
+%We take the REAL values to be estimated exactly by their medians.
 
@@
 
+Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+
 \end{document}

### 2019-03-25 code/create_models/analyzeVocabSize.R
revising

@@ -1,0 +1,1 @@
+data = read.csv("vocab_sizes.tsv", sep="\t")

### 2019-03-25 code/create_models/collectVocabSizes.py

@@ -1,0 +1,10 @@
+with open("../../ud_languages.txt", "r") as inFile:
+  languages = inFile.read().strip().split("\n")
+
+import subprocess
+with open("vocab_sizes.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "OOV_Perc", "VocabSize", "MostFrequentOOVFreq", "HapaxAmongOOV"])
+  for language in languages:
+    print >> outFile, (subprocess.check_output(["../python27", "vocab_size.py", language])).strip()
+
+

### 2019-03-25 code/create_models/corpusIterator.py

@@ -12,7 +12,7 @@ def readUDCorpus(language, partition):
       files = []
       while len(files) == 0:
         if len(basePaths) == 0:
-           print "No files found"
+           print >> sys.stderr, "No files found"
            raise IOError
         basePath = basePaths[0]
         del basePaths[0]
@@ -21,11 +21,11 @@ def readUDCorpus(language, partition):
       data = []
       for name in files:
         if "Sign" in name:
-           print "Skipping "+name
+           print >> sys.stderr, "Skipping "+name
            continue
         assert ("Sign" not in name)
         if "Chinese-CFL" in name:
-           print "Skipping "+name
+           print >> sys.stderr, "Skipping "+name
            continue
         suffix = name[len("UD_"+language):]
         if name == "UD_French-FTB":
@@ -35,7 +35,7 @@ def readUDCorpus(language, partition):
         subDirFiles = os.listdir(subDirectory)
         partitionHere = partition
         if (name in ["UD_North_Sami", "UD_Irish", "UD_Buryat-BDT", "UD_Armenian-ArmTDP"]) and partition == "dev" and (not language.endswith("-Adap")):
-            print "Substituted test for dev partition"
+            print >> sys.stderr, "Substituted test for dev partition"
             partitionHere = "test"
         elif language.endswith("-Adap"):
           if (name in ["UD_Kazakh-KTB", "UD_Cantonese-HK", "UD_Naija-NSC", "UD_Buryat-BDT", "UD_Thai-PUD", "UD_Breton-KEB", "UD_Faroese-OFT", "UD_Amharic-ATT", "UD_Kurmanji-MG", "UD_Upper_Sorbian-UFAL", "UD_Bambara-CRB", "UD_Erzya-JR"]):
@@ -45,7 +45,7 @@ def readUDCorpus(language, partition):
             
         candidates = filter(lambda x:"-ud-"+partitionHere+"." in x and x.endswith(".conllu"), subDirFiles)
         if len(candidates) == 0:
-           print "Did not find "+partitionHere+" file in "+subDirectory
+           print >> sys.stderr, "Did not find "+partitionHere+" file in "+subDirectory
            continue
         if len(candidates) == 2:
            candidates = filter(lambda x:"merged" in x, candidates)
@@ -66,7 +66,7 @@ def readUDCorpus(language, partition):
                        assert False
               data = data + newData
         except IOError:
-           print "Did not find "+dataPath
[…]

### 2019-03-25 code/create_models/vocab_size.py

@@ -1,0 +1,164 @@
+#/u/nlp/bin/stake.py -g 11.5g -s run-stats-pretrain2.json "python readDataDistEnglishGPUFree.py"
+
+
+
+# TODO also try other optimizers
+
+import random
+import sys
+
+objectiveName = "LM"
+
+language = sys.argv[1]
+
+
+
+posUni = set() #[ "ADJ", "ADP", "ADV", "AUX", "CONJ", "DET", "INTJ", "NOUN", "NUM", "PART", "PRON", "PROPN", "PUNCT", "SCONJ", "SYM", "VERB", "X"] 
+
+posFine = set() #[ "``", ",", ":", ".", "''", "$", "ADD", "AFX", "CC",  "CD", "DT", "EX", "FW", "GW", "HYPH", "IN", "JJ", "JJR",  "JJS", "-LRB-", "LS", "MD", "NFP", "NN", "NNP", "NNPS", "NNS",  "PDT", "POS", "PRP", "PRP$", "RB", "RBR", "RBS", "RP", "-RRB-", "SYM", "TO", "UH", "VB", "VBD", "VBG", "VBN", "VBP", "VBZ",  "WDT", "WP", "WP$", "WRB", "XX" ]
+
+
+
+deps = ["acl", "acl:relcl", "advcl", "advmod", "amod", "appos", "aux", "auxpass", "case", "cc", "ccomp", "compound", "compound:prt", "conj", "conj:preconj", "cop", "csubj", "csubjpass", "dep", "det", "det:predet", "discourse", "dobj", "expl", "foreign", "goeswith", "iobj", "list", "mark", "mwe", "neg", "nmod", "nmod:npmod", "nmod:poss", "nmod:tmod", "nsubj", "nsubjpass", "nummod", "parataxis", "punct", "remnant", "reparandum", "root", "vocative", "xcomp"] 
+
+#deps = ["acl", " advcl", " advmod", " amod", " appos", " aux", " case cc", " ccompclf", " compound", " conj", " cop", " csubjdep", " det", " discourse", " dislocated", " expl", " fixed", " flat", " goeswith", " iobj", " list", " mark", " nmod", " nsubj", " nummod", " obj", " obl", " orphan", " parataxis", " punct", " reparandum", " root", " vocative", " xcomp"]
+
+
+from math import log, exp
+from random import random, shuffle, randint
+
[…]

### 2019-03-25 code/preparePaper/samplesNumber_ground.py

@@ -1,0 +1,77 @@
+# ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
+
+with open("../../ud_languages.txt", "r") as inFile:
+   languages = inFile.read().strip().split("\n")
+
+languages = sorted(list(set(languages)))
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data[-1]) == 1 and data[-1][0] == "":
+       del data[-1]
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+def h(frame, line, name):
+    return line[frame[0][name]]
+
+
+#with open("tradeoff/effectSize.tsv", "r") as inFile:
+#   effectSize = readTSV(inFile)
+
+#with open("tradeoff/effectSize_diff.tsv", "r") as inFile:
+#   effectSize_diff = readTSV(inFile)
+
+with open("../../results/tradeoff/stats-onlyWordForms-boundedVocab_GROUND.tsv", "r") as inFile:
+   stats = readTSV(inFile)
+languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
+
+
+
+with open("../corpusSizes.tsv", "r") as inFile:
+  corpusSizes = readTSV(inFile) #dict([x.split("\t") for x in inFile.read().strip().split("\n")])
+languageKey_corpusSizes = dict([(h(corpusSizes, line, "Language"), line) for line in corpusSizes[1]])
+
+
+
+with open("../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
+   listener_curve = readTSV(inFile)
+
+languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
+
+output = []
+
+for language in languages:
+  #   line = languageKey[language]
+     line_listener = languageKey_listener_curve[language]
[…]

### 2019-03-25 toy-experiment.tex

We illustrate the linguistic predictions of Proposition \ref{prop:lower-bound} by reanalyzing the data from \cite{fedzechkina-human-2017}.
This is a miniature artificial language study that showed a bias for Dependency Length Minimization in production in artificial language learning.
Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in~(\ref{eq:memory-bound}).

As ~(\ref{eq:memory-bound}) is invariant under reversal of the language, we only consider the head-final version of her artificial language.
The language has consistent head-final order, and uses case marking on objects.
The relevant production targets are transitive sentences where one of the two arguments is much longer than the other, due to the presence of a PP modifier, as shown in Table~\ref{tab:artificial}.
The language has variable order of subjects and objects; for the production targets, the B versions produce much longer dependencies than the A versions.
Dependency Length Minimization thus predicts that speakers are more likely to use the A versions.
\cite{fedzechkina-human-2017} confirmed this experimentally.

\begin{table}
	\textbf{A Orders: Short Dependencies}

		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun\end{tabular}} & \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & Verb \\
		\end{tabular}
\\
\\

	\textbf{B Orders: Long Dependencies}

		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			 \fbox{\begin{tabular}{l}Noun\end{tabular}} &  \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
		\end{tabular}}  &  \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
		\end{tabular}} & Verb \\
		\end{tabular}

			\caption{Production targets in the artificial mini language from \cite{fedzechkina-human-2017}. The language has head-final order, with free variation between SO and OS orders. When one of the arguments is much longer than the other, placing the longer one first (`A' orders) shortens syntactic dependencies, compared to `B' orders.}\label{tab:artificial}

\end{table}

In this section, we show that our bound on speaker memory makes the same prediction, without reference to syntactic structure or specific memory architectures.

We constructed one language consisting of the A versions, and one language consisting of the B versions.
Following the experimental setup of \cite{fedzechkina-human-2017}, we assigned equal probability to the two possible configurations per language, and used a separate set of nouns (inanimate nouns) for the embedded noun in the long phrase.

We interpreted each of the two languages as a stationary processes, extending infinitely in both directions, by concatenating independent samples drawn from the language.
			We computed (\ref{eq:memory-bound}) from a chain of 1000 independently sampled sentences, for each of the two versions of the toy language.
			Figure~\ref{fig:toy-mis} (left) shows the curve of the conditional mutual information $I_t$ as a function of the distance $t$.
			The curves differ at $t=2$ and $t=5$: 
			About 0.073 nats of predictive information that are at distance $t=2$ in the A orders are moved to $t=5$ in the B orders.
%			In this sense, A orders have greater information locality than the B orders.
			%What is responsible for this difference?
			The source of the difference lies in predicting the presence and absence of a case marker on the second argument -- i.e., whether to anticipate a subject or object.
			In the A orders, considering the last two words is sufficient to make this decision.
			In the B orders, it is necessary to consider the word before the long second constituent, which is five words in the past.

			The total amounts of predictive information -- corresponding to the area under the curve -- are the same, indicating that both languages are equally predictable.
			However, we will see that the memory demands are different:
			Figure~\ref{fig:toy-mis} (right) shows $t\cdot I_t$ as a function of $t$.
			As $I_t$ decays faster in A orders, the total area under the curve now differs between A and B, and is larger in B.
			%This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.

%While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
%That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.

In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
%In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
For reaching optimal surprisal, Order A requires strictly less memory.
Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.

It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/toy-mis.pdf}
\includegraphics[width=0.45\textwidth]{toy/toy-t-mis.pdf}
%
	\caption{Left: Decay of Conditional Mutual Information, as a function of the distance $t$, for the two versions in the artificial language. The areas under the two curves are identical, corresponding to the fact that both orders are equally predictable. However, mutual Information decays faster in Order A.\ \ \ Right: $t I_t$, as a function of $t$. The area under the B curve is larger, corresponding to larger memory demand for this order.}\label{fig:toy-mis}
\end{figure*}
%Both languages have the same overall entropy rate, but they differ in the distribution of predictive information.
%
%plot of $I_t$
%
%The areas under the curves are identical.
%
%good version
%
%%CONTEXT LENGTH 0   2.06856758681  9997.93143241   0.0
%%CONTEXT LENGTH 1   0.29195106479  1.77661652202   1.77661652202
%%CONTEXT LENGTH 2   0.0729865489103  0.21896451588   2.21454555377
%%CONTEXT LENGTH 3   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 4   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 5   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 6   0.0729865489103  0.0   2.21454555377
%
%bad version
%
%%CONTEXT LENGTH 0   2.06937301571  9997.93062698   0.0
%%CONTEXT LENGTH 1   0.291673373027  1.77769964269   1.77769964269
%%CONTEXT LENGTH 2   0.145830550934  0.145842822093   2.06938528687
%%CONTEXT LENGTH 3   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 4   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 5   0.0729152754672  0.0729152754672   2.43396166421
%%CONTEXT LENGTH 6   0.0729152754672  0.0   2.43396166421
%%CONTEXT LENGTH 7   0.0729152754672  0.0   2.43396166421
%%
%
%
%
%%
%%grammar:
%%
%%S $\rightarrow$ Obj Subj V (1/2) | Subj Obj V (1/2)
%%
%%Obj $\rightarrow$ NP di
%%
%%Subj $\rightarrow$ NP
%%
%%NP $\rightarrow$ N (3/4) | PP NP (1/8) | Adj NP (1/8)
%%
%%PP $\rightarrow$ NP P
%%
%
%

### 2019-03-25 slice-hists_REAL_0.tex
slices

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-03-25 slice-hists_REAL_1.tex
slices

English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-03-25 code/preparePaper/sliceHistTables_REAL.py

@@ -1,0 +1,77 @@
+# ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
+
+with open("../../ud_languages.txt", "r") as inFile:
+   languages = inFile.read().strip().split("\n")
+
+languages = sorted(list(set(languages)))
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data[-1]) == 1 and data[-1][0] == "":
+       del data[-1]
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+def h(frame, line, name):
+    return line[frame[0][name]]
+
+
+#with open("tradeoff/effectSize.tsv", "r") as inFile:
+#   effectSize = readTSV(inFile)
+
+#with open("tradeoff/effectSize_diff.tsv", "r") as inFile:
+#   effectSize_diff = readTSV(inFile)
+
+with open("../../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
+   stats = readTSV(inFile)
+languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
+
+
+entries = []
+
+for language in languages:
+#   line = languageKey[language]
+   components = [language.replace("_"," ").replace("-Adap", "")]
+   #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
+   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}" )
+   entries.append(components)
+
+ROWS_PER_PART = 4
+COLUMNS = 4
+entries += [["",""] for x in range((COLUMNS-(len(entries)%COLUMNS))%COLUMNS)]
+
+outputRows = []
+if True:
[…]

### 2019-03-25 paper.tex
revising

@@
 For each of these languages, we pooled all available corpora in one dataset.
+We excluded corpora that primarily contain text created by non-native speakers.
 Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
@@
 For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags.
-This affected languages with larger corpora (TODO).
+Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
+This applied to 37 languages, affecting an average of 11~\% of words in this languages.
 We believe that this modeling limitation does not affect our results for the following reasons.
 First, this affects the same words in real and counterfactually ordered sentences.
-Second, excluded words are extremely infrequent, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset.
-This means that any model would only be able to extract very limited information on these words from the available data, likely \emph{less} than what is provided by the part-of-speech tag.
+Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
+Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
+This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
 Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
@@
 
-In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
+%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
 
@@
 
-In Figure~\ref{tab:slices}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{tab:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
+In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
+
+In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
 
@@
 \begin{table}
+\begin{longtable}{l|lll||l|lllllllllllllll}
+	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
+\input{boot_g_REAL.tex}
+\end{longtable}
+	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
+\end{table}
+
+\begin{table}
 \begin{longtable}{ccccccccccccccclll}
@@
 \input{medians_REAL_3.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{slice-hists_REAL_0.tex}
+\end{longtable}
+	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{slice-hists_REAL_1.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{slice-hists_REAL_2.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{slice-hists_REAL_3.tex}
 \end{longtable}
@@
 \begin{table}
+\begin{longtable}{l|ll||l|llllllllllllll}
+	Language & Base. & MLE & Language & Base. & MLE \\ \hline
+\input{samplesNumber_ground.tex}
+\end{longtable}
+	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{table}
+
+\begin{table}
 \begin{longtable}{ccccccccccccccclll}
@@
 \end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+	\caption{Experiment 3.
+Medians: For each memory budget, we provide the median surprisal for real and random languages.
 Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
@@
 
-\subsection{Experiment 2b: Fixed Word Orders}
+\section{Experiment 3: Fixed Word Orders}
 

### 2019-03-25 paper.tex
add intro

@@
 
+Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
+Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
+Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
+
+At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
+For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
+
+Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
+Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
+
+Our main new concept is the idea of a \key{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
+Analogous results also hold for language production by resource-constrained speakers.
+We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
+
+The remainder of this paper is structured as follows.
+In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
+In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
+In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \key{information locality}: whenever utterance elements that predict each other are close to each other.
+We argue that information locality is a good model of the effects of memory constraints on language processing.
+In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
+In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
+Section~\ref{sec:conclusion} concludes.
+
+\section{Background}
+\label{sec:background}
+% More detailed lit review
+
 A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unaccebtability of multiple center embeddings in English to limitations of human working memory.
+An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
 \cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
@@
 
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
 All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
@@
 
+%TODO mention prior work
+%
+%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
+%
+%- explanation of greenberg universals
+%
+%- dependency length minimization
+%
+%- \cite{gibson-linguistic-1998}
+%
+
+%\paragraph{\cite{berwick-grammatical-1986}}
+%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
+
+%also (CITE): operations in a specific incremental parsing model
+
 \section{Memory-Surprisal Tradeoff}
+\label{sec:ms-tradeoff}
 

### 2019-04-03 paper-short.tex
short paper

--- 2019-03-24 paper.tex
@@
 
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
+\title{TODO title}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
+\date{2019}
 
@@
 
-%
-\begin{abstract}
+Languages order words
 
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
+Languages have different word orders
 
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
+test hypothesis that these grammars represent different solutions to efficient computation.
 
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
+For several decades, basically everyone has been saying something like this on a theoretical level, from Chomsky (approximately since the 90s?) to Hawkins.
 
-\section{Introduction}
+This has been difficult to test for two reasons:
 
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unaccebtability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
+(1) Need a lot of data representing the distribution of underlying hierarchical structures that people communicate,
 
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
+(2) Need computational model of efficiency in computation.
 
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
+We use information-theory to for the first time really test (2).
+We use information-theoretical lower bounds on the difficulty in language processing.
 
-%also (CITE): operations in a specific incremental parsing model
+surprisal as processing effort
 
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
+memory: how many bits of memory about past input does a listener need to represent.
 
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
+related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory.
 
-\section{Memory-Surprisal Tradeoff}
+memory-surprisal tradeoff
 
-With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal
+conditional MI
 
-In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.
+theorem
 
-We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
-%Informally, we expect that both speaker and listener will be affected by memory demands:
-%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
-For the listener, predicting the next word well requires maintaining information about the past.
-For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
-For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
-Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.
+what we did
 
-TODO say at some point that we're studying sentence-internal memory
+results
 
-\subsection{Theoretical Results}
+some discussion
 
-We now introduce our main theoretical result on memory-surprisal tradeoffs.
-
-\paragraph{Setting}
-We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-
-We model the sequence as a probabilistic sequence; that is, given a context $x_{<t}$, the next word is distributed according to a distribution $p(x_t|x_{<t})$.
-
-% $(X_t)_{t \in \mathbb{Z}}$
-%We think of the language as a stochastic process, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
-%As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
-%This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
-%We make the assumption that this process is \emph{stationary}.
-%Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
-%Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
-%In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-%Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
-
-%The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
-%We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
-%The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
-%\begin{equation}\label{eq:markov-speaker}
-%X_{>t} \bot X_{\leq t} | S_t
-%\end{equation}
-%This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
-
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
-	\caption{Illustration of (\ref{eq:listener-markov}).
-As the utterance unfolds, the listener maintains a memory state.
-After receiving word $X_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $X_t$.}\label{fig:listener-markov}
-\end{figure}
-
-There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $L_t$ is determined by the last word $X_t$, and the prior memory state $L_{t-1}$.
-As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
-This is formalized as a statement about conditional probabilities:
-	\begin{equation}\label{eq:listener-markov}
-p(L_1| (X_{t})_t, L_0)   = p((X_{t})_t| L_0, X_1)
-%p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.	
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
-This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-For the study of memory in sentence processing, this seems fair. Discuss this more.
-
-%
-%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-%Formally: 
-%\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot X_{>t} | X_{\leq t}
-%\end{equation}
-%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-%
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-%\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-%\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
-
-%The listener can trade off memory and future surprisal:
-%A listener who chooses to store less memory will exerience higher surprisal in the future.
-%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-%Consider a listener who uses $J$ bits of memory on average.
-%What can we say about the listener's surprisal?
-
-\paragraph{Conditional Mutual Information}
-
-We will use the concept of \emph{conditional mutual information} (Figure~\ref{fig:it-surprisal-intuition}).
-%For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
-%\begin{equation}
-%	I[X,Y] := H[X] - H[X|Y]
-%\end{equation}
-%For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-%\begin{equation}
-%I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-%\end{equation}
-%We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-%	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-	I_t := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-\end{equation}
-This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
-	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}\label{fig:it-surprisal-intuition}
-\end{figure*}
-
-We return to the information and memory curves in Figure~\ref{fig:basic}.
-In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
-This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
-Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
-We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
-As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-Then
-
-\begin{thm}\label{prop:suboptimal}
-	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
-	\begin{equation}\label{eq:memory}
-		\sum_{t=1}^T t I_t
-	\end{equation}
-bits of memory on average.
-Then this listener will incur surprisal at least
-	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-	on average.
-\end{thm}
-The proof is given in the appendix (REF).
-
-%An intuitive (non-rigorous) argument goes as follows:
-%For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
-
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
-%
-	\caption{Left: $I_t$ as a function of $t$, for two different processes.
-$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
-Left: $t \cdot I_t$ as a function of $t$ for the same processes.
-}\label{fig:basic}
-\end{figure*}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
-\end{center}
-	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
-\end{figure}
-
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-Then interplate linearly (justify this in appendix).
-We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
-We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
-Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
-In the given example, $T=4$.
-By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
-	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
-Recall that the red process had a faster decay of conditional mutual information.
-Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
-	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
-\end{figure}
-
-%that the model does not have an `absolute clock' built into it.
-
-Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
-% and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-\paragraph{Information Locality}
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-\paragraph{Center Embeddings}
-
-\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-
-%\cite{gibson-linguistic-1998}
-
-%first center-embeddings, POS level model
-
-%Later work found evidence that cross-serial dependencies are less taxing.
-
-%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies
-
-\paragraph{Other Psycholinguistic Predictions}
-
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries.
-
-\paragraph{Speakers}
-
-% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 52 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.
-
-In our model, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE) and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags.
-This affected languages with larger corpora (TODO).
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, excluded words are extremely infrequent, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset.
-This means that any model would only be able to extract very limited information on these words from the available data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[X_t | X_0, X_1, ..., X_{t-1}] - P_\theta[X_t | X_1, ..., X_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[X_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
-%\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slices}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{tab:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{longtable}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{longtable}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{longtable}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{longtable}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\subsection{Experiment 2b: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Speaker Memory}
-
-%The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
-TODO
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
-%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-Our results apply both to speakers and listeners.
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-%\begin{proposition}
-%Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
-%\end{proposition}
-%
-%\begin{proof}[Proof Sketch]
-%	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
-%
-%	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
-We consider incremental Earley parsing \citep{earley-efficient-1970}.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
-For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
-%	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
-%\end{proof}
-%
-%
-%
-
-\section{Conclusion}
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{longtable}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Proof of the Theorem}
-
-%\begin{proof}
-	We denote the listener's memory state at time $t$, after hearing $X_{<t} = ... X_{t-2} X_{t-1}$ by $L_t$.
-	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[X_t | L_t]$.
-	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]$.
-By the assumption of stationarity, we can rewrite this expression as
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]\right) 
-\end{align*}
-	\begin{lemma}
-For any $t$, the following inequality holds:
-%	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
-	\begin{equation}
-%H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-H[X_t | L_t] \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{equation}
-	\end{lemma}
-	\begin{proof}[Proof of the Lemma]
-
-By Equation~\ref{eq:listener-markov}:
-	\begin{equation}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-we have
-	\begin{align*}
-p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
-p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
-	\end{align*}
-Therefore
-\begin{align*}
-p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) & = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) \\
-& = \sum_{X_{<1}} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1}) \\
-&= \sum_{X_{<1}} p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, X_1, X_{2 \dots t-1}) \\
-&= p(X_t|L_0,  X_1, X_{2 \dots t-1}) 
-\end{align*}
-Thus
-\begin{equation}
-(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
-\end{equation}
-Thus
-	\begin{equation}
-H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
-	\end{equation}
-
-Finally, iteratively applying this inequality, we get:
-		\begin{align*}
-		H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq H[X_t| X_{t-2, t-1}, L_{t-2}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{align*}
-
-	\end{proof}
-
-	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-	Plugging this inequality into the equality above:
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[X_t|X_{1\dots t-1}, L_0] - \operatorname{H}[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(\operatorname{H}[X_{1\dots T} | L_0] - \operatorname{H}[X_{1\dots T} | X_{\leq 0}]\right)  \\
-	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
-\end{align*}
-	The first term $I[X_{1\dots T}|X_{\leq 0}]$ can be rewritten in terms of $I_t$:
-	\begin{align*}
-		I[X_{1\dots T}|X_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[X_i, X_j | X_{j+1}...X_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
-	\end{align*}
-	Therefore
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
-	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
-\end{align*}
-	$I[X_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
-	\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
-		&= \sum_{t > T} I_t
-\end{align*}
-	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[X_t|L_t] \geq \operatorname{H}[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
-%\end{proof}
-
-Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+this subsumes previous ideas about locality in language
 

### 2019-04-03 paper.tex
short paper

@@
 
-Our main new concept is the idea of a \key{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
+Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
 Analogous results also hold for language production by resource-constrained speakers.
@@
 In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \key{information locality}: whenever utterance elements that predict each other are close to each other.
+In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
 We argue that information locality is a good model of the effects of memory constraints on language processing.

### 2019-04-05 medians_0.tex
branching entropies

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\

### 2019-04-05 medians_1.tex
branching entropies

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\

### 2019-04-05 medians_3.tex
branching entropies

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} &  &
+\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} &  &
  \\

### 2019-04-05 medians_REAL_1.tex
branching entropies

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\

### 2019-04-05 medians_ngrams_0.tex
branching entropies

@@
-Afrikaans & Amharic & Arabic & Armenian
+Afrikaans & Arabic & Basque & Bulgarian
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{ngrams/figures/Afrikaans-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Arabic-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Basque-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Bulgarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
-Basque & Breton & Bulgarian & Buryat
+Catalan & Chinese & Croatian & Czech
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{ngrams/figures/Catalan-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Chinese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Croatian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Czech-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
-Cantonese & Catalan & Chinese & Croatian
+Danish & Dutch & Estonian & Finnish
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{ngrams/figures/Danish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Dutch-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Estonian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Finnish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\ 
-Czech & Danish & Dutch & Estonian
+French & German & Greek & Hebrew
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{ngrams/figures/French-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/German-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Greek-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hebrew-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
  \\

### 2019-04-05 code/preparePaper/medianTables.py

@@ -45,7 +45,7 @@ for language in languages:
 #   line = languageKey[language]
    components = [language.replace("_"," ").replace("-Adap", "")]
    #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
-   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}" )
+   components.append( "\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}" )
    entries.append(components)
 
 ROWS_PER_PART = 4

### 2019-04-05 neural/plot_wordsOnly_histByMem_onlyREAL.R

@@ -1,0 +1,81 @@
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+
+# Plots CIs for the quantile
+
+
+fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t") %>% filter(Type != "GROUND")
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    UnigramCE = mean(dataL$UnigramCE)
+    data = fullData %>% filter(Language == language)
+    barWidth = (max(data$MI) - min(data$MI))/30
+    plot = ggplot(data, aes(x=UnigramCE-MI, fill=Type, color=Type)) + theme_classic() + theme(legend.position="none")   + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled..))      + geom_bar(data = data %>% filter(Type %in% c("REAL_REAL", "GROUND")) %>% group_by(Type) %>% summarise(MI=mean(MI)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
+    ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf", sep=""))
+    return(plot)
+}
+
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
+plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
[…]

### 2019-04-05 neural/plot_wordsOnly_mediansAndQuantiles.R

@@ -1,0 +1,109 @@
+# DOES not produce reasonable results. Better to just provide empirical quantile plots.  
+
+library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+
+# Plots CIs for the quantile
+
+# Plots medians with confidence intervals
+
+fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") 
+
+randomRuns = read.csv("../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
+
+library(MASS)
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    UnigramCE = mean(dataL$UnigramCE)
+
+    dataR = randomRuns %>% filter(Language == language)
+    data = fullData %>% filter(Language == language)
+    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) 
+    plot = plot + geom_line(size=2)
+    plot = plot + theme_classic() + theme(legend.position="none") 
+    plot = plot + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
+
+quantiles = data.frame(Memory = c(), MI = c(), Density = c())
+for(i in (1:max(dataR$Position))) {
+   dataRP = dataR  %>% filter(Position == i )
+   quants = quantile(dataRP$Surprisal, (1:9)/10)
+   dens = data.frame(MI=quants, Quantile=(1:9)/10)
+   quantiles = rbind(quantiles, dens %>% mutate(Memory = dataRP$Memory[[1]]))
+}
+quantiles$Type = "RANDOM_BY_TYPE"
+
+plot = plot + geom_line(data = quantiles, aes(x=Memory, y=UnigramCE-MI, group=Quantile), linetype = "dotted", size=0.5) 
+
+
+
[…]

### 2019-04-05 neural/plot_wordsOnly_mediansAndQuantiles_onlyREAL.R

@@ -1,0 +1,109 @@
+# DOES not produce reasonable results. Better to just provide empirical quantile plots.  
+
+library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+
+# Plots CIs for the quantile
+
+# Plots medians with confidence intervals
+
+fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
+
+randomRuns = read.csv("../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
+
+library(MASS)
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    UnigramCE = mean(dataL$UnigramCE)
+
+    dataR = randomRuns %>% filter(Language == language)
+    data = fullData %>% filter(Language == language)
+    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) 
+    plot = plot + geom_line(size=2)
+    plot = plot + theme_classic() + theme(legend.position="none") 
+    plot = plot + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
+
+quantiles = data.frame(Memory = c(), MI = c(), Density = c())
+for(i in (1:max(dataR$Position))) {
+   dataRP = dataR  %>% filter(Position == i )
+   quants = quantile(dataRP$Surprisal, (1:9)/10)
+   dens = data.frame(MI=quants, Quantile=(1:9)/10)
+   quantiles = rbind(quantiles, dens %>% mutate(Memory = dataRP$Memory[[1]]))
+}
+quantiles$Type = "RANDOM_BY_TYPE"
+
+plot = plot + geom_line(data = quantiles, aes(x=Memory, y=UnigramCE-MI, group=Quantile), linetype = "dotted", size=0.5) 
+
+
+
[…]

### 2019-04-05 neural/plot_wordsOnly_mediansAndSmoothed_onlyREAL.R

@@ -1,0 +1,111 @@
+# DOES not produce reasonable results. Better to just provide empirical quantile plots.  
+
+library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+
+# Plots CIs for the quantile
+
+# Plots medians with confidence intervals
+
+fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
+
+randomRuns = read.csv("../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
+
+library(MASS)
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    library(tidyr)
+    library(dplyr)
+    library(ggplot2)
+    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    UnigramCE = mean(dataL$UnigramCE)
+
+    dataR = randomRuns %>% filter(Language == language)
+    data = fullData %>% filter(Language == language)
+    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) 
+    plot = plot + geom_line(size=2)
+    plot = plot + theme_classic() + theme(legend.position="none") 
+    plot = plot + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
+
+densities = data.frame(Memory = c(), MI = c(), Density = c())
+for(i in (1:max(dataR$Position))) {
+   dataRP = dataR  %>% filter(Position == i )
+   dens = density((dataRP)$Surprisal + mvrnorm(n=nrow(dataRP), 0, 0.01)[,1])
+   dens = data.frame(x=dens$x, y=dens$y)
+   densities = rbind(densities, dens %>% rename(MI=x, Density=y) %>% mutate(Memory = dataRP$Memory[[1]]))
+}
+densities$Type = "RANDOM_BY_TYPE"
+
+
+
[…]

### 2019-04-05 paper-short.tex

@@
 \usepackage{showlabels}
-\usepackage{fullpage}
+%\usepackage{fullpage}
 \usepackage{pslatex}
@@
 
-Languages order words
+All human languages express hierarchical thoughts into sentences, i.e., linear strings of words (CITE).
+Languages have different word orders: English orders the object after the verb, Japanese places it before the verb.
+Explaining the variation in the ways languages  order underlying hierarchical structures into linear strings of words has been the focus of a large body of linguistic research (CITE).
+We test the hypothesis that these grammars represent different solutions to the problem of efficient computation.
+Many linguists have suggested something like this on a theoretical level~\citep{chomsky2005three,hauser2002faculty, berwick1984grammatical,hawkins1994performance}.
+However, testing this hypothesis has been difficult for two resons:
+First, because this requires a large amount of data representing the distribution of underlying hierarchical structures that people communicate,
+and
+second, because this presupposes a computatioal model of processing efficiency.
+Such large-scale data has recently become available in the form of corpora annotated with underlying heirarchical structures across dozens of languages~\citep{nivre-universal-2017}.
 
-Languages have different word orders
+We use information theory to quantify efficiency of human language processing.
+Psycholinguistic research has uncovered two basic contributors to human processing effort: Surprisal and Memory.
 
-test hypothesis that these grammars represent different solutions to efficient computation.
+Memory has been harder to quantify
 
-For several decades, basically everyone has been saying something like this on a theoretical level, from Chomsky (approximately since the 90s?) to Hawkins.
+Here, for the first time, we prove a tradeoff between memory and surprisal.
+We quantify memory as the number of bits about prior input that a listener represents.
 
-This has been difficult to test for two reasons:
+related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \citep{PhysRevLett.115.098701}
 
-(1) Need a lot of data representing the distribution of underlying hierarchical structures that people communicate,
+A listener who invests more memory resources to create representations of the past will incur lower average surprisal.
 
-(2) Need computational model of efficiency in computation.
+We predict that languages order information in such a way as to create an optimal tradeoff.
 
-We use information-theory to for the first time really test (2).
-We use information-theoretical lower bounds on the difficulty in language processing.
+%We study language as a 
+Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
+\begin{equation}
+	I_t := I[X_t, X_0 | X_1, ..., X_{t-1}] = H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, ..., X_{t-1}] 
+\end{equation}
+We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
 
-surprisal as processing effort
+(visualization)
 
-memory: how many bits of memory about past input does a listener need to represent.
+We estimated the memory-surprisal tradeoff for 54 human languages, using standard neural-network models~\citep{hochreiter-long-1997}.
+We compared against chance by the following method.
+TODO
 
-related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory.
-
-memory-surprisal tradeoff
-
-conditional MI
-
-theorem
-
-what we did
+TODO results
 
@@
 
-this subsumes previous ideas about locality in language
+relation between our formalization of memory and models typically proposed in psycholinguistics and psychology
+
+- not all models of memory assume that listeners utilize the statistics to compress memory, though that would be more efficient.
+
+- this subsumes previous ideas about locality in language
+
+%Data for (1) has become available with UD, and there is Richard's 2015 paper and other DLM work, but the relation between what they investigate (DLM) and processing is only heuristic.
+%futrell2015largescale
+
+%We use information-theory to for the first time really test (2).
+%We use information-theoretical lower bounds on the difficulty in language processing.
+
+\bibliographystyle{apalike}
+
+\bibliography{literature}
 

### 2019-04-07 paper-short.tex
revising

--- 2019-04-03 paper-short.tex
@@
 
-\title{TODO title}
+\title{???? The Structure of Human Language Reflects Limitations of Working Memory}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
-Languages order words
+Human languages express hierarchical thoughts into sentences, i.e., linear strings of words.
+We show that the ways languages do this reflect an underlying optimization principle:
+Human languages have evolved towards robustness against limitations in human working memory.
 
-Languages have different word orders
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
+Explaining the variation in the ways languages  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
+We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
 
-test hypothesis that these grammars represent different solutions to efficient computation.
+Theoretical linguists have long suggested that the structure of human language reflects a need for efficient processing under resource limitations~\citep{berwick1984grammatical,hawkins1994performance} (also general efficiency mentioned outside functional work, e.g.
+\citep{chomsky2005three,hauser2002faculty}).
 
-For several decades, basically everyone has been saying something like this on a theoretical level, from Chomsky (approximately since the 90s?) to Hawkins.
+We directly test this hypothesis using large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.
 
-This has been difficult to test for two reasons:
+We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
+During comprehension, this listener will represent some amount of information about past words in their short-term memory.
+We consider the question of how many bits of information this listener has to carry from one instant to the next.
+We expect that a listener's ease of comprehension will be modulated by the amount of memory held about previous context:
+A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load.
 
-(1) Need a lot of data representing the distribution of underlying hierarchical structures that people communicate,
+%We use information theory to quantify the tradeoff between memory load and ease of understanding.
+Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted:
+%Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
+%The surprisal of a word in context is the degree to which it cannot be predicted:
+\begin{equation}\label{eq:surp}
+	-\log p(x_i|x_{1...i-1})
+\end{equation}
+where $x_i$ is the $i$-th word in a given utterance (or discourse).
+%This quantity is known to be a strong linear predictor of human processing effort.
 
-(2) Need computational model of efficiency in computation.
+A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
+We denote a listener's memory representation at time $t$ by $l_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
+The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
+This listener will experience processing effort
+\begin{equation}
+	S :=	-\log p(x_t|l_t)
+\end{equation}
+upon hearing the next word $x_t$.
+%We quantify memory as the number of bits about prior input that a listener represents.
 
-We use information-theory to for the first time really test (2).
-We use information-theoretical lower bounds on the difficulty in language processing.
+We prove a theorem relating memory to the experienced comprehension difficulty $S$.
+Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
 
-surprisal as processing effort
+Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
+We compared against chance by the following method.
+For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
+These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences.
+These random grammars apply consistent grammatical rules similar to human language.
 
-memory: how many bits of memory about past input does a listener need to represent.
+Results are show in Figure~\ref{fig:results}.
+Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.001$).
 
-related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory.
+At extremely low memory capacities, real and baseline languages are equally efficient.
+However, the curves soon diverge, and real language provides a more efficient tradeoff.
+The point of divergence is given by that memory capacity that fully exhausts information provided by the immediately preceding word, that is, when the capacity of a Markov model in approximating language is exhausted.
 
-memory-surprisal tradeoff
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/full-results.png}
+	\caption{}\label{fig:results}
+\end{figure}
 
-conditional MI
+Beyond human language, our theorem applies to other stochastic processes.
+To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.
 
-theorem
+% from the introduction to paper.tex:
+%Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
+%Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
+%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
+%
+%At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
+%For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
+%
+%Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
+%Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
+%
+%Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
+Analogous results also hold for language production by resource-constrained speakers.
+We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
+%
+%
 
-what we did
+%However, testing this hypothesis has been difficult for two reasons:
+%First, this requires a large amount of data representing the distribution of underlying hierarchical structures that people communicate,
+%and
+%second, because this presupposes a computatioal model of processing efficiency.
+%Large-scale data has recently become available in the form of corpora annotated with underlying hierarchical structures across dozens of languages~\citep{nivre-universal-2017}.
 
-results
+%We test the idea that language is optimized for processing with limited resources, in particular with limited memory resources.
 
-some discussion
+%Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
+%\begin{equation}
+%I_t := I[X_t, X_0 | X_1, ..., X_{t-1}] = H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, ..., X_{t-1}] 
+%\end{equation}
+%We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
 
-this subsumes previous ideas about locality in language
+%
+%Memory refers to information about previous parts of a sentence maintained in short-term memory.
+%While memory limitations have been shown experimentally to affect human language processing, there is no consensus on the architecture used to represent memories.
+%
+%Here, for the first time, we provide an information-theoretic integration of these two aspects of processing, leading to a general theory of the computational efficiency of language.
+%
+%We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
+%This listener will experience surprisal
+%\begin{equation}
+%	S :=	-\log P(X_t|L_t)
+%\end{equation}
+%upon hearing the next word.
+%%We quantify memory as the number of bits about prior input that a listener represents.
+%The minimal amount of information stored is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
+%%related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \citep{PhysRevLett.115.098701}
+%We note that there is a tradeoff between average memory, $\E[S]$ and memory $H[L_t]$: A listener who invests more memory resources to create representations of the past will incur lower average surprisal.
+%
+%We predict that languages order information in such a way as to create an optimal tradeoff.
+
+%We study language as a
+
+%Data for (1) has become available with UD, and there is Richard's 2015 paper and other DLM work, but the relation between what they investigate (DLM) and processing is only heuristic.
+%futrell2015largescale
+
+%We use information-theory to for the first time really test (2).
+%We use information-theoretical lower bounds on the difficulty in language processing.
+
+\bibliographystyle{apalike}
+
+\bibliography{literature}
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
+\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
+	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
+Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
+In the given example, $T=4$.
+By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
+\end{figure}
+
+%(up to ~2500 words including references, notes and captions
+% abstract
+% introductory paragraph
+%up to four figures or tables
+%about 30 references
 

### 2019-04-07 medians_REAL_0.tex

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\

### 2019-04-07 medians_REAL_3.tex

@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
+\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} &  &
+\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} &  &
  \\

### 2019-04-09 paper-short.tex
revising

@@
 
-\title{???? The Structure of Human Language Reflects Limitations of Working Memory}
+\title{The Structure of Human Language Reflects Limitations of Working Memory}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
+%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+
 Human languages express hierarchical thoughts into sentences, i.e., linear strings of words.
-We show that the ways languages do this reflect an underlying optimization principle:
-Human languages have evolved towards robustness against limitations in human working memory.
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
+Explaining the variation in the ways languages use to order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
+We show that the ways in which languages order information reflect an underlying optimization principle: Human languages have evolved towards robustness against limitations in human working memory.
 
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
-Explaining the variation in the ways languages  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
-We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
+%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
+The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\citep{berwick1984grammatical,hawkins1994performance};
+however, it has been hard to test this idea without ad-hoc modeling assumptions.
+%Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?
 
-Theoretical linguists have long suggested that the structure of human language reflects a need for efficient processing under resource limitations~\citep{berwick1984grammatical,hawkins1994performance} (also general efficiency mentioned outside functional work, e.g.
-\citep{chomsky2005three,hauser2002faculty}).
-
-We directly test this hypothesis using large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.
+We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.
 
@@
 We consider the question of how many bits of information this listener has to carry from one instant to the next.
-We expect that a listener's ease of comprehension will be modulated by the amount of memory held about previous context:
+Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
+Thus, we expect that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
 A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load.
+We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
 
 %We use information theory to quantify the tradeoff between memory load and ease of understanding.
-Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted:
+Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context:
 %Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
@@
 A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
-We denote a listener's memory representation at time $t$ by $l_t$, which is some -- generally imperfect -- summary of the context $X_{<t}$.
+We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
 The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
@@
 \begin{equation}
-	S :=	-\log p(x_t|l_t)
+	S :=	-\log p(x_t|L_t)
 \end{equation}
 upon hearing the next word $x_t$.
+One should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.
 %We quantify memory as the number of bits about prior input that a listener represents.
-
-We prove a theorem relating memory to the experienced comprehension difficulty $S$.
+We prove a theorem describing how the shape of this tradeoff relates to language structure. %relating memory to the experienced comprehension difficulty $S$.
 Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
@@
 These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences.
-These random grammars apply consistent grammatical rules similar to human language.
+These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
 
 Results are show in Figure~\ref{fig:results}.
-Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.001$).
+Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.0001$).
+We replicated this result with different methods of estimating chance (see SI).
 
@@
 
-Beyond human language, our theorem applies to other stochastic processes.
+Things to perhaps mention:
+
+- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?
+
+- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \citep{chomsky2005three,hauser2002faculty}.
+
+- Beyond human language, our theoretical result applies to other stochastic processes.
 To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.

### 2019-04-16 paper-short.tex
start working on abstract

@@
 
+\begin{abstract}
+  %  the Nature template:
+  % 1-2 sentences basic introduction, for anyone
+  Each of the \~6000 natural languages currently spoken in the world embodies a different system for encoding hierarchically-structured thoughts into linear strings of words.
+  By characterizing the commonalities and differences among these systems, we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
+  % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
+  In linguistics and cognitive science, there have been two
+
+  % 1 sentence general problem
+  % 1 sentence here we show
+  Here we show that the order of words across human languages is structured to enable efficient transfer of information under the constraint that language must be comprehended and produced by agents with limited short-term memory capacities.
+  % 2-3 sentences how main result adds to knowledge
+  Using concepts from information theory, we prove a theorem showing that any efficient communication system operating under memory constraints will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
+  We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
+  Next, we directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO real languages and in random baseline languages, finding that the real languages require less memory.
+  % 1-2 sentences general context
+
+  % 2-3 sentences broader perspective
+  Our results suggest a view of natural language as a maximally efficient code for communication, subject to the particular information-processing constraints of the human mind. 
+\end{abstract}
+
 %Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
 
-Human languages express hierarchical thoughts into sentences, i.e., linear strings of words.
+Human languages express hierarchical thoughts into sentences, i.e. linear strings of words.
 Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
 Explaining the variation in the ways languages use to order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
-We show that the ways in which languages order information reflect an underlying optimization principle: Human languages have evolved towards robustness against limitations in human working memory.
+We show that the ways in which languages order information reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
 

### 2019-04-17 paper-short.tex
revising

--- 2019-04-09 paper-short.tex
@@
 
+%TODO
+%Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
+\begin{equation}
+	\operatorname{I}_t := \operatorname{I}[X_t, X_0 | X_1, \dots, X_{t-1}] = \operatorname{H}[X_t|X_1, \dots, X_{t-1}] - \operatorname{H}[X_t|X_0, \dots, X_{t-1}] 
+\end{equation}
+%We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
+
+%
+%\begin{figure}
+%\includegraphics[width=\textwidth]{figures/theorem.png}
+%	\caption{Left: }\label{fig:theorem}
+%\end{figure}
+%
+
+%\caption{A speaker (left) utters a sentence. A comprehender (right) maintains a noisy representation of prior words, and incurs surprisal upon hearing the next word.}
+
 Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
@@
 For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
-These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences.
+These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}).
 These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/grammars.png}
+	\caption{Grammars specify the rules by which languages linearize underlying hierarchical structures into strings of words.
+In this example, the first grammar corresponds to English; the others are counterfactual.
+(\textbf{TODO} this draft is currently just copied from the Greenberg Universals paper, probably have to make it look a bit different.
+Also, adapt it so it looks more like `hierarchical structure' than `dependency graphs'.
+And add subjects.) }\label{fig:grammars}
+\end{figure}
 
@@
 To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.
+
+- Our results provide a new perspective on several previous results: (1) DLM, (2) DLT, (3) decay of MI over distance.
 
@@
 
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
+%
+%
+%\begin{figure}
+%\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
+%\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
+%	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
 Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
@@
 By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
+%\end{figure}
+%
 

### 2019-04-17 paper-short.tex
add some comments

@@
   Next, we directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO real languages and in random baseline languages, finding that the real languages require less memory.
+	% TODO i'm wondering whether we should first mention the big result, and then say, well it also subsumes previously proposed constraints.
   % 1-2 sentences general context
@@
 
-Human languages express hierarchical thoughts into sentences, i.e. linear strings of words.
+Human languages express hierarchically-structured thoughts into sentences, i.e. linear strings of words.
 Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
@@
 
-We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
+In order to test this hypothesis, it is necessary to consider an individual listener's language comprehension process.
+%We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
+During comprehension, this listener is processing a stream of words uttered by an interlocutor.
 During comprehension, this listener will represent some amount of information about past words in their short-term memory.
@@
 A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load.
+(TODO need be clearer here.
+need to make really explicit and obvious why memory improves understanding.)
 We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
+
+Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
 
@@
 %This quantity is known to be a strong linear predictor of human processing effort.
+
+TODO Make sure letters are intutive.
 
@@
 %TODO
-%Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
+Let $I_t$ be the Conditional Mutual Information of two words $X_0, X_t$ conditioned on the intervening words:
 \begin{equation}
@@
 \end{equation}
+This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
+It is a statistical property of the language.
+Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.
+
+Consider a listener who invests $M$ bits of memory into representing past input.
+We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $M$.
+Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
+
+By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.
+
 %We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
+
+%TODO Underdiscussed: the theoretical result.
+
+%maybe multiple sentences in def of It?
+%explicit arcs making clear going back 1 word in the past, 2 words in the past? maybe matching between arcs and colors instead of the arrows?
+
+%define math, AND give people intuition.
 
@@
 Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
-We compared against chance by the following method.
+We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
 For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
-These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}).
-These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
+These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
+%These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
 
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/grammars.png}
-	\caption{Grammars specify the rules by which languages linearize underlying hierarchical structures into strings of words.
+%\begin{figure}
+%\includegraphics[width=\textwidth]{figures/grammars.png}
+%	\caption{Grammars specify the rules by which languages linearize underlying hierarchical structures into strings of words.
 In this example, the first grammar corresponds to English; the others are counterfactual.
@@
 And add subjects.) }\label{fig:grammars}
-\end{figure}
+%\end{figure}
+%
 
@@
 \end{figure}
+
+say s.th. about estimates of human working memory. only if the last word is the only thing do things not make a difference.
 

### 2019-04-22 paper-short.tex
revising

@@
 \newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
 
@@
 \newtheorem{lemma}[theorem]{Lemma}
+
+\definecolor{Red}{RGB}{255,0,0}
+\newcommand{\red}[1]{\textcolor{Red}{#1}}
+\newcommand{\jd}[1]{\textcolor{Red}{[jd: #1]}} 
+\newcommand{\mhahn}[1]{\textcolor{Red}{[mhahn: #1]}} 
+\newcommand{\rljf}[1]{\textcolor{Red}{[rljf: #1]}}
 
@@
 
-Human languages express hierarchically-structured thoughts into sentences, i.e. linear strings of words.
+Human languages express hierarchically-structured thoughts into sentences, i.e. linear strings of words. \jd{too complicated for the general audience}
 Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
@@
 During comprehension, this listener is processing a stream of words uttered by an interlocutor.
-During comprehension, this listener will represent some amount of information about past words in their short-term memory.
-We consider the question of how many bits of information this listener has to carry from one instant to the next.
+This listener will represent information about past words in their short-term memory.
+We quantify the amount of memory resources that are used by the number of bits.
+We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
 Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
@@
 need to make really explicit and obvious why memory improves understanding.)
-We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
+We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load. \jd{why?}
+
+For example,
+
+- example? something like `John threw the trash out'
+- something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.
+- center embedding: `The physician who consulted the cardiologist checked the files in her office' -- (Fedorenko, Gibson, and Rohde (2006))
 
 Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
+
+A maximally efficient listener will incur surprisal exactly at the entropy rate of the language, i.e., at the rate at which the language provides information.
 
@@
 
+- can refer to general things such as Friston
+
 TODO Make sure letters are intutive.
@@
 upon hearing the next word $x_t$.
-One should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.
+One should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}
+
 %We quantify memory as the number of bits about prior input that a listener represents.
 We prove a theorem describing how the shape of this tradeoff relates to language structure. %relating memory to the experienced comprehension difficulty $S$.
-Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
+%Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
+This theorem captures formally that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
 
@@
 By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.
+
+An immediate consequence of this theoretical result is a general relation between memory and locality:
+A language will enable more efficient tradeoffs if $I_t$ decays faster as $t$ grows, which corresponds to stronger concentration of predictive information in the recent past.
 

### 2019-04-22 paper-short.tex
figure

@@
 
+- discontinuous idioms
 - example? something like `John threw the trash out'
+- particle verb, with completely idiomatic meaning.
+`give up' `the climber gave the attempt to reach the peak of Mount Everest up' could have figure `he gave the attempt to sculpture out of marble up'.
+understanding what the word means in contet requires understanding of the past.
+- 
 - something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.

### 2019-04-22 paper-short.tex
some prose changes

@@
   % 1-2 sentences basic introduction, for anyone
-  Each of the \~6000 natural languages currently spoken in the world embodies a different system for encoding hierarchically-structured thoughts into linear strings of words.
+  Each of the \~6000 natural languages currently spoken in the world embodies a different system for encoding thoughts into linear strings of words.
   By characterizing the commonalities and differences among these systems, we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
@@
   Here we show that the order of words across human languages is structured to enable efficient transfer of information under the constraint that language must be comprehended and produced by agents with limited short-term memory capacities.
+  We directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO real languages and in random baseline languages, finding that the real languages require less memory.
   % 2-3 sentences how main result adds to knowledge
@@
   We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
-  Next, we directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO real languages and in random baseline languages, finding that the real languages require less memory.
 	% TODO i'm wondering whether we should first mention the big result, and then say, well it also subsumes previously proposed constraints.
@@
 
-Human languages express hierarchically-structured thoughts into sentences, i.e. linear strings of words. \jd{too complicated for the general audience}
+Human languages express thoughts into sentences, i.e. linear strings of words. \jd{too complicated for the general audience}
+For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
+% Run with this example
+
+Each language embodies a different strategy for encoding thoughts into strings of words.
+
 Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.

### 2019-04-23 paper-short.tex
revising

@@
 Human languages express thoughts into sentences, i.e. linear strings of words. \jd{too complicated for the general audience}
-For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
+%For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
 % Run with this example
@@
 We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
+
 Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
-Thus, we expect that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
+Understanding what a word means often requires consideration of context. 
+Consider the sentence `The climber gave the attempt to reach the peak up.'
+Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
+In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
+For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
 A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
-(TODO need be clearer here.
-need to make really explicit and obvious why memory improves understanding.)
-We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load. \jd{why?}
+We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
 
-For example,
-
-- discontinuous idioms
-- example? something like `John threw the trash out'
-- particle verb, with completely idiomatic meaning.
+%
+%For example, 
+%- discontinuous idioms
+%- example? something like `John threw the trash out'
+%- particle verb, with completely idiomatic meaning.
 `give up' `the climber gave the attempt to reach the peak of Mount Everest up' could have figure `he gave the attempt to sculpture out of marble up'.
 understanding what the word means in contet requires understanding of the past.
-- 
-- something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.
-- center embedding: `The physician who consulted the cardiologist checked the files in her office' -- (Fedorenko, Gibson, and Rohde (2006))
+%- something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.
+%- center embedding: `The physician who consulted the cardiologist checked the files in her office' -- (Fedorenko, Gibson, and Rohde (2006))
+%Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
 
-Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
-
-A maximally efficient listener will incur surprisal exactly at the entropy rate of the language, i.e., at the rate at which the language provides information.
+%A maximally efficient listener will incur surprisal exactly at the entropy rate of the language, i.e., at the rate at which the language provides information.
 
@@
 
-- can refer to general things such as Friston
-
-TODO Make sure letters are intutive.
+%- can refer to general things such as Friston
+%TODO Make sure letters are intutive.
 
@@
 upon hearing the next word $x_t$.
-One should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}
+For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
+Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.
+%Since $L_t$ is some summary representation, 
+%Therefore, ne should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}
+
+%For instance, in `The climber gave the attempt to reach the peak up', the word \emph{up} is less surprising to a listener who has maintained some information about the word \emph{gave} in their memory.
+Thus, there is a tradeoff between memory and surprisal:
+A listener can achieve lower levels of processing effort, at the expense of increased memory load.
 
@@
 Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
-
 By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.

### 2019-04-25 paper-short.tex
revising

@@
 
-In order to test this hypothesis, it is necessary to consider an individual listener's language comprehension process.
-%We consider language understanding from the perspective of a listener processing a stream of words uttered by an interlocutor.
-During comprehension, this listener is processing a stream of words uttered by an interlocutor.
-This listener will represent information about past words in their short-term memory.
-We quantify the amount of memory resources that are used by the number of bits.
-We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
-
-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
-Understanding what a word means often requires consideration of context. 
-Consider the sentence `The climber gave the attempt to reach the peak up.'
-Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
-In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
-For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
-A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
-We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
-
-%
-%For example, 
-%- discontinuous idioms
-%- example? something like `John threw the trash out'
-%- particle verb, with completely idiomatic meaning.
-`give up' `the climber gave the attempt to reach the peak of Mount Everest up' could have figure `he gave the attempt to sculpture out of marble up'.
-understanding what the word means in contet requires understanding of the past.
-%- something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.
-%- center embedding: `The physician who consulted the cardiologist checked the files in her office' -- (Fedorenko, Gibson, and Rohde (2006))
-%Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
-
-%A maximally efficient listener will incur surprisal exactly at the entropy rate of the language, i.e., at the rate at which the language provides information.
-
-%We use information theory to quantify the tradeoff between memory load and ease of understanding.
-Experimental research has established that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context:
-%Experimental research has firmly established Surprisal as an accurate metric of human processing effort in language comprehesion.
-%The surprisal of a word in context is the degree to which it cannot be predicted:
+\mhahn{smooth out}We consider an individual listener's language comprehension process, processing a stream of words uttered by an interlocutor.
+Experimental research has established that listeners form probabilistic expectations about the upcoming words.
+Consider example...
+Human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
+Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
+For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
 \begin{equation}\label{eq:surp}
@@
 where $x_i$ is the $i$-th word in a given utterance (or discourse).
-%This quantity is known to be a strong linear predictor of human processing effort.
 
-%- can refer to general things such as Friston
-%TODO Make sure letters are intutive.
-
+In reality, a listener will have bounded capacity to represent prior context.
 A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
 We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
-The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
-This listener will experience processing effort
-\begin{equation}
+This listener will experience surprisal
+\begin{equation}\label{eq:surp-listener}
 	S :=	-\log p(x_t|L_t)
@@
 upon hearing the next word $x_t$.
-For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
-Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.
-%Since $L_t$ is some summary representation, 
-%Therefore, ne should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}
+%The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
+%For a listener who invests less bits into representing information $L_t$ about prior context, $L_t$ will be a more coarse-grained representation of past observations.
+%Thus, the predictions made by such a listener will be on average less accurate, and such a listener will incur greater processing effort $S$.
 
-%For instance, in `The climber gave the attempt to reach the peak up', the word \emph{up} is less surprising to a listener who has maintained some information about the word \emph{gave} in their memory.
-Thus, there is a tradeoff between memory and surprisal:
+Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents less information about past input.
+Consequently, there is a tradeoff between memory and surprisal:
 A listener can achieve lower levels of processing effort, at the expense of increased memory load.
+
+Formally, we quantify memory as the number of bits about prior input that a listener represents.
 
@@
 This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
-It is a statistical property of the language.
+It is a statistical property of the language, and can be estimated from large-scale text data.
 Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.
@@
 An immediate consequence of this theoretical result is a general relation between memory and locality:
-A language will enable more efficient tradeoffs if $I_t$ decays faster as $t$ grows, which corresponds to stronger concentration of predictive information in the recent past.
+A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.
 
@@
 - Our results provide a new perspective on several previous results: (1) DLM, (2) DLT, (3) decay of MI over distance.
+
+-This listener will represent information about past words in their short-term memory.
+We quantify the amount of memory resources that are used by the number of bits.
+We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
+
+-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
+Understanding what a word means often requires consideration of context. 
+Consider the sentence `The climber gave the attempt to reach the peak up.'
+Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
+In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
+For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
+A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
+We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
+
+%
+%For example, 
+%- discontinuous idioms
+%- example? something like `John threw the trash out'
+%- particle verb, with completely idiomatic meaning.
+`give up' `the climber gave the attempt to reach the peak of Mount Everest up' could have figure `he gave the attempt to sculpture out of marble up'.
+understanding what the word means in contet requires understanding of the past.
+%- something with a relative clause `The child that was playing on the street had a red ball' -- should not forget the `child'.
+%- center embedding: `The physician who consulted the cardiologist checked the files in her office' -- (Fedorenko, Gibson, and Rohde (2006))
+%Make clear that Understanding requires knowing the past. (introduce prediction earlier?) Example? Then formally introduce prediction.
+
+%A maximally efficient listener will incur surprisal exactly at the entropy rate of the language, i.e., at the rate at which the language provides information.
+
+%We use information theory to quantify the tradeoff between memory load and ease of understanding.
+%This quantity is known to be a strong linear predictor of human processing effort.
+
+%- can refer to general things such as Friston
+%TODO Make sure letters are intutive.
+
+%Since $L_t$ is some summary representation, 
+%Therefore, ne should expect that processing effort $S$ increases for a listener who invests less bits into representing information about prior context.\jd{why? this is important.}
+
+%For instance, in `The climber gave the attempt to reach the peak up', the word \emph{up} is less surprising to a listener who has maintained some information about the word \emph{gave} in their memory.
 

### 2019-04-25 paper-short.tex
revising

@@
 This theorem captures formally that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures-gdrive/theorem.png}
+	\caption{The statement of the theoretical result.}\label{fig:results}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures-gdrive/tradeoff.png}
+	\caption{}\label{fig:results}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures-gdrive/counterfactual-languages.png}
+	\caption{Estimating chance by constructing counterfactual grammars and languages.}\label{fig:results}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{figures-gdrive/mi-distance.png}
+	\caption{}\label{fig:results}
+\end{figure}
 

### 2019-04-28 paper-short.tex
revising

--- 2019-04-25 paper-short.tex
@@
-\documentclass[11pt,letterpaper]{article}
+\documentclass[12pt]{article}
 
 \usepackage{showlabels}
-%\usepackage{fullpage}
+\usepackage{fullpage}
 \usepackage{pslatex}
@@
 \usepackage{rotating}
-\usepackage{natbib}
 \usepackage{amssymb}
@@
 \frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
 
@@
 
+\usepackage{scicite}
+
+\usepackage{times}
+
+\topmargin 0.0cm
+\oddsidemargin 0.2cm
+\textwidth 16cm 
+\textheight 21cm
+\footskip 1.0cm
+
+\newenvironment{sciabstract}{%
+\begin{quote} \bf}
+{\end{quote}}
+
+\date{}
+
+%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%
+
 \begin{document}
+
+\baselineskip24pt
 
@@
 
-%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+%Since short-term memory of this kind is known to be highly limited in capacity \cite{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+
+%The manuscript should start with a brief introduction describing the paper’s significance.
+%The introduction should provide sufficient background information to make the article intelligible to readers in other disciplines, and sufficient context that the significance of the experimental findings is clear.
+%Technical terms should be defined. Symbols, abbreviations, and acronyms should be defined the first time they are used.
+%All tables and figures should be cited in numerical order.
+All data must be shown either in the main text or in the Supplementary Materials or must be available in an established database with accession details provided in the acknowledgements section.
+References to unpublished materials are not allowed to substantiate significant conclusions of the paper.
+
+% introductory paragraph
 
@@
 %We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\citep{berwick1984grammatical,hawkins1994performance};
+The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance};
 however, it has been hard to test this idea without ad-hoc modeling assumptions.
@@
 
-We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\citep{nivre-universal-2017}.
+We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}.
 
-\mhahn{smooth out}We consider an individual listener's language comprehension process, processing a stream of words uttered by an interlocutor.
-Experimental research has established that listeners form probabilistic expectations about the upcoming words.
-Consider example...
-Human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{figures-gdrive/communication.png}
+	\caption{Memory and Language Comprehension.
+(1) A speaker produces a sequence of words.
+(2) A listener maintains a representation of the words received so far.
+The listener can represent these at higher (left) or lower (right) levels of precision.
+(3) Throughout communication, the listener generates probabilistic expectations about the next word.
+Higher precision of memory representations leads to more precise predictions.
+(4) When receiving the next word, the listener incurs surprisal depending on the predictions.
+Higher levels of fidelity in memory lead to lower surprisal on average.
+}\label{fig:communication}
+\end{figure}
+
+We consider the process of language comprehension in a person listening to an interlocutor, processing a stream of words uttered by the other person (Figure~\ref{fig:communication}).
+Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words.
+%, and (3) words are easier to process when they are highly expected in context.
+Importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
+More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.
+The psycholinguistic literature has established that humans differ in their capacity to store such representations, and that capacity limitations can create difficulties in language comprehension.
+
+Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
 Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
@@
 where $x_i$ is the $i$-th word in a given utterance (or discourse).
-
-In reality, a listener will have bounded capacity to represent prior context.
+%
+%In reality, a listener will have bounded capacity to represent prior context.
 A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
-We denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
+As in Figure~\ref{fig:communication}, we denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
 This listener will experience surprisal
@@
 
-Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents less information about past input.
+Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
 Consequently, there is a tradeoff between memory and surprisal:
@@
 
-Formally, we quantify memory as the number of bits about prior input that a listener represents.
+%Formally, we quantify memory as the number of bits about prior input that a listener represents.
 
@@
 
+\begin{figure}
+	(a)
+\includegraphics[width=0.4\textwidth]{figures-gdrive/mi-distance.png}
+	(b)
+\includegraphics[width=0.2\textwidth]{figures-gdrive/theorem.png}
+	(c)
+\includegraphics[width=0.2\textwidth]{figures-gdrive/tradeoff.png}
+	\caption{
+		(a) Conditional mutual information $I_t$ describes how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
+		(b) Theoretical result: We plot It (top) and tIt (bottom) as functions of t.
+A listener using B bits memory (bottom) to represent prior observations will incur at least A bits of extra surprisal beyond the entropy rate (top).
+		(c) 
+}\label{fig:theorem}
+\end{figure}
+
 %TODO
@@
 \end{equation}
-This quantity measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
+This quantity, visualized in Figure~\ref{fig:theorem} (a), measures how much predictive information is provided by the next word $X_t$ by the word $t$ steps in the past.
 It is a statistical property of the language, and can be estimated from large-scale text data.
-Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$.
-
-Consider a listener who invests $M$ bits of memory into representing past input.
-We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $M$.
+Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$ (Figure~\ref{fig:theorem} (b)):
+Consider a listener who invests $B$ bits of memory into representing past input.
+We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $B$.
 Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
@@
 %We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
-
-%TODO Underdiscussed: the theoretical result.
-
 %maybe multiple sentences in def of It?
-%explicit arcs making clear going back 1 word in the past, 2 words in the past? maybe matching between arcs and colors instead of the arrows?
-
 %define math, AND give people intuition.
 
-%
-%\begin{figure}
-%\includegraphics[width=\textwidth]{figures/theorem.png}
-%	\caption{Left: }\label{fig:theorem}
-%\end{figure}
-%
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{figures-gdrive/counterfactual-languages.png}
+	\caption{Estimating chance by constructing counterfactual grammars and languages.}\label{fig:grammars}
+\end{figure}
 
-%\caption{A speaker (left) utters a sentence. A comprehender (right) maintains a noisy representation of prior words, and incurs surprisal upon hearing the next word.}
-
-Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\citep{hochreiter-long-1997}.
+Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\cite{hochreiter-long-1997}.
 We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
 For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
-These counterfactual versions use randomly created grammar for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
+These counterfactual versions use randomly created grammars for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
 %These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
@@
 
-- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?
+%- Potentially cite \cite{PhysRevLett.115.098701} as a reference saying that physical systems have bounded memory?
 
-- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \citep{chomsky2005three,hauser2002faculty}.
+- Computational efficiency has been mentioned in mainstream syntax, though it's not so clear what it means there \cite{chomsky2005three,hauser2002faculty}.
 
@@
 
--This listener will represent information about past words in their short-term memory.
-We quantify the amount of memory resources that are used by the number of bits.
-We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
-
--Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
-Understanding what a word means often requires consideration of context. 
-Consider the sentence `The climber gave the attempt to reach the peak up.'
-Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
-In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
-For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
-A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
-We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
+%
+%-This listener will represent information about past words in their short-term memory.
+%We quantify the amount of memory resources that are used by the number of bits.
+%We consider the question of how many bits of information this listener has to carry from one instant to the next.  %\jd{why?}
+%
+%-Human language is well known to have a large amount of context-dependence in the meanings and forms of words.
+%Understanding what a word means often requires consideration of context. 
+%Consider the sentence `The climber gave the attempt to reach the peak up.'
+%Understanding the word \emph{up} requires establishing that it is related to the word \emph{gave} -- neither of them has the meaning of `give up' by itself.
+%In order to understand this sentence correctly, a listener has to store some information about the occurrence of the word \emph{gave} until reaching \emph{up}.
+%For language comprehension, this means that a listener's ease of comprehension will be modulated by how much memory they invest into representing information about past context:
+%A listener who remembers more of the past context should have less difficulty understanding, at the cost of increased working memory load. 
+%We thus expect there to be a tradeoff between a listener's ease of language understanding and their working memory load.
+%
 
@@
 %Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
+%Since short-term memory of this kind is known to be highly limited in capacity \cite{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
+%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \cite{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
 %
@@
 %second, because this presupposes a computatioal model of processing efficiency.
-%Large-scale data has recently become available in the form of corpora annotated with underlying hierarchical structures across dozens of languages~\citep{nivre-universal-2017}.
+%Large-scale data has recently become available in the form of corpora annotated with underlying hierarchical structures across dozens of languages~\cite{nivre-universal-2017}.
 
@@
 %The minimal amount of information stored is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
-%%related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \citep{PhysRevLett.115.098701}
+%%related to work in physics on the ability of physical systems to represent memory, representing more bits costs more memory \cite{PhysRevLett.115.098701}
 %We note that there is a tradeoff between average memory, $\E[S]$ and memory $H[L_t]$: A listener who invests more memory resources to create representations of the past will incur lower average surprisal.
@@
 
-\bibliographystyle{apalike}
+\bibliographystyle{Science}
 
@@
 
+\appendix{SI}
+
+\section{Materials and Methods}
+
+\subsection{Experimental Design}
+\subsection{Statistical Analysis}
+
 \end{document}

### 2019-05-02 paper-short.tex
minor edits and added some comments to paper

@@
   % 1-2 sentences basic introduction, for anyone
-  Each of the \~6000 natural languages currently spoken in the world embodies a different system for encoding thoughts into linear strings of words.
-  By characterizing the commonalities and differences among these systems, we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
+  Each of the roughly 6000 natural languages currently spoken in the world embodies a different system for encoding thoughts into linear strings of words.
+  By characterizing the commonalities and differences among these systems we can gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence. 
   % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
-  In linguistics and cognitive science, there have been two
+  In linguistics and cognitive science, there have been two \jd{unfinished sentence}
 
@@
   Here we show that the order of words across human languages is structured to enable efficient transfer of information under the constraint that language must be comprehended and produced by agents with limited short-term memory capacities.
-  We directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO real languages and in random baseline languages, finding that the real languages require less memory.
+  We directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO \jd{random TODO} real languages and in random baseline languages, finding that the real languages require less memory.
   % 2-3 sentences how main result adds to knowledge
@@
 
-Human languages express thoughts into sentences, i.e. linear strings of words. \jd{too complicated for the general audience}
+Human languages express thoughts into sentences, i.e. linear strings of words. 
 %For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
 % Run with this example
+\jd{is the commented out example sth that will make it in? it does seem like an example would be good here}
 
-Each language embodies a different strategy for encoding thoughts into strings of words.
+Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}
 
 Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
-Explaining the variation in the ways languages use to order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (CITE).
-We show that the ways in which languages order information reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
+Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
+We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
 
@@
 The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance};
-however, it has been hard to test this idea without ad-hoc modeling assumptions.
+however, it has been hard to test this idea without ad-hoc modeling assumptions. \jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
 %Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?
 
-We directly test this hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}.
+We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}.
 
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{figures-gdrive/communication.png}
-	\caption{Memory and Language Comprehension.
+\centering
+\includegraphics[width=0.9\textwidth]{figures-gdrive/communication.png}
+	\caption{Memory and language comprehension.
 (1) A speaker produces a sequence of words.
@@
 Higher levels of fidelity in memory lead to lower surprisal on average.
-}\label{fig:communication}
+\jd{what do the numbers (1) (2) etc refer to?
+also: image is blurry, at least on my screen}}
+	\label{fig:communication}
 \end{figure}
 
-We consider the process of language comprehension in a person listening to an interlocutor, processing a stream of words uttered by the other person (Figure~\ref{fig:communication}).
-Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words.
+We consider the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
+Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words \jd{CITE}.
 %, and (3) words are easier to process when they are highly expected in context.
@@
 More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.
-The psycholinguistic literature has established that humans differ in their capacity to store such representations, and that capacity limitations can create difficulties in language comprehension.
+The psycholinguistic literature has established that humans differ in their capacity to store such representations \jd{CITE}, and that capacity limitations can create difficulties in language comprehension \jd{CITE}.
 
-Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context.
-Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times.
+Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context \jd{CITE roger}.
+Surprisal is known to be a strong linear predictor of processing effort as measured by per-word log-transformed reading times.
 For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
@@
 Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
+\jd{include an additional sentence on the DPI?}
 Consequently, there is a tradeoff between memory and surprisal:
-A listener can achieve lower levels of processing effort, at the expense of increased memory load.
+A listener can achieve lower levels of processing effort at the expense of increased memory load.
 
@@
 		(a) Conditional mutual information $I_t$ describes how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
-		(b) Theoretical result: We plot It (top) and tIt (bottom) as functions of t.
-A listener using B bits memory (bottom) to represent prior observations will incur at least A bits of extra surprisal beyond the entropy rate (top).
-		(c) 
+		(b) Theoretical result: We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$.
+A listener using $B$ bits memory (bottom) to represent prior observations will incur at least $A$ bits of extra surprisal beyond the entropy rate (top).
+		(c) \jd{FILL OUT}
 }\label{fig:theorem}
@@
 It is a statistical property of the language, and can be estimated from large-scale text data.
-Our theoretical results describes how the tradeoff between memory and comprehension difficulty relates to $I_t$ (Figure~\ref{fig:theorem} (b)):
+Our theoretical results describe how the tradeoff between memory and comprehension difficulty relates to $I_t$ (Figure~\ref{fig:theorem} (b)):
 Consider a listener who invests $B$ bits of memory into representing past input.
-We then consider the smallest $T$ such the area under the curve of $t I_t$, to the left of $T$, has size $B$.
+We then consider the smallest $T$ such that the area under the curve of $t I_t$, to the left of $T$, has size $B$.
 Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{figures-gdrive/counterfactual-languages.png}
-	\caption{Estimating chance by constructing counterfactual grammars and languages.}\label{fig:grammars}
+\centering
+\includegraphics[width=\textwidth]{figures-gdrive/counterfactual-languages.png}
+	\caption{Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me.
+also looks like there's sth in the background on the left?}}\label{fig:grammars}
 \end{figure}
@@
 These counterfactual versions use randomly created grammars for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
+\jd{refer to SI here and make sure to include a lot more information in the SI about how the grammars were created}
 %These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
@@
 
-Results are show in Figure~\ref{fig:results}.
+Results are shown in Figure~\ref{fig:results}.
 Human languages lead to a more efficient tradeoff than expected by chance ($p < 0.0001$).
@@
 \includegraphics[width=\textwidth]{figures/full-results.png}
-	\caption{}\label{fig:results}
+	\caption{\jd{FILL OUT; make less blurry}}\label{fig:results}
 \end{figure}

### 2019-05-02 paper-short.tex
from meeting

@@
 
+\mhahn{all three of the next are like first sentences}
 Human languages express thoughts into sentences, i.e. linear strings of words. 
@@
 
-Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}
+Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}\mhahn{make this sentence nonredundant}
+\mhahn{give an example}
 
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb.
+These strategies can be seen as sets of rules that differ in the way ...
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \mhahn{make more explicit here}.
 Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
@@
 The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance};
-however, it has been hard to test this idea without ad-hoc modeling assumptions. \jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
+A lot of progress has been mdade on understanding. Evidence for the idea comes from various different frameworks.
+The only problem with these is it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tied to specific parsing algorithms.
+Maybe citation that explicitly criticizes thigns like DLM for this?
+%however, it has been hard to test this idea without ad-hoc modeling assumptions.
+\jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
+\mhahn{this is a strong critique. not clear what kinds of ad-hoc modeling this is about}
+e.g. previously based on some grammatical formalism and a linking to ideas about parsing algorithms.
+
 %Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?
+For the first time, we test this hypothesis by applying...
 
-We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}.
+We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages
+\mhahn{Emphasize minimality of assumptions.}
+\mhahn{Say more about the steps that this will entail, so the reader won't get lost}
 
@@
 \jd{what do the numbers (1) (2) etc refer to?
-also: image is blurry, at least on my screen}}
+also: image is blurry, at least on my screen}\mhahn{make reference to $L_t$ inside the caption.
+you and the surprisals visually better aligned.
+maybe can color-code?
+}}
 	\label{fig:communication}
@@
 
+We begin by considering the process of language understanding.
+\mhahn{this is too sudden, why still unclear}
 We consider the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
@@
 \end{equation}
+\mhahn{S= but then S also appears below}
 where $x_i$ is the $i$-th word in a given utterance (or discourse).
@@
 
-Information theory, specifically the Data Processing Inequality, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
+Information theory, specifically the Data Processing Inequality \mhahn{add reference}, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
 \jd{include an additional sentence on the DPI?}
@@
 A listener can achieve lower levels of processing effort at the expense of increased memory load.
+
+\mhahn{reader will have lost the plot by now. strogn promise in the beginning}
 
@@
 	\caption{
-		(a) Conditional mutual information $I_t$ describes how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
-		(b) Theoretical result: We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$.
+		(a) Conditional mutual information $I_t$ captures how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
+		(b) Theoretical result:\mhahn{cofusing.
+the result is not the plotting} We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$.
 A listener using $B$ bits memory (bottom) to represent prior observations will incur at least $A$ bits of extra surprisal beyond the entropy rate (top).
-		(c) \jd{FILL OUT}
+\mhahn{there is t and T.
+COnfusing}
+		(c) \jd{FILL OUT} \mhahn{make the colors in (c) difefrenfr from this in (b).
+Do want to make clear what the mapping from (b) to (c) is.
+Maybe can even have a pair of the two (b) curves, and then show what they map on.}
 }\label{fig:theorem}
@@
 We then consider the smallest $T$ such that the area under the curve of $t I_t$, to the left of $T$, has size $B$.
-Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.
+Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.\mhahn{maybe result first, and then say what $T$ is}
 By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.
+\mhahn{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
 
@@
 A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.
+\mhahn{not taken up again. maybe this is together with locality discussion. after getting the tradeoff, would be good to go straight into the 54 languages}
 
@@
 \includegraphics[width=\textwidth]{figures-gdrive/counterfactual-languages.png}
-	\caption{Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me.
+	\caption{\mhahn{Can say more here to explain how it works} Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me.
 also looks like there's sth in the background on the left?}}\label{fig:grammars}
 \end{figure}
+
+\mhahn{move this where appropriate}
+, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}
 
@@
 These counterfactual versions use randomly created grammars for linearizing hierarchical structures into sentences (see Figure~\ref{fig:grammars}), akin to those found in human languages.
+See SI for details.
+\mhahn{maybe even more. hard concept to grasp. say more about the representation format, that it's as atheoretical as possible. encodes the ways heads are encoded to their dependents}
 \jd{refer to SI here and make sure to include a lot more information in the SI about how the grammars were created}
@@
 
+\mhahn{frame it the other way around. point is: they are NOT equally efficient. Afterthought: The only point where theyare equally efficient is at extremely low capacities}
 At extremely low memory capacities, real and baseline languages are equally efficient.
@@
 The point of divergence is given by that memory capacity that fully exhausts information provided by the immediately preceding word, that is, when the capacity of a Markov model in approximating language is exhausted.
+\mhahn{can one rephrase this?}
 

### 2019-05-02 paper-short.tex
revising

@@
 A lot of progress has been mdade on understanding. Evidence for the idea comes from various different frameworks.
-The only problem with these is it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tied to specific parsing algorithms.
-Maybe citation that explicitly criticizes thigns like DLM for this?
+However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\mhahn{TODO make this better}
+\mhahn{Maybe there is a citation that explicitly mentions this as a limitation of  thigngs like DLM}
 %however, it has been hard to test this idea without ad-hoc modeling assumptions.
@@
 
-We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages
+We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages.
 \mhahn{Emphasize minimality of assumptions.}
@@
 
-We begin by considering the process of language understanding.
-\mhahn{this is too sudden, why still unclear}
-We consider the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
+We begin this by considering the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
 Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words \jd{CITE}.
@@
 \begin{equation}\label{eq:surp}
-	-\log p(x_i|x_{1...i-1})
+	S :=	-\log p(x_i|x_{1...i-1})
 \end{equation}
-\mhahn{S= but then S also appears below}
 where $x_i$ is the $i$-th word in a given utterance (or discourse).
@@
 \begin{equation}\label{eq:surp-listener}
-	S :=	-\log p(x_t|L_t)
+	S_L :=	-\log p(x_t|L_t)
 \end{equation}
@@
 
-\mhahn{reader will have lost the plot by now. strogn promise in the beginning}
+\mhahn{reader will have lost the plot by now. strogn promise in the beginning. can solve this by foreshadowing}
 
@@
 		(a) Conditional mutual information $I_t$ captures how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
-		(b) Theoretical result:\mhahn{cofusing.
-the result is not the plotting} We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$.
+		(b) Here we illustrate our theoretical result.
+We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$.
 A listener using $B$ bits memory (bottom) to represent prior observations will incur at least $A$ bits of extra surprisal beyond the entropy rate (top).
@@
 
-An immediate consequence of this theoretical result is a general relation between memory and locality:
-A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.
-\mhahn{not taken up again. maybe this is together with locality discussion. after getting the tradeoff, would be good to go straight into the 54 languages}
+\mhahn{Alternative:}
+A listener will experience average surprisal at least
+\begin{equation}
+	H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t
+\end{equation}
+where $T$ is chosen to be the smallest $T$ for which $\sum_{t=1}^T t I_t \geq B$.
+\mhahn{End alternative.}
 
@@
 
-\mhahn{move this where appropriate}
-, with expert annotations of underlying hierarchical structures~\cite{nivre-universal-2017}
-
-Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages, using standard estimation methods for language modeling based on neural networks~\cite{hochreiter-long-1997}.
+Based on our theoretical result, we estimated the memory-surprisal tradeoff for 54 human languages.
+We estimated $I_t$ using standard estimation methods for language modeling based on neural networks~\cite{hochreiter-long-1997}.
 We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
+We draw on the Universal Dependncies project, which has constructed expert annotations of underlying hierarchical structures for dozens of languages~\cite{nivre-universal-2017}.
 For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.
@@
 
-\mhahn{frame it the other way around. point is: they are NOT equally efficient. Afterthought: The only point where theyare equally efficient is at extremely low capacities}
-At extremely low memory capacities, real and baseline languages are equally efficient.
-However, the curves soon diverge, and real language provides a more efficient tradeoff.
+The only point where real and baseline languages are equally efficient is at extremely low memory capacities.
 The point of divergence is given by that memory capacity that fully exhausts information provided by the immediately preceding word, that is, when the capacity of a Markov model in approximating language is exhausted.
@@
 \end{figure}
+
+- Our results provide a new perspective on several previous results: (1) DLM, (2) DLT, (3) decay of MI over distance.
+
+An immediate consequence of this theoretical result is a general relation between memory and locality:
+A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.
+\mhahn{not taken up again. maybe this is together with locality discussion. after getting the tradeoff, would be good to go straight into the 54 languages}
 
@@
 To the extent that they may be affected by bounded memory in computation, we predict they should also enable efficient tradeoffs between memory and surprisal.
-
-- Our results provide a new perspective on several previous results: (1) DLM, (2) DLT, (3) decay of MI over distance.
 

### 2019-05-02 paper-short.tex
revising

@@
 \newcommand{\jd}[1]{\textcolor{Red}{[jd: #1]}} 
-\newcommand{\mhahn}[1]{\textcolor{Red}{[mhahn: #1]}} 
+\newcommand{\note}[1]{\textcolor{Red}{[note: #1]}} 
 \newcommand{\rljf}[1]{\textcolor{Red}{[rljf: #1]}}
@@
 
-\mhahn{all three of the next are like first sentences}
+\note{all three of the next are like first sentences}
 Human languages express thoughts into sentences, i.e. linear strings of words. 
@@
 
-Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}\mhahn{make this sentence nonredundant}
-\mhahn{give an example}
+Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}\note{make this sentence nonredundant}
+\note{give an example}
 
 These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \mhahn{make more explicit here}.
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
 Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
@@
 A lot of progress has been mdade on understanding. Evidence for the idea comes from various different frameworks.
-However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\mhahn{TODO make this better}
-\mhahn{Maybe there is a citation that explicitly mentions this as a limitation of  thigngs like DLM}
+However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\note{TODO make this better}
+\note{Maybe there is a citation that explicitly mentions this as a limitation of  thigngs like DLM}
 %however, it has been hard to test this idea without ad-hoc modeling assumptions.
 \jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
-\mhahn{this is a strong critique. not clear what kinds of ad-hoc modeling this is about}
+\note{this is a strong critique. not clear what kinds of ad-hoc modeling this is about}
 e.g. previously based on some grammatical formalism and a linking to ideas about parsing algorithms.
@@
 We directly test the efficient processing hypothesis by applying a novel information-theoretic analysis to large-scale data from 54 languages.
-\mhahn{Emphasize minimality of assumptions.}
-\mhahn{Say more about the steps that this will entail, so the reader won't get lost}
+\note{Emphasize minimality of assumptions.}
+\note{Say more about the steps that this will entail, so the reader won't get lost}
 
@@
 \jd{what do the numbers (1) (2) etc refer to?
-also: image is blurry, at least on my screen}\mhahn{make reference to $L_t$ inside the caption.
+also: image is blurry, at least on my screen}\note{make reference to $L_t$ inside the caption.
 you and the surprisals visually better aligned.
@@
 
-Information theory, specifically the Data Processing Inequality \mhahn{add reference}, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
+Information theory, specifically the Data Processing Inequality \note{add reference}, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
 \jd{include an additional sentence on the DPI?}
@@
 
-\mhahn{reader will have lost the plot by now. strogn promise in the beginning. can solve this by foreshadowing}
+\note{reader will have lost the plot by now. strogn promise in the beginning. can solve this by foreshadowing}
 
@@
 A listener using $B$ bits memory (bottom) to represent prior observations will incur at least $A$ bits of extra surprisal beyond the entropy rate (top).
-\mhahn{there is t and T.
+\note{there is t and T.
 COnfusing}
-		(c) \jd{FILL OUT} \mhahn{make the colors in (c) difefrenfr from this in (b).
+		(c) \jd{FILL OUT} \note{make the colors in (c) difefrenfr from this in (b).
 Do want to make clear what the mapping from (b) to (c) is.
@@
 We then consider the smallest $T$ such that the area under the curve of $t I_t$, to the left of $T$, has size $B$.
-Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.\mhahn{maybe result first, and then say what $T$ is}
+Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.\note{maybe result first, and then say what $T$ is}
 By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.
-\mhahn{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
+\note{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
 
-\mhahn{Alternative:}
+\note{Alternative:}
 A listener will experience average surprisal at least
@@
 where $T$ is chosen to be the smallest $T$ for which $\sum_{t=1}^T t I_t \geq B$.
-\mhahn{End alternative.}
+\note{End alternative.}
 
@@
 \includegraphics[width=\textwidth]{figures-gdrive/counterfactual-languages.png}
-	\caption{\mhahn{Can say more here to explain how it works} Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me.
+	\caption{\note{Can say more here to explain how it works} Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me.
 also looks like there's sth in the background on the left?}}\label{fig:grammars}
@@
 See SI for details.
-\mhahn{maybe even more. hard concept to grasp. say more about the representation format, that it's as atheoretical as possible. encodes the ways heads are encoded to their dependents}
+\note{maybe even more. hard concept to grasp. say more about the representation format, that it's as atheoretical as possible. encodes the ways heads are encoded to their dependents}
 \jd{refer to SI here and make sure to include a lot more information in the SI about how the grammars were created}
@@
 The point of divergence is given by that memory capacity that fully exhausts information provided by the immediately preceding word, that is, when the capacity of a Markov model in approximating language is exhausted.
-\mhahn{can one rephrase this?}
+\note{can one rephrase this?}
 
@@
 A language will enable more efficient tradeoffs if $I_t$ decays very fast as $t$ grows, which corresponds to strong concentration of predictive information in the recent past.
-\mhahn{not taken up again. maybe this is together with locality discussion. after getting the tradeoff, would be good to go straight into the 54 languages}
 

### 2019-05-03 paper-short.tex
some changes

@@
   % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
-  In linguistics and cognitive science, there have been two \jd{unfinished sentence}
+  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \citep{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.
 
@@
   % 1 sentence here we show
-  Here we show that the order of words across human languages is structured to enable efficient transfer of information under the constraint that language must be comprehended and produced by agents with limited short-term memory capacities.
-  We directly show that languages are optimized to minimize memory usage by calculating lower bounds on the amount of memory resources required for language comprehension in TODO \jd{random TODO} real languages and in random baseline languages, finding that the real languages require less memory.
+  Here we give evidence for this view by showing that human languages structure information in time so as to enable efficient communication under the constraint that language must be comprehended and produced by agents with limited short-term memory resources.
+  We directly show that languages minimize memory usage by calculating information-theoretic bounds on the amount of memory resources required for language comprehension in 54 real languages and in random baseline languages, finding that the real languages require less memory.
   % 2-3 sentences how main result adds to knowledge
-  Using concepts from information theory, we prove a theorem showing that any efficient communication system operating under memory constraints will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
+  To explain how languages achieve this result, we prove a theorem showing that any efficient communication system operating under memory constraints will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
   We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
@@
   % 2-3 sentences broader perspective
-  Our results suggest a view of natural language as a maximally efficient code for communication, subject to the particular information-processing constraints of the human mind. 
+  Our results suggest that human languages can be modeled as solutions to the problem of maximally efficient communication subject to the particular information-processing constraints of the human mind.
 \end{abstract}

### 2019-05-10 paper-short.tex
add text

@@
   % 2-3 sentences how main result adds to knowledge
-  To explain how languages achieve this result, we prove a theorem showing that any efficient communication system operating under memory constraints will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
+  To explain how languages achieve this result, we prove a theorem showing that any efficient communication system operating under constraints on memory capacity will exhibit a property called information locality: symbols which predict each other should be produced close to each other in time.
   We show that information locality subsumes several principles of word order previously proposed in the linguistics and cognitive science literature, and explains a number of the previously-documented universal patterns in word order across languages.
@@
 %We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance};
-A lot of progress has been mdade on understanding. Evidence for the idea comes from various different frameworks.
+The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
+Here
+
+\rljf{While it's important to point out that our results are much less theory-bound than other models, I don't think the ``previous results are all tied to previous theories'' framing is strong.}
+
+%Typically, such theories are developed in terms of specific mechanistic models of human language processing; for example, \citet{hawkins1994performance} shows that word order patterns can result from optimization for efficient parsing under a particular parsing algorithm.
+
+A lot of progress has been made on understanding. Evidence for the idea comes from various different frameworks.
 However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\note{TODO make this better}
@@
 
-We begin this by considering the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
-Experimental research has established that such a listener (1) maintains information about the words received so far, and (2) forms probabilistic expectations about the upcoming words \jd{CITE}.
-%, and (3) words are easier to process when they are highly expected in context.
+We begin developing our model by considering the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
+Experimental research has established that such a listener (1) maintains information about the words received so far, (2) forms probabilistic expectations about the upcoming words, and (3) words are easier to process when they are highly expected in context \citep{hale2001probabilistic,levy2008expectation} \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}.
+
+Experimental research has shown that human processing effort on a given word is strongly predicted by its \textbf{surprisal}, the degree to which it cannot be predicted from the preceding context \citep{hale2001probabilistic,levy2008expectation}.
+Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times \citep{smith2013effect}.
+For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
+\begin{equation}\label{eq:surp}
+	S :=	-\log_2 p(x_i|x_{1...i-1})
+\end{equation}
+where $x_i$ is the $i$-th word in a given utterance (or discourse).
+In information-theoretic terms, surprisal is equivalent to the \textbf{information content} of a word, measured in bits, and the fact that the per-word reading time appears to be linear in the surprisal suggests that human language comprehension is best characterized in terms of a scalar rate of information processing in terms of bits per second.
+The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \citep{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \citep{egner2010expectation}, auditory and music perception \citep{agres2018information}, and motor planning \citep{wolpert2001motor}.
+
+However, Equation~\ref{eq:surp} can only represent processing difficulty for a listener with unbounded memory resources, who can accurately remember the complete history of previous words $x_{1...i-1}$ leading up to the current word $x_i$.
+More realistically, human listeners have fixed-capacity memory resources that are being deployed during language comprehension, meaning that their predictions about upcoming words may be inaccurate compared to Eq.~\ref{eq:surp}.
+If $m_{i-1}$ is a listener's memory state before processing word $x_i$, then the true surprisal experienced by the listener will be given by \ref{eq:lossy-surp}:
+\begin{equation}
+  \label{eq:lossy-surp}
+  S_M := -\log_2 p(x_i|m_{i-1}),
+\end{equation}
+which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof). % Gibbs' inequality
+The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with surprisal \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
+
+Here we propose to model human language comprehension by treating the listener's memory $m_{i-1}$ as a lossily compressed representation of the context $x_{1...i-1}$, and the processing effort associated with each word using Eq.~\ref{eq:lossy-surp}.
+
 Importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
@@
 Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context \jd{CITE roger}.
-Surprisal is known to be a strong linear predictor of processing effort as measured by per-word log-transformed reading times.
-For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
-\begin{equation}\label{eq:surp}
-	S :=	-\log p(x_i|x_{1...i-1})
-\end{equation}
-where $x_i$ is the $i$-th word in a given utterance (or discourse).
-%
-%In reality, a listener will have bounded capacity to represent prior context.
 A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).

### 2019-05-10 paper-short.tex
add text

@@
 Here we propose to model human language comprehension by treating the listener's memory $m_{i-1}$ as a lossily compressed representation of the context $x_{1...i-1}$, and the processing effort associated with each word using Eq.~\ref{eq:lossy-surp}.
+Suppose that the listener's memory has a fixed information-carrying capacity $k$,
 
-Importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
+\rljf{I am feeling the urge to reframe everything in terms similar to my NYU talk, where we say the processing effort of a language is its entropy rate (an unavoidable cost) plus excess cost arising from memory distortion, which should be minimized.
+This feels like a very general theory, which can be explained graphically.
+Linguists and psycholinguists will not understand it at first, but maybe they will if we include the stuff at the end about how this relates to dependency locality in word order and processing.
+The entropy rate framing would allow us to define more useful terms for explaining our results.
+For example above we have $x_i$, $x_{1...i-1}$, and $m_{i-1}$ as values, but really they are best thought of as random variables, becuase all of our results are about entropy and mutual information, not pointwise surprisal or pointwise MI.
+Maybe we could even make some kind of call-out to the competence/performance distinction: entropy rate is the best achievable efficiency in language processing given a perfect listener (like competence), but the memory-distorted cross-entropy rate gives us the true difficulty (performance).
+Tagline: Human languages minimize the divergence between competence and performance given fixed processing resources.}
+
+importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
 More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.

### 2019-05-10 paper-short.tex
changes

--- 2019-05-03 paper-short.tex
@@
   % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
-  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \citep{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.
+  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \cite{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.
 

### 2019-05-15 paper-short.tex

@@
 \usepackage{xcolor}
+\usepackage{natbib}
 \usepackage{url}
@@
 \newcommand{\thetal}[0]{{\theta_{LM}}}
+\newcommand{\key}[1]{\textbf{#1}}
 
@@
   % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
-  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \cite{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.
+  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension.
 
@@
   % 1 sentence here we show
-  Here we give evidence for this view by showing that human languages structure information in time so as to enable efficient communication under the constraint that language must be comprehended and produced by agents with limited short-term memory resources.
-  We directly show that languages minimize memory usage by calculating information-theoretic bounds on the amount of memory resources required for language comprehension in 54 real languages and in random baseline languages, finding that the real languages require less memory.
+  Here we give a novel formalization of this view in terms of information theory, and demonstrate that human languages structure information in time so as to enable efficient communication under the constraint that language must be comprehended and produced by agents with limited short-term memory resources.
+  We directly show that languages minimize memory usage by calculating information-theoretic bounds on the amount of working memory resources required for language comprehension in 54 real languages and in random baseline languages, finding that the real languages require less memory.
   % 2-3 sentences how main result adds to knowledge
@@
 % introductory paragraph
+Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
+A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
+Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
+We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
 
-\note{all three of the next are like first sentences}
-Human languages express thoughts into sentences, i.e. linear strings of words. 
-%For example, if your mountain climber friend Bob decided to turn back near the summit of Mt. Everest, you might say \emph{Bob gave the attempt to climb Mt. Everest up}.
-% Run with this example
-\jd{is the commented out example sth that will make it in? it does seem like an example would be good here}
-
-Each language embodies a different strategy for encoding thoughts into strings of words. \jd{will this sentence and the next both survive?}\note{make this sentence nonredundant}
-\note{give an example}
+The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \citep{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
 
@@
 
+% Nice phrases from Sims (2018):
+% biological information processing should seek to maximize performance subject to constraints on information processing capacity.
+% The current results can be concisely stated as follows:
+% Rate-distortion theory provides analytical and numerical tools for solving such constrained optimization problems (12–14).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We begin developing our model by considering the process of language comprehension in a person processing a stream of words uttered by an interlocutor (Figure~\ref{fig:communication}).
-Experimental research has established that such a listener (1) maintains information about the words received so far, (2) forms probabilistic expectations about the upcoming words, and (3) words are easier to process when they are highly expected in context \citep{hale2001probabilistic,levy2008expectation} \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}.
-
-Experimental research has shown that human processing effort on a given word is strongly predicted by its \textbf{surprisal}, the degree to which it cannot be predicted from the preceding context \citep{hale2001probabilistic,levy2008expectation}.
-Surprisal is known to be a strong linear predictor of processing effort as measured by per-word reading times \citep{smith2013effect}.
-For an idealized listener with unbounded memory resources, surprisal will be given by the negative log-probability of a word given its preceding context:
+We begin developing our model by considering the process of language comprehension illustrated in Figure~\ref{fig:communication}, where the listener Alice is processing a stream of words uttered by an interlocutor Bob (Figure~\ref{fig:communication}).
+Experimental research has established that (1) listeners maintain information about the words received so far, (2) listeners forms probabilistic expectations about the upcoming words, and (3) words are easy to process to the extent that they are predictable in context \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}.
+Furthermore, the exact mathematical form of the relationship between predictability and processing difficulty (as measured e.g.
+in reading times) seems to be given by the \key{information content} of a word in context \citep{hale2001probabilistic,levy2008expectation,smith2013effect}.
+For an idealized listener with unbounded memory resources, information content will be given by the negative log-probability of a word given its preceding context:
 \begin{equation}\label{eq:surp}
@@
 where $x_i$ is the $i$-th word in a given utterance (or discourse).
-In information-theoretic terms, surprisal is equivalent to the \textbf{information content} of a word, measured in bits, and the fact that the per-word reading time appears to be linear in the surprisal suggests that human language comprehension is best characterized in terms of a scalar rate of information processing in terms of bits per second.
+Information content is measured in bits and gives the amount of information communicated by the word.
 The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \citep{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \citep{egner2010expectation}, auditory and music perception \citep{agres2018information}, and motor planning \citep{wolpert2001motor}.
+The average information-processing difficulty associated with a \emph{language} would then be the expected information content of each word in context, which is equivalent to an information-theoretic quantity called \key{entropy rate}: \rljf{the def.
+below is true only for a stationary process but maybe that's too much detail}
+\begin{equation}\label{eq:entropy-rate}
+  h := \mathbb{E}[-\log_2 p(x_i|x_{1...i-1})].
+\end{equation}
+The entropy rate of a language (seen as a stochastic process) is thus a measurement of a component of the information-processing cost associated with comprehending the language.
 
 However, Equation~\ref{eq:surp} can only represent processing difficulty for a listener with unbounded memory resources, who can accurately remember the complete history of previous words $x_{1...i-1}$ leading up to the current word $x_i$.
-More realistically, human listeners have fixed-capacity memory resources that are being deployed during language comprehension, meaning that their predictions about upcoming words may be inaccurate compared to Eq.~\ref{eq:surp}.
+More realistically, human listeners have fixed-capacity memory resources that are being deployed during language comprehension, meaning that their predictions about upcoming words will be inaccurate compared to Eq.~\ref{eq:surp}.
 If $m_{i-1}$ is a listener's memory state before processing word $x_i$, then the true surprisal experienced by the listener will be given by \ref{eq:lossy-surp}:
@@
 \end{equation}
-which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof). % Gibbs' inequality
-The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with surprisal \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
+which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof).
+The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with raw information content \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
+The average processing difficulty associated with a language then comes out to a \key{cross entropy}:
 
@@
 This feels like a very general theory, which can be explained graphically.
-Linguists and psycholinguists will not understand it at first, but maybe they will if we include the stuff at the end about how this relates to dependency locality in word order and processing.
+Linguists and psycholinguists will not understand it at first, but maybe they will if we include the stuff about how this relates to dependency locality in word order and processing.
 The entropy rate framing would allow us to define more useful terms for explaining our results.
@@
 Tagline: Human languages minimize the divergence between competence and performance given fixed processing resources.}
+
+\rljf{Here we recast these theories in terms of information theory and the theory of statistical complexity...}
 

### 2019-05-15 short-paper-si.tex
SI

--- 2019-03-25 paper.tex
@@
 
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
+\title{SI}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
+\date{2019}
 
@@
 
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
-
-\section{Introduction}
-
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unaccebtability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-\section{Memory-Surprisal Tradeoff}
-
-With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal
-
-In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.
-
-We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
-%Informally, we expect that both speaker and listener will be affected by memory demands:
-%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
-For the listener, predicting the next word well requires maintaining information about the past.
-For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
-For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
-Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.
-
-TODO say at some point that we're studying sentence-internal memory
-
-\subsection{Theoretical Results}
-
-We now introduce our main theoretical result on memory-surprisal tradeoffs.
-
-\paragraph{Setting}
-We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-
-We model the sequence as a probabilistic sequence; that is, given a context $x_{<t}$, the next word is distributed according to a distribution $p(x_t|x_{<t})$.
-
-% $(X_t)_{t \in \mathbb{Z}}$
-%We think of the language as a stochastic process, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
-%As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
-%This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
-%We make the assumption that this process is \emph{stationary}.
-%Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
-%Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
-%In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-%Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
-
-%The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
-%We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
-%The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
-%\begin{equation}\label{eq:markov-speaker}
-%X_{>t} \bot X_{\leq t} | S_t
-%\end{equation}
-%This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
-
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
-	\caption{Illustration of (\ref{eq:listener-markov}).
-As the utterance unfolds, the listener maintains a memory state.
-After receiving word $X_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $X_t$.}\label{fig:listener-markov}
-\end{figure}
-
-There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $L_t$ is determined by the last word $X_t$, and the prior memory state $L_{t-1}$.
-As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
-This is formalized as a statement about conditional probabilities:
-	\begin{equation}\label{eq:listener-markov}
-p(L_1| (X_{t})_t, L_0)   = p((X_{t})_t| L_0, X_1)
-%p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.	
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
-This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-For the study of memory in sentence processing, this seems fair. Discuss this more.
-
-%
-%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-%Formally: 
-%\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot X_{>t} | X_{\leq t}
-%\end{equation}
-%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-%
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-%\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-%\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
-
-%The listener can trade off memory and future surprisal:
-%A listener who chooses to store less memory will exerience higher surprisal in the future.
-%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-%Consider a listener who uses $J$ bits of memory on average.
-%What can we say about the listener's surprisal?
-
-\paragraph{Conditional Mutual Information}
-
-We will use the concept of \emph{conditional mutual information} (Figure~\ref{fig:it-surprisal-intuition}).
-%For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
-%\begin{equation}
-%	I[X,Y] := H[X] - H[X|Y]
-%\end{equation}
-%For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-%\begin{equation}
-%I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-%\end{equation}
-%We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-%	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-	I_t := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-\end{equation}
-This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
-	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}\label{fig:it-surprisal-intuition}
-\end{figure*}
-
-We return to the information and memory curves in Figure~\ref{fig:basic}.
-In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
-This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
-Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
-We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
-As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-Then
-
-\begin{thm}\label{prop:suboptimal}
-	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
-	\begin{equation}\label{eq:memory}
-		\sum_{t=1}^T t I_t
-	\end{equation}
-bits of memory on average.
-Then this listener will incur surprisal at least
-	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-	on average.
-\end{thm}
-The proof is given in the appendix (REF).
-
-%An intuitive (non-rigorous) argument goes as follows:
-%For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
-
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
-%
-	\caption{Left: $I_t$ as a function of $t$, for two different processes.
-$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
-Left: $t \cdot I_t$ as a function of $t$ for the same processes.
-}\label{fig:basic}
-\end{figure*}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
-\end{center}
-	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
-\end{figure}
-
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-Then interplate linearly (justify this in appendix).
-We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
-We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
-Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
-In the given example, $T=4$.
-By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
-	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
-Recall that the red process had a faster decay of conditional mutual information.
-Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
-	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
-\end{figure}
-
-%that the model does not have an `absolute clock' built into it.
-
-Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
-% and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-\paragraph{Information Locality}
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-\paragraph{Center Embeddings}
-
-\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-
-%\cite{gibson-linguistic-1998}
-
-%first center-embeddings, POS level model
-
-%Later work found evidence that cross-serial dependencies are less taxing.
-
-%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies
-
-\paragraph{Other Psycholinguistic Predictions}
-
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries.
-
-\paragraph{Speakers}
-
-% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
+\section{Materials and Methods}
 
@@
 
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
 \paragraph{Model and Parameter Estimation}
@@
 
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
 \paragraph{Choice of Hyperparameters}
@@
 
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
 \paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
@@
 
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
 For each language, we collected data from the actual orderings and from several random grammars.
@@
 Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
 
@@
 
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
 We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
 We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
@@
 Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
 
@@
 
-\paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Speaker Memory}
-
-%The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
-TODO
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
-%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-Our results apply both to speakers and listeners.
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-%\begin{proposition}
-%Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
-%\end{proposition}
-%
-%\begin{proof}[Proof Sketch]
-%	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
-%
-%	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
-We consider incremental Earley parsing \citep{earley-efficient-1970}.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
-For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
-%	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
-%\end{proof}
-%
-%
-%
-
-\section{Conclusion}
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
 \section{N-Gram Models}
@@
 
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
 \section{Character-Level Modeling}
@@
 
-%\begin{proof}
 	We denote the listener's memory state at time $t$, after hearing $X_{<t} = ... X_{t-2} X_{t-1}$ by $L_t$.
@@
 
-	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
 	Plugging this inequality into the equality above:

### 2019-05-29 paper-short.tex

@@
 \usepackage{xcolor}
-\usepackage{natbib}
+%\usepackage{natbib}
 \usepackage{url}
@@
 \newcommand{\rljf}[1]{\textcolor{Red}{[rljf: #1]}} 
+\newcommand{\mhahn}[1]{\textcolor{Red}{[mhahn: #1]}}
 
@@
 
-\baselineskip24pt
+%\baselineskip24pt
 
@@
 
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \citep{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
+The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
 
@@
 
-A lot of progress has been made on understanding. Evidence for the idea comes from various different frameworks.
-However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\note{TODO make this better}
-\note{Maybe there is a citation that explicitly mentions this as a limitation of  thigngs like DLM}
+%A lot of progress has been made on understanding. Evidence for the idea comes from various different frameworks.
+%However, it's been hard to concluseively take this as evidence for efficient processing under resource limitations because tests have been tied to specific algorithmic models of language processing.\note{TODO make this better}
+%\note{Maybe there is a citation that explicitly mentions this as a limitation of  thigngs like DLM}
 %however, it has been hard to test this idea without ad-hoc modeling assumptions.
-\jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
-\note{this is a strong critique. not clear what kinds of ad-hoc modeling this is about}
-e.g. previously based on some grammatical formalism and a linking to ideas about parsing algorithms.
+%\jd{like what? why is the information-theoretic approach not subject to this ad-hoc-ness?}
+%\note{this is a strong critique. not clear what kinds of ad-hoc modeling this is about}
+%e.g. previously based on some grammatical formalism and a linking to ideas about parsing algorithms.
 
 %Alternative?: there has been debate about what shapes human language. role played by UG/innate constraints? environment? cognition?
+
 For the first time, we test this hypothesis by applying...
@@
 maybe can color-code?
-}}
+} \mhahn{problem suggests that memory has to be contiguous substrings}}
 	\label{fig:communication}
@@
 Furthermore, the exact mathematical form of the relationship between predictability and processing difficulty (as measured e.g.
-in reading times) seems to be given by the \key{information content} of a word in context \citep{hale2001probabilistic,levy2008expectation,smith2013effect}.
+in reading times) seems to be given by the \key{information content} of a word in context \cite{hale2001probabilistic,levy2008expectation,smith2013effect}.
 For an idealized listener with unbounded memory resources, information content will be given by the negative log-probability of a word given its preceding context:
@@
 Information content is measured in bits and gives the amount of information communicated by the word.
-The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \citep{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \citep{egner2010expectation}, auditory and music perception \citep{agres2018information}, and motor planning \citep{wolpert2001motor}.
+The idea that processing effort is determined by surprisal links language comprehension to theories of perception and brain function that are organized around the idea of prediction and \key{predictive coding} \cite{friston2009predictive,friston2010free,clark2013whatever}, which has had empirical success in other domains such as perception \cite{egner2010expectation}, auditory and music perception \cite{agres2018information}, and motor planning \cite{wolpert2001motor}.
 The average information-processing difficulty associated with a \emph{language} would then be the expected information content of each word in context, which is equivalent to an information-theoretic quantity called \key{entropy rate}: \rljf{the def.
-below is true only for a stationary process but maybe that's too much detail}
+below is true only for a stationary process but maybe that's too much detail} \mhahn{yes, the whole theory assumes stationarity.
+not yet sure where to say this}
 \begin{equation}\label{eq:entropy-rate}
@@
 which will on average be larger than Eq.~\ref{eq:surp} (see SI for proof).
-The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with raw information content \citep{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
+The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with raw information content \cite{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.
 The average processing difficulty associated with a language then comes out to a \key{cross entropy}:
-
-Here we propose to model human language comprehension by treating the listener's memory $m_{i-1}$ as a lossily compressed representation of the context $x_{1...i-1}$, and the processing effort associated with each word using Eq.~\ref{eq:lossy-surp}.
-Suppose that the listener's memory has a fixed information-carrying capacity $k$,
 
@@
 
-importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
-More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.
-The psycholinguistic literature has established that humans differ in their capacity to store such representations \jd{CITE}, and that capacity limitations can create difficulties in language comprehension \jd{CITE}.
+Here we propose to model human language comprehension by treating the listener's memory $m_{i-1}$ as a lossily compressed representation of the context $x_{1...i-1}$, and the processing effort associated with each word using Eq.~\ref{eq:lossy-surp}.
+Suppose that the listener's memory has a fixed information-carrying capacity $k$, then this listener has to compress the context into a memory representation $m_{i-1}$ with at most $k$ bits: $H[m_{i-1}] \leq k$.
 
-Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context \jd{CITE roger}.
-A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
-As in Figure~\ref{fig:communication}, we denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
-This listener will experience surprisal
-\begin{equation}\label{eq:surp-listener}
-	S_L :=	-\log p(x_t|L_t)
-\end{equation}
-upon hearing the next word $x_t$.
+Information theory implies that the processing cost $S_M$ trades off with the memory capacity $k$:
+Processing cost will be lower for a listener who represents past input at higher levels of precision.
+
+%importantly, the listener can represent the prior context at higher or lower levels of precision (Figure~\ref{fig:communication}).
+%More detailed memory representations lead to a higher working memory load, but also enable more precise predictions.
+%The psycholinguistic literature has established that humans differ in their capacity to store such representations \jd{CITE}, and that capacity limitations can create difficulties in language comprehension \jd{CITE}.
+%
+%Experimental research has shown that human processing effort on a given word is strongly predicted by its surprisal, the degree to which it cannot be predicted from the preceding context \jd{CITE roger}.
+%A listener with limited memory of prior context will not be able to compute the full conditional distribution~(\ref{eq:surp}).
+%As in Figure~\ref{fig:communication}, we denote a listener's memory representation at time $t$ by $L_t$, which is some -- generally imperfect -- summary of the preceding context $\dots x_{t-3}x_{t-2}x_{t-1}$.
+%This listener will experience surprisal
+%\begin{equation}\label{eq:surp-listener}
+%	S_L :=	-\log p(x_t|L_t)
+%\end{equation}
+%upon hearing the next word $x_t$.
 %The amount of information stored by the listener is the number of bits in $L_t$, i.e., its entropy $M := H[L_t]$.
@@
 
-Information theory, specifically the Data Processing Inequality \note{add reference}, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
+%
+%Information theory, specifically the Data Processing Inequality \note{add reference}, implies that the average surprisal $S$ will be higher for a listener who represents past input at lower levels of precision.
 \jd{include an additional sentence on the DPI?}
-Consequently, there is a tradeoff between memory and surprisal:
-A listener can achieve lower levels of processing effort at the expense of increased memory load.
-
-\note{reader will have lost the plot by now. strogn promise in the beginning. can solve this by foreshadowing}
+%Consequently, there is a tradeoff between memory and surprisal:
+%A listener can achieve lower levels of processing effort at the expense of increased memory load.
+%
+%	\note{reader will have lost the plot by now. strogn promise in the beginning. can solve this by foreshadowing}
 
@@
 We then consider the smallest $T$ such that the area under the curve of $t I_t$, to the left of $T$, has size $B$.
-Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$.\note{maybe result first, and then say what $T$ is}
+Such a listener will experience average surprisal at least $H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t$. %\note{maybe result first, and then say what $T$ is}
 By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.
-\note{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
-
-\note{Alternative:}
-A listener will experience average surprisal at least
-\begin{equation}
-	H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t
-\end{equation}
-where $T$ is chosen to be the smallest $T$ for which $\sum_{t=1}^T t I_t \geq B$.
-\note{End alternative.}
+%\note{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
+%
+%
+%\note{Alternative:}
+%A listener will experience average surprisal at least
+%\begin{equation}
+%	H[X_t| X_{<t}] + \sum_{t=T+1}^\infty I_t
+%\end{equation}
+%where $T$ is chosen to be the smallest $T$ for which $\sum_{t=1}^T t I_t \geq B$.
+%\note{End alternative.}
 
@@
 See SI for details.
-\note{maybe even more. hard concept to grasp. say more about the representation format, that it's as atheoretical as possible. encodes the ways heads are encoded to their dependents}
-\jd{refer to SI here and make sure to include a lot more information in the SI about how the grammars were created}
+%\note{maybe even more. hard concept to grasp. say more about the representation format, that it's as atheoretical as possible. encodes the ways heads are encoded to their dependents}
+%\jd{refer to SI here and make sure to include a lot more information in the SI about how the grammars were created}
 %These random grammars apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
@@
 The point of divergence is given by that memory capacity that fully exhausts information provided by the immediately preceding word, that is, when the capacity of a Markov model in approximating language is exhausted.
-\note{can one rephrase this?}
+%\note{can one rephrase this?}
 
@@
 \includegraphics[width=\textwidth]{figures/full-results.png}
-	\caption{\jd{FILL OUT; make less blurry}}\label{fig:results}
+	\caption{Tradeoffs between memory (x axis) and processing cost (y axis) in 54 languages, for real orderings (blue) and counterfactual baseline grammars (red).
+\jd{FILL OUT; make less blurry}}\label{fig:results}
 \end{figure}

### 2019-06-04 short-paper-si.tex
paper
SI

@@
 
+\section{Locality in a model with Memory Retrieval}
+
+Here we describe an information-theoretic formalization of a model that has both a limited WM and an unlimited STM.
+
+We can explicitly explain how this covers the McElree ideas and the Lewis and Vasishth ACT-R model.
+
+Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
+
+\begin{thm}
+Average working memory $H[L_t] \leq \sum_{t=1}^T t I_t$, average retrieval $H[R_t] \leq \sum_{t=T+1}^S I_t$ (per word), leads to average surprisal increase at least $\geq \sum_{t>S} I_t$.
+\end{thm}
+
+\begin{proof}
+The negative surprisal gap is
+$\leq \frac{1}{T}(I[X_1\dots X_T|(M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t) \leq  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) = H[R_t] - \sum_{t>T} I_t$
+
+\end{proof}
+
+We can also think of this as a multi-objective optimization:
+
+minimize $\lambda_1 WM + \lambda_2 Retrieval$ to achieve a given surprisal level.
+
+If $\frac{\lambda_2}{\lambda_1} \rightarrow \infty$ (retrievals get more expensive), recover previous model.
+
+If $\frac{\lambda_2}{\lambda_1} \rightarrow 0$ (retrievals get cheaper), locality effect gets weaker, and disappears in the limit\footnote{(Of course, even in this limit, there might be additional factors that may still favor locality in a specific implementation of memory -- e.g., in ACT-R, decay and interference are less problematic if there is locality.)}
+
+Objective
+$$\min_{T} \lambda_1 \sum_{t=1}^T t I_t + \lambda_2 \sum_{t=T+1}^S I_t$$
+has solution $T \approx \frac{\lambda_2}{\lambda_1}$.\footnote{Can do simple proof using the continuous-$T$-version.}
+
+Consequence: As long as retrievals are more expensive than keeping the same amount in WM, locality is predicted.
+
 \end{document}

### 2019-08-05 introduction.tex
revising

--- 2019-04-03 paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
+\mhahn{pasted from the other draft}
+Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
+A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
+Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
+We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
 
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
+The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
 
-\usepackage{tikz-dependency}
-\usepackage{longtable}
+These strategies can be seen as sets of rules that differ in the way ...
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
+Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
+We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
 
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
+%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
+The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
+Here
 
-\usepackage{multirow}
-
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-
-\usepackage{amsthm}
-
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
-
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
-
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
-
-\begin{document}
-
-\maketitle
-
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
-
-\section{Introduction}
-
+\mhahn{original intro}
 Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
@@
 Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
-% More detailed lit review
-
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
-
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
-
-With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal
-
-In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.
-
-We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
-%Informally, we expect that both speaker and listener will be affected by memory demands:
-%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
-For the listener, predicting the next word well requires maintaining information about the past.
-For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
-For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
-Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.
-
-TODO say at some point that we're studying sentence-internal memory
-
-\subsection{Theoretical Results}
-
-We now introduce our main theoretical result on memory-surprisal tradeoffs.
-
-\paragraph{Setting}
-We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-
-We model the sequence as a probabilistic sequence; that is, given a context $x_{<t}$, the next word is distributed according to a distribution $p(x_t|x_{<t})$.
-
-% $(X_t)_{t \in \mathbb{Z}}$
-%We think of the language as a stochastic process, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
-%As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
-%This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
-%We make the assumption that this process is \emph{stationary}.
-%Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
-%Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
-%In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-%Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
-
-%The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
-%We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
-%The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
-%\begin{equation}\label{eq:markov-speaker}
-%X_{>t} \bot X_{\leq t} | S_t
-%\end{equation}
-%This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
-
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
-	\caption{Illustration of (\ref{eq:listener-markov}).
-As the utterance unfolds, the listener maintains a memory state.
-After receiving word $X_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $X_t$.}\label{fig:listener-markov}
-\end{figure}
-
-There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $L_t$ is determined by the last word $X_t$, and the prior memory state $L_{t-1}$.
-As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
-This is formalized as a statement about conditional probabilities:
-	\begin{equation}\label{eq:listener-markov}
-p(L_1| (X_{t})_t, L_0)   = p((X_{t})_t| L_0, X_1)
-%p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.	
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
-This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-For the study of memory in sentence processing, this seems fair. Discuss this more.
-
-%
-%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-%Formally: 
-%\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot X_{>t} | X_{\leq t}
-%\end{equation}
-%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-%
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-%\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-%\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
-
-%The listener can trade off memory and future surprisal:
-%A listener who chooses to store less memory will exerience higher surprisal in the future.
-%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-%Consider a listener who uses $J$ bits of memory on average.
-%What can we say about the listener's surprisal?
-
-\paragraph{Conditional Mutual Information}
-
-We will use the concept of \emph{conditional mutual information} (Figure~\ref{fig:it-surprisal-intuition}).
-%For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
-%\begin{equation}
-%	I[X,Y] := H[X] - H[X|Y]
-%\end{equation}
-%For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-%\begin{equation}
-%I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-%\end{equation}
-%We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-%	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-	I_t := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-\end{equation}
-This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
-	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}\label{fig:it-surprisal-intuition}
-\end{figure*}
-
-We return to the information and memory curves in Figure~\ref{fig:basic}.
-In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
-This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
-Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
-We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
-As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-Then
-
-\begin{thm}\label{prop:suboptimal}
-	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
-	\begin{equation}\label{eq:memory}
-		\sum_{t=1}^T t I_t
-	\end{equation}
-bits of memory on average.
-Then this listener will incur surprisal at least
-	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-	on average.
-\end{thm}
-The proof is given in the appendix (REF).
-
-%An intuitive (non-rigorous) argument goes as follows:
-%For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
-
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
-%
-	\caption{Left: $I_t$ as a function of $t$, for two different processes.
-$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
-Left: $t \cdot I_t$ as a function of $t$ for the same processes.
-}\label{fig:basic}
-\end{figure*}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
-\end{center}
-	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
-\end{figure}
-
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-Then interplate linearly (justify this in appendix).
-We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
-We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
-Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
-In the given example, $T=4$.
-By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
-	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
-Recall that the red process had a faster decay of conditional mutual information.
-Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
-	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
-\end{figure}
-
-%that the model does not have an `absolute clock' built into it.
-
-Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
-% and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-\paragraph{Information Locality}
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-\paragraph{Center Embeddings}
-
-\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-
-%\cite{gibson-linguistic-1998}
-
-%first center-embeddings, POS level model
-
-%Later work found evidence that cross-serial dependencies are less taxing.
-
-%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies
-
-\paragraph{Other Psycholinguistic Predictions}
-
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries.
-
-\paragraph{Speakers}
-
-% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 52 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain text created by non-native speakers.
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.
-
-In our model, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE) and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
-This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[X_t | X_0, X_1, ..., X_{t-1}] - P_\theta[X_t | X_1, ..., X_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[X_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
-%\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{longtable}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{longtable}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{longtable}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{longtable}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{longtable}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{longtable}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{longtable}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\section{Experiment 3: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Speaker Memory}
-
-%The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
-TODO
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
-%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-Our results apply both to speakers and listeners.
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-%\begin{proposition}
-%Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
-%\end{proposition}
-%
-%\begin{proof}[Proof Sketch]
-%	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
-%
-%	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
-We consider incremental Earley parsing \citep{earley-efficient-1970}.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
-For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
-%	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
-%\end{proof}
-%
-%
-%
-
-\section{Conclusion}
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{longtable}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Proof of the Theorem}
-
-%\begin{proof}
-	We denote the listener's memory state at time $t$, after hearing $X_{<t} = ... X_{t-2} X_{t-1}$ by $L_t$.
-	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[X_t | L_t]$.
-	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]$.
-By the assumption of stationarity, we can rewrite this expression as
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]\right) 
-\end{align*}
-	\begin{lemma}
-For any $t$, the following inequality holds:
-%	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
-	\begin{equation}
-%H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-H[X_t | L_t] \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{equation}
-	\end{lemma}
-	\begin{proof}[Proof of the Lemma]
-
-By Equation~\ref{eq:listener-markov}:
-	\begin{equation}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-we have
-	\begin{align*}
-p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
-p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
-	\end{align*}
-Therefore
-\begin{align*}
-p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) & = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) \\
-& = \sum_{X_{<1}} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1}) \\
-&= \sum_{X_{<1}} p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, X_1, X_{2 \dots t-1}) \\
-&= p(X_t|L_0,  X_1, X_{2 \dots t-1}) 
-\end{align*}
-Thus
-\begin{equation}
-(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
-\end{equation}
-Thus
-	\begin{equation}
-H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
-	\end{equation}
-
-Finally, iteratively applying this inequality, we get:
-		\begin{align*}
-		H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq H[X_t| X_{t-2, t-1}, L_{t-2}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{align*}
-
-	\end{proof}
-
-	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-	Plugging this inequality into the equality above:
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[X_t|X_{1\dots t-1}, L_0] - \operatorname{H}[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(\operatorname{H}[X_{1\dots T} | L_0] - \operatorname{H}[X_{1\dots T} | X_{\leq 0}]\right)  \\
-	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
-\end{align*}
-	The first term $I[X_{1\dots T}|X_{\leq 0}]$ can be rewritten in terms of $I_t$:
-	\begin{align*}
-		I[X_{1\dots T}|X_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[X_i, X_j | X_{j+1}...X_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
-	\end{align*}
-	Therefore
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
-	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
-\end{align*}
-	$I[X_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
-	\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
-		&= \sum_{t > T} I_t
-\end{align*}
-	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[X_t|L_t] \geq \operatorname{H}[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
-%\end{proof}
-
-Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
-
-\end{document}

### 2019-08-05 paper.tex
revising

@@
 \newcommand\mhahn[1]{{\color{red}(#1)}}
+\newcommand\note[1]{{\color{red}(#1)}}
+\newcommand\jd[1]{{\color{red}(#1)}}
+\newcommand\rljf[1]{{\color{red}(#1)}}
+\newcommand{\key}[1]{\textbf{#1}}
 
@@
 %	For each positive integer $t$, define
-%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
+%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
 %i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
@@
 %Then this comprehender will incur average surprisal at least
-%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
+%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
 %\end{thm}
@@
 
+\input{introduction.tex}
+
+\mhahn{pasted from the other draft}
+Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
+A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
+Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
+We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
+
+The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
+
+These strategies can be seen as sets of rules that differ in the way ...
+Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
+Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
+We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
+
+%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
+The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
+Here
+
+\mhahn{original intro}
 Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
@@
 
-With minimal assumptions, we will use information theory to derive a tradeoff between \emph{listener memory} and (listener) surprisal
-
-In the second part of the paper, we examine whether word orders in natural language optimise this tradeoff.
-
-We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
-%Informally, we expect that both speaker and listener will be affected by memory demands:
-%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
-For the listener, predicting the next word well requires maintaining information about the past.
-For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
-For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
-Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.
-
-TODO say at some point that we're studying sentence-internal memory
-
-\subsection{Theoretical Results}
-
-We now introduce our main theoretical result on memory-surprisal tradeoffs.
-
-\paragraph{Setting}
-We formalize a language as a probabilistic sequence of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-
-We model the sequence as a probabilistic sequence; that is, given a context $x_{<t}$, the next word is distributed according to a distribution $p(x_t|x_{<t})$.
-
-% $(X_t)_{t \in \mathbb{Z}}$
-%We think of the language as a stochastic process, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
-%As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
-%This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
-%We make the assumption that this process is \emph{stationary}.
-%Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
-%Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
-%In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-%Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
-
-%The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
-%We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
-%The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
-%\begin{equation}\label{eq:markov-speaker}
-%X_{>t} \bot X_{\leq t} | S_t
-%\end{equation}
-%This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.
-
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
-	\caption{Illustration of (\ref{eq:listener-markov}).
-As the utterance unfolds, the listener maintains a memory state.
-After receiving word $X_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $X_t$.}\label{fig:listener-markov}
-\end{figure}
-
-There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $L_t$ is determined by the last word $X_t$, and the prior memory state $L_{t-1}$.
-As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $X_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
-This is formalized as a statement about conditional probabilities:
-	\begin{equation}\label{eq:listener-markov}
-p(L_1| (X_{t})_t, L_0)   = p((X_{t})_t| L_0, X_1)
-%p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $X_1$.	
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
-This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-For the study of memory in sentence processing, this seems fair. Discuss this more.
-
-%
-%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
-%Formally: 
-%\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot X_{>t} | X_{\leq t}
-%\end{equation}
-%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
-%
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
-%\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot X_{<t} | X_{t-1}, L_{t-1}
-%\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
-
-%The listener can trade off memory and future surprisal:
-%A listener who chooses to store less memory will exerience higher surprisal in the future.
-%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
-
-%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
-%Consider a listener who uses $J$ bits of memory on average.
-%What can we say about the listener's surprisal?
-
-\paragraph{Conditional Mutual Information}
-
-We will use the concept of \emph{conditional mutual information} (Figure~\ref{fig:it-surprisal-intuition}).
-%For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
-%\begin{equation}
-%	I[X,Y] := H[X] - H[X|Y]
-%\end{equation}
-%For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
-%\begin{equation}
-%I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
-%\end{equation}
-%We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
-\begin{equation}
-%	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-	I_t := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
-\end{equation}
-This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
-That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
-This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.
-
-\begin{figure*}
-\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
-	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}\label{fig:it-surprisal-intuition}
-\end{figure*}
-
-We return to the information and memory curves in Figure~\ref{fig:basic}.
-In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
-This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
-Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
-We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
-As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
-Then
-
-\begin{thm}\label{prop:suboptimal}
-	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
-	\begin{equation}\label{eq:memory}
-		\sum_{t=1}^T t I_t
-	\end{equation}
-bits of memory on average.
-Then this listener will incur surprisal at least
-	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
-	on average.
-\end{thm}
-The proof is given in the appendix (REF).
-
-%An intuitive (non-rigorous) argument goes as follows:
-%For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
-
-\begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
-%
-	\caption{Left: $I_t$ as a function of $t$, for two different processes.
-$I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past.
-Left: $t \cdot I_t$ as a function of $t$ for the same processes.
-}\label{fig:basic}
-\end{figure*}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
-\end{center}
-	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
-\end{figure}
-
-The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$.
-Then interplate linearly (justify this in appendix).
-We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
-We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
-	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
-Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal.
-In the given example, $T=4$.
-By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
-	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.
-Recall that the red process had a faster decay of conditional mutual information.
-Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
-	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
-Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
-\end{figure}
-
-%that the model does not have an `absolute clock' built into it.
-
-Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model.
-% and the mechanisms by which it computes predictions.
-In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
-Memory representations do not have to be rational or optimal for this bound to hold:
-It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
-
-\paragraph{Information Locality}
+\input{tradeoff.tex}
+
+\section{Information Locality}\label{sec:info-locality}
+
+TODO derive information locality
 
@@
 
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
+%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
 The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
@@
 %\paragraph{Discussion}
-%We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
+%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
 %The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
@@
 
-\paragraph{Center Embeddings}
-
-\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-
+%\paragraph{Center Embeddings}
+%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
 %\cite{gibson-linguistic-1998}
-
-%first center-embeddings, POS level model
-
-%Later work found evidence that cross-serial dependencies are less taxing.
-
-%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies
-
-\paragraph{Other Psycholinguistic Predictions}
-
+%\paragraph{Other Psycholinguistic Predictions}
 % RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-
-\paragraph{Speakers}
-
-% RF: what matters for the speaker is not I[X_t, X_0 | X_1, …, X_{t-1}], but I[X_t, X_0 | X_1, …, X_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
+%\paragraph{Speakers}
+% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
 This changes the interpretation of the mutual information.
@@
 \paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 52 languages.
+We considered all languages for which there are Universal Dependencies 2.3 treebanks with a total of at least 500 sentences of training data.
+We excluded data from historical languages.\footnote{Ancient Greek, Classical Chinese, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
+This resulted in 54 languages.
 
 For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain text created by non-native speakers.
+We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
 Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
@@
 We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.
-
-In our model, this parameter defines how dependents are ordered relative to their head:
+Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.\mhahn{give an example to make clear what this refers to}
+
+Following Gildea and colleagues, this parameter defines how dependents are ordered relative to their head:
 Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
@@
 \subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE) and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
+To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
 %We model language on the level of individual word forms.
@@
 
-The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
+The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
 \begin{equation}
-H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]
+H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
 \end{equation}
@@
 \begin{equation}
-	-\log P_\theta[X_t | X_0, X_1, ..., X_{t-1}] - P_\theta[X_t | X_1, ..., X_{t-1}]
+	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
 \end{equation}
@@
 
-We estimate the unigram entropy $H[X_0]$ by averaging over all models.
+We estimate the unigram entropy $H[w_0]$ by averaging over all models.
 
@@
 %Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-
 %We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
 %For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
@@
 In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{longtable}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{longtable}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{longtable}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{longtable}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{longtable}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{longtable}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{longtable}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{longtable}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{longtable}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{longtable}
-	\caption{Quantiles (part 3)}
-\end{table}
 
@@
 \section{Discussion}
-
-\subsection{Speaker Memory}
-
-%The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).
-TODO
 
@@
 \paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
+In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[w_t, w_0 | w_{1\dots t-1}]$.
+Prior work has studied the decay of \emph{unconditional} mutual information $I[w_t, w_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
 
@@
 While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$.
-%%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
+Consider the constant process where with probability 1/2 all $w_t = 0$, and with probability 1/2 all $w_t = 1$.
+%%$w_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
 The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
@@
 This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal.
+To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $w_t$ is equal to the XOR of $w_{t-1}$ and $w_{t-N}$, such that each $w_t$ has $Bernoulli(1/2)$ as its marginal.
 The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
@@
 \paragraph{Speakers vs Listeners}
-Our results apply both to speakers and listeners.
+can we say something about this
 
@@
 
-%\begin{proposition}
-%Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
-%\end{proposition}
-%
-%\begin{proof}[Proof Sketch]
-%	TODO Actually, Proposition 2 in \cite{chi-statistical-1999} is enough, no need to talk about how to do parsing.
-%
-%	We provide an algorithm that performs generation or optimal prediction and has finite average memory use.
-We consider incremental Earley parsing \citep{earley-efficient-1970}.
-Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$).
-For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
-%	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
-%\end{proof}
-%
-%
-%
-
 \section{Conclusion}
+
+TODO
 
@@
 
+\begin{figure}
+\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\caption{Histogram}\label{fig:hist-real}
+\end{figure}
+
+\begin{table}
+\begin{longtable}{l|ll||l|llllllllllllll}
+	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
+\input{corpusSizes.tex}
+\end{longtable}
+	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
+\end{table}
+
+\begin{table}
+\begin{longtable}{l|ll||l|llllllllllllll}
+	Language & Base. & Real & Language & Base. & Real \\ \hline
+\input{samplesNumber.tex}
+\end{longtable}
+	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{table}
+
+\begin{table}
+\begin{longtable}{l|lll||l|lllllllllllllll}
+	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
+\input{boot_g_REAL.tex}
+\end{longtable}
+	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_0.tex}
+\end{longtable}
+	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_1.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_2.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{longtable}{ccccccccccccccclll}
+\input{medians_REAL_3.tex}
+\end{longtable}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{slice-hists_REAL_0.tex}
+\end{tabular}
+	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{slice-hists_REAL_1.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{slice-hists_REAL_2.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{slice-hists_REAL_3.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_REAL_0.tex}
+\end{tabular}
+	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_REAL_1.tex}
+\end{tabular}
+	\caption{Quantiles (part 2)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_REAL_2.tex}
+\end{tabular}
+	\caption{Quantiles (part 3)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{l|ll||l|llllllllllllll}
+	Language & Base. & MLE & Language & Base. & MLE \\ \hline
+\input{samplesNumber_ground.tex}
+\end{tabular}
+	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{medians_0.tex}
+\end{tabular}
+	\caption{Experiment 3.
+Medians: For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{medians_1.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{medians_2.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{medians_3.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{medianDiff_0.tex}
+\end{tabular}
+	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
+Lower values indicate lower surprisal compared to baselines.
+Solid lines indicate sample means.
+Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{medianDiff_1.tex}
+\end{tabular}
+	\caption{Median Differences (Part 2)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{medianDiff_2.tex}
+\end{tabular}
+	\caption{Median Differences (Part 3)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{medianDiff_3.tex}
+\end{tabular}
+	\caption{Median Differences (Part 4)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_0.tex}
+\end{tabular}
+	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_1.tex}
+\end{tabular}
+	\caption{Quantiles (part 2)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{quantiles_2.tex}
+\end{tabular}
+	\caption{Quantiles (part 3)}
+\end{table}
+
 \section{N-Gram Models}
@@
 \begin{table}
-\begin{longtable}{ccccccccccccccclll}
+\begin{tabular}{ccccccccccccccclll}
 \input{medians_ngrams_0.tex}
-\end{longtable}
+\end{tabular}
 	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
@@
 \begin{table}
-\begin{longtable}{ccccccccccccccclll}
+\begin{tabular}{ccccccccccccccclll}
 \input{medians_ngrams_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
 \input{medians_ngrams_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
 \input{medians_ngrams_3.tex}
-\end{longtable}
+\end{tabular}
 	\caption{Medians (cont.)}
@@
 
-\section{Proof of the Theorem}
-
-%\begin{proof}
-	We denote the listener's memory state at time $t$, after hearing $X_{<t} = ... X_{t-2} X_{t-1}$ by $L_t$.
-	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[X_t | L_t]$.
-	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]$.
-By the assumption of stationarity, we can rewrite this expression as
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]\right) 
-\end{align*}
-	\begin{lemma}
-For any $t$, the following inequality holds:
-%	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]$$
-	\begin{equation}
-%H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-H[X_t | L_t] \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{equation}
-	\end{lemma}
-	\begin{proof}[Proof of the Lemma]
-
-By Equation~\ref{eq:listener-markov}:
-	\begin{equation}
-p((X_{t})_t| L_0, L_1, X_1)   = p((X_{t})_t| L_0, X_1)
-	\end{equation}
-we have
-	\begin{align*}
-p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) \\
-p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1})   = p(X_{<1}| L_0, X_1, X_{2 \dots t-1})
-	\end{align*}
-Therefore
-\begin{align*}
-p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) & = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) \\
-& = \sum_{X_{<1}} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, L_1, X_1, X_{2 \dots t-1}) \\
-&= \sum_{X_{<1}} p(X_t|X_{<1}, L_0,  X_1, X_{2 \dots t-1}) p(X_{<1}| L_0, X_1, X_{2 \dots t-1}) \\
-&= p(X_t|L_0,  X_1, X_{2 \dots t-1}) 
-\end{align*}
-Thus
-\begin{equation}
-(X_t) \rightarrow (L_0, X_1, X_{2 \dots t-1})   \rightarrow   (L_1, X_{2 \dots t-1})
-\end{equation}
-Thus
-	\begin{equation}
-H[X_t| X_{2 \dots t-1}, L_{1}] \geq H[X_t|X_{1 \dots t-1}, L_0]
-	\end{equation}
-
-Finally, iteratively applying this inequality, we get:
-		\begin{align*}
-		H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq H[X_t| X_{t-2, t-1}, L_{t-2}] \geq ... \geq H[X_t|X_{1 \dots t-1}, L_0]
-		\end{align*}
-
-	\end{proof}
-
-	% = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-%	$p(X_t|L_0, X_1, X_{2 \dots t-1}; L_1, X_{2 \dots t-1}) = p(X_t|L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, L_0, L_1, X_1, X_{2 \dots t-1}) = \int dX_{<1} p(X_t|X_{<1}, X_1, X_{2 \dots t-1})  = \int dX_{<1} p(X_t|L_0, X_{<1}, X_1, X_{2 \dots t-1}) = p(X_t|L_0, X_1, X_{2 \dots t-1}) $
-
-	Plugging this inequality into the equality above:
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[X_t|X_{1\dots t-1}, L_0] - \operatorname{H}[X_t | X_{1\dots t-1}, X_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(\operatorname{H}[X_{1\dots T} | L_0] - \operatorname{H}[X_{1\dots T} | X_{\leq 0}]\right)  \\
-	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
-\end{align*}
-	The first term $I[X_{1\dots T}|X_{\leq 0}]$ can be rewritten in terms of $I_t$:
-	\begin{align*}
-		I[X_{1\dots T}|X_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[X_i, X_j | X_{j+1}...X_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
-	\end{align*}
-	Therefore
-\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
-	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
-\end{align*}
-	$I[X_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
-	\begin{align*}
-	\operatorname{H}[X_t | L_t] - \operatorname{H}[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
-		&= \sum_{t > T} I_t
-\end{align*}
-	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[X_t|L_t] \geq \operatorname{H}[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
-%\end{proof}
-
-Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+\section{Formal Analysis and Proofs}
+\input{formal-analysis.tex}
 

### 2019-08-05 writeup/formal-analysis.tex
directories

In this section, we prove the theorem described above.

We first make explicit how we formalize language processing for proving the theorem.

\paragraph{Language as a Stochastic Process}
We represent language as a stochastic process:
We formalize a language as a probabilistic sequence of words $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}

We model the sequence as a probabilistic sequence; that is, given a context $w_{<t}$, the next word is distributed according to a distribution $p(w_t|w_{<t})$.

The assumption of infinite length is for mathematical convenience and does not affect the substance of our results:
As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.

We make the assumption that this process is \emph{stationary}.
Formally, this means that the conditional distribution $P(w_t|w_{<t})$ does not depend on $t$, it only depends on the actual sequence $w_{<t}$.
Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.

\begin{figure}
\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
	\caption{Illustration of (\ref{eq:listener-markov}). As the utterance unfolds, the listener maintains a memory state. After receiving word $w_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $w_t$.}\label{fig:listener-markov}
\end{figure}

\paragraph{Listener's Memory}
We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.

There are no assumptions about the memory architecture and the nature of its computations.
We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
At a given point in time, the listener's memory state $L_t$ is determined by the last word $w_t$, and the prior memory state $L_{t-1}$.
As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
This is formalized as a statement about conditional probabilities:
	\begin{equation}\label{eq:listener-markov}
		p(L_1| (w_{t})_{t \in \mathbb{Z}}, L_0)   = p((w_{t})_{t \in \mathbb{Z}} | L_0, w_1)
%p((w_{t})_t| L_0, L_1, w_1)   = p((w_{t})_t| L_0, w_1)
	\end{equation}
This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $w_1$.	
As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
For the study of memory in sentence processing, this seems fair. Discuss this more.

%
%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
%Formally: 
%\begin{equation}\label{eq:listener-markov-1}
%L_t \bot w_{>t} | w_{\leq t}
%\end{equation}
%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
%
%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $w_{t-1}$ and $L_{t-1}$:
%\begin{equation}\label{eq:listener-markov-2}
%L_t \bot w_{<t} | w_{t-1}, L_{t-1}
%\end{equation}
%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.

%The listener can trade off memory and future surprisal:
%A listener who chooses to store less memory will exerience higher surprisal in the future.
%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.

%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
%Consider a listener who uses $J$ bits of memory on average.
%What can we say about the listener's surprisal?

\begin{thm}\label{prop:suboptimal}
	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
	\begin{equation}\label{eq:memory}
		\sum_{t=1}^T t I_t
	\end{equation}
bits of memory on average.
Then this listener will incur surprisal at least
	$$H[w_t|w_{<t}] + \sum_{t > T} I_t$$
	on average.
\end{thm}
The proof is given in the appendix (REF).

\subsection{Proof of the Theorem}

%\begin{proof}
	We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $L_t$.
	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | L_t]$.
	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]$.
By the assumption of stationarity, we can rewrite this expression as
\begin{align*}
	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]\right) 
\end{align*}
	\begin{lemma}
For any $t$, the following inequality holds:
%	$$H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]$$
	\begin{equation}
%H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]
H[w_t | L_t] \geq H[w_t|w_{1 \dots t-1}, L_0]
		\end{equation}
	\end{lemma}
	\begin{proof}[Proof of the Lemma]

By Equation~\ref{eq:listener-markov}:
	\begin{equation}
p((w_{t})_t| L_0, L_1, w_1)   = p((w_{t})_t| L_0, w_1)
	\end{equation}
we have
	\begin{align*}
p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) = p(w_t|w_{<1}, L_0,  w_1, w_{2 \dots t-1}) \\
p(w_{<1}| L_0, L_1, w_1, w_{2 \dots t-1})   = p(w_{<1}| L_0, w_1, w_{2 \dots t-1})
	\end{align*}
Therefore
\begin{align*}
p(w_t|L_0, w_1, w_{2 \dots t-1}; L_1, w_{2 \dots t-1}) & = p(w_t|L_0, L_1, w_1, w_{2 \dots t-1}) \\
& = \sum_{w_{<1}} p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) p(w_{<1}| L_0, L_1, w_1, w_{2 \dots t-1}) \\
&= \sum_{w_{<1}} p(w_t|w_{<1}, L_0,  w_1, w_{2 \dots t-1}) p(w_{<1}| L_0, w_1, w_{2 \dots t-1}) \\
&= p(w_t|L_0,  w_1, w_{2 \dots t-1}) 
\end{align*}
Thus we have a Markov chain
\begin{equation}
(w_t) \rightarrow (L_0, w_1, w_{2 \dots t-1})   \rightarrow   (L_1, w_{2 \dots t-1})
\end{equation}
Thus, by the Data Processing Inequality,
	\begin{equation}
H[w_t| w_{2 \dots t-1}, L_{1}] \geq H[w_t|w_{1 \dots t-1}, L_0]
	\end{equation}

Finally, iteratively applying this inequality, we get:
		\begin{align*}
		H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq H[w_t| w_{t-2, t-1}, L_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]
		\end{align*}

	\end{proof}

	% = \int dw_{<1} p(w_t|w_{<1}, w_1, w_{2 \dots t-1})  = \int dw_{<1} p(w_t|L_0, w_{<1}, w_1, w_{2 \dots t-1}) = p(w_t|L_0, w_1, w_{2 \dots t-1}) $

%	$p(w_t|L_0, w_1, w_{2 \dots t-1}; L_1, w_{2 \dots t-1}) = p(w_t|L_0, L_1, w_1, w_{2 \dots t-1}) = \int dw_{<1} p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) = \int dw_{<1} p(w_t|w_{<1}, w_1, w_{2 \dots t-1})  = \int dw_{<1} p(w_t|L_0, w_{<1}, w_1, w_{2 \dots t-1}) = p(w_t|L_0, w_1, w_{2 \dots t-1}) $

	Plugging this inequality into the equality above:
\begin{align*}
	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, L_0] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
	& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | L_0] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
	& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|L_0]\right) 
\end{align*}
	The first term $I[w_{1\dots T}|w_{\leq 0}]$ can be rewritten in terms of $I_t$:
	\begin{align*}
		I[w_{1\dots T}|w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
	\end{align*}
	Therefore
\begin{align*}
	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|L_0]\right) 
	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[w_{1\dots T}|L_0]\right) \\
\end{align*}
	$I[w_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
	\begin{align*}
	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
		&= \sum_{t > T} I_t
\end{align*}
	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|L_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
%\end{proof}

Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.

### 2019-08-05 writeup/tables/medianDiff_1.tex
directories

English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-05 writeup/tables/medians_REAL_2.tex
directories

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tables/medians_ngrams_2.tex
directories

Persian & Polish & Romanian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Thai & Turkish & Ukrainian & Uyghur
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Vietnamese &  &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} &  &  & 
 \\

### 2019-08-05 writeup/tables/medians_small.tex
directories

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tables/medians_small_wide.tex
directories

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat & Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.05\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian & Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.05\textwidth]{neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak & Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.05\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.05\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tables/quantiles_1.tex
directories

Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-05 writeup/tables/quantiles_REAL_0.tex
directories

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tables/quantiles_REAL_1.tex
directories

Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tables/slice-hists_REAL_2.tex
directories

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2019-08-05 writeup/tradeoff.tex
directories

%We consider a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
%%Informally, we expect that both speaker and listener will be affected by memory demands:
%%rom the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
%For the listener, predicting the next word well requires maintaining information about the past.
%For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
%For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
%Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.
%
%
%TODO say at some point that we're studying sentence-internal memory

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{figures-gdrive/communication.png}
	\caption{Memory and language comprehension. (1) A speaker produces a sequence of words. (2) A listener maintains a representation of the words received so far. The listener can represent these at higher (left) or lower (right) levels of precision. (3) Throughout communication, the listener generates probabilistic expectations about the next word. Higher precision of memory representations leads to more precise predictions. (4) When receiving the next word, the listener incurs surprisal depending on the predictions. Higher levels of fidelity in memory lead to lower surprisal on average. \jd{what do the numbers (1) (2) etc refer to? also: image is blurry, at least on my screen}\note{make reference to $L_t$ inside the caption. you and the surprisals visually better aligned. maybe can color-code? } \mhahn{problem suggests that memory has to be contiguous substrings}}
	\label{fig:communication}
\end{figure}

We begin developing our model by considering the process of language comprehension illustrated in Figure~\ref{fig:communication}, where the listener Alice is processing a stream of words uttered by an interlocutor Bob (Figure~\ref{fig:communication}).

Experimental research has established that (1) listeners maintain information about the words received so far, (2) listeners forms probabilistic expectations about the upcoming words, and (3) words are easy to process to the extent that they are predictable in context \rljf{Good source for other cites here would be to dig around in Luke \& Christianson's review of predictability effects}. Furthermore, the exact mathematical form of the relationship between predictability and processing difficulty (as measured e.g. in reading times) seems to be given by the \key{information content} of a word in context \cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

Surprisal theory \citep{hale-probabilistic-2001, levy-expectation-based-2008} posits that the processing effort on a word $w_t$ in context $w_1\dots w_{t-1}$ is proportional to the \key{surprisal} of the word in context:
\begin{equation}
S = -\log \operatorname{P}(w_t|w_1\dots w_{t-1})\label{eq:surp}.
\end{equation}
Experimental work has confirmed that surprisal is a reliable and linear predictor of processing effort as reflected in reading times \citep{smith-effect-2013}.

The average processing difficulty associated with a language then comes out to its entropy rate:
\begin{equation}
	S = \operatorname{H}(w_t|w_1\dots w_{t-1}).
\end{equation}

However, surprisal theory as presented above cannot in principle account for effects of memory limitations on online processing, because Equation~\ref{eq:surp} represents surprisal as experienced by an idealized listener who accurately remembers the entire history of previous words $w_{1...t-1}$.
More realistically, human listeners deploy memory resources that maintain imperfect representations of the preceding context \citep{lewis-activation-based-2005, futrell-noisy-context-2017}.
If $m_{t-1}$ is a listener's memory state after hearing $w_1\dots w_{t-1}$, then the true surprisal experienced by the listener will be:
\begin{equation}
  \label{eq:lossy-surp}
  S_M := -\log_2 \operatorname{P}(x_t|m_{t-1}),
\end{equation}
which must be larger than Eq.~\ref{eq:surp} on average (see below for proof).
This modified surprisal $S_M$ in Eq.~\ref{eq:lossy-surp} can account for well-known memory effects on online processing such as dependency locality effects and structural forgetting \citep{futrell-noisy-context-2017,futrell2019information}.
The limitations of memory have well-documented empirical effects on language processing, causing difficulty above and beyond the difficulty associated with raw information content \cite{gibson1998linguistic,gibson1999memory,gibson2000dependency,vasishth2005activationbased,levy2013memory}.

The average processing difficulty associated with a language then comes out to a \key{cross entropy}:
\begin{equation}
	S = \operatorname{H}(w_t|m_{t-1}).
\end{equation}

These considerations imply a \emph{tradeoff between memory and surprisal}:
A listener maintaining higher-precision memory representations $m_{t-1}$ will, on average, incur lower surprisal, at the cost of higher memory load.
The idea of the memory-surprisal tradeoff is visualized in Fig.~\ref{fig:tradeoff}: for each desired level of average surprisal, there is a minimum number of bits of information which must be stored about context.
The shape of the trade-off is determined by the language, and in particular its word order:
some languages enable more efficient trade-offs than others by forcing a listener to store more bits in memory to achieve the same level of average surprisal.

Here we propose to model human language comprehension by treating the listener's memory $m_{i-1}$ as a lossily compressed representation of the context $x_{1...i-1}$, and the processing effort associated with each word using Eq.~\ref{eq:lossy-surp}.
Suppose that the listener's memory has a fixed information-carrying capacity $k$, then this listener has to compress the context into a memory representation $m_{i-1}$ with at most $k$ bits: $H[m_{i-1}] \leq k$.
Information theory implies that the processing cost $S_M$ trades off with the memory capacity $k$:
Processing cost will be lower for a listener who represents past input at higher levels of precision.

We prove a theorem describing how the shape of this tradeoff relates to language structure. %relating memory to the experienced comprehension difficulty $S$.
%Informally, the theorem says that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.
This theorem captures formally that languages support more efficient tradeoffs between short-term memory and comprehension difficulty when words are strongly predictive of their neighboring words.

We derive the precise form of the memory-surprisal tradeoff in Theorem 1.
Let $\operatorname{I}_t$ be the conditional mutual information between words that are $t$ steps apart, conditioned on the intervening words: 
\begin{equation}
	\operatorname{I}_t := \operatorname{I}[w_t, w_0 | w_1, \dots, w_{t-1}] = \operatorname{H}[w_t|w_1, \dots, w_{t-1}] - \operatorname{H}[w_t|w_0, \dots, w_{t-1}] 
\end{equation}

This quantity measures how much predictive information the word $t$ steps in the past contains about the current word, above and beyond the information contained in the intervening words.

This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.

This quantity, visualized in Figure~\ref{fig:theorem} (a), measures how much predictive information is provided by the next word $w_t$ by the word $t$ steps in the past.
It is a statistical property of the language, and can be estimated from large-scale text data.

Our theoretical results describe how the tradeoff between memory and comprehension difficulty relates to $I_t$ (Figure~\ref{fig:theorem} (b)):
Consider a listener who invests $B$ bits of memory into representing past input.
We then consider the smallest $T$ such that the area under the curve of $t I_t$, to the left of $T$, has size $B$.
Such a listener will experience average surprisal at least $H[w_t| w_{<t}] + \sum_{t=T+1}^\infty I_t$. %\note{maybe result first, and then say what $T$ is}
By tracing out all values $T >0$, one can obtain a bound on the tradeoff curve for any possible listener.

This is formalized in the following theorem:

\begin{thm}\label{prop:suboptimal}
Let $T$ be a positive integer, and consider a listener using at most $\sum_{t=1}^T t \operatorname{I}_t$ bits of memory on average.
Then this listener will incur average surprisal at least
$\operatorname{H}[w_t|w_{<t}] + \sum_{t > T} \operatorname{I}_t$.
\end{thm}

We provide a rigorous proof, making only minimal mathematical assumptions, in Appendix Section X.

An intuitive (non-rigorous) argument for this theorem goes as follows:
For each bit of information that is held in memory, we count the number of timesteps from when it is first entered into memory to when it is forgotten.
If a word $w_{T-t}$, $t$ steps in the past, contains one bit of information about the current word $w_T$, and if this information is not redundant with information in the intervening words, then this bit of information had to be maintained in memory for $t$ steps.
The amount of such information is $I_t$. This explains the factor $t$.

The theorem allows us to estimate the extra surprisal associated with each amount of memory capacity for a language.
The quantities $\operatorname{I}_t$ can be estimated as the difference between the cross-entropy of language models that have access to the last $t-1$ or $t$ words.
Given such estimates of $\operatorname{I}_t$, we estimate tradeoff curves as in Figure~\ref{fig:tradeoff} by tracing out $T=1, 2, \dots$.

Our result is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model. % and the mechanisms by which it computes predictions.
In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
Memory representations do not have to be rational or optimal for this bound to hold:
It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.

\begin{figure}
	(a)
\includegraphics[width=0.4\textwidth]{figures-gdrive/mi-distance.png}
	(b)
\includegraphics[width=0.2\textwidth]{figures-gdrive/theorem.png}
	(c)
\includegraphics[width=0.2\textwidth]{figures-gdrive/tradeoff.png}
	\caption{
		(a) Conditional mutual information $I_t$ captures how much predictive information about the next word is provided, on average, by the word $t$ steps in the past.
		(b) Here we illustrate our theoretical result. We plot $I_t$ (top) and $tI_t$ (bottom) as functions of $t$. A listener using $B$ bits memory (bottom) to represent prior observations will incur at least $A$ bits of extra surprisal beyond the entropy rate (top). \note{there is t and T. COnfusing}
		(c) \jd{FILL OUT} \note{make the colors in (c) difefrenfr from this in (b). Do want to make clear what the mapping from (b) to (c) is. Maybe can even have a pair of the two (b) curves, and then show what they map on.}
}\label{fig:theorem}
\end{figure}

%TODO
%Let $I_t$ be the Conditional Mutual Information of two words $w_0, w_t$ conditioned on the intervening words:
%\note{hard to swallow for general audience. maybe can just use more straightforward less mathy language.}
%
%
%\note{Alternative:}
%A listener will experience average surprisal at least
%\begin{equation}
%	H[w_t| w_{<t}] + \sum_{t=T+1}^\infty I_t
%\end{equation}
%where $T$ is chosen to be the smallest $T$ for which $\sum_{t=1}^T t I_t \geq B$.
%\note{End alternative.}

%We prove that, for any integer $T > 0$, a listener encoding $\sum_{t=1}^T t I_t$ bits of memory will incur average surprisal at least $H[w_t| w_{<t}] + \sum_{t=T+1}^\infty I_t$.
%maybe multiple sentences in def of It?
%define math, AND give people intuition.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures-gdrive/counterfactual-languages.png}
	\caption{\note{Can say more here to explain how it works} Estimating chance by constructing counterfactual grammars and languages.\jd{this figure is again blurry for me. also looks like there's sth in the background on the left?}}\label{fig:grammars}
\end{figure}

We now introduce our main theoretical result on memory-surprisal tradeoffs.

\begin{figure}
	\begin{center}
\includegraphics[width=0.45\textwidth]{figures/interpolate-curve.png}
\end{center}
	\caption{Estimating memory-surprisal tradeoff using the Theorem: We trace out the memory and surprisal values for all $T=1, 2, ...$, and linearly interpolate the curve.}\label{fig:interpolate}
\end{figure}

The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[w_t|w_{<t}] + \sum_{t > T} I_t$.
Then interplate linearly (justify this in appendix).
We obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
%
	\caption{Left: $I_t$ as a function of $t$, for two different processes. $I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past. Left: $t \cdot I_t$ as a function of $t$ for the same processes. }\label{fig:basic}
\end{figure*}

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
	\caption{Illustration for Proposition~\ref{prop:suboptimal}. Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal. In the given example, $T=4$. By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}. Recall that the red process had a faster decay of conditional mutual information. Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
\end{figure}

%\subsection{Formal Analysis and Proofs}

### 2019-08-06 writeup/first-draft/paper.tex

\documentclass[11pt,letterpaper]{article}

\usepackage{pslatex}
%\usepackage{latexsym}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage{rotating}
\usepackage{natbib}
\usepackage{amssymb}

\usepackage{tikz-dependency}
\usepackage{longtable}

\newcommand{\R}[0]{\mathbb{R}}
\newcommand{\E}[0]{\mathbb{E}}
\newcommand{\Ff}[0]{\mathcal{F}}

\usepackage{multirow}

\newcommand{\soft}[1]{}
\newcommand{\nopreview}[1]{}
\newcommand\comment[1]{{\color{red}#1}}
\newcommand\mhahn[1]{{\color{red}(#1)}}

\usepackage{amsthm}

\newcommand{\thetad}[0]{{\theta_d}}
\newcommand{\thetal}[0]{{\theta_{LM}}}

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{thm}[theorem]{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}

\frenchspacing
%\def\baselinestretch{0.975}

%\emnlpfinalcopy
%\def\emnlppaperid{496}

\title{Word Orders Trade Off Predictability and Memory Load Across Languages}%  Word Order Minimizes  Across Languages
\author{Michael Hahn, Judith Degen, Richard Futrell}
\date{2018}

\begin{document}

\maketitle

%
\begin{abstract}

Are languages optimized for processing with limited memory?
Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners. %assuming only that listeners perform incremental prediction.
Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners. % producing language and listeners predicting input.
\end{abstract}

%
%
%\begin{thm}\label{prop:suboptimal}
%	For each positive integer $t$, define
%$I_t := I[X_t, X_0 | X_{1\dots t-1}]$,
%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
%bits of memory on average.
%Then this comprehender will incur average surprisal at least
%	$H[X_t|X_{<t}] + \sum_{t > T} I_t$.
%\end{thm}
%

\section{Introduction}

A wide range of work has argued that natural language orders information in ways that reduce memory effort.
An early example is \cite{miller-finitary-1963}, who attributed the unaccebtability of multiple center embeddings in English to limitations of human working memory.
\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.

%TODO mention prior work
%
%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
%
%- explanation of greenberg universals
%
%- dependency length minimization
%
%- \cite{gibson-linguistic-1998}
%

%\paragraph{\cite{berwick-grammatical-1986}}
%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.

%also (CITE): operations in a specific incremental parsing model

All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.

We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
We will consider a general setting of a speaker producing sentences from a language, and a listener performing incremental prediction.
Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
Unlike many previous theories, our results apply to both speakers and listeners.
We will then use corpus data from over 50 languages to provide evidence that their word orders help lower memory cost.

\section{Two Memory-Surprisal Tradeoffs}

With minimal assumptions, we will use information theory to derive two tradeoffs:

(1) a tradeoff between \emph{speaker memory} and (listener) surprisal

(2) a tradeoff between \emph{listener memory} and (listener) surprisal

In the second part of the paper, we examine whether word orders in natural language optimise these two tradeoffs.

\subsection{Memory and Locality in General Sequence Models}

We consider a \emph{speaker} who produces sentences and a \emph{listener} who, as the speaker's utterance unfolds, engages in incremental prediction.
Informally, we expect that both speaker and listener will be affected by memory demands:
From the speaker's perspective, producing well-formed sentences will require information about what she has uttered so far.
From the listener's perspective, predicting the next word well requires maintaining information about the past.
For the listener, the quality of prediction is measured by the average \emph{surprisal} experienced.
For a fixed language, we can ask how much information about the past (1) the speaker has to maintain to produce well-formed utterances, and (2) the listener has to maintain to incur a minimal amount of surprisal.
Utilizing the tools of information theory, we quantify memory in \emph{bits}, obtaining bounds that hold across different models of memory architecture and ways of quantifying memory load.

We now make this more formal:
We think of the language as a stochastic process $(X_t)_{t \in \mathbb{Z}}$, that is, a probability distribution over sequences of words $\dots x_{-2} x_{-1} x_0 x_{1} x_{2} \dots$.
As the elements are indexed by the full set of integers, we think of the process as extending infinitely into both the past and the future; that is, we model a setting where the speaker has been speaking and will continue to speak for an infinite amount of time.
This assumption is for mathematical convenience and does not affect the substance of our results: As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
Second, we make the assumption that this process is \emph{stationary}.
Formally, this means that the conditional distribution $P(X_t|X_{<t})$ does not depend on $t$, it only depends on the actual sequence $X_{<t}$.
Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.

The speaker maintains a memory state $S_t$, which evolves over the timesteps $t$.
We will not make any assumptions about the content of $S_t$ and how the speaker computes it.
The only assumption is that the speaker's past and future utterances are independent when conditioning on the memory state:
\begin{equation}\label{eq:markov-speaker}
X_{>t} \bot X_{\leq t} | S_t
\end{equation}
This means that the memory state $S_t$ functions as a `bottleneck', encoding all information about the past that influence the speaker's choices in the future.

What can we say about the amount of information that the speaker has to store when producing utterances -- that is, how many bits does it take to encode the speaker's  memory states $S_t$ and $L_t$?
Our first theoretical result sheds light on this; we will show that the amount of bits needed to encode the memory state is related to the decay of mutual information over distances.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/listener-it.png}
	\caption{(TODO adapt to format) Intuitive explanation of $I_t$}
\end{figure*}

We will use the concept of \emph{conditional mutual information}.
For two random variables $X, Y$, the mutual information $I[X,Y]$ is the reduction in entropy about $X$ by knowing $Y$:
\begin{equation}
	I[X,Y] := H[X] - H[X|Y]
\end{equation}
For three random variables $X, Y, Z$, the mutual information between $X$ and $Y$, conditional on $Z$, is this entropy reduction when $Z$ is already known:
\begin{equation}
I[X,Y|Z] := H[X|Z] - H[X|Z,Y]
\end{equation}
We will use the specific form where $X$ and $Y$ are observations $X_0$, $X_t$ of the process $(X_t)_{t \in \mathbb{Z}}$, and $Z$ is a block of intervening observations $X_1, \dots, X_{t-1}$:
\begin{equation}
	I[X_t, X_0 | X_1, \dots, X_{t-1}] := H[X_t| X_1, \dots, X_{t-1}] - H[X_t| X_0, X_1, \dots, X_{t-1}]
\end{equation}
This is equal to the reduction in uncertainty about the $t$-th observation when knowing the $0$-th observation, in addition to the block of intervening observations.
That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.

Using this concept, we can now state our first main result, concerning the memory load of a speaker: %, and of a listener performing optimal incremental prediction:

\begin{proposition}\label{prop:lower-bound}
	A speaker producing samples from $(X_t)_{t \in \mathbb{Z}}$ needs to maintain at least
\begin{equation}\label{eq:memory-bound}
M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}
bits of memory on average.

\end{proposition}

Also can relax this to imprecise production: A speaker with low memory must deviate strongly from the language, measured in KL divergence.

%that the model does not have an `absolute clock' built into it.
The quantity $M$ is equal to the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$, which is often referred to as the \emph{excess entropy} of the process (CITE).

Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
This formalizes a general, assumption-free, link between memory and locality in language production.
In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
%incremental prediction.

The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model. % and the mechanisms by which it computes predictions.
In particular, while the relation to psycholinguistic and psychological models of how memory works will be interesting to explore, our result applies to any such model.
Memory representations do not have to be rational or optimal for this bound to hold:
It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.

We will write $I_t$ as an abbreviation for $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$.
The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
%This confirms the intuition that finiteness of memory entails that the contribution of long-term

We have treated the process $(X_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
The analysis is not even restricted to discrete timesteps: We can replace the sum in~(\ref{eq:memory-bound}) with an integral to get a continuous-time version.

We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
However, $I_t$ decays more quickly for Process A (red).
This means that predictive information about an observation is concentrated more strongly in the recent past.
In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
Note that the area under the curve is equal to (\ref{eq:memory-bound}).
This area is smaller for the red process, as $I_t$ decays more quickly there.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
%
	\caption{Left: $I_t$ as a function of $t$, for two different processes. $I_t$ decays faster for the red process: Predictive information about the present observation is concentrated more strongly in the recent past. Left: $t \cdot I_t$ as a function of $t$ for the same processes. The area under the curves is given by (\ref{eq:memory-bound}). This area is smaller for the red process, as $I_t$ decays more quickly there.   }\label{fig:basic}
\end{figure*}

\subsection{Memory and Dependency Length}

We illustrate the linguistic predictions of Proposition \ref{prop:lower-bound} by reanalyzing the data from \cite{fedzechkina-human-2017}.
This is a miniature artificial language study that showed a bias for Dependency Length Minimization in production in artificial language learning.
Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in~(\ref{eq:memory-bound}).

As ~(\ref{eq:memory-bound}) is invariant under reversal of the language, we only consider the head-final version of her artificial language.
The language has consistent head-final order, and uses case marking on objects.
The relevant production targets are transitive sentences where one of the two arguments is much longer than the other, due to the presence of a PP modifier, as shown in Table~\ref{tab:artificial}.
The language has variable order of subjects and objects; for the production targets, the B versions produce much longer dependencies than the A versions.
Dependency Length Minimization thus predicts that speakers are more likely to use the A versions.
\cite{fedzechkina-human-2017} confirmed this experimentally.

\begin{table}
	\textbf{A Orders: Short Dependencies}

		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun\end{tabular}} & \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			\fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
					\end{tabular}} & \fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & Verb \\
		\end{tabular}
\\
\\

	\textbf{B Orders: Long Dependencies}

		\begin{tabular}{ccc}
			Subject & Object & Verb \\
			 \fbox{\begin{tabular}{l}Noun\end{tabular}} &  \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun-Case
		\end{tabular}}  &  \fbox{\begin{tabular}{l}Verb\end{tabular}}  \\
		\end{tabular}
\\
		\begin{tabular}{ccc}
			Object & Subject & Verb \\
			\fbox{\begin{tabular}{l}Noun-Case\end{tabular}} & \fbox{\begin{tabular}{llllll}
				\fbox{\begin{tabular}{llll} Adjective &Noun &Postposition\end{tabular}} & Noun
		\end{tabular}} & Verb \\
		\end{tabular}

			\caption{Production targets in the artificial mini language from \cite{fedzechkina-human-2017}. The language has head-final order, with free variation between SO and OS orders. When one of the arguments is much longer than the other, placing the longer one first (`A' orders) shortens syntactic dependencies, compared to `B' orders.}\label{tab:artificial}

\end{table}

In this section, we show that our bound on speaker memory makes the same prediction, without reference to syntactic structure or specific memory architectures.

We constructed one language consisting of the A versions, and one language consisting of the B versions.
Following the experimental setup of \cite{fedzechkina-human-2017}, we assigned equal probability to the two possible configurations per language, and used a separate set of nouns (inanimate nouns) for the embedded noun in the long phrase.

We interpreted each of the two languages as a stationary processes, extending infinitely in both directions, by concatenating independent samples drawn from the language.
			We computed (\ref{eq:memory-bound}) from a chain of 1000 independently sampled sentences, for each of the two versions of the toy language.
			Figure~\ref{fig:toy-mis} (left) shows the curve of the conditional mutual information $I_t$ as a function of the distance $t$.
			The curves differ at $t=2$ and $t=5$: 
			About 0.073 nats of predictive information that are at distance $t=2$ in the A orders are moved to $t=5$ in the B orders.
%			In this sense, A orders have greater information locality than the B orders.
			%What is responsible for this difference?
			The source of the difference lies in predicting the presence and absence of a case marker on the second argument -- i.e., whether to anticipate a subject or object.
			In the A orders, considering the last two words is sufficient to make this decision.
			In the B orders, it is necessary to consider the word before the long second constituent, which is five words in the past.

			The total amounts of predictive information -- corresponding to the area under the curve -- are the same, indicating that both languages are equally predictable.
			However, we will see that the memory demands are different:
			Figure~\ref{fig:toy-mis} (right) shows $t\cdot I_t$ as a function of $t$.
			As $I_t$ decays faster in A orders, the total area under the curve now differs between A and B, and is larger in B.
			This area corresponds to the lower bound in (\ref{eq:memory-bound}), and is 2.21 nats in A orders, and 2.43 nats in B orders.
			While (\ref{eq:memory-bound}) is a general lower bound, it can be proven that this bound is actually tight in the case of this specific example.\footnote{This can be shown by computing the causal states and then showing that the crypticity is zero, both of which is tractable in the case of this small-scale artificial language.}
That is, a speaker who optimally allocates memory resources will spend 2.21 nats in A orders, and 2.43 nats in B orders.
It is important to stress that, even though we computed this value by considering the number of words impacting predictions at a given point in time, this bound holds independently of the actual implementation and architecture of memory and predictions.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{toy/toy-mis.pdf}
\includegraphics[width=0.45\textwidth]{toy/toy-t-mis.pdf}
%
	\caption{Left: Decay of Conditional Mutual Information, as a function of the distance $t$, for the two versions in the artificial language. The areas under the two curves are identical, corresponding to the fact that both orders are equally predictable. However, mutual Information decays faster in Order A.\ \ \ Right: $t I_t$, as a function of $t$. The area under the B curve is larger, corresponding to larger memory demand for this order.}\label{fig:toy-mis}
\end{figure*}
%Both languages have the same overall entropy rate, but they differ in the distribution of predictive information.
%
%plot of $I_t$
%
%The areas under the curves are identical.
%
%good version
%
%%CONTEXT LENGTH 0   2.06856758681  9997.93143241   0.0
%%CONTEXT LENGTH 1   0.29195106479  1.77661652202   1.77661652202
%%CONTEXT LENGTH 2   0.0729865489103  0.21896451588   2.21454555377
%%CONTEXT LENGTH 3   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 4   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 5   0.0729865489103  0.0   2.21454555377
%%CONTEXT LENGTH 6   0.0729865489103  0.0   2.21454555377
%
%bad version
%
%%CONTEXT LENGTH 0   2.06937301571  9997.93062698   0.0
%%CONTEXT LENGTH 1   0.291673373027  1.77769964269   1.77769964269
%%CONTEXT LENGTH 2   0.145830550934  0.145842822093   2.06938528687
%%CONTEXT LENGTH 3   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 4   0.145830550934  0.0   2.06938528687
%%CONTEXT LENGTH 5   0.0729152754672  0.0729152754672   2.43396166421
%%CONTEXT LENGTH 6   0.0729152754672  0.0   2.43396166421
%%CONTEXT LENGTH 7   0.0729152754672  0.0   2.43396166421
%%
%
%
%
%%
%%grammar:
%%
%%S $\rightarrow$ Obj Subj V (1/2) | Subj Obj V (1/2)
%%
%%Obj $\rightarrow$ NP di
%%
%%Subj $\rightarrow$ NP
%%
%%NP $\rightarrow$ N (3/4) | PP NP (1/8) | Adj NP (1/8)
%%
%%PP $\rightarrow$ NP P
%%
%
%

\subsection{Languages can trade off Memory Load and Predictability}
%\subsection{Speakers can can minimize Memory Load at Fixed Information Rate}

We have derived a general lower bound on the memory load of a speaker in Proposition~\ref{prop:lower-bound}.
We now relate the speaker's memory to the surprisal experienced by the listener.

The average surprisal of a language $(X_t)_{t \in \mathbb{Z}}$, as experienced by a listener performing optimal prediction, is equal to the entropy rate $H[X_t|X_{<t}]$.
This can be written as the unigram entropy minus the mutual information between a single observation $X_t$ and the past:
\begin{equation}
	H[X_t|X_{<t}] = H[X_t] - I[X_t, X_{<t}]
\end{equation}
where the unigram entropy $H[X_t]$ only depends on the frequencies of individual words, independently of the language's word order.

We will refer to the mutual information $I[X_t, X_{<t}]$ as the \emph{predictability} of the process.
The predictability $P := I[X_t, X_{<t}]$ is a lower bound on $M = I[X_{\geq t}, X_{<t}]$, and thus on speaker memory:
\begin{equation}
P \leq M
\end{equation}
and thus\footnote{More generally 
$H[X_t|X_{<t}] \geq \frac{1}{T} \left(H[X_{1\dots T}] - M\right)$ for any $T$. Thus, the following relation holds for any $T$: $$T \cdot \text{ListenerSurprisal} \geq H[X_{1\dots T}] - \text{SpeakerMemory}$$}
\begin{equation}
	Listener\ Surprisal \geq	H[X_t|X_{<t}] \geq H[X_t] - M \geq H[X_t] - Speaker\ Memory
\end{equation}
This means that, when we consider different languages that share a single unigram distribution, memory and surprisal trade off:
Languages that are highly predictable  require a good amount of speaker memory; a language that can be produced with very small memory cannot be very predictable.
In particular, a language that requires zero speaker memory must have zero predictability and maximal average surprisal.

We thus obtain a tradeoff between two different types of costs facing speakers and listeners: The \emph{speaker's} memory trades off with the \emph{listeners'} surprisal.

We can visualize this by placing languages into a \emph{memory-surprisal-plane} (Figure~\ref{fig:toy-mis}). Each language is assigned its entropy rate as its $y$ coordinate, and the memory load necessary to achieve optimal surprisal as its $x$ coordinate.
We will continue to assume that all languages share a single unigram distribution -- this will in particular apply when studying different word order variations of the same language.

The plane can be separated into an \emph{achievable} region, populated by languages, and an \emph{unachievable} region in which no languages can occur.
No language can have zero surprisal with zero memory -- the corner of the plane thus must fall in the unachievable region.
The boundary between the two regions is the \emph{Pareto frontier}.
Languages that fall on the boundary optimally trade off surprisal and memory; they cannot further decrease one without increasing the other.
The smallest possible values for this boundary consist of the points satisfying $P=M$, indicated by a straight diagonal line in Figure~\ref{fig:toy-mis}.
Only part of the achievable region will be populated by languages that could plausibly be natural languages: For instance, we may expect natural languages to have long-term dependencies that will prevent them from being at the absolute Pareto frontier.
We hypothesize that, if languages optimize the tradeoff between memory and surprisal, they will be close to the optimal boundary of a region populated by a `comparison class' of languages that share essential characteristics of natural language, but have other word order rules.
%When We will specify a class of baseline languages that we will argue to form a reasonable comparison class for natural language.
%Within this, we hypothesize that natural language is close to the Pareto frontier.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/pareto-scheme.pdf}
%
	\caption{Memory-Surprisal Tradeoff. The straight line represents the Pareto frontier for unconstrained languages. Languages resembling natural languages inhabit a subset of the achievable region (grey oval). Its lower-left boundary represents the Pareto frontier relevant for natural language. We hypothesize that natural language lies close to this frontier.   }\label{fig:toy-mis}
\end{figure}

\subsection{Listeners can trade off Memory Load and Surprisal}\label{sec:listener}

%TODO redo curves so that x-axis is surprisal, y-axis the necessary memory

We have analyzed the memory load of a speaker and related it to the listener's surprisal.
We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
Similar to the listener maintains a memory state $L_t$; again, there are no assumptions about the memory architecture and the nature of its computations.
%We assume that the listener has no access to the speaker's state beyond what the speaker has already uttered.
We only make two basic assumptions about the flow of information:

First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
Formally: 
\begin{equation}\label{eq:listener-markov-1}
L_t \bot X_{>t} | X_{\leq t}
\end{equation}
This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.

Second, we assume that $L_t$ contains no information about the past beyond what is contained in $X_{t-1}$ and $L_{t-1}$:
\begin{equation}\label{eq:listener-markov-2}
L_t \bot X_{<t} | X_{t-1}, L_{t-1}
\end{equation}
This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.

The listener can trade off memory and future surprisal:
A listener who chooses to store less memory will exerience higher surprisal in the future.
A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.

We will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
Consider a listener who uses $J$ bits of memory on average.
What can we say about the listener's surprisal?
We return to the information and memory curves in Figure~\ref{fig:basic}.
In the graph of $t \cdot I_t$, we look for the first $T$ such that the area under the curve to the left of $T$ has size $\geq J$.
This is illustrated in Figure~\ref{fig:listener-tradeoff-decay} (right).
Then let $\epsilon$ be the area under the curve of $I_t$ to the right of $T$ (Figure~\ref{fig:listener-tradeoff-decay}, left).
We can prove that such a listener must incur surprisal at least $\epsilon$ greater than a listener with perfect memory.
As before, we write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$
Then

\begin{proposition}\label{prop:suboptimal}
	Let $T$ be a positive integer, and consider a listener using at most
$$\sum_{t=1}^T t I_t$$
bits of memory on average.
Then this listener will incur surprisal at least
	$$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
	on average.
\end{proposition}

The proposition gives us a lower bound on the listener's memory-surprisal curve: Taking all pairs of memory $\sum_{t=1}^T t I_t$ and surprisal $H[X_t|X_{<t}] + \sum_{t > T} I_t$, we obtain a curve in memry-surprisal plane, which is a lower bound on the memory demands of any listener at a given surprisal level.
We visualize this for the two processes from Figure~\ref{fig:basic} in Figure~\ref{fig:listener-tradeoff}.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
	\caption{Illustration for Proposition~\ref{prop:suboptimal}. Listeners can trade off memory and surprisal: A listener only investing memory of the amount given by the black area on the right will incur at least the black area on the left in additional surprisal. In the given example, $T=4$. By varying $T$, the two areas describe the listener's memory-surprisal tradeoff curve.}\label{fig:listener-tradeoff-decay}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}. Recall that the red process had a faster decay of conditional mutual information. Correspondingly, this figure shows that a listener can achieve lower surprisal at the same level of memory load.}\label{fig:listener-tradeoff}
\end{figure}

In Figure~\ref{fig:toy-listener-tradeoff}, we show the resulting curve for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
In this language, speaker memory coincides with the memory demand of a listener who performs optimally in incremental prediction.
The curve shows that, at any desired level of surprisal, Order A requires at most as much memory as Order B.
For reaching optimal surprisal, Order A requires strictly less memory.
Thus, in this case, the listener's surprisal-memory tradeoff is optimized by the orders predicted by Dependency Length Minimization.

\begin{figure}
\includegraphics[width=0.45\textwidth]{toy/toy-mem-surp.pdf}
	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}. Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
\end{figure}

\paragraph{Center Embeddings}

\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.

%\cite{gibson-linguistic-1998}

%first center-embeddings, POS level model

%Later work found evidence that cross-serial dependencies are less taxing.

%now cross-serial: POS level model things are the same. But lexical level, assume PMI and dependencies. Then listener's tradeoff is better for cross-serial dependencies

\subsection{Relation to Theories of Sentence Processing}

\paragraph{Lossy-Context Surprisal}
\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
In particular, they consider loss models that delete, erase, or replace words in the past.
Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
A listener will incur surprisal
$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.

\paragraph{Relation to Dependency Locality}
The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.

\paragraph{Center Embeddings, ...}
Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures

Miller and Chomsky 1963

Frzier 1985 local nonterminal count

Rambow and Joshi 1994 using TAG

Marcus 1980 deterministic parsing

(Sabrina Gerth, Memory Limitations in Sentence Processing)

\subsection{Relation to other work}

\paragraph{Statistical Complexity}
There are deep connections between our formalization of speaker and listener memory and studies of dynamic systems in the Physics literature.
Our formalization of listener memory, under the assumption of optimal prediction, is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
The quantity in (\ref{eq:memory-bound}) is equal to the \emph{excess entropy}, and it has long been known to bound statistical and generative complexity \citep{crutchfield-inferring-1989}.
However, the link between memory and locality provided by our series expansion in (\ref{eq:memory-bound}) appears to be a novel contribution.

\cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.

\paragraph{Decay of Mutual Information}
In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[X_t, X_0 | X_{1\dots t-1}]$.
Prior work has studied the decay of \emph{unconditional} mutual information $I[X_t, X_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.

The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
Consider the constant process where with probability 1/2 all $X_t = 0$, and with probability 1/2 all $X_t = 1$. %%$X_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR. This one has bounded nonzero $J$, but zero unconditional MI. To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $X_t$ is equal to the XOR of $X_{t-1}$ and $X_{t-N}$, such that each $X_t$ has $Bernoulli(1/2)$ as its marginal. The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}

\paragraph{Long-range dependencies in text}    % excess entropy
\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
Our work contrasts with this work in that we are interested in dependencies within sentences.

\subsection{Discussion}

\paragraph{Speakers vs Listeners}
Our results apply both to speakers and listeners.

\paragraph{Decay vs Interference}
Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
This is not the case:
A long dependency can be maintained perfectly with low average memory:
Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.

\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
In fact, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory (and thus in particular finite excess entropy):

\begin{proposition}
Any language generated by a PCFG, with rational weights, and finitely many terminals and nonterminals, has finite average memory demand in production and incremental prediction.
\end{proposition}

\begin{proof}[Proof Sketch]
	We provide an algorithm that performs generation or optimal prediction and has finite average memory use. We consider incremental Earley parsing \citep{earley-efficient-1970}. Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory (where the constants depend on the grammar but not $n$). For storing probabilities, the storage will be similarly bounded as long as the weights of the grammar are rational.
	The result then follows from Proposition 2 in \cite{chi-statistical-1999}, which says that any function that is polynomial in the length of words drawn from a PCFG has finite expectation.\footnote{Using Proposition 2 of \cite{chi-statistical-1999}, one can also directly show that excess entropy is finite for any PCFG (without assumption about the weights), if we interpret it as giving rise to a stationary process consisting of a sequence of words independently drawn from the language.}
\end{proof}

\section{Heavy NP Shift}

Phenomenon linked to locality/memory (Hawkins)

Is the listener's memory-surprisal tradeoff optimized by Heavy NP Shift?

Penn Treebank

Trained 20 language models counterfactual version

then estimate memory on real and reverse orderings

3) Describe the key dependent variable(s) specifying how they will be measured.
Collect all instances of a VP consisting of a verb, and (in some order) an NP and a PP, from the Penn Treebank (WSJ) training set. Calculate the listener's memory-surprisal tradeoff on the words belonging to the NP and the PP.

5) Specify exactly which analyses you will conduct to examine the main question/hypothesis.
(1) Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs? Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP? This will be tested by bootstrapping over language models and examples.
(2) Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?
Also tested by bootstrapping over language models and examples.

6) Describe exactly how outliers will be defined and handled, and your precise rule(s) for excluding observations.
Language models with overall validation average surprisal > 10 will be excluded (this didn't happen).

7) How many observations will be collected or what will determine sample size?
20 LSTM language models are constructed on a version of Penn Treebank where for all verb-initial VPs, the order of the remaining daughter constituents is shuffled.

Is the listener's memory-surprisal tradeoff significantly better on real as opposed to inverted order of the NPs and PPs?

Is this separately true for cases where NP-PP is the real ordering, and for cases where it is PP-NP?

Is the listeners memory-surprisal tradeoff significantly better when placing the shorter constituent (of this NP and PP) before the longer one?

%
%
%\section{NP Structure}
%
%- adjective order
%
%- noted by ...: Det - Num - Adj
%
%
%This is a case where the predictions differ from DepL
%
%
%E.g., if the choice of determiners, numerals, and adjectives are statistically independent when conditioning on the noun, will predict the effect.
%Computationally verify the effect.

\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}

We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
We will define a class of baseline languages that serve as a comparison class of word orders that natural language can be compared to.

Ordering grammars are defined as in the Optimization paper: For each UD relation type, define a weight. To get random baseline languages, randomize the weights.
All grammars are valuated deterministically.

We found that, across treebanks, random grammars that have more fine-grained rules or are evaluated nondeterministically do worse in both memory-surprisal tradeoffs.
Thus, random grammars with by-relation weights constitute a stronger baseline.

%Things that one may ponder: (1) what are even stronger baselines? (2) probably should do a control experiment using maximum-likelihood fits

%\subsection{Prediction on the level of POS}
%As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences. %, following~\cite{futrell-noisy-context-2017}.
%As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.
%
%%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
%%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
%%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
%%
%%%\mhahn{TODO results across languages}
%%
%%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
%
%Results for four languages are shown in Figure~\ref{fig:memory-results}.
%Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
%%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).
%
%
%%%For half of the languages
%%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
%%For 37 \% of the languages, it also requires less memory than the original language.
%
%%We found that, as predicted, memory usage in real languages was correlated with word order freedom:
%
%
%%\begin{figure*}
%%\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%%%
%%\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
%%\end{figure*}
%%

\subsection{Method}
The small size of available syntactic data for many languages means that estimating mutual information is problematic.

In general, we expect a bias-variance tradeoff:

A method that estimates on a coarse-grained level (e.g. POS) will result in estimates with relatively higher fidelity and lower variance, potentially making it easier to localize the actual language in the space of counterfactual languages.
However, it introduces \emph{bias} by describing a simplified version of the language.

We propose to therefore compute the quantities on levels less likely to suffer from data sparseness: part-of-speech and morphology.

In the UD treebanks, each word has part-of-speech and morphological information.
The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token that indicates that information related to the word ends.
We further provide the word's lemma after its POS tag, if it is one of the 50 most frequent lemmas of the language.
Each POS tag and morphological feature-value pair is encoded as an atomic token.
The rationale for including frequent lemmas is that such words often are strongly grammaticalized and may provide grammatical information (and frequent words are less likely to be affected by data sparseness).

To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling sequences such as natural language.
We will discuss alternative estimation methods below.

During incremental prediction, the LSTM sequentially reads these tokens and computes a distribution over the next word.
The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).

Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.

Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
Therefore, to compute this quantity, we compute the differences between the surprisals
$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.
We cut $T$ off at 30.

\paragraph{Alternative Models}
Recent research has proven strong theoretical learnability guarantees for neural network models (CITE), supporting their success in practice.
They also provide the strongest predictors of the surprisal effect on human reading times \citep{frank-insensitivity-2011, goodkind-predictive-2018}.

In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:

First, an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations. Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent. Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.

A statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar. The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model. One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations. The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.

\subsection{Data}
We considered all languages for which there are Universal Dependencies 2.2 treebanks with a total of at least 400 sentences of training data.
We excluded data from historical languages.\footnote{Ancient Greek, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
We excluded four languages for a later confirmatory experiment (English, Korean, Russian). % Irish
This resulted in X languages.

For most language, we used the predefined data split.
For some languages, where there was no training partition, or where the training partition was smaller than the test partition, we redefined the split to obtain more training data:
For these languages, we randomly pooled all the available partitions, used 100 sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}

\subsection{Setup}
The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
This biases the hyperparameters towards favoring counterfactual grammars.

This experiment is not fully confirmatory, since some data was collected for some languages before the predictions and the setup of the experiment were finalized.

For each language, we collected data from the actual orderings and from random grammars.
We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
We collected data from the actual and random orderings in proportion one to three.
The stopping criterion will be described below.

\paragraph{Speaker Memory: Distance from Pareto Frontier}
We want to test: Are languages closer to the Pareto frontier than baseline languages tend to be?

For this, we need a way of quantifying distance to the Pareto frontier.
We suggest to measure the distance as the fraction of baseline languages that are better on both dimensions.

More formally, fixing a natural language, let $P_0$ be the distribution of baseline languages mapped to the surprisal-memory plane, viewed as a probability distribution over the positive quadrant of the plane $\mathbb{R}_+^2$.
For any point $x \in \mathbb{R}_+^2$, let $A_x$ be the fraction of baseline languages that are better on both dimensions.
Similarly, let $B_x$ be the fraction of baseline languages that are worse on both dimensions.
Using the probability distribution $P_0$, we can write:
$$A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$$
$$B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$$
%Ways of using this to test and quantify for distance (I think they are all reasonable and worth reporting in SI. Question is which ones to rely on mostly for the main part):
Futher, let $P_1$ be the distribution in the surprisal-memory plane that is generated by the results obtained for the real orderings, with different random initializations.
For each language, we \emph{test} the hypothesis that the language optimizes the tradeoff by considering the null hypothesis that the real orderings do not Pareto-dominate more baseline orderings than there are baseline orderings dominating them:
$$\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$$

We propose to \emph{quantify} closeness to the frontier using the following two statistics:

\begin{enumerate}
%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$

		This has the following interpretation: Among orderings that either dominate or are dominated by the real orderings, what fraction is dominated by the real orerings?
		This quantity is $1$ if the real orderings are not Pareto-dominated by any baseline order, but do dominate some baseline orders.
		The null hypothesis defined in our test amounts to the inequality $\E_{x \sim P_1}E_x \leq 0.5$.

A drawback of this quantity is that it is meaningless when the real orders do not dominate, and are not dominated by, any real order.
The following quantity is always maningful:

	\item Compute $A_{x'}$ for each $x'$ from the baseline distribution $P_0$, and then calculate the percentile rank of $\E_{x \sim P_1}[A_x]$ among these values:
		$$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$$
This statistic describes how many random orders achieve closer distance to the frontier as measured by $A_x$.
		Stated differently, it quantifies how `extraordinary' the value $\E_{x \sim P_1}[A_x]$ is in relation to values typically attained for $x' \sim P_0$.
The statistic is close to zero if $\E_{x \sim P_1}[A_x]$ is smaller than for almost all samples from $P_0$.

\end{enumerate}

We will obtain confidence intervals for the statistics via bootstrapping, and use a confidence interval for $B_x-A_x$ for testing $B_x \leq A_x$.
In the appendix, we report simulations that justify the use of bootstrapping in this situation.

\paragraph{Listener Memory: Surprisal-Memory Tradeoff}
We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
For each model for each language, we interpolate the surprisal-memory tradeoff curves given by Proposition~\ref{prop:suboptimal} linearly.
For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.

We test by considering the null hypothesis that $\E_{x \sim P_1}[W(x)] \leq 1/2$ where $P_1$ is the distribution over values obtained for real orderings.

We use a bootstrapped confidence interval for $\E[W]$ for testing and for quantifying the degree of optimization.

%- bootstrapping
%- subsampling
%- permutation test / rank test ??
%

\paragraph{Stopping Criterion}

For each language, we collected at least 20 data points for real and baseline orderings each, and continued collecting until the CI for $D$ had width $\leq 0.3$, and the CI for $W$ had width $\leq 0.15$.
We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.

%- at least 30 samples from both baseline and real
%
%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.

%
%(1) is bigram MI always greater in real languages?
%
%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?

\subsection{Results}

Results are reported in Appendix~\ref{sec:results-table}.

\paragraph{Surprisal and Speaker Memory}
In all but X languages, the null hypothesis $\E_{x \sim P_1} [B_x] \leq \E_{x \sim P_1}[A_x]$ was rejected.
Also, in all but X languages, $D < 0.5$.

\paragraph{Surprisal and Listener Memory}
In all but X languages $\E[W] > 0.5$ significantly.

\section{Experiments on Further Treebanks}

Held-out:

- UD languages:

-- English, Korean, Russian

-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment)

-- character-level Russian

- other treebanks

-- Penn treebank \citep{marcus-building-1993}

-- spoken English (T{\"u}ba-E/S)

-- spoken German (T{\"u}ba-D/S)

-- spoken Japanese (T{\"u}ba-J/S)

-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017}

-- Chinese treebank \citep{xue-chinese-2013}

-- another Chinese dependency treebank LDC2012T05

-- furher (2.2/2.3 Japanese treebanks)

Due to the sizes of these treebanks, can also do experiment with full word forms.

\bibliographystyle{apalike}
\bibliography{literature}

\appendix

\section{Proofs}

\begin{proof}
	The number of bits of memory is given by the entropy $H[S_t]$.
	In view of (\ref{eq:markov-speaker}) and the Data Processing Inequality, this is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
	I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I[X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}] \\
	&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I[X_i, X_j|X_{j+1}...X_{i-1}] \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
where we used stationarity of the process in the last step.
\end{proof}

\begin{proof}
We denote the listener's memory state after hearing $X_{<t}$ by $L_t$.
The average number of bits required to encode this state is $H[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $H[X_t | L_t]$.
The difference between the listener's surprisal and optimal surprisal is thus at least $H[X_t | L_t] - H[X_t | X_{<t}]$.
By stationarity, we can rewrite this expression as
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | L_t] - H[X_t | X_{<t}]\right) 
\end{align*}
For any $t$,
	$$H[X_t | L_t] \geq H[X_t| X_{t-1}, L_{t-1}] \geq ... \geq H[X_t|X_{1 \dots T}, L_0]$$
	This can be derived from the two conditions \ref{eq:listener-markov-1} and \ref{eq:listener-markov-2} (TODO explain how).

Therefore
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq   \frac{1}{T} \left(H[X_{1\dots T} | L_0] - H[X_{1\dots T} | X_{\leq 0}]\right)  \\
	& = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|L_0]\right) 
\end{align*}
	The first term $I[X_{1\dots T}|X_{\leq 0}]$ can be rewritten in terms of $I_t$:
\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|L_0]\right) 
	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[X_{1\dots T}|L_0]\right) \\
\end{align*}
	$I[X_{1\dots T}|L_0]$ is at most $H[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
	\begin{align*}
	H[X_t | L_t] - H[X_t | X_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
		&= \sum_{t > T} I_t
\end{align*}
	Rearranging shows that the listener's surprisal is at least $H[X_t|L_t] \geq H[X_t | X_{<t}] + \sum_{t > T} I_t$, as claimed.
\end{proof}

\section{Bootstrapping for Distance to Frontier}

TODO write up

We conducted simulations in the following setting:

sample two ellipses such that

(1) $\mu_1 \sim Unif([-0.5, 0.5]^2)$

(2) $\mu_2 \sim Unif([-1.5, 1.5]^2)$

$A_1, A_2 \sim Unif(\mathbb{R}^2 \times \mathbb{R}^2)$, $\Sigma_i = A_i A_i^T$ subject to the wo eigenvalues differing by a factor of $\leq 2$, and $\Sigma_1$ having area 1, $\Sigma_2$ having area 1/20.

Computed ground-truth $A_x, B_x$ from 3000 samples from both ellipses.

Bootstrapping: get 0.001, 0.999

Results:

For $A_x-B_x$: good coverage

For $E_x$: intervals are unreliable if $A_x + B_x$ is small

For $D_x$: good coverage

\section{Results (Morphology on UD Treebanks)}\label{sec:results-table}

Statistics reported:

\begin{enumerate}
%	\item CIs for $C_x := B_x - A_x$ may be a resonable stopping criterion
	\item $A_x := P_0(\{X : X_1 \leq x_1 \wedge X_2 \leq x_2\})$
	\item $B_x := P_0(\{X : X_1 \geq x_1 \wedge X_2 \geq x_2\})$
	\item  $E := \frac{\E_{x \sim P_1}[A_x]}{\E_{x \sim P_1}[A_x+B_x]}$
	\item Compute $A_{x'}$ for each $x'$ in the baseline distribution, and then calculate the percentile rank of $A_x$ among these values:
$D := P_0(A_X \leq \E_{x \sim P_1}[A_x])$
\item For each sample $x$ from real orderings, we look at the proportion $W(x)$ of samples from the baseline languages that are below $x$ throughout the entire range where both curves are defined.

\end{enumerate}

For each language, we show the number of sentences in the training partition, the surprisal-speaker memory tradeoff plane, the listener's surprisal-memory tradeoff curves (smoothed means), and the statistics defined above with bootstraped CIs.

Languages where the stopping criterion hasn't been reached are marked with asterisks.

\begin{longtable}{lllllllllllllll}
Afrikaans  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.28]  \\ 
1315  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
  &    &    &  $E_x$  &  0.42  &  [0.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.99  &  [0.96, 1.0]  \\ [10.25ex] \hline
Amharic-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.16  &  [0.06, 0.29]  \\ 
974  &    &    &  $B_x-A_x$  &  0.1  &  [0.04, 0.18]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Arabic*  &\\ [10.25ex] \hline
Armenian-Adap*  &\\ [10.25ex] \hline
Basque  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Basque-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.03, 0.15]  \\ 
5396  &    &    &  $B_x-A_x$  &  0.82  &  [0.76, 0.88]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Breton-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.03, 0.22]  \\ 
788  &    &    &  $B_x-A_x$  &  0.32  &  [0.2, 0.42]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Bulgarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.22]  \\ 
8907  &    &    &  $B_x-A_x$  &  0.56  &  [0.44, 0.7]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Buryat-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.09  &  [0.02, 0.19]  \\ 
808  &    &    &  $B_x-A_x$  &  0.28  &  [0.19, 0.42]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Cantonese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.08, 0.31]  \\ 
550  &    &    &  $B_x-A_x$  &  0.25  &  [0.17, 0.36]  \\ 
  &    &    &  $E_x$  &  0.92  &  [0.85, 0.98]  \\ 
  &    &    &  $W_x$  &  0.94  &  [0.86, 1.0]  \\ [10.25ex] \hline
Catalan  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Catalan-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.03, 0.25]  \\ 
13123  &    &    &  $B_x-A_x$  &  0.44  &  [0.33, 0.57]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Chinese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Chinese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.04, 0.23]  \\ 
3997  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.19]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.95, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Croatian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Croatian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.03, 0.23]  \\ 
7689  &    &    &  $B_x-A_x$  &  0.13  &  [0.05, 0.23]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Czech  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Czech-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.05, 0.27]  \\ 
102993  &    &    &  $B_x-A_x$  &  0.66  &  [0.56, 0.77]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.94  &  [0.86, 0.99]  \\ [10.25ex] \hline
Danish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Danish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.26]  \\ 
4383  &    &    &  $B_x-A_x$  &  0.15  &  [0.06, 0.25]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Dutch  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Dutch-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.01, 0.15]  \\ 
18310  &    &    &  $B_x-A_x$  &  0.04  &  [0.01, 0.08]  \\ 
  &    &    &  $E_x$  &  0.88  &  [0.73, 1.0]  \\ 
  &    &    &  $W_x$  &  0.98  &  [0.96, 1.0]  \\ [10.25ex] \hline
Estonian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.48]  \\ 
6959  &    &    &  $B_x-A_x$  &  0.26  &  [0.18, 0.36]  \\ 
  &    &    &  $E_x$  &  0.76  &  [0.67, 0.84]  \\ 
  &    &    &  $W_x$  &  0.72  &  [0.65, 0.79]  \\ [10.25ex] \hline
Faroese-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.13  &  [0.05, 0.26]  \\ 
1108  &    &    &  $B_x-A_x$  &  0.34  &  [0.24, 0.45]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Finnish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Finnish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.01, 0.12]  \\ 
27198  &    &    &  $B_x-A_x$  &  0.95  &  [0.91, 0.99]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
French*  &\\ [10.25ex] \hline
German  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/German-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.44  &  [0.37, 0.51]  \\ 
13814  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.29]  \\ 
  &    &    &  $E_x$  &  0.66  &  [0.6, 0.74]  \\ 
  &    &    &  $W_x$  &  0.63  &  [0.57, 0.69]  \\ [10.25ex] \hline
Greek  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Greek-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.2]  \\ 
1662  &    &    &  $B_x-A_x$  &  0.59  &  [0.48, 0.73]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hebrew  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\ 
5241  &    &    &  $B_x-A_x$  &  0.19  &  [0.08, 0.3]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hindi  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hindi-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.07  &  [0.0, 0.24]  \\ 
13304  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Hungarian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.55  &  [0.44, 0.65]  \\ 
910  &    &    &  $B_x-A_x$  &  0.07  &  [0.0, 0.15]  \\ 
  &    &    &  $E_x$  &  0.62  &  [0.51, 0.74]  \\ 
  &    &    &  $W_x$  &  0.83  &  [0.76, 0.89]  \\ [10.25ex] \hline
Indonesian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.43]  \\ 
4477  &    &    &  $B_x-A_x$  &  0.27  &  [0.22, 0.33]  \\ 
  &    &    &  $E_x$  &  0.82  &  [0.77, 0.87]  \\ 
  &    &    &  $W_x$  &  0.62  &  [0.57, 0.66]  \\ [10.25ex] \hline
Japanese  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Japanese-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.05  &  [0.0, 0.18]  \\ 
7164  &    &    &  $B_x-A_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kazakh-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.08  &  [0.0, 0.24]  \\ 
947  &    &    &  $B_x-A_x$  &  0.29  &  [0.2, 0.38]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Kurmanji-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.09, 0.36]  \\ 
634  &    &    &  $B_x-A_x$  &  0.1  &  [0.05, 0.16]  \\ 
  &    &    &  $E_x$  &  0.86  &  [0.72, 0.95]  \\ 
  &    &    &  $W_x$  &  0.96  &  [0.92, 1.0]  \\ [10.25ex] \hline
Latvian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Latvian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.51  &  [0.41, 0.62]  \\ 
4124  &    &    &  $B_x-A_x$  &  -0.05  &  [-0.11, 0.01]  \\ 
  &    &    &  $E_x$  &  0.37  &  [0.24, 0.53]  \\ 
  &    &    &  $W_x$  &  0.66  &  [0.58, 0.74]  \\ [10.25ex] \hline
Naija-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.1  &  [0.02, 0.25]  \\ 
848  &    &    &  $B_x-A_x$  &  0.33  &  [0.21, 0.47]  \\ 
  &    &    &  $E_x$  &  1.0  &  [0.99, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
North Sami  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.36  &  [0.28, 0.45]  \\ 
2257  &    &    &  $B_x-A_x$  &  0.15  &  [0.08, 0.23]  \\ 
  &    &    &  $E_x$  &  0.73  &  [0.62, 0.81]  \\ 
  &    &    &  $W_x$  &  0.36  &  [0.28, 0.43]  \\ [10.25ex] \hline
Norwegian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.14  &  [0.04, 0.3]  \\ 
29870  &    &    &  $B_x-A_x$  &  0.63  &  [0.48, 0.78]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Persian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Persian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.07, 0.32]  \\ 
4798  &    &    &  $B_x-A_x$  &  0.12  &  [0.06, 0.2]  \\ 
  &    &    &  $E_x$  &  0.93  &  [0.77, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [0.99, 1.0]  \\ [10.25ex] \hline
Polish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Polish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.39  &  [0.3, 0.5]  \\ 
6100  &    &    &  $B_x-A_x$  &  0.08  &  [-0.0, 0.15]  \\ 
  &    &    &  $E_x$  &  0.6  &  [0.5, 0.69]  \\ 
  &    &    &  $W_x$  &  0.31  &  [0.23, 0.39]  \\ [10.25ex] \hline
Romanian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Romanian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.24]  \\ 
8664  &    &    &  $B_x-A_x$  &  0.15  &  [0.07, 0.24]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Serbian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Serbian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.06  &  [0.01, 0.15]  \\ 
2935  &    &    &  $B_x-A_x$  &  0.19  &  [0.13, 0.26]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Slovak  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovak-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.21  &  [0.13, 0.3]  \\ 
8483  &    &    &  $B_x-A_x$  &  0.48  &  [0.39, 0.56]  \\ 
  &    &    &  $E_x$  &  0.94  &  [0.9, 0.97]  \\ 
  &    &    &  $W_x$  &  0.89  &  [0.83, 0.94]  \\ [10.25ex] \hline
Slovenian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.31  &  [0.22, 0.39]  \\ 
7532  &    &    &  $B_x-A_x$  &  0.44  &  [0.35, 0.56]  \\ 
  &    &    &  $E_x$  &  0.85  &  [0.79, 0.92]  \\ 
  &    &    &  $W_x$  &  0.79  &  [0.72, 0.85]  \\ [10.25ex] \hline
Spanish* &\\ [10.25ex] \hline
Swedish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.11  &  [0.02, 0.29]  \\ 
7041  &    &    &  $B_x-A_x$  &  0.06  &  [0.01, 0.12]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Tamil  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Tamil-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.04, 0.29]  \\ 
400  &    &    &  $B_x-A_x$  &  0.17  &  [0.1, 0.28]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Thai-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.15  &  [0.06, 0.28]  \\ 
900  &    &    &  $B_x-A_x$  &  0.19  &  [0.12, 0.27]  \\ 
  &    &    &  $E_x$  &  0.99  &  [0.97, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Turkish  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Turkish-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.18  &  [0.05, 0.35]  \\ 
3685  &    &    &  $B_x-A_x$  &  0.03  &  [0.01, 0.06]  \\ 
  &    &    &  $E_x$  &  1.0  &  [1.0, 1.0]  \\ 
  &    &    &  $W_x$  &  1.0  &  [1.0, 1.0]  \\ [10.25ex] \hline
Ukrainian  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.12  &  [0.03, 0.22]  \\ 
4506  &    &    &  $B_x-A_x$  &  0.18  &  [0.1, 0.24]  \\ 
  &    &    &  $E_x$  &  0.94  &  [0.87, 1.0]  \\ 
  &    &    &  $W_x$  &  0.99  &  [0.97, 1.0]  \\ [10.25ex] \hline
Uyghur-Adap  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-entropy-memory.pdf}}  &  \multirow{4}{*}{\includegraphics[width=0.25\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf}}  &  $D_x$  &  0.17  &  [0.06, 0.3]  \\ 
1656  &    &    &  $B_x-A_x$  &  0.0  &  [-0.0, 0.01]  \\ 
  &    &    &  $E_x$  &  0.9  &  [0.0, 1.0]  \\ 
  &    &    &  $W_x$  &  0.0  &  [0.0, 0.0]  \\ [10.25ex] \hline
Vietnamese* &\\ [10.25ex] \hline
\end{longtable}

\section{Thoughts}
Upper bound: Use $TH[X_t] - \sum_{t=1}^T (T-t) I_t = T (H[X_t] - \sum_{t=1}^T I_t) + \sum_{t=1}^T t I_t$ bits of memory and incur $\sum_{t>T} I_t$ in additioal surprisal.

\begin{tabular}{cccccccccccccc}
Afrikaans & Amharic & Arabic & Armenian & Basque & Breton & Bulgarian & Buryat \\
\includegraphics[width=0.14\textwidth]{figures/Afrikaans-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Amharic-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Arabic-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Armenian-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Basque-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Breton-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Bulgarian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Buryat-Adap-listener-surprisal-memory.pdf}
\\
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & Estonian \\
\includegraphics[width=0.14\textwidth]{figures/Cantonese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Catalan-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Chinese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Croatian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Czech-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Danish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Dutch-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Estonian-listener-surprisal-memory.pdf}
\\
Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian \\
\includegraphics[width=0.14\textwidth]{figures/Faroese-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Finnish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/French-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/German-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Greek-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hebrew-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hindi-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Hungarian-listener-surprisal-memory.pdf}
\\
Indonesian & Japanese & Kazakh & Kurmanji & Latvian & Naija & North Sami & Norwegian \\
\includegraphics[width=0.14\textwidth]{figures/Indonesian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Japanese-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kazakh-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Kurmanji-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Latvian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Naija-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/North_Sami-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Norwegian-listener-surprisal-memory.pdf}
\\
Persian & Polish & Romanian & Serbian & Slovak & Slovenian & Spanish & Swedish \\
\includegraphics[width=0.14\textwidth]{figures/Persian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Polish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Romanian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Serbian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovak-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Slovenian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Spanish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Swedish-listener-surprisal-memory.pdf}
\\
Tamil & Thai & Turkish & Ukrainian & Uyghur & Vietnamese \\
\includegraphics[width=0.14\textwidth]{figures/Tamil-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Thai-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Turkish-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Ukrainian-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Uyghur-Adap-listener-surprisal-memory.pdf} & \includegraphics[width=0.14\textwidth]{figures/Vietnamese-listener-surprisal-memory.pdf}
\\\end{tabular}

\end{document}

\appendix
\section{Bounding Memory in Imperfect Modeling}
We derived the nonlocality $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?

The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
(Other models would incur even higher average surprisal, without reduction in memory cost.)
The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.

This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
%Naturally, we expect that lower surprisal rates will require more memory.

%\subsection{Lower bound on approximation with n-th order Markov models}
%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
%We know $\sum_{t>T} I_{t}$ 
%If a Markov model has order $n$

There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
\begin{proposition}\label{prop:markov}
The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
\end{proposition}

In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.

\subsection{Lower bound on imperfect approximation} 
Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.

As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
How much greater will their surprisal be? In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$. We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1. In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$

The following result holds for any process:
\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%Choose any $T \in \mathbb{N}$.
Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
\end{proposition}
%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$. It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
%An equivalent way of stating this bound is:
Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}

\begin{proof}
Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
As explained above, without loss of generality, we can assume that the model is Bayes optimal. %: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
By definition, the difference between the model's surprisal and optimal surprisal is
\footnote{The first inequality seems really crude. In general, we cannot do better. If we know more about the approximating model, we might be able to improve this. Also, consider the relation to the other inequality. Can they be attained at the same time?}
\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
\end{align*}
\end{proof}
%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$. Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
%
%\begin{corollary}
%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%
%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%
%\end{corollary}
%
%

Here is way of phrasing the same statement in terms of memory:
\begin{corollary}
Let $\epsilon > 0$. Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.

\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
%What is the optimal $T$?
%It turns out that the optimal $T$ 
Searching for the optimal $T$ in the bound, we find:
\begin{corollary}
Consider any memory requirement $J \in [0,M)$.
Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
That is, it cannot do better than a $T$-th order Markov model.
\end{corollary}
Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.

% a $T$-th order Markov

Here is a way of rephrasing this in terms of memory demand:
\begin{corollary}
Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$. Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
%
%Do these results agree with the idea of information locality decreasing memory load?
%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
%

\begin{example}
%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
%
Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.

Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
and, plugging back into the bound, get that the model's surprisal is at least
$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
\end{example}

%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns

%\begin{corollary}[Approximation by N-Gram Models, lower bound]
%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
%Then
%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
%
%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
%\end{corollary}
%
%\begin{proof}
%
%\end{proof}
%
%

\subsection{Upper Bounds on imperfect approximation}
$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy. Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$. Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else. We consider the problem of coding $X$ with small memory to predict $Y$ optimally. We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$. In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large. The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}

However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.

%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).

\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
\end{proposition}
\begin{proof}
In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
\end{proof}
%Compared with the lower bound from the previous section, we see:
Computing the memory cost of a an n-th order Markov model, we can conclude:
\begin{corollary}
(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.

(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$

\end{corollary}
\begin{proof}
By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
This model can be implemented using 
$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
This bound is really crude. We can compute
\begin{align*}
H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
%&\leq I/\epsilon H[X] + I \\
%& = I(1+H[X]/\epsilon) \\
\end{align*}
but does this gives us anything?}
The second statement is a rephrasing of the first.
\end{proof}
As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
In the case of natural language, $H[X]$ is a reasonably small constant.

%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
%\begin{corollary}[Approximation by $n$-gram models]
%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
%\end{corollary}

%\subsection{Summary}

%
%\paragraph{Invariance properties under certain transformations of the sequence}
%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
%%Second, 
%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
%If we measure memory also at all new boundaries, $I$ can at most increase.
%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
%
%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
%

\section{Crypticity}

%We let $X_t$ be a process of observations (e.g. words), and $M_t$ be the internal memory states of the listener, which we can take to be a function of $X_{\leq t}$.
Under the assumption of optimality oif predictions and minimality of memory cost, we can take $M_t$ to be an equivalence class of pasts that induce the same distribution over futures (need to be careful for general processes. we here want to add the assumption that, since we are interested in sentence processing, memory is reset entirely at sentence or discourse boundaries, avoiding cases where the math wouldn't work so nicely).

Nonlocality:
\begin{equation}\label{eq:locality-cost}
L := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}

Crypticity:
\begin{equation}\label{eq:ambiguity-cost}
A := H[M_0|X_{> 0}]
\end{equation}

Note that both terms are nonnegative by definition.

\begin{proof}
By the definitions of conditional entropy and mutual information,
$H[M_0] = H[M_0 | X_{>0}] + I[M_0, X_{>0}]$
The first part is Ambiguity cost.
Now we show that the second part is equal to Locality cost.
As we assume that the listeners predictions are optimal, $I[M_0, X_{>0}] = I[X_{\leq 0}, X_{>0}]$.
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
\end{proof}

\end{document}

\section{Normalized Excess Entropy}

We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$

Why?

(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity

(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?

\section{Computing a tighter lower bound}

Here is a method that takes a quadratic amount of time to compute:

Take all pairs of positions in the corpus, and for each pair $s, t$ compute
$$P(X_{t, ..., t+T}|T_{s-T, ..., s-1})$$
After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t, ..., t+T}|T_{s-T, ..., s-1}) + \log P(X_{t, ..., t+T}|T_{t-T, ..., t-1})\right) < \epsilon$$
Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) lower bound of memory for modeling with KL loss $\epsilon$.
However, this bound might still not be tight, unless we require that $\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$.

Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).

It should be feasible to apply this to POS-level language modeling.

ACTUALLY

Take all pairs of positions in the corpus, and for each pair $s, t$ compute
$$P(X_{t}|T_{s-T, ..., s-1})$$
After fixing $\epsilon > 0$, find a partitioning $\Pi$ of the positions $T, ..., N-T$ ($N$ corpus length) such that
$$\frac{1}{N-2T} \sum_{t=T}{N-T} \left( - \log \frac{1}{|\Pi(t)|} \sum_{s \in \Pi(t)} P(X_{t}|T_{s-T, ..., s-1}) + \log P(X_{t}|T_{t-T, ..., t-1})\right) < \epsilon$$
and
$\Pi(s+1)$ is determined by $\Pi(s)$ and $X_{s+1}$
Taking the minimum entropy of any such $P$ should be a consistent (as $T$ and the data go to infinity, assuming that the model is already correct, or converges sufficiently strongly) estimator of a tight lower bound of memory for modeling with KL loss $\epsilon$.

Unlike excess entropy, it is not left-right invariant (which is good, because memory is not left-right invariant).

It should be feasible to apply this to POS-level language modeling.

Is this a consistent estimator?

\section{Pointwise Bounds}

\begin{proposition}
Let $F := X_{>0}$, $P := X_{\leq X}$. Then pointwise cost for an optimal representation (in terms of average memory cost) is
%$$C(p) := \log P\left{p' : P(F|p') = P(F|p)$$
\end{proposition}

There are several pointwise versions of excess entropy.
Here is one:
$$D(p) := \E_{f|p} pmi(f,p) = D_{KL}(P(f|p) || P(f))$$

\begin{proposition}
$D(p) \leq C(p)$
\end{proposition}

\begin{proof}
\begin{align*}
D(p) &= \sum_f p(f|p) \log \frac{p(f|p)}{p(f)} \\
&= \sum_f p(f|p) (\log p(p|f) - \log p(p)) \\
&= \left( \sum_{f} p(f|p) \log p(p|f)\right) - \log p(p)
\end{align*}
and the left quantity is $\leq 0$.
\end{proof}

$D$ can be estimated at each specific point by sampling continuations.

TODO $D(p)$ can be seen as a formalization of DLT storage cost

\begin{corollary}
If $\pi(p)$ is the equivalence class of the past, then
pointwise memory cost is
$$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
\end{corollary}
Taking the expectation over $p$, the left component is the excess entropy, while the right component is the gap between excess entropy and state entropy:
$$\E{f,p} \log(\pi(p)|f) = H[\pi(P) | F]$$

$H[\pi(P)] = I[P,F] + H[\pi(P) | F] = I[\pi(P), F] + H[\pi(p) | F]$

Memory = Non-Locality + Difficulty of Guessing Past from Future

I will call excess entropy `locality cost', and the second term `ambiguity cost'.

This decomposition also holds if $\pi$ is lossy. If $\pi$ is lossy, the non-locality term has to decrease.

Informally $$\E_{f|p} \left[ pmi(f,p) -  \log(\pi(p)|f)\right]$$
decomposes into a locality/storage cost term $\E_{f|p} pmi(f,p)$ (can be seen as a formalization of DLT storage cost), and $- \E_{f|p} \log(\pi(p)|f)$.
The latter component quantifies how hard it is to guess the storage content from the future (difficulty of guessing past from future, given the past).
One possible interpretation is in terms of noisy-channel memory:
Under memory constraints, more frequent/typical pasts will be more likely to be imputed based on the more recent observations. `Preventing a wrong noisy-channel inference takes extra memory effort.'
Can also relate this to prefix ambiguity, in the case where the ambiguity doesn't always turn up: E.g., if it is hard to predict from a continuation whether the prefix had been temporarily ambiguous.

Interpreting Ambiguity Cost:

- not knowing which things will be relevant later is costly

- one instantiation: not knowing which referents will be referred back to later

- one instantiation: ambiguous prefixes are costly

can you show that languages also optimize ambiguity cost?

\section{Masha's experiments}

- DepL experiment: excess entropy, we can even trace the pointwise profile

- MaxOP: initial NP can give rise to three memory states: subj, obj, indeterminate. The remainder of the sentence does not reveal whether the first NP had a case marker (if it did reveal that, this would increase locality cost). Ambiguity cost can be decreased by avoiding the indeterminate state (but there are other, even better minima! -- so this doesn't predict as much as MaxOP).

\section{Estimating Memory Cost and Ambiguity Cost}

by actually estimating a memory representation: estimate a variable that is independent of the future when conditioning on the past, and also: independent of past when conditioning on last word and last state of memory

if memory is deterministic, just say that it is a function of $X_t, M_{t-1}$.

if memory is randomized:

formally: $M_t$ must have the properties (1) $M_t \bot X_{<t} | (M_{t-1}, X_t)$, (2) $M_t \bot X_{>t} | X_{\leq t}$.

In this case, cost measure should be MI with past, not entropy

randomized memory is useful because we can apply gradient descent, solving an optimization problem to provide an upper bound on memory cost

- solving a discrete optimization problem (probably not super scalable)

- variational: stochastic continuous memory representations and take memory cost to be $I[Memory, Past]$ (also done in IB)

(noise the LSTM hidden state in each time step)

calculate upper bound $D_{KL}(P(Memory|Past) || P_{prior}(Memory)) \geq I[Memory, Past]$, estimate the tradeoff curve between this and cross-entropy (the resulting curve is a variational upper bound on the real curve).

- other ways of minimizing MI with past (don't require noising of the hidden state):

-- adversarial: make the past hard to predict

-- dual lower bound on MI

both of these methods are ways of doing IB

\section{Incremental Prediction under Memory Limitations}

%\subsection{Noisy-Context Surprisal}
%
%\citet{futrell-noisy-context-2017} describe a psycholinguistic noisy-channel model of human language processing under memory limitations where words are progressively deleted from memory. %with progressive probability.
%They show that processing as described by this model favors word order where linguistic material that has high mutual information will appear closer together, in the sense that the processing cost, measured by the surprisal $-\log P(w_t | w_1 ... w_{t-1})$ will be
%$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
%where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
%Given that $f$ is decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
%
%%In the most extreme case of such a model, a bigram model, we have:
%%$$-\log P_{Model}(w_t | w_1 ... w_{t-1}) = -\log P(w_t) - pmi(w_t, w_{t-1})$$
%%where $PMI(X,Y) = \log P(X,Y) - \log P(X) - \log P(Y)$ is the pointwise mutual information.
%%
%%This directly applies to n-gram models: Here, words are indeed `forgotten' deterministically when they are more than $n$ words in the past.
%
%%\citet{futrell-noisy-context-2017} show that this model derives a range of psycholinguistic locality effects.
%%It also applies to models with strictly bounded context, such as n-gram models -- in this case, past words beyond a fixed distance threshold are entirely erased from memory.
%%However, it is less clear that human language processing (or state-of-the-art NLP techniques such as LSTMs) are affected by such loss -- it is not clear that humans entirely forget previous words, and perfectly remember the other words.
%
%\subsection{Memory in General Sequence Models}

We consider a general incremental model that maintains a memory state and, in each time step, computes a distribution over the vocabulary based on its memory state, stochastically emits a word from this distribution, and updates its memory state depending on the chosen word.
We are making the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
This model describes aspects of human language processing, if we accept that humans perform incremental prediction.
(Typical NLP models such as n-gram models and LSTM language models also fit into this framework.)
%N-gram models and LSTM language models both fit into this framework.
%On the other hand, LSTM models with random access to prior words do not fit into this framework, as they rely on past states in addition to their internal memory state at time $t$.

In this general framework, the minimal amount of bits needed to encode the model's internal memory state is directly determined by the decay of mutual information over distances:
The faster mutual information between words decays as their distance increases, the less memory is required.
More precisely:

\begin{proposition}
Modeling a stationary process  $(X_t)_{t \in \mathbb{Z}}$ in this framework %of incremental modeling described above,
 requires maintaining at least
\begin{equation}\label{eq:memory-bound}
M := \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{equation}
bits of memory,
where
$\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ is the mutual information of the observations $X_t$ and $X_0$, conditioned on the intervening observations.
\end{proposition}
%that the model does not have an `absolute clock' built into it.
This quantity is also known as \emph{excess entropy} (CITE).

\begin{proof}
%We assume that the process is stationary, that is, that there is no `absolute clock' built into the model. Formally, this means that, for any set of indices $t_1 < t_2 < \dots < t_n$, the joint distribution $(X_{t_1}, X_{t_2}, ..., X_{t_n})$ only depends on the differences $(t_{i+1}-t_i)_{i=1,...,n-1}$, not on the absolute values $t_i$.
The number of bits of memory is lower bounded by the mutual information between past and future observations, $I[(X_i)_{i>0}, (X_i)_{i \leq 0}]$. 
Then the formula is a consequence of the chain rule for mutual information:
\begin{align*}
I[(X_i)_{i > 0}, (X_i)_{i \leq 0}] &= \sum_{i=1}^\infty I(X_i, (X_i)_{i \leq 0}|X_1, ..., X_{i-1}) \\
&= \sum_{i=1}^\infty \sum_{j=0}^{-\infty} I(X_i, X_j|X_{j+1}...X_{i-1}) \\
& = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]
\end{align*}
%In the last step, we have made the technical assumption that the process is stationary, that is, that the rules by which the model manipulates its memory and outputs words do not change over time.
%that the model does not have an `absolute clock' built into it.
\end{proof}
That is, we measure the amount of statistical dependency of observations that are $t$ steps apart, controlling for any information that is redundant with intervening observations.
This quantifies how much information needs to be carried across $t$ timesteps without any possibility for `guessing' it from intervening observations.

Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.

The result in (\ref{eq:memory-bound}) is entirely information-theoretic and applies to \emph{any} physical encoding of the past, entirely independent of the implementation of the model and the mechanisms by which it computes predictions.
In particular, while the relation to psycholinguistic and psychological models of how memory works would be interesting to explore, our result applies to any such model.
Memory representations do not have to be rational or optimal for this bound to hold:
It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.

%as long the model describes the process $(X_t)_{t \in \mathbb{Z}}$

%\section{Estimating Memory Load in Prediction}

%We want to apply (\ref{eq:memory-bound}) on corpus data to compute the memory load of language for incremental prediction.

For our purposes, it is important to note that we can estimate (\ref{eq:memory-bound}) from a statistical model assigning probabilities to sentences.
(TODO summarise prior work attemtpting to estimate the same quantity, if there is any such work).

\section{Bounding Memory in Imperfect Modeling}
We derived $M$ as a lower bound on the memory of a model that does optimally well in predicting the input process.
What happens when we consider imperfect (or near-optimal) prediction -- which may be a more realistic model for human processing?

The goal of this section is to show that $M$ is not just a lower bound on the memory required for perfect modeling, but also can be used to derive lower and upper bounds on imperfect modeling by suboptimal models.
To study imperfect modeling, we think of the process $(X_t)$ as fixed, and of a model as a function $f$ mapping histories $X_{<t}$ to memory representations, and computing a probability distribution over the vocabulary for $X_t$ based on the memory representation $f(X_{<t})$.
The memory representation $f(X_{<t})$ is a random variable whose entropy is the memory cost of the model.
In general, a model could use any mapping from memory states to distributions over the vocabulary to make its predictions.
However, we will focus on Bayes-optimal models, i.e., models that correctly compute the distribution of $X_t$ given the memory representation $f(X_{<t})$.
(Other models would incur even higher average surprisal, without reduction in memory cost.)
The average surprisal experienced by such a model is $H[X_t | f(X_{<t})]$, and is lower bounded by the true surprisal, $H[X_t|X_{<t}]$.

This section provides some results relating the model's memory cost $H[f(X_{<t})]$ to the attained surprisal $H[X_t | f(X_{<t})]$.
%Naturally, we expect that lower surprisal rates will require more memory.

%\subsection{Lower bound on approximation with n-th order Markov models}
%Let us assume that there is some $T$ such that $I_t = 0$ for $t > 0$ (e.g., $T$ is the length of the longest sentence in the corpus).
%We know $\sum_{t>T} I_{t}$ 
%If a Markov model has order $n$

There is a direct and obvious connection between the decay of mutual information and approximation by $n$-th order Markov models:
\begin{proposition}\label{prop:markov}
The Bayes-optimal $n$-th order Markov model will experience average surprisal (that is, cross-entropy) exactly $$H[X_t|X_{<t}] + \sum_{t > n} I[X_t, X_0 | X_{1\dots t-1}]$$
\end{proposition}

In the rest of this section, we think about further bounds for the case when we have information about memory, but not necessarily about the decay of mutual informations or the nature of the approximating model.

\subsection{Lower bound on imperfect approximation} 
Here we lower-bound the surprisal experienced by a model about which we only know that it has memory $J$ less than $M$, and where have no further knowledge about its structure.

As one might hope, models with memory $J$ less than $M$ will necessarily show greater average surprisal than a perfect model (simply because they cannot be perfect models by the previous proposition).
How much greater will their surprisal be? In general, if a model has memory much smaller than $M$, we cannot conclude that it must have very high average surprisal.\footnote{Example: consider, for each $T \in \mathbb{N}$, a process where $I_{T} = 1$, and $I_{t} = 0$ for $t \neq T$. We have $M = T$, but, even as $T \rightarrow \infty$, a model with zero memory achieves a KL loss of only 1. In this case, Proposition \ref{prop:lower-bound} yields the correct result.}
However, we can get such a bound in relation to the contribution of long-term dependencies to $M$.
%The state I've arrived at so far is still a bit complicated, but there is a reasonable final result for a wide class of processes.
%For smaller contribution of long-term dependencies is to $M$, we can lower bound the model's surprisal more strongly.
We write $$I_t := I[X_t, X_0 | X_{1\dots t-1}]$$

The following result holds for any process:
\begin{proposition}[Approximation by Models with Small Memory, lower bound]~\label{prop:lower-bound}
%For any $T \in \mathbb{N}$, a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T$$
%Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%$$\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$$
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%Choose any $T \in \mathbb{N}$.
Any model with memory $J$ will experience average surprisal (that is, cross-entropy) at least
$$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%$$H[X_t|X_{<t}] + \sup_{T\in \mathbb{N}}\ \left[\frac{M-J}{T} - \sum_{t > T} (t/T - 1) I_t\right]$$
\end{proposition}
%The second term confirms the intuition that the model's surprisal should grow with the difference $M-J$, but the strength of the bound depends on the choice of $T$.\footnote{The bound can be vacuous for small choices of $T$: If we choose $T = 1$, the bound simply evaluates to the trivial lower bound $H[X_t] - J$, which may be even weaker than the likewise trivial lower bound $H[X_t|X_{<t}]$. It will become clear below that the bound is always nonvacuous when $T$ is large enough.}
%An equivalent way of stating this bound is:
Finding the optimal $T$ will require knowledge of the decay profile $I_t.$\footnote{For any fixed $T$, one can find a decay profile for which the bound would end up vacuous at that $T$, but meaningful for a larger $T$.}
Note that this bound only requires knowledge of $J$ and $I_t$, but no further knowledge of the process and the model.
I think that the bound is also optimal among all bounds that only require knowledge of $J$ and $I_t$, in the sense that for every choice of $I_t$ and $J$, there is a process and a model for which the bound is attained.\footnote{There can be no general nonvacuous bound relying only on $M$ and $J$, as shown in a previous footnote.}

\begin{proof}
Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
As explained above, without loss of generality, we can assume that the model is Bayes optimal. %: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
By definition, the difference between the model's surprisal and optimal surprisal is
\footnote{The first inequality seems really crude. In general, we cannot do better. If we know more about the approximating model, we might be able to improve this. Also, consider the relation to the other inequality. Can they be attained at the same time?}
\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& H[X_t | f(X_{<t})] - H[X_t | X_{<t}] =  \frac{1}{T} \sum_{t=1}^{T} \left(H[X_t | f(X_{<t})] - H[X_t | X_{<t}]\right) \\
& \geq   \frac{1}{T} \left(H[X_{1\dots T} | f(X_{\leq 0})] - H[X_{1\dots T} | X_{\leq 0}]\right)  = \frac{1}{T} \left(I[X_{1\dots T}|X_{\leq 0}] - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& = \frac{1}{T} \left(\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) = \frac{1}{T} \left(M - \sum_{t > T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{\leq 0})]\right) \\
& \geq \frac{1}{T} \left(M + \sum_{t > T} (T-t) I_t - J\right) 
\end{align*}
\end{proof}
%We can restate this bound in the following two, maybe more insightful, ways:\footnote{Note that (1) and (2) make clear that the bound is guaranteed to be larger than the basic bound $H[X_t|X_{<t}]$ when $T$ is chosen sufficiently large, but the gain over that trivial bound diminishes as $T \rightarrow \infty$. Finding an optimal $T$ that makes the bound as large as possible will depend on the process in question.}
%
%\begin{corollary}
%(1) A model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] +\sup_{T\in \mathbb{N}}\ \left[ \frac{\left(\sum_{t=1}^T t I_t\right)-J}{T} + \sum_{t > T} I_t\right]$$
%
%(2) Let $\epsilon > 0$, $T \in \mathbb{N}$ such that $\sum_{t > T} t I_t < \epsilon$.
%Then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $$H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T} + \sum_{t > T} I_t$$
%
%\end{corollary}
%
%

Here is way of phrasing the same statement in terms of memory:
\begin{corollary}
Let $\epsilon > 0$. Then it takes at least (and possibly more than) $$\sup_{T \in \mathbb{N}} \left(\sum_{t=1}^T t I_t\right) + T\left(\sum_{t > T} I_t - \epsilon\right)$$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

\paragraph{Implication 1: Less information locality requires more memory at the same level of prediction error}
Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
Then, when fixing $J$, the lower bound on surprisal provided by the proposition is higher in the case of $Y$ than in the case of $X$.
While this does not mean that every such $X$ will require less memory than every such $Y$ (since this is only a lower bound), it does entail an asymptotic version of this:
If we fix $X$ and decrease the information locality of $Y$ indefinitely strongly, $Y$ will ultimately require more memory than $X$, at the same level of experienced surprisal.

\paragraph{Implication 2: With suboptimal memory, no model can do better than a Markov model}
%What is the optimal $T$?
%It turns out that the optimal $T$ 
Searching for the optimal $T$ in the bound, we find:
\begin{corollary}
Consider any memory requirement $J \in [0,M)$.
Let $T$ be the smallest $T \in \{1,2,\dots\}$ such that $$\left(\sum_{t=1}^T t I_t\right) \geq J$$
Informally, we choose $T$ such that a memory capacity of $\approx J$ is sufficient for perfectly modeling all statistical dependencies up to length $T$ when ignoring any longer dependencies.
Then any model with memory $J$ will experience average surprisal at least $$H[X_t|X_{<t}] + \sum_{t > T} I_t$$
That is, it cannot do better than a $T$-th order Markov model.
\end{corollary}
Informally, this means that, irrespective of the architecture of the approximating model, it cannot do better than a model that spends all its capacity on modeling short-range dependencies, and fully ignores long-range ones.

% a $T$-th order Markov

Here is a way of rephrasing this in terms of memory demand:
\begin{corollary}
Let $\epsilon > 0$, then let $T$ be the largest $T$ such that $\sum_{t > T} I_t > \epsilon$. Then it takes at least (and possibly more than) $\left(\sum_{t=1}^T t I_t\right)$ bits of memory to achieve surprisal $\leq H[X_t|X_{<t}] + \epsilon$.
\end{corollary}

%These statements show that the bound on the model's surprisal becomes stronger if $I_t$ decays faster and the convergence of the partial sums $\sum_{t=1}^T t I_t$ to $M$ (as $T \rightarrow \infty$) becomes faster.
%For natural language, where $I_t$ seems to really decay quite rapidly and where (in this work) we're focusing on information within sentences only, we should expect that these bounds would end up reasonably strong.
%
%Do these results agree with the idea of information locality decreasing memory load?
%%If we consider two processes with the same entropy rate and unigram entropy, but where $I_t$ decays strictly faster for one than the other, then for any $T$, the bound will be greater for the process that has less information locality -- confirming that processes with less information locality require more memory in approximation (TODO work out):
%Assume $X$ and $Y$ are processes with mutual informations $I_t^X, I_t^Y$, such that they agree on the entropy rate and the unigram entropy.
%Assume that $X$ has greater information locality than $Y$ in the (quite strong) sense that, for any $T$, $$\sum_{t>T} I_t^X \leq \sum_{t>T} I_t^Y$$
%Then the lower bound that we derived for surprisal of a model with memory $J$ is stronger (higher) for the process $Y$ that has less information locality.\footnote{Because this is true for any $T$, and thus also for the supremum over all $T$'s.}
%Thus, at least broadly speaking, processes with more information locality can be modeled with lower loss at the same memory cost.
%

\begin{example}
%It follows instantly from the bounds that, if $I_t$ is zero for $t > T$ (i.e., the process is $T$-order Markov), average surprisal of the model is lower-bounded by $H[X_t|X_{<t}] + \frac{M-J}{T}$.\footnote{I think that this bound is optimal: If all we require about $I_t$ is that it is zero for $t > T$, and have no other knowledge about $I_t$, then this bound should be attained by some processes where $I_T = M/T$, and all other $I_t$'s are zero.}
%
Let us consider a class of processes that might be a reasonable model of language: Assume $I_t = \alpha t^{-\beta}$ for some $\alpha > 0$, $\beta >2$ (for smaller $\beta$, we would have $M = +\infty$).
We can bound $M = \sum_{t=1}^\infty \alpha t^{-\beta+1} \geq \int_1^\infty \alpha t^{-\beta+1} dt = \frac{\alpha}{\beta-2}$, and similarly $M \leq \alpha + \frac{\alpha}{\beta-2} = \alpha \frac{\beta-1}{\beta-2}$.

Then we can lower-bound the surprisal of a model with memory $J$ by $H[X_t|X_{<t}]$ plus (TODO algebra errors may lie ahead)
$$\frac{1}{T} \int_1^T \alpha t^{-\beta+1} dt - j/T + \int_T^\infty \alpha t^{-\beta} = T^{-1}[\frac{\alpha}{\beta-2} - J] + T^{-\beta+1} [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]$$
To get an optimal $T$, we can take derivatives and maximize this expression, which gives us (assuming that $[\frac{\alpha}{\beta-2} - J] > 0$, otherwise the maximum will be assumed at $T=1$?)
$$T := \left(\frac{(1-\beta)[ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]}{[\frac{\alpha}{\beta-2} - J]}\right)^{1/(\beta-2)}$$
and, plugging back into the bound, get that the model's surprisal is at least
$$ H[X_t|X_{<t}] + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta-1}{\beta-2}}}{\left((1-\beta) [ \frac{\alpha}{\beta-1} - \frac{\alpha}{\beta-2}]\right)^{1/(\beta-2)}} + \frac{[\frac{\alpha}{\beta-2} - J]^{\frac{\beta+1}{\beta-2}}}{(\beta-1)^{\frac{\beta+1}{\beta-2}} [ \frac{\alpha}{\beta-2} - \frac{\alpha}{\beta-1}]^{\frac{\beta+1}{\beta-2}-1}}$$
which grows monotonically, with slightly superlinear speed, as the approximating model's memory $J$ decreases.
Furthermore, if information locality in the process decreases -- that is, $\beta$ decreases but we adjust $\alpha$ to keep $I[X_t|X_{<t}]$ constant -- it seems that the factor $(\beta-1)$ in both denominators will increase the bound, confirming the intuition that approximating a process with less information locality will require more memory.
\end{example}

%Informally, the reason for this phenomenon is that large values of $M$ could reflect large amounts of information at a small distance, but also small amoutns

%\begin{corollary}[Approximation by N-Gram Models, lower bound]
%Let $\epsilon > 0$, $ \in \mathbb{N}$ such that $\sum_{t > T} t I_t/P < \epsilon$.
%Then
%$$\frac{\sum_{M/P < t < T} I_t}{P} \geq $$
%
%Any $n$-gram model will experience average surprisal (that is, cross-entropy) $$H[X_t|X_{<t}] + \sum_{t > n} I_t$$
%\end{corollary}
%
%\begin{proof}
%
%\end{proof}
%
%

\subsection{Upper Bounds on imperfect approximation}
$M$ is only a lower bound on memory need, not necessarily identical to it: When we require that the model can optimally model the data, then there are examples where the memory need is much larger than $M$, as proved by~\cite[Section 6]{harsha-communication-nodate} (though it's not clear to me whether they are relevant to language).\footnote{Here is a much simpler example sufficient for our need: Let $Z$ be some random variable with nonzero, finite entropy. Let $Z'$ be an independent copy of $Z$, and let $B \sim Bernoulli(p)$. Then define $(X,Y) := (Z,Z)$ if $B=1$ and $(Z, Z')$ else. We consider the problem of coding $X$ with small memory to predict $Y$ optimally. We have $I[X,Y] = pH[Z]$, but any method of predicting $Y$ from a (possibly randomized) function $f(X)$ of $X$ will be suboptimal unless $H[f(X)] \geq H[X]$. In particular, as $p$ can be made small and $H[Z]$ can be made large, the gap between $I[X,Y]$ and the entropy of any mediating storage representation $f(X)$ can be arbitrarily large. The situation changes when encoding and decoding can share a random string (this randomness is then also averaged over when computing the KL-loss), as shown by \cite{harsha-communication-nodate}, but this doesn't seem to model the setting we're interested in.}

However, this needn't be a problem: As long as we only require \emph{near}-optimal prediction and work in the setting of \emph{sequence prediction}, memory need can be bounded in a way that grows linearly in $M$.
We do this using a result of~\cite{sharan-prediction-2016} showing that the loss of $n$-th order Markov models can be bounded in terms of $M$, and noting that we can compute the storage cost of $n$-th order Markov models.

%By a theorem of (CITE), lower excess entropy translates into better approximation by modeling with fixed context window (n-gram models).

\begin{proposition}[Approximation by Markov Models, upper bound. \cite{sharan-prediction-2016}]
The Bayes-optimal $n$-th order Markov model achieves average surprisal $\leq H[X_t|X_{<t}] + M/n$.
%For any $\epsilon > 0$, an $M/\epsilon$-order Markov model can achieve KL loss $\leq \epsilon$.
\end{proposition}
\begin{proof}
In their Proposition 1, they showed that, for any $\epsilon > 0$, a $M/\epsilon$-order Markov model achieves KL loss $\leq \epsilon$, which gives our statement by rearranging.
(Their result follows directly from Proposition~\ref{prop:markov}, which is also how they prove it.)
\end{proof}
%Compared with the lower bound from the previous section, we see:
Computing the memory cost of a an n-th order Markov model, we can conclude:
\begin{corollary}
(1) For any $\epsilon > 0$, there is a model achieving surprisal $H[X_t|X_{<t}] + \epsilon$ using $\leq M \frac{H[X]}{\epsilon}$ bits of memory.

(2) For any $J < M$, there is a model with memory $J$ that achieves average surprisal at most $$H[X_t|X_{<t}] + \frac{M}{J} H[X_t]$$

\end{corollary}
\begin{proof}
By the proposition, we can achieve a surprisal of $H[X_t|X_{<t}] + \epsilon$ with an $M/\epsilon$-gram model.
This model can be implemented using 
$H[X_1 ... X_{M/\epsilon}] \leq M \frac{H[X]}{\epsilon}$ bits of memory.\footnote{
This bound is really crude. We can compute
\begin{align*}
H[X_1 ... X_{M/\epsilon}] &= \frac{M}{\epsilon} H[X] - \sum_{t=1}^{M/\epsilon} (M/\epsilon-t) I_t
%&= I/\epsilon H[X] - I/\epsilon \left(\sum_{t=1}^{I/\epsilon} I_t\right) + \sum_{t=1}^{I/\epsilon} t I_t \\
%&\leq I/\epsilon H[X] + I \\
%& = I(1+H[X]/\epsilon) \\
\end{align*}
but does this gives us anything?}
The second statement is a rephrasing of the first.
\end{proof}
As expected, this bound on surprisal increases as the available memory $J$ decreases in relation to $M$.
In the case of natural language, $H[X]$ is a reasonably small constant.

%Combining with (REF), we can characterize the surprisal of an $n$-gram model in terms of $M$ as
%\begin{corollary}[Approximation by $n$-gram models]
%The average surprisal $S_n$ of the Bayes-optimal $n$-gram model is bounded as
%$$H[X_t|X_{<t}] + \frac{M}{n} - \left(\sum_{t > n} (t/n - 1) I_t\right)\ \leq\ S_n\ \leq\ H[X_t|X_{<t}] + \frac{M}{n}$$
%\end{corollary}

%\subsection{Summary}

%
%\paragraph{Invariance properties under certain transformations of the sequence}
%%Because $I$ is the mutual information between past and future observations, it is invariant under various transformations of the sequence:
%%First, if, say a filler word is randomly added into places in the sequence, the entropy rate will increase but $I$ will stay the same.
%%Second, 
%If we split tokens (deterministically and invertibly) into subtokens, but only measure the memory that is transmitted across the boundaries that were token boundaries before this transformation, $I$ will stay the same.
%If we measure memory also at all new boundaries, $I$ can at most increase.
%On the other hand, if we merge tokens (deterministically and invertibly), $I$ can at most decrease.
%
%%These invariance properties are remarkable since $I$, at the same time, is a measure of the \emph{locality} of information and the \emph{decay} of mutual information over distances.
%

\section{Normalized Excess Entropy}

We quantify information (non-)locality by $\frac{I[X_{\geq t}, X_{<t}]}{I[X_t, X_{<t}]}$

Why?

(1) The memory bound is then exactly the amount of information that needs to be transmitted from the past for one word times this quantity

(2) In the plot of decay of MIs, this is the center of mass: At what distance is the predictive information about the current word stored on average?

\section{Excess Entropy and Hierarchical Structure}

Finiteness of excess entropy is not in conflict with the presence of hierarchical structure and long-range dependencies in language:
Average memory demand can be small if the probability mass assigned to long dependencies is small.

Indeed, perhaps surprisingly, languages defined by PCFGs always have finite average memory (and thus in particular finite excess entropy):

\begin{proposition}
Any language generated by a PCFG\footnote{With rational weights, and finitely many terminals and nonterminals} has finite average memory demand.

\end{proposition}

\begin{proof}
We can consider incremental Earley parsing. Disregarding probabilities, parsing a word of length $n$ takes $O(n^2)$ bits of memory on average. For storing probabilities, the storage will be similarly bounded.
The result follows from (CITE).
\end{proof}

\begin{example}
Consider $a^nb^n$ with the probability assignment given by the PCFG $X \rightarrow \epsilon$ ($1-p$), $X \rightarrow aXb$ ($p$), for some $0<p<1$.

To compute  excess entropy, we construct a stationary process by considering concatenations of i.i.d. samples from the language.

%Also, the associated stationary process has finite excess entropy.

\end{example}

\section{Related Work}

derivation of information locality from lossy-context surprisal

Crutchfield, L{\"o}hr: construct HMMs that provide upper bounds on memory cost.
Should we be studying such upper bounds? I think that the lower bound that we study here is more useful due to the bounds it provides on imperfect modeling. 
The example from \cite{harsha-communication-nodate} shows that the true minimum memory cost may be a lot larger than the lower bound $M$ (it is exponential in $M$ for the family of processes that they considered), while -- for fixed KL loss and unigram entropy -- the memory required for imperfect modeling is $O(M)$, as we saw in the previous section.

Debowski: compute excess entropy in text

\cite{sharan-prediction-2016}: link between excess entropy and approximability by $n$-th order Markov models %study approximability by n-th order Markov models

\bibliographystyle{apalike}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

\section{Memory and Cross-Linguistic Word-Orders}

Do word orders in language minimize memory need as quantified by (\ref{eq:memory-bound}), when compared to other possible orderings?

\subsection{Prediction on the level of POS}
As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we first compute (\ref{eq:memory-bound}) on the level of POS tags -- that is, we compute memory need for incremental prediction of POS sequences. %, following~\cite{futrell-noisy-context-2017}.
As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly for the empirical distribution over sentences in a given corpus.

%We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
%As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
%This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.
%
%%\mhahn{TODO results across languages}
%
%%We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.

Results for four languages are shown in Figure~\ref{fig:memory-results}.
Averaged across languages, natural language requires less memory than 87 \% of random baseline ordering models.
%Languages which require less memory than all baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).

%%For half of the languages
%For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
%For 37 \% of the languages, it also requires less memory than the original language.

%We found that, as predicted, memory usage in real languages was correlated with word order freedom:

\begin{figure*}
\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%
\caption{Memory demands of real, optimized (for ease of LSTM language modeling), and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
\end{figure*}

\subsection{Incorporating Morphology}
In many languages, morphology may have a big impact on prediction, but it is entirely ignored if we compute memory need for predicting POS tags.
Accounting for morphology requires a more sophisticated statistical model of language that is able to encode morphosyntatic generalizations.

The following models might be options, considering the NLP literature:

\begin{enumerate}
\item an n-gram model: Here, the problem is that n-grams do not express any morphosyntactic generalizations. Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent. Both the small scale of many UD corpora and free word order in many languages with rich morphology thus seem to make such models unattractive.

\item a statistical grammar: It may sound natural to model morphosyntax with a hierarchical grammar. The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model. One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations. The second option will make this a machine learning model that, for our purposes, does not seem to be superior to (3). % (it kept many NLP people busy for the 2000s...)
\item a recurrent neural network (in particular, LSTM): NLP research in the past years shows that such models consistently fit language data better than (1) and (2) (and any other model that does not have a neural component). The model family seems rich enough to learn morphosyntactic generalizations and deal with free word order.
\end{enumerate}

(3) clearly seems to be the best choice.

As LSTMs are sequential models, the question is how to encode the input, including morphological information, into a sequence of tokens.
%On option would be to encode words as sequences of characters, or morphemes.
The UD treebanks encode morphology by providing, for each word in a sentence, a sequence of feature-value pairs (e.g., NUMBER=Singular).
We encode each word into (1) its POS tag, (2) the sequence of these morphological feature-value pairs, (3) a special end-of-word token.
Each POS tag and morphological feature-value pair is encoded as an atomic token.

During incremental prediction, the LSTM sequentially reads these tokens and predicts the next one.
The probability that the LSTM assigns to a word -- encoded as a bundle of POS and morphological features -- in a context is the product of the contextual probabilities of the POS tag, the feature-value pairs, and the end-of-word token (the purpose of the end-of-word token is to make this possible -- we couldn't compute the probability of a word directly without this).

Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.

Note that the quantities $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ in~(\ref{eq:memory-bound}) are equal to difference between conditional entropies $H[X_t|X_1, ..., X_{t-1}] - H[X_t|X_0, X_1, ..., X_{t-1}]$.
Therefore, to compute this quantity, we compute the differences between the surprisals
$-\log P[X_t | X_0, X_1, ..., X_{t-1}]$ and $-\log P[X_t | X_1, ..., X_{t-1}]$ for each word in the held-out partition, and take the average over these.

\section{Measuring Information Locality via Normalized Excess Entropy}

%We consider the following question: For a fixed budget of predictability, $I[X_t|X_{<t}]$, we want to minimize excess entropy -- that is, we want to

How do we compare excess entropies for models with different predictabilities?
We can take the quotient of excess entropy and predictability:
$$\frac{\sum_{t=1}^\infty t \cdot I[X_t,X_0 | X_{1...t-1}]}{\sum_{t=1}^\infty I[X_t,X_0|X_{1...t-1}]}$$
Why is this a good measure of locality?

\begin{enumerate}
\item Minimizing this quantity (subject to constraints) is equivalent to minizing excess entropy subject to fixed predictability (and these constraints).
\item This quantity stays invariant when we multiply each $I[X_t,X_0 | X_{1...t-1}]$ with a fixed factor. This makes it a natural measure of the decay of $I[X_t,X_0 | X_{1...t-1}]$ as $t \rightarrow \infty$.
\item This quantity is $\geq 1$, and it takes on the value $1$ if and only if the process is modeled perfectly by a bigram model (is a first-order Markov process).

For fixed entropy rate, the excess entropy is minimized by exactly the Markov processes of this entropy rate.
\end{enumerate}

\bibliographystyle{acl}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

Research has shown that dependency lengths are minimized across a wide range of languages, and it has been suggested that this may be a universal.

This raises the question of what causes this phenomenon.

What explains this?

Hawkins: TODO

Gildea and Jaeger: Gildea & Jaeger (2015) also perform optimization experiments, searching over possible grammars
optimized for dependency length, average log probability, or some weighted combination of the two.
The authors find that optimizing one criterion does not generally lead to good performance on the
other,  but that the observed languages are generally closer to some optimal frontier that trades off
dependency length against average log probability.  (from Temperley 2018)

Futrell and Levy: noisy context surprisal explains

Here, we use state-of-the art language modeling techniques to address this question:
On 50 diverse languages, we ... 
Resulting languages show shorter dependency length.

\subsection{Ordering Model}

As in work studying dependency length minimization in language (e.g. Futrell et al. 2015, Gildea and Jaeger 2015), we specify a parametric dependency tree ordering model.
%While the precise parameterization of the models is not necessarily crucial, it is crucial that the models are probabilistic, as this allows us to use gradient-based optimization.
%As we want to compare models optimized for language modeling with models 
We design our model to be particularly simple to make its parameters interpretable and comparable across models.

For each triple $\tau = $(Head POS, Dependency Label, Dependent POS), we have (1) a Direction Parameter $a_\tau$, and (2) a Distance Parameter $b_\tau$.
%An Ordering Model is an assignment of values $a_\tau, b_\tau \in \R$ to each of these triples.
Given a dependency tree, define a distribution over linearizations as follows:

\begin{enumerate}
\item Choose which dependents follow or precede their heads: Given a head $h$ and dependents $\{d_1, ..., d_k\}$, independently sample $O_i \sim Bernoulli(\sigma(a_{\tau_i}))$, where $i = 1, ..., k$, $\sigma$ is the inverse-logit function, and $\tau_i = (\text{POS of }h, \text{Label of }h \rightarrow d_i, \text{POS of }d_i)$. Order $d_i$ to the left of $h$ if $O_i = 0$, otherwise to the right of $h$.

\item %For any set of siblings $d_1, ..., d_k$ that are on the same side of their common head, the probability that $d_i$ is closest to the head, among these siblings, is $$\frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$$

Siblings $d_1, ..., d_k$ that were assigned to the same side of their head are ordered from the head outwards by iterating the following procedure:

Sample $i \in \{1, ..., k\}$ from the distribution $p_i = \frac{\exp{b_{\delta(d_i)}}}{\sum_{j=1}^k \exp{b_{\delta(d_j)}}}$.
Then put $d_i$ closest to the head, and remove $d_i$ from the set, so that $k-1$ siblings remain.
Repeat this procedure until only one sibling remains, which is then ordered farthest away from the head.
\end{enumerate}

\mhahn{TODO: Maybe I should create a figure to illustrate this?}

As (Futrell 2015, Gildea and Jaeger 2015),  we only consider projective linearizations.

one similar model is the one from Gildea \& Temperley (2007), where each dependent has a weight in [-1,1] which defines its position wrt the head and the other dependents.

Yours is more general in that it’s probabilistic; theirs is deterministic.

I also have a short EMNLP paper Futrell & Gibson (2015) which derives ordering models from generative dependency models such as Eisner Model C and the Dependency Model with Valence: the Eisner Model C linearization model seems like it would end up being the same as yours with $a_tau$ but not $b_tau$

Kristina Gulordava made an improved version of those models but I can’t find her paper right now; I’m pretty sure it was at CMCL

Another previous work is the reordering model from Galactic Dependencies: Wang & Eisner (2016)

There’s also a small but focused literature on more general linearization models, with a shared task in Belz et al. (2011)

I think the state of the art linearization model in terms of BLEU score is ZGen: https://github.com/SUTDNLP/ZGen but these highly optimized NLP models probably won’t work for a scientific study like you’re doing because of their complexity.

%One issue with most of these models (glidea & temperley, my model, galactic dependencies) is they only model order constraints in “local subtrees” consisting of heads and their immediate dependents. That may not always work (but it’s hard to think of a counterexample other than heavy NP shift!). Also, nearly all models that I’m aware of, including yours, only generate projective linearizations.

The only exceptions I know of are some things by Bernd Bohnet, who’s now at Google: http://www.aclweb.org/anthology/D12-1085. And the way his system does it strikes me as hacky. %Another issue with some models (glidea & temperley, my model based on Eisner Model C, your model) is they don’t model interactions among dependents: it may be that if some dependent A is present that changes the order preference for some other dependent B, but this isn’t captured. 
%
%There’s not a lot of work using linearization models to test scientific hypotheses. The only ones I can really think of are Gildea & Temperley and my stuff and other people who have used that method.

A big problem with that work is that there’s no way to quantify goodness of fit: I can say that real dependency length is minimized beyond some baseline but what I’d like to do is make a model that incorporates dependency length minimization and show that it fits the real orders better than a model without dependency length minimization. It seems like your method provides a way to do that. 
%
%On minimizing dependency length, I don’t know if you know about the algorithmic work on this problem. There’s polynomial time solutions to the minimal-dependency-length linearization problem subject to various constraints. Chung (1984) has the solution when there are no constraints, Gildea & Temperley (2007) have it under the constraint of projectivity, and Park & Levy (2009) do it under the constraint of mild nonprojectivity (gap degree 1). These wouldn’t fit into your gradient descent framework though.
%
%Future work may extend this to non-projective ones.

\subsection{Optimization Framework}

We are interested in finding ordering models $\thetad$ that minimize the cross-entropy loss:
$$-\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P(w_{i+1}|w_i)$$
where linearized sentences ${\bf w}$ are obained by ordering trees from the treebank according to the ordering model $\thetad$, and $P$ is a language model appropriate to this distribution of linearized sentences. %the distribution defined by $\thetad$.
%In the limiting case where there are no memory limitations, and $P$ is exactly the distribution defined by $\thetad$, this is the entropy of the language, and does not actually depend on the ordering model.
Note that the original word orders from the treebank do not enter this objective -- they in particular do not influence the language model $P$.

As we simultaneously look for an ordering model optimized for surprisal of the language model, and a language model that optimizes surprisal given the ordering model, this looks like a chicken-egg problem.
Fortunately, we can reduce this to an unconstrained optimization problem:
The appropriate language model for a given language is the one that minimizes expected surprisal.
Therefore, our minimization problem is equivalent to jointly minimizing over ordering models $\thetad$ and language models $\thetal$:
\begin{equation}\label{eq:surp-final-obj}
\min_{\thetad, \thetal} -\E_{{\bf w} \sim \thetad} \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})
\end{equation}

We solve~(\ref{eq:surp-final-obj}) by performing stochastic gradient descent over both sets of parameters $\thetad, \thetal$ simultaneously:
In each step, we sample a dependency tree from the treebank, then sample an ordering from the current setting of $\thetad$ to obtain a linearized sentence ${\bf w}$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator
$${- \partial_\thetad \left(\log P_\thetad({\bf w})\right) \cdot \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i}) \choose  - \partial_\thetal \sum_{i=1}^{\#{\bf w}} \log P_\thetal(w_{i+1}|w_{1\dots i})}$$
for the gradient ${ \partial_\thetad R(\thetad, \thetal) \choose \partial_\thetal R(\thetad, \thetal)}$.
This unbiased estimator is the ordinary gradient estimator for $\thetal$, and the REINFORCE estimator for $\thetad$~\cite{williams-simple-1992}.

\paragraph{Implemention}
To reduce the variance of the gradient estimator, we apply reward attribution (CITE) and use a word-dependent baseline.
We add an entropy reward (CITE) for the direction parameters $a_\tau$, weighted with $\alpha$.
In preliminary experiments, we found $\alpha=0.001$ and a learning rate of $0.01$ with momentum $0.9$ for $\thetad$ to give stable results.
We apply early stopping using a Monte Carlo estimate of the loss  (\ref{eq:surp-final-obj}) on the development set.
The hyperparameters for the LSTMs were as in the first experiment; however, we explicitly added input POS embeddings with 20 dimensions to encourage generalization in languages with small training corpora.
Embeddings were randomly initialized, as before.

\paragraph{Setup}
We carried this out on all Universal Dependencies\footnote{\url{http://universaldependencies.org/}} 2.1 treebanks for which a training set was available, a total of 50 languages.
For each language where multiple treebanks with training sets were available, we pooled their training sets; similarly for development sets.

\section{Results}

%\subsection{Correlation with Maximum-Likelihood Fit}
\subsection{Comparison with Ground-Truth Orderings}

\begin{figure}
\includegraphics[width=0.48\textwidth]{figures/freedom-lstm.pdf}
\caption{Optimizing word order for LSTM language modeling explains more of the language's actual rules in languages that have constrained word order.}\label{fig:lstm-languages}
\end{figure}

We first ask to what extent the ordering rules of the actual languages, as manifested in the corpora, can be recovered by optimizing for predictability.

Since we are interested in locality effects, we compare the distance parameters inferred by optimizing for language modeling with distance parameters inferred from the actual orderings via maximum-likelihood with an $\ell_2$ penalty.

We correlate the weights both at the level of dependency triples, and averaged at the levels of dependency labels, head POS, and dependent POS.
The correlations are significant at $p < 0.01$ for X languages.
In Figure~\ref{fig:lstm-languages}, we plot the per-language correlations on the level of triples against word order freedom.
The optimized languages explain more of the ground-truth ordering preferences in languages with low word order freedom ($r =$TODO, $p <$ TODO), in accordance with our previous result that LSTMs showed a stronger bias towards languages with fixed word order.

\subsection{Information Locality}

We predict that orderings optimized for prediction show information locality. %ease prediction with a limited memory. %show steeper decay of mutual information than random orderings.
We quantify Information Locality by computing the memory needs by the formula~(\ref{eq:memory-bound}).
As mutual information cannot be realiably estimated on many of the smaller corpora in the UD project, we compute MI on the level of POS tags, following~\cite{futrell-noisy-context-2017}.
As the summation in~(\ref{eq:memory-bound}) terminates when $t$ is greater than the length of any sentence, we can compute this quantity exactly.

We computed these quantitities for the optimized languages, the original languages, and for baseline  languages. For each language, we created 50 baseline ordering models, for which $a_\tau, b_\tau$ where drawn uniformly from $[-1, 1]$.
As mutual information between words is decreased if the language exhibits additional randomness (as our stochastic ordering models do) -- which decreases memory demands but increases entropy -- we hold the entropy of our languages constant by replacing each sampling step with taking the maximum -- that is, we take the zero temperature limit.
This ensures that all versions of a given language have the same entropy, and makes the numbers for memory demand comparable.

Results for four languages are shown in Figure~\ref{fig:memory-results}.
We find that, averaged across all 50 languages, optimized models require less memory than 93 \% of the respective baseline models.
In comparison, again averaged across languages, natural language requires less memory than 87 \% of baseline models.
%For half of the languages
For 75 \% of all languages, the LSTM-optimized version takes less memory than 95 \% of the baseline models.
For 37 \% of the languages, it also requires less memory than the original language.

We found that, as predicted, memory usage in real languages was correlated with word order freedom:
Languages which took less memory than all associated baseline languages had significantly lower word order freedom ($t = -4.061$, $p = 0.00018$).

\begin{figure*}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Arabic.pdf}
%%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Japanese.pdf}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-English.pdf}
%
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Korean.pdf}
%\includegraphics[width=0.48\textwidth]{figures/memory-demand-upos-Basque.pdf}
\includegraphics[width=0.98\textwidth]{figures/memory-demand-upos-facets.pdf}
%
\caption{Memory demands of real, optimized, and baseline versions of languages computed using~(\ref{eq:memory-bound}): Optimizing for ease of language modeling results in languages that require less memory in incremental prediction. On the level of POS tags, languages with fixed word order minimize memory need, while languages with very free word order, such as Basque, have similar memory requirements to baseline languages.}\label{fig:memory-results}
\end{figure*}

\subsection{Dependency Length Minimization}

\begin{figure}
\includegraphics[width=0.48\textwidth]{figures/depLength-facet.pdf}
\caption{Optimizing for LSTM language modeling (blue) shortens dependency lengths compared to baseline (green), though not always as strongly as natural language (red) does.}\label{fig:dep-lengths}
% Optimizing for LSTM modeling (green) tends to shorten more strongly than optimizing for bigram surprisal (red). \mhahn{choose languages}}
\end{figure}

A prominent manifestation of locality in language is dependency length minimization:
Languages have shorter dependencies than would be expected at random~\cite{futrell-large-scale-2015}. %, and under reasonable baselines.

%\cite{futrell-noisy-context-2017} 
As previous work (CITE) has argued that dependency length minimization can be derived from human memory limitations, we expect that languages optimized for language modeling should show short dependencies.
We computed per-sentence dependency lengths for the optimized ordering rules, creating a counterfactual corpus by sampling a linearization for each tree.
%As before, we sample linearizations at zero tempterature to eliminate nondeterminism that may lengthen average dependency length.
%We sampled trees with temperature $\tau = 1/100$ to eliminate nondeterminism (which may lengthen average dependency length).

%As a baseline, we also computed dependency lengths for ordering models that were trained with dependency length as the objective.
%(This differs from (CITE), who computed the per-sentence absolute minimum. Our baseline is more realistic in that it respects grammatical consistency across sentences).
As in the previous experiment and similar to \cite{futrell-large-scale-2015}, we considered ordering models with randomized $a_\tau, b_\tau$ as baselines.

In every one of the 50 languages, the optimized models show shorter dependencies than the baseline languages.
Results from four languages are shown in Figure~\ref{fig:dep-lengths}.
In 20 \% of the languages, they exceed the actual ground-truth orderings in dependency length minimization.

\subsection{Qualitative Predictions Across Languages}

We have seen that, for languages with fixed word order, optimizing for language models explains a significant amount of word order preferences.
We now examine what kinds of rules LSTMs prefer across languages. %take the grand average across languages.

We averaged distance weights across all languages, both for the optimized models, and for the maximum-likelihood models.
Within each language, we averaged across dependents, weighted by frequency of the dependency-dependent combination in the language.

The correlations between these averages across 33 dependency types was 0.50 \mhahn{TODO: I'll bootstrap over languages}, i.e., half of the variance in cross-linguistic ordering preferences as captured by $b_\tau$ could be explained as being optimized for language modeling. 
We then considered specifically the averages separately for dependencies headed by nouns and verbs, respectively.
We considered the five dependencies that bind most strongly and least strongly, that is, where $b_\tau$ was smallest or greatest.

%For nouns, we found that dependents typically considered to be part of the noun phrase (determiners, numerals, adjectives, modifiers, compound parts) bind strongly to the noun both in optimized and real languages.
%Copulas and adpositions, which in UD are considered dependents of the noun, bind very strongly in optimized models, while they are generally not considered part of the noun phrase in the linguistic literature -- nonetheless they tend to bind relatively close to the noun even in natural language.
%Intriguingly, the functions of adpositions and determiners, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.
%In both optimized and real languages, among the dependents that tend to be less close to the noun are relative clauses, arguments of copulas, adverbs, and coordination -- these are elements that are not part of the noun phrase as typically understood in the linguistic literature, and they also tend to be longer.

The closest dependencies were determiners, classifiers, case markers/adpositions, and numerals.
The farthest dependencies were relative clauses, conjuncts, adjectives, and adverbial clauses.
Intriguingly, the functions of determiners, classifiers, and adpositions, which bind most strongly in optimized models, are expressed morphologically on the noun in many languages, suggesting that grammaticalization into morphological markers is favored by locality considerations.

Turning now to dependencies headed by verbs, expletives, auxiliaries, markers, and adjectives bind most strongly. % bind very strongly to the verb both in optimized and in real languages, even they are often not understood as dependents of the verb in the linguistic literature.
Again, intriguingly, typical functions of auxiliaries and markers are expressed morphologically on the verb in many languages. %, suggesting again that grammaticalization into morphological markers may be favored by locality.
Perhaps counter to expectation, arguments (subjects, objects) are not generally closer than adjuncts, neither in optimized nor in real languages.
Similar to noun phrases, adjunct clauses stand out farther in optimized languages.
The farthest dependents are two kinds of nominal modifiers, dislocated elements, and objects.

\section{Related Work and Discussion}

\citet{gildea-human-2015} showed that a sample of five natural languages optimizes trigram surprisal and dependency length compared to random baselines, and also performed optimization with these objectives.
Our work links these findings to general properties of sequence models including LSTMs, and studies a much broader sample of languages. % and studies a much broader sample of languages. %, and (3) investigating qualitative linguistic predictions.

Our work relates to recent work showing that neural language models tend to focus strongly on recent words for prediction~\cite{daniluk-frustratingly-2017, khandelwal-sharp-2018}.
Our results suggests that this is explained both by general properties of sequence models with bounded memory, and by natural language favoring incremental prediction with such models.

Recent work has disucssed the adequacy of sequence models for language with its hierarchical structure (CITE). Our work adds to this by showing that some of the architectural biases in linear sequence modeling turn out to favor natural language, arguably because they are also present in human language processing. %natural language.

%Our findings highlight that there is a psycholinguistic explanation for the reasonable success of local methods, such as n-gram models, in NLP.
%We have shown that word order regularities in languages can partly be recovered from unordered data by optimizing for predictability.

\section{Conclusion}

We first showed that LSTM language models are biased towards orderings of information as found in natural language, and that this bias is stronger for languages with fixed word orders.
This is grounded in general properties of sequence modeling, which favors sequences exhibiting locality as found in natural language.
Asking which orderings would be optimal for LSTM language modeling, we then used reinforcement learning to create counterfactual versions of 50 diverse languages that are optimized for ease of language modeling.
We showed that, across corpora, LSTMs favor word orders that exhibit information locality and have short syntactic dependencies, similar to natural languages.
In languages with fixed word order, a significant amount of the actual word order preferences can be recovered by optimizing for language modeling, suggesting that word order rules in these languages can partly be explained as making prediction easier.

%- If we could adapt the language's word order rules to make LSTM language modeling easiest, what would the resulting language look like?

%- previous work has shown language has locality

%- we showed that LSTM language models are biased towards orders typically chosen in natural language, and that this preference increases as memory is restricted, and increases if memory is unbounded

%- this bias was shown to be caused particularly by function words -- words primarily carrying grammatical information.

%- we then used reinforcement learning to directly investigate what kinds of word orders are preferred by incremental language models. We chose LSTMs and bigram models as two extremes.

%- Across 50 languages, we showed that LSTMs prefer word orders that require little memory to model and have short syntactic dependencies, similar to natural language.

%- Using an information theoretic measure of memory need, we showed that word orders chosen in many natural language require less memory than would be expected at random

%- Our quantification of memory demands is entirely information-theoretic and makes no architectural assumptions beyond incrementality of sequence modeling.
%
%- We also showed that inferred ordering preferences explain a significant amount of ordering preferences in natural language, both for specific languages and across languages.
%
%- Our work has implications both for NLP and for syntactic typology. 
%
%
%Bounded context representations
%
%
%We observe that incrementally predicting language under memory limitations -- as instantiated by both recurrent language models and, arguably, human language processing, favors word orders that exhibit Information Locality: 
%
%We showed that recurrent language models indeed show a bias towards information locality.
%We then showed that information locality makes nontrivial and partly accurate typological predictions for word order across languages.

\section{More on Memory}

- tradeoff between memory and cross-entropy: let $M$ be a model with memory $Mem(M)$. For all $t$,

$Mem(M) + t CE[M,X] \geq Mem(X) + t H_{rate}[X]$

since there is a compression scheme for the sequences in $X$ of average length $t$ with size $Mem(M) + t CE[M,X]$, but the optimal scheme has length $Mem(X) + t H_{rate}[X]$

In particular, if $M$ is a lossy context encoding scheme and word-by-word prediction is optimal w.r.t. this, the word-by-word KL loss is lower bounded as
$$(CE[M,X] - H_{rate}[X]) \geq (Mem(X) - Mem(M))$$
Of course, $Mem(M)$ might be large, in which case this bound will be vacuous.
We could define the actual information content in $Mem(X)$ that is relevant to $X$ using cross-entropies -- this might then be larger or smaller than the actual information content, depending on the efficiency of the encoding (e.g., it could ignore certain kinds of context observations and thus require unnecessarily much memory).

\bibliographystyle{acl}
%\bibliography{extracted}
\bibliography{literature}

\end{document}

\begin{align*}
& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
\end{align*}

Is this useful?
If we know $I_t$, we can choose $T$ so as to make this bound as tight as possible.
To show the usefulness of this method, we consider the case where $I_t = \frac{\alpha}{t^\beta}$, for $\alpha > 0$, $\beta > 2$ (for smaller $\beta$, $M$ evaluates to $+\infty$).
Then $M = \alpha \sum_{t=1}^\infty \frac{1}{t^{\beta-1}}$, which is bounded as $\frac{\alpha}{\beta-1} = \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt \leq M \leq \alpha + \alpha \int_1^\infty \frac{1}{t^{\beta-1}} dt = \alpha + \frac{\alpha}{\beta-1} = \frac{\alpha\beta}{\beta-1}$.
The bound from the proposition evaluates to
\begin{align*}
\sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T &= \sum_{t \leq T} \frac{\alpha}{T t^{\beta-1}} + \sum_{t > T} \frac{\alpha}{t^\beta} - J/T \\
& \geq \int_1^T \frac{\alpha}{T t^{\beta-1}} dt + \int_T^\infty \frac{\alpha}{t^\beta} dt - J/T \\
& = \frac{\alpha (\beta-1)}{T} - \frac{\alpha(\beta-1)}{T^{\beta-1}} + \frac{\alpha\beta}{T^{\beta-1}} - J/T \\
&= \frac{1}{T} (\alpha(\beta-1) - J) + \frac{\alpha}{T^{\beta-1}}  \\
&\geq \frac{1}{T} (M(\beta-1) - \alpha - J) + \frac{\alpha}{T^{\beta-1}}  \\
%&\geq \frac{1}{T} (M - J) + \frac{M}{(\beta-1)T^{\beta-1}} \\
%& \geq M-J \ \ \ \ \ \ \  \text{(by setting $T=1$)} 
\end{align*}
%whose derivative by $T$ is
%$$\frac{-1}{T^2} (\alpha(\beta-1) - J) - \frac{\alpha(\beta-1)}{T^{\beta}}  $$
%which is zero at
%$$T = \left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)} $$
%Substituting back, we get the bound
%$$\frac{1}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{1/(\beta-2)}} (\alpha(\beta-1) - J) + \frac{\alpha}{\left(- \frac{\alpha(\beta-1)}{ (\alpha(\beta-1) - J)    } \right)^{(\beta-1)/(\beta-2)}}$$

%
%\begin{proposition}
%Let $\epsilon > 0$, $T \in \mathbb{N}$.
%If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$. TODO
%\end{proposition}
%\begin{proof}
%Let $f$ be the function mapping histories $X_{<t}$ to memory representations of the model. By assumption, $f(X_{<t})$ can be encoded with up to $J$ bits, so $H[f(X_{<t})] \leq J$.
%Without loss of generality, we can assume that the model is Bayes optimal: the model's distribution over $X_t$ after reading $X_{<t}$ is equal to $P(X_t | f(X_{<t}))$ (any other model would incur even higher surprisal).
%
%By definition, the difference between the model's surprisal and optimal surprisal is
%\begin{align*}
%& - \E_{X} \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t})\ \ \ \ \ \ \text{(for any $t$, due to stationarity)} \\
%& = - \frac{1}{T} \E_{X} \sum_{t=1}^T \log P(X_t | f(X_{<t})) - \log P(X_t | X_{<t}) \\
%& \geq  - \frac{1}{T} \E_{X} \log P(X_{1\dots T} | f(X_{<0})) - \log P(X_{1\dots T} | X_{<0})  \\
%& = \frac{1}{T} \left(H[X_{1\dots T}|f(X_{<0})] -  H[X_{1\dots T}|X_{<0}]\right) \\
%& = \frac{1}{T} (I[X_{1\dots T}|X_{<0}] - I[X_{1\dots T}|f(X_{<0})]) \\
%& = \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - I[X_{1\dots T}|f(X_{<0})]) \\
%& \geq \frac{1}{T} (\sum_{t \leq T} t I_t + T \sum_{t > T} I_t - J) \\
%& = \sum_{t \leq T} \frac{t}{T} I_t + \sum_{t > T} I_t - J/T \\
%%& \geq \frac{1}{T} \sum_{t \leq T}  I_t + \sum_{t > T} I_t - J/T \\
%%& \geq \frac{1}{T} (TODO - I[X_{1\dots T}|f(X_{<0})]) \\
%%& \geq \frac{1}{T} (TODO M - J - \epsilon) \\
%\end{align*}
%\end{proof}

%If we consider ratios between the memory bound and the predictability $I[X_t, X_{<t}]$, we can also express this as follows:
%\begin{corollary}
%Let $\epsilon > 0$, and let $P := I[X_t, X_{<t}]$ the predictability.
%Then If $\sum_{t > T} t I[X_t, X_0 | X_{1\dots t-1}] < \epsilon$, then a model with memory $J$ will experience average surprisal (that is, cross-entropy) at least $H[X_t|X_{<t}] + \frac{M-J-\epsilon}{T}$.
%\end{corollary}
%
%\begin{proof}
%$\frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{P} = \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > 0} I[X_t, X_0 | X_{1\dots t-1}]}  \leq \frac{\sum_{t > M/\epsilon} t I[X_t, X_0 | X_{1\dots t-1}]}{\sum_{t > M/\epsilon} I[X_t, X_0 | X_{1\dots t-1}]}< \epsilon$
%
%Let $T := M/\epsilon$.
%Then 
%\end{proof}
%
%

%\subsection{Incorporating All Lemmas}

%\subsection{Prediction on Raw Word Forms}

%\section{information locality and dependency length minimization}

%\subsection{Languages can trade off Memory Load and Predictability; Information Locality}
%
%%The quantity $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$
%
%We will describe a geometric view on $M$ that gives us a  quantitative measure of Information Locality.
%
%By information locality, we refer to the degree to which words that are predictive of each other are closer together.
%Stated differently, information in past observations that helps predict $X_t$ should be concentrated in recent observations $X_{t'}$ ($t' < t$) with $t'$ close to $t$.
%
%Information locality is maximised if all contextual information predicting $X_t$ is contained in $X_{t-1}$, in which case $J = I[X_t, X_{t-1}]$.
%
%To quantify information locality more generally, we consider the relation between the quantity
%$$J = I[X_{\geq t}, X_{<t}] = \sum_{t=1}^\infty\ t \cdot \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
%and the quantity
%$$P = I[X_t, X_{< t}] = \sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$$
%The first one quantifies how much information is transmitted between past and future, the second one captures the average predictability of single words from their prior context.
%Evidently, $P \leq J$ -- higher degrees of predictability will require more information to be stored from past observations.
%Stated differently, there is a tradeoff between predictability and memory: If the future is easier to predict from the past (at the same unconditional entropy), more memory is required to store all the relevant predictive information from the past.
%
%Minimizing $J$ unconditionally corresponds to an entirely unpredictable process.
%In contrast, if we minimize $J$ while holding $P$ constant, we obtain a formalization of the requirement of optimal information locality: This requires placing the predictive information closer to the present observation, keeping the total amount of predictive information constant..
%
%If there are no further constraints on the minimization, the minimum is attained when $P = J$.
%This is the case when all contextual information predicting $X_t$ is contained in $X_{t-1}$ -- this minimizes memory load, as only information from the last word needs to be stored.
%If $P$ is fixed, and $J$ increases, information locality decreases -- corresponding to increasing memory load.
%
%To understand this more generally, let us consider $\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]$ as a function of $t$.
%This function describes the amount of bits predictive of $X_t$ contained in $X_0$, the observation $t$ steps in the fast.
%The faster this function decays, the lower $J$ is.
%Information locality can be interpreted as requiring this function to decay fast -- $J$ is absolutely minimized if the function jumps to zero for $t > 1$.
%
%If this is a principle describing natural language, we expect that there will be further constraints, resulting in values of $J$ larger than $P$.
%We will study this in detail in the next section.
%
%
%We can observe that the area under the curve is equal to $P$ -- this quantifies the total amount of contextual information predicting the current observation.
%This suggests to consider the ratio
%$$L := \frac{J}{P}$$
%First, note that minimizing $J$ with $P$ fixed is equivalent to minimizing $L$.
%Second, expanding this quantity, we see
%$$L = \sum_{t=1}^\infty t \frac{\operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}{\sum_{t=1}^\infty\ \operatorname{I}[X_t, X_0 | X_1, ..., X_{t-1}]}$$
%That is, $L$ describes the $J$-coordinate of the center of mass of the area under the curve in (REF).
%A different way of understanding it is as the average distance of past information predictive of the current observation.
%With this interpretation of $L$, we can understand it as a direct fornalization of the notion of information locality.
%
%Let  us consider a second visualization by plotting $J$ against $P$.
%The area that is achievable by processes is the area below the diagonal.
%Information locality is maximized by processes on the diagonal; the distance from the diagonal quantifies the ratio $J/P$ and thus information locality.
%
%

### 2019-08-06 writeup/paper-short.tex

--- 2019-05-29 paper-short.tex
@@
   % 2-3 sentences detailed background, for scientists in related disciplines (linguistics, cogsci, computer science, info theory)
+%  In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension \cite{gibson2019efficiency,florian,hawkins,piantadosi2011word,zipf1949}.
   In linguistics and cognitive science, there has been a persistent suggestion that the properties of human languages can be explained in terms of maximizing communicative efficiency under the cognitive constraints involved in language production and comprehension.
@@
 We compared these against chance, that is, to tradeoffs that would be expected for languages that apply consistent grammatical rules as human language does, but have no bias towards any kind of processing efficiency or robustness.
-We draw on the Universal Dependncies project, which has constructed expert annotations of underlying hierarchical structures for dozens of languages~\cite{nivre-universal-2017}.
+We draw on the Universal Dependencies project, which has constructed expert annotations of underlying hierarchical structures for dozens of languages~\cite{nivre-universal-2017}.
 For each language, we constructed counterfactual versions that differ from real languages in their word order, holding all other properties constant.

### 2019-08-06 writeup/tables/medianDiff_0.tex

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-06 writeup/tables/medianDiff_2.tex

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIAN_DIFFS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-06 writeup/tables/medians_2.tex

Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-06 writeup/tables/medians_ngrams_1.tex

Hindi & Hungarian & Japanese & Norwegian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Hindi-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hungarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Japanese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Norwegian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Polish & Romanian & Slovak & Slovenian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Polish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Romanian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovak-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovenian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
 \\ 
Spanish & Swedish &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Spanish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Swedish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} &  & 
 \\

### 2019-08-06 writeup/tables/slice-hists_REAL_3.tex

Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf} &  & 
 \\

### 2019-08-06 writeup/background.tex
code

--- 2019-08-05 paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
-
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
-
-\usepackage{tikz-dependency}
-\usepackage{longtable}
-
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
-
-\usepackage{multirow}
-
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-\newcommand\note[1]{{\color{red}(#1)}}
-\newcommand\jd[1]{{\color{red}(#1)}}
-\newcommand\rljf[1]{{\color{red}(#1)}}
-\newcommand{\key}[1]{\textbf{#1}}
-
-\usepackage{amsthm}
-
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
-
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
-
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
-
-\begin{document}
-
-\maketitle
-
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
-
-\section{Introduction}
-
-\input{introduction.tex}
-
-\mhahn{pasted from the other draft}
-Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
-A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
-Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
-We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
-
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
-
-These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
-Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
-We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
-
-%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
-Here
-
-\mhahn{original intro}
-Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
-Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
-
-At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
-For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
-
-Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
-Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
-
-Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
-Analogous results also hold for language production by resource-constrained speakers.
-We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
-
-The remainder of this paper is structured as follows.
-In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
-In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
-We argue that information locality is a good model of the effects of memory constraints on language processing.
-In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
-In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
-Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
-% More detailed lit review
-
 A wide range of work has argued that natural language orders information in ways that reduce memory effort.
@@
 %also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-%\paragraph{Center Embeddings}
-%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-%\cite{gibson-linguistic-1998}
-%\paragraph{Other Psycholinguistic Predictions}
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-%\paragraph{Speakers}
-% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.3 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Classical Chinese, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 54 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.\mhahn{give an example to make clear what this refers to}
-
-Following Gildea and colleagues, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
-This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[w_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
-%\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\section{Experiment 3: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[w_t, w_0 | w_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[w_t, w_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $w_t = 0$, and with probability 1/2 all $w_t = 1$.
-%%$w_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $w_t$ is equal to the XOR of $w_{t-1}$ and $w_{t-N}$, such that each $w_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-can we say something about this
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
-\end{document}

### 2019-08-06 writeup/discussion.tex
code

--- 2019-08-05 paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
-
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
-
-\usepackage{tikz-dependency}
-\usepackage{longtable}
-
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
-
-\usepackage{multirow}
-
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-\newcommand\note[1]{{\color{red}(#1)}}
-\newcommand\jd[1]{{\color{red}(#1)}}
-\newcommand\rljf[1]{{\color{red}(#1)}}
-\newcommand{\key}[1]{\textbf{#1}}
-
-\usepackage{amsthm}
-
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
-
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
-
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
-
-\begin{document}
-
-\maketitle
-
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
-
-\section{Introduction}
-
-\input{introduction.tex}
-
-\mhahn{pasted from the other draft}
-Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
-A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
-Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
-We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
-
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
-
-These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
-Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
-We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
-
-%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
-Here
-
-\mhahn{original intro}
-Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
-Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
-
-At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
-For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
-
-Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
-Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
-
-Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
-Analogous results also hold for language production by resource-constrained speakers.
-We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
-
-The remainder of this paper is structured as follows.
-In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
-In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
-We argue that information locality is a good model of the effects of memory constraints on language processing.
-In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
-In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
-Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
-% More detailed lit review
-
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
-
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-%\paragraph{Center Embeddings}
-%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-%\cite{gibson-linguistic-1998}
-%\paragraph{Other Psycholinguistic Predictions}
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-%\paragraph{Speakers}
-% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.3 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Classical Chinese, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 54 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.\mhahn{give an example to make clear what this refers to}
-
-Following Gildea and colleagues, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
-This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[w_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
-%\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\section{Experiment 3: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-
-\section{Discussion}
-
 \subsection{Relation to Theories of Sentence Processing}
@@
 
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
-\end{document}
+%\paragraph{Center Embeddings}
+%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
+%\cite{gibson-linguistic-1998}
+%\paragraph{Other Psycholinguistic Predictions}
+% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
+%\paragraph{Speakers}
+% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
+This changes the interpretation of the mutual information.
+For the listener, it’s just redundancy.
+For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
+Then the speaker’s pressure is to keep conceptually related things close.

### 2019-08-06 writeup/main-experiment.tex
code

--- 2019-08-05 paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
-
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
-
-\usepackage{tikz-dependency}
-\usepackage{longtable}
-
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
-
-\usepackage{multirow}
-
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-\newcommand\note[1]{{\color{red}(#1)}}
-\newcommand\jd[1]{{\color{red}(#1)}}
-\newcommand\rljf[1]{{\color{red}(#1)}}
-\newcommand{\key}[1]{\textbf{#1}}
-
-\usepackage{amsthm}
-
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
-
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
-
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
-
-\begin{document}
-
-\maketitle
-
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
-
-\section{Introduction}
-
-\input{introduction.tex}
-
-\mhahn{pasted from the other draft}
-Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
-A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
-Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
-We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
-
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
-
-These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
-Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
-We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
-
-%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
-Here
-
-\mhahn{original intro}
-Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
-Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
-
-At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
-For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
-
-Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
-Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
-
-Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
-Analogous results also hold for language production by resource-constrained speakers.
-We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
-
-The remainder of this paper is structured as follows.
-In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
-In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
-We argue that information locality is a good model of the effects of memory constraints on language processing.
-In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
-In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
-Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
-% More detailed lit review
-
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
-
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-%\paragraph{Center Embeddings}
-%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-%\cite{gibson-linguistic-1998}
-%\paragraph{Other Psycholinguistic Predictions}
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-%\paragraph{Speakers}
-% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
 We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
@@
 \paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
+We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at 40 evenly spaced points of memory (selected individually for each language), using the binomial PDF.
+\mhahn{some standard stats reference}
 We consider the medians over all runs for the real language, and over all baselines grammars.
@@
 We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
+\mhahn{the resulting plots are not very intuitive, might scrap this}
 
@@
 %CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
+\mhahn{might scrap the CI}
 
@@
 
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
+We derive a CI for the quantile, taking the REAL values to be estimated exactly by their medians.
 
@@
 
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
+At a fixed memory value $\mu$, let $n_+$ be the better baseline samples, $n_-$ the worse (or equal) ones.
 
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
+We want to get a confidence bound $q$ on $P_{X \sim Baseline}(X < x_{real})$.
 
@@
 We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
 
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
+\includegraphics[width=0.95\textwidth]{../code/analyze_neural/figures/surprisal-branching-entropy-REAL.pdf}
 	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.5\textwidth]{../code/analyze_neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 	\caption{Histogram}\label{fig:hist-real}
@@
 \paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[w_t, w_0 | w_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[w_t, w_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $w_t = 0$, and with probability 1/2 all $w_t = 1$.
-%%$w_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $w_t$ is equal to the XOR of $w_{t-1}$ and $w_{t-N}$, such that each $w_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-can we say something about this
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
-\end{document}

### 2019-08-06 writeup/paper.tex
code

--- 2019-08-05 paper.tex
@@
 
-\mhahn{pasted from the other draft}
-Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
-A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
-Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
-We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
-
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
-
-These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
-Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
-We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
-
-%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
-Here
-
-\mhahn{original intro}
-Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
-Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
-
-At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
-For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
-
-Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
-Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
-
-Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
-Analogous results also hold for language production by resource-constrained speakers.
-We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
-
-The remainder of this paper is structured as follows.
-In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
-In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
-We argue that information locality is a good model of the effects of memory constraints on language processing.
-In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
-In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
-Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
+\section{Background}\label{sec:background}
 % More detailed lit review
 
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
+\input{background.tex}
 
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
+\section{Memory-Surprisal Tradeoff}\label{sec:ms-tradeoff}
 
@@
 
-%\paragraph{Center Embeddings}
-%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-%\cite{gibson-linguistic-1998}
-%\paragraph{Other Psycholinguistic Predictions}
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-%\paragraph{Speakers}
-% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
 \section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
 
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.3 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Classical Chinese, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 54 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.\mhahn{give an example to make clear what this refers to}
-
-Following Gildea and colleagues, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
-This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[w_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
-%\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\section{Experiment 3: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
+\input{main-experiment.tex}
 
@@
 
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[w_t, w_0 | w_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[w_t, w_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $w_t = 0$, and with probability 1/2 all $w_t = 1$.
-%%$w_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $w_t$ is equal to the XOR of $w_{t-1}$ and $w_{t-N}$, such that each $w_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-can we say something about this
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
+\input{discussion.tex}
 
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.95\textwidth]{../code/analyze_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 \caption{Histogram}\label{fig:hist-real}
@@
 	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
+\input{tables/corpusSizes.tex}
 \end{longtable}
@@
 	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
+\input{tables/samplesNumber.tex}
 \end{longtable}
@@
 	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
+\input{tables/boot_g_REAL.tex}
 \end{longtable}
@@
 \begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
+\input{tables/medians_REAL_0.tex}
 \end{longtable}
 	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$).
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
@@
 \begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
+\input{tables/medians_REAL_1.tex}
 \end{longtable}
@@
 \begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
+\input{tables/medians_REAL_2.tex}
 \end{longtable}
@@
 \begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
+\input{tables/medians_REAL_3.tex}
 \end{longtable}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
+\input{tables/slice-hists_REAL_0.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
+\input{tables/slice-hists_REAL_1.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
+\input{tables/slice-hists_REAL_2.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
+\input{tables/slice-hists_REAL_3.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+%\begin{table}
+%\begin{tabular}{cccccccccccccccccc}
+%\input{tables/quantiles_REAL_0.tex}
+%\end{tabular}
+%	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+%\end{table}
+%
+%\begin{table}
+%\begin{tabular}{cccccccccccccccccc}
+%\input{tables/quantiles_REAL_1.tex}
+%\end{tabular}
+%	\caption{Quantiles (part 2)}
+%\end{table}
+%
+%\begin{table}
+%\begin{tabular}{cccccccccccccccccc}
+%\input{tables/quantiles_REAL_2.tex}
+%\end{tabular}
+%	\caption{Quantiles (part 3)}
+%\end{table}
+%
+%
+%
+
+\begin{table}
+\begin{tabular}{l|ll||l|llllllllllllll}
+	Language & Base. & MLE & Language & Base. & MLE \\ \hline
+\input{tables/samplesNumber_ground.tex}
+\end{tabular}
+	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{tables/medians_0.tex}
+\end{tabular}
+	\caption{Experiment 3.
+Medians: For each memory budget, we provide the median surprisal for real and random languages.
+Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
+Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{tables/medians_1.tex}
 \end{tabular}
@@
 \begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{tables/medians_2.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccclll}
+\input{tables/medians_3.tex}
+\end{tabular}
+	\caption{Medians (cont.)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{tables/medianDiff_0.tex}
+\end{tabular}
+	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
+Lower values indicate lower surprisal compared to baselines.
+Solid lines indicate sample means.
+Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{tables/medianDiff_1.tex}
+\end{tabular}
+	\caption{Median Differences (Part 2)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{tables/medianDiff_2.tex}
+\end{tabular}
+	\caption{Median Differences (Part 3)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{ccccccccccccccccll}
+\input{tables/medianDiff_3.tex}
+\end{tabular}
+	\caption{Median Differences (Part 4)}
+\end{table}
+
+\begin{table}
 \begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
+\input{tables/quantiles_noAssumption_0.tex}
 \end{tabular}
@@
 \begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
+\input{tables/quantiles_noAssumption_1.tex}
 \end{tabular}
@@
 \begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_2.tex}
+\input{tables/quantiles_noAssumption_2.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
+\input{tables/medians_ngrams_0.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
+\input{tables/medians_ngrams_1.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
+\input{tables/medians_ngrams_2.tex}
 \end{tabular}
@@
 \begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
+\input{tables/medians_ngrams_3.tex}
 \end{tabular}

### 2019-08-06 writeup/tables/quantiles_noAssumption_0.tex
code

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\

### 2019-08-06 writeup/tables/quantiles_noAssumption_1.tex
code

Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\

### 2019-08-06 writeup/tables/quantiles_noAssumption_2.tex
code

Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.25\textwidth]{neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} &  & 
 \\

### 2019-08-06 writeup/toy-experiment.tex
directories
code

--- 2019-03-25 toy-experiment.tex
@@
 \begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/toy-mis.pdf}
-\includegraphics[width=0.45\textwidth]{toy/toy-t-mis.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/toy-mis.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/toy-t-mis.pdf}
 %
@@
 %
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/toy-mem-surp.pdf}
+	\caption{Tradeoff between listener memory and surprisal, for the two versions of the artificial language from \cite{fedzechkina-human-2017}.
+Language A requires less memory at the same level of surprisal.}\label{fig:toy-listener-tradeoff}
+\end{figure}

### 2019-08-06 writeup/tradeoff.tex
code

@@
 \begin{figure*}
-\includegraphics[width=0.45\textwidth]{toy/decay.pdf}
-\includegraphics[width=0.45\textwidth]{toy/memory.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/decay.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/memory.pdf}
 %
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/add-surp.pdf}
-\includegraphics[width=0.45\textwidth]{toy/lower-mem.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/add-surp.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/lower-mem.pdf}
 	\caption{Illustration for Proposition~\ref{prop:suboptimal}.
@@
 \begin{figure}
-\includegraphics[width=0.45\textwidth]{toy/listener-tradeoff.pdf}
+\includegraphics[width=0.45\textwidth]{../code/toy/figures/listener-tradeoff.pdf}
 	\caption{Listener's memory-surprisal tradeoff for the two processes in Figure~\ref{fig:basic}.

### 2019-08-06 code/analyze_neural/order_freedom/order_freedom.R

@@ -1,4 +1,4 @@
-fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
+fullData = read.csv("../../../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
     library(tidyr)
     library(dplyr)
     library(ggplot2)
@@ -7,7 +7,7 @@ fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", se
     data = data %>% mutate(MI_z = (MI-mean)/sd)
     barWidth = 0.05 #(max(data$MI) - min(data$MI))/30
 
-branching = read.csv("../code/branching_entropy/branching_entropy.tsv", sep="\t")
+branching = read.csv("../../branching_entropy/branching_entropy.tsv", sep="\t")
 data = merge(data, branching, by=c("Language"))
 
 dataReal = data %>% filter(Type %in% c("REAL_REAL")) %>% group_by(Language, Type) %>% summarise(MI_z=mean(MI_z), MIDiff = mean(MI-mean), BranchingEntropy=mean(BranchingEntropy)) %>% mutate(y=1)
@@ -17,9 +17,11 @@ plot = ggplot(dataReal, aes(x=-MIDiff, y=BranchingEntropy)) + geom_point()
 plot = plot + theme_classic() 
 plot = plot + theme(legend.position="none")   
 plot = plot + geom_text(aes(label=Language), hjust=0.8, vjust=1.1)
-#plot = plot + xlim(-2.5, 0.5)
+plot = plot + xlab("Surprisal Difference")
+plot = plot + ylab("Branching Direction Entropy")
+plot = plot + theme(axis.title=element_text(size=30))
 
-ggsave(plot, file=paste("figures/surprisal-branching-entropy-REAL.pdf", sep=""))
+ggsave(plot, file=paste("../figures/surprisal-branching-entropy-REAL.pdf", sep=""), width=15, height=7)
 
 
 
@@ -32,7 +34,7 @@ plot = plot + theme(legend.position="none")
 plot = plot + geom_text(aes(label=Language), hjust=0.8, vjust=1.1)
 #plot = plot + xlim(-2.5, 0.5)
 
-ggsave(plot, file=paste("figures/surprisal-branching-entropy-GROUND.pdf", sep=""))
+ggsave(plot, file=paste("../figures/surprisal-branching-entropy-GROUND.pdf", sep=""))
 
 
 

### 2019-08-06 code/analyze_neural/plot_wordsOnly_histByMem.R

@@ -1,12 +1,12 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 
-fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
+fullData = read.csv("../../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
-    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    dataL = read.csv(paste("../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     barWidth = (max(data$MI) - min(data$MI))/30
@@ -15,62 +15,9 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     return(plot)
 }
 
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Basque")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Bulgarian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Croatian")
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_histByMem_onlyREAL.R

@@ -1,81 +1,36 @@
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 # Plots CIs for the quantile
 
-
-fullData = read.csv("../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t") %>% filter(Type != "GROUND")
+fullData = read.csv("../../results/tradeoff/listener-curve-histogram_byMem.tsv", sep="\t") %>% filter(Type != "GROUND")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
-    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    dataL = read.csv(paste("../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     barWidth = (max(data$MI) - min(data$MI))/30
     plot = ggplot(data, aes(x=UnigramCE-MI, fill=Type, color=Type)) + theme_classic() + theme(legend.position="none")   + geom_density(data= data%>%filter(Type == "RANDOM_BY_TYPE"), aes(y=..scaled..))      + geom_bar(data = data %>% filter(Type %in% c("REAL_REAL", "GROUND")) %>% group_by(Type) %>% summarise(MI=mean(MI)) %>% mutate(y=1),  aes(y=y, group=Type), width=barWidth, stat="identity", position = position_dodge()) 
-    ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-HIST_byMem_onlyWordForms_boundedVocab_REAL.pdf", sep=""))
+
+
+    plot = plot + ylab("Density")
+    plot = plot + xlab("Surprisal")
+
+    plot = plot + theme(axis.text.x = element_text(size=20))  
+    plot = plot + theme(axis.text.y = element_text(size=20))  
+    plot = plot + theme(axis.title.x = element_text(size=20))  
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_means.R

@@ -1,11 +1,11 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 
-fullData = read.csv("../results/tradeoff/listener-curve-means.tsv", sep="\t")
+fullData = read.csv("../../results/tradeoff/listener-curve-means.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
     dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
@@ -14,62 +14,9 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     return(plot)
 }
 
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Basque")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Bulgarian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Croatian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Estonian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hebrew")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Japanese")
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_medianDiff.R

@@ -1,14 +1,12 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 # Plots median difference with confidence intervals
 
-fullData = read.csv("../results/tradeoff/listener-curve-ci-median_diff.tsv", sep="\t")
+fullData = read.csv("../../results/tradeoff/listener-curve-ci-median_diff.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
-#    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
- #   UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     plot = ggplot(data, aes(x=Memory, y=-EmpiricalMedianDiff, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=-MedianDiff_Lower), linetype="dashed") + geom_line(aes(x=Memory, y=-MedianDiff_Upper), linetype="dashed")
     plot = plot + xlab("Memory") + ylab("Difference in Median Surprisal")
@@ -17,62 +15,9 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     return(plot)
 }
 
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_mediansAndQuantiles.R

@@ -1,26 +1,20 @@
-# DOES not produce reasonable results. Better to just provide empirical quantile plots.  
-
 library(tidyr)
-    library(dplyr)
-    library(ggplot2)
+library(dplyr)
+library(ggplot2)
 
 # Plots CIs for the quantile
 
 # Plots medians with confidence intervals
 
-fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") 
+fullData = read.csv("../../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") 
 
-randomRuns = read.csv("../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
+randomRuns = read.csv("../../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
 
 library(MASS)
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
-    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    dataL = read.csv(paste("../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
-
     dataR = randomRuns %>% filter(Language == language)
     data = fullData %>% filter(Language == language)
     plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) 
@@ -28,82 +22,24 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     plot = plot + theme_classic() + theme(legend.position="none") 
     plot = plot + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
 
-quantiles = data.frame(Memory = c(), MI = c(), Density = c())
-for(i in (1:max(dataR$Position))) {
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_mediansAndQuantiles_onlyREAL.R

@@ -1,109 +1,56 @@
-# DOES not produce reasonable results. Better to just provide empirical quantile plots.  
 
 library(tidyr)
-    library(dplyr)
-    library(ggplot2)
+library(dplyr)
+library(ggplot2)
 
-# Plots CIs for the quantile
+# This is for the median and its CI
+fullData = read.csv("../../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
 
-# Plots medians with confidence intervals
-
-fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
-
-randomRuns = read.csv("../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
+# This is for quantifying the quantiles
+randomRuns = read.csv("../../results/tradeoff/listener-curve-interpolated.tsv", sep="\t") %>% filter(Type == "RANDOM_BY_TYPE")
 
 library(MASS)
 
-memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
-    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+make_plot = function(language) {
+    dataL = read.csv(paste("../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
 
     dataR = randomRuns %>% filter(Language == language)
     data = fullData %>% filter(Language == language)
+
+    # Plot Median
     plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) 
     plot = plot + geom_line(size=2)
     plot = plot + theme_classic() + theme(legend.position="none") 
+
+    # Add CI around the median
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_quantile.R

@@ -1,20 +1,18 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 # Plots CIs for the quantile
 
-fullData_ConfidenceLowerBound = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", sep="\t")
-fullData_ConfidenceLowerBound_05 = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05.tsv", sep="\t")
+fullData_ConfidenceLowerBound = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", sep="\t")
+fullData_ConfidenceLowerBound_05 = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05.tsv", sep="\t")
 
-fullData_BinomialTest = read.csv("../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t")
+fullData_BinomialTest = read.csv("../../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
     data = fullData_ConfidenceLowerBound %>% filter(Language == language)
     data2 = fullData_BinomialTest %>% filter(Language == language)
     data3 = fullData_ConfidenceLowerBound_05 %>% filter(Language == language)
-    #data = merge(data, data2, by=c("Language", "Position", "Type"))
-    #data$Memory = data$Memory.x
     plot = ggplot(data, aes(x=Memory, y=LowerConfidenceBound, fill=Type, color=Type))
     plot = plot + geom_line(size=1, linetype="dotted") 
     plot = plot + geom_line(data=data3, size=2, linetype="dashed") 
@@ -28,67 +26,13 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     plot = plot + ylab("Quantile")
     plot = plot + xlab("Memory")
     plot = plot + theme(text = element_text(size=20))
-   # + geom_line(aes(x=Memory, y=-MedianDiff_Upper))
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_quantile_noAssumption.R

@@ -1,0 +1,39 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+
+# Plots CIs for the quantile
+
+# ../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption-05.tsv
+fullData_ConfidenceLowerBound = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption.tsv", sep="\t")
+fullData_ConfidenceLowerBound_05 = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption-05.tsv", sep="\t")
+
+fullData_BinomialTest = read.csv("../../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t")
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    data = fullData_ConfidenceLowerBound %>% filter(Language == language)
+    data2 = fullData_BinomialTest %>% filter(Language == language)
+    data3 = fullData_ConfidenceLowerBound_05 %>% filter(Language == language)
+    plot = ggplot(data, aes(x=Memory, y=LowerConfidenceBound, fill=Type, color=Type))
+    plot = plot + geom_line(size=1, linetype="dotted") 
+    plot = plot + geom_line(data=data3, size=2, linetype="dashed") 
+    plot = plot + geom_line(data=data2, aes(x=Memory, y=BetterEmpirical), size=2)
+    data2 = data2 %>% mutate(pValue_print = ifelse(round(pValue,5) == 0, "p<0.00001", paste("p=", round(pValue,5), sep="")))
+    plot = plot + geom_text(data=data2 %>% filter(Position %% 9 == 0, Type == "REAL_REAL"), aes(x=Memory, y=BetterEmpirical+0.1, label=pValue_print), size=3)
+    plot = plot + geom_text(data=data2 %>% filter(Position %% 9 == 0, Type == "GROUND"), aes(x=Memory, y=BetterEmpirical+0.05, label=pValue_print), size=3)
+    plot = plot + theme_classic() 
+    plot = plot + theme(legend.position="none") 
+    plot = plot + ylim(0,1.1)
+    plot = plot + ylab("Quantile")
+    plot = plot + xlab("Memory")
+    plot = plot + theme(text = element_text(size=20))
[…]

### 2019-08-06 code/analyze_neural/plot_wordsOnly_quantile_onlyREAL.R

@@ -1,13 +1,13 @@
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 # Plots CIs for the quantile
 
-fullData_ConfidenceLowerBound = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", sep="\t") %>% filter(Type != "GROUND")
-fullData_ConfidenceLowerBound_05 = read.csv("../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05.tsv", sep="\t") %>% filter(Type != "GROUND")
+fullData_ConfidenceLowerBound = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile.tsv", sep="\t") %>% filter(Type != "GROUND")
+fullData_ConfidenceLowerBound_05 = read.csv("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05.tsv", sep="\t") %>% filter(Type != "GROUND")
 
-fullData_BinomialTest = read.csv("../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t") %>% filter(Type != "GROUND")
+fullData_BinomialTest = read.csv("../../results/tradeoff/listener-curve-binomial-test.tsv", sep="\t") %>% filter(Type != "GROUND")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     data = fullData_ConfidenceLowerBound %>% filter(Language == language)
@@ -27,67 +27,13 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     plot = plot + ylim(0,1.1)
     plot = plot + ylab("Quantile")
     plot = plot + xlab("Memory")
-    # + geom_line(aes(x=Memory, y=-MedianDiff_Upper))
     ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL.pdf", sep=""), height=3.5, width=4.5)
     return(plot)
 }
 
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
[…]

### 2019-08-06 code/medians/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py

@@ -1,3 +1,7 @@
+# This script is used to compute `Confidence Interval for Medians'
+
+# It is called by code/medians/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_All.py
+
 # Better than yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MaxControl.py by modeling the median of REAL
 
 import sys
@@ -21,7 +25,7 @@ def readTSV(x):
       for i in range(len(data)):
           data[i][column] = vals[i]
     return (header, data)
-with open("../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+with open("../../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
      data = readTSV(inFile)
 #print(len(data))
 
@@ -135,32 +139,11 @@ for typ, interpolated in interpolatedByTypes.iteritems():
              bestCI = (lower, upper, coverage)
 #       print(typ, i, bestCI)
           # median is in [lower, upper]
-
+       empirical_median = float(interpolated[:,i].median())
+       assert bestCI[0] <= empirical_median
        # TODO don't trust the CI
-       print "\t".join(map(str,[language, typ, i, float(xPoints[i]), float(interpolated[:,i].median()), bestCI[0], bestCI[1], bestCI[2]]))
+       print "\t".join(map(str,[language, typ, i,                float(xPoints[i]), empirical_median,     bestCI[0], bestCI[1], bestCI[2]]))
+                             # "Language", "Type", "Position",  "Memory",           "MedianEmpirical",    "MedianLower", "MedianUpper", "Level"
 
 
 
-
-
-       #quit()
-
-#       comparison = interpolatedByTypes["RANDOM_BY_TYPE"] < minReal.unsqueeze(0)
-#       comparisonReverse = interpolatedByTypes["RANDOM_BY_TYPE"] > minReal.unsqueeze(0)
-#    
-#     #  print(comparison.size())
-#       comparisonMean = comparison.float().sum(dim=0)
-#       comparisonReverseMean = comparisonReverse.float().sum(dim=0)
-#    
-#
-#
[…]

### 2019-08-06 code/preparePaper/quantiles_noAssumption.py

@@ -1,0 +1,90 @@
+# ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
+
+with open("../../ud_languages.txt", "r") as inFile:
+   languages = inFile.read().strip().split("\n")
+
+languages = sorted(list(set(languages)))
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data[-1]) == 1 and data[-1][0] == "":
+       del data[-1]
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+def h(frame, line, name):
+    return line[frame[0][name]]
+
+
+#with open("tradeoff/effectSize.tsv", "r") as inFile:
+#   effectSize = readTSV(inFile)
+
+#with open("tradeoff/effectSize_diff.tsv", "r") as inFile:
+#   effectSize_diff = readTSV(inFile)
+
+with open("../../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
+   stats = readTSV(inFile)
+languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
+
+
+
+with open("../corpusSizes.tsv", "r") as inFile:
+  corpusSizes = readTSV(inFile) #dict([x.split("\t") for x in inFile.read().strip().split("\n")])
+languageKey_corpusSizes = dict([(h(corpusSizes, line, "Language"), line) for line in corpusSizes[1]])
+
+
+
+
+with open("../../results/tradeoff/listener-curve-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
+   listener_curve = readTSV(inFile)
+
+languageKey_listener_curve = dict([(h(listener_curve, line, "language"), line) for line in listener_curve[1]])
+
+entries = []
+
+for language in languages:
+#   line = languageKey[language]
+
[…]

### 2019-08-06 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py

@@ -21,7 +21,7 @@ def readTSV(x):
       for i in range(len(data)):
           data[i][column] = vals[i]
     return (header, data)
-with open("../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+with open("../../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
      data = readTSV(inFile)
 #print(len(data))
 
@@ -137,6 +137,8 @@ for real in ["REAL_REAL", "GROUND"]:
      # The goal here is to provide a confidence lower bound on worseRandomCount
      if worseRandomCount == 0:
          print "\t".join(map(str,[language, real, i, 0.0, 0.0, float(xPoints[i])]))
+#                              ["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+
          bound = 0
      else:
    #     print(hereRandom.size())
@@ -161,9 +163,10 @@ for real in ["REAL_REAL", "GROUND"]:
              p1 = math.pow(1-unaccountedFor, worseRandomCount)
           else:
              p1 = 1-(scipy.stats.binom_test(x=sameRandomCount+betterRandomCount, n=worseRandomCount+sameRandomCount+betterRandomCount, p=unaccountedFor, alternative="greater"))
-        #  print(unaccountedFor, p1)
           if p1 < 0.001:
              print "\t".join(map(str,[language, real, i, 1-float(percentile), p1, float(xPoints[i])]))
+#                ["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+             assert 1-float(percentile) <= (worseRandomCount) / (sameRandomCount + betterRandomCount), (unaccountedFor, 1-float(percentile), (worseRandomCount) / (sameRandomCount + betterRandomCount))
              hasFound =True 
              break
          if hasFound:
@@ -171,4 +174,8 @@ for real in ["REAL_REAL", "GROUND"]:
         assert hasFound, (worseRandomCount, sameRandomCount, betterRandomCount)
         if not hasFound:
              print "\t".join(map(str,[language, real, i, 0.0, 0.0, float(xPoints[i])]))
+#                                  ["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+
+
+
 

### 2019-08-06 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py

@@ -21,7 +21,7 @@ def readTSV(x):
       for i in range(len(data)):
           data[i][column] = vals[i]
     return (header, data)
-with open("../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+with open("../../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
      data = readTSV(inFile)
 #print(len(data))
 
@@ -136,13 +136,15 @@ for real in ["REAL_REAL", "GROUND"]:
 
      # The goal here is to provide a confidence lower bound on worseRandomCount
      if worseRandomCount == 0:
+         print "\t".join(map(str,[language, real, i, 0.0, 0.0, float(xPoints[i])]))
+#                              ["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+
          bound = 0
      else:
    #     print(hereRandom.size())
         largestPossibleValue = xPoints[i]
         #print(i, largestPossibleValue, medians[real][i], hereRandom)
         # if the median is at most the best sample
-        targetLevel = 0.05
         #unaccountedForTarget = 1-math.pow(targetLevel, 1/(float(hereRandom.size()[0])))
         randomBestWorse = float(hereRandom[int(worseRandomCount)-1])
         assert randomBestWorse < float(medians[real][i])
@@ -153,13 +155,15 @@ for real in ["REAL_REAL", "GROUND"]:
           assert unaccountedFor > 0
           assert unaccountedFor < 1
           
-          percentile = unaccountedFor #* (largestPossibleValue - medians[real][i] + 0.0001) / (largestPossibleValue - randomBestWorse + 0.0001)
+          percentile = unaccountedFor 
           if sameRandomCount+betterRandomCount == 0:
              p1 = math.pow(1-unaccountedFor, worseRandomCount)
           else:
              p1 = 1-(scipy.stats.binom_test(x=sameRandomCount+betterRandomCount, n=worseRandomCount+sameRandomCount+betterRandomCount, p=unaccountedFor, alternative="greater"))
-          if p1 < 0.05:
-             print "\t".join(map(str,[language, real, 1-float(percentile), p1]))
+          if p1 < 0.001:
+             print "\t".join(map(str,[language, real, i, 1-float(percentile), p1, float(xPoints[i])]))
[…]

### 2019-08-06 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_05.py

@@ -1,0 +1,169 @@
+# Better than yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MaxControl.py by modeling the median of REAL
+
+import sys
+
+language = sys.argv[1]
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data) < 2:
+       return (header, [])
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+with open("../../results/raw/word-level/"+language+"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", "r") as inFile:
+     data = readTSV(inFile)
+#print(len(data))
+
+
+
+#data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
+
+import torch
+
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+
+matrix = [[g(data, "ModelID", i), g(data, "Distance", i), g(data, "ConditionalMI", i), g(data, "UnigramCE", i)] for i in range(len(data[1]))]
+
+matrixByType = {}
+misByType = {}
+#unigramCEByType = {}
+for i in range(len(data[1])):
+    typ = g(data, "Type", i)
+#    print(i,typ, len(data[1]))
+    if typ not in matrixByType:
+        matrixByType[typ] = []
+    matrixByType[typ].append(matrix[i])
+for typ in matrixByType:
+   matrixByType[typ] = torch.FloatTensor(matrixByType[typ]).view(-1, 19, 4)
+   #print(matrixByType[typ][0])
+   misByType[typ] = matrixByType[typ][:,:,2]
+#print(misByType["RANDOM_BY_TYPE"])
+
+#data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
+distance = torch.FloatTensor(range(1,20))
[…]

### 2019-08-06 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_05_All.py

@@ -1,0 +1,13 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption-05.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_05.py", language]).strip()
+
+

### 2019-08-06 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_All.py

@@ -1,0 +1,13 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py", language]).strip()
+
+

### 2019-08-06 writeup/control-ngrams.tex
paper

--- 2019-08-06 writeup/paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
-
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
-
-\usepackage{tikz-dependency}
-\usepackage{longtable}
-
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
-
-\usepackage{multirow}
-
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-\newcommand\note[1]{{\color{red}(#1)}}
-\newcommand\jd[1]{{\color{red}(#1)}}
-\newcommand\rljf[1]{{\color{red}(#1)}}
-\newcommand{\key}[1]{\textbf{#1}}
-
-\usepackage{amsthm}
-
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
-
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
-
-\frenchspacing
-%\def\baselinestretch{0.975}
-
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
-
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
-
-\begin{document}
-
-\maketitle
-
-%
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
+%bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])
+%bounds.append(["gamma", int, 1, 2, 3, 4, 5, 8, 10, 15, 20, 25, 30]) # , 200, 300
+%bounds.append(["delta", int, 0.1, 0.2, 0.5, 1.0 , 2.0, 3.0, 4.0, 5.0, 8.0, 10.0]) #, 1024]) #, 1024]) # 64, 128,
+%bounds.append(["cutoff", int, 2,3,4,5,6,7,8,9,10]) #,7,8,9,10]) #, 1024]) #, 1024]) # 64, 128,
 %
 
-\section{Introduction}
-
-\input{introduction.tex}
-
-\section{Background}\label{sec:background}
-% More detailed lit review
-
-\input{background.tex}
-
-\section{Memory-Surprisal Tradeoff}\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-\input{main-experiment.tex}
-
-\section{Discussion}
-
-\input{discussion.tex}
-
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{../code/analyze_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{tables/corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{tables/samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{tables/boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$).
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_0.tex}
-%\end{tabular}
-%	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-%\end{table}
-%
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_1.tex}
-%\end{tabular}
-%	\caption{Quantiles (part 2)}
-%\end{table}
-%
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_2.tex}
-%\end{tabular}
-%	\caption{Quantiles (part 3)}
-%\end{table}
-%
-%
-%
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{tables/samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
+We use Interpolated Kneser-Ney.
+Hyperparameters are tuned with the same strategy as for the neural network models.
 
@@
 \end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
-\end{document}

### 2019-08-06 writeup/paper.tex
paper

@@
 
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
+\input{control-ngrams.tex}
 
 %-- English, Korean, Russian
+%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
+%-- character-level Russian
+%\section{Character-Level Modeling}
+%\section{Non-UD Dependency Treebanks}
+%- other treebanks
+%-- spoken Japanese (T{\"u}ba-J/S)
+%-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
+%-- another Chinese dependency treebank LDC2012T05
+%Due to the sizes of these treebanks, can also do experiment with full word forms.
 %
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
 %
-%-- character-level Russian
+%\section{Constituency Treebank}
 %
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
+%-- Penn treebank \citep{marcus-building-1993}
+%
+%-- spoken English (T{\"u}ba-E/S)
+%
+%-- spoken German (T{\"u}ba-D/S)
+%
+%-- Chinese treebank \citep{xue-chinese-2013}
 

### 2019-08-12 writeup/control-ngrams.tex
code

@@
 \end{table}
+
+% ../../writeup/tables/quantiles_REAL_NGRAMS_0.tex
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{tables/quantiles_REAL_NGRAMS_0.tex}
+\end{tabular}
+	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
+At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
+Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{tables/quantiles_REAL_NGRAMS_1.tex}
+\end{tabular}
+	\caption{Quantiles (part 2)}
+\end{table}
+
+\begin{table}
+\begin{tabular}{cccccccccccccccccc}
+\input{tables/quantiles_REAL_NGRAMS_2.tex}
+\end{tabular}
+	\caption{Quantiles (part 3)}
+\end{table}

### 2019-08-12 writeup/discussion.tex
code

@@
-\subsection{Relation to Theories of Sentence Processing}
+\subsection{Other Models of Sentence Processing}
 
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
+\paragraph{Early Models}
 
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
 Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
@@
 
-\subsection{Relation to other work}
+\paragraph{Dependency Locality}
+The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
+Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
+In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
+In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
+
+also surprisal -- integration cost
+
+\paragraph{Cue-Based Retrieval}
+
+\paragraph{Lossy-Context Surprisal}
+\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
+In particular, they consider loss models that delete, erase, or replace words in the past.
+Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
+A listener will incur surprisal
+$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
+where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
+Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
+Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
+
+\subsection{Statistical Studies of Language}
 
@@
 
+\mhahn{Q what are some of the things that need to go here?}
+
 \paragraph{Speakers vs Listeners}

### 2019-08-12 writeup/main-experiment.tex
code

@@
 We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
+We compare to counterfactual reorderings that differ from the actual languages in their word order rules, i.e., the rules that determine how syntactic structures are ordered into strings of words.\mhahn{where introduce the idea of word order}.
+To create such counterfactual reorderings, we parameterize these word order rules using a method introduced by Gildea and colleagues \citep{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
 
@@
 
+\mhahn{at some point the whole idea of syntactic structures have to be introduced}
+
 \subsection{Data}
 We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
+The Universal Dependencies project has compiled such annotated corpora for several dozen languages~\citep{nivre-universal-2017}.
+These are annotated in the format of Dependency Grammar.
 
@@
 A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
+Dependency trees can be defined in terms of many different syntactic theories \citep{corbett1993heads}.
 Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
+\mhahn{currently this is taken from the Greenberg paper}
 
@@
 For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
+We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL for English, CFL for Chinese}
 Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
@@
 \subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
+To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{We note that the state-of-the-art is now achieved by other models when extremely large amounts of training data (on the order of several dozen GB of texts) are available.
+In comparison, LSTMs remain state of the art on small/medium training data such as the UD corpora.} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
 We provide data from alternative estimation methods in the SI.
-
-%\paragraph{Data}
-
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
 
@@
 This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
+x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution $p(w_t|w_1...w_{t-1})$ over the next word $w_{t}$ given preceding words $w_1 ...
+w_{t-1}$.
+Our model specification follows (CITE someone who used exactly the same model).
 
@@
 %We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
+At the beginning of training, the parameters $\theta$ are randomly initialized to some setting $\theta_0$.
 
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
+The training corpus is chopped into word sequences $w_1 ...
+w_{T_{max}}$ of length ${T_{max}}$, where ${T_{max}}$ is the highest $T$ for which we estimate $I_T$.
+%  (${T_{max}} = 20$ in our experiments).
+We use Stochastic Gradient Descent to optimize the parameters $\theta$ so as to minimize the surprisal
 \begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
+	\frac{1}{T_{max}} \sum_{i=1}^{T_{max}} \log p_\theta(w_i|w_1...w_{i-1})
 \end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
 When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
+
 Once all sequences have been processed, we start another pass through the training data.
 Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
+After each pass through the training data, the average surprisal~(\ref{eq:train}) at the current parameter setting $\theta$ is evaluated on the held-out partition.
 We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
 
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
+In our experiments, we chose $T_{max} = 20$.
+Prior work has found that the probabilities $p(w_t|w_1...w_{t-1})$ are dominated by a small number of preceding words \citep{daniluk2017frustratingly}, suggesting that $I_t$ will be close to zero for $t$ greater than 20.
 
@@
 
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
+The LSTM model has a set of numerical \emph{hyperparameters} that need to be specified before parameter estimation, such as the number of recurrent neurons and the learning rate.
+%namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
+For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
+We optimized the hyperparameters to minimize average surprisal~(\ref{eq:train}) on the held-out partition resulting at the end of parameter estimation, on languages generated from random word order grammars.
+This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis that real orders result in better memory-surprisal tradeoffs than counterfactual orders.
 
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
+Due to reasons of computational efficiency, neural language models can only process a bounded number of distinct words in a single language (CITE).
 For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
 Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
+This applied to 37 languages, affecting an average of 11~\% of words in these languages.
 We believe that this modeling limitation does not affect our results for the following reasons.
@@
 
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
 \paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
 
-The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
+The quantity $I_t$ in~(\ref{eq:memory-bound}) is equal to the difference 
 \begin{equation}
-H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
+\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}] = H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
 \end{equation}
-For each word in the held-out partition, we compute the difference
+This quantity can be estimated as follows.
+After estimating the parameter vector $\theta$, we compute the following difference for each word $w_t$ ($t$ ranging from $T_{max}$ up to the length of the held-out partition) in the held-out partition:
 \begin{equation}
-	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
+	\frac{1}{|HeldOut|-t} \sum_{i=t}^{|HeldOut|} \log P_\theta[w_i | w_{i-t}, w_{i-t+1}, ..., w_{i-1}] - P_\theta[w_i | w_{i-t+1}, ..., w_{i-1}]
 \end{equation}
@@
 
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
+%We then used linear interpolation to interpolate the surprisal value for memory values in between these values.
+(TODO make a figure).\mhahn{mention linear interpolation already when explaining how to get tradeoff from theorem}
+%This is justified theoretically (TODO maybe discuss when introducing the theorem).
 
 We estimate the unigram entropy $H[w_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
 
@@
 We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
+\mhahn{Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).}
 
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at 40 evenly spaced points of memory (selected individually for each language), using the binomial PDF.
-\mhahn{some standard stats reference}
-We consider the medians over all runs for the real language, and over all baselines grammars.
+\paragraph{Median Surprisal and Confidence Interval for Medians}
+We compute surprisal at 40 evenly spaced points of memory (selected individually for each language), over all runs for the real language, and over all baselines grammars.
+We then compute the median surprisal over all runs for the real language, and over all baselines grammars.
+We compute a confidence interval for this median surprisal, using the binomial PDF. \mhahn{some standard stats reference}
+This interval is exact, without parametric assumptions or asymptotic approximations.
 
@@
 
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-\mhahn{the resulting plots are not very intuitive, might scrap this}
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
+%\paragraph{CI for Median Difference}
+%We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
+%\mhahn{the resulting plots are not very intuitive, might scrap this}
+%
+%% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
 
 \paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
+At five evenly spaced memory values $\mu$ (chosen per language), we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
 Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
 For each level $\mu$ of memory, we consider the null hypothesis that
@@
 \end{equation}
+In this test, we approximate the surprisal value of the real language by the \emph{sample median}.
 %In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
@@
 %CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-\mhahn{might scrap the CI}
+%\mhahn{might scrap the CI}
 
@@
 
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
+To provide estimates of the degree of optimization of real orders, we also compute what \emph{fraction} of baseline languages provides better tradeoffs than the real language.
+Specifically, for each level of memory $\mu$, we estimate what percentage of baseline languages have lower (or equal) surprisal than the real language.
 This is described in Figure~\ref{fig:quantile-pointwise}.
+We use the Binomial Test to derive a confidence lower bound for this quantile (CITE).
+Again, we take the REAL values to be estimated exactly by their medians.
 
-We derive a CI for the quantile, taking the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-At a fixed memory value $\mu$, let $n_+$ be the better baseline samples, $n_-$ the worse (or equal) ones.
-
-We want to get a confidence bound $q$ on $P_{X \sim Baseline}(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
+%We want to create a CI at each Memory value for the quantile.
+%
+%At a fixed memory value $\mu$, let $n_+$ be the better baseline samples, $n_-$ the worse (or equal) ones.
+%
+%We want to get a confidence bound $q$ on $P_{X \sim Baseline}(X < x_{real})$.
+%Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
+%Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
+%We minimize $q$ subject to $p < 0.05$.
+%This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
+%
+%Also the following does not assume unimodality, and ends up getting about the same intervals
+%% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
 
 \paragraph{Global Quantile Estimate}
+
+To provide a stopping criterion, we additionally need a global measure of the degree of optimization of the real language.
+
+Due to the use of bootstrapping, the confidence intervals are not exact.
 
@@
 For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
 
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{../code/analyze_neural/figures/surprisal-branching-entropy-REAL.pdf}
+\includegraphics[width=0.95\textwidth]{../code/visualize_neural/figures/surprisal-branching-entropy-REAL.pdf}
 	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
@@
 
+\mhahn{One thing that has to be discussed: the absolute values differ between languages}
+
+\section{Experiment 2: Free Word Orders}
+
+Here we match the branching entropies of the baseline grammars with those of the real language.
+
 \section{Experiment 3: Fixed Word Orders}
 
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
+A possible concern is that the optimization observed in Experiment 1 might be an artifact of the grammar formalism.
+To address this possibility, we repeated the experiment comparing baseline grammars to grammars that represent the word order of the real languages as faithfully as is possible with the grammar formalism.
+%We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
+%This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
 
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{../code/analyze_neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 	\caption{Histogram}\label{fig:hist-real}
 \end{figure}
+
+\subsection{Setup}
+
+Everything is identical to Experiment 1, except that we replaced the real orderings with orderings generated from grammars fitted to the word orders found in the real language.
 
@@
 \paragraph{Results}
+\mhahn{to what extent do we have to present the large results figures separately for Experiment 1 and this experiment?
+or maybe we just provide some summary statistics separately, but provide the figures together.
+or all the big figures go into the SI?
+I don't mind, I just don't know what people in that audience would expect.}
+
+%
+%\subsection{Discussion: Alternative Models}
+%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
+%
+%A traditional model uses n-gram models.
+A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
+Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
+Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
+%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
+%
+%A second option is to construct a statistical grammar, such as PCFG.
+%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
+%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
+%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
+%
+
+%\subsection{Data}
+%\subsection{Setup}
+%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
+%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
+%This biases the hyperparameters towards favoring counterfactual grammars.
+
+%\subsection{Setup}
+
+%\paragraph{Data}
+%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
+%%
+%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
+%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
+%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
+%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
+%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
+%
+%%Such networks estimate the probability of a word in context as follows.
+%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
+%It then computes a sequence of hidden and cell states by the following recurrent computation:
+%\begin{align*}
+%	h_1, c_1 &:= 0 \\
+%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
+%	\dots \\
+%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
+%\end{align*}
+%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
+%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
+%
+%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
+%We obtain per-word predictions of the next word by computing
+%\begin{align*}
+%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
+%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
+%\end{align*}
+%where the softmax transformation normalizes vectors into probability distributions as follows
+%\begin{equation}
+%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
+%\end{equation}
+%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
+%\begin{equation}
+%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+%\end{equation}
+%and thus the surprisal is estimated as
+%\begin{equation}
+%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
+%\end{equation}
+%We discuss the choice of the numerical parameters in the next section.
+%
+
+%We collected data from the actual and random orderings in proportion one to two.
+%The stopping criterion will be described below.
+
+%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
+%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
+%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
+%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
+
+%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
+
+%- at least 30 samples from both baseline and real
+%
+%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
+
+%
+%(1) is bigram MI always greater in real languages?
+%
+%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
+
+%Training progresses in a series of parameter update steps, constructing updated parameters $\theta_0, \theta_1, \theta_2, \dots$.
+%In the $n$-th update step, we first randomly select a word sequence $w_1 ...
+w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
+%We then update the parameter vector:
+%\mhahn{maybe better to just say we use SGD}
+%\begin{equation}\label{eq:train}
+%	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
+%\end{equation}
+%where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.

### 2019-08-12 writeup/paper.tex
code

@@
 
-%
 \begin{abstract}
-
 Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
@@
 We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
 \end{abstract}
-
-%
-%
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
-%
 
@@
 \section{Information Locality}\label{sec:info-locality}
-
+\mhahn{not sure about this section}
 TODO derive information locality
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{../code/analyze_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 \caption{Histogram}\label{fig:hist-real}

### 2019-08-12 writeup/tables/medians_ngrams_0.tex
code

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2019-08-12 writeup/tables/medians_ngrams_1.tex
code

@@
-Hindi & Hungarian & Japanese & Norwegian
+English & Erzya & Estonian & Faroese
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Hindi-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hungarian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Japanese-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Norwegian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
  \\ 
-Polish & Romanian & Slovak & Slovenian
+Finnish & French & German & Greek
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Polish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Romanian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovak-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovenian-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
  \\ 
-Spanish & Swedish &  & 
+Hebrew & Hindi & Hungarian & Indonesian
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Spanish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Swedish-listener-surprisal-memory-MEDIANS_QUANTILES_onlyWordForms_boundedVocab.pdf} &  &
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
  \\ 
+Italian & Japanese & Kazakh & Korean
+ \\ 
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+ \\

### 2019-08-12 writeup/tables/medians_ngrams_2.tex
code

@@
-Persian & Polish & Romanian & Serbian
+Kurmanji & Latvian & Maltese & Naija
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+ \\ 
+North Sami & Norwegian & Persian & Polish
+ \\ 
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+ \\ 
+Portuguese & Romanian & Russian & Serbian
+ \\ 
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
  \\
-Thai & Turkish & Ukrainian & Uyghur
- \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
- \\ 
-Vietnamese &  &  & 
- \\ 
-\includegraphics[width=0.25\textwidth]{ngrams/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} &  &  & 
- \\

### 2019-08-12 writeup/tables/medians_ngrams_3.tex
code

Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.25\textwidth]{../code/analyze_ngrams/visualize/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} &  & 
 \\

### 2019-08-12 writeup/tables/quantiles_REAL_NGRAMS_0.tex
code

Afrikaans & Amharic & Arabic & Armenian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Bambara & Basque & Breton & Bulgarian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Buryat & Cantonese & Catalan & Chinese
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Croatian & Czech & Danish & Dutch
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
English & Erzya & Estonian & Faroese
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\

### 2019-08-12 writeup/tables/quantiles_REAL_NGRAMS_1.tex
code

Finnish & French & German & Greek
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Hebrew & Hindi & Hungarian & Indonesian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Italian & Japanese & Kazakh & Korean
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\

### 2019-08-12 writeup/tables/quantiles_REAL_NGRAMS_2.tex
code

Portuguese & Romanian & Russian & Serbian
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Slovak & Slovenian & Spanish & Swedish
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Thai & Turkish & Ukrainian & Urdu
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}
 \\ 
Uyghur & Vietnamese &  & 
 \\ 
\includegraphics[width=0.25\textwidth]{ngrams/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} & \includegraphics[width=0.25\textwidth]{ngrams/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf} &  & 
 \\

### 2019-08-12 writeup/toy-experiment.tex
code

@@
-We illustrate the linguistic predictions of Proposition \ref{prop:lower-bound} by reanalyzing the data from \cite{fedzechkina-human-2017}.
+We illustrate the linguistic predictions of our theoretical results on the Memory-Surprisal tradeoff by reanalyzing the data from \cite{fedzechkina-human-2017}.
 This is a miniature artificial language study that showed a bias for Dependency Length Minimization in production in artificial language learning.
-Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in~(\ref{eq:memory-bound}).
+Due to the controlled setting, it is possible to exactly compute the speaker's memory as given in Theorem~\ref{prop:suboptimal}.
 
-As ~(\ref{eq:memory-bound}) is invariant under reversal of the language, we only consider the head-final version of her artificial language.
+As the bound in Theorem~\ref{prop:suboptimal} is invariant under reversal of the language, we only consider the head-final version of her artificial language.
 The language has consistent head-final order, and uses case marking on objects.

### 2019-08-12 code/analyze_ngrams/visualize/plot_medians.R

@@ -1,13 +1,13 @@
 
 # Plots medians with confidence intervals
 
-fullData = read.csv("../results/tradeoff/ngrams/listener-curve-ci-median.tsv", sep="\t")
+fullData = read.csv("../../../results/tradeoff/ngrams/listener-curve-ci-median.tsv", sep="\t")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     library(tidyr)
     library(dplyr)
     library(ggplot2)
-    dataL = read.csv(paste("../results/raw/ngrams/",language,"_ngrams_decay_after_tuning.tsv", sep=""), sep="\t")
+    dataL = read.csv(paste("../../../results/raw/ngrams/",language,"_ngrams_decay_after_tuning.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
     plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
@@ -17,62 +17,12 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     return(plot)
 }
 
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Arabic")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Catalan")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Czech")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Dutch")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Finnish")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("French")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("German")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Hindi")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Norwegian")
-plot = memListenerSurpPlot_onlyWordForms_boundedVocab("Spanish")
[…]

### 2019-08-12 code/analyze_ngrams/visualize/plot_wordsOnly_quantile_onlyREAL_Ngrams.R

@@ -1,0 +1,40 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+
+# Plots CIs for the quantile
+
+fullData_ConfidenceLowerBound = read.csv("../../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption-Ngrams.tsv", sep="\t") %>% filter(Type != "GROUND")
+#fullData_ConfidenceLowerBound_05 = read.csv("../../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-05-Ngrams.tsv", sep="\t") %>% filter(Type != "GROUND")
+
+fullData_BinomialTest = read.csv("../../../results/tradeoff/listener-curve-binomial-test-ngrams.tsv", sep="\t") %>% filter(Type != "GROUND")
+
+memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
+    data = fullData_ConfidenceLowerBound %>% filter(Language == language)
+    data2 = fullData_BinomialTest %>% filter(Language == language)
+#    data3 = fullData_ConfidenceLowerBound_05 %>% filter(Language == language)
+    #data = merge(data, data2, by=c("Language", "Position", "Type"))
+    #data$Memory = data$Memory.x
+    plot = ggplot(data, aes(x=Memory, y=LowerConfidenceBound, fill=Type, color=Type))
+    plot = plot + geom_line(size=1, linetype="dotted") 
+ #   plot = plot + geom_line(data=data3, size=2, linetype="dashed") 
+    plot = plot + geom_line(data=data2, aes(x=Memory, y=BetterEmpirical), size=2)
+    data2 = data2 %>% mutate(pValue_print = ifelse(round(pValue,5) == 0, "p<0.00001", paste("p=", round(pValue,5), sep="")))
+    plot = plot + geom_text(data=data2 %>% filter(Position %% 9 == 0, Type == "REAL_REAL"), aes(x=Memory, y=BetterEmpirical+0.1, label=pValue_print), size=3)
+    plot = plot + geom_text(data=data2 %>% filter(Position %% 9 == 0, Type == "GROUND"), aes(x=Memory, y=BetterEmpirical+0.05, label=pValue_print), size=3)
+    plot = plot + theme_classic() 
+    plot = plot + theme(legend.position="none") 
+    plot = plot + ylim(0,1.1)
+    plot = plot + ylab("Quantile")
[…]

### 2019-08-12 code/medians/interpolate/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Interpolate_Ngrams_All.py

@@ -5,7 +5,7 @@ import subprocess
 
 # ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
 
-with open("../../results/tradeoff/ngrams/listener-curve-interpolated.tsv", "w") as outFile:
+with open("../../../results/tradeoff/ngrams/listener-curve-interpolated.tsv", "w") as outFile:
   print >> outFile, "\t".join(["Language", "Type", "FileName", "Position", "Memory", "Surprisal"])
   for language in languages:
      print(language)

### 2019-08-12 code/medians/median-CI/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_Ngrams.py

@@ -22,7 +22,7 @@ def readTSV(x):
           data[i][column] = vals[i]
     return (header, data)
 try:
-  with open("../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
+  with open("../../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
      data = readTSV(inFile)
 except IOError:
   print >> sys.stderr, ("ERROR nothing for this language? "+language)

### 2019-08-12 code/medians/median-CI/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_Ngrams_All.py

@@ -7,7 +7,7 @@ import subprocess
 
 # ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
 
-with open("../../results/tradeoff/ngrams/listener-curve-ci-median.tsv", "w") as outFile:
+with open("../../../results/tradeoff/ngrams/listener-curve-ci-median.tsv", "w") as outFile:
   print >> outFile, "\t".join(["Language", "Type", "Position", "Memory", "MedianEmpirical", "MedianLower", "MedianUpper", "Level"])
   for language in languages:
      print(language)

### 2019-08-12 code/pointwise-test/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_All_Ngrams.py

@@ -1,0 +1,13 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../results/tradeoff/listener-curve-binomial-test-ngrams.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "Position", "Memory", "BetterEmpirical", "pValue"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_Ngrams.py", language]).strip()
+
+

### 2019-08-12 code/pointwise-test/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_Ngrams.py

@@ -1,4 +1,4 @@
-# Better than yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MaxControl.py by modeling the median of REAL
+# This one is mentioned in the paper source
 
 import sys
 
@@ -21,13 +21,8 @@ def readTSV(x):
       for i in range(len(data)):
           data[i][column] = vals[i]
     return (header, data)
-try:
-  with open("../../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
+with open("../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
      data = readTSV(inFile)
-except IOError:
-  print >> sys.stderr, ("ERROR nothing for this language? "+language)
-  quit()
-
 #print(len(data))
 
 
@@ -50,26 +45,21 @@ for i in range(len(data[1])):
     if typ not in matrixByType:
         matrixByType[typ] = []
     matrixByType[typ].append(matrix[i])
-MAX_DISTANCE = -1
 for typ in matrixByType:
-   tensorized = torch.FloatTensor(matrixByType[typ])
-   MAX_DISTANCE = int(tensorized[:,1].max())
-   matrixByType[typ] = tensorized.view(-1, MAX_DISTANCE, 4)
+   matrixByType[typ] = torch.FloatTensor(matrixByType[typ]).view(-1, 19, 4)
    #print(matrixByType[typ][0])
    misByType[typ] = matrixByType[typ][:,:,2]
 #print(misByType["RANDOM_BY_TYPE"])
-if MAX_DISTANCE == -1:
-  print >> sys.stderr, ("ERROR nothing for this language? "+language)
-  quit()
+
 #data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
-distance = torch.FloatTensor(range(1,1+MAX_DISTANCE))
+distance = torch.FloatTensor(range(1,20))
 
 cumMIs = {}
 cumMems = {}
 cumInterpolated = {}
 maximalMemory = 0
 for typ, mis in misByType.iteritems():
-  mask = torch.FloatTensor([[1 if j <= i else 0 for j in range(MAX_DISTANCE)] for i in range(MAX_DISTANCE)])
+  mask = torch.FloatTensor([[1 if j <= i else 0 for j in range(19)] for i in range(19)])
[…]

### 2019-08-12 code/preparePaper/quantiles_REAL_Ngrams.py

@@ -1,6 +1,4 @@
 # ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
-import sys
-import os
 
 with open("../../ud_languages.txt", "r") as inFile:
    languages = inFile.read().strip().split("\n")
@@ -30,27 +28,18 @@ def h(frame, line, name):
     return line[frame[0][name]]
 
 
-#with open("tradeoff/effectSize.tsv", "r") as inFile:
-#   effectSize = readTSV(inFile)
-
-#with open("tradeoff/effectSize_diff.tsv", "r") as inFile:
-#   effectSize_diff = readTSV(inFile)
-
 
 
 entries = []
 
 for language in languages:
 #   line = languageKey[language]
-   components = [language.replace("_"," ").replace("-Adap", "")]
-   #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
-   if not os.path.isfile("../analyze_ngrams/visualize/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf"):
-      print("No data for "+language)
-      continue
-   components.append( "\includegraphics[width=0.25\\textwidth]{../code/analyze_ngrams/visualize/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}" )
-   entries.append(components)
 
-ROWS_PER_PART = 4
+   components = [language.replace("_"," ").replace("-Adap", "")]
+   components.append( "\includegraphics[width=0.25\\textwidth]{ngrams/figures/"+language+"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL_NGRAMS.pdf}" )
+   entries.append(components) #   print >> outFile, ("  &  ".join([str(x) for x in components]) + "  \\\\ [10.25ex] \\hline" )
+
+ROWS_PER_PART = 5
 COLUMNS = 4
 entries += [["",""] for x in range((COLUMNS-(len(entries)%COLUMNS))%COLUMNS)]
 
@@ -66,7 +55,7 @@ if True:
        outputRows.append( " \\\\ ")
 
 for part in range(len(entries)/(COLUMNS*ROWS_PER_PART)+1):
-  with open("../../writeup/tables/medians_ngrams_"+str(part)+".tex", "w") as outFile:
[…]

### 2019-08-12 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_All_Ngrams.py

@@ -1,0 +1,13 @@
+from ud_languages import languages
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../results/tradeoff/listener-curve-binomial-confidence-bound-quantile-noAssumption-Ngrams.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "Position", "LowerConfidenceBound", "Level", "Memory"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["./python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_Ngrams.py", language]).strip()
+
+

### 2019-08-12 code/quantiles/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption_Ngrams.py

@@ -21,13 +21,8 @@ def readTSV(x):
       for i in range(len(data)):
           data[i][column] = vals[i]
     return (header, data)
-try:
-  with open("../../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
+with open("../../results/raw/ngrams/"+language+"_ngrams_decay_after_tuning.tsv", "r") as inFile:
      data = readTSV(inFile)
-except IOError:
-  print >> sys.stderr, ("ERROR nothing for this language? "+language)
-  quit()
-
 #print(len(data))
 
 
@@ -50,26 +45,21 @@ for i in range(len(data[1])):
     if typ not in matrixByType:
         matrixByType[typ] = []
     matrixByType[typ].append(matrix[i])
-MAX_DISTANCE = -1
 for typ in matrixByType:
-   tensorized = torch.FloatTensor(matrixByType[typ])
-   MAX_DISTANCE = int(tensorized[:,1].max())
-   matrixByType[typ] = tensorized.view(-1, MAX_DISTANCE, 4)
+   matrixByType[typ] = torch.FloatTensor(matrixByType[typ]).view(-1, 19, 4)
    #print(matrixByType[typ][0])
    misByType[typ] = matrixByType[typ][:,:,2]
 #print(misByType["RANDOM_BY_TYPE"])
-if MAX_DISTANCE == -1:
-  print >> sys.stderr, ("ERROR nothing for this language? "+language)
-  quit()
+
 #data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
-distance = torch.FloatTensor(range(1,1+MAX_DISTANCE))
+distance = torch.FloatTensor(range(1,20))
 
 cumMIs = {}
 cumMems = {}
 cumInterpolated = {}
 maximalMemory = 0
 for typ, mis in misByType.iteritems():
-  mask = torch.FloatTensor([[1 if j <= i else 0 for j in range(MAX_DISTANCE)] for i in range(MAX_DISTANCE)])
+  mask = torch.FloatTensor([[1 if j <= i else 0 for j in range(19)] for i in range(19)])
   cumulativeMI = torch.matmul(mis, mask.t())
   #print("MIs", mis[0])
   #print("Cum MI", cumulativeMI[0])
@@ -119,11 +109,61 @@ for typ, mis in misByType.iteritems():
[…]

### 2019-08-25 writeup/background.tex
update intro and background

@@
 A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
+This strand of research has brought together linguistic typologists documenting language universals, corpus linguists studying production preferences, psycholinguistics studying processing difficulty, and computer scientists studying formal properties of grammars.
+
+An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory; following work has studied how different grammars induce different memory requirements in terms of the number of symbols that must be stored at each point to produce or parse a sentence \citet{yngve1960model,yngve1961depth,abney1991memory,gibson1991computational,resnik1992left}.
+
 \cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
@@
 We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
+
+% Production argument
+% Idea: A producer wants to approximate a language conditional on a production target G.
+% Ie producer finds m_t to minimize loss: D[ w_t | w_{<t}, g   ||   w_t  | m_t  ] + a * H[m_t]
+% That is, the memory has to contain both the previous words and the current production goal.
+% The loss comes out to I[ w_t : w_{<t},g | m_t] + a * H[m_t]
+%                       = I[w_t : w_{<t} | m_t] + I[w_t : g | w_{<t}, m_t] + a * H[m_t]
+% Now let's decompose the memory m_t into two parts:
+% the part about the previous words: call this r_t
+% the part about the current goal: call this g_t. And let's assume H[g|g_t]=0, i.e. the memory stores
+% all the information about the goal.
+% Then H[m_t] = H[r_t] + H[g_t|r_t].
+% Now we have loss:
+%     I[w_t : w_{<t} | r_t, g] + I[w_t : g | w_{<t}, r_t, g] + a H[r_t, g_t]
+%   = I[w_t : w_{<t} | r_t, g] + a H[r_t, g_t] 
+% Now we have H[r_t, g_t] >= H[r_t] >= \sum t I_t.
+% And then we have
 

### 2019-08-25 writeup/introduction.tex
update intro and background

%Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
%A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
Why is human language the way it is? 
In this paper, we argue that the grammars of human languages can be modeled mathematically as solutions to the problem of maximally efficient communication among agents subject to particular resource constraints on information processing during speaking and listening.
We formalize these constraints using information theory.
Using these concepts, we show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension, and
we show that our formalization allows us to explain some of the universal syntactic properties of languages.

The suggestion that the structure of human language reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\citep{yngve1961,berwick1984grammatical,hawkins1994performance,jaeger2011language,gibson2019efficiency}, and has been summed up in \citeauthor{hawkins1994efficiency}'s (\citeyear{hawkins1994efficiency}) \key{Performance--Grammar Correspondence Hypothesis} (PGCH), which holds that grammars are structured so that the typical utterance is easy to produce and comprehend under performance constraints.
The idea has also been controversial, with prominent claims that efficiency in this sense has had no or only a minimal effect on the structure of human language \citep{chomsky}.

Among theories of language based on processing efficiency, some of the most successful have been those dealing with resource constraints on working memory as deployed during language production and comprehension.
An extensive body of work from psycholinguistics has documented that limitations in incremental memory cause measurable difficulty in production and comprehension \citep{gibson1998syntactic,gibson1999memory,gibson2000dependency,lewis2005activationbased,bartek2011search,nicenboim2015working}. % TODO are there cites for production?
Complementing this work, a number of researchers in linguistic typology and corpus linguistics have shown how various universal properties of grammars can be explained in terms of memory-based processing difficulty \citep{hawkins2004efficiency,hawkins2014crosslinguistic,ferrericancho2006syntactic,gildea2010grammars,futrell2015largescale,liu2017dependency,temperley2018dependency}.
We will see that one of the most successful theories in this domain, the principle of \key{dependency length minimization}, falls out of our formalization of processing efficiency as a special case.
%Our work formalizes and generalizes this previous work.

We formalize the notion of memory constraints in language processing in terms of what we call the \key{memory--surprisal tradeoff}: the idea that it is possible to achieve greater ease of word-by-word comprehension, and greater accuracy in word-by-word production, at the cost of investing more computational resources into remembering previous words.
The shape of this tradeoff depends on the grammar of a language, and in particular its word order properties.
We characterize memory resources in a theory-neutral, information-theoretic manner based on rate--distortion theory \citep{cover2006elements} and the theory of statistical complexity \citep{crutchfield1989inferring}. % cite the crutchfield nature review, and predictive info bottleneck

%For each amount of memory invested, a language permits a certain level of ease of processing 
%The resulting tradeoff of memory and surprisal of depends on the word order properties of a language. Analogous results also hold for language production by resource-constrained speakers. We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.

The remainder of this paper is structured as follows. In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing. In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}. In Section~\ref{sec:info-locality}, we prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other. In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}. In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders. Section~\ref{sec:conclusion} concludes.

%These strategies can be seen as sets of rules that differ in the way ...
%Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
%Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
%We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.

%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.

%\mhahn{original intro}
%Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
%Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
%Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
%Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.

%At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
%For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.

### 2019-09-03 writeup/control-ngrams.tex

@@
 
-We use Interpolated Kneser-Ney.
-Hyperparameters are tuned with the same strategy as for the neural network models.
+We use a version of Kneser-Ney Smoothing.
+For a sequence $w_1\dots w_k$, $N(w_{1\dots k})$ is the number of times $w_{1\dots k}$ occurs in the training set.
+The unigram probabilities are estimated as
+\begin{equation}
+	p_1(w_t) :=   \frac{N(w_t) + \delta}{|Train| + |V| \cdot \delta}
+\end{equation}
+where $\delta \in \mathbb{R}_+$ is a hyperparameter.
+Here $|Train|$ is the number of tokens in the training set, $|V|$ is the number of types occurring in train or held-out data.
+Higher-order probabilities $p_t(w_t|w_{0 \dots t-1})$ are estimated recursively as follows.
+Let $\gamma > 0$ be a hyperparameter.
+If $N(w_{0 \dots t-1}) < \gamma$, set $p_t(w_t|w_{0 \dots t-1}) := p_{t-1}(w_t|w_{1\dots t-1})$.
+Otherwise, we interpolate between $n$-th order and lower-order estimates:
+\begin{equation}
+	p_t(w_t|w_{0 \dots t-1}) :=  \frac{\operatorname{max}(N(w_{0\dots t}) - \alpha, 0.0) + \alpha \cdot \#\{w : N(w_{0 \dots t-1}w) > 0\} \cdot p_{t-1}(w_t|w_{1\dots t-1})}{N(w_{0\dots t-1})}
+\end{equation}
+where $\alpha \in [0,1]$ is also a hyperparameter.
+
+Hyperparameters $\alpha, \gamma, \delta$ are tuned with the same strategy as for the neural network models.
 

### 2019-10-10 writeup/formal-analysis.tex
code

@@
 \paragraph{Language as a Stochastic Process}
-We represent language as a stochastic process:
-We formalize a language as a probabilistic sequence of words $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $x_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+We represent language as a stochastic process of words $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
+The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
 
-We model the sequence as a probabilistic sequence; that is, given a context $w_{<t}$, the next word is distributed according to a distribution $p(w_t|w_{<t})$.
+%We model the sequence as a probabilistic sequence; that is, given a context $w_{<t}$, the next word is distributed according to a distribution $p(w_t|w_{<t})$.
 
@@
 As the utterance unfolds, the listener maintains a memory state.
-After receiving word $w_t$, the listener computes their new memory state $L_t$ based on the previous memory state $L_{t-1}$ and the new word $w_t$.}\label{fig:listener-markov}
+After receiving word $w_t$, the listener computes their new memory state $m_t$ based on the previous memory state $m_{t-1}$ and the new word $w_t$.}\label{fig:listener-markov}
 \end{figure}
@@
 We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $L_t$.
+As the speaker's utterance unfolds, the listener maintains a memory state $m_t$.
 
@@
 We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $L_t$ is determined by the last word $w_t$, and the prior memory state $L_{t-1}$.
-As a consequence, $L_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $L_{t-1}$.
+At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$.
+As a consequence, $m_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $m_{t-1}$.
 This is formalized as a statement about conditional probabilities:
 	\begin{equation}\label{eq:listener-markov}
-		p(L_1| (w_{t})_{t \in \mathbb{Z}}, L_0)   = p((w_{t})_{t \in \mathbb{Z}} | L_0, w_1)
-%p((w_{t})_t| L_0, L_1, w_1)   = p((w_{t})_t| L_0, w_1)
+		p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
 	\end{equation}
-This says that $L_1$ contains no information about the utterances beyond what is contained in $L_0$ and $w_1$.	
+This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
 As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
 This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-For the study of memory in sentence processing, this seems fair. Discuss this more.
+\mhahn{For the study of memory in sentence processing, this seems fair. Discuss this more.}
 
@@
 %\begin{equation}\label{eq:listener-markov-1}
-%L_t \bot w_{>t} | w_{\leq t}
+%m_t \bot w_{>t} | w_{\leq t}
 %\end{equation}
@@
 %
-%Second, we assume that $L_t$ contains no information about the past beyond what is contained in $w_{t-1}$ and $L_{t-1}$:
+%Second, we assume that $m_t$ contains no information about the past beyond what is contained in $w_{t-1}$ and $m_{t-1}$:
 %\begin{equation}\label{eq:listener-markov-2}
-%L_t \bot w_{<t} | w_{t-1}, L_{t-1}
+%m_t \bot w_{<t} | w_{t-1}, m_{t-1}
 %\end{equation}
-%This means that any information about the past in $L_t$ has to be contained in $L_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
+%This means that any information about the past in $m_t$ has to be contained in $m_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
 
@@
 %A listener who chooses to store less memory will exerience higher surprisal in the future.
-%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $L_t$ contains all predictive information about the future that is contained in the past.
+%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $m_t$ contains all predictive information about the future that is contained in the past.
 
@@
 
-%\begin{proof}
-	We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $L_t$.
-	The average number of bits required to encode this state is $\operatorname{H}[L_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-	As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | L_t]$.
-	The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]$.
-By the assumption of stationarity, we can rewrite this expression as
-\begin{align*}
-	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}] &=  \frac{1}{T} \sum_{t=1}^{T} \left(\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]\right) 
-\end{align*}
+We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
+The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
+As described above, we assume
+\begin{equation}\label{eq:listener-markov}
+	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
+\end{equation}
+that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.
+As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+
+The average number of bits required to encode this state is $\operatorname{H}[m_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
+As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | m_t]$.
+The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.
+By the assumption of stationarity, we can, for any positive integer $T$, rewrite this expression as
+\begin{equation}\label{eq:byStation}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}] =  \frac{1}{T} \sum_{t'=1}^{T} \left(\operatorname{H}[w_{t'} | m_{t'}] - \operatorname{H}[w_{t'} | w_{<t'}]\right) 
+\end{equation}
 \begin{lemma}
-For any $t$, the following inequality holds:
-%	$$H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]$$
+For any positive integer $t$, the following inequality holds:
 \begin{equation}
-%H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]
-H[w_t | L_t] \geq H[w_t|w_{1 \dots t-1}, L_0]
+H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
 \end{equation}
@@
 \begin{proof}[Proof of the Lemma]
-
-By Equation~\ref{eq:listener-markov}:
+	By Bayes' Theorem
+\begin{align*}
+	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
+\begin{align*}
+p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+So we have a Markov chain
 \begin{equation}
-p((w_{t})_t| L_0, L_1, w_1)   = p((w_{t})_t| L_0, w_1)
-	\end{equation}
-we have
-	\begin{align*}
-p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) = p(w_t|w_{<1}, L_0,  w_1, w_{2 \dots t-1}) \\
-p(w_{<1}| L_0, L_1, w_1, w_{2 \dots t-1})   = p(w_{<1}| L_0, w_1, w_{2 \dots t-1})
-	\end{align*}
-Therefore
-\begin{align*}
-p(w_t|L_0, w_1, w_{2 \dots t-1}; L_1, w_{2 \dots t-1}) & = p(w_t|L_0, L_1, w_1, w_{2 \dots t-1}) \\
-& = \sum_{w_{<1}} p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) p(w_{<1}| L_0, L_1, w_1, w_{2 \dots t-1}) \\
-&= \sum_{w_{<1}} p(w_t|w_{<1}, L_0,  w_1, w_{2 \dots t-1}) p(w_{<1}| L_0, w_1, w_{2 \dots t-1}) \\
-&= p(w_t|L_0,  w_1, w_{2 \dots t-1}) 
-\end{align*}
-Thus we have a Markov chain
-\begin{equation}
-(w_t) \rightarrow (L_0, w_1, w_{2 \dots t-1})   \rightarrow   (L_1, w_{2 \dots t-1})
+(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
 \end{equation}
@@
 \begin{equation}
-H[w_t| w_{2 \dots t-1}, L_{1}] \geq H[w_t|w_{1 \dots t-1}, L_0]
+H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
 \end{equation}
-
 Finally, iteratively applying this inequality, we get:
 \begin{align*}
-		H[w_t | L_t] \geq H[w_t| w_{t-1}, L_{t-1}] \geq H[w_t| w_{t-2, t-1}, L_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, L_0]
+H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
 \end{align*}
-
 \end{proof}
-
-	% = \int dw_{<1} p(w_t|w_{<1}, w_1, w_{2 \dots t-1})  = \int dw_{<1} p(w_t|L_0, w_{<1}, w_1, w_{2 \dots t-1}) = p(w_t|L_0, w_1, w_{2 \dots t-1}) $
-
-%	$p(w_t|L_0, w_1, w_{2 \dots t-1}; L_1, w_{2 \dots t-1}) = p(w_t|L_0, L_1, w_1, w_{2 \dots t-1}) = \int dw_{<1} p(w_t|w_{<1}, L_0, L_1, w_1, w_{2 \dots t-1}) = \int dw_{<1} p(w_t|w_{<1}, w_1, w_{2 \dots t-1})  = \int dw_{<1} p(w_t|L_0, w_{<1}, w_1, w_{2 \dots t-1}) = p(w_t|L_0, w_1, w_{2 \dots t-1}) $
-
-	Plugging this inequality into the equality above:
+Plugging this inequality into Equation~\ref{eq:byStation} above:
 \begin{align*}
-	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, L_0] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
-	& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | L_0] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
-	& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|L_0]\right) 
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
+& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
+& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|m_1]\right) 
 \end{align*}
@@
 \begin{align*}
-	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|L_0]\right) 
-	%	&= \frac{1}{T} \left(\sum_{t=1}^\infty t I_t - \sum_{t > T}  (T-t) I_t- I[w_{1\dots T}|L_0]\right) \\
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|m_1]\right) 
 \end{align*}
-	$I[w_{1\dots T}|L_0]$ is at most $\operatorname{H}[L_0]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
+$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
 \begin{align*}
-	\operatorname{H}[w_t | L_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
 &= \sum_{t > T} I_t
 \end{align*}
-	Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|L_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
+Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
 %\end{proof}
 
-Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+%Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.

### 2020-01-14 writeup/main-experiment.tex
code

@@
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
+We now investigate whether word orders as found in natural language reflect optimizization for the memory-surprisal tradeoff.
 We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
@@
 
-\subsection{Counterfactual Ordering Grammars}
+\subsection{Baselines: Counterfactual Ordering Grammars}
 We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
@@
 \mhahn{One thing that has to be discussed: the absolute values differ between languages}
-
-\section{Experiment 2: Free Word Orders}
-
-Here we match the branching entropies of the baseline grammars with those of the real language.
-
-\section{Experiment 3: Fixed Word Orders}
-
-A possible concern is that the optimization observed in Experiment 1 might be an artifact of the grammar formalism.
-To address this possibility, we repeated the experiment comparing baseline grammars to grammars that represent the word order of the real languages as faithfully as is possible with the grammar formalism.
-%We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-%This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\subsection{Setup}
-
-Everything is identical to Experiment 1, except that we replaced the real orderings with orderings generated from grammars fitted to the word orders found in the real language.
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-\mhahn{to what extent do we have to present the large results figures separately for Experiment 1 and this experiment?
-or maybe we just provide some summary statistics separately, but provide the figures together.
-or all the big figures go into the SI?
-I don't mind, I just don't know what people in that audience would expect.}
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-%\paragraph{Data}
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model 
-%%
-%\textbf{TODO I'm describing this in a lot of detail. Alternatively, we can say this is a standard NLP method and refer to the NLP literature for the definition.}
-%The first component of such a model is an \emph{embedding matrix} $W_{emb} \in \mathbb{R}^{|V| \times d_{emb}}$, where the \emph{vocabulary} $\mathcal{V}$ is a set, containing the words that occur in the corpus, and $d_{emb} \in \mathbb{N}$ is a fixed parameter.
-%This matrix assigns a $d_{emb}$-dimensional vector to each word occurring in the corpus.
-%The second component is an LSTM cell $f_{LSTM}$, a nonlinear transformation mapping an \emph{input} vector $x_{i} \in \mathbb{R}^{d_{emb}}$ a \emph{hidden state} $h_i \in \mathbb{R}^{d_{LSTM}}$ and a \emph{cell state} $c_i \in \mathbb{R}^{d_{LSTM}}$ to a new pair of hidden state and cell states $h_{i+1}, c_{i+1} \in \mathbb{R}^{d_{LSTM}}$.
-%The LSTM cell $f_{LSTM}$ is parameterized by a matrix of numerical parameters $W_{LSTM}$.
-%
-%%Such networks estimate the probability of a word in context as follows.
-%Given a sequence of input words $w_1, ..., w_n \in V$, the model first retrieves fixed-dimensionality vector representations $x_1, ..., x_n$, where $x_i$ is the row of $W_{emb}$ corresponding to the word $w_i$.
-%It then computes a sequence of hidden and cell states by the following recurrent computation:
-%\begin{align*}
-%	h_1, c_1 &:= 0 \\
-%	h_2, c_2 &:= f_{LSTM}(x_1, h_1, c_1) \\
-%	\dots \\
-%	h_{n+1}, c_{n+1} &:= f_{LSTM}(x_n, h_n, c_n) \\
-%\end{align*}
-%The vector $h_i$ encodes the result of reading the words $w_1, ..., w_{i-1}$.
-%We will write $LSTM(w_1, ..., w_{i-1})$ for $h_i$.
-%
-%The third component of the recurrent language model is the matrix $W_{output} \in \mathbb{R}^{|V| \times d_{LSTM}}$.
-%We obtain per-word predictions of the next word by computing
-%\begin{align*}
-%	s_i := W_{output} h_i \in \mathbb{R}^{|V|} \\
-%	p_i := \operatorname{softmax}(s_i)\in \mathbb{R}^{|V|} 
-%\end{align*}
-%where the softmax transformation normalizes vectors into probability distributions as follows
-%\begin{equation}
-%	\operatorname{softmax}(x)_i := \frac{\exp(x_i)}{\sum_{j=1}^{|V|} \exp(x_j)}
-%\end{equation}
-%Finally, the probability of the word $w_n$ in the context $w_1, ..., w_{n-1}$ is computed as
-%\begin{equation}
-%	p_\theta(w_n|w_1...w_{n-1}) := \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%and thus the surprisal is estimated as
-%\begin{equation}
-%- \log	p_\theta(w_n|w_1...w_{n-1}) := -\log \frac{\exp((p_n)_{w_n})}{\sum_{w \in V} \exp(x_w)}
-%\end{equation}
-%We discuss the choice of the numerical parameters in the next section.
-%
-
-%We collected data from the actual and random orderings in proportion one to two.
-%The stopping criterion will be described below.
-
-%Due to the randomness both in the sequence of training examples and the random initialization of the network weights, the results of the parameter estimation procedure will vary when run multiple times, especially on smaller datasets.
-%Informally, due to the finiteness of the dataset, multiple parameter settings are compatible with the available training data.
-%Consequently, memory-surprisal tradeoffs estimated on held-out sets will also show some variation.
-%Therefore, we collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-
-%- at least 30 samples from both baseline and real
-%
-%- for the language-level tradeoff curve, either the fraction is zero or the bootstrapped CI has width $\leq 0.2$.
-
-%
-%(1) is bigram MI always greater in real languages?
-%
-%(2) is the tradeoff curve always lower than for deterministic simple grammar? for deterministic complex grammars? for stochastic simple/complex grammars?
-
-%Training progresses in a series of parameter update steps, constructing updated parameters $\theta_0, \theta_1, \theta_2, \dots$.
-%In the $n$-th update step, we first randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-%We then update the parameter vector:
-%\mhahn{maybe better to just say we use SGD}
-%\begin{equation}\label{eq:train}
-%	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-%\end{equation}
-%where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.

### 2020-01-14 writeup/si.tex
code

--- 2019-08-12 writeup/paper.tex
@@
 
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
+\title{Supplementary Information for: Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
 \author{Michael Hahn, Judith Degen, Richard Futrell}
@@
 
-\begin{abstract}
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-\end{abstract}
-
-\section{Introduction}
-
-\input{introduction.tex}
-
-\section{Background}\label{sec:background}
-% More detailed lit review
-
-\input{background.tex}
-
-\section{Memory-Surprisal Tradeoff}\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-\mhahn{not sure about this section}
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-\input{main-experiment.tex}
-
-\section{Discussion}
-
-\input{discussion.tex}
-
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
+\begin{figure}[!htbp]
 \includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{longtable}{l|ll||l|llllllllllllll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{longtable}{l|lll||l|lllllllllllllll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{longtable}{ccccccccccccccclll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{tabular}{ccccccccccccccclll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{tabular}{l|ll||l|llllllllllllll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{tabular}{ccccccccccccccccll}
@@
 
-\begin{table}
+\begin{table}[!htbp]
 \begin{tabular}{cccccccccccccccccc}
@@
 
+\bibliographystyle{apalike}
+\bibliography{literature}
+
+\appendix
+
 \end{document}

### 2020-01-14 writeup/tradeoff.tex
code

@@
 
+Assumptions:
+
+The memory representations at time $t+1$ are entirely determined by the memory representations at time $t$ and the word $w_t$.
+
 The theorem allows us to estimate the extra surprisal associated with each amount of memory capacity for a language.
@@
 It provides a \emph{lower bound} on the amount of information that needs to be stored -- other memory representations will always need to store at least as much information.
+
+There are theories of retrieval where the main bottleneck lies not in capacity, but in difficulty of retrieval. See General Discussion.
 
@@
 %\subsection{Formal Analysis and Proofs}
+
+\subsection{Information Locality}\label{sec:info-locality}
+
+Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
+This formalizes a general, assumption-free, link between memory and locality in language production.
+In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
+%incremental prediction.
+
+%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
+The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
+In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
+%This confirms the intuition that finiteness of memory entails that the contribution of long-term
+
+%\paragraph{Discussion}
+%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
+%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
+
+We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
+We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
+The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
+In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
+However, $I_t$ decays more quickly for Process A (red).
+This means that predictive information about an observation is concentrated more strongly in the recent past.
+In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
+Note that the area under the curve is equal to (\ref{eq:memory-bound}).
+This area is smaller for the red process, as $I_t$ decays more quickly there.

### 2020-01-14 writeup/si.tex
code

@@
 \usepackage{amssymb}
+
+\usepackage{caption}
 
@@
 
-\begin{figure}[!htbp]
+\begin{center}
 \includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
+\captionof{figure}{Histogram}\label{fig:hist-real}
+\end{center}
+
+\section{Formal Analysis and Proofs}
+
+In this section, we prove the theorem described above.
+
+\subsection{Mathematical Assumptions}
+
+We first make explicit how we formalize language processing for proving the theorem.
+
+\paragraph{Ingredient 1: Language as a Stationary Stochastic Process}
+We represent language as a stochastic process of words $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
+The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+
+%We model the sequence as a probabilistic sequence; that is, given a context $w_{<t}$, the next word is distributed according to a distribution $p(w_t|w_{<t})$.
+
+The assumption of infinite length is for mathematical convenience and does not affect the substance of our results:
+As we restrict our attention to the processing of individual sentences, which have finite length, we will actually not make use of long-range and infinite contexts.
+
+We make the assumption that this process is \emph{stationary}.
+Formally, this means that the conditional distribution $P(w_t|w_{<t})$ does not depend on $t$, it only depends on the actual sequence $w_{<t}$.
+Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
+In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
+Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
+
+\begin{figure}
+\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
+	\caption{Illustration of (\ref{eq:listener-markov}).
+As the utterance unfolds, the listener maintains a memory state.
+After receiving word $w_t$, the listener computes their new memory state $m_t$ based on the previous memory state $m_{t-1}$ and the new word $w_t$.}\label{fig:listener-markov}
 \end{figure}
 
-\begin{table}[!htbp]
+\paragraph{Ingredient 2: Flow of Information}
+We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
+As the speaker's utterance unfolds, the listener maintains a memory state $m_t$.
+
+There are no assumptions about the memory architecture and the nature of its computations.
+We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
+At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$.
+As a consequence, $m_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $m_{t-1}$.
+This is formalized as a statement about conditional probabilities:
+	\begin{equation}\label{eq:listener-markov}
+		p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
+	\end{equation}
+This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
+As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
+\mhahn{For the study of memory in sentence processing, this seems fair. Discuss this more.}
+
+%
+%First, we assume that the listener's internal state cannot depend on the future beyond its dependency on the past.
+%Formally: 
+%\begin{equation}\label{eq:listener-markov-1}
+%m_t \bot w_{>t} | w_{\leq t}
+%\end{equation}
+%This means that the listener has no access to the speaker's state beyond what the speaker has already uttered.
+%
+%Second, we assume that $m_t$ contains no information about the past beyond what is contained in $w_{t-1}$ and $m_{t-1}$:
+%\begin{equation}\label{eq:listener-markov-2}
+%m_t \bot w_{<t} | w_{t-1}, m_{t-1}
+%\end{equation}
+%This means that any information about the past in $m_t$ has to be contained in $m_{t-1}$ -- formalizing the idea that a listener can only remember aspects of the past by keeping them in memory, and that memories of the past cannot `spontaneously' form later in the future.
+
+%The listener can trade off memory and future surprisal:
+%A listener who chooses to store less memory will exerience higher surprisal in the future.
+%A listener can achieve minimal surprisal -- that is, the lowest average surprisal that any model could achieve by predicting the future from the past -- if and only if $m_t$ contains all predictive information about the future that is contained in the past.
+
+%We now describe the memory-surprisal tradeoff. will describe this tradeoff, and show that listener memory is linked to locality in a way similar to speaker memory.
+%Consider a listener who uses $J$ bits of memory on average.
+%What can we say about the listener's surprisal?
+
+\subsection{Proof of the Theorem}
+
+We restate the theorem:
+
+\begin{thm}\label{prop:suboptimal}
+	Let $T$ be any positive integer ($T \in \{1, 2, 3, ...\}$), and consider a listener using at most
+	\begin{equation}\label{eq:memory}
+		\sum_{t=1}^T t I_t
+	\end{equation}
+bits of memory on average.
+Then this listener will incur surprisal at least
+	$$H[w_t|w_{<t}] + \sum_{t > T} I_t$$
+	on average.
+\end{thm}
+
+We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
+The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
+As described above, we assume
+\begin{equation}\label{eq:listener-markov}
+	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
+\end{equation}
+that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.
+As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+
+The average number of bits required to encode this state is $\operatorname{H}[m_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
+As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | m_t]$.
+The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.
+By the assumption of stationarity, we can, for any positive integer $T$, rewrite this expression as
+\begin{equation}\label{eq:byStation}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}] =  \frac{1}{T} \sum_{t'=1}^{T} \left(\operatorname{H}[w_{t'} | m_{t'}] - \operatorname{H}[w_{t'} | w_{<t'}]\right) 
+\end{equation}
+
+We first show a lemma:
+
+\begin{lemma}
+For any positive integer $t$, the following inequality holds:
+\begin{equation}
+H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
+\end{equation}
+\end{lemma}
+\begin{proof}[Proof of the Lemma]
+	By Bayes' Theorem
+\begin{align*}
+	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
+\begin{align*}
+p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+So we have a Markov chain
+\begin{equation}
+(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
+\end{equation}
+Thus, by the Data Processing Inequality,
+\begin{equation}
+H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
+\end{equation}
+Finally, iteratively applying this inequality, we get:
+\begin{align*}
+H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
+\end{align*}
+\end{proof}
+Plugging this inequality into Equation~\ref{eq:byStation} above:
+\begin{align*}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
+& = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
+& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|m_1]\right) 
+\end{align*}
+The first term $I[w_{1\dots T}|w_{\leq 0}]$ can be rewritten in terms of $I_t$:
+\begin{align*}
+I[w_{1\dots T}|w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
+\end{align*}
+Therefore
+\begin{align*}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|m_1]\right) 
+\end{align*}
+$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
+\begin{align*}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - \sum_{t=1}^T t I_t\right) \\
+&= \sum_{t > T} I_t
+\end{align*}
+Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
+%\end{proof}
+
+%Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+
+\subsection{Locality in a model with Memory Retrieval}
+
+Here we show that our information-theoretic analysis is compatible with models placing the main bottleneck in the difficulty of retrieval.
+We describe an information-theoretic formalization of a model that has both a limited WM and an unlimited STM.
+
+We consider a model that maintains both a small working memory $m_t$ and an unlimited short-term memory $s_t$.
+
+Predictions are made based on working memory $m_t$, incurring surprisal $H[w_t|m_t]$.
+In each time step, there is some amount of communication between $m_t$ and $s_t$, corresponding to retrieval operations.
+We model this using a variable $r_t$ representing the information that is retrieved from $s_t$.
+
+This is instantiated by ACT-R.
+We can explicitly explain how this covers the McElree ideas and the Lewis and Vasishth ACT-R model.
+
+This model has two bottlenecks:
+The working memory capacity, which we model as $H[m_t]$, and the amount of information that is added through retrieval, modeled as $H[r_t|m_t]$.
+
+%Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
+
+\begin{thm}
+	Assume the average working memory $H[m_t]$ is bounded as $H[L_t] \leq \sum_{t=1}^T t I_t$, and the average amount of retrieval is bounded as $H[R_t] \leq \sum_{t=T+1}^S I_t$ (per word).
+	Then $H[w_t|m_t] \geq H[w_t|x_{<t}] + \sum_{t>S} I_t$.
+\end{thm}
+
+\begin{proof}
+The negative surprisal gap is
+$\leq \frac{1}{T}(I[X_1\dots X_T|(M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t) \leq  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) = H[R_t] - \sum_{t>T} I_t$
+
+\end{proof}
+
+We can also think of this as a multi-objective optimization, aiming to minimize $\lambda_1 WM + \lambda_2 Retrieval$ to achieve a given surprisal level.
+
+If $\frac{\lambda_2}{\lambda_1} \rightarrow \infty$ (retrievals get more expensive), recover previous model.
+
+If $\frac{\lambda_2}{\lambda_1} \rightarrow 0$ (retrievals get cheaper), locality effect gets weaker, and disappears in the limit\footnote{(Of course, even in this limit, there might be additional factors that may still favor locality in a specific implementation of memory -- e.g., in ACT-R, decay and interference are less problematic if there is locality.)}
+
+Objective
+$$\min_{T} \lambda_1 \sum_{t=1}^T t I_t + \lambda_2 \sum_{t=T+1}^S I_t$$
+has solution $T \approx \frac{\lambda_2}{\lambda_1}$.\footnote{Can do simple proof using the continuous-$T$-version.}
+
+Consequence: As long as retrievals are more expensive than keeping the same amount in WM, locality is predicted.
+
+\section{Corpus Size per Language}
+
+\begin{center}
 \begin{longtable}{l|ll||l|llllllllllllll}
@@
 \end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
+	\captionof{table}{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
+\end{center}
 
-\begin{table}[!htbp]
+\section{Samples Drawn per Language}
+
+\begin{center}
 \begin{longtable}{l|ll||l|llllllllllllll}
@@
 \end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
+	\captionof{figure}{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{longtable}{l|lll||l|lllllllllllllll}
@@
 \end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
+	\captionof{figure}{Bootstrapped estimates for $G$.}\label{tab:boot-g}
+\end{center}
 
-\begin{table}[!htbp]
+\section{Detailed Results per Language}
+
+\subsection{Median Surprisal per Memory Budget}
+
+\begin{center}
 \begin{longtable}{ccccccccccccccclll}
@@
 \end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+	\captionof{figure}{Medians: For each memory budget, we provide the median surprisal for real and random languages.
 Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$).
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{longtable}{ccccccccccccccclll}
@@
 \end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
+	\captionof{figure}{Medians (cont.)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{longtable}{ccccccccccccccclll}
@@
 \end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
+	\captionof{figure}{Medians (cont.)}
+\end{center}
 
-\begin{table}[!htbp]
+\subsection{Surprisal at Maximum Memory}
+
+\begin{center}
 \begin{tabular}{ccccccccccccccclll}
@@
 \end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
+	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccclll}
@@
 \end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
+	\captionof{figure}{Medians (cont.)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccclll}
@@
 \end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
+	\captionof{figure}{Medians (cont.)}
+\end{center}
 
@@
 %\end{tabular}
-%	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+%	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
 Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-%\end{table}
+%\end{center}
 %
@@
 %\end{tabular}
-%	\caption{Quantiles (part 2)}
-%\end{table}
+%	\captionof{figure}{Quantiles (part 2)}
+%\end{center}
 %
@@
 %\end{tabular}
-%	\caption{Quantiles (part 3)}
-%\end{table}
+%	\captionof{figure}{Quantiles (part 3)}
+%\end{center}
 %
@@
 
-\begin{table}[!htbp]
+\subsection{Samples Drawn (Experiment 3)}
+
+\begin{center}
 \begin{tabular}{l|ll||l|llllllllllllll}
@@
 \end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
+	\captionof{figure}{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+\end{center}
 
-\begin{table}[!htbp]
+\subsection{Medians (Experiment 3)}
+
+\begin{center}
 \begin{tabular}{ccccccccccccccclll}
@@
 \end{tabular}
-	\caption{Experiment 3.
+	\captionof{figure}{Experiment 3.
 Medians: For each memory budget, we provide the median surprisal for real and random languages.
@@
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccclll}
@@
 \end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
+	\captionof{figure}{Medians (cont.)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccccll}
@@
 \end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
+	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
 Lower values indicate lower surprisal compared to baselines.
@@
 Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccccll}
@@
 \end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
+	\captionof{figure}{Median Differences (Part 2)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccccll}
@@
 \end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
+	\captionof{figure}{Median Differences (Part 3)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{ccccccccccccccccll}
@@
 \end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
+	\captionof{figure}{Median Differences (Part 4)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{cccccccccccccccccc}
@@
 \end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
 Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{cccccccccccccccccc}
@@
 \end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
+	\captionof{figure}{Quantiles (part 2)}
+\end{center}
 
-\begin{table}[!htbp]
+\begin{center}
 \begin{tabular}{cccccccccccccccccc}
@@
 \end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
+	\captionof{figure}{Quantiles (part 3)}
+\end{center}
+
+\section{Details for Neural Network Models}
 
@@
 
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
 \bibliographystyle{apalike}
@@
 
-\appendix
+%\appendix
 

### 2020-01-15 writeup/freedom-control.tex

--- 2019-08-05 paper.tex
@@
-\documentclass[11pt,letterpaper]{article}
+Here we match the branching entropies of the baseline grammars with those of the real language.
 
-\usepackage{showlabels}
-\usepackage{fullpage}
-\usepackage{pslatex}
-%\usepackage{latexsym}
-\usepackage[english]{babel}
-\usepackage[utf8]{inputenc}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{graphicx}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage{rotating}
-\usepackage{natbib}
-\usepackage{amssymb}
+A possible concern is that the optimization observed in Experiment 1 might be an artifact of the grammar formalism.
+To address this possibility, we repeated the experiment comparing baseline grammars to grammars that represent the word order of the real languages as faithfully as is possible with the grammar formalism.
+%We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
+%This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
 
-\usepackage{tikz-dependency}
-\usepackage{longtable}
+\begin{enumerate}
+\item Baselines with matched order freedom
+\item Represent real languages by approximations within the formalism
+\item Simulation taking information structure into account (for one language)
+\end{enumerate}
 
-\newcommand{\R}[0]{\mathbb{R}}
-\newcommand{\E}[0]{\mathbb{E}}
-\newcommand{\Ff}[0]{\mathcal{F}}
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+	\caption{Histogram}\label{fig:hist-real}
+\end{figure}
 
-\usepackage{multirow}
+\subsection{Setup}
 
-\newcommand{\soft}[1]{}
-\newcommand{\nopreview}[1]{}
-\newcommand\comment[1]{{\color{red}#1}}
-\newcommand\mhahn[1]{{\color{red}(#1)}}
-\newcommand\note[1]{{\color{red}(#1)}}
-\newcommand\jd[1]{{\color{red}(#1)}}
-\newcommand\rljf[1]{{\color{red}(#1)}}
-\newcommand{\key}[1]{\textbf{#1}}
+Everything is identical to Experiment 1, except that we replaced the real orderings with orderings generated from grammars fitted to the word orders found in the real language.
 
-\usepackage{amsthm}
+\paragraph{Fitting Ordering Grammars to Actual Orders}
+We create ordering grammars that are fit to the actual orderings of each language.
+These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
 
-\newcommand{\thetad}[0]{{\theta_d}}
-\newcommand{\thetal}[0]{{\theta_{LM}}}
+We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
+We parameterized probabilistic ordering grammars as follows.
+For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
+Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
+Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
+184) without replacement.
 
-\newcounter{theorem}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{thm}[theorem]{Theorem}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{lemma}[theorem]{Lemma}
+Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
+We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
 
-\frenchspacing
-%\def\baselinestretch{0.975}
+We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
+Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
+Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
 
-%\emnlpfinalcopy
-%\def\emnlppaperid{496}
+Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
 
-\title{Crosslinguistic Word Orders Enable an Efficient Tradeoff between Memory and Surprisal}
-\author{Michael Hahn, Judith Degen, Richard Futrell}
-\date{2018}
+We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
+Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
+However, discrete optimization is computationally challenging.
 
-\begin{document}
-
-\maketitle
+\paragraph{Results}
+\mhahn{to what extent do we have to present the large results figures separately for Experiment 1 and this experiment?
+or maybe we just provide some summary statistics separately, but provide the figures together.
+or all the big figures go into the SI?
+I don't mind, I just don't know what people in that audience would expect.}
 
 %
-\begin{abstract}
-
-Online memory limitations are well-established as a factor impacting sentence processing and have been argued to account for crosslinguistic word order regularities.
-Building off expectation-based models of language processing, we provide an information-theoretic formalization of these memory limitations.
-We introduce the idea of a memory-surprisal tradeoff: comprehenders can achieve lower average surprisal per word at the cost of storing more information about past context.
-We show that the shape of the tradeoff is determined in part by word order.
-In particular, languages will enable more efficient tradeoffs when they exhibit information locality: when predictive information about a word is concentrated in the word’s recent past.
-We show evidence from corpora of 52 real languages showing that languages allow for more efficient memory-surprisal tradeoffs than random baseline word order grammars.
-
-%Are languages optimized for processing with limited memory?
-%Memory limitations are well-established as a factor impacting sentence processing, and have been argued to account for crosslinguistic word order regularities.
-%Computational models of language processing implementing memory constraints take various forms and make different kinds of assumptions about the architecture underlying human language processing.
-%We first  establish general information-theoretic lower bounds on memory that hold in any model of language processing, applying both to speakers and listeners.
-%assuming only that listeners perform incremental prediction.
-%Applying these results to corpora from over 50 languages, we then provide evidence that word orders in human language are optimized for memory demands of speakers and listeners.
-% producing language and listeners predicting input.
-\end{abstract}
-
+%\subsection{Discussion: Alternative Models}
+%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
 %
+%A traditional model uses n-gram models.
+A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
+Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
+Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
+%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
 %
-%\begin{thm}\label{prop:suboptimal}
-%	For each positive integer $t$, define
-%$I_t := I[w_t, w_0 | w_{1\dots t-1}]$,
-%i.e., the mutual information between words at distance $t$, controlling for redundancy with the intervening words.
-%	Let $T$ be a positive integer, and consider a comprehender using at most $\sum_{t=1}^T t I_t$
-%bits of memory on average.
-%Then this comprehender will incur average surprisal at least
-%	$H[w_t|w_{<t}] + \sum_{t > T} I_t$.
-%\end{thm}
+%A second option is to construct a statistical grammar, such as PCFG.
+%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
+%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
+%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
 %
 
-\section{Introduction}
+%\subsection{Data}
+%\subsection{Setup}
+%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
+%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
+%This biases the hyperparameters towards favoring counterfactual grammars.
 
-\input{introduction.tex}
-
-\mhahn{pasted from the other draft}
-Each human language is a system for encoding an infinite variety of thoughts into sentences, i.e. linear strings of words.
-A primary goal of the field of linguistics is to characterize and explain the commonalities and differences among these systems, and so to gain insight into the cognitive machinery that underpins this fundamental aspect of human intelligence.
-Here we argue that human languages can be modeled as solutions to the problem of maximally efficient communication among agents with particular information-processing constraints active during speaking and listening.
-We formalize these constraints using information theory, and show novel large-scale quantitative evidence from 54 languages that information is structured in time so as to minimize memory load during language comprehension.
-
-The idea that the universal properties of human languages are best explained with reference to communicative efficiency has a long history \cite{gabelentz1901sprachwissenschaft,zipf1949human,hockett1960origin,givon1991markedness,hawkins1994performance,hawkins2004efficiency,hawkins2014crosslinguistic,croft2001functional,haspelmath2008parametric,jaeger2011language,gibson2019efficiency}.
-
-These strategies can be seen as sets of rules that differ in the way ...
-Languages differ considerably in the rules they apply to order information: English orders the object after the verb, Japanese places it before the verb \note{make more explicit here}.
-Explaining the variation in the strategies languages use to  order underlying hierarchical structures into linear strings of words has been one of the foci of linguistic research (\jd{CITE}).
-We show that these strategies reflect an underlying optimization principle: human languages are adapted to limitations in human working memory.
-
-%We test the hypothesis that human languages represent different solutions to the problem of efficient computation with constrained working memory.
-The suggestion that the structure of human language partly reflects a need for efficient processing under resource limitations has been present in the theoretical and functional linguistics literature for decades~\cite{berwick1984grammatical,hawkins1994performance}, where theories have been developed in terms of specific mechanistic and heuristic models of human language processing.
-Here
-
-\mhahn{original intro}
-Since the 1950s, it has been a persistent suggestion that human language processing is shaped by a resource bottleneck in short-term memory.
-Language is produced and comprehended incrementally in a way that crucially requires both speaker and listener to use an active memory store to keep track of what was previously said.
-Since short-term memory of this kind is known to be highly limited in capacity \citep{miller1956magical}, it makes sense for these capacity limits to comprise a major constraint on production and comprehension.
-Indeed, a great deal of work in sentence processing has focused on characterizing the effects of memory constraints on language processing \citep{gibson-linguistic-1998,lewis-activation-based-2005,levy2013memory}.
-
-At the same time, the field of functional linguistics has argued that these resource constraints not only affect online language processing, they also shape the form of human language itself.
-For example, the Performace--Grammar Correspondence Hypothesis (PGCH) of \citet{hawkins1994performance} holds that forms which are practically easier to produce and comprehend end up becoming part of the grammars of languages, and that this process can explain several of the universal properties of human languages originally documented by \citet{greenberg-universals-1963}.
-
-Here we take up the question of how to characterize short-term memory capacity limitations in language processing for both speakers and listeners, and the question of whether natural language grammars are shaped by these limitations.
-Whereas previous theories were based on specific mechanistic models of memory, our theory is purely information-theoretic, meaning that our predictions will hold independently across a wide variety of implementations and architectures.
-
-Our main new concept is the idea of a \emph{memory--surprisal tradeoff}: it is possible for a listener to achieve greater ease of word-by-word comprehension at the cost of investing more computational resources into remembering previous words, and the particular shape of the resulting tradeoff of depends on the word order properties of a language.
-Analogous results also hold for language production by resource-constrained speakers.
-We show evidence that the preferred word orders of natural languages are those that enable efficient memory--surprisal tradeoffs.
-
-The remainder of this paper is structured as follows.
-In Section~\ref{sec:background}, we give a brief review of previous work on the effects of short-term memory on languages and language processing.
-In Section~\ref{sec:ms-tradeoff}, we describe the memory--surprisal tradeoff and how it results from rate--distortion theory, the theory of optimal information processing under resource constraints \citep{cover2006elements}.
-In Section~\ref{sec:info-locality}, We prove that word orders enable efficient processing in terms of the memory--surprisal tradeoff when they exhibit \emph{information locality}: whenever utterance elements that predict each other are close to each other.
-We argue that information locality is a good model of the effects of memory constraints on language processing.
-In Section~\ref{sec:experiment1}, languages which have previously been shown to be preferred in artificial language experiments are exactly those that enable efficient memory--surprisal tradeoffs \citep{fedzechkina2017human}.
-In Section~\ref{sec:experiment2}, we show that word orders of natural languages as found in dependency corpora \citep{ud} enable more efficient memory--surprisal tradeoffs than baseline word orders.
-Section~\ref{sec:conclusion} concludes.
-
-\section{Background}
-\label{sec:background}
-% More detailed lit review
-
-A wide range of work has argued that natural language orders information in ways that reduce memory effort.
-An early example is \cite{miller-finitary-1963}, who attributed the unacceptability of multiple center embeddings in English to limitations of human working memory.
-\cite{hawkins-efficiency-2003} provides cross-linguistic evidence that word orders are optimized for processing based on local contexts.
-Further work has found computational, corpus-based evidence that memory limitations impact language structure and production.
-In particular, languages have been shown to shorten the length of syntactic dependency lengths \citep{futrell-large-scale-2015}.
-Dependency length can be linked to memory use in certain models of incremental syntactic parsing, and increases processing difficulty in theories of memory in sentence processing \citep{gibson-linguistic-1998}.
-\cite{gildea-human-2015} further provide evidence from five languages that word orders optimize predictability from local contexts.
-\cite{futrell-noisy-context-2017} provide evidence that language shows information locality, i.e., elements with higher mutual infomation are closer together, which is predicted by their model of Lossy-Context Surprisal.
-
-All these models of memory in sentence processing, and derived measures of efficiency for memory allocation, require specific assumptions about the architecture of memory.
-This leaves open the question whether such assumptions are necessary, or whether word orders across languages are optimized for memory independently of the implementation and architecture of human language processing.
-
-We approach this question by first providing general information-theoretic lower bounds on memory load that will hold independently of the architecture of memory representations.
-We will consider a general setting of a listener performing incremental prediction.
-Our result immediatly entails a link between boundedness of memory and locality, which had been stipulated or derived from assumptions about memory architecture in previous models \citep{gibson-linguistic-1998, lewis-activation-based-2005, futrell-noisy-context-2017}.
-We will then use corpus data from over 52 languages to provide evidence that their word orders help lower memory cost.
-
-%TODO mention prior work
-%
-%-  center embeddings dispreferred/impossible to process \cite{miller-finitary-1963}
-%
-%- explanation of greenberg universals
-%
-%- dependency length minimization
-%
-%- \cite{gibson-linguistic-1998}
-%
-
-%\paragraph{\cite{berwick-grammatical-1986}}
-%They considered incremental parsing, or, more precisely, incremental recognition of grammaticality.
-
-%also (CITE): operations in a specific incremental parsing model
-
-\section{Memory-Surprisal Tradeoff}
-\label{sec:ms-tradeoff}
-
-\input{tradeoff.tex}
-
-\section{Information Locality}\label{sec:info-locality}
-
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
-\section{Experiment 1: Memory and Dependency Length}
-
-\input{toy-experiment.tex}
-
-%\paragraph{Center Embeddings}
-%\cite{miller-finitary-1963} attributed the unacceptability of multiple center-embedding to memory limitations.
-%\cite{gibson-linguistic-1998}
-%\paragraph{Other Psycholinguistic Predictions}
-% RF: the fact that you would get locality effects given medium WM capacity, but not very high or very low WM capacity, as Bruno Nicenboim found. And maybe some speaker-listener asymmetries. 
-%\paragraph{Speakers}
-% RF: what matters for the speaker is not I[w_t, w_0 | w_1, …, w_{t-1}], but I[w_t, w_0 | w_1, …, w_{t-1}, G] where G is some representation of the speaker’s goal (like in the van Dijk paper).
-This changes the interpretation of the mutual information.
-For the listener, it’s just redundancy.
-For the speaker, it’s redundancy *conditional on the goal*—which you could interpret as something like conceptual relatedness of linguistic elements.
-Then the speaker’s pressure is to keep conceptually related things close.
-
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
-
-We now investigate whether word orders as found in natural language optimize the two memory-surprisal tradeoffs.
-We compare the memory-surprisal tradeoffs of 52 actual languages to those of counterfactual reorderings.
-We cannot just compare to random orderings of individual syntactic trees, as such languages would not have word order regularities.
-Therefore, we compare to counterfactual word order grammars.
-
-\begin{figure}
-\centering
-\begin{dependency}[theme = simple]
-   \begin{deptext}[column sep=1em]
-	   I \&	   wrote \& risāla \& li \& sadīq  \\
-   \end{deptext}
-	%   \deproot{3}{ROOT}
-   \depedge{1}{2}{obj}
-	%   \depedge[edge start x offset=-6pt]{2}{5}{ATT}
-   \depedge{1}{4}{obl}
-   \depedge{4}{3}{case}
-   %\depedge[arc angle=50]{7}{6}{ATT}
-\end{dependency}
-	\caption{TODO Dependencies example}\label{fig:dependency}
-\end{figure}
-
-\subsection{Data}
-We draw on corpora annotated with syntactic structures.
-The Universal Dependencies project has compiled dependency corpora for several dozen languages~\citep{nivre-universal-2017}.
-
-\paragraph{Dependency Grammar}
-In dependency corpora, sentences are annotated with \emph{dependency trees} (Figure~\ref{fig:dependency}).
-These are directed trees describing the grammatical relations among words.
-For example, the arcs labeled ``obj'' represent that the noun in question is the \emph{direct object} if the verb, rather than e.g.
-the subject or an indirect object.
-A dependency arc is drawn from a \emph{head} (e.g. TODO in Figure TODO) to a \emph{dependent} (e.g. TODO).
-Dependency trees can be defined in terms of many different syntactic theories \cite{corbett1993heads}.
-Although there are some differences in how different formalisms would draw trees for certain sentences, there is broad enough agreement about dependency trees that it has been possible to develop large-scale dependency-annotated corpora of text from dozens of languages \cite{nivre2017universal}.
-
-\paragraph{Corpora}
-We considered all languages for which there are Universal Dependencies 2.3 treebanks with a total of at least 500 sentences of training data.
-We excluded data from historical languages.\footnote{Ancient Greek, Classical Chinese, Coptic, Gothic, Latin, Old Church Slavonic, Old French.}
-This resulted in 54 languages.
-
-For each of these languages, we pooled all available corpora in one dataset.
-We excluded corpora that primarily contain code-switched text\footnote{Hindi English corpus}, or text created by non-native speakers.\footnote{ESL, CFL}
-Universal Dependencies corpora have a predefined split into \emph{training}, \emph{held-out} (also known as \emph{development}), and \emph{test} partitions.
-While larger corpora have all three partitions, smaller corpora often have only some of these partitions.
-For most language, we used the predefined data split, separately pooling data from the different partitions. %We did not use the test partitions.
-For some languages with little data, there is no predefined training partition, or the training partition is smaller than the other partitions.
-In these cases, we redefined the split to obtain more training data:
-For these languages, we pooled all the available partitions, used 100 randomly selected sentences as held-out data, and used the remainder as training data.\footnote{This affects Amharic, Armenian, Breton, Buryat, Cantonese, Faroese, Kazakh, Kurmanji, Naija, Thai, and Uyghur.}
-For each language, we used the training and held-out sets for estimating the memory-surprisal tradeoff (see Section~\ref{sec:method}).
-We provide the sizes of the resulting datasets in Table~\ref{tab:corpora}.
-
-\subsection{Counterfactual Ordering Grammars}
-We define ordering grammars, small models of the rules by which languages order syntactic structures into sentences.
-Our formalism of ordering grammars adapts the method of \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015} to the setting of dependency corpora.
-
-Universal Dependencies defines 37 universal syntactic relations that are used to label dependency arcs across all corpora.
-These relations encode cross-linguistically meaningful relations such as subjects, objects, and adjectival modifiers.
-We define ordering grammars by assigning a parameter $a_\tau \in [-1,1]$ to every one of these 37 universal syntactic relations.
-Relations sometimes have language-specific subtypes; we do not distinguish these subtypes.\mhahn{give an example to make clear what this refers to}
-
-Following Gildea and colleagues, this parameter defines how dependents are ordered relative to their head:
-Given a head and a set of dependents, we order each dependents by the parameter $a_\tau$ assigned to the syntactic relation linking it to the head.
-Dependents with negative weights are placed to the left of the head; dependents with positive weights are placed to the right.
-
-Ordering grammars describe languages that have consistent word order:
-For instance, the subject is consistently ordered before or after the verb, depending on whether the parameter for the verb-subject dependency is positive or negative.
-
-We define baseline grammars by randomly sampling the parameters $a_\tau$.
-Such baseline grammars define languages that have consistent word order, but do not exhibit any systematic correlations between the orderings of different dependents.
-
-\paragraph{Discussion}
-In actual languages, the ordering of words is largely determined by the syntactic relations (CITE).
-However, certain kinds of rules cannot be modeled by our word order grammars, such as rules sensitive to the category of the dependent (e.g., differences between nominal and pronominal objects).
-Word order freedom also is not modeled.
-In this sense, ordering grammars represent approximations to the kinds of ordering rules found in natural language \cite{gildea-optimizing-2007, gildea-grammars-2010, gildea-human-2015}.
-
-\subsection{Estimating Memory-Surprisal Tradeoff}\label{sec:method}
-To estimate mutual informations, we use LSTM recurrent neural networks, the basis of the state of the art in modeling natural language (CITE)\footnote{While Transformers are now better at modeling when large training sets comprising millions of sentences per language are available, they have not reached the level of LSTMs on small/medium training data such as the UD corpora} and predicting the surprisal effect on reading times~\citep{frank-insensitivity-2011, goodkind-predictive-2018}.
-%We model language on the level of individual word forms.
-We provide data from alternative estimation methods in the SI.
+%\subsection{Setup}
 
 %\paragraph{Data}
-
 %Given a sequence of input words $w_1, ..., w_n \in V$, the model 
@@
 
-\paragraph{Model and Parameter Estimation}
-We use a recurrent neural network with Long-Short-Term Memory cells~\citep{hochreiter-long-1997} (CITE for Neural LM).
-This architecture takes as input a sequence $x_1 ...
-x_N$ of words, and at each time step $t=1, ..., N$, calculates a probability distribution over the next word $w_{t}$ given preceding words $w_1 ...
-w_{t-1}$: $p(w_t|w_1...w_{t-1})$.
-
-The network is parameterized by a vector $\theta$ of weights determining how the activations of neurons propagate through the network~\citep{hochreiter-long-1997}.
-Given a corpus, the numeral parameters of the LSTM are chosen so as to minimize the average surprisal across the training corpus.
-%We can think of the LSTM parameters as forming one large vector $\theta$.
-At the beginning of training, the parameters $\theta$ are randomly initialized.
-
-The training corpus is chopped into word sequences $w_1 ... w_T$ of length $T$ ($T = 20$ in our experiments).
-If $\theta_n$ consists of the LSTM parameters after $n$ training steps, we randomly select a word sequence $w_1 ...
-w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
-We then update the parameter vector:
-\begin{equation}\label{eq:train}
-	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
-\end{equation}
-where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.
-When calculating the parameter update, we use three standard methods of regularization that have been shown to improve neural language modeling: dropout~\citep{srivastava-dropout:-2014}, word dropout, and word noising~\citep{xie2017data}.
-In this process, the word sequences are sampled without replacement.
-Once all sequences have been processed, we start another pass through the training data.
-Before each pass through the training data, the order of sentences of the training data is shuffled, and the corpus is again chopped into sequences of length $T$.
-
-After each pass through the training data, the average surprisal at the current parameter setting $\theta_n$ is evaluated on the held-out partition.
-We terminate training once this held-out  surprisal does not improve over the one computed after the previous pass any more.
-
-%\paragraph{Regularization}
-%Dropout, Word dropout, Word noising
-%Any statistical estimation problem faces the 
-%The quality of neural network models is further improved by regularization, improving generalization to the full data distribution.
-%In the case of neural networks, the most successful regularization methods typically take the form of random modifications to the input and internal activations when computing the gradients in the training updates~(\ref{eq:train}).
-%First, we apply \emph{dropout} both to the input vectors $x_i$ and to the output vectos $h_i$, randomly setting individual elements to zero with rate $p_{embedding} \in [0,1]$.
-%Second, we apply \emph{word dropout}, randomly zeroing out entire input vectors $x_i$ at rate $p_{word}$.
-
-\paragraph{Choice of Hyperparameters}
-
-The LSTM model has a set of numerical \emph{hyper-parameters} that need to be specified before parameter estimation, namely the dimensionalities of the embeddings $d_{emb}$, the dimensionality of the hidden states $d_{LSTM}$, and the number of LSTM layers $d_{layer}$, the learning rate $\alpha$, and the regularization parameters (dropout rate $p_{embedding}$, word dropout rate $p_{word}$, word noising rate $p_{noising}$).
-We choose these parameters so as to minimize the average surprisal on the held-out partition resulting at the end of parameter estimation.
-
-For each corpus, we used Bayesian optimization using the Expected Improvement acquisition function \citep{snoek-practical-2012} to find a good setting of the hyperparameters.
-We optimized the hyperparameters to minimize average surprisal on languages generated from random word order grammars.
-This biases the hyperparameters towards modeling counterfactual grammars better, biasing them \emph{against} our hypothesis.
-
-For computational efficiency, neural language models can only process a bounded number of distinct words in a single language.
-For each corpus, we limited the number of distinct processed words to the $N=10,000$ most common words in the training corpus, a common choice for neural language models (CITE).
-Following (CITE), we represented other words by their part-of-speech tags as annotated in the corpora.
-This applied to 37 languages, affecting an average of 11~\% of words in this languages.
-We believe that this modeling limitation does not affect our results for the following reasons.
-First, this affects the same words in real and counterfactually ordered sentences.
-Second, all excluded words are extremely infrequent in the available data, occurring less than 10 times (except for Czech and Russian, the languages for which we have by far the largest datasets).
-Many of the excluded words occur only once in the dataset (78 \% on average across the affected languages).
-This means that any model would only be able to extract very limited information about these words from the available training data, likely \emph{less} than what is provided by the part-of-speech tag.
-Third, traditional N-gram models, which do not have this limitation, provide results in qualitative agreement with the neural network-based estimates (see SI).
-
-%Given a corpus, we estimate a language model by training the LSTM on a training partition to maximize the data likelihood, and stopping training once the data likelihood on a held-out partition drops.
-%The hyperparameters (number of hidden units, learning rate, etc.) are chosen so as to maximize the likelihood of the held-out partition.
-%This procedure helps prevent overfitting the model to the training data, and ensure that it generalizes to unseen data.
-
-\paragraph{Estimating the Memory-Surprisal Tradeoff Curve}
-
-The quantity $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$ in~(\ref{eq:memory-bound}) is equal to the difference 
-\begin{equation}
-H[w_t|w_1, ..., w_{t-1}] - H[w_t|w_0, w_1, ..., w_{t-1}]
-\end{equation}
-For each word in the held-out partition, we compute the difference
-\begin{equation}
-	-\log P_\theta[w_t | w_0, w_1, ..., w_{t-1}] - P_\theta[w_t | w_1, ..., w_{t-1}]
-\end{equation}
-and take the average over these.
-We cut $T$ off at 20, as this is the length of the sequences processed by the model.
-
-We then used linear interpolation to interpolate the surprisal value for memory values in between these values. (TODO make a figure).
-This is justified theoretically (TODO maybe discuss when introducing the theorem).
-
-We estimate the unigram entropy $H[w_0]$ by averaging over all models.
-
-%
-%\subsection{Discussion: Alternative Models}
-%In view of the NLP literature, the following are the main other options that exist for estimating mutual information and probabilities in sequences:
-%
-%A traditional model uses n-gram models.
-A challenge of n-gram models is that they do not express any morphosyntactic generalizations.
-Furthermore, standard n-gram models do not express any generalizations about pairs of words that are not adjacent -- e.g., encoding a generalization about morphological agreement between two words is hard for such a model to capture if the two words are not always adjacent.
-Both the small scale of available corpora in many languages and free word order in many languages with rich morphology thus seem to make such models unattractive.
-%We evaluate our hypothesis using n-gram models in SI Section X, confirming the conclusions obtained from neural models.
-%
-%A second option is to construct a statistical grammar, such as PCFG.
-%The challenge is to encode statistical morphosyntactic generalizations, and to decide which independence assumptions to put into the model.
-%One can either decide on a language-specific basis which generalizations to put in (laborious and might introduce bias), or choose a general model family that is rich enough to learn generalizations.
-%The second option will make this a machine learning model that, for our purposes, does not seem to be superior to a recurrent neural network.
-%
-
-%\subsection{Data}
-%\subsection{Setup}
-%The recurrent neural network architecture has a range of adjustable parameters such as the number of neurons.
-%For each language, we used Bayesian optimization using the Expected Improvement acquisition function (CITE) \citep{snoek-practical-2012} to find a good setting of the hyperparameters, taking average surprisal on random grammars as the objective.
-%This biases the hyperparameters towards favoring counterfactual grammars.
-
-%\subsection{Setup}
-
-For each language, we collected data from the actual orderings and from several random grammars.
-We collect multiple samples for the actual orderings to control for variation due to the random initialization of the neural network.
-For each of the random grammars, we collect one sample.
-Data is collected according to a precision-based stopping criterion described in Section (REF).
 %We collected data from the actual and random orderings in proportion one to two.
@@
 
-\subsection{Statistics}
-
-We now describe how we compared memory-surprisal tradeoffs between real and baseline languages.
-
-%For each sample, we estim
-%For each language, we viewed surprisal as a function of memory.
-%We used linear interpolation to obtain surprisal values for all levels of memory.
-
-We want to test whether languages' surprisal-memory tradeoffs better than those of most baseline languages.
-%We view surprisal as a function of memory load, so that the curve is defined on all of $\mathbb{R}_+$.
-We compare real and baseline languages by evaluating which languages result in lower surprisal at the same level of memory.
-We now describe the statistics we use to quantifying the difference between real and baseline languages.
-We do everything in a frequentist framework (null hypothesis testing \& confidence intervals), as we can do exact tests and confidence intervals without parametric assumptions.
-Maybe we can explain how the tests \& CIs also have reasonable Bayesian interpretations (for the specific methods used here, rejection of the null should guarantee that the posterior of the null hypothesis is small under a wide range of priors.).
-
-\paragraph{Confidence Interval for Medians}
-We use a (nonparametric and nonasymptotic) confidence interval for the median surprisal at each memory value, using the binomial test.
-We consider the medians over all runs for the real language, and over all baselines grammars.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py
-
-\paragraph{CI for Median Difference}
-We create (nonparametric and nonasymptotic) confidence interval for the difference between real and baseline median surprisals at each memory value.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianDiffCI.py
-
-\paragraph{Pointwise Significance Test}
-For each memory value $\mu$, we do a nonparametric and nonasymptotic significance hypothesis test against the null hypothesis that at least half of the baseline grammars have lower surprisal than the actual language (Figure~\ref{fig:nhst-pointwise}).
-Formally, let $W_-(\mu)$ be the proportion of baseline languages that have strictly lower surprisal than the real language at memory level $\mu$.
-We take the real language to be represented by the \emph{sample median}.
-For each level $\mu$ of memory, we consider the null hypothesis that
-\begin{equation}
-	W_-(\mu) \leq 0.5
-\end{equation}
-%In reality, we do not observe the curve of the real language exactly, but as noisy samples from our computational estimator.
-We use the Binomial Test.
-
-%This statistic also should have a reasonable Bayesian interpretation:
-% E.g., if the random samples are unimodal, and we do inference over only the median (location family),
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/nhst.png}
-\end{center}
-	\caption{Illustration for the pointwise null-hypothesis significance test.
-At a given level of memory, we test against the null hypothesis that at least half of the baseline orders provide lower surprisal than the real language.}\label{fig:nhst-pointwise}
-\end{figure}
-
-%For each memory value $\mu$, we do a significance test (nonparametric and nonasymptotic).
-%\begin{equation}
-%	W_+(\mu) \geq W_-(\mu)
+%Training progresses in a series of parameter update steps, constructing updated parameters $\theta_0, \theta_1, \theta_2, \dots$.
+%In the $n$-th update step, we first randomly select a word sequence $w_1 ...
+w_T$ from the training corpus, and use the LSTM using the current parameter setting $\theta_n$ to compute the per-word surprisals.
+%We then update the parameter vector:
+%\mhahn{maybe better to just say we use SGD}
+%\begin{equation}\label{eq:train}
+%	\theta_{n+1} := \theta_n + \alpha \partial_\theta \left(\sum_{i=1}^T \log p_\theta(w_i|w_1...w_{i-1})\right)
 %\end{equation}
-%We use the empirical median for the real language.
-
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single.py
-
-%We take the REAL values to be estimated exactly by their medians.
-
-\paragraph{Pointwise Quantile Estimate}
-%CI for quantile: % yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections.py
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile.png}
-\end{center}
-	\caption{Illustration for the quantile estimate.
-At each level of memory, we provide an estimate of the percentage of baseline languages that have lower surprisal than the real language.}\label{fig:quantile-pointwise}
-\end{figure}
-
-For each level of memory, we estimate what percentage of baseline languages have lower surprisal than the real language.
-This is described in Figure~\ref{fig:quantile-pointwise}.
-
-We derive a confidence interval under the assumption that the distribution of baseline languages is unimodal.
-
-We derive a CI for the quantile, under the assumption that the baseline distribution is unimodal. We take the REAL values to be estimated exactly by their medians.
-
-We want to create a CI at each Memory value for the quantile.
-
-Let $n_+$ be the better baseline samples, $n_-$ the worse ones.
-
-Let $\theta$ the best baseline sample that is worse than the real median.
-
-We want to get a confidence bound $q$ on $P(X < x_{real})$.
-
-Let $p := P(N_+ \leq n_+ | N_+ + N_-; q)$.
-Then output $(0, q)$ as a level $p$ CI for the parameter $P(X < x_{real})$.
-
-We minimize $q$ subject to $p < 0.05$.
-
-Once we have $q$, we can make it a bit better under the assumption that the baseline distribution is unimodal.
-
-This CI is exact in the sense that it does not involve asymptotic approximations or parametric assumptions, but it is extremely conservative.
-
-Also the following does not assume unimodality, and ends up getting about the same intervals
-% yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_UnimodalBoundOnQuantile_BothDirections_NoAssumption.py
-
-\paragraph{Global Quantile Estimate}
-
-\begin{figure}
-	\begin{center}
-\includegraphics[width=0.45\textwidth]{figures/quantile-global.png}
-\end{center}
-	\caption{Illustration for the global quantile estimate. For each sample for the real language, we compare the memory-surprisal curve to all baselines.}\label{fig:quantile-global}
-\end{figure}
-
-For each sample $x$ from real orderings, we look at the proportions $N_+(x)$ of samples from the baseline languages that are more optimal than $x$ throughout the entire range where both curves are defined, and the proportion $N_-(x)$ of baseline samples that are consistently less optimal.
-
-%We consider the null hypothesis that, on average, not more baseline languages are consistently less optimal than are consistently more optimal than the real orderings:
-%\begin{equation}
-%	\E_{x \sim P_1}[W_+(x)] \geq \E_{x \sim P_1}[W_-(x)]
-%\end{equation}
-
-We estimate the quotient
-\begin{equation}\label{eq:g}
-	G :=	\frac{\E_{x \sim P_1}[W_+(x)]}{\E_{x \sim P_1}[W_+(x) + W_-(x)]}
-\end{equation}
-where $P_1$ is the distribution over values obtained for real orderings.
-We use a bootstrapped confidence interval for $\E[G]$ for quantifying the degree of optimization.
-For bootstrapping, we separately resample samples from the real language and from the baseline grammars.
-
-Unlike the other statistics, this one provides a global measure of the degree of optimization of the real language.
-Due to the use of bootstrapping, the confidence intervals are not exact.
-
-%- bootstrapping
-%- subsampling
-%- permutation test / rank test ??
-%
-
-\subsection{Number of Samples}
-Training neural language models is computationally costly.
-Therefore, we used a precision-based stopping criterion to adaptively choose a sample size for each language.
-Precision-based stopping criteria offer a way to adaptively choose sample size without biasing results (CITE).
-
-For each language, we first collected 10 data points for real orderings and 10 data points for baseline orderings.
-We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.\footnote{Due to a scripting error, a much higher number of samples was generated for Erzya.}
-
-This procedure was parallelized on several machines.
-In the case where the stopping criterion was reached for a language while several machines were still computing samples for this language, we did not discard those samples.
-Consequently, more samples were collected than necessary to reach the stopping criterion; however, in a way that does not bias our results towards or against our hypothesis.
-
-%Due to parallelization of this procedure, it often produced more samples than required, when the stopping criterion .
-%We chose these thresholds based on preliminary simulations which had suggested that these widths were achievable at acceptable computational cost.
-%For each language, we collected at least 5 data points for real orderings and at least 10 data points for baseline orderings.
-%We continued obtaining new data points until the CI for $G$ had width $\leq 0.15$, or there were 100 samples from $P_1$ and 300 samples from $P_2$.
-%Up to the end, we chose the next sample to be from $P_0$ with probability 2/3, and $P_1$ otherwise.
-
-\subsection{Results}
-
-The numbers of samples taken per language are provided in Table~\ref{tab:samples}.
-
-%In Figure~\ref{tab:plain-results} (TODO), we show the estimated memory-surprisal tradeoff curves for all samples.
-
-In Figure~\ref{tab:medians}, we show the medians for real and baseline languages.
-
-Descriptively, the real language provides better tradeoffs than the median of the baselines across languages, with four exceptions (Latvian, North Sami, Polish, Slovak).
-
-In Figure~\ref{tab:slice-hists-real}, we show the distribution of surprisals achieved at the maximal memory value for real and random languages.
-
-In Figure~\ref{fig:hist-real}, we show surprisals at maximum memory, after z-transforming for each individual language and then aggregating.
-
-In Table \ref{tab:median_diffs}, we show the differences in median surprisal, as a function of memory.
-
-In Table~\ref{tab:boot-g}, we report the bootstrap estimates and confidence intervals for G~(\ref{eq:g}).
-$\E[G]$ was not estimated to be significantly above $>5$ for four languages: Latvian, North Sami, Polish, and Slovak.
-
-In Table~\ref{tab:quantiles}, we show the quantiles.
-
-\subsection{Discussion}
-
-We have found that 48 out of 52 languages provide better memory-surprisal tradeoffs than random baselines with consistent but counterfactual word order rules.
-
-Four languages provide exceptions; these are Latvian (Baltic), North Sami (Uralic), Polish and Slovak (both Slavic).
-All four languages have strong word order freedom (CITE).
-Freedom of word order plausibly makes sentences less predictable, as the same syntactic structure can receive different surface realizations.
-We thus hypothesized that freedom of word order impacts the memory-surprisal tradeoff, and that languages with more strongly fixed word order should display more optimal memory-surprisal tradeoffs.
-%We hypothesized that freedom of word order freedom may be responsible for the difference between these languages and the other languages.
-
-To test this hypothesis, we examined the correlation between word order freedom and the surprisal difference between real and baseline orderings.
-To quantify word order freedom, we used a corpus-based estimate, the \emph{branching direction entropy}~\citep{futrell-quantifying-2015}.
-This is the entropy of the ordering (head-first or dependent-first) of dependencies conditioned on the dependency label and the part-of-speech label of head and dependent.
-%\cite{futrell-quantifying-2015} showed that this me
-These two quantities are plotted in Figure~\ref{fig:hist-real}.
-We found that branching direction entropy was strongly correlated with the surprisal difference between real and baseline orderings (Spearman correlations -0.58, $p = 7.414e-6$).
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/surprisal-branching-entropy-REAL.pdf}
-	\caption{Surprisal Difference vs Branching Direction Entropy.}\label{fig:hist-real}
-\end{figure}
-
-\section{Experiment 3: Fixed Word Orders}
-
-We test this hypothesis by comparing baseline languages to \emph{fixed-order} versions of the real languages.
-This enables us to tease apart the impact of the languages' word order rules from the impact of word order freedom.
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-GROUND-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-	\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\paragraph{Fitting Ordering Grammars to Actual Orders}
-We create ordering grammars that are fit to the actual orderings of each language.
-These grammars faithfully represent the ordering rules if the actual language, to the extent that is possible in the formalism of ordering grammars.
-
-We construct these grammars by constructing \emph{probabilistic ordering grammars}, and setting the parameters to maximize the \emph{likelihood} of the actually observed orderings.
-We parameterized probabilistic ordering grammars as follows.
-For each relation type $\tau$, we introduce a \emph{direction parameter} $a_\tau \in [0,1]$ and a \emph{distance parameter} $b_\tau \in \mathbb{R}$.
-Each dependent is ordered on the left of its head with probability $a_\tau$ and to the right with probability $1-a_\tau$. 
-Then for each set of co-dependents $\{s_1, \dots , s_n\}$ placed on one side of a head, their order outward from the head is determined by iteratively sampling from the distribution $\operatorname{softmax}(b_{\tau_1}, \dots, b_{\tau_n})$ (\cite{goodfellow2016deep}, p.
-184) without replacement.
-
-Given a dependency tree, a probabilistic ordering grammar assigns a probability distribution over the possible projective linearizations of that tree.
-We use gradient descent to find parameters $a_\tau, b_\tau$ so as to maximize the overall likelihood of the orders in the actual corpus.
-
-We convert probabilistic ordering grammars into ordinary ordering grammars by the following method.
-Let $A_-$ be those relations $\tau$ where $a_\tau > 0.5$, similarly for $A_+$ those here $a_\tau \geq 0.5$.
-Then we order all relations in $A_-$ by $b_\tau$ in \emph{decreasing} order, and those in $A_+$ by $b_\tau$ in \emph{increasing} order.
-
-Then ordering a tree following the converted version is equivalent to greedily choosing the highest-probability linearization for the dependents of each head in a tree.
-
-We choose this method since maximum-likelihood grammars can be constructed with simple gradient descent.
-Another option would be to use some kind of discrete optimization method to approximate the original orders without a probabilistic method.
-However, discrete optimization is computationally challenging.
-
-\paragraph{Results}
-
-\section{Discussion}
-
-\subsection{Relation to Theories of Sentence Processing}
-
-\paragraph{Lossy-Context Surprisal}
-\citet{futrell-noisy-context-2017} describe a processing model where listeners make predictions (and incur surprisal) based on lossy memory representations.
-In particular, they consider loss models that delete, erase, or replace words in the past.
-Under the assumption that loss affects words more strongly that are further in the past, they derive a principle of information locality:
-A listener will incur surprisal
-$$ -\log P(w_t) - \sum_{j=1}^{t-1} f(i-j) pmi(w_i; w_j) + R$$
-where the `survival probability' $f(d)$ decreases as the distance $d$ between two words increases, and $R$ is a remainder term that can be argued to be small.
-Given that $f$ is assumed to be decreasing, this prediction loss will be smaller when words with high mutual information are closer together in the input.
-Our Proposition~\ref{prop:suboptimal} can be seen as an analogous result for general models of memory.
-
-\paragraph{Relation to Dependency Locality}
-The quantity described in Proposition~\ref{prop:lower-bound} is formally similar to Storage Cost in the Dependency Locality Theory (DLT) \citep{gibson-linguistic-1998}: Storage cost at a given timestep is defined as the number of predictions that are held in memory.
-Storage cost only considers predictions that are certain, and each prediction takes an equal amount of memory.
-In contrast, the result in Proposition~\ref{prop:lower-bound} can be seen as weighting predictions by their certainty and the amount of predictive information.
-In this sense, DLT storage cost can be seen as an approximation to Proposition~\ref{prop:lower-bound}.
-
-also surprisal -- integration cost
-
-\paragraph{Center Embeddings, ...}
-Yngve 1960, had a complexity measure, but doesn't work well for left-branching structures
-
-Miller and Chomsky 1963
-
-Frzier 1985 local nonterminal count
-
-Rambow and Joshi 1994 using TAG
-
-Marcus 1980 deterministic parsing
-
-(Sabrina Gerth, Memory Limitations in Sentence Processing)
-
-\subsection{Relation to other work}
-
-\paragraph{Statistical Complexity}
-There are deep connections between our formalization of listener memory and studies of dynamic systems in the Physics literature.
-%Speaker memory corresponds to \emph{Generative Complexity} \cite{loehr-non-sufficient-2008, loehr-predictive-2010}.
-The tradeoff between listener memory and surprisal is formally equivalent to the \emph{Recursive Information Bottleneck} considered by \cite{still-information-2014}.
-In the limit of optimal prediction and minimal surprisal, our formalization of listener memory is equivalent to the notion of \emph{Statistical Complexity} \citep{crutchfield-inferring-1989}.
-In the limit $T \rightarrow \infty$, the quantity in (\ref{eq:memory}) is equal to the \emph{excess entropy}, which is known to bound statistical complexity \citep{crutchfield-inferring-1989}.
-However, the link between memory and information locality provided by our Proposition~\ref{prop:suboptimal} appears to be a novel contribution.
-Relatedly, \cite{sharan-prediction-2016} shows a link between excess entropy and approximability by $n$-th order Markov models, noting that processes with low excess entropy can be approximated well with Markov models of low order.
-
-also information-theoretic studies of memory capacity
-
-\paragraph{Decay of Mutual Information}
-In Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal}, we showed a close link between memory and the decay of \emph{conditional} mutual information $I_t := I[w_t, w_0 | w_{1\dots t-1}]$.
-Prior work has studied the decay of \emph{unconditional} mutual information $I[w_t, w_0]$ in natural language \citep{ebeling-entropy-1994,lin-critical-2017}, and linked it to locality and memory \citep{futrell-noisy-context-2017}.
-
-The decay of unconditional mutual information is less closely linked to memory requirements than conditional mutual information:
-While the decay of conditional mutual informations provides a lower bound on memory need, unconditional mutual information does not:
-Consider the constant process where with probability 1/2 all $w_t = 0$, and with probability 1/2 all $w_t = 1$.
-%%$w_t = c$, where $c$ is random but independent of $t$ for each specific draw from the process.
-The unconditional mutual information is 1 at all distances, so does not decay at all, but the process only requires 1 bit of memory.
-Conversely, one can construct processes where the unconditional mutual informations are 0 for all $t$, but where $P > 0$ and this predictive information is actually spread out over arbitrarily large distances (that is, the ratio of memory $M$ and predictability $P$ can be made arbitrarrily large).\footnote{First, consider the process (called X by REF) consisting of 2 random bits and their XOR.
-This one has bounded nonzero $J$, but zero unconditional MI.
-To get unbounded $J$, consider the following process for any $N \in \mathbb{N}_{>2}$: Every $w_t$ is equal to the XOR of $w_{t-1}$ and $w_{t-N}$, such that each $w_t$ has $Bernoulli(1/2)$ as its marginal.
-The unconditional mutual information between any two timesteps is zero, but modeling the process requires $N$ bits of memory.}
-
-\paragraph{Long-range dependencies in text}    % excess entropy
-\cite{debowski-excess-2011} has studied the excess entropy of language across long ranges of text, in particular studying whether it is finite. % compute excess entropy in text
-Our work contrasts with this work in that we are interested in dependencies within sentences.
-
-\subsection{Discussion}
-
-\paragraph{Speakers vs Listeners}
-can we say something about this
-
-\paragraph{Decay vs Interference}
-Work has suggested that interference and memory overload is more appropriate than decay \cite[p. 408]{lewis-activation-based-2005} for modeling locality and memory in sentence processing.
-The bounds in Propositions~\ref{prop:lower-bound} and \ref{prop:suboptimal} hold for any type of memory model, and are thus compatible with decay- or interference-based models.
-The formula in (\ref{eq:memory-bound}) might suggest that boundedness of memory entails that memory has to decay.
-This is not the case:
-A long dependency can be maintained perfectly with low average memory:
-Informally, if every sentence is $N$ words long and has one long-distance dependency spanning the entire sentence, this dependency can be modeled perfectly with a memory cost that is independent of $N$.
-In contrast, if every symbol strongly and non-redundantly depends on the character $T$ steps in the past, with $T$ large, this will create a memory cost proportional to $T$.
-
-\paragraph{Memory and Hierarchical Structure; Finiteness of Memory}
-Processing nontrivial hierarchical structures typically requires unbounded amounts of memory.
-However, crucially, the \emph{average} memory demand for prediction can be finite, if the probability mass assigned to long dependencies is small.
-For instance, languages defined by Probabilistic Context Free Grammars (PCFG) always have finite average memory.
-The reason is that PCFGs assign low probabilities to long sequences.\footnote{Proposition 2 in \cite{chi-statistical-1999} implies that words drawn from a PCFG have finite expected length.
-This implies that average memory demands are finite.}
-
-\section{Conclusion}
-
-TODO
-
-\bibliographystyle{apalike}
-\bibliography{literature}
-
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.95\textwidth]{neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_REAL_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{quantiles_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%-- English, Korean, Russian
-%
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%
-%-- character-level Russian
-%
-
-\section{Character-Level Modeling}
-
-\section{Non-UD Dependency Treebanks}
-
-- other treebanks
-
--- spoken Japanese (T{\"u}ba-J/S)
-
--- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-
--- another Chinese dependency treebank LDC2012T05
-
-Due to the sizes of these treebanks, can also do experiment with full word forms.
-
-\section{Constituency Treebank}
-
--- Penn treebank \citep{marcus-building-1993}
-
--- spoken English (T{\"u}ba-E/S)
-
--- spoken German (T{\"u}ba-D/S)
-
--- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
-\end{document}
+%where $\alpha \in \mathbb{R}_+$ is the \emph{learning rate}.

### 2020-01-15 writeup/paper.tex

@@
 
-\section{Information Locality}\label{sec:info-locality}
-\mhahn{not sure about this section}
-TODO derive information locality
-
-Due to the factor $t$ inside each term of the sum, carrying the same amount of information over longer distances requires more memory -- that is, modeling long statistical dependencies is more costly in terms of memory than modeling shorter ones.
-This formalizes a general, assumption-free, link between memory and locality in language production.
-In Section~\ref{sec:listener}, we will extend this analysis to listeners performing incremental prediction.
-%incremental prediction.
-
-%We will write $I_t$ as an abbreviation for $\operatorname{I}[w_t, w_0 | w_1, ..., w_{t-1}]$.
-The proposition implies that memory is decreased if $I_t$ decreases quickly as $t \rightarrow \infty$ -- that is, if the contributions of long-term dependencies in the process are small.
-In particular, memory load can only be finite if $I_t$ decreases fast enough for the infinite sum to converge to a finite value.
-%This confirms the intuition that finiteness of memory entails that the contribution of long-term
-
-%\paragraph{Discussion}
-%We have treated the process $(w_t)_t$ as a discrete-time process whose time steps correspond to words, but this is immaterial to the analysis.
-%The analysis is not even restricted to discrete timesteps: We can replace thes sum in Proposition~\ref{prop:suboptimal} with an integral to get a continuous-time version.
-
-We illustrate Proposition~\ref{prop:lower-bound} in Figure~\ref{fig:basic}.
-We consider two processes A and B, where $I_t := 5t^{-1.5}$ for $A$ and $I_t := 3.5 t^{-2.5}$ for $B$.
-The curves of $I_t$, as a function of the distance $t$, are shown in Figure~\ref{fig:basic} (left).
-In both cases, $I_t$ converges to zero as $t$ grows to infinity. 
-However, $I_t$ decays more quickly for Process A (red).
-This means that predictive information about an observation is concentrated more strongly in the recent past.
-In Figure~\ref{fig:basic} (right), we show $t\cdot I_t$ as a function of $t$.
-Note that the area under the curve is equal to (\ref{eq:memory-bound}).
-This area is smaller for the red process, as $I_t$ decays more quickly there.
-
 \section{Experiment 1: Memory and Dependency Length}
@@
 
-\section{Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
+\section{Experiment 2: Large-Scale Evidence that natural language optimize Memory-Surprisal Tradeoff}
 
 \input{main-experiment.tex}
+
+\section{Experiment 3: Controlling for Order Freedom}
+
+\input{freedom-control.tex}
 
@@
 
-\appendix
-
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\caption{Histogram}\label{fig:hist-real}
-\end{figure}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Training & Held-Out & 	Language & Training & Held-Out\\ \hline
-\input{tables/corpusSizes.tex}
-\end{longtable}
-	\caption{Languages, with the number of training and held-out sentences available.}\label{tab:corpora}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|ll||l|llllllllllllll}
-	Language & Base. & Real & Language & Base. & Real \\ \hline
-\input{tables/samplesNumber.tex}
-\end{longtable}
-	\caption{Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{longtable}{l|lll||l|lllllllllllllll}
-	Language & Mean & Lower & Upper & Language & Mean & Lower & Upper \\ \hline
-\input{tables/boot_g_REAL.tex}
-\end{longtable}
-	\caption{Bootstrapped estimates for $G$.}\label{tab:boot-g}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_0.tex}
-\end{longtable}
-	\caption{Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$).
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_1.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_2.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_3.tex}
-\end{longtable}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_0.tex}
-\end{tabular}
-	\caption{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_0.tex}
-%\end{tabular}
-%	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-%\end{table}
-%
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_1.tex}
-%\end{tabular}
-%	\caption{Quantiles (part 2)}
-%\end{table}
-%
-%\begin{table}
-%\begin{tabular}{cccccccccccccccccc}
-%\input{tables/quantiles_REAL_2.tex}
-%\end{tabular}
-%	\caption{Quantiles (part 3)}
-%\end{table}
-%
-%
-%
-
-\begin{table}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{tables/samplesNumber_ground.tex}
-\end{tabular}
-	\caption{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_0.tex}
-\end{tabular}
-	\caption{Experiment 3.
-Medians: For each memory budget, we provide the median surprisal for real and random languages.
-Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median.
-Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_0.tex}
-\end{tabular}
-	\caption{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
-Lower values indicate lower surprisal compared to baselines.
-Solid lines indicate sample means.
-Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_1.tex}
-\end{tabular}
-	\caption{Median Differences (Part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_2.tex}
-\end{tabular}
-	\caption{Median Differences (Part 3)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_3.tex}
-\end{tabular}
-	\caption{Median Differences (Part 4)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
-Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
-At five evenly spaced memory levels, we provide a p-value for the null hypothesis that the actual population mean is $0.5$ or less.
-Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
-
-\section{N-Gram Models}
-
-\input{control-ngrams.tex}
-
-%-- English, Korean, Russian
-%-- UD$\_$Polish-LFG (released in 2.2, not included in original experiment) (13,744 sentences)
-%-- character-level Russian
-%\section{Character-Level Modeling}
-%\section{Non-UD Dependency Treebanks}
-%- other treebanks
-%-- spoken Japanese (T{\"u}ba-J/S)
-%-- another Vietnamese dependency treebank \citep{nguyen-bktreebank:-2017} (5,639 sentences)
-%-- another Chinese dependency treebank LDC2012T05
-%Due to the sizes of these treebanks, can also do experiment with full word forms.
-%
-%
-%\section{Constituency Treebank}
-%
-%-- Penn treebank \citep{marcus-building-1993}
-%
-%-- spoken English (T{\"u}ba-E/S)
-%
-%-- spoken German (T{\"u}ba-D/S)
-%
-%-- Chinese treebank \citep{xue-chinese-2013}
-
-\section{Formal Analysis and Proofs}
-\input{formal-analysis.tex}
-
 \end{document}

### 2020-01-16 code/preparePaper/output/medians_small.tex
code

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2020-01-16 code/analysis/visualize_neural/plot_wordsOnly_medians_onlyREAL.R

@@ -6,76 +6,77 @@
 
 # Plots medians with confidence intervals
 
-fullData = read.csv("../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
+fullData = read.csv("../../../results/tradeoff/listener-curve-ci-median.tsv", sep="\t") %>% filter(Type != "GROUND")
 
 memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     library(tidyr)
     library(dplyr)
     library(ggplot2)
-    dataL = read.csv(paste("../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+    dataL = read.csv(paste("../../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
-    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")
-    ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf", sep=""))
+    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")    +  theme(axis.title.x=element_blank(),
+        axis.title.y=element_blank(), axis.text = element_text(size=20))
[…]

### 2020-01-16 code/preparePaper/medianTables_Small.py

@@ -1,6 +1,6 @@
 # ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
 
-with open("../../ud_languages.txt", "r") as inFile:
+with open("../ud_languages.txt", "r") as inFile:
    languages = inFile.read().strip().split("\n")
 
 languages = sorted(list(set(languages)))
@@ -45,7 +45,7 @@ for language in languages:
 #   line = languageKey[language]
    components = [language.replace("_"," ").replace("-Adap", "")]
    #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
-   components.append( "\includegraphics[width=0.1\\textwidth]{neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}" )
+   components.append( "\includegraphics[width=0.1\\textwidth]{../code/analysis/visualize_neural/figures/"+language+"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf}" )
    entries.append(components)
 
 ROWS_PER_PART = 7
@@ -66,7 +66,7 @@ if True:
        outputRows.append( " \\\\ ")
 
 part=0
-with open("../../medians_small.tex", "w") as outFile:
+with open("output/medians_small.tex", "w") as outFile:
       for row in range(len(outputRows)):
 #          if COLUMNS*(row) >= len(outputRows):
  #             break

### 2020-01-16 code/preparePaper/output/medians_small_mle.tex
code

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab.pdf}
 \\

### 2020-01-16 code/analysis/visualize_neural/plot_wordsOnly_medians_onlyREAL.R

@@ -1,6 +1,6 @@
-    library(tidyr)
-    library(dplyr)
-    library(ggplot2)
+library(tidyr)
+library(dplyr)
+library(ggplot2)
 
 # Plots CIs for the quantile
 
@@ -15,8 +15,7 @@ memListenerSurpPlot_onlyWordForms_boundedVocab = function(language) {
     dataL = read.csv(paste("../../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
     UnigramCE = mean(dataL$UnigramCE)
     data = fullData %>% filter(Language == language)
-    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")    +  theme(axis.title.x=element_blank(),
-        axis.title.y=element_blank(), axis.text = element_text(size=20))
+    plot = ggplot(data, aes(x=Memory, y=UnigramCE-MedianEmpirical, fill=Type, color=Type)) + geom_line(size=2)+ theme_classic() + theme(legend.position="none") + geom_line(aes(x=Memory, y=UnigramCE-MedianLower), linetype="dashed") + geom_line(aes(x=Memory, y=UnigramCE-MedianUpper), linetype="dashed")    +  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), axis.text = element_text(size=20))
     ggsave(plot, file=paste("figures/",language,"-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab_REAL.pdf", sep=""), height=4, width=4)
     return(plot)
 }

### 2020-01-16 code/preparePaper/output/quantiles_small.tex

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
 \\

### 2020-01-21 code/preparePaper/output/quantiles_small.tex
code

@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}
+\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}
  \\

### 2020-01-21 code/preparePaper/quantileTables_Small.py

@@ -45,7 +45,8 @@ for language in languages:
 #   line = languageKey[language]
    components = [language.replace("_"," ").replace("-Adap", "")]
    #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
-   components.append( "\includegraphics[width=0.1\\textwidth]{../code/analysis/visualize_neural/figures/"+language+"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_noAssumption.pdf}" )
+   components.append( "\includegraphics[width=0.1\\textwidth]{../code/analysis/visualize_neural/figures/"+language+"-listener-surprisal-memory-QUANTILES_onlyWordForms_boundedVocab_REAL-equalOrBigger.pdf}" )
+
 
    entries.append(components)
 

### 2020-02-05 writeup/si.tex

@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{../code/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+\includegraphics[width=0.5\textwidth]{../code/analysis/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
 \captionof{figure}{Histogram}\label{fig:hist-real}
@@
 As described above, we assume
+\begin{equation}
+	m_t = M(m_{t-1}, w_{t-1})
+\end{equation}
+\footnote{Alternatively we could admit nondeterministic memory encodings, and require
 \begin{equation}\label{eq:listener-markov}
@@
 \end{equation}
-that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.
+that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.}
 As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
@@
 
-We first show a lemma:
-
-\begin{lemma}
-For any positive integer $t$, the following inequality holds:
+%NOTE the proof could be made simpler by taking $m_t$ to be a deterministic function of $x_t$, $m_{t-1}$, rather than talking about conditional independence
+%
+%We first show a lemma:
+%
+%\begin{lemma}
+%For any positive integer $t$, the following inequality holds:
+%\begin{equation}
+%H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
+%\end{equation}
+%\end{lemma}
+%
+%\begin{proof}[Proof of the Lemma]
+Because $m_t$ is determined by $(w_{1 \dots t-1}, m_1)$:
+\begin{equation}
+	m_t = M(m_{t-1}, w_{t-1}) = M(M(m_{t-2}, w_{t-2}), w_{t-1}) = M(M(M(m_{t-3}, w_{t-3}), w_{t-2}), w_{t-1}) = \dots
+\end{equation}
+the Data Processing inequality entails the following inequality for every positive integer $t$:
 \begin{equation}
@@
 \end{equation}
-\end{lemma}
-\begin{proof}[Proof of the Lemma]
-	By Bayes' Theorem
-\begin{align*}
-	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
-\end{align*}
-By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
-\begin{align*}
-p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
-\end{align*}
-So we have a Markov chain
-\begin{equation}
-(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
-\end{equation}
-Thus, by the Data Processing Inequality,
-\begin{equation}
-H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
-\end{equation}
-Finally, iteratively applying this inequality, we get:
-\begin{align*}
-H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
-\end{align*}
-\end{proof}
+
+%\end{proof}
+
+%\begin{proof}[Proof of the Lemma]
+%	By Bayes' Theorem
+%\begin{align*}
+%	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
+%\end{align*}
+%By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
+%\begin{align*}
+%p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
+%\end{align*}
+%So we have a Markov chain
+%\begin{equation}
+%(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
+%\end{equation}
+%Thus, by the Data Processing Inequality,
+%\begin{equation}
+%H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
+%\end{equation}
+%Finally, iteratively applying this inequality, we get:
+%\begin{align*}
+%H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
+%\end{align*}
+%\end{proof}
 Plugging this inequality into Equation~\ref{eq:byStation} above:
@@
 & = \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_1] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
-& = \frac{1}{T} \left(I[w_{1\dots T}|w_{\leq 0}] - I[w_{1\dots T}|m_1]\right) 
+& = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, m_1]\right) 
 \end{align*}
-The first term $I[w_{1\dots T}|w_{\leq 0}]$ can be rewritten in terms of $I_t$:
+The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
 \begin{align*}
-I[w_{1\dots T}|w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
+I[w_{1\dots T}, w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
 \end{align*}
@@
 \begin{align*}
-\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}|m_1]\right) 
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}, m_1]\right) 
 \end{align*}
@@
 
-%Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
+Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
+
+In ACT-R, each retrieval operation takes time. If each chunk has a bounded amount of information, then this corresponds to a bottleneck in $H[r_t|m_t]$
+
+Formally:
+
+$m_t = f(x_t, m_{t-1}, r_t)$
+
+$r_t = g(m_{t-1}, s_{t-1})$
+
+$s_t = h(m_{t-1}, s_{t-1})$
 
@@
 \begin{proof}
-The negative surprisal gap is
-$\leq \frac{1}{T}(I[X_1\dots X_T|(M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t) \leq  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) = H[R_t] - \sum_{t>T} I_t$
+	As before using the Data Processing inequality:
+\begin{align*}
+	-(\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]) \leq & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - I[X_1\dots X_T, X_{<1}]\right) \\
+= & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
+	= & \frac{1}{T}\left(I[X_{1\dots T}, M_0] + I[X_{1\dots T}, R_1|M_0] + \dots + I[X_{1\dots T}, R_T|M_0, R_{1 \dots T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
+	\leq & \frac{1}{T}\left(H[M_0] + H[R_1] + \dots + H[R_{T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
+	\leq &  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) \\
+	= & H[R_t] - \sum_{t>T} I_t
+\end{align*}
 
@@
 Consequence: As long as retrievals are more expensive than keeping the same amount in WM, locality is predicted.
+
+\subsection{Results for Language Production}
+
+Here we show results linking memory and locality in production.
+
+First, we consider a setting in which a speaker produces sentences with bounded memory, and analyze the deviation of the produced distribution from the actual distribution of the language.
+
+The first setting does not account for the fact that language is produced aiming for some communicative goal.
+We therefore now assume that the speaker has a communicative goal $G$ in mind.
+This goal $G$ stays constant during production process for a sentence, and we count how much memory is needed in addition to the goal $G$.
+We assume that there is a distribution of sentences expressing goals $G$:
+\begin{equation}
+	P(sentence|G)
+\end{equation}
+and assume that the speaker aims to match this distribution
+\begin{equation}
+\mathbb{E}_G[D_{KL}((language|G)||(produced|G))]
+\end{equation}
+We can analyze this model by adding conditioning w.r.t. $G$ throughout the analysis of the previous case.
+Specifically, we need $I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]$.
+
+Take $I_t$ conditioned on $G$: only count statistical dependencies to the degree that they are not redundant with the goal
 

### 2020-02-07 writeup/si.tex
code

@@
 We describe an information-theoretic formalization of a model that has both a limited WM and an unlimited STM.
-
 We consider a model that maintains both a small working memory $m_t$ and an unlimited short-term memory $s_t$.
-
 Predictions are made based on working memory $m_t$, incurring surprisal $H[w_t|m_t]$.
@@
 We model this using a variable $r_t$ representing the information that is retrieved from $s_t$.
+Formally, the memory state is determined not just by the input $x_t$ and the previous memory state $m_{t-1}$, but also by the retrieved information:
+\begin{equation}
+	m_t = f(x_t, m_{t-1}, r_t) 
+\end{equation}
+The retrieval operation is jointly determined by working memory and short-term memory:
+\begin{equation}\label{eq:rt}
+	r_t = g(m_{t-1}, s_{t-1}) 
+\end{equation}
+Finally, the short-term memory can incorporate information from the last word and the working memory:
+\begin{equation}
+	s_t = h(x_{t-1}, m_{t-1}, s_{t-1}) 
+\end{equation}
+
+In the model of \cite{lewis-activation-based-2005}, the working memory corresponds to the \emph{buffers}, which together hold a small and fixed number of individual chunks.
+The short-term memory contains an unbounded number of chunks.
+
+In ACT-R, each retrieval operation is initiated through a query constructed based on the current state of the buffers; it returns a chunk from short-term memory that is then placed into a buffer.
+In our formalization, $r_t$ reflects the totality of all retrieval operations that are made during the processing of $x_{t-1}$; they happen after $x_{t-1}$ has been observed but before $x_t$ has.
+Equation~\ref{eq:rt} reflects that these retrieved chunks are determined by working memory and short-term memory.
+
+In ACT-R, the time it takes to process a word is determined primarily by the time spent retrieving chunks, which is determined by the number of retrieval operations and the time it takes to complete each retrieval operation.
+A retrieval operation takes longer if more chunks are similar to the retrieval cue.
+
+If the information content of each chunk is bounded, then a bound on $H[r_t]$ corresponds to a bound on the number of retrieval operations.
+
+The ACT-R model of \cite{lewis-activation-based-2005} does not have an explicit surprisal cost.
+Instead, surprisal effects are interpreted as arising because, in less constraining contexts, the parser is more likely to
 
@@
 Formally:
-
-$m_t = f(x_t, m_{t-1}, r_t)$
-
-$r_t = g(m_{t-1}, s_{t-1})$
-
-$s_t = h(m_{t-1}, s_{t-1})$
 
@@
 Take $I_t$ conditioned on $G$: only count statistical dependencies to the degree that they are not redundant with the goal
+
+\section{Example where window model is not optimal}
 

### 2020-02-07 writeup/control-ngrams.tex
code

@@
-%bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])
-%bounds.append(["gamma", int, 1, 2, 3, 4, 5, 8, 10, 15, 20, 25, 30]) # , 200, 300
-%bounds.append(["delta", int, 0.1, 0.2, 0.5, 1.0 , 2.0, 3.0, 4.0, 5.0, 8.0, 10.0]) #, 1024]) #, 1024]) # 64, 128,
-%bounds.append(["cutoff", int, 2,3,4,5,6,7,8,9,10]) #,7,8,9,10]) #, 1024]) #, 1024]) # 64, 128,
-%
-
+\subsection{Method}
 We use a version of Kneser-Ney Smoothing.
-For a sequence $w_1\dots w_k$, $N(w_{1\dots k})$ is the number of times $w_{1\dots k}$ occurs in the training set.
+For a sequence $w_1\dots w_k$, let $N(w_{1\dots k})$ be the number of times $w_{1\dots k}$ occurs in the training set.
 The unigram probabilities are estimated as
@@
 Let $\gamma > 0$ be a hyperparameter.
-If $N(w_{0 \dots t-1}) < \gamma$, set $p_t(w_t|w_{0 \dots t-1}) := p_{t-1}(w_t|w_{1\dots t-1})$.
-Otherwise, we interpolate between $n$-th order and lower-order estimates:
+If $N(w_{0 \dots t-1}) < \gamma$, set
+\begin{equation}
+	p_t(w_t|w_{0 \dots t-1}) := p_{t-1}(w_t|w_{1\dots t-1})
+\end{equation}
+Otherwise, we interpolate between $t$-th order and lower-order estimates:
 \begin{equation}
@@
 where $\alpha \in [0,1]$ is also a hyperparameter.
+(CITE) show that this definition results in a well-defined probability distribution, i.e., $\sum_{w \in V} p_t(w|w_{0 \dots t-1}) = 1$.
+
+%Note that this definition guarantees $\sum_{w \in V} p_t(w|w_{0 \dots t-1}) = 1$, because
+%
+%$\sum_{w \in V} \operatorname{max}(N(w_{0\dots t-1}w) - \alpha, 0.0) + \alpha \cdot \#\{w' : N(w_{0 \dots t-1}w') > 0\} \cdot p_{t-1}(w|w_{1\dots t-1})$
+%
+%$\sum_{w \in V : N(w_{0\dots t-1} w) > 0} (N(w_{0\dots t}) - \alpha) + \alpha \cdot \sum_{w \in V} \#\{w' : N(w_{0 \dots t-1}w') > 0\} \cdot p_{t-1}(w|w_{1\dots t-1})$
+%
+%$\sum_{w \in V : N(w_{0\dots t-1} w) > 0} (N(w_{0\dots t}) - \alpha) + \alpha \cdot \#\{w' : N(w_{0 \dots t-1}w') > 0\}$
+%
+%$\sum_{w \in V} : N(w_{0\dots t-1} w) = N(w_{0\dots t-1})$
 
@@
 
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
+\subsection{Results}
+
+\begin{center}
+\begin{longtable}{ccccccccccccccclll}
 \input{tables/medians_ngrams_0.tex}
-\end{tabular}
-	\caption{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
+\input{tables/medians_ngrams_1.tex}
+\input{tables/medians_ngrams_2.tex}
+\input{tables/medians_ngrams_3.tex}
+\end{longtable}
+	\captionof{figure}{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
 Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians_ngrams}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_1.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_2.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_ngrams_3.tex}
-\end{tabular}
-	\caption{Medians (cont.)}
-\end{table}
+\end{center}
 
@@
 
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
+\begin{center}
+\begin{longtable}{cccccccccccccccccc}
 \input{tables/quantiles_REAL_NGRAMS_0.tex}
-\end{tabular}
-	\caption{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+\input{tables/quantiles_REAL_NGRAMS_1.tex}
+\input{tables/quantiles_REAL_NGRAMS_2.tex}
+\end{longtable}
+	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
 Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{table}
+\end{center}
 
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_REAL_NGRAMS_1.tex}
-\end{tabular}
-	\caption{Quantiles (part 2)}
-\end{table}
-
-\begin{table}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_REAL_NGRAMS_2.tex}
-\end{tabular}
-	\caption{Quantiles (part 3)}
-\end{table}
+%bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])
+%bounds.append(["gamma", int, 1, 2, 3, 4, 5, 8, 10, 15, 20, 25, 30]) # , 200, 300
+%bounds.append(["delta", int, 0.1, 0.2, 0.5, 1.0 , 2.0, 3.0, 4.0, 5.0, 8.0, 10.0]) #, 1024]) #, 1024]) # 64, 128,
+%bounds.append(["cutoff", int, 2,3,4,5,6,7,8,9,10]) #,7,8,9,10]) #, 1024]) #, 1024]) # 64, 128,
+%

### 2020-02-07 writeup/si.tex
code

@@
 \paragraph{Ingredient 2: Flow of Information}
-We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
-As the speaker's utterance unfolds, the listener maintains a memory state $m_t$.
-
+%We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
+%As the speaker's utterance unfolds, the listener maintains a memory state $m_t$.
 There are no assumptions about the memory architecture and the nature of its computations.
 We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$.
+At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$:
+\begin{equation}
+	m_t = M(m_{t-1}, w_t)
+\end{equation}
 As a consequence, $m_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $m_{t-1}$.
-This is formalized as a statement about conditional probabilities:
-	\begin{equation}\label{eq:listener-markov}
-		p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
-	\end{equation}
-This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
+%This is formalized as a statement about conditional probabilities:
+%\begin{equation}\label{eq:listener-markov}
+%p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
+%\end{equation}
+%This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
 As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
@@
 Plugging this inequality into Equation~\ref{eq:byStation} above:
-\begin{align*}
+\begin{align}\label{eq:plugged}
 \operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t-1}, m_1] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
@@
 & = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, m_1]\right) 
-\end{align*}
+\end{align}
 The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
@@
 
-Here we show that our information-theoretic analysis is compatible with models placing the main bottleneck in the difficulty of retrieval.
-We describe an information-theoretic formalization of a model that has both a limited WM and an unlimited STM.
-We consider a model that maintains both a small working memory $m_t$ and an unlimited short-term memory $s_t$.
+Here we show that our information-theoretic analysis is compatible with models placing the main bottleneck in the difficulty of retrieval \citep{mcelree2000sentence,lewis-activation-based-2005,nicenboim2018models,vasishth2019computational}.
+We extend our model of memory in incremental prediction to capture key aspects of the models described by \citet{lewis-activation-based-2005,nicenboim2018models,vasishth2019computational}.
+
+The ACT-R model of \cite{lewis-activation-based-2005} assumes a small working memory consisting of \emph{buffers} and a \emph{control state}, which together hold a small and fixed number of individual \emph{chunks}.
+It also assumes a large short-term memory that contains an unbounded number of chunks.
+This large memory store is accessed via \emph{cue-based retrieval}: a query is constructed based on the current state of the buffers; a chunk that matches this query is then selected from the memory storage and placed into one of the buffers.
+
+\paragraph{Formal Model}
+We extend our information-theoretic analysis by considering a model that maintains both a small working memory $m_t$ -- corresponding to the buffers -- and an unlimited short-term memory $s_t$.
 Predictions are made based on working memory $m_t$, incurring surprisal $H[w_t|m_t]$.
-In each time step, there is some amount of communication between $m_t$ and $s_t$, corresponding to retrieval operations.
+When processing a word $x_t$, there is some amount of communication between $m_t$ and $s_t$, corresponding to retrieval operations.
 We model this using a variable $r_t$ representing the information that is retrieved from $s_t$.
@@
 
-In the model of \cite{lewis-activation-based-2005}, the working memory corresponds to the \emph{buffers}, which together hold a small and fixed number of individual chunks.
-The short-term memory contains an unbounded number of chunks.
-
+\paragraph{Cost of Retrieval}
+%In the ACT-R model of \cite{lewis-activation-based-2005}, the working memory corresponds to the \emph{buffers}, which together hold a small and fixed number of individual chunks.
+%The short-term memory contains an unbounded number of chunks.
 In ACT-R, each retrieval operation is initiated through a query constructed based on the current state of the buffers; it returns a chunk from short-term memory that is then placed into a buffer.
@@
 
-In ACT-R, the time it takes to process a word is determined primarily by the time spent retrieving chunks, which is determined by the number of retrieval operations and the time it takes to complete each retrieval operation.
-A retrieval operation takes longer if more chunks are similar to the retrieval cue.
-
+In the model of \cite{lewis-activation-based-2005}, the time it takes to process a word is determined primarily by the time spent retrieving chunks, which is determined by the number of retrieval operations and the time it takes to complete each retrieval operation.
 If the information content of each chunk is bounded, then a bound on $H[r_t]$ corresponds to a bound on the number of retrieval operations.
 
+In the model of \cite{lewis-activation-based-2005}, a retrieval operation takes longer if more chunks are similar to the retrieval cue, whereas, in the direct-access model \citep{mcelree2000sentence,nicenboim2018models,vasishth2019computational}, retrieval operations take a constant amount of time.
+There is no direct counterpart to differences in retrieval times and similarity-based inhibition as in the activation-based model in our formalization.
+Our formalization thus more closely matches the direct-access model, though it might be possible to incorporate aspects of the activation-based model in our formalization.
+
+\paragraph{Role of Surprisal}
 The ACT-R model of \cite{lewis-activation-based-2005} does not have an explicit surprisal cost.
-Instead, surprisal effects are interpreted as arising because, in less constraining contexts, the parser is more likely to
+Instead, surprisal effects are interpreted as arising because, in less constraining contexts, the parser is more likely to make decisions that then turn out to be incorrect, leading to additional correcting steps.
+%We see this as an algorithmic-level implementation of the justification for surprisal theory provided by \citet{levy2008expectation}:
+We view this as an algorithmic-level implementation of a surprisal cost $H[x_t|m_{t-1}]$:
+If the word $x_t$ is unexpected given the current state of the working memory -- i.e., buffers and control states -- then their current state must provide insufficient information to constrain the actual syntactic state of the sentence, meaning that the parsing steps made to integrate $x_t$ will include more backtracking and correction steps.
+Thus, we argue that cue-based retrieval models predict that the surprisal $- \log P(x_t|m_{t-1})$ will be part of the cost of processing word $x_t$.
 
-This is instantiated by ACT-R.
-We can explicitly explain how this covers the McElree ideas and the Lewis and Vasishth ACT-R model.
+%This is instantiated by ACT-R.
+%We can explicitly explain how this covers the McElree ideas and the Lewis and Vasishth ACT-R model.
+%This model has two bottlenecks:
+%The working memory capacity, which we model as $H[m_t]$, and the amount of information that is added through retrieval, modeled as $H[r_t|m_t]$.
+%Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
+%In ACT-R, each retrieval operation takes time. If each chunk has a bounded amount of information, then this corresponds to a bottleneck in $H[r_t|m_t]$
 
-This model has two bottlenecks:
-The working memory capacity, which we model as $H[m_t]$, and the amount of information that is added through retrieval, modeled as $H[r_t|m_t]$.
-
-Bounding retrieval = bounding the precision and/or content of retrieved items. Explain more how this relates to McElree and ACT-R.
-
-In ACT-R, each retrieval operation takes time. If each chunk has a bounded amount of information, then this corresponds to a bottleneck in $H[r_t|m_t]$
-
-Formally:
+\paragraph{Theoretical Result}
+We now show an extension of our theoretical result in the setting of the retrieval-based model described above.
 
 \begin{thm}
-Assume the average working memory $H[m_t]$ is bounded as $H[L_t] \leq \sum_{t=1}^T t I_t$, and the average amount of retrieval is bounded as $H[R_t] \leq \sum_{t=T+1}^S I_t$ (per word).
-Then $H[w_t|m_t] \geq H[w_t|x_{<t}] + \sum_{t>S} I_t$.
+	Assume the average working memory $\operatorname{H}[m_t]$ is bounded as $\operatorname{H}[L_t] \leq \sum_{t=1}^T t I_t$, and the average amount of retrieved information is bounded as $\operatorname{H}[R_t] \leq \sum_{t=T+1}^S I_t$ (per word).
+	Then $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t|x_{<t}] + \sum_{t>S} I_t$.
 \end{thm}
@@
 \begin{proof}
-	As before using the Data Processing inequality:
+The proof is a generalization of the proof above.
+	As in (\ref{eq:plugged}), the Data Processing Inequality entails:
 	\begin{align*}
 	-(\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]) \leq & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - I[X_1\dots X_T, X_{<1}]\right) \\
+	\end{align*}
+	Now, using the same calculation as above, this can be rewritten as:
+	\begin{align*}
 = & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
 	= & \frac{1}{T}\left(I[X_{1\dots T}, M_0] + I[X_{1\dots T}, R_1|M_0] + \dots + I[X_{1\dots T}, R_T|M_0, R_{1 \dots T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
+	\end{align*}
+	Applying the assumptions about the information contained in working memory and retrievals, we bound this as follows:
+	\begin{align*}
 	\leq & \frac{1}{T}\left(H[M_0] + H[R_1] + \dots + H[R_{T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
 	\leq &  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) \\
-	= & H[R_t] - \sum_{t>T} I_t
+	= & H[R_t] - \sum_{t>T} I_t \\
+		\leq & \sum_{t=T+1}^S I_t - \sum_{t>T} I_t \\
+		= & - \sum_{t>S} I_t
 \end{align*}
@@
 
-We can also think of this as a multi-objective optimization, aiming to minimize $\lambda_1 WM + \lambda_2 Retrieval$ to achieve a given surprisal level.
+\paragraph{Information Locality}
+We now show that this result predicts information locality provided that retrieving information is more expensive than keeping the same amount of information in working memory.
+For this, we formalize the problem of finding an optimal memory strategy as a multi-objective optimization, aiming to minimize
+\begin{equation}
+	\lambda_1 H[m_t] + \lambda_2 H[r_t]
+\end{equation}
+to achieve a given surprisal level.
+What is the optimal division of labor between keeping information in working memory and recovering them through retrieval?
+The problem
+\begin{equation}\label{eq:multi-obj-t}
+	\min_{T} \lambda_1 \sum_{t=1}^T t I_t + \lambda_2 \sum_{t=T+1}^S I_t
+\end{equation}
+has solution $T \approx \frac{\lambda_2}{\lambda_1}$. %\footnote{Can do simple proof using the continuous-$T$-version.}
+This means that, as long as retrievals are more expensive than keeping the same amount of information in working memory (i.e., $\lambda_2 > \lambda_1$), the optimal strategy stores information from recent words in working memory.
+Due to the factor $t$ inside $\sum_{t=1}^T t I_t$, the bound~(\ref{eq:multi-obj-t}) will be reduced when $I_t$ decays faster, i.e., there is strong information locality.
 
-If $\frac{\lambda_2}{\lambda_1} \rightarrow \infty$ (retrievals get more expensive), recover previous model.
+The assumption that retrieving information is more difficult than storing it is reasonable for cue-based retrieval models, as retrieval suffers from similarity-based interference effects due to the unstructured nature of the storage~\citep{lewis-activation-based-2005}.
+A model that maintains no information in its working memory, i.e.
+$H[m_t]=0$, would correspond to a cue-based retrieval model that stores nothing in its buffers and control states, and relies entirely on retrieval to access past information.
+Given the nature of representations assumed in models~\citep{lewis-activation-based-2005}, such a model would seem to be severely restricted in its ability to parse language.
 
-If $\frac{\lambda_2}{\lambda_1} \rightarrow 0$ (retrievals get cheaper), locality effect gets weaker, and disappears in the limit\footnote{(Of course, even in this limit, there might be additional factors that may still favor locality in a specific implementation of memory -- e.g., in ACT-R, decay and interference are less problematic if there is locality.)}
-
-Objective
-$$\min_{T} \lambda_1 \sum_{t=1}^T t I_t + \lambda_2 \sum_{t=T+1}^S I_t$$
-has solution $T \approx \frac{\lambda_2}{\lambda_1}$.\footnote{Can do simple proof using the continuous-$T$-version.}
-
-Consequence: As long as retrievals are more expensive than keeping the same amount in WM, locality is predicted.
+%If $\frac{\lambda_2}{\lambda_1} \rightarrow \infty$ (retrievals get more expensive), recover previous model.
+%If $\frac{\lambda_2}{\lambda_1} \rightarrow 0$ (retrievals get cheaper), locality effect gets weaker, and disappears in the limit\footnote{(Of course, even in this limit, there might be additional factors that may still favor locality in a specific implementation of memory -- e.g., in ACT-R, decay and interference are less problematic if there is locality.)}
 
@@
 \begin{center}
-\begin{tabular}{ccccccccccccccclll}
+\begin{longtable}{ccccccccccccccclll}
 \input{tables/slice-hists_REAL_0.tex}
-\end{tabular}
+\input{tables/slice-hists_REAL_1.tex}
+\input{tables/slice-hists_REAL_2.tex}
+\input{tables/slice-hists_REAL_3.tex}
+\end{longtable}
 	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_1.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_2.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_3.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
 \end{center}
@@
 \begin{center}
-\begin{tabular}{ccccccccccccccclll}
+\begin{longtable}{ccccccccccccccclll}
 \input{tables/medians_0.tex}
-\end{tabular}
+\input{tables/medians_1.tex}
+\input{tables/medians_2.tex}
+\input{tables/medians_3.tex}
+\end{longtable}
 	\captionof{figure}{Experiment 3.
@@
 \begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_1.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_2.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccclll}
-\input{tables/medians_3.tex}
-\end{tabular}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccccll}
+\begin{longtable}{ccccccccccccccccll}
 \input{tables/medianDiff_0.tex}
-\end{tabular}
+\input{tables/medianDiff_1.tex}
+\input{tables/medianDiff_2.tex}
+\input{tables/medianDiff_3.tex}
+\end{longtable}
 	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
@@
 \begin{center}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_1.tex}
-\end{tabular}
-	\captionof{figure}{Median Differences (Part 2)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_2.tex}
-\end{tabular}
-	\captionof{figure}{Median Differences (Part 3)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{ccccccccccccccccll}
-\input{tables/medianDiff_3.tex}
-\end{tabular}
-	\captionof{figure}{Median Differences (Part 4)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{cccccccccccccccccc}
+\begin{longtable}{cccccccccccccccccc}
 \input{tables/quantiles_noAssumption_0.tex}
-\end{tabular}
+\input{tables/quantiles_noAssumption_1.tex}
+\input{tables/quantiles_noAssumption_2.tex}
+\end{longtable}
 	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{center}
-
-\begin{center}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_1.tex}
-\end{tabular}
-	\captionof{figure}{Quantiles (part 2)}
-\end{center}
-
-\begin{center}
-\begin{tabular}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_2.tex}
-\end{tabular}
-	\captionof{figure}{Quantiles (part 3)}
 \end{center}

### 2020-02-08 code/optimization/dlm/optimizeDependencyLength_AllLanguages.py
code

@@ -28,8 +28,8 @@ while len(languages) > 0:
      with open(relevantPath+name, "r") as inFile:
        for line in inFile:
            line = line.split("\t")
-           if line[5] == "obj":
-             dhWeight = float(line[4])
+           if line[2] == "obj":
+             dhWeight = float(line[1])
              if dhWeight < 0:
                 negCount += 1
              elif dhWeight > 0:

### 2020-02-08 code/optimization/evaluate/analyze.R

@@ -1,5 +1,6 @@
 library(dplyr)
 library(tidyr)
+data_dlm = read.csv("/u/scr/mhahn/deps/locality_optimized_dlm/manual_output_funchead_coarse_depl/English_optimizeDependencyLength.py_model_3137767.tsv", sep="\t")
 data_i1 = read.csv("/u/scr/mhahn/deps/locality_optimized_i1/English_optimizeGrammarForI1_6_Long.py_model_876472065.tsv", sep="\t")
 data_neural = read.csv("/u/scr/mhahn/deps/locality_optimized_neural/manual_output_funchead_langmod_coarse_best_ud/English_optimizePredictability.py_model_2194458393.tsv", sep="\t")
 data_ground = read.csv("~/scr/CODE/memory-surprisal/results/manual_output_ground_coarse/English_inferWeightsCrossVariationalAllCorpora_NoPunct_NEWPYTORCH_Coarse.py_model_3723683.tsv", sep="\t")
@@ -7,10 +8,13 @@ data_ground = data_ground %>% rename(CoarseDependency = Dependency)
 
 data_b = merge(data_i1, data_ground, by=c("CoarseDependency"))
 data_n = merge(data_neural, data_ground, by=c("CoarseDependency"))
+data_d = merge(data_dlm, data_ground, by=c("CoarseDependency"))
 
 cor.test(data_b$DistanceWeight, data_b$Distance_Mean_NoPunct)$p.value
 cor.test(data_n$DistanceWeight, data_n$Distance_Mean_NoPunct)$p.value
+cor.test(data_d$DistanceWeight, data_d$Distance_Mean_NoPunct)$p.value
 cor.test(data_b$DH_Weight, data_b$DH_Mean_NoPunct)$p.value
 cor.test(data_n$DH_Weight, data_n$DH_Mean_NoPunct)$p.value
+cor.test(data_d$DH_Weight, data_d$DH_Mean_NoPunct)$p.value
 
 

### 2020-02-08 code/optimization/evaluate/runNgramLocalityAll_ForAllModels.py

@@ -1,0 +1,49 @@
+import sys
+import subprocess
+import os
+import random
+
+language = sys.argv[1]
+
+models = []
+#DLM
+import glob
+
+dlm = glob.glob("/u/scr/mhahn/deps/locality_optimized_dlm/manual_output_funchead_coarse_depl/"+language+"_optimizeDependencyLength.py_model_*.tsv")
+i1  = glob.glob("/u/scr/mhahn/deps/locality_optimized_i1/"+language+"_optimizeGrammarForI1_*.py_model_*.tsv")
+neural = glob.glob("/u/scr/mhahn/deps/locality_optimized_neural/manual_output_funchead_langmod_coarse_best_ud/"+language+"_optimizePredictability_OnlyWords.py_model_*.tsv")
+ground = glob.glob("~/scr/CODE/memory-surprisal/results/manual_output_ground_coarse/"+language+"_inferWeightsCrossVariationalAllCorpora_NoPunct_NEWPYTORCH_Coarse.py_model_*.tsv")
+models = dlm + i1 + neural + ground
+print(models)
+#quit()
+
+
+
+version = "yWithMorphologySequentialStreamDropoutDev_Ngrams_Log.py"
+argumentNames = ["alpha", "gamma", "delta", "cutoff"]
+
+
+BASE_DIR = "/u/scr/mhahn/deps/memory-need-ngrams/"
+
+for model in models:
+
+
+    filenames = ([x for x in os.listdir(BASE_DIR) if x.startswith("search-"+language+"_yWithMo") and len(open(BASE_DIR+x, "r").read().split("\n"))>=30])
+    if len(filenames) == 0:
+       assert False
+#    assert len(filenames) == 1
+    with open(BASE_DIR+filenames[0], "r") as inFile:
+        params = next(inFile).strip().split("\t")[2:]
+        assert len(params) == 4
+        params[-1] = 10
+    params2 = []
+    for i in range(len(params)):
+      params2.append("--"+argumentNames[i])
+      params2.append(params[i])
+    paramsString = " ".join([str(x) for x in params2])
+#    command = ["/u/nlp/anaconda/ubuntu_16/envs/py27-mhahn/bin/python2.7", "yHyperParamSearchGPUs_CorPost_Automated_OnlyWordForms_Slurm_Ngrams.py", language, "1", "2", "NONE", "RANDOM_BY_TYPE", "0.02", "30"]
[…]

### 2020-02-08 code/optimization/evaluate/yWithMorphologySequentialStreamDropoutDev_Ngrams_Log.py

@@ -462,7 +462,7 @@ for k in range(0,args.cutoff):
    devSurprisalTable.append(surprisal)
    print("Surprisal", surprisal, len(itos))
 
-outpath = TARGET_DIR+"/estimates-"+args.language+"_"+__file__+"_model_"+str(myID)+"_"+args.model+".txt"
+outpath = TARGET_DIR+"/estimates-"+args.language+"_"+__file__+"_model_"+str(myID)+"_"+args.model.split("/")[-1]+".txt"
 print(outpath)
 with open(outpath, "w") as outFile:
          print >> outFile, " ".join(sys.argv)

### 2020-02-08 code/optimization/yHyperParamSearchGPUs_CorPost_Automated_OnlyWordForms_Slurm_OptimizeForI1_5.py

@@ -183,9 +183,13 @@ DIRPATH = "/u/scr/mhahn/deps/locality_optimized_i1/"
 def getResult(i):
 #   return theirXPs[i][0]
    if runningProcesses[i].poll() is not None:
+     try:
       with open(DIRPATH+"/report_optimizeGrammarForI1_6_EvaluateGrammar.py_"+language+"_"+version+"_model_"+str(theirIDs[i])+".tsv", "r") as inFile:
          loss = min(map(float, next(inFile).strip().split(" ")))
          return loss
+     except IOError:
+         print("Could not open file")
+         return 10
    else:
       return None 
 

### 2020-02-08 code/preparePaper/output/auc_small.tex

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Afrikaans-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Amharic-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Arabic-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Armenian-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Bambara-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Basque-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Breton-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Bulgarian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Buryat-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Cantonese-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Catalan-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Chinese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Croatian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Czech-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Danish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Dutch-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/English-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Erzya-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Estonian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Faroese-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Finnish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/French-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/German-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Greek-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hebrew-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hindi-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hungarian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Indonesian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Italian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Japanese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Kazakh-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Korean-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Kurmanji-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Latvian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Maltese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Naija-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/North_Sami-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Norwegian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Persian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Polish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Portuguese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Romanian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Russian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Serbian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Slovak-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Slovenian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Spanish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Swedish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Thai-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Turkish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Ukrainian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Urdu-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Uyghur-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Vietnamese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab_REAL.pdf}
 \\

### 2020-02-08 writeup/si.tex

@@
 The first term $I[w_{1\dots T}, w_{\leq 0}]$ can be rewritten in terms of $I_t$:
-\begin{align*}
+\begin{align}\label{eq:i-expanded}
 I[w_{1\dots T}, w_{\leq 0}] &= \sum_{i=1}^T \sum_{j=-1}^{-\infty} I[w_i, w_j | w_{j+1}...w_{i-1}] = \sum_{t=1}^T t I_t + T \sum_{t > T} I_t
-\end{align*}
+\end{align}
 Therefore
@@
 It also assumes a large short-term memory that contains an unbounded number of chunks.
-This large memory store is accessed via \emph{cue-based retrieval}: a query is constructed based on the current state of the buffers; a chunk that matches this query is then selected from the memory storage and placed into one of the buffers.
+This large memory store is accessed via \emph{cue-based retrieval}: a query is constructed based on the current state of the buffers and the control state; a chunk that matches this query is then selected from the memory storage and placed into one of the buffers.
 
 \paragraph{Formal Model}
-We extend our information-theoretic analysis by considering a model that maintains both a small working memory $m_t$ -- corresponding to the buffers -- and an unlimited short-term memory $s_t$.
+We extend our information-theoretic analysis by considering a model that maintains both a small working memory $m_t$ -- corresponding to the buffers and the control state -- and an unlimited short-term memory $s_t$.
 Predictions are made based on working memory $m_t$, incurring surprisal $H[w_t|m_t]$.
@@
 We model this using a variable $r_t$ representing the information that is retrieved from $s_t$.
-Formally, the memory state is determined not just by the input $x_t$ and the previous memory state $m_{t-1}$, but also by the retrieved information:
+In our formalization, $r_t$ reflects the totality of all retrieval operations that are made during the processing of $x_{t-1}$; they happen after $x_{t-1}$ has been observed but before $x_t$ has.
+
+The working memory state is determined not just by the input $x_t$ and the previous working memory state $m_{t-1}$, but also by the retrieved information:
 \begin{equation}
@@
 \end{equation}
-The retrieval operation is jointly determined by working memory and short-term memory:
+The retrieval operation is jointly determined by working memory, short-term memory, and the previous word:
 \begin{equation}\label{eq:rt}
-	r_t = g(m_{t-1}, s_{t-1}) 
+	r_t = g(x_{t-1}, m_{t-1}, s_{t-1}) 
 \end{equation}
-Finally, the short-term memory can incorporate information from the last word and the working memory:
+Finally, the short-term memory can incorporate any -- possibly all -- information from the last word and the working memory:
 \begin{equation}
@@
 \end{equation}
+While $s_t$ is unconstrained, there are constraints on the capacity of working memory $\operatorname{H}[m_t]$ and the amount of retrieved information $\operatorname{H}[r_t]$.
+Placing a bound on $\operatorname{H}[m_t]$ reflects the fact that the buffers can only hold a small and fixed number of chunks \citep{lewis-activation-based-2005}.
 
 \paragraph{Cost of Retrieval}
-%In the ACT-R model of \cite{lewis-activation-based-2005}, the working memory corresponds to the \emph{buffers}, which together hold a small and fixed number of individual chunks.
-%The short-term memory contains an unbounded number of chunks.
-In ACT-R, each retrieval operation is initiated through a query constructed based on the current state of the buffers; it returns a chunk from short-term memory that is then placed into a buffer.
-In our formalization, $r_t$ reflects the totality of all retrieval operations that are made during the processing of $x_{t-1}$; they happen after $x_{t-1}$ has been observed but before $x_t$ has.
-Equation~\ref{eq:rt} reflects that these retrieved chunks are determined by working memory and short-term memory.
-
+%In ACT-R, each retrieval operation is initiated through a query constructed based on the current state of the buffers; it returns a chunk from short-term memory that is then placed into a buffer.
+%Equation~\ref{eq:rt} reflects that these retrieved chunks are determined by working memory and short-term memory.
 In the model of \cite{lewis-activation-based-2005}, the time it takes to process a word is determined primarily by the time spent retrieving chunks, which is determined by the number of retrieval operations and the time it takes to complete each retrieval operation.
@@
 We view this as an algorithmic-level implementation of a surprisal cost $H[x_t|m_{t-1}]$:
-If the word $x_t$ is unexpected given the current state of the working memory -- i.e., buffers and control states -- then their current state must provide insufficient information to constrain the actual syntactic state of the sentence, meaning that the parsing steps made to integrate $x_t$ will include more backtracking and correction steps.
+If the word $x_t$ is unexpected given the current state of the working memory -- i.e., buffers and control states -- then their current state must provide insufficient information to constrain the actual syntactic state of the sentence, meaning that the parsing steps made to integrate $x_t$ are likely to include more backtracking and correction steps.
 Thus, we argue that cue-based retrieval models predict that the surprisal $- \log P(x_t|m_{t-1})$ will be part of the cost of processing word $x_t$.
@@
 \begin{thm}
-	Assume the average working memory $\operatorname{H}[m_t]$ is bounded as $\operatorname{H}[L_t] \leq \sum_{t=1}^T t I_t$, and the average amount of retrieved information is bounded as $\operatorname{H}[R_t] \leq \sum_{t=T+1}^S I_t$ (per word).
-	Then $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t|x_{<t}] + \sum_{t>S} I_t$.
+Let $0 < S \leq T$ be positive integers such that the average working memory cost $\operatorname{H}[m_t]$ is bounded as
+	\begin{equation}
+		\operatorname{H}[m_t] \leq \sum_{t=1}^T t I_t
+	\end{equation}
+	and the average amount of retrieved information is bounded as
+	\begin{equation}
+		\operatorname{H}[r_t] \leq \sum_{t=T+1}^S I_t
+	\end{equation} %(per word).
+	Then the surprisal cost is lower-bounded as
+	\begin{equation}
+		\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t|x_{<t}] + \sum_{t>S} I_t
+	\end{equation}
 \end{thm}
@@
 The proof is a generalization of the proof above.
-	As in (\ref{eq:plugged}), the Data Processing Inequality entails:
+	For any positive integer $t$, $m_t$ is determined by $w_{1\dots t}, m_0, r_0, \dots, r_t$.
+	Therefore, the Data Processing Inequality entails:
+	\begin{equation}
+		\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t|w_{1\dots t}, m_0, r_0, \dots, r_t]
+	\end{equation}
+	As in~(\ref{eq:plugged}), this leads to
+\begin{align}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \sum_{t=1}^T ( \operatorname{H}[w_t|w_{1\dots t}, m_0, r_0, \dots, r_t] - \operatorname{H}[w_t | w_{1\dots t-1}, w_{\leq 0}]  )    \\
+& \geq \frac{1}{T} \left(\operatorname{H}[w_{1\dots T} | m_0, r_0, \dots, r_T] - \operatorname{H}[w_{1\dots T} | w_{\leq 0}]\right)  \\
+	& = \frac{1}{T} \left(I[w_{1\dots T}, w_{\leq 0}] - I[w_{1\dots T}, (m_0, r_0, \dots, r_T)]\right) 
+\end{align}
+	Now, using the calculation from (\ref{eq:i-expanded}), this can be rewritten as:
 	\begin{align*}
-	-(\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]) \leq & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - I[X_1\dots X_T, X_{<1}]\right) \\
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]= & \frac{1}{T}\left(\sum_{t=1}^T t I_t + T \sum_{t>T} I_t - I[X_1\dots X_T, (M_0, R_1, ..., R_T)]\right) \\
+		= & \frac{1}{T}\left(\sum_{t=1}^T t I_t + T \sum_{t>T} I_t - I[X_{1\dots T}, M_0] - \sum_{t=1}^T I[X_{1\dots T}, R_t|M_0, r_{1\dots t-1}]\right) \\
 	\end{align*}
-	Now, using the same calculation as above, this can be rewritten as:
-	\begin{align*}
-= & \frac{1}{T}\left(I[X_1\dots X_T, (M_0, R_1, ..., R_T)] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
-	= & \frac{1}{T}\left(I[X_{1\dots T}, M_0] + I[X_{1\dots T}, R_1|M_0] + \dots + I[X_{1\dots T}, R_T|M_0, R_{1 \dots T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
-	\end{align*}
-	Applying the assumptions about the information contained in working memory and retrievals, we bound this as follows:
-	\begin{align*}
-	\leq & \frac{1}{T}\left(H[M_0] + H[R_1] + \dots + H[R_{T-1}] - \sum_{t=1}^T t I_t - T \sum_{t>T} I_t\right) \\
-	\leq &  \frac{1}{T}(T\cdot H[R_t] - T \sum_{t>T} I_t) \\
-	= & H[R_t] - \sum_{t>T} I_t \\
-		\leq & \sum_{t=T+1}^S I_t - \sum_{t>T} I_t \\
-		= & - \sum_{t>S} I_t
-\end{align*}
+	Due to the inequalities $I[X_{1\dots T}, M_0] \leq \operatorname{H}[M_0]$ and $\operatorname{I}[X_{1\dots T}, R_t|M_0, r_{1\dots t-1}] \leq \operatorname{H}[R_t]$, this can be bounded as
+	\begin{align}
+\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]		\geq & \frac{1}{T}\left(\sum_{t=1}^T t I_t  + T \sum_{t>T} I_t-H[M_0] - \sum_{t=1}^T H[R_t]\right) \\
+	\end{align}
+	Finally, this reduces as
+	\begin{align}
+	\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]		\geq &  \frac{1}{T}(T \sum_{t>T} I_t - T\cdot H[R_t]) \\
+	= & \sum_{t>T} I_t- H[R_t]  \\
+		\geq & \sum_{t>T} I_t - \sum_{t=T+1}^S I_t \\
+		= &  \sum_{t>S} I_t
+\end{align}
 
@@
 \end{equation}
-to achieve a given surprisal level.
-What is the optimal division of labor between keeping information in working memory and recovering them through retrieval?
+to achieve a given surprisal level, for some setting of $\lambda_1, \lambda_2 > 0$ describing the relative cost of storage and retrieval.
+What is the optimal division of labor between keeping information in working memory and recovering it through retrieval?
 The problem
@@
 has solution $T \approx \frac{\lambda_2}{\lambda_1}$. %\footnote{Can do simple proof using the continuous-$T$-version.}
-This means that, as long as retrievals are more expensive than keeping the same amount of information in working memory (i.e., $\lambda_2 > \lambda_1$), the optimal strategy stores information from recent words in working memory.
+This means that, as long as retrievals are more expensive than keeping the same amount of information in working memory (i.e., $\lambda_2 > \lambda_1$), the optimal strategy stores information from the last $T > 1$ words in working memory.
 Due to the factor $t$ inside $\sum_{t=1}^T t I_t$, the bound~(\ref{eq:multi-obj-t}) will be reduced when $I_t$ decays faster, i.e., there is strong information locality.
@@
 \section{Example where window model is not optimal}
+
+Here we provide an example of a stochastic process where a window-based memory encoding is not optimal, but the bound provided by our theorem still holds.
+
+Let $k$ be some positive integer.
+Consider a process
+$x_{t+1} = (v_{t+1}, w_{t+1}, y_{t+1}, z_{t+1})$
+where
+\begin{enumerate}
+	\item The first two components consist of fresh random bits. Formally, $v_{t+1}$ is an independent draw from $Bernoulli(0.5)$, independent from all preceding observations $x_{\leq t}$.
+		Second, let $w_{t+1}$ consist of $2k$ many such independent random bits (so that $H[w_{t+1}] = 2k$)
+	\item The third component \emph{deterministically} copies the first bit from $2k$ steps earlier. Formally, $y_{t+1}$ is equal to the first component of $x_{t-2k+1}$
+	\item The fourth component \emph{stochastically} copies the second part (consisting of $2k$ random bits) from one step earlier.
+Formally, each component $z_{t+1}^{(i)}$ is determined as follows: First take a sample $u_{t+1}^{(i)}$ from $Bernoulli(\frac{1}{4k})$, independent from all preceding observations.
+		If $u_{z+1}^{(i)}=1$, set $z_{t+1}^{(i)}$ to be equal to the second component of $w_{t}^{(i)}$.
+		Otherwise, let $z_{t+1}^{(i)}$ be a fresh draw from $Bernoulli(0.5)$.
+\end{enumerate}
+
+Predicting observations optimally requires taking into account observations from the $2k$ last time steps.
+
+We show that, when approximately predicting with low memory capacities, a window-based approach does \emph{not} in general achieve an optimal memory-surprisal tradeoff.
+
+Consider a model that predicts $x_{t+1}$ from only the last observation $x_t$, i.e., uses a window of length one.
+The only relevant piece of information in this past observation is $w_t$, which stochastically influences $z_{t+1}$.
+Storing this costs $2k$ bit of memory as $w_t$ consists of $2k$ draws from $Bernoulli(0.5)$.
+How much does it reduce the surprisal of $x_{t+1}$?
+Due to the stochastic nature of $z_{t+1}$, it reduces the surprisal only by about $I[x_{t+1}, w_t] = I[z_{t+1}, w_t] < 2k \cdot \frac{1}{2k} = 1$, i.e., surprisal reduction is strictly less than one bit.
+\footnote{We can evaluate $I[z_{t+1}, w_t]$ as follows.
+Set $l = k/4$.
+Write $z, w$ for any of the $2k$ components of $z_{t+1}, w_t$, respectively.
+First, calculate $p(z = 1|w=1) = 1/l + (1-1/l) \frac{1}{2} = 1/(2l) + 1/2 = \frac{1+l}{2l}$ and $p(z = 0|w=1) = (1-1/l) \frac{1}{2} = 1/2 - 1/2l = \frac{l-1}{2l}$.
+Then $I[Z, W] = D_{KL}(p(z|w=1)||p(z)) = \frac{1+l}{2l} \log \frac{\frac{1+l}{2l}}{1/2} + \frac{l-1}{2l} \log \frac{\frac{l-1}{2l}}{1/2}  = \frac{1+l}{2l} \log \frac{1+l}{l} + \frac{l-1}{2l} \log \frac{l-1}{l}  \leq \frac{1+l}{l} \log \frac{1+l}{l} =  (1+1/l) \log (1+1/l)  \leq  (1+1/l) (1/l) = 1/l + 1/l^2 < 2/l = \frac{1}{2k}.$
+%Then $H[Z|w=1] = - \frac{1+k}{2k} \log \frac{1+k}{2k} - \frac{k-1}{2k} \log \frac{k-1}{2k}  = - (1/2 + 1/2k) \log \frac{1+k}{2k} - (1/2 - 1/2k) \log \frac{k-1}{2k}$ %= - \frac{1+k}{2k} \log (1+k) - \frac{k-1}{2k} \log (k-1) + \log 2k = - \frac{1+k}{2k} \log (1+k) - \frac{k-1}{2k} \log (k-1) + \log 2k$.
+}
+
+We show that there is an alternative model that strictly improves on this window-based model:
+Consider a memory encoding model that encodes each of $v_{t-2k+1}, \dots, v_{t}$, which costs $2k$ bits of memory -- as the window-based model did.
+Since $y_{t+1} = v_{t-2k+1}$, this model achieves a surprisal reduction of $H[v_{t-2k+1}] = 1$ bit, strictly more than the window-based model.
+
+This result does not contradict our theorem because the theorem only provides \emph{bounds} across models, which are not necessarily achieved by a given window-based model.
+In fact, for the process described here, no memory encoding function $M$ can exactly achieve the theoretical bound described by the theorem.
 

### 2020-02-13 writeup/si.tex
code

@@
 the Data Processing inequality entails the following inequality for every positive integer $t$:
-\begin{equation}
-	H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
-\end{equation}
 
-%\end{proof}
-
-%\begin{proof}[Proof of the Lemma]
-%	By Bayes' Theorem
-%\begin{align*}
-%	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
-%\end{align*}
-%By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
-%\begin{align*}
-%p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
-%\end{align*}
-%So we have a Markov chain
-%\begin{equation}
-%(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
-%\end{equation}
-%Thus, by the Data Processing Inequality,
-%\begin{equation}
-%H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
-%\end{equation}
-%Finally, iteratively applying this inequality, we get:
-%\begin{align*}
-%H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
-%\end{align*}
-%\end{proof}
 	Plugging this inequality into Equation~\ref{eq:byStation} above:
@@
 
-%Justify linear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+%Justifylinear interpolation: The curve is convex, which is shown by `time-sharing': Use one code $\lambda$ fraction of times, and the other code $1-\lambda$ fraction of times.
+
+\subsection{For nondeterministic encoding functions}
+
+We relax condition (X) to
+\begin{equation}\label{eq:listener-markov-nondeterministic}
+p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
+\end{equation}
+This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.
+
+The one place in the proof where (X) plays a role is the proof of the inequality:
+\begin{equation}
+	H[w_t | m_t] \geq H[w_t|w_{1 \dots t-1}, m_1]
+\end{equation}
+
+We show that this inequality still holds under the relaxed condition (X):
+\begin{proof}
+	By Bayes' Theorem
+\begin{align*}
+	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
+\begin{align*}
+p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
+\end{align*}
+So we have a Markov chain
+\begin{equation}
+(w_t) \rightarrow (m_0, w_{0 \dots t-1})   \rightarrow   (m_1, w_{1 \dots t-1})
+\end{equation}
+Thus, by the Data Processing Inequality,
+\begin{equation}
+H[w_t| w_{1 \dots t-1}, m_{1}] \geq H[w_t|w_{0 \dots t-1}, m_0]
+\end{equation}
+Finally, iteratively applying this inequality, we get:
+\begin{align*}
+H[w_t | m_t] \geq H[w_t| w_{t-1}, m_{t-1}] \geq H[w_t| w_{t-2, t-1}, m_{t-2}] \geq ... \geq H[w_t|w_{1 \dots t-1}, m_1]
+\end{align*}
+\end{proof}
 
@@
 Here we show results linking memory and locality in production.
+We show that results similar to our main theorem hold for the tradeoff between a speaker's memory and the accuracy with which they match the distribution of the language.
 
+\paragraph{Speaker aims to match language distribution}
 First, we consider a setting in which a speaker produces sentences with bounded memory, and analyze the deviation of the produced distribution from the actual distribution of the language.
 
+We consider a speaker who maintains memory representations and incrementally produces based on these representations:
+\begin{equation}
+	p_{speaker}(x_t|X_{<t}) = p(x_t|m_t) \\
+\end{equation}
+We show a tradeoff between the memory capacity $\operatorname{H}[m_t]$ and the KL-divergence between the actual language statistics and the speaker's production distribution:
+\begin{equation}
+D_{KL}(P_{language}||P_{produced})  := \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p_{speaker}(x_t|X_{<t})} \\
+\end{equation}
+\begin{thm}
+If a speaker maintains memory
+	\begin{equation}
+		\operatorname{H}[m_t] \leq \sum_{i=1}^T tI_t
+	\end{equation}
+	then 
+\begin{equation}
+	D_{KL}(P_{language}||P_{produced}) \geq \sum_{t=T+1}^\infty I_t
+\end{equation}
+\end{thm}
+
+While this bound only considers the production of a single word, it immediately entails a bound on the production accuracy for sequences:
+\begin{equation}
+	D_{KL}(P_{language}(X_1\dots X_t|X_{\leq 0})||P_{produced}(X_1\dots X_t|X_{\leq 0}))  = t \cdot D_{KL}(P_{language}(X_1|X_{\leq 0})||P_{produced}(X_1|X_{\leq 0}))
+\end{equation}
+
+\begin{proof}
+First note
+\begin{align}
+D_{KL}(P_{language}||P_{produced}) & = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p_{speaker}(x_t|X_{<t})} \\
+& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log \frac{p(x_t|X_{<t})}{p(x_t|M(X_{<t}))} \\
+& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|X_{<t}) - \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|M(X_{<t})) \\
+	& = \E_{X_{<t}} \sum_{x_t} p(x_t|X_{<t}) \log p(x_t|X_{<t}) + S_M(x_t|X_{<t})
+\end{align}
+In the last line, the first term is a constant independent of $M$.
+
+Then the proof for the listener case transfers to this setting.
+\end{proof}
+
+TODO limitations of this as a model of production
+
+\paragraph{Speaker aims to match, conditional on goal}
 The first setting does not account for the fact that language is produced aiming for some communicative goal.
@@
 We can analyze this model by adding conditioning w.r.t. $G$ throughout the analysis of the previous case.
-Specifically, we need $I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]$.
+Specifically, we need
+\begin{equation}
+I_t^G := I[X_t, X_0|X_1, \dots, X_{t-1}, G]
+\end{equation}
 
 Take $I_t$ conditioned on $G$: only count statistical dependencies to the degree that they are not redundant with the goal
+
+\begin{thm}
+If a speaker maintains memory
+	\begin{equation}
+		\operatorname{H}[m_t] \leq \sum_{i=1}^T tI_t^G
+	\end{equation}
+	then 
+\begin{equation}
+\E_G	D_{KL}(P_{language}(\cdot|G)||P_{produced}(\cdot|G)) \geq \sum_{t=T+1}^\infty I_t^G
+\end{equation}
+\end{thm}
+
+\begin{proof}
+This is entirely analogous to the previous proof.
+\end{proof}
+
+are there conditions under which this is close to $I_t$?
+
+\paragraph{Pragmatic Speaker}
+
+would need an assumption on the density of goals in the space of sequences.
+
+Note 
+\begin{equation}
+	D_{KL}(P_{language}(x_{1\dots t})||P_{produced}(x_{1\dots t}))  := \sum_{x_{1\dots t}} p(x_t|X_{1\dots t}) \log \frac{p(x_t|X_{1\dots t})}{p_{speaker}(x_t|X_{1\dots t})} \geq t D_KL(x_t||...)
+\end{equation}
+
+$H[G|Produced] - H[G|Language]$
+
+\section{Proof of Left-Right Invariance}
+
+Here we show that the bound provided by our theorem is invariant under reversal of the process.
+That is: Given a process $(X_t)_{t \in \mathbb{Z}}$, we define its reverse process $(Y_t)_{t \in \mathbb{Z}}$ by $Y_t := X_t$.
+We claim that the theorem provides the same bounds for the memory-surprisal tradeoff curves.
+To prove this, we note:
+\begin{equation}
+	I[X_t, X_0|X_{1\dots t-1}] = I[Y_{-t}, Y_0|Y_{1-t\dots -1}] = I[Y_0, Y_t|Y_{1\dots t-1}] = I[Y_t, Y_0|Y_{1\dots t-1}]
+\end{equation}
+The first step follows from the definition of $Y$.
+The second step follows from the fact that $X_t$, and thus also $Y_t$, is stationary, and thus adding $t$ to each index in the expression does not change the resulting value.
+The third step uses the fact that mutual information is symmetric.
 
@@
 
+\section{Optimization}
+We cannot optimize for AUC, because we cannot construct an unbiased gradient estimator
+$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s$
+
+Softmax gradient update corresponds to adding $\alpha$ to the target logit and removing $\alpha$ from all other logits.
+Equivalently, add a certain dynamic amount to the logcount of the target and the total logcount.
+
+Counting corresponds to adding $1$ to the target probability.
+
 \bibliographystyle{apalike}

### 2020-02-14 writeup/si.tex

@@
 
-\begin{center}
-\includegraphics[width=0.5\textwidth]{../code/analysis/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
-\captionof{figure}{Histogram}\label{fig:hist-real}
-\end{center}
+%
+%\begin{center}
+%\includegraphics[width=0.5\textwidth]{../code/analysis/visualize_neural/figures/full-REAL-listener-surprisal-memory-HIST_z_byMem_onlyWordForms_boundedVocab.pdf}
+%\captionof{figure}{Histogram}\label{fig:hist-real}
+%\end{center}
+%
+%
 
@@
 
-In this section, we prove the theorem described above.
+In this section, we prove Theorem 1.
 
@@
 We make the assumption that this process is \emph{stationary}.
-Formally, this means that the conditional distribution $P(w_t|w_{<t})$ does not depend on $t$, it only depends on the actual sequence $w_{<t}$.
+Formally, this means that the conditional distribution $P(w_t|w_{<t})$ does not depend on $t$, it only depends on the (semi-infinite) context sequence $w_{<t}$.
 Informally, this says that the process has no `internal clock', and that the statistical rules of the language do not change at the timescale we are interested in.
@@
 
-\begin{figure}
-\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
-	\caption{Illustration of (\ref{eq:listener-markov}).
+%
+%
+%
+%\begin{figure}
+%\includegraphics[width=0.45\textwidth]{figures/markov-condition.png}
+%	\caption{Illustration of (\ref{eq:listener-markov}).
 As the utterance unfolds, the listener maintains a memory state.
 After receiving word $w_t$, the listener computes their new memory state $m_t$ based on the previous memory state $m_{t-1}$ and the new word $w_t$.}\label{fig:listener-markov}
-\end{figure}
+%\end{figure}
+%
 
-\paragraph{Ingredient 2: Flow of Information}
+\paragraph{Ingredient 2: Postulates about Processing}
+The second ingredient consists of the three postulates described in the main paper.
 %We now analyze memory from the perspective of the listener, who needs to maintain information about the past to predict the future.
 %As the speaker's utterance unfolds, the listener maintains a memory state $m_t$.
-There are no assumptions about the memory architecture and the nature of its computations.
-We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
-At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$:
-\begin{equation}
-	m_t = M(m_{t-1}, w_t)
-\end{equation}
-As a consequence, $m_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $m_{t-1}$.
-%This is formalized as a statement about conditional probabilities:
-%\begin{equation}\label{eq:listener-markov}
-%p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
+There are no further assumptions about the memory architecture and the nature of its computations.
+%We only make a basic assumption about the flow of information (Figure~\ref{fig:listener-markov}):
+%At a given point in time, the listener's memory state $m_t$ is determined by the last word $w_t$, and the prior memory state $m_{t-1}$:
+%\begin{equation}
+%	m_t = M(m_{t-1}, w_t)
 %\end{equation}
-%This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
-This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
-\mhahn{For the study of memory in sentence processing, this seems fair. Discuss this more.}
+%As a consequence, $m_t$ contains no information about the process beyond what is contained in the last word observed $w_{t-1}$ and in the memory state before that word was observed $m_{t-1}$.
+%%This is formalized as a statement about conditional probabilities:
+%%\begin{equation}\label{eq:listener-markov}
+%%p(m_1| (w_{t})_{t \in \mathbb{Z}}, m_0)   = p(m_1 | m_0, w_1)
+%%\end{equation}
+%%This says that $m_1$ contains no information about the utterances beyond what is contained in $m_0$ and $w_1$.	
+%As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+%This is a simplification, as the listener could obtain information about the speaker from other sources, such as their common environment (weather, ...).
+%\mhahn{For the study of memory in sentence processing, this seems fair. Discuss this more.}
 
@@
 
-We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
-We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
-As described above, we assume
-\begin{equation}
-	m_t = M(m_{t-1}, w_{t-1})
-\end{equation}
-\footnote{Alternatively we could admit nondeterministic memory encodings, and require
-\begin{equation}\label{eq:listener-markov}
-	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
-\end{equation}
-that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.}
-As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
+%We formalize a language as a stationary stochastic process $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
+%The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+%We denote the listener's memory state at time $t$, after hearing $w_{<t} = ... w_{t-2} w_{t-1}$ by $m_t$.
+%As described above, we assume
+%\begin{equation}
+%	m_t = M(m_{t-1}, w_{t-1})
+%\end{equation}
+%\footnote{Alternatively we could admit nondeterministic memory encodings, and require
+%\begin{equation}\label{eq:listener-markov}
+%	p(m_{t+1}| (w_{t'})_{t' \in \mathbb{Z}}, m_t)   = p(m_{t+1} | m_t, w_{t})
+%\end{equation}
+%that is, $m_{t+1}$ contains no information about the utterances beyond what is contained in $m_t$ and $w_{t}$.}
+%As a consequence, the listener has no knowledge of the speaker's state beyond the information provided in their prior communication.
 
-The average number of bits required to encode this state is $\operatorname{H}[m_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
-As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | m_t]$.
-The difference between the listener's surprisal and optimal surprisal is thus at least $\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.
+%The average number of bits required to encode this state is $\operatorname{H}[m_t]$, which by assumption is at most $\sum_{t=1}^T t I_t$.
+%As the listener's predictions are made on the basis of her memory state, her average surprisal is at least $\operatorname{H}[w_t | m_t]$.
+\begin{proof}
+The difference between the listener's surprisal and optimal surprisal is $\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.\footnote{A listener whose predictions are not optimal given $m_t$ can only incur even higher surprisal.}
 By the assumption of stationarity, we can, for any positive integer $T$, rewrite this expression as
@@
 \end{equation}
-
 %NOTE the proof could be made simpler by taking $m_t$ to be a deterministic function of $x_t$, $m_{t-1}$, rather than talking about conditional independence
@@
 the Data Processing inequality entails the following inequality for every positive integer $t$:
-
+\begin{align}\label{eq:plugged}
+\operatorname{H}[w_t | m_t]& \geq \operatorname{H}[w_t|w_{1\dots t-1}, m_1]
+\end{align}
 Plugging this inequality into Equation~\ref{eq:byStation} above:
@@
 Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
-%\end{proof}
+\end{proof}
 
@@
 
-We relax condition (X) to
+We have been assuming that $m_t$ is a deterministic function of $x_t$ and $m_{t-1}$.
+Here, we show that this assumption can be relaxed to stochastic encoding functions.
+
+We relax condition (X) to the following requirement, for all values of $m_1, (w_{t})_{t \in \mathbb{Z}}, m_0$:
 \begin{equation}\label{eq:listener-markov-nondeterministic}
@@
 
-We show that this inequality still holds under the relaxed condition (X):
+We show that this inequality still holds under the relaxed condition (\ref{eq:listener-markov-nondeterministic}):
 \begin{proof}
@@
 \begin{align*}
-	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1})
+	p(w_t|m_0, m_1, w_{0\dots t-1}) &= \frac{p(m_1|m_0, w_{0\dots t})}{p(m_1|m_0, w_{0\dots t-1})} \cdot p(w_t|m_0, w_{0\dots t-1}) \\
+ &= \frac{p(m_1|m_0, w_{0})}{p(m_1|m_0, w_{0})} \cdot p(w_t|m_0, w_{0\dots t-1}) \\
+ &= p(w_t|m_0, w_{0\dots t-1}) \\
 \end{align*}
-By Equation~\ref{eq:listener-markov}, the quotient on the RHS is equal to $1$, so
-\begin{align*}
-p(w_t|m_0, m_1, w_{0\dots t-1}) = p(w_t|m_0, w_{0\dots t-1})
-\end{align*}
+	where the second equation follows from (\ref{eq:listener-markov-nondeterministic}).
 So we have a Markov chain
@@
 \end{equation}
-Finally, iteratively applying this inequality, we get:
+Finally, iteratively applying this reasoning, we conclude:
 \begin{align*}

### 2020-02-15 code/preparePaper/output/it_small.tex
It plot

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-it_REAL.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-it_REAL.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-it_REAL.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-it_REAL.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-it_REAL.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-it_REAL.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-it_REAL.pdf}
 \\

### 2020-02-15 writeup/control-ngrams.tex
It plot

@@
 % ../../writeup/tables/quantiles_REAL_NGRAMS_0.tex
-
-\begin{center}
-\begin{longtable}{cccccccccccccccccc}
-\input{tables/quantiles_REAL_NGRAMS_0.tex}
-\input{tables/quantiles_REAL_NGRAMS_1.tex}
-\input{tables/quantiles_REAL_NGRAMS_2.tex}
-\end{longtable}
-	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+%
+%\begin{center}
+%\begin{longtable}{cccccccccccccccccc}
+%\input{tables/quantiles_REAL_NGRAMS_0.tex}
+%\input{tables/quantiles_REAL_NGRAMS_1.tex}
+%\input{tables/quantiles_REAL_NGRAMS_2.tex}
+%\end{longtable}
+%	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
 Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{center}
-
+%\end{center}
+%
 %bounds.append(["alpha", float, 0.95, 1.0]) # + [x/20.0 for x in range(15, 21)])

### 2020-02-15 writeup/si.tex
It plot

@@
 
-\subsection{Median Surprisal per Memory Budget}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_0.tex}
-\end{longtable}
-	\captionof{figure}{Medians: For each memory budget, we provide the median surprisal for real and random languages.
+%\subsection{Median Surprisal per Memory Budget}
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/medians_REAL_0.tex}
+%\end{longtable}
+%	\captionof{figure}{Medians: For each memory budget, we provide the median surprisal for real and random languages.
 Solid lines indicate sample medians, dashed lines indicate 95 $\%$ confidence intervals for the population median, dotted lines indicate empirical quantiles ($10\%, 20\%, \dots, 80\%, 90\%$).
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{center}
+%\end{center}
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/medians_REAL_1.tex}
+%\end{longtable}
+%	\captionof{figure}{Medians (cont.)}
+%\end{center}
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/medians_REAL_2.tex}
+%\end{longtable}
+%	\captionof{figure}{Medians (cont.)}
+%\end{center}
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/medians_REAL_3.tex}
+%\end{longtable}
+%	\captionof{figure}{Medians (cont.)}
+%\end{center}
+%
 
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_1.tex}
-\end{longtable}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_2.tex}
-\end{longtable}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_REAL_3.tex}
-\end{longtable}
-	\captionof{figure}{Medians (cont.)}
-\end{center}
-
-\subsection{Surprisal at Maximum Memory}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/slice-hists_REAL_0.tex}
-\input{tables/slice-hists_REAL_1.tex}
-\input{tables/slice-hists_REAL_2.tex}
-\input{tables/slice-hists_REAL_3.tex}
-\end{longtable}
-	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
-\end{center}
+%
+%\subsection{Surprisal at Maximum Memory}
+%
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/slice-hists_REAL_0.tex}
+%\input{tables/slice-hists_REAL_1.tex}
+%\input{tables/slice-hists_REAL_2.tex}
+%\input{tables/slice-hists_REAL_3.tex}
+%\end{longtable}
+%	\captionof{figure}{Histograms: Surprisal, at maximum memory.}\label{tab:slice-hists-real}
+%\end{center}
+%
 
@@
 
-\subsection{Samples Drawn (Experiment 3)}
-
-\begin{center}
-\begin{tabular}{l|ll||l|llllllllllllll}
-	Language & Base. & MLE & Language & Base. & MLE \\ \hline
-\input{tables/samplesNumber_ground.tex}
-\end{tabular}
-	\captionof{figure}{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
-\end{center}
-
-\subsection{Medians (Experiment 3)}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccclll}
-\input{tables/medians_0.tex}
-\input{tables/medians_1.tex}
-\input{tables/medians_2.tex}
-\input{tables/medians_3.tex}
-\end{longtable}
-	\captionof{figure}{Experiment 3.
+%
+%
+%\subsection{Samples Drawn (Experiment 3)}
+%
+%
+%
+%\begin{center}
+%\begin{tabular}{l|ll||l|llllllllllllll}
+%	Language & Base. & MLE & Language & Base. & MLE \\ \hline
+%\input{tables/samplesNumber_ground.tex}
+%\end{tabular}
+%	\captionof{figure}{Experiment 3: Samples drawn per language according to the precision-dependent stopping criterion.}\label{tab:samples}
+%\end{center}
+%
+%
+%
+%\subsection{Medians (Experiment 3)}
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccclll}
+%\input{tables/medians_0.tex}
+%\input{tables/medians_1.tex}
+%\input{tables/medians_2.tex}
+%\input{tables/medians_3.tex}
+%\end{longtable}
+%	\captionof{figure}{Experiment 3.
 Medians: For each memory budget, we provide the median surprisal for real and random languages.
@@
 Green: Random baselines; blue: real language; red: maximum-likelihood grammars fit to real orderings.}\label{tab:medians}
-\end{center}
-
-\begin{center}
-\begin{longtable}{ccccccccccccccccll}
-\input{tables/medianDiff_0.tex}
-\input{tables/medianDiff_1.tex}
-\input{tables/medianDiff_2.tex}
-\input{tables/medianDiff_3.tex}
-\end{longtable}
-	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
+%\end{center}
+%
+%
+%
+%
+%
+%\begin{center}
+%\begin{longtable}{ccccccccccccccccll}
+%\input{tables/medianDiff_0.tex}
+%\input{tables/medianDiff_1.tex}
+%\input{tables/medianDiff_2.tex}
+%\input{tables/medianDiff_3.tex}
+%\end{longtable}
+%	\captionof{figure}{Median Differences between Real and Baseline: For each memory budget, we provide the difference in median surprisal between real languages and random baselines; for real orders (blue) and maximum likelihood grammars (red).
 Lower values indicate lower surprisal compared to baselines.
@@
 Dashed lines indicate 95 $\%$ confidence intervals.}\label{tab:median_diffs}
-\end{center}
-
-\begin{center}
-\begin{longtable}{cccccccccccccccccc}
-\input{tables/quantiles_noAssumption_0.tex}
-\input{tables/quantiles_noAssumption_1.tex}
-\input{tables/quantiles_noAssumption_2.tex}
-\end{longtable}
-	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
+%\end{center}
+%
+%
+%
+%
+%
+%
+%
+%\begin{center}
+%\begin{longtable}{cccccccccccccccccc}
+%\input{tables/quantiles_noAssumption_0.tex}
+%\input{tables/quantiles_noAssumption_1.tex}
+%\input{tables/quantiles_noAssumption_2.tex}
+%\end{longtable}
+%	\captionof{figure}{Quantiles: At a given memory budget, what percentage of the baselines results in higher listener surprisal than the real language?
 Solid curves represent sample means, dashed lines represent 95 \% confidence bounds; dotted lines represent 99.9 \% confidence bounds.
@@
 Confidence bounds and p-values are obtained using an exact nonparametric method (see text).}\label{tab:quantiles}
-\end{center}
+%\end{center}
+%
+%
 
@@
 \section{Optimization}
-We cannot optimize for AUC, because we cannot construct an unbiased gradient estimator
-$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s$
+We cannot optimize for AUC using the method of (CITE), because we cannot construct an unbiased gradient estimator
+$AUC \propto \sum_{t} (\sum_{s \leq t} I_s) tI_t = \sum_{s \leq t}  tI_t I_s = I_1^2 + 2 I_1 I_2 + 2 I_2^2 + \dots$
+
+As a surrogate, we propose to maximize $I_1$.
+This corresponds
+
+It provides an accurate approximation to the AUC if $I_s$ is small for $s > 1$, which holds for the n-gram based estimator.
+
+If $I_s < \epsilon$ for $s > 1$, then 
+\begin{equation}
+	I_1^2 \leq \sum_{1\leq s \leq t \leq T}  tI_t I_s \leq I_1^2 + I_1 \epsilon T^2 + \epsilon^2 T^2
+\end{equation}
+That means, $I_t^2$ is an accurate approximation of the AUC when $\epsilon T^2$ is small.
 

### 2020-02-15 code/analysis/visualize_neural/plot_wordsOnly_It.R

@@ -1,0 +1,31 @@
+library(tidyr)
+library(dplyr)
+library(ggplot2)
+
+# Plots CIs for the quantile
+
+# Plots medians with confidence intervals
+
+fullData = read.csv("../../../results/tradeoff/listener-curve-ci-median.tsv", sep="\t")
+
+languages = read.csv("languages.tsv")$Language
+
+
+for(language in languages) {
+  dataL = read.csv(paste("../../../results/raw/word-level/",language,"_decay_after_tuning_onlyWordForms_boundedVocab.tsv", sep=""), sep="\t")
+  
+  dataL = dataL %>% group_by(Distance, Type) %>% summarise(ConditionalMI = median(ConditionalMI))
+  
+  dataL = dataL %>% filter(Type %in% c("RANDOM_BY_TYPE", "REAL_REAL"))
+  
+  plot = ggplot(dataL %>% filter(Distance < 5), aes(x=Distance, y=ConditionalMI, color=Type)) + geom_line(size=2)
+  plot = plot + theme_bw()
+  plot = plot + theme(axis.title.x=element_blank(), axis.title.y=element_blank(), axis.text = element_text(size=20))
+  plot = plot + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), legend.position="none") 
+  plot = plot + theme(axis.line = element_line(colour = "black"),
+	      panel.grid.major = element_blank(),
+	          panel.grid.minor = element_blank(),
+	          panel.border = element_blank(),
+		      panel.background = element_blank()) 
+  ggsave(plot, file=paste("figures/",language,"-it_REAL.pdf", sep=""), height=4, width=4)
+}

### 2020-02-15 code/preparePaper/siTables_small.py

@@ -1,0 +1,80 @@
+# ./python27 tradeoffPrepareTable_OnlyWordForms_BoundedVocab.py > ../results-table-word-level.tex
+
+with open("../ud_languages.txt", "r") as inFile:
+   languages = inFile.read().strip().split("\n")
+
+languages = sorted(list(set(languages)))
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data[-1]) == 1 and data[-1][0] == "":
+       del data[-1]
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+def h(frame, line, name):
+    return line[frame[0][name]]
+
+
+#with open("tradeoff/effectSize.tsv", "r") as inFile:
+#   effectSize = readTSV(inFile)
+
+#with open("tradeoff/effectSize_diff.tsv", "r") as inFile:
+#   effectSize_diff = readTSV(inFile)
+
+with open("../../results/tradeoff/stats-onlyWordForms-boundedVocab_REAL.tsv", "r") as inFile:
+   stats = readTSV(inFile)
+languageKey_stats = dict([(h(stats, line, "Language"), line) for line in stats[1]])
+
+
+entries = []
+
+for language in languages:
+#   line = languageKey[language]
+   components = [language.replace("_"," ").replace("-Adap", "")]
+   #components.append( "\\multirow{4}{*}{\includegraphics[width=0.25\\textwidth]{neural/figures/"+language+"-entropy-memory.pdf}}")
+
+# user@user-X510UAR:~/memory-surprisal/code/analysis/visualize_neural$ evince figures/Amharic-Adap-it_REAL.pdf 
+
+   components.append( "\includegraphics[width=0.1\\textwidth]{../code/analysis/visualize_neural/figures/"+language+"-it_REAL.pdf}" )
+   entries.append(components)
+
+ROWS_PER_PART = 7
+COLUMNS = 9
+while len(entries) % COLUMNS != 0:
[…]

### 2020-02-16 code/analysis/medians/interpolate/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Interpolate_PCFG.py
code

@@ -1,0 +1,102 @@
+# Better than yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MaxControl.py by modeling the median of REAL
+
+import sys
+
+language = sys.argv[1]
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data) < 2:
+       return (header, [])
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+try:
+  with open("../../../../results/raw/pcfg/"+language+"_pcfg_decay_after_tuning.tsv", "r") as inFile:
+     data = readTSV(inFile)
+except IOError:
+  print >> sys.stderr, ("ERROR nothing for this language? "+language+"    "+"../../../results/raw/pcfg/"+language+"_pcfg_decay_after_tuning.tsv")
+  quit()
+
+import torch
+
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+
+matrix = [[g(data, "ModelID", i), g(data, "Distance", i), g(data, "ConditionalMI", i), g(data, "UnigramCE", i)] for i in range(len(data[1]))]
+
+matrixByType = {}
+misByType = {}
+for i in range(len(data[1])):
+    typ = g(data, "Type", i)
+    if typ not in matrixByType:
+        matrixByType[typ] = []
+    matrixByType[typ].append(matrix[i])
+MAX_DISTANCE = -1
+for typ in matrixByType:
+   tensorized = torch.FloatTensor(matrixByType[typ])
+   MAX_DISTANCE = int(tensorized[:,1].max())
+   matrixByType[typ] = tensorized.view(-1, MAX_DISTANCE, 4)
+   #print(matrixByType[typ][0])
+   misByType[typ] = matrixByType[typ][:,:,2]
+if MAX_DISTANCE == -1:
[…]

### 2020-02-16 code/analysis/medians/interpolate/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Interpolate_PCFG_All.py

@@ -1,0 +1,14 @@
+with open("../../../ud_languages.txt", "r") as inFile:
+  languages = inFile.read().strip().split("\n")
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../../../results/tradeoff/pcfg/listener-curve-interpolated.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "FileName", "Position", "Memory", "Surprisal"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["../python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_Interpolate_PCFG.py", language]).strip()
+
+

### 2020-02-16 code/analysis/medians/median-CI/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_PCFG.py

@@ -1,0 +1,176 @@
+# Better than yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MaxControl.py by modeling the median of REAL
+
+import sys
+
+language = sys.argv[1]
+
+def readTSV(x):
+    header = next(x).strip().split("\t")
+    header = dict(zip(header, range(len(header))))
+    data = [y.strip().split("\t") for y in x]
+    if len(data) < 2:
+       return (header, [])
+    for column in range(len(header)):
+      try:
+        vals=  [int(y[column]) for y in data]
+      except ValueError:
+        try:
+          vals=  [float(y[column]) for y in data]
+        except ValueError:
+          vals=  [y[column] for y in data]
+      for i in range(len(data)):
+          data[i][column] = vals[i]
+    return (header, data)
+try:
+  with open("../../../../results/raw/pcfg/"+language+"_pcfg_decay_after_tuning.tsv", "r") as inFile:
+     data = readTSV(inFile)
+except IOError:
+  print >> sys.stderr, ("ERROR nothing for this language? "+language)
+  quit()
+
+#print(len(data))
+
+
+
+#data = data %>% group_by(ModelID) %>% mutate(CumulativeMemory = cumsum(Distance*ConditionalMI), CumulativeMI = cumsum(ConditionalMI))
+
+import torch
+
+def g(frame, name, i):
+    return frame[1][i][frame[0][name]]
+
+matrix = [[g(data, "ModelID", i), g(data, "Distance", i), g(data, "ConditionalMI", i), g(data, "UnigramCE", i)] for i in range(len(data[1]))]
+
+matrixByType = {}
+misByType = {}
+#unigramCEByType = {}
+for i in range(len(data[1])):
+    typ = g(data, "Type", i)
+#    print(i,typ, len(data[1]))
+    if typ not in matrixByType:
+        matrixByType[typ] = []
+    matrixByType[typ].append(matrix[i])
+MAX_DISTANCE = -1
+for typ in matrixByType:
+   tensorized = torch.FloatTensor(matrixByType[typ])
+   MAX_DISTANCE = int(tensorized[:,1].max())
+   matrixByType[typ] = tensorized.view(-1, MAX_DISTANCE, 4)
[…]

### 2020-02-16 code/analysis/medians/median-CI/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_PCFG_All.py

@@ -1,0 +1,16 @@
+#from ud_languages import languages
+
+with open("../../../ud_languages.txt", "r") as inFile:
+  languages = inFile.read().strip().split("\n")
+
+import subprocess
+
+# ./python27 yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_HistogramsByMem_All.py > ../results/tradeoff/listener-curve-histogram_byMem.tsv
+
+with open("../../../../results/tradeoff/pcfg/listener-curve-ci-median.tsv", "w") as outFile:
+  print >> outFile, "\t".join(["Language", "Type", "Position", "Memory", "MedianEmpirical", "MedianLower", "MedianUpper", "Level"])
+  for language in languages:
+     print(language)
+     print >> outFile, subprocess.check_output(["../python27", "yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI_PCFG.py", language]).strip()
+
+

### 2020-02-16 code/preparePaper/output/medians_small_pcfg.tex
code

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
 \\

### 2020-03-29 writeup/si.tex
code

@@
 
+% https://tex.stackexchange.com/questions/24132/overline-outside-of-math-mode
+\makeatletter
+\newcommand*{\textoverline}[1]{$\overline{\hbox{#1}}\m@th$}
+\makeatother
+
 \frenchspacing
@@
 We represent language as a stochastic process of words $\dots w_{-2} w_{-1} w_0 w_{1} w_{2} \dots$, extending indefinitely both into the past and into the future.
-The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, ..., any other kind of unit.}
+The symbols $w_i$ belong to a common set, representing the words of the language.\footnote{Could also be phonemes, sentences, or any other kind of unit.}
 
@@
 In reality, the statistical rules of language do change: They change as language changes over generations, and they also change between different situations -- e.g., depending on the interlocutor at a given point in time.
-Given that we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes, stationarity seems to be a reasonable assumption to make.
+However, we are interested in memory needs in the processing of \emph{individual sentences}, at a timescale of seconds or minutes.
+At this level, the statistical regularities of language do not change, making stationarity a reasonable modeling assumption.
 
@@
 \begin{proof}
-The difference between the listener's surprisal and optimal surprisal is $\operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.\footnote{A listener whose predictions are not optimal given $m_t$ can only incur even higher surprisal.}
+The difference between the listener's average surprisal $S_M$ and optimal surprisal $S_\infty$ is $S_M - S_\infty = \operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]$.\footnote{A listener whose predictions are not optimal given $m_t$ can only incur even higher surprisal.}
 By the assumption of stationarity, we can, for any positive integer $T$, rewrite this expression as
@@
 Therefore
-\begin{align*}
+\begin{align}\label{eq:surp-diff-expanded}
 \operatorname{H}[w_t | m_t] - \operatorname{H}[w_t | w_{<t}]& \geq \frac{1}{T} \left(\sum_{t=1}^T t I_t + T \sum_{t > T} I_t - I[w_{1\dots T}, m_1]\right) 
-\end{align*}
-$I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus, the expression above is bounded by
+\end{align}
+The term $I[w_{1\dots T}|m_1]$ is at most $\operatorname{H}[m_1]$, which is at most $\sum_{t=1}^T t I_t$ by assumption. Thus,  (\ref{eq:surp-diff-expanded}) implies the following:
 \begin{align*}
@@
 \end{align*}
-Rearranging shows that the listener's surprisal is at least $\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t$, as claimed.
+Rearranging yields
+\begin{align}
+\operatorname{H}[w_t|m_t] \geq \operatorname{H}[w_t | w_{<t}] + \sum_{t > T} I_t
+\end{align}
+as claimed.
 \end{proof}
@@
 Here, we show that this assumption can be relaxed to stochastic encoding functions.
-
-We relax condition (X) to the following requirement, for all values of $m_1, (w_{t})_{t \in \mathbb{Z}}, m_0$:
+That is, we show that our result holds if
+$m_t$ is not a deterministic function of $w_{t-1}$ and $m_{t-1}$, but includes randomness in processing.
+
+To formalize that setting, we relax Comprehension Postulate 1 to the following requirement, for all values of $m_1, (w_{t})_{t \in \mathbb{Z}}, m_0$:
 \begin{equation}\label{eq:listener-markov-nondeterministic}
@@
 
-The one place in the proof where (X) plays a role is the proof of the inequality:
+The one place in the proof where Comprehension Postulate 1 plays a role is the proof of the inequality:
 \begin{equation}
@@
 
-\section{Example where window model is not optimal}
+\section{Examples with Analytical Calculations}
+
+Here, we provide examples of the theorem in settings where analytical calculations are possible.
+
+\subsection{Example I: Tight Bounds in Artificial Language}
+
+Here we provide explicit calculations for the artificial language simulation, showing that the bounds provided by the theorem are tight in this case.
+
+TODO
+
+\subsection{Example II: Window-Based Model not Optimal}
 
 Here we provide an example of a stochastic process where a window-based memory encoding is not optimal, but the bound provided by our theorem still holds.
+This is an example where the bound provided by the theorem is loose: while it bounds the memory-surprisal tradeoff of all possible listeners, the bound is `optimistic', meaning that no mathematically possible memory encoding function $M$ can exactly achieve the bound.
 
@@
 
+\subsection{Example III: Tight Bound for Retrieval Model}
+
+Here, we provide an example where our bound is tight for the retrieval-based model even though it is quite loose for the capacity model.
+That means, while no memory encoding function can exactly achieve the bound in the capacity-bounded setting, there are memory encoding functions that exactly achieve the bound in the retrieval-based setting.
+
+Let $k$ be a positive integer.
+Consider a process $x_{t+1} = (y_{t+1}, z_{t+1}, u_{t+1}, v_{t+1})$ where
+\begin{enumerate}
+    \item $y_{t+1}$ consists of $2k$ random bits.
+    \item $z_{t+1}$ is a draw from $Bernoulli(\frac{1}{4k})$.
+    \item $u_{t+1}$ consists of $2k$ random bits if $z_t = 0$ and is equal to $y_{t-2k+1}$ else.
+    \item $v_{t+1} := z_t$ 
+\end{enumerate}
+Predicting observations $x_{t+1}$ optimally requires storing $y_{t-2k+1}, \dots, y_{t}$ and $z_t$.
+This amounts to $(2k+1)\cdot 2k + H_2[1/4k]$ bits of memory in the capacity-based model.
+However, $I[u_{t+1}, y_{t-2k+1}|z_{t+1}] \leq 1/k$ (TODO).
+Therefore, the theorem bounds the memory cost only by $HM \geq \sum_{t=1}^\infty t I_t = 1$.
+The bound provided by the theorem is therefore loose in this case.
+
+However, it is tight for the retrieval-based model:
+We use $s_t$ to store $y_{t-2k+1}, \dots, y_{t}$, and we use the working memory $m_{t+1}$ to store $z_t$.
+Then, if $z_t = 1$, we retrieve $r_t = g(x_{t-1}, m_{t-1}, s_{t-1}) := y_{t-2k+1}$.
+The cost of storing $z_t$ is $H_2[1/4k]$, and the cost of retrieving $r_t$ is $\frac{1}{4k} \cdot 2k$.
+
+In total, $H[m_t] = H_2[1/4k]$ and $H[r_t] = 1/k$.
+
+Taking, in the theorem, $T=1$ and $S\rightarrow\infty$, we obtain $H[m_t] \geq H_2[1/4k]$ and $H[r_t] \geq 1/k$.
+Thus, the bound is tight for both working memory and retrieval costs.
+
 \section{Corpus Size per Language}
@@
 
-\section{Optimization}
+\section{Chart Parsing Control}
+
+LSTMs and n-gram models are linear sequence models that might incorporate biases towards linear order as opposed to hierarchical structure.
+Here we use chart parsing to show that the results also hold when estimating $I_t$ using a model that is based on hierarchical structure and incorporates no bias towards linear closeness.
+
+We use PCFGs.
+PCFG surprisal is often computed in psycholinguistic research using approximate incremental parsers, but these might themselves incorporate some biases towards linear closeness.
+We instead opt for exact inference for PCFGs using exact chart parsing.
+
+\subsection{Deriving PCFGs from Dependency Corpora}
+
+We binarize, and give assign nonterminal labels to intermediate nodes based on (1) the POS of the head, (2) the lexical identity of the head, (3) the dependency label linking head and dependent.
+We binarize so that left children branch off before right children.
+
+The preterminals are labeled by POS tag and lexical identity.
+
+It is necessary to reduce the number of preterminals and nonterminals, both to deal with data sparsity, and to make chart parsing tractable.
+In our implementation for calculating $I_t$ (see below), we found that up to 700 nonterminals were compatible with efficient inference.
+(For comparison, the Berkeley parser uses X nonterminals for its English grammar, but employs a highly optimized coarse-to-fine strategy.)
+
+We reduced the number of nonterminals as follows:
+(1) For words with frequency below a threshold parameter, we did not record lexical identity in preterminals and nonterminals.
+(2) Nonterminals that only differ in the relation label were merged if their frequency fell below a threshold parameter,
+(2) Nonterminals that only differ in the head's lexical identity were merged if their frequency fell below a threshold parameter.
+
+Words occurring less than 3 times in the dataset were replaced by OOV.
+
+Alternative: merge-and-split, but that would have taken too long to run on all the corpora.
+
+We chose the threshold parameters for (1)-(3) separately for each language by sampling 15 configurations, and choosing the one that minimized estimated surprisal (see below) on a sampled baseline grammar, while resulting in at most 700 nonterminals and preterminals.
+
+mention approaches in the literature
+
+An alternative avoiding binarization would be to use the Earley parser, but that would have made it less feasible to parallelize processing on GPUs (see below).
+
+\subsection{Estimating $I_t$ with Chart Parsing}
+algorithm, cite Goodman's thesis
+
+Calculating $I_t$ requires estimating entropies $H[X_1, \dots, X_t]$, and thus probabilities $P(X_1, \dots, X_t)$.
+This is challenging because it requires marginalization over possible positions in a sequence.
+
+There is a known extension of the CKY algorithm that calculates \emph{prefix} probabilities
+\begin{equation}
+P[\#, X_1, \dots, X_t] := \sum_N \sum_{Y_{1\dots N}} P(\#, X_1, \dots, X_t, Y_{1\dots N}, \#)
+\end{equation}
+(here, $\#$ denotes the beginning/end of a sentence), that is, the probability mass assigned to all sentences starting with the given prefix $X_1, \dots, X_t$.
+
+However, simultaneously summing over possible left \emph{and} right continuations is more challenging.\footnote{(CITE) describe a method for calculating infix probabilities, but that method computes something subtly different from the quantity required here, and it is also computationally costly.}
+We approach this by restricting the summation on the left to prefixes of a fixed length:
+\begin{equation}
+    \sum_{Y_1\dots Y_N} P(\#, Y_1 \dots Y_N, X_1, \dots, X_t)
+\end{equation}
+and estimating
+\begin{equation}
+    P(X_t|X_1\dots X_{t-1}) \approx \E_{Y_1\dots Y_N} P(X_t|\#, Y_1 \dots Y_N, X_1, \dots, X_{t-1})
+\end{equation}
+Under certain conditions on the PCFG, this approximation converges to the tru value for sufficiently large values of $N$.\footnote{TODO say something about Markov chain convergence: For each $t$, consider for each nonterminal $\tau$ the number $n_\tau$ of nodes with this nonterminal dominating $w_\tau$.
+This is a Markov chain.}
+Empirically, we found that the values already became essentially stationary at $N\geq 5$.
+
+The resulting algorithm is shown in X.
+
+For computational efficiency, we estimated $I_t$ for $t=1, \dots 5$, finding $I_t$ to be very close to zero for higher $t$.
+
+We ran the algorithm on all contiguous sequences of length $T=5$.
+Following (CITE), we took advantage of GPU parallelization for implementation, processing 1,000 sequences in parallel.
+
+\subsection{Results}
+We computed $I_t$ for the MLE grammar and for five random baseline grammars.
+
+We did not run this on the observed orderings, as these may have crossing branches, making binarization difficult and thus rendering comparison with baselines less straightforward.
+
+Figure~\ref{fig:resu-pcfg}
+
+Limitations: Absolute numbers aren't comparable with other models because there are many OOVs (they are necessary because the number of non- and preterminals has to be kept low).
+Also, the amount of exploited predictive information $\sum_t I_t$ is much lower than in the other models.
+Agrees with the observation that PCFG independence assumptions are inadequate, and that chart parsers have not historically reached good perplexities (parsers with good perplexities such as Roark Parser and RNNGs do not make these independence assumptions, but also do not allow efficient exact chart parsing).
+Nonetheless, the experiment confirms the finding with a model that is based on hierarchical syntactic structure while enabling exact inference.
+
+\begin{center}
+\includegraphics[width=\textwidth]{results-table-pcfg.pdf}
+\captionof{figure}{PCFG estimator, comparing fitted grammars (blue) with baselines (red)}\label{fig:resu-pcfg}
+\end{center}
+
+\section{Morphology}
+
+\subsection{Japanese}
+
+Here, we describe how we determined the Japanese verb suffixes described in the main paper.
+
+%Japanese orthography does not indicate word boundaries, and there is no universally accepted segmentation.
+We determined a set of frequent morphemes as follows.
+We selected all morphemes occurring in the dataset at least 50 times and annotated their meaning/function.
+Among these, three morphemes are treated as independent words, not suffixes, by \cite{kaiser2013japanese} (\textit{dekiru} `be able to', \textit{naru} `become', \textit{yoo} `as if'); we excluded these.
+Furthermore, passive and potential markers are formally identical for many verbs; we included both here.
+
+We list the morphemes according to the order extracted according to the model.
+%While this ordering matches almost all observed forms, there are some orderings that do not fit into this pattern, see below.
+
+Note that there  is no universally accepted segmentation for Japanese suffixes; we follow the UD tokenization in choosing which suffixes to segment.\footnote{The biggest difference to some other treatments is that the ending -\textit{u}/-\textit{ru} is viewed as part of the preceding morpheme that appears in some environments due to allomorphic variation, while it is viewed as a nonpast suffix in some other treatments \citep[p.116]{hasegawa2014japanese}.}
+
+\begin{enumerate}
+\item VALENCE: causative (-\textit{(s)ase}-).
+\cite[142]{hasegawa2014japanese} \cite[Chapter 13]{kaiser2013japanese}.
+In the UD data, this is lemmatized as \textit{saseru}, \textit{seru} (190 occurrences).
+\item VOICE: passive (-\textit{are}-, -\textit{rare}-) \cite[152]{hasegawa2014japanese} \cite[Chapter 12]{kaiser2013japanese}  In the UD data, this is lemmatized as \textit{rareru}, \textit{reru} ($\approx$ 2000 occurrences).
+\item MOOD, MODALITY:
+\begin{enumerate}
+\item potential (allomorphs -\textit{are}-, -\textit{rare}-, -\textit{e}-). In the UD data, this is lemmatized as \textit{rareru}, \textit{reru}, \textit{eru}, \textit{keru}.
+This is formally identical to the passive morpheme for many verbs (\cite[346]{vaccari1938complete}, \cite[398]{kaiser2013japanese}).
+
+%atsuka-e-nai `can't handle' %扱えない GSD treebank
+%araso-e-nai `can't dispute' % 争えない GSD treebank
+
+%how is this ordered relative to PASSIVE?
+In UD, they don't seem to be always separated (ambiguity of reru/rareru.
+\cite[346]{vaccari1938complete}: identical to the passive for one class, or -eru for other verb class).
+%\cite[398]{kaiser2013japanese}
+
+\item politeness -\textit{mas}- (allomorphs -\textit{masu}-, -\textit{mashi}-, -\textit{mase}-) \cite[190]{kaiser2013japanese}. % 見できます, 見ましたい
+In the UD data, this is lemmatized as \textit{masu} ($\approx$ 600 occurrences).
+\item MODALITY: desiderative -\textit{ta}- (allomorphs: -\textit{tai}, -\textit{taku}-, -\textit{taka}-) (85 occurrences) \cite[238]{kaiser2013japanese}. %mi-mashi-tai `I want to see'
+\end{enumerate}
+\item NEGATION: negation -\textit{na}- (allomorphs: -\textit{nai}, -\textit{n}-, -\textit{nakat}-). 
+%also -mai for negative+volition.
+Lemmatized as \textit{nai} (630 occurrences).
+\item TENSE/ASPECT/MOOD:
+\begin{enumerate}
+\item -\textit{ta} for past (4K occurrences) \cite[211]{kaiser2013japanese}
+\item -\textit{yoo} for hortative, future, and similar meanings \cite[229]{kaiser2013japanese}. This is lemmatized as \textit{u} (92 occurrences).
+\end{enumerate}
+%Q how do-nai- and -yoo interact? Should -yoo be placed together with desiderative?
+\item -\textit{te} derives a nonfinite form \cite[186]{kaiser2013japanese}. (4K occurrences)
+\end{enumerate}
+
+%When labeling morphemes, we consider the lemmas, and unite saseru and seru to CAUSATIVE and reru+rareru+eru+keru to PASSIVE/POTENTIAL.
+
+%EXAMPLES:
+
+We provide examples illustrating the relative ordering of different morphemes.
+Note that passive and potential markers do not co-occur.
+We omit examples with -te; it always follows other suffixes that are compatible with it.
+
+\begin{tabular}{lllllllll|lllllll}
+%     & VALENCE   & VOICE    & \multicolumn{3}{c}{MOOD/MODALITY} & NEGATION & TENSE \\
+Stem & Caus. & Pass. & Pot. & Polite. & Desid. & Neg. & TAM & \\ \hline
+mi &          &      & &            &          & naka     & tta    & did not see \cite[153]{vaccari1938complete}\\
+mi &          &      & &            & taku     & nai      &         & I do not wish to see \cite[98]{vaccari1938complete} \\
+mi &          &      & &            & taku     & naka    & tta       & I did not wish to see \cite[98]{vaccari1938complete} \\
+%oyog & ase    &      & &            &          &         & ta    &     & made swim \\
+tat & ase     & rare & &            &          &         & ta         & was made to stand up \cite[396]{kaiser2013japanese} \\
+waraw  &       & are & &            &           &         & ta        & was laughed at \cite[384]{kaiser2013japanese} \\
+mi     &       & rare& & mase       &          & n        &           & is not seen \cite[337]{vaccari1938complete} \\
+mi     &       & rare& & mash      &           &          & yoo      & will be seen \cite[337]{vaccari1938complete} \\
+de     &       &    & &            &           & naka     & roo      & will not go out \cite[170]{vaccari1938complete} \\
+mi     &       &    & e & mase       &           & n        &          & cannot see \cite[349]{vaccari1938complete} \\
+%mi     &       &    & & mashi      & tai       &          &      &    & want to see
+\end{tabular}
+
+%Not all combinations are possible. 
+%`was not seen' is miraremasen deshita \cite[337]{vaccari1938complete}, not MIRU-RERU-MASE-NAI-TA
+%Negation is not directly combined with -yoo.
+
+%Some relatively frequent morphemes that are below the threshold include:
+
+%\begin{enumerate}
+%\item nakereba `if not' (cf \cite[9.3.1.4]{kaiser2013japanese})
+%\item tsudzukeru (continuative, 30 occurrences)
+%\item yasui `easy to'
+%\item hajimeru (inchoative, 21)
+%\item naru (`become', 102). Maybe this should be excluded, not treated as an affix in the books.
+%\item yooda (180) EVIDENTIAL?
+yooda.
+but past -ta- can be after it: iruyoodatta IRU-YOODA-TA (Kaiser et al 9.5.6.1.1.1) vs misenakattayooda MISERU-NAI-TA-YOODA `it didn't seem' (GSD treebank).
+Maybe this should be thought of better as an SCONJ at least when including -ni, Vaccari p.
+238-241.
+%\item beki Necessitative \cite[248]{kaiser2013japanese}
+%\item soo `likely to' \cite[258]{kaiser2013japanese}, probably not AUX in UD
+%\item dekiru potential (180), 
+%\end{enumerate}
+
+% volitional 加わろう kawawar-oo
+% volitional polite 加わりましょう kawawar-i-mashy-oo
+% volitional neg 加わるまい(+), 加わらない[よう/こと]にしよう kawawaru-mai, kawawara-nai-yoo-nishyoo
+% volitional neg polite 加わりますまい, 加わらない[よう/こと]にしましょう  kawawari-masu-mai, kawawar-anai-yoo-nishimashyoo
+
+%  verb-valence-voice-aspect-tense-mood-modality-person-number
+
+% http://nihongo.monash.edu/cgi-bin/wwwjdic?1W%B2%F1%B5%C4%A4%CB%B2%C3%A4%EF%A4%EB_v5r
+% passive negative
+% 加わられない kuwawar-are-nai
+
+% passive negative polite
+% 加わられません kuwawar-are-mase-n
+
+% causative polite
+% 加わらせます kuwawar-ase-masu
+
+% causative negative polite
+% 加わらせません kuwawar-ase-mase-n
+% 加わらしません kuwawar-ashi-mase-n
+
+% da/de/na (lemmatized as da): copula (?). na after adjectives.
+
+%honorific -masu
+
+%-s-are-naka-tta ?passive+negative.past?
+
+% -u/-ru (non-past) -- not segmented off in UD?
+
+% suru - reru - ta
+% suru -masu - ta
+% reru - ta
+% suru - ta
+% masu - ta
+% rareru - ta
+% masu - nai - deshita
+% suru - reru
+% suru - nai
+% masu - nai
+% reru - masu - ta
+% dekiru - masu
+
+% TODO in excluding -te-, have to deal with -de- as in 読んで yonde (Vaccari p. 109)
+
+%(('し/動詞/し', 'た/助動詞/た'), 14389)
+%(('る/語尾/る',), 13880)
+%(('あ/動詞/あ', 'る/語尾/る'), 11481)
+%(('っ/語尾/っ', 'た/助動詞/た'), 9378)
+%(('さ/動詞/さ', 'れ/助動詞/れ', 'た/助動詞/た'), 7143)
+%(('さ/動詞/さ', 'れ/助動詞/れ', 'る/語尾/る'), 1966)
+%(('られ/助動詞/られ', 'る/語尾/る'), 1156)
+%(('だっ/助動詞/だっ', 'た/助動詞/た'), 1151)
+%(('られ/助動詞/られ', 'た/助動詞/た'), 1114)
+%(('ん/語尾/ん', 'だ/助動詞/だ'), 854)
+%((), 839)
+%(('い/語尾/い', 'た/助動詞/た'), 761)
+%(('わ/語尾/わ', 'れ/助動詞/れ', 'た/助動詞/た'), 715)
+%(('で/助動詞/で', 'あ/動詞/あ', 'る/語尾/る'), 688)
+%(('さ/語尾/さ', 'れ/助動詞/れ', 'た/助動詞/た'), 599)
+%(('わ/語尾/わ', 'れ/助動詞/れ', 'る/語尾/る'), 595)
+%(('れ/助動詞/れ', 'る/語尾/る'), 563)
+%(('く/語尾/く',), 485)
+
+% ('できる', 'ます'): 11, ('かもしれる', 'ます', 'ない'): 1, 
+%('ない', 'た'): 29, ('ちゃう', 'ます', 'た'): 2, 
+%('する', 'ます'): 23, ('られる', 'ます'): 4, ('できる',): 18, ('する', 'せる', 'た'): 12, ('たい', 'ない'): 1, ('する', 'ます', 'ない'): 4, ('せる', 'ない'): 1, ('らしい',): 2, ('出来る', 'ます', 'ない'): 1, ('かもしれる', 'ない'): 1, ('れる', 'ない', 'た'): 1, ('ない', 'だめ', 'だ'): 1, ('する', 'れる', 'そうだ'): 2, ('える',): 5, ('かねる', 'ない'): 1, ('れる', 'ようだ', 'なる', 'た'): 3, ('せる', 'ます', 'ない'): 2, ('続ける', 'た'): 1, ('られる', 'ようだ', 'なる', 'ます', 'た'): 1, ('できる', 'ます', 'た'): 7, ('する', 'せる'): 2, ('やすい', 'なる', 'た'): 1, ('くださる', 'ます'): 1, ('ようだ', 'なる', 'ます', 'た'): 2, ('える', 'ようだ', 'なる', 'た'): 1, ('する', 'がたい'): 1, ('する', 'れる', 'ます', 'た'): 6, ('ざるを得る', 'ます', 'ない'): 1, ('ようだ', 'なる', 'た'): 10, ('する', 'れる', 'ます'): 3, ('られる', 'ます', 'た'): 4, ('える', 'た'): 1, ('する', 'ます', 'う'): 1, ('する', 'ます', 'ない', 'でした'): 1, ('済み',): 1, ('ようだ',): 2, ('出す', 'た'): 1, ('出す',): 1, ('する', 'た', 'みたいだ'): 1, ('させる', 'た'): 2, ('きる', 'た'): 1, ('する', 'たい'): 2, ('続ける',): 2, ('ざるをえる', 'ます', 'ない'): 1, ('める', 'ます'): 6, ('せる', 'ます'): 1, ('ようだ', 'なる'): 7, ('られる', 'ようだ', 'なる', 'た'): 1, ('始める', 'た'): 2, ('ない', 'なる'): 2, ('かねる', 'ます'): 1, ('始める',): 2, ('する', 'れる', 'ない', 'た'): 1, ('える', 'ます'): 3, ('た', 'そうだ'): 4, ('れる', 'ます', 'ない'): 1, ('する', 'ようだ', 'なる'): 4, ('する', 'える', 'ない'): 1, ('ける', 'ます'): 3, ('できる', 'ない', 'た'): 1, ('する', 'ない', 'た'): 2, ('出来る', 'ます'): 1, ('れる', '始める'): 1, ('ます', 'う'): 4, ('する', '始める', 'た', 'そうだ'): 1, ('する', 'れる', 'た', 'ようだ'): 1, ('できる', 'まい'): 1, ('た', 'らしい'): 1, ('する', 'ようだ', 'なる', 'た'): 2, (' える', 'ます', 'ない'): 1, ('ける',): 2, ('られる', 'た', 'そうだ'): 1, ('られる', 'そうだ'): 1, ('れる', 'にくい'): 1, ('する', 'やすい'): 1, ('ける', 'た'): 1, ('する', 'れる', 'ます', 'ない'): 2, ('そ うだ',): 1, ('ない', 'た', 'ようだ'): 1, ('ます', 'たー'): 1, ('できる', 'た'): 3, ('べる',): 1, ('する', 'れる', 'ようだ', 'なる', 'ます', 'た'): 1, ('ない', 'ようだ'): 1, ('ける', 'ます', 'う'): 1, ('ちゃう', 'ます'): 1, ('する', 'そうだ'): 1, ('たい', 'なる'): 1, ('ようだ', 'なる', 'ます'): 1, ('たい', 'ない', 'た'): 1, ('やすい',): 3, ('かける',): 1, ('ない', 'なる', 'ようだ', 'なる', 'た'): 1, ('ない', 'そうだ'): 1, ('する', 'たい', 'ない', 'らしい'): 1, ('する', 'ちゃう', 'ます'): 1, ('た', 'ようだ'): 1, ('られる', 'たい', 'ない'): 1, ('える', 'ない'): 1, ('てる', 'ない'): 1, ('する', '易い', 'なる', 'た'): 1, ('れる', 'ない'): 1})
+
+% nai (negation)
+
+\subsection{Sesotho}
+
+Here, we describe how we determined the Sesotho verb prefixes and suffixes.
+
+Sesotho has composite forms consisting of an inflected auxiliary followed by an inflected verb.
+Both verbs carry subject agreement.
+While they are annotated as a unit in the Demuth corpus, they are treated as separate words in grammars \citep{doke1967textbook,guma1971outline}.
+We separated these, taking the main verb to start at its subject agreement prefix.
+We only considered main verbs for the experiments here.
+
+Forms in child utterances are annotated with well-formed adult forms; we took these here.
+
+In the Demuth corpus, each morpheme is annotated; a one- or two-letter key indicates the type of morpheme (e.g. subject agreement, TAM marker).
+We classified morphemes by this annotation.
+
+According to \cite{demuth1992acquisition}, affixes in the Sesotho verb have the following order:
+\begin{enumerate}
+    \item Subject agreement
+    \item Tense/aspect
+    \item Object agreement
+    \item Verb stem
+    \item `Extension'/perfect/passive markers, where `extension' refers to causative, neuter/stative, reversive, etc.
+    \item Mood
+\end{enumerate}
+We refined this description by considering all morpheme types occurring at least 50 times in the corpus.
+
+As in Japanese, morphemes show different forms depending on their environment, and the corpus contains some instances of fused neighboring morphemes that were not segmented further.
+
+\paragraph{Prefixes}
+
+\begin{enumerate}
+    \item Subject agreement:
+
+    This morpheme encodes agreement with the subject, for person, number, and noun class (the latter only in the 3rd person) \cite[\textsection 395]{doke1967textbook} \cite[p. 162]{guma1971outline}.
+
+    In the Demuth corpus, this is annotated as \textit{sm} (17K occurrences) for ordinary forms, and \textit{sr} (193 occurrences) for forms used in relative clauses.
+
+    \item Negation:
+
+    In various TAM forms, negation is encoded with a morpheme -\textit{sa}- in this position (362 occurrences) \cite[p. 172]{guma1971outline} \cite[\textsection 429]{doke1967textbook}.
+    Common allomorphs in the corpus include \textit{ska}, \textit{seka}, \textit{sa}, \textit{skaba}.
+
+%    sa used in Negative Dependent Present (Paroz 1946, p.31) and Perfect (ibid p. 80). E.g. Sec 13.38 p.172 in Guma 1971. -sa- and -se- Lombard 1985 p. 108, appear in different moods.
+    %What is the order of NEGATION and TENSE/ASPECT?
+
+    \item Tense/Aspect/Mood (13K occurrences)
+
+    Tense/aspect marker (t\^{} 13K) \cite[p. 165]{guma1971outline}
+
+    Common TAM markers in this position in the corpus include, with the labels provided in the Demuth corpus:
+
+    \begin{itemize}
+    \item -tla-, -tlo-, -ilo- future \cite[\textsection 410--412]{doke1967textbook}
+    \item -a- present \cite[\textsection 400]{doke1967textbook}
+    \item -ka- potential \cite[\textsection 422--428]{doke1967textbook}
+    \item -sa- persistive \cite[\textsection 413--418]{doke1967textbook}
+    \item -tswa- recent past \cite[\textsection 404--406]{doke1967textbook} 
+    %\item 
+    %..
+    %Past subjunctive
+
+    %Potential -ka- (cannot be combined with negation marker in the previous slot, \cite[\textsection 424]{doke1967textbook}).
+    \end{itemize}
+
+    %Some of the common markers in the corpus are t\^{}f for future and t\^{}pf for perfect.
+    %(e.g., ile- for remote past \cite[\textsection 177]{doke1967textbook})
+
+    In the corpus, TAM prefixes are often fused with the subsequent object marker.
+
+    \item OBJECT agreement (labeled om, 6K occurrences) or reflexive (labeled rf, 751 occurrences).
+
+    Similar to subject agreement, object agreement denotes person, number, and noun class features of the object.
+    Unlike subject agreement, it is optional \cite[\textsection 459]{doke1967textbook}.
+
+    Object agreement and reflexive marking are mutually exclusive \cite[p. 165]{guma1971outline}.
+\end{enumerate}
+
+In addition to these morphemes used in finite verbs, there is an \textit{infinitive} prefix \textit{ho}- (labeled `if', 314 occurrences) \cite[\textsection\textsection 378-384]{doke1967textbook}, which is compatible with TAM markers \cite[\textsection 379]{doke1967textbook} and object agreement \cite[\textsection 382]{doke1967textbook}.
+%Additionally FINITENESS (?) Conditional morpheme (\emph{ho}, `if', 314), comes before object agreement.
+
+%In agreement with Bybee's hierarchy, subject agreement is encoded in a position further away than TAM features.
+
+%TODO ha- negation is a prefix coming before the subject agreement prefix (Guma 1971, p. 164).  ga- in Lomard 1985 p. 108.
+
+%Also some merging between tense and object markers. for the lemma-based version, have to construct a version where all of this merging is undone (e.g. including a slash, or t\^{}.om2s)
+
+\paragraph{Verb Suffixes in Sesotho}
+
+Again, we extracted morpheme types occurring at least 50 times.
+
+\begin{enumerate}
+\item Reversive: (labeled \textit{rv}, 214 occurrences),  \cite[\textsection 345]{doke1967textbook}.
+
+    This suffix changes semantics.
+    Examples: tlama `bind' -- tlam\textoverline{o}lla `loosen', etsa `do' -- ets\textoverline{o}lla `undo' \citep[\textsection 346]{doke1967textbook}.
+    Such suffixes are found across Bantu languages \cite{schadeberg2003derivation}.
+
+    %General references (comparative Bantu):
+%Schadeberg, Thilo C. 2003. Derivations. In Derek Nurse and Gerard Philippson (eds.), The Bantu languages, 71–89. London: Routledge.
+%Doke, Clement. 1935. Bantu linguistic terminology. London: Longman, Green \& Co
+%Guthrie, Malcolm. 1962. The status of radical extensions in Bantu languages. Journal of African Languages 1(3). 202–220.
+
+    \item VALENCE:
+    \begin{enumerate}
+    \item causative (labeled \textit{c}, 1K occurrences), -\textit{isa} (with morphophonological changes) \cite[\textsection 325]{doke1967textbook}
+
+    \item neuter (labeled \textit{nt}, 229 occurrences), -\textit{eha}, -\textit{ahala} \cite[\textsection 307]{doke1967textbook}
+
+    The neuter suffix reduces valence: lahla `throw away' -- lahlela `get lost', s\textoverline{e}nya `to damage' -- s\textoverline{e}nyeha `to get damaged' \citep[\textsection 308]{doke1967textbook}.
+
+    \item applicative
+    (labeled \textit{ap}, 2K occurrences) -el- \cite[\textsection 310]{doke1967textbook}
+
+    The applicative suffix increases valence: b\textoverline{o}lela `to say' b\textoverline{o}lella `to say to (s.o.)' \citep[\textsection 310]{doke1967textbook}.
+
+    %\cite[p. 109]{lombard1969handbook}
+
+%    Example from Demuth corpus:
+
+ %   I am lighting for you
+
+  %  ke-a-u-khan-ts-ets-a
+
+   % sm1s-t\^{}p-om2s-shine-c-ap-m\^{}in
+
+%    Leave my blanket alone then
+
+ %   tl-oh-el-a kobo ya-ka he
+
+  %  come-rv-ap-m\^{}i blanket(9,10) 9-1s ij
+
+    \item Perfective/Completive -\textit{ella} (annotated cl, 66 occurrences) \cite[\textsection 336]{doke1967textbook}
+
+    This does not actually change valence, but it is formally a reduplication of the applicative suffix \cite[\textsection 336]{doke1967textbook}, and as such its ordering behavior patterns with that of valence suffixes, in particular, it is placed before the passive suffix.\footnote{Example from the Demuth corpus: u-neh-el-ets-w-a-ng t\^{}p.om2s-give-ap-cl-p-m\^{}in-wh `What is it that you want passed to you?'.}
+
+    \item Reciprocal -\textit{ana} (annotated rc, 103 times)  \cite[\textsection 338]{doke1967textbook}
+
+    This reduces valence: rata `to love' -- ratana `to love another' \citep[\textsection 338]{doke1967textbook}.
+
+    \end{enumerate}
+
+    Some of these suffixes can be stacked, e.g., see \cite[\textsection 345]{doke1967textbook} for reversive+causative, and \cite[\textsection 314-315]{doke1967textbook} for applicative suffixes applied to other valence affixes.\footnote{Example of reciprocal+applicative from Demuth corpus: ba-arol-el-an-a sm2-t\^{}p\_divide-ap-rc-m\^{}in `Do they share?'}
+
+    Some other suffixes documented in the literature do not occur frequently or are not annotated in the corpus (e.g., the associative suffix \cite[textsection 343]{doke1967textbook}).
+
+    \item VOICE: passive -\textit{w}- (labeled \textit{p}, 1K occurrences) \cite[\textsection 300]{doke1967textbook}
+
+    %\cite[p. 114]{lombard1969handbook}
+
+    %Passive follows applicative: sho-el-oa die-APPL-PASS \cite[\textsection 324]{doke1967textbook}
+
+    \item TENSE: tense (labeled t\^{}, 3K occurrences) .
+
+    The only tense suffix is the perfect affix -\textit{il}-, which has a range of allomorphs depending on the preceding stem and valence/voice suffixes, if present \cite[\textsection 369]{doke1967textbook}, \cite[p.
+167]{guma1971outline}.
+% \cite[p.
+116]{lombard1969handbook}.
+    Common morphs in the Demuth corpus are -il- and -its-.
+
+    \item MOOD: Mood (labeled m\^{}, 37K occurrences)
+
+    In the Demuth corpus, the following mood endings are labeled (the analysis provided by \cite{demuth1992acquisition} is different from that provided by \cite{doke1967textbook}, meaning the citations are only approximate):
+
+    \begin{enumerate}
+    \item Imperative (labeled IMP) \cite[\textsection 386--387]{doke1967textbook}: singular (-e, labeled IMP) \cite[\textsection 386]{doke1967textbook} and plural (-ang, labeled IMP.PL) \cite[\textsection 386]{doke1967textbook}.
+
+    Similar subjunctive SBJV1 -\textit{e} (singular), -\textit{eng} (plural).
+
+    \item IND (-a, -e) and NEG  (-e, -a) \cite[\textsection 394--421]{doke1967textbook}.
+
+    \item subjunctive SBJV2 (-e, -a)  \cite[\textsection 444--455]{doke1967textbook}
+    \end{enumerate}
+
+    %We note that the analysis of these suffixes as mood endings follows \cite{demuth1992acquisition}.
+
+    % Demuth:
+    %The mood of infinitives is m^in if affirmative indicative, m^x if neg in indicative assertive, declarative, affirmative statements 
+    %pt participial in compound tenses, subordinate clauses, relative clauses 
+    %x negative in negative utterances ha ke-tseb-e ng sm1s-t^p_v^know-m^x I don’t know 
+    %i imperative m-ph-e om1s-v^give-m^i Give me. 
+    %ip imp plural bon-ang v^see-m^ip Look! 
+    %s subjunctive ere ke-bon-e  ht sm1s-t^p_v^see-m^s Let me see. 
+    %sp plural subjunctive ha re-y-eng ht sm1p-t^p_v^go-m^sp Let’s go.
+
+    \item Interrogative (labeled \textit{wh}, 2K times) and relative (labeled \textit{rl}, 857 times) markers -\textit{ng}.
+
+    The interrogative marker -\textit{ng} is a clitic form of \textit{eng} `what' according to \cite[p.
+168]{guma1971outline}, \cite[\textsection 160, 320, 714]{doke1967textbook}; it is treated as a suffix in the Demuth corpus.
+
+    The relative marker -\textit{ng} is affixed to verbs in relative clauses are marked with -ng \cite[\textsection 271, 793]{doke1967textbook}.
+
+\end{enumerate}
+
+Examples from \cite{demuth1992acquisition}:
+
+\begin{tabular}{lllllllll|lllllll}
+Sbj. & Ng. & TAM & Obj. & V & Val. & Voice & Tense & Mood \\ \hline
+o       &          &     &        & pheh &         &       & il    & e  & (Thabo) cooked (food) (\cite{demuth1992acquisition} (15)) \\
+ke      &          &     & e      & f   &          & uw    &      & e   & (I) was given (the book) (\cite{demuth1992acquisition} (26c)) \\
+o       &          &     &        & pheh & el      &      &       & a & (Thabo) cooks (food for Mpho) (\cite{demuth1992acquisition} (41))\\
+o       &          &     &        & pheh & el      & w    &      & a & (Mpho) is being cooked (food) (\cite{demuth1992acquisition} (42))
+\end{tabular}
+
+\section{Some theoretical thoughts (scratch area)}
+\subsection{Learnability bounds}
+
+\paragraph{Upper bounds via n-gram models:}
+Want to show that, if we know a process has stronger locality, there is a learning algorithm with lower sample complexity.
+
+\begin{thm}
+The class of processes with fixed bounds on $H_0$ on $EE$ can be learned up to KL loss $2\epsilon$ with ... samples with prob ...
+\end{thm}
+
+We want to learn a process to average KL distance $2\epsilon$.
+Assume excess entropy is $\leq I$, then only need to learn $N := I/\epsilon$-gram-model for KL loss $\epsilon$.
+
+(Or, to get a better bound, take $N$ so that $\sum_{t=N}^\infty I_t < \epsilon$.).
+
+So $N = \sum_{t=1}^\infty t I_t/\epsilon$.
+How many samples are needed to learn this up to $\epsilon$? Probably that will scale with the block entropy, which is $N H_0 + \sum_{t=1}^N t I_t$, with $H_0$ the unigram entropy.
+So the sample complexity would seem to scale with
+$$\exp\left(H_0 \sum_{t=1}^\infty t I_t/\epsilon + \sum_{t=1}^{\sum_{t=1}^\infty t I_t/\epsilon} t I_t\right)$$
+
+Want to learn $P(Y|X)$ up to $\mathbb{E}_x D_{KL}(P(y|x)||\hat P(y|x)) \leq \epsilon$.
+How many i.i.d. samples from $(X,Y)$ do we need?
+
+Assume the distribution of $Z$ has `low' entropy.
+How many i.i.d. samples do we need to get a good estimate of $P(Z)$?
+Something scaling with $\exp(H(Z))$?
+
+Convergence
+
+references:
+
+\url{https://arxiv.org/pdf/1904.02291.pdf} leads to:
+
+$$Pr(D_{KL}(\hat P(z)||P(z)) \geq \epsilon) \leq e^{-n\epsilon} \left(\frac{e\epsilon n}{|V|-1}\right)^{|V|-1}$$
+if $V$ is the set of values of $Z$, when $\epsilon > \frac{|V|-1}{n}$.
+
+Is it possible to get something similar but with $|V|$ replaced with $H[X]+|V_Y|$?
+
+\begin{thm}(must be something standard)
+Let $X$ be an RV. Then $1-\epsilon$ of its probability mass is concentrated on at most
+???
+values.
+\end{thm}
+
+\begin{proof}
+Assume no way of covering $1-\epsilon$ probability mass takes less than $K$ values.
+That is (ordering $p_i$ by magnitude downwards):
+$$\sum_{i=1}^K p_i \leq 1-\epsilon$$
+
+Want to show that $H[X]$ cannot be too small.
+
+First, note
+
+$(1-\epsilon) - (K-1) \frac{1-\epsilon}{|V|-K} \geq p_1 \geq \dots \geq p_K \geq \frac{1-\epsilon}{|V|-K}$
+
+or (reformulating slightly)
+
+$(1-\epsilon) (1 - (K-1) \frac{1}{|V|-K}) \geq p_1 \geq \dots \geq p_K \geq \frac{1-\epsilon}{|V|-K}$
+
+$H[X] \geq \sum_{i=1}^K \frac{1-\epsilon}{|V|-K} \log \frac{1}{(1-\epsilon) (1 -  \frac{K-1}{|V|-K})} = K \frac{1-\epsilon}{|V|-K} \log \frac{1}{(1-\epsilon) (1 -  \frac{K-1}{|V|-K})} = K \frac{1-\epsilon}{|V|-K} \log \frac{1}{(1-\epsilon) \frac{|V|-K-K+1}{|V|-K})} = K \frac{1-\epsilon}{|V|-K} \log \frac{{|V|-K}}{(1-\epsilon) (|V|-2K+1)}$
+
+And so
+
+$H[X] \geq  K \frac{1-\epsilon}{|V|-K} \log \frac{{|V|-K}}{(1-\epsilon) (|V|-2K+1)} =  (1-\epsilon) \frac{K}{|V|-K} \log \frac{{|V|-K}}{(1-\epsilon) (|V|-2K+1)} $
+If $K << |V|$, this is close to zero.
+
+-----
+
+So
+$H[X] \geq K \frac{1-\epsilon}{|V|-K} \log \frac{{|V|-K}}{(1-\epsilon) (|V|-K)} = \frac{K}{|V|-K} (1-\epsilon) \log \frac{1}{(1-\epsilon)}$
+
+So
+
+$\frac{K}{|V|-K} \leq H[X] \frac{1}{(1-\epsilon) \log \frac{1}{(1-\epsilon)}}$
+
+However, this bound does not seem very useful, as it never allows one to conclude something like $K << |V|$.
+
+Or, a better bound:
+
+$H[X] \geq  - (1-\epsilon) \log \left[ (1-\epsilon)  \left(\frac{|V|-2K+1}{|V|-K}\right) \right] + \epsilon \log \frac{1}{\epsilon}$
+
+So If $K << |V|$, then $H[X] \geq ... \approx -(1-\epsilon) \log (1-\epsilon)  + \epsilon \log \frac{1}{\epsilon}$ .
+
+\end{proof}
+
+%Then
+%$$H[X] \leq (1-\epsilon) \log \frac{K}{1-\epsilon} + \epsilon \log \frac{|V|-K}{\epsilon}$$
+%So
+%$$H[X] \leq (1-\epsilon) \log \frac{K}{1-\epsilon} + \epsilon \log \frac{|V|}{\epsilon}$$
+%So
+%$$H[X] - \epsilon \log \frac{|V|}{\epsilon} \leq (1-\epsilon) \log \frac{K}{1-\epsilon}$$
+%So
+%$$\exp\left(\frac{H[X] - \epsilon \log \frac{|V|}{\epsilon}}{1-\epsilon}\right) (1-\epsilon) \leq K$$
+%Taking the contrapositive, if
+%$$\exp\left(\frac{H[X] - \epsilon \log \frac{|V|}{\epsilon}}{1-\epsilon}\right) (1-\epsilon) > K$$
+%then $$\sum_{i=1}^K p_i > 1-\epsilon$$
+
+To get intuition, if $I_t = \alpha \cdot t^{-3}$, then excess entropy $I=\alpha \pi^2/6$, and (bounds are a bit crude here)
+\begin{align}
+& H_0 \sum_{t=1}^\infty t \alpha t^{-3}/\epsilon + \sum_{t=1}^{\sum_{t=1}^\infty t \alpha t^{-3}/\epsilon} t \alpha \cdot t^{-3} \\
+= & H_0 \alpha/\epsilon \cdot \sum_{t=1}^\infty  t^{1-3} + \sum_{t=1}^{\sum_{t=1}^\infty t \alpha t^{-3}/\epsilon} t \alpha \cdot t^{-3} \\
+\leq & H_0 \alpha/\epsilon \cdot \pi^2/6 + \sum_{t=1}^{\alpha/\epsilon \cdot \pi^2/6} t \alpha \cdot t^{-3} \\
+= & \frac{H_0 \alpha \pi^2}{6 \epsilon} + \alpha \sum_{t=1}^{\alpha/\epsilon \cdot \pi^2/6}  t^{-2} \leq  \frac{H_0 \alpha \pi^2}{6 \epsilon} + \alpha \pi^2/6  \leq  \left(\frac{H_0}{\epsilon} + 1\right) \frac{\alpha \pi^2}{6} \\
+\end{align}
+
+Question: Can there be meaningful lower bounds in terms of locality?
+
+Question: Can we give a lower bound by considering that learning sequences itself requires short-term memory?
+
+try to derive something for Trading Value and Information paper
+
+\subsection{Optimization}
+Area under $t - H_t$ curve up to $T$: $\sum_{t=1}^T H_t = T H + \sum_{t=1}^T t I_t$
+
 We cannot optimize for AUC using the method of (CITE), because we cannot construct an unbiased gradient estimator

### 2020-03-29 writeup/si.tex
code

@@
 \usepackage{pslatex}
+\usepackage{hyperref}
 %\usepackage{latexsym}
@@
 \maketitle
+
+\tableofcontents
 
@@
 
+\subsection{Results: Japanese}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-it.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-memsurp.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-auc-hist.pdf}
+\end{center}
+	\caption{Japanese verb suffixes, measuring prediction on the level of phonemes, for real (blue), random (green), and approximately optimized (red) orderings.
+Left: $I_t$ as a function of $t$.
+Center: Memory-surprisal tradeoff.
+Right: Areas under the curve for the memory-surprisal tradeoff.}\label{fig:jap-phon}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-it.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-memsurp.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-auc-hist.pdf}
+\end{center}
+	\caption{Japanese verb suffixes, measuring prediction on the level of morphemes, for real (blue), random (green), and approximately optimized (red) orderings.
+Left: $I_t$ as a function of $t$.
+Center: Memory-surprisal tradeoff.
+Right: Areas under the curve for the memory-surprisal tradeoff.}\label{fig:jap-morph}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\begin{tabular}{c||llll}
+             &       Pairs & Full \\ \hline\hline
+Optimized for Phoneme Prediction   &   0.976 (SD 0.007) & 0.971 (SD 0.011) \\
+Optimized for Morpheme Prediction  &   0.873 (SD 0.154) & 0.85 (SD 0.184) \\
+Random Baseline     &  0.519 (SD 0.177) & 0.559 (SD 0.202) \\
+\end{tabular}
+\end{center}
+\caption{Accuracy of approximately optimized orderings, and of random baseline orderings, in predicting verb suffix order in Japanese.
+`Pairs' denotes the rate of pairs of morphemes that are ordered correctly, and `Full' denotes the rate of verb forms where order is predicted entirely correctly.
+We show means and standard deviations over different runs of the optimization algorithm (`Optimized'), and over different random orderings (`Random').}\label{fig:acc-japanese}
+\end{figure}
+
+\subsection{Results: Sesotho}
+
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-it.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-memsurp.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-auc-hist.pdf}
+\end{center}
+	\caption{Sesotho verb affixes, measuring prediction on the level of phonemes, for real (blue), random (green), and approximately optimized (red) orderings.
+Left: $I_t$ as a function of $t$.
+Center: Memory-surprisal tradeoff.
+Right: Areas under the curve for the memory-surprisal tradeoff.}\label{fig:jap-phon}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-it.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-memsurp.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-auc-hist.pdf}
+\end{center}
+	\caption{Sesotho verb affixes, measuring prediction on the level of morphemes, for real (blue), random (green), and approximately optimized (red) orderings.
+Left: $I_t$ as a function of $t$.
+Center: Memory-surprisal tradeoff.
+Right: Areas under the curve for the memory-surprisal tradeoff.}\label{fig:jap-morph}
+\end{figure}
+
+%cp ~/memory-surprisal/code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/* figures/
+%cp ~/memory-surprisal/code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/* figures/
+
+\begin{figure}
+\begin{tabular}{cc||ll|ll}
+             &              & \multicolumn{2}{c}{Prefixes}    & \multicolumn{2}{|c}{Suffixes} \\
+             &              & Pairs & Full & Pairs & Full \\ \hline\hline
+             Phonemes   &   Optimized  &  0.987 (SD 0.004) & 0.983 (SD 0.006) & 0.914 (SD 0.131) & 0.906 (SD 0.14) \\
+           &   Random  &  0.468 (SD 0.321) & 0.364 (SD 0.353) & 0.459 (SD 0.224) & 0.409 (SD 0.235) \\ \hline
+Morphemes  &   Optimized  &  0.997 (SD 0.0) & 0.995 (SD 0.0) & 0.804 (SD 0.0) & 0.721 (SD 0.0) \\
+           &   Random  &  0.521 (SD 0.331) & 0.43 (SD 0.349) & 0.468 (SD 0.233) & 0.401 (SD 0.251) \\
+\end{tabular}
+\caption{Accuracy of approximately optimized orderings, and of random baseline orderings, in predicting verb affix order in Sesotho.
+`Pairs' denotes the rate of pairs of morphemes that are ordered correctly, and `Full' denotes the rate of verb forms where order is predicted entirely correctly.
+We show means and standard deviations over different runs of the optimization algorithm (`Optimized'), and over different random orderings (`Random').}\label{fig:acc-sesotho}
+\end{figure}
+
 \section{Some theoretical thoughts (scratch area)}

### 2020-03-29 writeup/si.tex

@@
 		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-auc-hist.pdf}
+
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-it-heldout.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-memsurp-heldout.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byPhonemes-auc-hist-heldout.pdf}
 \end{center}
@@
 		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-auc-hist.pdf}
+
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-it-heldout.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-memsurp-heldout.pdf}
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Japanese/tradeoffs/figures/Japanese-suffixes-byMorphemes-auc-hist-heldout.pdf}
 \end{center}
@@
 Center: Memory-surprisal tradeoff.
-Right: Areas under the curve for the memory-surprisal tradeoff.}\label{fig:jap-morph}
+Right: Areas under the curve for the memory-surprisal tradeoff.
+Top: Simple estimator, Bottom: Heldout estimator}\label{fig:jap-morph}
 \end{figure}

### 2020-03-30 writeup/si.tex
code

@@
 \begin{center}
-\begin{tabular}{c||llll}
-             &       Pairs & Full \\ \hline\hline
-Optimized for Phoneme Prediction   &   0.976 (SD 0.007) & 0.971 (SD 0.011) \\
-Optimized for Morpheme Prediction  &   0.873 (SD 0.154) & 0.85 (SD 0.184) \\
-Random Baseline     &  0.519 (SD 0.177) & 0.559 (SD 0.202) \\
+\begin{tabular}{ccc||llll}
+	& &              &       Pairs & Full \\ \hline\hline
+	Tokens &	Simple & Optimized for Phoneme Prediction   &   0.976 (SD 0.006) & 0.972 (SD 0.01) \\
+	&	& Optimized for Morpheme Prediction  &   0.93 (SD 0.01) & 0.919 (SD 0.008) \\ \hline
+	&	Heldout & Optimized for Phoneme Prediction   &   0.962 (SD 0.001) & 0.957 (SD 0.002) \\
+	&	& Optimized for Morpheme Prediction  &   0.954 (SD 0.009) & 0.945 (SD 0.012) \\ \hline
+	&	&	Random Baseline    &  0.504 (SD 0.0) & 0.414 (SD 0.0) \\ \hline \hline
+	Types & Simple & Optimized for Phoneme Prediction   &   0.968 (SD 0.006) & 0.961 (SD 0.011) \\
+	& & Optimized for Morpheme Prediction  &   0.903 (SD 0.014) & 0.883 (SD 0.011) \\ \hline
+	& Heldout & Optimized for Phoneme Prediction   &   0.946 (SD 0.002) & 0.937 (SD 0.002) \\
+	& & Optimized for Morpheme Prediction  &   0.938 (SD 0.011) & 0.922 (SD 0.014) \\ \hline
+	& & Random Baseline    &  0.482 (SD 0.0) & 0.379 (SD 0.0) \\
 \end{tabular}
@@
 `Pairs' denotes the rate of pairs of morphemes that are ordered correctly, and `Full' denotes the rate of verb forms where order is predicted entirely correctly.
-We show means and standard deviations over different runs of the optimization algorithm (`Optimized'), and over different random orderings (`Random').}\label{fig:acc-japanese}
+We show means and standard deviations over different runs of the optimization algorithm (`Optimized'), and over different random orderings (`Random').
+`Tokens' resultsare obtained by counting each form by the number of occurrences in the data set; `Types' results count each form only once.
+`Simple' models are optimized for in-sample AUC, `Heldout' models are optimized for heldout AUC.}\label{fig:acc-japanese}
 \end{figure}
@@
 \includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-auc-hist.pdf}
+
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-it-heldout.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-memsurp-heldout.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byPhonemes-auc-hist-heldout.pdf}
 \end{center}
@@
 \includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-auc-hist.pdf}
+
+		\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-it-heldout.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-memsurp-heldout.pdf}
+\includegraphics[width=0.3\textwidth]{../code/ngram-control/create_models_ngrams/morph/Sesotho/tradeoffs/figures/Sesotho-suffixes-byMorphemes-auc-hist-heldout.pdf}
 \end{center}
@@
 \begin{figure}
-\begin{tabular}{cc||ll|ll}
-             &              & \multicolumn{2}{c}{Prefixes}    & \multicolumn{2}{|c}{Suffixes} \\
-             &              & Pairs & Full & Pairs & Full \\ \hline\hline
-             Phonemes   &   Optimized  &  0.987 (SD 0.004) & 0.983 (SD 0.006) & 0.914 (SD 0.131) & 0.906 (SD 0.14) \\
-           &   Random  &  0.468 (SD 0.321) & 0.364 (SD 0.353) & 0.459 (SD 0.224) & 0.409 (SD 0.235) \\ \hline
-Morphemes  &   Optimized  &  0.997 (SD 0.0) & 0.995 (SD 0.0) & 0.804 (SD 0.0) & 0.721 (SD 0.0) \\
-           &   Random  &  0.521 (SD 0.331) & 0.43 (SD 0.349) & 0.468 (SD 0.233) & 0.401 (SD 0.251) \\
+\begin{tabular}{cccc||ll|ll}
+	&& &              & \multicolumn{2}{c}{Prefixes}    & \multicolumn{2}{|c}{Suffixes} \\
+	  &&   &              & Pairs & Full & Pairs & Full \\ \hline\hline
+	Tok. & Simple & Phon.   &   Opt.  &  0.987 (SD 0.004) & 0.983 (SD 0.006) & 0.914 (SD 0.131) & 0.906 (SD 0.14) \\
+	& &           &   Rand.  &  0.322 (SD 0.253) & 0.195 (SD 0.24) & 0.588 (SD 0.155) & 0.554 (SD 0.167) \\
+	& & Morph.  &   Opt.  &  0.997 (SD 0.0) & 0.995 (SD 0.0) & 0.804 (SD 0.0) & 0.721 (SD 0.0) \\
+	& &           &   Rand.  &  0.672 (SD 0.305) & 0.604 (SD 0.338) & 0.423 (SD 0.204) & 0.332 (SD 0.211) \\ \hline
+	&	Heldout & Phon.   &   Opt.  &  0.993 (SD 0.0) & 0.989 (SD 0.0) & 0.792 (SD 0.102) & 0.734 (SD 0.13) \\
+	 &&  &   Rand.  &  0.322 (SD 0.253) & 0.195 (SD 0.24) & 0.588 (SD 0.155) & 0.554 (SD 0.167) \\
+	&&Morph.  &   Opt.  &  0.988 (SD 0.0) & 0.989 (SD 0.0) & 0.756 (SD 0.014) & 0.676 (SD 0.017) \\
+	  && &   Rand.  &  0.672 (SD 0.305) & 0.604 (SD 0.338) & 0.423 (SD 0.204) & 0.332 (SD 0.211) \\ \hline\hline
+	Typ. & Simple & Phon.   &   Opt.  &  0.979 (SD 0.006) & 0.972 (SD 0.009) & 0.92 (SD 0.113) & 0.912 (SD 0.116) \\
+	& &           &   Rand.  &  0.338 (SD 0.248) & 0.205 (SD 0.219) & 0.592 (SD 0.137) & 0.55 (SD 0.149) \\
+	& & Morph.  &   Opt.  &  0.994 (SD 0.0) & 0.99 (SD 0.0) & 0.841 (SD 0.0) & 0.754 (SD 0.0) \\
+	& &           &   Rand.  &  0.666 (SD 0.302) & 0.588 (SD 0.329) & 0.438 (SD 0.213) & 0.329 (SD 0.22) \\ \hline
+	&Heldout&Phon.   &   Opt.  &  0.988 (SD 0.0) & 0.981 (SD 0.0) & 0.831 (SD 0.079) & 0.77 (SD 0.107) \\
+	&&           &   Rand.  &  0.338 (SD 0.248) & 0.205 (SD 0.219) & 0.592 (SD 0.137) & 0.55 (SD 0.149) \\
+	&&Morph.  &   Opt.  &  0.98 (SD 0.0) & 0.98 (SD 0.0) & 0.783 (SD 0.021) & 0.698 (SD 0.025) \\
+	&&           &   Rand.  &  0.666 (SD 0.302) & 0.588 (SD 0.329) & 0.438 (SD 0.213) & 0.329 (SD 0.22) \\
 \end{tabular}
@@
 `Pairs' denotes the rate of pairs of morphemes that are ordered correctly, and `Full' denotes the rate of verb forms where order is predicted entirely correctly.
-We show means and standard deviations over different runs of the optimization algorithm (`Optimized'), and over different random orderings (`Random').}\label{fig:acc-sesotho}
+We show means and standard deviations over different runs of the optimization algorithm (`Opt.'), and over different random orderings (`Random').
+`Tokens' resultsare obtained by counting each form by the number of occurrences in the data set; `Types' results count each form only once.
+`Simple' models are optimized for in-sample AUC, `Heldout' models are optimized for heldout AUC.}\label{fig:acc-sesotho}
 \end{figure}

### 2020-05-01 code/preparePaper/output/medians_small_pcfg.tex
code

@@
-Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
+Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
-Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
+Breton & Bulgarian & Buryat & Cantonese & Catalan & Chinese
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
-Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
+Croatian & Czech & Danish & Dutch & English & Erzya
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
-Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
+Estonian & Faroese & Finnish & French & German & Greek
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
-North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
+Hebrew & Hindi & Hungarian & Indonesian & Italian & Japanese
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
-Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
+Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
+North Sami & Norwegian & Persian & Polish & Portuguese & Romanian
+ \\ 
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+ \\ 
+Russian & Serbian & Slovak & Slovenian & Spanish & Swedish
+ \\ 
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+ \\ 
+Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
+ \\ 
+\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+ \\

### 2020-05-01 code/analysis/medians/median-CI/yStudyTradeoff_Bootstrap_Parallel_OnlyWordForms_BoundedVocab_BinomialTest_Single_MedianCI.py

@@ -65,14 +65,24 @@ maximalMemory = 0
 for typ, mis in misByType.iteritems():
   mask = torch.FloatTensor([[1 if j <= i else 0 for j in range(19)] for i in range(19)])
   cumulativeMI = torch.matmul(mis, mask.t())
-  #print("MIs", mis[0])
-  #print("Cum MI", cumulativeMI[0])
+  print(typ, mis)
+  
+  print("MIs", mis[0])
+  print("Cum MI", cumulativeMI[0])
   cumulativeMemory = torch.matmul(distance*mis, mask.t())
-  #print("Cum Mem", cumulativeMemory[0])
+  print("Cum Mem", cumulativeMemory[0])
   cumMIs[typ] = cumulativeMI
   cumMems[typ] = cumulativeMemory
+  for line1, line2 in zip(mis, cumulativeMemory):
+      print("--")
+      print(line1)
+      print(line2)
+      if float(torch.max(line2)) > 10:
+          break
   maximalMemory = max(maximalMemory, float(torch.max(cumulativeMemory)))
-
+  print(maximalMemory)
+  assert maximalMemory < 10
+assert False
 #print("MAXIMAL MEMORY", maximalMemory)
 
 

### 2020-05-02 code/preparePaper/output/auc_small_MLE.tex

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Afrikaans-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Amharic-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Arabic-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Armenian-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Bambara-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Basque-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Breton-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Bulgarian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Buryat-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Cantonese-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Catalan-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Chinese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Croatian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Czech-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Danish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Dutch-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/English-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Erzya-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Estonian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Faroese-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Finnish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/French-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/German-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Greek-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hebrew-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hindi-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Hungarian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Indonesian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Italian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Japanese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Kazakh-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Korean-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Kurmanji-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Latvian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Maltese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Naija-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/North_Sami-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Norwegian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Persian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Polish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Portuguese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Romanian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Russian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Serbian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Slovak-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Slovenian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Spanish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Swedish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Thai-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Turkish-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Ukrainian-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Urdu-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Uyghur-Adap-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/medians/auc/figures/Vietnamese-listener-surprisal-memory-HIST_AUC_onlyWordForms_boundedVocab.pdf}
 \\

### 2020-05-10 writeup/control-ngrams.tex
code

@@
-\subsection{Method}
-We use a version of Kneser-Ney Smoothing.
+Here we show that the results of Study 2 remain robust when estimating surprisal with a simple n-gram model instead of recurrent neural networks.
+
+\subsubsection{Method}
+We use a version of Kneser-Ney Smoothing \citep{kneser-improved-1995}.
 For a sequence $w_1\dots w_k$, let $N(w_{1\dots k})$ be the number of times $w_{1\dots k}$ occurs in the training set.
@@
 where $\alpha \in [0,1]$ is also a hyperparameter.
-(CITE) show that this definition results in a well-defined probability distribution, i.e., $\sum_{w \in V} p_t(w|w_{0 \dots t-1}) = 1$.
+\citet{kneser-improved-1995} show that this definition results in a well-defined probability distribution, i.e., $\sum_{w \in V} p_t(w|w_{0 \dots t-1}) = 1$.
 
@@
 
-Hyperparameters $\alpha, \gamma, \delta$ are tuned with the same strategy as for the neural network models.
+Hyperparameters $\alpha, \gamma, \delta$ are tuned using the held-out set, with the same strategy as for the neural network models.
 
-\subsection{Results}
+\subsubsection{Results}
 
@@
 \end{longtable}
-	\captionof{figure}{Medians (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
+	\captionof{figure}{Memory-surprisal tradeoff curves (estimated using n-gram models): For each memory budget, we provide the median surprisal for real and random languages.
 Solid lines indicate sample medians for ngrams, dashed lines indicate 95 \% confidence intervals for the population median.

### 2020-05-10 writeup/compile.sh

@@ -1,0 +1,10 @@
+pdflatex auc-table_MLE.tex
+pdflatex auc-table.tex
+pdflatex it-table.tex
+pdflatex langs-figure.tex
+pdflatex quantiles-table.tex
+pdflatex results-table-mle.tex
+pdflatex results-table-pcfg.tex
+pdflatex results-table.tex
+pdflatex results-table-word-level.tex
+

### 2020-05-23 code/preparePaper/output/it_small_mle.tex

Afrikaans & Amharic & Arabic & Armenian & Bambara & Basque & Breton & Bulgarian & Buryat
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Afrikaans-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Amharic-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Arabic-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Armenian-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bambara-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Basque-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Breton-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Bulgarian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Buryat-Adap-it_mle.pdf}
 \\ 
Cantonese & Catalan & Chinese & Croatian & Czech & Danish & Dutch & English & Erzya
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Cantonese-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Catalan-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Chinese-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Croatian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Czech-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Danish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Dutch-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/English-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Erzya-Adap-it_mle.pdf}
 \\ 
Estonian & Faroese & Finnish & French & German & Greek & Hebrew & Hindi & Hungarian
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Estonian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Faroese-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Finnish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/French-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/German-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Greek-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hebrew-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hindi-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Hungarian-it_mle.pdf}
 \\ 
Indonesian & Italian & Japanese & Kazakh & Korean & Kurmanji & Latvian & Maltese & Naija
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Indonesian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Italian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Japanese-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kazakh-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Korean-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Kurmanji-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Latvian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Maltese-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Naija-Adap-it_mle.pdf}
 \\ 
North Sami & Norwegian & Persian & Polish & Portuguese & Romanian & Russian & Serbian & Slovak
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/North_Sami-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Norwegian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Persian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Polish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Portuguese-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Romanian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Russian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Serbian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovak-it_mle.pdf}
 \\ 
Slovenian & Spanish & Swedish & Thai & Turkish & Ukrainian & Urdu & Uyghur & Vietnamese
 \\ 
\includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Slovenian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Spanish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Swedish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Thai-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Turkish-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Ukrainian-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Urdu-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Uyghur-Adap-it_mle.pdf} & \includegraphics[width=0.1\textwidth]{../code/analysis/visualize_neural/figures/Vietnamese-it_mle.pdf}
 \\

### 2020-05-23 code/preparePaper/output/medians_small_pcfg.tex

@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Afrikaans-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Amharic-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Arabic-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Armenian-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Bambara-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Basque-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Breton-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Bulgarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Buryat-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Cantonese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Catalan-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Chinese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Croatian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Czech-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Danish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Dutch-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/English-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Erzya-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Estonian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Faroese-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Finnish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/French-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/German-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Greek-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Hebrew-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Hindi-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Hungarian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Indonesian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Italian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Japanese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Kazakh-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Korean-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Kurmanji-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Latvian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Maltese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Naija-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/North_Sami-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Norwegian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Persian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Polish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Portuguese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Romanian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Russian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Serbian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Slovak-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Slovenian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Spanish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Swedish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\ 
@@
  \\ 
-\includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg/analyze_pcfg/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
+\includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Thai-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Turkish-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Ukrainian-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Urdu-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Uyghur-Adap-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf} & \includegraphics[width=0.1\textwidth]{../code/pcfg-control/analyze_pcfg/figures/Vietnamese-listener-surprisal-memory-MEDIANS_onlyWordForms_boundedVocab-pcfg.pdf}
  \\
